### 2015-05-18 Harris_2015/create_data.Rmd
Add Harris_2015 analyses

# Setup

```{r, cache = TRUE}
devtools::load_all()
species_range = c(5, 10, 20)        # Numbers of species to evaluate
site_range = c(20, 100, 500, 2500)  # Numbers of sites to evaluate
reps = 24                           # Number of times to repeat combination of sites and species
min_gibbs = 1000                    # Always perform at least this many Gibbs samples

n_env = 2                           # Number of environmental variables in the model
```

# Functions

A function for simulating known $\alpha$ and $\beta$ coefficients.

```{r}
make_coefficients = function(n_spp){

  # Most of the n-choose-2 interactions are assumed to be weak; a few
  # are much stronger.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

  # On average, three-quarters of the simulated interactions are
  # negative.
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(.75, .25),
    replace = TRUE
  )

  # Alpha coefficients are normally distributed
  a = rnorm(n_spp, -1)

  # rosalia stores betas in the upper triangle of a matrix and alphas
  # along the diagonal.
  c(upper = b, diag = a)
}
```

# Simulate and save the simulated data

```{r, cache = TRUE}
for(n_spp in species_range){
  cat("\n", n_spp, "\n")

  for(n_sites in site_range){
    cat("\n")
    for(rep in 1:reps){
      cat(".")

      # Make coefficients using the function above
      par = make_coefficients(n_spp)

      # Save the "true" betas
      truth = par[1:choose(n_spp, 2)]

      # The "true" alphas will only be used during simulation
      alpha = par[-(1:choose(n_spp, 2))]

      # Start beta empty, fill in upper triangle, then fill in lower tri with its transpose
      # to make a symmetric matrix of beta coefficients
      beta = matrix(0, n_spp, n_spp)
      beta[upper.tri(beta)] = truth
      beta = beta + t(beta)

      # Environmental states, simulated from a Gaussian for each site and variable
      env = matrix(rnorm(n_sites * n_env), ncol = n_env)

      # Species only respond to the environment in the second half of the simulations
      sd = 2 * (rep > reps/2)
      alpha_env = matrix(rnorm(n_spp * n_env, sd = sd), nrow = n_env)

      # Simulate the landscape from known process
      i = 0 # Gibbs iteration counter starts at zero
      give_up = FALSE # don't give up simulating unless needed
      incomplete = TRUE # Sampling isn't complete yet

      # Landscape starts as if betas were all zero
      x = matrix(
        plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
        nrow = n_sites, 
        ncol = n_spp
      )

      # Gibbs sampling until enough iterations have passed *and* no species are totally absent
      while(incomplete & !give_up){
        i = i + 1
        for(j in 1:n_spp){
          x[,j] = rbinom(
            nrow(x),
            size = 1,
            prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
          )
        }

        # Simulation is "incomplete"" if insufficient Monte Carlo samples have been collected
        # or if one or more species are missing from the landscape.
        incomplete = i < min_gibbs | min(colSums(x)) == 0

        # If, after 100*minimum iterations, no valid landscapes have been produced, give
        # up on this set of coefficients.
        give_up = i > min_gibbs * 100 
      }

      if(give_up){
        print("giving up")
        next
      }else{
        file_stem = paste(n_spp, n_sites, rep, sep = "-")

        # Save the simulated landscape in the fakedata folder
        write.csv(
          x, 
          file = paste0("fakedata/matrices/", file_stem, ".csv")
        )

        # Ulrich's "Pairs" software rejects empty ("degenerate") sites
        x_subset = x[rowSums(x) != 0, colSums(x) != 0]

        # "Pairs" expects the data to be transposed relative to the other three
        # methods' inputs.
        write.table(
          t(x_subset), 
          file = paste0("fakedata/matrices/", file_stem, "-transposed.txt"), 
          quote = FALSE
        )

        # save the "true" beta values
        write(
          truth, 
          file = paste0("fakedata/truths/", file_stem, ".txt"), 
          ncolumns = length(truth)
        )
      }
    }
  }
}
```

# Housekeeping

In order to run the Pairs analyses more quickly, I moved the simulated data matrices
into 8 folders to run simultaneously on 8 processor cores.

```{r}
transposed_files = dir("fakedata/matrices", pattern = "transposed", full.names = FALSE)

rep_num = as.integer(gsub(".*-.*-(.*)-transposed.txt", "\\1", transposed_files))

core_num = (rep_num %%8) + 1
for(i in 1:8){
  dir.create(paste0("fakedata/matrices/", i), showWarnings = FALSE)
}

file.rename(
  paste0("fakedata/matrices/", transposed_files), 
  paste0("fakedata/matrices/", core_num, "/", transposed_files)
)

for(i in 1:8){
  write(
    paste0(
      c("comment", sample(dir(paste0("fakedata/matrices/", i), pattern = "transposed"))), 
      collapse = "\r\n"
    ), 
    file = paste0("fakedata/matrices/", i, "/", "list.txt")
  )
}

# Include the .exe file in each folder so it can run on the corresponding files
for(i in 1:8){
  file.copy("fakedata/matrices/pairs.exe", paste0("fakedata/matrices/", i, "/", "pairs.exe"))
}
```

### 2015-05-18 Harris_2015/models.Rmd
Add Harris_2015 analyses

---
output: html_document
---
# Simulating landscapes from known coefficients

The following function draws random coefficients for a Markov network of a pre-specified size.
The probability distribution was chosen so that most species pairs interact weakly, a few
act strongly, and that most of the interactions were negative rather than positive.

```{r, cache=TRUE}
make_coefficients = function(n_spp){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

  # Species interactions are negative 3/4 of the time in 
  # these simulations.
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(.75, .25),
    replace = TRUE
  )

  # Species' intercepts are normally distributed
  a = rnorm(n_spp, -1)

  # Return the simulated values.
  # The rosalia function stores pairwise parameters in the upper
  # triangle of an n-by-n matrix and stores the species intercepts
  # along the diagonal, so these values are named accordingly.
  c(upper = b, diag = a)
}
```

I used the following function to generate "true" parameters with the method above, 
simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo,
and save the results to a "fakedata" folder.

```{r, cache = TRUE}
library(rosalia)

simulate_data = function(n_spp, n_sites, n_rep, min_gibbs = 1000, reps = 24, n_env = 2){

  # Determine the "true" parameters for the simulated assemblage
  par = make_coefficients(n_spp)

  # "True" interaction strengths, to save for later
  truth = par[1:choose(n_spp, 2)]

  # "True" intercepts, possibly adjusted below by environment
  alpha = par[-(1:choose(n_spp, 2))]

  # Turn the interaction values into an n-by-n matrix
  # Start with empty matrix; fill in upper triangle; 
  # then fill in lower triangle with its transpose
  beta = matrix(0, n_spp, n_spp)
  beta[upper.tri(beta)] = truth
  beta = beta + t(beta)

  # Environmental states are normally distributed
  env = matrix(rnorm(n_sites * n_env), ncol = n_env)

  # Species only respond to the environment in the second half of the simulations
  # (i.e. their responses are sampled from a Gaussian with a mean and standard deviation 
  # of zero in the first half)
  sd = 2 * (rep > reps/2)
  alpha_env = matrix(rnorm(n_spp * n_env, sd = sd), nrow = n_env)

  # Simulate the landscape from known process with Gibbs sampling
  i = 0 # Gibbs iteration counter starts at zero
  give_up = FALSE # don't give up simulating unless things run for too long
  incomplete = TRUE # Sampling isn't complete yet

  # Landscape starts as if betas were all zero
  x = matrix(
    plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
    nrow = n_sites, 
    ncol = n_spp
  )

  # Gibbs sampling until enough iterations have passed *and* no species are totally absent
  while(incomplete & !give_up){
    i = i + 1
    for(j in 1:n_spp){
      x[,j] = rbinom(
        nrow(x),
        size = 1,
        prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
      )
    }

    # If fewer than min_gibbs iterations have passed, or if a species is completely absent,
    # then keep sampling
    incomplete = i < min_gibbs | min(colSums(x)) == 0

    # But give up sampling if the sampling goes on 100x longer than the minimum
    give_up = i > min_gibbs * 100
  }

  if(give_up){
    print("giving up")
    next
  }else{
    # Save the results in a "fake data" folder

    file_stem = paste(n_spp, n_sites, rep, sep = "-")

    # Save the matrix of presence/absence observations
    write.csv(
      x, 
      file = paste0("fakedata/matrices/", file_stem, ".csv")
    )

    # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
    x_subset = x[rowSums(x) != 0, colSums(x) != 0]

    # Gotelli and Ulrich's Pairs method expects the data matrices to be transposed,
    # So I save them separately
    write.table(
      t(x_subset), 
      file = paste0("fakedata/matrices/", file_stem, "-transposed.txt"), 
      quote = FALSE
    )

    # Save the "true" species interactions
    write(
      truth, 
      file = paste0("fakedata/truths/", file_stem, ".txt"), 
      ncolumns = length(truth)
    )
  }
}
```

# Model fitting

## R-compatible methods

Three of the four methods are described in this subsection.  The "Pairs" software
needs to be analyzed outise of R and is described below.

### Rosalia

The `rosalia` function finds (penalized) maximum likelihood estimates for the
$\alpha$ and $\beta$ parameters in a Markov network using the `BFGS` method of
the `optim` function for convex optimization.  For these analyses,
I penalized the likelihood with logistic priors on both sets of parameters to stabilize
the model's estimates. In order to ensure convergence, I increased the number of
`BFGS` iterations from 100 (the default) to 1000.

The following function returns a vector of $\beta$ coefficients.

```{r, cache=TRUE, eval = FALSE}
library("rosalia")

fit_rosalia = function(x){
  fit = rosalia(
    x = x, 
    prior = make_logistic_prior(scale = 1), 
    maxit = 1E3
  )

  # Return the upper triangle of the interaction matrix,
  # which has all the beta values.
  relist(fit$par)$upper
}

```

### Partial covariance

The partial covariance is normally computed as the matrix inverse of
the covariance times `-1`.  The `ridgeS` function adds a regularization parameter,
`lambda`, that stabilizes the estimates and makes it possible to 
calculate a partial covariance matrix even when two species are perfectly 
correlated in the observed data.

```{r, cache=TRUE}
fit_partial = function(x){
  library("rags2ridges") # rags2ridges can be installed from CRAN with 
                         # install.packages("rags2ridges")

  # The effective size of the penalty should shrink as more data becomes available
  # and overwhelms it
  fit = -ridgeS(cov(x), lambda = 1 / (5 * nrow(x)))

  # all pairwise estimtes can be found in the upper triangle of the 
  # partial covariance matrix.
  fit[upper.tri(fit)]
}

```

### Sample covariance

The sample covariance provides a simpler estimate of species' interaction strengths.

```{r, cache=TRUE}
fit_cov = function(x){
  fit = cov(x)
  fit[upper.tri(fit)]
}
```

## Analyzing simulated landscapes with the first three models

This function takes a quoted model name (`"cov"`, `"rosalia"`, or `"partial"`) and applies the
corresponding function above to all the matrices in the `fakedata` folder.

It returns a data frame with information about each species pair's simulated landscape,
the "true" interaction strength between the pair, and a "predicted" value from the 
fitted model.

```{r, cache = TRUE}
fit_model = function(method_name, csv_names){
  fit = get(paste("fit", method_name, sep = "_"))
  out_list = lapply(
    csv_names, 
    function(x){
      file_info = as.integer(strsplit(x, "-|\\.csv")[[1]])

      full_path = paste0("fakedata/matrices/", x)

      truth = read.table(
        paste0(
          "fakedata/truths/",
          paste(file_info, collapse = "-"),
          ".txt"
        )
      )

      data.frame(
        n_spp = file_info[1], 
        n_loc = file_info[2],
        n_rep = file_info[3],
        truth = unlist(truth),
        predicted = fit(read.csv(full_path, row.names = 1)),
        method = method_name,
        stringsAsFactors = FALSE
      )
    }
  )
  do.call(rbind, out_list)
}
```

```{r, cache= TRUE}
cov_results = fit_model("cov", dir("fakedata/matrices"))
```

```{r, cache = TRUE}
partial_results = fit_model("partial", dir("fakedata/matrices"))
```

## Settings for fitting a Pairs model

The Pairs program, written by Dr. Werner Ulrich, can be downloaded from 
[his web site](http://www.keib.umk.pl/pairs/?lang=en) as a Windows executable, but no source
code or scripts are available to include here.

I ran the software with the following options:

* Batch mode
* Sequential swap ("s")
* Printing all pairs ("y")
* C-score co-occurrence measure ("c")
* Default confidence limits (0.05)
* Default iterations (100)
* Maximum of 20 species

## Code for parsing the results from Pairs

The Pairs software produces a text file with two tables of results for each data matrix.

For the analyses in this paper, I needed to match the (negative) Z-scores associated with each
species pair with the "true" interaction parameters used to simulate the corresponding
data matrix.  I used the following code to do so.

The Z-scores, (which are based on pairwise C-scores), have the opposite sign of the other metrics
assessed here, so I multiplied them by negative one after importing them.

```{r, cache=TRUE}

raw = readLines("Pairs-results.txt")

# Pull out the file ID (a sequence of digits separated by dashes) 
# from the lines that begin with ">"
ids = gsub("^.[^0-9]*([0-9]*-[0-9]*-[0-9]*).*$", "\\1", grep("^>", raw, value = TRUE))

splitted_ids = strsplit(ids, "-")

# All the tables in `raw` have "MeanScore" in their headers, but I don't
# want the tables that have "ObsNumber" in them.
header_rows = grep("MeanScore", raw)
wrong_header_rows = grep("ObsNumber", raw)

# Find the start and ending lines for the data tables of interest
starts = header_rows + 1
ends = c(
  # The table of interest ends 3 lines before the next table starts
  wrong_header_rows[-1] - 3,
  length(raw)
)

# Loop through the tables and pull out negative Z-scores for each
# species pair, then line them up with the "truth" values.
# Finally, concatenate the results into a giant data.frame
pairs_results = do.call(
  rbind,
  lapply(
    1:length(ids), 
    function(i){

      # Split the table by whitespace
      split_table = strsplit(raw[starts[[i]]:ends[[i]]], "\\s+")

      # Z-scores are the 14th element of each splitted row
      z_score = sapply(split_table, function(x) as.numeric(x[[14]]))

      # The first species' ID is in the third element of each splitted row
      species_1 = sapply(split_table, function(x) as.numeric(x[[3]]))

      # The second species' ID is in the fourth element of each splitted row
      species_2 = sapply(split_table, function(x) as.numeric(x[[4]]))

      truth = as.numeric(
        strsplit(
          readLines(
            paste0(c("fakedata/truths/", ids[i], ".txt"), collapse = "")
          ),
          " "
        )[[1]]
      )

      # The number of species is the first element of each id
      n_spp = as.numeric(splitted_ids[[i]][1])
      n_loc = splitted_ids[[i]][2] # Number of locations is second
      n_rep = splitted_ids[[i]][3] # Replicate number locations is second

      truthmat = matrix(0, n_spp, n_spp)
      truthmat[upper.tri(truthmat)] = truth
      truthmat = truthmat + t(truthmat)

      # Output a data.frame with the "true" parameters, the negative Z-scores
      # and information about the simulated landscape
      data.frame(
        n_spp = as.integer(n_spp), 
        n_loc = as.integer(n_loc),
        n_rep = as.integer(n_rep),
        truth = truthmat[cbind(species_1, species_2)],
        predicted = -z_score,
        method = "Null",
        stringsAsFactors = FALSE
      )
    }
  )
)

```

As noted in the manuscript, I manually altered one Z-score produced by the Pairs software because
it was such a strong outlier that it completely dominated the regression model described below.

```{r, cache = TRUE}
# The largest predicted Z-score was over 1000; the next largest was about 32
head(sort(pairs_results$predicted, decreasing = TRUE))
pairs_results$predicted[which.max(pairs_results$predicted)] = 32.47
```

# Model evaluation

For each method, I fit a linear regression model through the origin in order to put the different
models' estimates on a common scale.  
In these regressions, the weights associated with each species pair in the regression model are 
inversely proportional to the number of species in a given landscape.  In this way, I ensured that 
each landscape (as opposed to each species pair) contributed equally to the regression.

For each species pair, I recorded the residuals between this regression estimate and the "true"
interaction strength used to simulate the original data.

```{r, cache = TRUE}

results = do.call(
  rbind,
  lapply(
    list(cov_results, partial_results, pairs_results),
    function(x){
      residuals = residuals(
        lm(
          truth ~ predicted + 0, 
          data = x, 
          weights = 1/choose(n_spp, 2),
          y = TRUE
        )
      )
      cbind(x, residuals = residuals)
    }
  )
)

# The second half of the simulated landscapes had an environmental component
results$env = results$n_rep > (max(results$n_rep) / 2)

```

I then calculated $R^2$ for each method on each type of landscape (i.e. landscapes with each level
of richness, each number of sites, and presence/absence of environmental variables)

method as one minus the sum of squared residuals divided by
the squared deviations from zero (e.g. `1 - sum(pairs_fit$residuals^2) / sum(pairs_fit$truth^2)` for
the `pairs_results`).

```{r, cache = TRUE}
library(dplyr)

summarized_results = results %>% group_by_("n_spp", "n_loc", "method", "env") %>% 
  do_(
    ~data.frame(r2 = 1 - sum(.$residuals^2) / sum(.$truth^2))
  )

head(summarized_results) # Show the first 6 rows
```

```{r, cache = FALSE}
library(ggplot2)

basic_plot = ggplot(aes(n_loc, r2, color = method, shape = method), data = summarized_results) + 
  geom_line(size = .5) + 
  geom_point(size = 2, fill = "white") + 
  facet_grid(env~n_spp) + 
  geom_hline(yintercept = 0, size = 1/2) +
  geom_vline(xintercept = 0, size = 1)

scaled_plot = basic_plot +
  scale_shape_manual(values = c(21, 25, 15, 18)) + 
  coord_cartesian(ylim = c(-0.1, 1.01)) + 
  ylab(expression(R^2)) + 
  xlab("Number of sites (log scale)") + 
  scale_x_log10(breaks = c(20, 100, 500, 2500), limits = c(20, 2.5E3)) +
  scale_color_manual(values = c("red3", "purple2", "sienna1", "mediumblue"))

scaled_plot + theme_bw(base_size = 11) + 
  theme(panel.margin = grid::unit(1.25, "lines")) + 
  theme(panel.border = element_blank(), axis.line = element_blank()) + 
  theme(
    panel.grid.minor = element_blank(), 
    panel.grid.major.y = element_line(color = "lightgray", size = 1/4), 
    panel.grid.major.x = element_blank()
  ) + 
  theme(strip.background = element_blank(), legend.key = element_blank()) + 
  theme(plot.margin = grid::unit(c(.01, .01, .75, .1), "lines")) + 
  #theme(legend.text = element_text(size = 11)) + 
  theme(
    axis.title.x = element_text(vjust = -0.2, size = 14), 
    axis.title.y = element_text(angle = 0, hjust = -.1, size = 14)
  )

```

### 2015-05-18 R/simulate_data.R
add simulation function

@@ -1,0 +1,20 @@
+simulate_data = function(par, possibilities, possible_cooc, n_sites){
+  # Assemblage probabilities under specified model --------------------------
+  
+  # Energies of all communities
+  E = colSums(-par * possible_cooc)
+  
+  # Partition function
+  Z = logSumExp(-E)
+  
+  # Probabilities of each community
+  p = exp(-E) / exp(Z)
+  
+  # generate x --------------------------------------------------------------
+  
+  # Randomly select row indices from the distribution described above
+  rows = sample.int(nrow(possibilities), size = n_sites, prob = p, replace = TRUE)
+  
+  # Return the selected rows
+  possibilities[rows, ]
+}

### 2015-05-18 Harris_2015/three_species.Rmd
Add 3-species code

# Evaluation of the three species in Figure 1

Load dependencies

```{r}
library(dplyr)
library(magrittr)
library(rosalia)
library(knitr)
library(rosalia)
```

```{r}
set.seed(1)
```

Generate the species interaction matrix from Figure 1.

```{r}
alpha = c(3, 2, 1)
beta = matrix(
  c(0, -3, -3, -3, 0, -1, -3, -1, 0), 
  nrow = 3, 
  dimnames = replicate(2, c("tree", "shrub1", "shrub2"), simplify = FALSE)
)

kable(beta)
```

```{r}
# Simulated community has three species at some of 100 locations
n_spp = 3
n_sites = 100

# Re-organize the coefficients for rosalia
truth = c(beta[upper.tri(beta)], alpha)

# Enumerate the 8 possible species assemblages
# using an internal rosalia function
possibilities = rosalia:::generate_possibilities(n_spp)

# In each of these 8 possible assemblages, which of the 6 possible
# species pairs *both* occur?
possible_cooc = sapply(
  1:2^n_spp,
  function(i){
    tcp = tcrossprod(possibilities[i, ])
    c(tcp[upper.tri(tcp)], diag(tcp))
  }
)
```

Simulate 1000 assemblages of 3 species across 100 sites.

Once a landscape has been simulated, test whether `rosalia`, the sample covariance, and the partial 
covariance got the sign of the shrub-shrub interaction correct.
```{r}
evaluation = sapply(
  1:1000,
  function(i){
    x = rosalia:::simulate_data(
      par = truth, 
      possibilities = possibilities, 
      possible_cooc = possible_cooc, 
      n_sites = n_sites
    )
    if(paste0("fakedata/three/", i, ".txt") %in% dir("fakedata/three", full.names = TRUE)){
      # Do nothing: simulated data was already written to disk
    }else{
      write.table(t(x[rowSums(x) > 0, ]), file = paste0("fakedata/three/", i, ".txt"))
    }

    c(
      cov = cov(x)[2, 3] < 0,
      partial = -solve(cov(x))[2, 3] < 0,
      rosalia = rosalia(x, trace = 0, hessian = FALSE)$par[[3]] < 0
    )
  })

```

After running the "Pairs" software on transposed versions of the files simulated above, 
we can import its results to see how the method performed.

Not

```{r}
Pairs.txt = readLines("fakedata/three/Pairs.txt")
z_score_column = grep("Z-Score", Pairs.txt, value = TRUE)[2] %>% 
  strsplit(" +") %>% 
  extract2(1) %>% 
  grep("Z-Score", .)

# A success for Pairs.exe implies Z>0 because C-scores are about absences, not presences

pairs_evaluation = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
    strsplit(" +") %>%
    sapply(. %>% extract(z_score_column) %>% as.numeric) %>%
    is_greater_than(0)
```

Percentage of correct sign estimates:
```{r, eval=FALSE}
mean(pairs_evaluation) * 100

rowMeans(evaluation) * 100
```

```{r, echo = FALSE}
out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation) * 100)

kable(data_frame(model = names(out), percent_correct = paste(out, "%")))
```

### 2015-05-18 Harris_2015/three_species.Rmd
minor fixes for three_species

@@
 
-Not
-
 ```{r}
 Pairs.txt = readLines("fakedata/three/Pairs.txt")
+
+# Find the column that contains the Z-score
 z_score_column = grep("Z-Score", Pairs.txt, value = TRUE)[2] %>% 
@@
 
-# A success for Pairs.exe implies Z>0 because C-scores are about absences, not presences
-
+# Find the right rows of the text file, split them on whitespace, pull out
+# the column identified above, and find its sign
 pairs_evaluation = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
@@
     sapply(. %>% extract(z_score_column) %>% as.numeric) %>%
-    is_greater_than(0)
+    is_greater_than(0) # Z>0 implies negative interactions because the Z-score
+                       # relates to the distribution of C-scores, which describe
+                       # the frequency with which species *do not* co-occur
 ```

### 2015-05-19 README.Rmd
package stuff

---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# rosalia: Exact inference for small binary Markov networks

Santa Rosalia, the patron saint of biodiversity, became a controversial symbol in the 1970's
and 1980's for ecological inferences regarding species' interactions. This package, which 
provides tools from statistical physics for drawing such inferences, is named in her honor.

# Installation:

The easiest way to install `rosalia` is with the `devtools` package, which can be installed from 
the R command line with `install.packages("devtools")`.

Once `devtools` is available, `rosalia` can be downloaded and installed with 
`devtools::install_github("davharris/rosalia")`

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

### 2015-05-19 R/exact.R
document priors

@@ -3,7 +3,15 @@
 #' @param location Numeric vector; see \code{\link{Logistic}}
 #' @param scale Numeric vector; see \code{\link{Logistic}}
 #' @param ...   Currently not used
-#' @return A `prior` object with a log density and a log gradient
+#' @return A `prior` object, which is a list of two functions:
+#' \item{log_d}{A function for calculating the log of the prior density for 
+#' each element of a vector. Calculated with \code{\link{dlogis}} with \code{log = TRUE}}
+#' \item{log_grad}{A function for calculating the gradient of the log-density for each element 
+#' of a vector.}
+#' @examples 
+#' p = make_logistic_prior(location = 0, scale = 2)
+#' curve(p$log_d(x), from = -5, to = 5)
+#' curve(p$log_grad(x), from = -5, to = 5)
 #' @export
 make_logistic_prior = function(location = 0, scale, ...){
   structure(
@@ -21,13 +29,21 @@ make_logistic_prior = function(location = 0, scale, ...){
 #' Make a flat prior for maximum likelihood estimation
 #' 
 #' @param ...   Currently not used
-#' @return A `prior` object with a constant log density and a log gradient equal to zero
+#' @return A `prior` object, which is a list of two functions:
+#' \item{log_d}{A function for calculating the log of the prior density for each element of a vector. 
+#' With a uniform prior, the log-density is always zero.}
+#' \item{log_grad}{A function for calculating the gradient of the log-density for each element of 
+#' a vector. For the uniform prior, this is always 0. With a uniform prior, this gradient is always zero.}
+#' @examples 
+#' p = make_flat_prior()
+#' curve(p$log_d(x), from = -5, to = 5)
+#' curve(p$log_grad(x), from = -5, to = 5)
 #' @export
-make_flat_log_prior = function(...){
+make_flat_prior = function(...){
   structure(
     list(
-      log_d = function(x){0},
-      log_grad = function(x){0}
+      log_d = function(x){rep(0, length(x))},
+      log_grad = function(x){rep(0, length(x))}
     ),
     class = "prior"
   )
@@ -119,7 +135,7 @@ find_observed_cooc = function(x){
 #' @export
 rosalia = function(
   x, 
-  prior = make_flat_log_prior(), 
+  prior = make_flat_prior(), 
   maxit = 100, 
   trace = 1, 
   hessian = FALSE,

### 2015-05-19 R/exact.R
Improve user-friendliness

@@ -1,13 +1,14 @@
 #' Make a logistic prior with specified scale
 #' 
-#' @param location Numeric vector; see \code{\link{Logistic}}
-#' @param scale Numeric vector; see \code{\link{Logistic}}
+#' @param location Numeric vector; see \code{\link[stats]{Logistic}}
+#' @param scale Numeric vector; see \code{\link[stats]{Logistic}}
 #' @param ...   Currently not used
-#' @return A `prior` object, which is a list of two functions:
+#' @return A \code{prior} object, which is a list of two functions:
 #' \item{log_d}{A function for calculating the log of the prior density for 
-#' each element of a vector. Calculated with \code{\link{dlogis}} with \code{log = TRUE}}
+#' each element of a vector. Calculated with \code{\link[stats]{dlogis}} using \code{log = TRUE}}
 #' \item{log_grad}{A function for calculating the gradient of the log-density for each element 
 #' of a vector.}
+#' @seealso \code{\link{make_flat_prior}}
 #' @examples 
 #' p = make_logistic_prior(location = 0, scale = 2)
 #' curve(p$log_d(x), from = -5, to = 5)
@@ -29,11 +30,12 @@ make_logistic_prior = function(location = 0, scale, ...){
 #' Make a flat prior for maximum likelihood estimation
 #' 
 #' @param ...   Currently not used
-#' @return A `prior` object, which is a list of two functions:
+#' @return A \code{prior} object, which is a list of two functions:
 #' \item{log_d}{A function for calculating the log of the prior density for each element of a vector. 
 #' With a uniform prior, the log-density is always zero.}
 #' \item{log_grad}{A function for calculating the gradient of the log-density for each element of 
 #' a vector. For the uniform prior, this is always 0. With a uniform prior, this gradient is always zero.}
+#' @seealso \code{\link{make_logistic_prior}}
 #' @examples 
 #' p = make_flat_prior()
 #' curve(p$log_d(x), from = -5, to = 5)
@@ -56,10 +58,10 @@ logSumExp = function(x){
 }
 
 
-# Generate all possible binary vectors of length n_spp
-generate_possibilities = function(n_spp){
[…]

### 2015-05-19 inst/tests/tests.R

@@ -1,17 +1,17 @@
 test_that("possibilities are correct", {
-  n_spp = 12
+  n_nodes = 12
   expect_equal(
-    strtoi(apply(generate_possibilities(n_spp), 1, paste0, collapse = ""), 2),
-    seq(0, 2^n_spp - 1)
+  strtoi(apply(generate_possibilities(n_nodes), 1, paste0, collapse = ""), 2),
+    seq(0, 2^n_nodes - 1)
   )
 })
 
 test_that("extracting rows from possibilities works",{
-  n_spp = 12
-  n_sites = 55
-  x = matrix(rbinom(n_spp * n_sites, size = 1, prob = 0.5), ncol = n_spp)
+  n_nodes = 12
+  n_samples = 55
+  x = matrix(rbinom(n_nodes * n_samples, size = 1, prob = 0.5), ncol = n_nodes)
   
-  possibilities = generate_possibilities(n_spp)
+  possibilities = generate_possibilities(n_nodes)
   
   rows = find_rows(x)
   
@@ -19,20 +19,20 @@ test_that("extracting rows from possibilities works",{
 })
 
 test_that("loss works with empty theta",{
-  n_spp = 12
-  n_sites = 55
+  n_nodes = 12
+  n_samples = 55
   
-  x = matrix(rbinom(n_spp * n_sites, size = 1, prob = 0.5), ncol = n_spp)
+  x = matrix(rbinom(n_nodes * n_samples, size = 1, prob = 0.5), ncol = n_nodes)
   
   parlist = as.relistable(list(
-    upper = rep(0, choose(n_spp, 2)),
-    diagonal = rep(0, n_spp)
+    upper = rep(0, choose(n_nodes, 2)),
+    diagonal = rep(0, n_nodes)
   ))
   
-  possibilities = generate_possibilities(n_spp)
+  possibilities = generate_possibilities(n_nodes)
   
   possible_cooc = sapply(
-    1:2^n_spp,
+    1:2^n_nodes,
     function(i){
       tcp = tcrossprod(possibilities[i, ])
       c(tcp[upper.tri(tcp)], diag(tcp))
@@ -43,13 +43,13 @@ test_that("loss works with empty theta",{
   
   expect_equal(
     nll(rows, possible_cooc = possible_cooc, par = unlist(parlist), skeleton = parlist),
-    -log(1/2^n_spp) * length(rows)
+    -log(1/2^n_nodes) * length(rows)
   )
 })
 
[…]

### 2015-05-19 R/exact.R
change default prior, minor tweaks

@@ -130,8 +130,18 @@ find_observed_cooc = function(x){
 
 #' Fit a Markov network to binary data
 #' 
-#' @param x        A binary matrix.
-#' @param prior    An object of class \code{prior}. By default, the prior is flat for maximum likelihood estimation
+#' @description 
+#' The optimization is performed using the \code{BFGS} method in the \code{\link[stats]{optim}}
+#' function.
+#' 
+#' @param x A binary matrix.  Each row corresponds to an observed sample and each
+#' column corresponds to one node in the observed network.
+#' @param prior    An object of class \code{prior}. The default prior, produced by
+#' \code{\link{make_logistic_prior}}, regularizes the parameter estimates. This is
+#' important in Markov networks because the number of parameters will often be large
+#' compared to the number of independent observations. The flat prior produced by 
+#' \code{\link{make_logistic_prior}} is unbiased, but can result in unbounded estimates
+#' if two nodes always share the same state in the data set.
 #' @param maxit,trace,hessian,...    Arguments passed to \code{\link{optim}}
 #' @param parlist  user-specified starting values (optional).
 #' @return a \code{list} with the following elements:
@@ -154,7 +164,7 @@ find_observed_cooc = function(x){
 #' @importFrom assertthat assert_that
 rosalia = function(
   x, 
-  prior = make_flat_prior(), 
+  prior = make_logistic_prior(location = 0, scale = 1), 
   maxit = 100, 
   trace = 1, 
   hessian = FALSE,

### 2015-05-19 inst/tests/tests.R

@@ -193,7 +193,7 @@ test_that("optimization works on small data", {
   rows = sample.int(nrow(possibilities), prob = p, size = n_samples, replace = TRUE)
   x = possibilities[rows, ]
   
-  out = rosalia(x, maxit = 1E4)
+  out = rosalia(x, maxit = 1E4, trace = 0)
 
   estimated_theta = out$beta
   diag(estimated_theta) = out$alpha

### 2015-05-20 Harris_2015/manuscript-materials/Appendices.Rmd
add manuscript materials

# Appendices

## 1. Simlulated community structures

I randomly assigned each species pair an interaction magnitude, $\beta_{ij}$, 
drawn from an exponential distribution with rate parameter 1.  I also randomly
assigned three quarters of the interaction coefficients to be negative, and the
remaining quarter of the coefficients to be positive.

For the simulated landscapes where the abiotic environment was constant across
locations, each species' $\alpha$ coefficient was drawn from a normal
distribution with mean -1 and standard deviation 1.  For the remaining
landscapes, two environmental variables, $x_1$ and $x_2$ were sampled from
independent standard normals with different values for each location. Each
species' response to these two environmental variables was a linear function
of these two environmental variables, with coefficients drawn from normal
distributions with mean 0 and standard deviation 2.  In this way,  species'
$\alpha$ coefficients (and thus their occurrence probabilities) depended on
external environmental factors.

Once the "true" coefficients had been generated for each site on the
landscape, I generated possible landscapes using Gibbs sampling.  In each
round of Gibbs sampling, I cycled through all the species, randomly updating
each one's presence/absence vector in response to its conditional occurrence
probability:

$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$

where the logistic function is $\frac{1}{1+e^{-x}}$ [@murphy_machine_2012]. After 1000 rounds of
sampling, I continued this procedure until I obtained a landscape matrix where
all of the species occurred at least once. I treated the resulting matrices as
"observed" data for analysis.

## 2. Log-likelihood gradient

The *energy* of a Markov network is defined as

$$E(y;\alpha,\beta) = -\sum_{i}\alpha_i y_i - \sum_{i\neq j}\beta_{ij}y_i y_j.$$

The energy function can be used to define the log-likelihood of a given $y$
vector as

$$\log\mathcal{L}(y;\alpha, \beta) = -E(y;\alpha, \beta) - \log(Z(\alpha, \beta)).$$

Here, $Z(\alpha, \beta)$ is the *partition function*, a scaling factor defined
as

$$Z(\alpha, \beta) = \sum_{y \in Y}e^{-E(y;\alpha, \beta)},$$

where $Y$ is the set of all possible $y$ vectors.

The partial derivative of the log-likelihood with respect to $\alpha_i$ is

$$\frac{\partial}{\partial \alpha_i} \log\mathcal{L}(y;\alpha, \beta) = y_i - p(y_i;\alpha, \beta).$$

The first term, $y_i$, is zero if species $i$ is absent in the observed
assemblage and one if it's present. The latter term, $p(y_i;\alpha, \beta)$,
describes the expected probability of observing species $i$ under the current
values of $\alpha$ and $\beta$. It comes from the derivative of the partition
function, as derived in (learning Boltzmann machines, Murphy 2012, etc.).
Following the gradient of $\alpha_i$ adjusts the expected probability of
observing species $i$ until it matches the observed value and the two terms in
the gradient cancel one another out.

The partial derivative of the log-likelihood with respect to $\beta_{ij}$
can be derived similarly as

$$\frac{\partial}{\partial \alpha_i} \log\mathcal{L}(y;\alpha, \beta) =y_i y_j - p(y_i y_j;\alpha, \beta).$$

Following this gradient adjusts the expected probability of co-occurrence
between species $i$ and species $j$ until this value matches the observed 
co-occurrence frequency and the two terms in the gradient cancel one another out.

# References

### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
add manuscript materials

*Running head:* Species interactions in Markov networks

**Title: Estimating species interactions from observational data with Markov
networks**

**Author:** David J. Harris: Population Biology; 1 Shields Avenue, Davis CA, 95616

***Abstract***

Inferring species interactions from observational data is one of the most
controversial tasks in community ecology. One difficulty is that a single
pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. A linear
approximation, based on partial covariances, also performed well as long as
the number of sampled locations exceeded the number of species in the data.
Indirect effects reliably caused a common null modeling approach to produce
incorrect inferences, however.

**Key words:** Ecological interactions; Occurrence data; Species associations;
Markov network; Markov random field; Ising model; Biogeography;
Presence–absence matrix; Null model

***Introduction***

Ecologists' intense interest in drawing inferences about species interactions
---especially competition---from presence-absence data has a long history
[@macarthur_population_1958; @diamond_island_1975; @connor_checkered_2013]. If
nontrophic species interactions are important drivers of community assembly patterns,
then we should expect to see their influence in our data sets. Despite decades
of work and several major controversies, however [@lewin_santa_1983;
@strong_ecological_1984; @gotelli_swap_2003], existing methods for detecting
competition's effects on community structure are unreliable
[@gotelli_empirical_2009]. More generally, it can be difficult to reason about
the complex web of direct and indirect interactions that contribute to the
structure of real assemblages, especially when these interactions  occur
against a background of other ecological processes such as dispersal and
environmental filtering [@connor_checkered_2013]. For this reason, it isn't
always clear what kinds of patterns would even constitute evidence of
competition, as opposed to some other biological process or random sampling
error [@roughgarden_competition_1983; @lewin_santa_1983].

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that would
be expected if *all* species on the landscape were independent
[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
against such a "null" background, however, rules out the possibility that the
overall association between two species could be driven by an outside force,
as shown in Figure 1. Even though the two shrub species in this example compete 
with one another for resources at a mechanistic level, they end up clustering
together on the landscape because they both grow best in areas that are not overshadowed
by trees. Due to fundamental constraints on the structure of valid
covariance matrices [@brown_constraints_2010], this kind of positive correlation
is almost inevitable, even when all the interactions are negative. For this
reason, knowing that the two shrubs co-occur more than expected by chance
under some null model (e.g. that their correlation in the data set is positive)
can not---by itself---tell us much about their direct effects on one another.

While the competition between the two shrubs in the previous example does not
leave the commonly-expected pattern in community structure (negative
association at the landscape level), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, *among shaded sites*, there will be a
deficit of co-occurrences, and *among unshaded sites*, there will also be such
a deficit. These *conditional* ("all-else-equal") relationships can be estimated with partial
correlations [@albrecht_spatial_2001], although this approach is best-suited
for Gaussian variables.  For non-Gaussian data, a more general approach is
needed.

In this paper, I introduce Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks,
which generalize partial correlations to non-Gaussian data, have been used in
many scientific fields to model associations between various kinds of
"particles". For example, a well-studied network called the Ising model has played
an important role in our understanding of physics [where nearby particles tend
to align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I suggest that ecologists could similarly treat
species as the interacting particles in this modeling framework. Doing so
would allow ecologists to simulate and study the landscape-level consequences
of arbitrary species interaction matrices, even when our observations are not
Gaussian. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously forced
researchers to rely on severe approximations that produced unintelligible
results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
Now that it is computationally feasible to fit these models
exactly, the approach has become worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov networks
work and how they can be used to simulate landscape-level data and to predict
the direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the "true" ecological structure is known, I
compare this approach with several existing methods for detecting species
interactions. Finally, I discuss opportunities for extending the approach
presented here to larger problems in community ecology.

***Methods***

***Conditional relationships and Markov networks.***
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships. 
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001], as opposed to the simple correlation, which
does not control for other variables. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs, 
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition). To the
extent that our observations are not multivariate Gaussian (e.g. 
presence-absence data), partial covariances do not describe variables' conditional
relationships exactly, and a more general approach is needed. Markov networks
allow researchers to generalize conditional relationships to non-Gaussian
networks, much as generalized linear models allow researchers to extend
regression to non-Gaussian response variables [@lee_learning_2012].

Markov networks give a probability value for every possible combination of
presences and absences in communities. For example, given a network with
binary outcomes (i.e. 0 for absence and 1 for presence), the relative
probability of observing a given presence-absence vector, $\vec{y}$, is given
by

$$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$$

Here, $\alpha_{i}$ is the amount that the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability, and controls
how often the two species will be found together (Figure 2A, Figure 2B).
$\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
networks. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

A major feature of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of $\beta$
coefficients [@murphy_machine_2012]. For example, if the coefficient linking
two mutualist species is $+2$, then---all else equal---the odds of observing
either species increase by a factor of $e^2$ when its partner is present
[@murphy_machine_2012]. Of course, if all else is *not* equal (e.g. Figure 1,
where the presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from this
expectation. For this reason, it is important to consider how coefficients'
effects propagate through the network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a *partition
function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). While the analyses in this paper involve exact computations, this
approach quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species the number of bookkeeping operations
required for exact inference quickly spirals exponentially into the billions.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, either through analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but these techniques are beyond the scope of this paper.

***Simulations.***
In order to compare different methods for drawing inferences from
observational data, I simulated two sets of landscapes using known
parameters.

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability distribution
defined by the interaction coefficients in that figure. Each of the methods
described below (a Markov network, two correlation-based methods and a null modeling 
approach) was then evaluated on its ability to correctly infer that all
three species interactions were negative, despite the frequent co-occurrences
of the two shrub species.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500, or
2500 observed communities (24 replicate simulations for each combination).
These simulated data sets span the range from small, single-observer data
sets to large collaborative efforts such as the North American Breeding Bird
Survey. As described in Appendix 1, I randomly drew the "true" coefficient
values for each replicate so that most species pairs interacted negligibly, a
few pairs interacted very strongly, and competition was three times more
common than facilitation. I then used Gibbs sampling to randomly generate
landscapes with varying numbers of species and sites via Markov chain Monte
Carlo (Appendix 1). For half of the simulated landscapes, I treated each
species' $\alpha$ coefficient as a constant, as described above. For the other
half, I treated the $\alpha$ coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix 1). The latter set of simulated landscapes provide an
important test of the methods' ability to distinguish co-occurrence patterns
that were generated from pairwise interactions among the observed species from
those that were generated by external forces like abiotic environmental
filtering. This task was made especially difficult because---as with most
analyses of presence-absence data for co-occurrence patterns---the inference
procedure did not have access to any information about the environmental or
spatial variables that helped shape the landscape [cf @connor_checkered_2013;
@blois_framework_2014].

***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
The previous sections involved known values of $\alpha$ and $\beta$. In most
real cases, however, the true values of these parameters are not known, and
ecologists  will need to estimate them from data. When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for
all of the $\alpha$ and $\beta$ coefficients by optimizing Equation 1
(Appendix 2).  Doing so allows us to find the unique set of $\alpha$ and
$\beta$ coefficients that would be expected to produce exactly the observed
occurrence frequencies and co-occurrence frequencies. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum.

For the analyses in this paper, I made a small modification to this maximum
likelihood procedure. Given the large number of parameters associated with
some of the networks to be estimated, I regularized the likelihood using a
logistic prior distribution [@gelman_weakly_2008] with a scale of 1 on the
$\alpha$ and $\beta$ terms.

The optimization procedures described here and in Appendix 2 are implemented
in the *rosalia* package^[source code available for download from
https://github.com/davharris/rosalia] for the R programming language
[@r_core_team_r_2015].

***Other inference techniques for comparison.***
After fitting Markov networks to the simulated landscapes described above, I
used several other techniques for inferring the sign and strength of marginal
associations between pairs of species.

The first two interaction measures were the simple and partial covariances
between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001]. Because partial covariances are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
regression with a Gaussian prior]. For these analyses, I set the ridge
parameter to 0.2 divided by the number of sites on the landscape.

The third method, described in @gotelli_empirical_2009, involved simulating
possible landscapes from a null model that retained the row and column sums of
the original matrix [@strong_ecological_1984]. Using the default options in the
Pairs software described in @gotelli_empirical_2009, I simulated  the null
distribution of scaled C-scores (a test statistic describing the number of
*non*-co-occurrences between two species).  The software then calculated a
$Z$-statistic for each species pair using this null distribution. After
multiplying this statistic by $-1$ so that positive values corresponded to
facilitation and negative values corresponded to competition, I used it as
another estimate of species interactions.

***Method evaluation.***
For the first simulated landscape (three species), I kept the evaluation
simple and qualitative: any method that reliably determined that the two shrub
species were negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using linear
regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the "true" $\beta$ coefficient for
each species pair against the model's estimate, re-weighting the pairs so that
each landscape contributed equally to the rescaled estimate^[The null model
generated one $Z$-score outlier greater than 1000, which dominated the
regression and squared error analyses. To reduce its influence on these
results, I changed its value to 32.5, which was the value of the next largest
$Z$-score in the null model's results.]. Then, for each combination of species
richness, landscape size, method, and presence/absence of environmental
filtering, I calculated the mean squared error associated with the rescaled
estimate across all 12 replicates. Finally, I calculated an $R^2$ value for
each landscape by comparing this squared error estimate with the squared error
associated with the null hypothesis that all interactions were zero.

***Results***

***Three species.***
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---
due to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations 94% of the time
and the randomization-based null model falsely reported such associations 100%
of the  time. The two methods for evaluating conditional relationships (Markov
networks and partial covariances), however, successfully controlled for the
indirect pathway via the tree species and each correctly identified the direct
negative interaction between the shrubs 92% of the time.

***Larger landscapes.***
The Markov network consistently outperformed all of the other techniques that
I evaluated (Figure 3). Of the remaining methods, partial covariances tended
to be most accurate, especially when 100 or more communities had been observed
on the landscape. $Z$-scores from the null model had the least predictive
utility overall, as indicated by $R^2$ values. In general, all the models'
accuracies declined when environmental filters were added, but the declines
were smaller when environmentally-induced correlations were spread out among a
larger number of pairwise interactions.

***Discussion***

The results presented above are very promising, as they show that Markov
networks can recover much of the variation in species' pairwise interaction
strengths from observational data, even when direct interactions are largely
overwhelmed by indirect effects (e.g. Figure 1) or environmental effects
(lower panels of Figure 3). For cases where it is infeasible to fit a Markov
network, these results also indicate that partial covariances---which can be
computed straightforwardly by linear regression---can often provide an
accurate approximation.

Apart from the environmental filters, the simulated landscapes presented here
represent the best-case scenario for these methods.  Future research should
thus examine these models' performance characteristics when the "true"
interaction matrices include guild structure or trophic levels, which could
make the $\beta$ coefficients much more difficult to infer (particularly for
linear approximations like the partial covariance approach;
@loh_structure_2013). On the other hand, ecologists may often have prior 
information about the nature of real species' interaction patterns from 
natural history or ecological experiments, which could substantially reduce
the probability and magnitude of error. The rosalia package has built-in
mechanisms for incorporating this kind of information, if it can be expressed
as a prior probability distribution or a penalty on the likelihood.

Additionally, it is important to note that, while partial correlations and
Markov networks both prevent us from mistaking marginal associations for
conditional ones, they cannot tell us the underlying biological mechanism.
Real species co-occurrence patterns will depend on a number of factors---especially
in taxa that emigrate in response to other species---and the
$\beta$ coefficients in Markov networks have to reduce this to a single
number. Thus, experiments and natural history knowledge will generally be
required to pin down the exact nature of the interaction (e.g. who outcompetes
whom).

Despite these limitations, the results with environmental filtering seem to
indicate that the method can be very robust.  Additionally, the fact that
Markov networks provide a likelihood function to optimize makes them highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>100 species) while simultaneously modeling each species'
response to the abiotic environment with complex, nonlinear functions. This
sort of approach, which combines multiple ecological processes, could help
ecologists to disentangle different factors behind the co-occurrence patterns
we observe in nature. Numerous other extensions are possible: similar networks
can be fit with continuous variables, count data, or both
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that could tip the balance between facilitation and
exploitation [@bruno_inclusion_2003]. Fully exploring these possibilities will
require more research into the various available approximations to the log-likelihood
and to its gradient, in order to balance efficiency, accuracy, and
the ability to generate confidence limits for statistical inference.

By providing precise quantitative expectations about the results of species
interactions, Markov networks have the potential for addressing long-standing
ecological questions. For example, Markov networks can provide a precise
answer to the question of how competition affects species' overall prevalence,
which was a major flash point for the null model debates in the 1980’s
[@strong_ecological_1984]. From Equation 1, one can derive the expected
prevalence of a species in the absence of biotic influences ($\frac{1}{1 +
e^{-\alpha}}$). Any significant difference between this value and the observed
prevalence can be attributed to the $\beta$ coefficients linking this species
to its facilitators and competitors (cf Figure 2D).

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for nearly a
century, and the models' properties, capabilities, and limits are well-understood
in a huge range of applications, from spatial lattices
[@gelfand_modelling_2005] to haphazard neural connections in the retina and
brain [@schneidman_weak_2006] to complex behavioral interactions among human
beings [@lee_statistical_2013]. Modeling species interactions using the same framework
would thus allow ecologists to tap into an enormous set of existing discoveries and
techniques for dealing with indirect effects, stability, and alternative
stable states.

Finally, the null models commonly used by ecologists for drawing inferences
about pairwise species interactions do not seem to distinguish between direct
and indirect interactions any better than simple correlation coefficients.
Null and neutral models can be very useful for clarifying our thinking about the numerical
consequences of species' richness and abundance patterns
[@harris_occupancy_2011; @xiao_strong_2015], but it is less clear whether deviations
from the null hypothesis can clearly implicate specific biological processes like 
competition.

***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, and N. J. Gotelli provided useful feedback on an earlier draft of
this work.

***References:***

***Figure captions***

***Figure 1.*** 
**A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, as indicated by their negative coefficient (circled), but this effect
is substantially weaker. **B.** In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in locations
without trees makes their occurrence vectors positively correlated (circled).
**C.** Controlling for the tree species' presence with a conditional method
such as a partial covariance or a Markov network allows us to correctly
identify the negative interaction between these two species (circled).

***Figure 2.***
**A.** A small Markov network with two species.  The depicted abiotic
environment favors the occurrence of both species ($\alpha >0$), particularly
species 2 ($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking
these two species implies that they co-occur less than expected under
independence.  **B.** Relative probabilities of all four possible presence-absence
combinations for Species 1 and Species 2. The exponent includes
$\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
species $2$ is present ($y_2 = 1$), and $\beta$ only when both are present
($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four relative
probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
probabilities, we can find the expected frequencies of all possible co-occurrence
 patterns between the two species of interest. **D.** If $\beta$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
occurrence rate and the deficit of co-occurrences would be eliminated.

***Figure 3.***
Proportion of variance in interaction coefficients explained by each method
with 5, 10, or 20 species arrayed across varying numbers of sampled locations
when environmental filtering was absent (top row) or present (bottom  row).

### 2015-05-20 Harris_2015/manuscript-materials/draft.tex
add manuscript materials

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\date{}

\begin{document}

\begin{spacing}{1.9}
\begin{flushleft}
\emph{Running head:} Species interactions in Markov networks

\textbf{Title: Estimating species interactions from observational data
with Markov networks}

\textbf{Author:} David J. Harris: Population Biology; 1 Shields Avenue,
Davis CA, 95616

\textbf{\emph{Abstract}}

Inferring species interactions from observational data is one of the
most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and
produce surprising indirect consequences. For example, two competing
species would ordinarily correlate negatively in space, but this effect
can be reversed in the presence of a third species that is capable of
outcompeting both of them when it is present. Here, I apply models from
statistical physics, called Markov networks or Markov random fields,
that can predict the direct and indirect consequences of any possible
species interaction matrix. Interactions in these models can be
estimated from observational data via maximum likelihood. Using
simulated landscapes with known pairwise interaction strengths, I
evaluated Markov networks and several existing approaches. The Markov
networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect
interactions or abiotic environmental effects largely overpowered them.
A linear approximation, based on partial covariances, also performed
well as long as the number of sampled locations exceeded the number of
species in the data. Indirect effects reliably caused a common null
modeling approach to produce incorrect inferences, however.

\textbf{Key words:} Ecological interactions; Occurrence data; Species
associations; Markov network; Markov random field; Ising model;
Biogeography; Presence--absence matrix; Null model

\textbf{\emph{Introduction}}

Ecologists' intense interest in drawing inferences about species
interactions ---especially competition---from presence-absence data has
a long history (MacArthur 1958, Diamond 1975, Connor et al. 2013). If
nontrophic species interactions are important drivers of community
assembly patterns, then we should expect to see their influence in our
data sets. Despite decades of work and several major controversies,
however (Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003),
existing methods for detecting competition's effects on community
structure are unreliable (Gotelli and Ulrich 2009). More generally, it
can be difficult to reason about the complex web of direct and indirect
interactions that contribute to the structure of real assemblages,
especially when these interactions occur against a background of other
ecological processes such as dispersal and environmental filtering
(Connor et al. 2013). For this reason, it isn't always clear what kinds
of patterns would even constitute evidence of competition, as opposed to
some other biological process or random sampling error (Lewin 1983,
Roughgarden 1983).

Most existing methods in this field compare the frequency with which two
putative competitors are observed to co-occur against the frequency that
would be expected if \emph{all} species on the landscape were
independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
species pair against such a ``null'' background, however, rules out the
possibility that the overall association between two species could be
driven by an outside force, as shown in Figure 1. Even though the two
shrub species in this example compete with one another for resources at
a mechanistic level, they end up clustering together on the landscape
because they both grow best in areas that are not overshadowed by trees.
Due to fundamental constraints on the structure of valid covariance
matrices (Brown et al. 2010), this kind of positive correlation is
almost inevitable, even when all the interactions are negative. For this
reason, knowing that the two shrubs co-occur more than expected by
chance under some null model (e.g.~that their correlation in the data
set is positive) can not---by itself---tell us much about their direct
effects on one another.

While the competition between the two shrubs in the previous example
does not leave the commonly-expected pattern in community structure
(negative association at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, \emph{among shaded
sites}, there will be a deficit of co-occurrences, and \emph{among
unshaded sites}, there will also be such a deficit. These
\emph{conditional} (``all-else-equal'') relationships can be estimated
with partial correlations (Albrecht and Gotelli 2001), although this
approach is best-suited for Gaussian variables. For non-Gaussian data, a
more general approach is needed.

In this paper, I introduce Markov networks (undirected graphical models
also known as Markov random fields; Murphy 2012) as a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov
networks, which generalize partial correlations to non-Gaussian data,
have been used in many scientific fields to model associations between
various kinds of ``particles''. For example, a well-studied network
called the Ising model has played an important role in our understanding
of physics (where nearby particles tend to align magnetically with one
another; Cipra 1987). In spatial contexts, these models have been used
to describe interactions between adjacent grid cells (Harris 1974,
Gelfand et al. 2005). In neurobiology, they have helped researchers
determine which neurons are connected to one another by modeling the
structure in their firing patterns (Schneidman et al. 2006). Following
recent work by Azaele et al. (2010) and Fort (2013), I suggest that
ecologists could similarly treat species as the interacting particles in
this modeling framework. Doing so would allow ecologists to simulate and
study the landscape-level consequences of arbitrary species interaction
matrices, even when our observations are not Gaussian. While ecologists
explored some related approaches in the 1980's (Whittam and
Siegel-Causey 1981), computational limitations had previously forced
researchers to rely on severe approximations that produced
unintelligible results (e.g. ``probabilities'' greater than one; Gilpin
and Diamond 1982). Now that it is computationally feasible to fit these
models exactly, the approach has become worth a second look.

The rest of the paper proceeds as follows. First, I discuss how Markov
networks work and how they can be used to simulate landscape-level data
and to predict the direct and indirect consequences of possible
interaction matrices. Then, using simulated data sets where the ``true''
ecological structure is known, I compare this approach with several
existing methods for detecting species interactions. Finally, I discuss
opportunities for extending the approach presented here to larger
problems in community ecology.

\textbf{\emph{Methods}}

\textbf{\emph{Conditional relationships and Markov networks.}}
Ecologists are often interested in inferring direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate
\emph{conditional} (``all-else-equal'') relationships, rather than
\emph{marginal} (``overall'') relationships. The most familiar
conditional relationship is the partial correlation, which indicates the
portion of the correlation between two species that remains after
controlling for other variables in the data set (Albrecht and Gotelli
2001), as opposed to the simple correlation, which does not control for
other variables. The example with the shrubs and trees in Figure 1 shows
how the two correlation measures can have opposite signs, and suggests
that the partial correlation is more relevant for drawing inferences
about species interactions (e.g.~competition). To the extent that our
observations are not multivariate Gaussian (e.g. presence-absence data),
partial covariances do not describe variables' conditional relationships
exactly, and a more general approach is needed. Markov networks allow
researchers to generalize conditional relationships to non-Gaussian
networks, much as generalized linear models allow researchers to extend
regression to non-Gaussian response variables (Lee and Hastie 2012).

Markov networks give a probability value for every possible combination
of presences and absences in communities. For example, given a network
with binary outcomes (i.e.~0 for absence and 1 for presence), the
relative probability of observing a given presence-absence vector,
\(\vec{y}\), is given by

\[p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\]

Here, \(\alpha_{i}\) is the amount that the presence of species \(i\)
contributes to the log-probability of \(\vec{y}\); it directly controls
the prevalence of species \(i\). Similarly, \(\beta_{ij}\) is the amount
that the co-occurrence of species \(i\) and species \(j\) contributes to
the log-probability, and controls how often the two species will be
found together (Figure 2A, Figure 2B). \(\beta\) thus acts as an analog
of the partial covariance, but for non-Gaussian networks. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce
assemblages that have many pairs of positively-associated species and
relatively few pairs of negatively-associated species (exactly as an
ecologist might expect).

A major feature of Markov networks is the fact that the conditional
relationships between species can be read directly off the matrix of
\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
linking two mutualist species is \(+2\), then---all else equal---the
odds of observing either species increase by a factor of \(e^2\) when
its partner is present (Murphy 2012). Of course, if all else is
\emph{not} equal (e.g.~Figure 1, where the presence of one competitor is
associated with release from another competitor), then species' marginal
association rates can differ from this expectation. For this reason, it
is important to consider how coefficients' effects propagate through the
network, as discussed below.

Estimating the marginal relationships predicted by a Markov network is
more difficult than estimating conditional relationships, because doing
so requires absolute probability estimates. Turning the relative
probability given by Equation 1 into an absolute probability entails
scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
ensures that the probabilities of all possible assemblages that could be
produced by the model sum to one (bottom of Figure 2B). While the
analyses in this paper involve exact computations, this approach quickly
becomes infeasible as the number of species increases: with \(2^N\)
possible assemblages of \(N\) species the number of bookkeeping
operations required for exact inference quickly spirals exponentially
into the billions. Numerous techniques are available for working with
Markov networks that keep the computations tractable, either through
analytic approximations (Lee and Hastie 2012) or Monte Carlo sampling
(Salakhutdinov 2008), but these techniques are beyond the scope of this
paper.

\textbf{\emph{Simulations.}} In order to compare different methods for
drawing inferences from observational data, I simulated two sets of
landscapes using known parameters.

The first set of simulated landscapes included the three competing
species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling exactly from a probability
distribution defined by the interaction coefficients in that figure.
Each of the methods described below (a Markov network, two
correlation-based methods and a null modeling approach) was then
evaluated on its ability to correctly infer that all three species
interactions were negative, despite the frequent co-occurrences of the
two shrub species.

I also simulated a second set of landscapes with five, ten, or twenty
potentially-interacting species on landscapes composed of 20, 100, 500,
or 2500 observed communities (24 replicate simulations for each
combination). These simulated data sets span the range from small,
single-observer data sets to large collaborative efforts such as the
North American Breeding Bird Survey. As described in Appendix 1, I
randomly drew the ``true'' coefficient values for each replicate so that
most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.
I then used Gibbs sampling to randomly generate landscapes with varying
numbers of species and sites via Markov chain Monte Carlo (Appendix 1).
For half of the simulated landscapes, I treated each species' \(\alpha\)
coefficient as a constant, as described above. For the other half, I
treated the \(\alpha\) coefficients as linear functions of two abiotic
environmental factors that varied from location to location across the
landscape (Appendix 1). The latter set of simulated landscapes provide
an important test of the methods' ability to distinguish co-occurrence
patterns that were generated from pairwise interactions among the
observed species from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013, Blois et al. 2014).

\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
presence-absence data.}} The previous sections involved known values of
\(\alpha\) and \(\beta\). In most real cases, however, the true values
of these parameters are not known, and ecologists will need to estimate
them from data. When the number of species is reasonably small, one can
compute exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients by optimizing Equation 1 (Appendix 2). Doing so
allows us to find the unique set of \(\alpha\) and \(\beta\)
coefficients that would be expected to produce exactly the observed
occurrence frequencies and co-occurrence frequencies. Fully-observed
Markov networks like the ones considered here have unimodal likelihood
surfaces (Murphy 2012), ensuring that this procedure will always
converge on the global maximum.

For the analyses in this paper, I made a small modification to this
maximum likelihood procedure. Given the large number of parameters
associated with some of the networks to be estimated, I regularized the
likelihood using a logistic prior distribution (Gelman et al. 2008) with
a scale of 1 on the \(\alpha\) and \(\beta\) terms.

The optimization procedures described here and in Appendix 2 are
implemented in the \emph{rosalia} package\footnote{source code available
  for download from https://github.com/davharris/rosalia} for the R
programming language (R Core Team 2015).

\textbf{\emph{Other inference techniques for comparison.}} After fitting
Markov networks to the simulated landscapes described above, I used
several other techniques for inferring the sign and strength of marginal
associations between pairs of species.

The first two interaction measures were the simple and partial
covariances between each pair of species' data vectors on the landscape
(Albrecht and Gotelli 2001). Because partial covariances are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on ridge regression {[}Wieringen and Peeters
(2014); i.e.~linear regression with a Gaussian prior{]}. For these
analyses, I set the ridge parameter to 0.2 divided by the number of
sites on the landscape.

The third method, described in Gotelli and Ulrich (2009), involved
simulating possible landscapes from a null model that retained the row
and column sums of the original matrix (Strong et al. 1984). Using the
default options in the Pairs software described in Gotelli and Ulrich
(2009), I simulated the null distribution of scaled C-scores (a test
statistic describing the number of \emph{non}-co-occurrences between two
species). The software then calculated a \(Z\)-statistic for each
species pair using this null distribution. After multiplying this
statistic by \(-1\) so that positive values corresponded to facilitation
and negative values corresponded to competition, I used it as another
estimate of species interactions.

\textbf{\emph{Method evaluation.}} For the first simulated landscape
(three species), I kept the evaluation simple and qualitative: any
method that reliably determined that the two shrub species were
negatively associated passed; other methods failed.

For the larger landscapes, I rescaled the four methods' estimates using
linear regression through the origin so that they all had a consistent
interpretation. In each case, I regressed the ``true'' \(\beta\)
coefficient for each species pair against the model's estimate,
re-weighting the pairs so that each landscape contributed equally to the
rescaled estimate\footnote{The null model generated one \(Z\)-score
  outlier greater than 1000, which dominated the regression and squared
  error analyses. To reduce its influence on these results, I changed
  its value to 32.5, which was the value of the next largest \(Z\)-score
  in the null model's results.}. Then, for each combination of species
richness, landscape size, method, and presence/absence of environmental
filtering, I calculated the mean squared error associated with the
rescaled estimate across all 12 replicates. Finally, I calculated an
\(R^2\) value for each landscape by comparing this squared error
estimate with the squared error associated with the null hypothesis that
all interactions were zero.

\textbf{\emph{Results}}

\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
relationship between the two shrub species was positive---despite their
competition for space at a mechanistic level--- due to indirect effects
of the dominant tree species. As a result, the covariance method falsely
reported positive associations 94\% of the time and the
randomization-based null model falsely reported such associations 100\%
of the time. The two methods for evaluating conditional relationships
(Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each
correctly identified the direct negative interaction between the shrubs
92\% of the time.

\textbf{\emph{Larger landscapes.}} The Markov network consistently
outperformed all of the other techniques that I evaluated (Figure 3). Of
the remaining methods, partial covariances tended to be most accurate,
especially when 100 or more communities had been observed on the
landscape. \(Z\)-scores from the null model had the least predictive
utility overall, as indicated by \(R^2\) values. In general, all the
models' accuracies declined when environmental filters were added, but
the declines were smaller when environmentally-induced correlations were
spread out among a larger number of pairwise interactions.

\textbf{\emph{Discussion}}

The results presented above are very promising, as they show that Markov
networks can recover much of the variation in species' pairwise
interaction strengths from observational data, even when direct
interactions are largely overwhelmed by indirect effects (e.g.~Figure 1)
or environmental effects (lower panels of Figure 3). For cases where it
is infeasible to fit a Markov network, these results also indicate that
partial covariances---which can be computed straightforwardly by linear
regression---can often provide an accurate approximation.

Apart from the environmental filters, the simulated landscapes presented
here represent the best-case scenario for these methods. Future research
should thus examine these models' performance characteristics when the
``true'' interaction matrices include guild structure or trophic levels,
which could make the \(\beta\) coefficients much more difficult to infer
(particularly for linear approximations like the partial covariance
approach; Loh and Wainwright (2013)). On the other hand, ecologists may
often have prior information about the nature of real species'
interaction patterns from natural history or ecological experiments,
which could substantially reduce the probability and magnitude of error.
The rosalia package has built-in mechanisms for incorporating this kind
of information, if it can be expressed as a prior probability
distribution or a penalty on the likelihood.

Additionally, it is important to note that, while partial correlations
and Markov networks both prevent us from mistaking marginal associations
for conditional ones, they cannot tell us the underlying biological
mechanism. Real species co-occurrence patterns will depend on a number
of factors---especially in taxa that emigrate in response to other
species---and the \(\beta\) coefficients in Markov networks have to
reduce this to a single number. Thus, experiments and natural history
knowledge will generally be required to pin down the exact nature of the
interaction (e.g.~who outcompetes whom).

Despite these limitations, the results with environmental filtering seem
to indicate that the method can be very robust. Additionally, the fact
that Markov networks provide a likelihood function to optimize makes
them highly extensible, even when it is inconvenient to compute the
likelihood exactly. For example, the mistnet software package for joint
species distribution modeling (Harris 2015) can fit \emph{approximate}
Markov networks to large species assemblages (\textgreater{}100 species)
while simultaneously modeling each species' response to the abiotic
environment with complex, nonlinear functions. This sort of approach,
which combines multiple ecological processes, could help ecologists to
disentangle different factors behind the co-occurrence patterns we
observe in nature. Numerous other extensions are possible: similar
networks can be fit with continuous variables, count data, or both (Lee
and Hastie 2012). There are even methods (Whittam and Siegel-Causey
1981, Tjelmeland and Besag 1998) that would allow the coefficient
linking two species in an interaction matrix to vary as a function of
the abiotic environment or of third-party species that could tip the
balance between facilitation and exploitation (Bruno et al. 2003). Fully
exploring these possibilities will require more research into the
various available approximations to the log-likelihood and to its
gradient, in order to balance efficiency, accuracy, and the ability to
generate confidence limits for statistical inference.

By providing precise quantitative expectations about the results of
species interactions, Markov networks have the potential for addressing
long-standing ecological questions. For example, Markov networks can
provide a precise answer to the question of how competition affects
species' overall prevalence, which was a major flash point for the null
model debates in the 1980's (Strong et al. 1984). From Equation 1, one
can derive the expected prevalence of a species in the absence of biotic
influences (\(\frac{1}{1 + e^{-\alpha}}\)). Any significant difference
between this value and the observed prevalence can be attributed to the
\(\beta\) coefficients linking this species to its facilitators and
competitors (cf Figure 2D).

This paper only scratches the surface of what Markov networks can do for
ecology. This family of models---particularly the Ising model for binary
networks---has been extremely well-studied in statistical physics for
nearly a century, and the models' properties, capabilities, and limits
are well-understood in a huge range of applications, from spatial
lattices (Gelfand et al. 2005) to haphazard neural connections in the
retina and brain (Schneidman et al. 2006) to complex behavioral
interactions among human beings (Lee et al. 2013). Modeling species
interactions using the same framework would thus allow ecologists to tap
into an enormous set of existing discoveries and techniques for dealing
with indirect effects, stability, and alternative stable states.

Finally, the null models commonly used by ecologists for drawing
inferences about pairwise species interactions do not seem to
distinguish between direct and indirect interactions any better than
simple correlation coefficients. Null and neutral models can be very
useful for clarifying our thinking about the numerical consequences of
species' richness and abundance patterns (Harris et al. 2011, Xiao et
al. 2015), but it is less clear whether deviations from the null
hypothesis can clearly implicate specific biological processes like
competition.

\textbf{\emph{Acknowledgements:}} This research was funded by a Graduate
Research Fellowship from the US National Science Foundation and
benefited greatly from discussions with A. Sih, M. L. Baskett, R.
McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A.
K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding,
and N. J. Gotelli provided useful feedback on an earlier draft of this
work.

\textbf{\emph{References:}}

\textbf{\emph{Figure captions}}

\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
species. The tree (top) tends not to co-occur with either of the two
shrub species, as indicated by the strongly negative coefficient linking
them. The two shrub species also compete with one another, as indicated
by their negative coefficient (circled), but this effect is
substantially weaker. \textbf{B.} In spite of the competitive
interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors
positively correlated (circled). \textbf{C.} Controlling for the tree
species' presence with a conditional method such as a partial covariance
or a Markov network allows us to correctly identify the negative
interaction between these two species (circled).

\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
species. The depicted abiotic environment favors the occurrence of both
species (\(\alpha >0\)), particularly species 2
(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
these two species implies that they co-occur less than expected under
independence. \textbf{B.} Relative probabilities of all four possible
presence-absence combinations for Species 1 and Species 2. The exponent
includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
constant \(Z\), ensures that the four relative probabilities sum to 1.
In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
we can find the expected frequencies of all possible co-occurrence
patterns between the two species of interest. \textbf{D.} If \(\beta\)
equaled zero (e.g.~if the species no longer competed for the same
resources), then the reduction in competition would allow each species
to increase its occurrence rate and the deficit of co-occurrences would
be eliminated.

\textbf{\emph{Figure 3.}} Proportion of variance in interaction
coefficients explained by each method with 5, 10, or 20 species arrayed
across varying numbers of sampled locations when environmental filtering
was absent (top row) or present (bottom row).

Albrecht, M., and N. J. Gotelli. 2001. Spatial and temporal niche
partitioning in grassland ants. Oecologia 126:134--141.

Azaele, S., R. Muneepeerakul, A. Rinaldo, and I. Rodriguez-Iturbe. 2010.
Inferring plant ecosystem organization from species occurrences. Journal
of theoretical biology 262:323--329.

Blois, J. L., N. J. Gotelli, A. K. Behrensmeyer, J. T. Faith, S. K.
Lyons, J. W. Williams, K. L. Amatangelo, A. Bercovici, A. Du, J. T.
Eronen, and others. 2014. A framework for evaluating the influence of
climate, dispersal limitation, and biotic interactions using fossil
pollen associations across the late Quaternary. Ecography 37:1095--1108.

Brown, I. H., E. Bedrick, S. Morgan, I.-L. E. C. Ernest, and I. F.
Kelly. 2010. Constraints on Negative Relationships: Mathematical Causes
and Ecological. The Nature of Scientific Evidence: Statistical,
Philosophical, and Empirical Considerations:298.

Bruno, J. F., J. J. Stachowicz, and M. D. Bertness. 2003. Inclusion of
facilitation into ecological theory. Trends in Ecology \& Evolution
18:119--125.

Cipra, B. A. 1987. An introduction to the Ising model. American
Mathematical Monthly 94:937--959.

Connor, E. F., M. D. Collins, and D. Simberloff. 2013. The checkered
history of checkerboard distributions. Ecology 94:2403--2414.

Diamond, J. M. 1975. The island dilemma: Lessons of modern biogeographic
studies for the design of natural reserves. Biological conservation
7:129--146.

Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
Selected Problems in Ecology. Entropy 15:5237--5276.

Gelfand, A. E., A. M. Schmidt, S. Wu, J. A. Silander, A. Latimer, and A.
G. Rebelo. 2005. Modelling species diversity through species level
hierarchical modelling. Journal of the Royal Statistical Society: Series
C (Applied Statistics) 54:1--20.

Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
Informative Default Prior Distribution for Logistic and Other Regression
Models. The Annals of Applied Statistics 2:1360--1383.

Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
non-randomness in species Co-occurrences on Islands. Oecologia
52:75--84.

Gotelli, N. J., and G. L. Entsminger. 2003. Swap algorithms in null
model analysis. Ecology:532--535.

Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
tool to identify non-random species associations. Oecologia
162:463--477.

Harris, D. J. 2015. Generating realistic assemblages with a Joint
Species Distribution Model. Methods in Ecology and Evolution.

Harris, D. J., K. G. Smith, and P. J. Hanly. 2011. Occupancy is
nine-tenths of the law: Occupancy rates determine the homogenizing and
differentiating effects of exotic species. The American naturalist
177:535.

Harris, T. E. 1974. Contact Interactions on a Lattice. The Annals of
Probability 2:969--988.

Lee, E. D., C. P. Broedersz, and W. Bialek. 2013. Statistical mechanics
of the US Supreme Court. arXiv:1306.5004 {[}cond-mat, physics:physics,
q-bio{]}.

Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
Models.

Lewin, R. 1983. Santa Rosalia Was a Goat Ecologists have for two decades
made assumptions about the importance of competition in community
organization; that idea is now under vigorous attack. Science
221:636--639.

Loh, P.-L., and M. J. Wainwright. 2013. Structure estimation for
discrete graphical models: Generalized covariance matrices and their
inverses. The Annals of Statistics 41:3022--3049.

MacArthur, R. H. 1958. Population ecology of some warblers of
northeastern coniferous forests. Ecology 39:599--619.

Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
MIT Press.

R Core Team. 2015. R: A Language and Environment for Statistical
Computing. R Foundation for Statistical Computing, Vienna, Austria.

Roughgarden, J. 1983. Competition and Theory in Community Ecology. The
American Naturalist 122:583--601.

Salakhutdinov, R. 2008. Learning and evaluating Boltzmann machines.
Technical Report UTML TR 2008-002, Department of Computer Science,
University of Toronto, Dept. of Computer Science, University of Toronto.

Schneidman, E., M. J. Berry, R. Segev, and W. Bialek. 2006. Weak
pairwise correlations imply strongly correlated network states in a
neural population. Nature 440:1007--1012.

Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.
Ecological communities: Conceptual issues and the evidence. Princeton
University Press.

Tjelmeland, H., and J. Besag. 1998. Markov Random Fields with
Higher-order Interactions. Scandinavian Journal of Statistics
25:415--433.

Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
Community Structure in Alaskan Seabird Colonies. Ecology 62:1515--1524.

Wieringen, W. N. van, and C. F. Peeters. 2014. Ridge Estimation of
Inverse Covariance Matrices from High-Dimensional Data. arXiv preprint
arXiv:1403.0904.

Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}

\end{document}

### 2015-05-20 Harris_2015/models.Rmd
Add 3-species code
save model results better
add manuscript materials

@@
   # along the diagonal, so these values are named accordingly.
-  c(upper = b, diag = a)
+  c(alpha  = a, beta = b)
 }
@@
   # "True" interaction strengths, to save for later
-  truth = par[1:choose(n_spp, 2)]
+  truth = par[1:n_spp]
 
   # "True" intercepts, possibly adjusted below by environment
-  alpha = par[-(1:choose(n_spp, 2))]
+  alpha = par[-(1:n_spp)]
 
@@
 
-  # Landscape starts as if betas were all zero
+  # Landscape starts as if betas were all zero. Each species' occurrence probability
+  # depends on its alpha value and on the environment (assuming alpha_env is not zero).
   x = matrix(
@@
     i = i + 1
+
+    # Each round of Gibbs sampling updates one species (column) across all sites
+    # according to its conditional probability (i.e. conditional on environment
+    # and the other species that are present).
     for(j in 1:n_spp){
@@
 
+You can use the code below on the same simulated landscapes that I used by unzipping
+the `fakedata.zip` file included with this file.
+
 ## R-compatible methods
@@
 
-```{r, cache=TRUE, eval = FALSE}
+```{r, cache=TRUE}
 library("rosalia")
@@
   # which has all the beta values.
-  relist(fit$par)$upper
+  fit$beta[upper.tri(fit$beta)]
 }
@@
 ```{r, cache = TRUE}
+library(parallel)
 fit_model = function(method_name, csv_names){
   fit = get(paste("fit", method_name, sep = "_"))
-  out_list = lapply(
+  out_list = mclapply(
     csv_names, 
@@
         truth = unlist(truth),
-        predicted = fit(read.csv(full_path, row.names = 1)),
+        predicted = fit(as.matrix(read.csv(full_path, row.names = 1))),
         method = method_name,
@@
 partial_results = fit_model("partial", dir("fakedata/matrices"))
+```
+
+```{r, cache = TRUE}
+rosalia_results = fit_model("rosalia", grep("^[^2]", dir("fakedata/matrices"), value = TRUE))
 ```
@@
 
+```{r}
+str(rosalia_results)
+```
+
 ```{r, cache = TRUE}
@@
   lapply(
-    list(cov_results, partial_results, pairs_results),
+    list(
+      rosalia_results, 
+      cov_results, 
+      partial_results, 
+      pairs_results
+    ),
     function(x){
@@
   )
+
+write.csv(summarized_results, "summarized_results.csv")
 
@@
 ```
+
+How often does each method get the sign of the interaction correct?
+```{r, cache = TRUE}
+sapply(
+  list(partial = partial_results, pairs = pairs_results, rosalia = rosalia_results, cov = cov_results), 
+  function(x){
+    signif(100 * mean(sign(x$predicted) == sign(x$truth)), 3)
+  }
+)
+```
+
+Compare with the percent correct one would obtain "by chance" if one knew that 75%
+of the true interactions were negative
+```{r, cache = TRUE}
+signif(
+   100 *mean(rbinom(10000, size = 1, prob = .75) == rbinom(10000, size = 1, prob = .75)),
+  3
+)
+```

### 2015-05-20 Harris_2015/manuscript-materials/pdf.sh

@@ -1,0 +1,11 @@
+# bibtex ids with colons in them (eg R package names) don't work. Remove the pattern :_
+sed -i "" "s/:_/_/g" My\ Library.bib
+
+pandoc -o draft.tex --template=mytemplate.latex --bibliography=My\ Library.bib --csl=ecology.csl draft.Rmd
+
+pandoc -o draft.pdf --template=mytemplate.latex --bibliography=My\ Library.bib --csl=ecology.csl draft.Rmd
+
+pandoc -o Appendices.pdf --bibliography=My\ Library.bib --csl=ecology.csl Appendices.Rmd
+
+open draft.pdf
+open Appendices.pdf

### 2015-05-20 R/exact.R
coerce to matrix, more docs

@@ -95,7 +95,7 @@ nlp = function(par, rows, possible_cooc, prior, ...){
 
 # Gradient of the negative log-likelihood
 nll_grad = function(par, possible_cooc, observed_cooc, n_samples, ...){
-  
+
   # Find energy for all possible co-occurrence patterns
   E = -c(par %*% possible_cooc)
   
@@ -132,7 +132,8 @@ find_observed_cooc = function(x){
 #' 
 #' @description 
 #' The optimization is performed using the \code{BFGS} method in the \code{\link[stats]{optim}}
-#' function.
+#' function. Note that the likelihood function includes \code{2^N} terms for \code{N} nodes,
+#' so parameter estimation can be very slow with more than about 20-25 nodes.
 #' 
 #' @param x A binary matrix.  Each row corresponds to an observed sample and each
 #' column corresponds to one node in the observed network.
@@ -175,6 +176,10 @@ rosalia = function(
   n_samples = nrow(x)
   
   assert_that(all(x == 1 | x == 0))
+  if(!is.matrix(x)){
+    message("coercing x to matrix")
+    x = as.matrix(x)
+  }
   
   if (n_nodes > 25) {
     message(

### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
more manuscript

@@
 together on the landscape because they both grow best in areas that are not overshadowed
-by trees. Due to fundamental constraints on the structure of valid
-covariance matrices [@brown_constraints_2010], this kind of positive correlation
-is almost inevitable, even when all the interactions are negative. For this
-reason, knowing that the two shrubs co-occur more than expected by chance
-under some null model (e.g. that their correlation in the data set is positive)
-can not---by itself---tell us much about their direct effects on one another.
+by trees. Thus, even significant deviations from independence cannot---by 
+themselves---tell us much about species' direct effects on one another.
 
@@
 deficit of co-occurrences, and *among unshaded sites*, there will also be such
-a deficit. These *conditional* ("all-else-equal") relationships can be estimated with partial
-correlations [@albrecht_spatial_2001], although this approach is best-suited
-for Gaussian variables.  For non-Gaussian data, a more general approach is
-needed.
+a deficit.
 
@@
 understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them from observational data. Markov networks,
+interactions, and for detecting them with observational data. Markov networks,
 which generalize partial correlations to non-Gaussian data, have been used in
@@
 Gaussian. While ecologists explored some related approaches in the 1980's
-[@whittam_species_1981], computational limitations had previously forced
-researchers to rely on severe approximations that produced unintelligible
+[@whittam_species_1981], computational limitations had previously imposed
+severe approximations that produced unintelligible
 results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
@@
 The most familiar conditional relationship
-is the partial correlation, which indicates the portion of the correlation
+is the partial correlation, which indicates the portion of the sample correlation
 between two species that remains after controlling for other variables in the
-data set [@albrecht_spatial_2001], as opposed to the simple correlation, which
-does not control for other variables. The example with the shrubs and trees in
+data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
 Figure 1 shows how the two correlation measures can have opposite signs, 
 and suggests that the partial correlation is more relevant
-for drawing inferences about species interactions (e.g. competition). To the
-extent that our observations are not multivariate Gaussian (e.g. 
-presence-absence data), partial covariances do not describe variables' conditional
-relationships exactly, and a more general approach is needed. Markov networks
-allow researchers to generalize conditional relationships to non-Gaussian
-networks, much as generalized linear models allow researchers to extend
-regression to non-Gaussian response variables [@lee_learning_2012].
+for drawing inferences about species interactions (e.g. competition). 
+Markov networks extend this approach to non-Gaussian data, much as
+generalized linear models do for linear regression [@lee_learning_2012].
 
@@
 
-A major feature of Markov networks is the fact that the conditional
+A major benefit of Markov networks is the fact that the conditional
 relationships between species can be read directly off the matrix of $\beta$
@@
 possible assemblages that could be produced by the model sum to one (bottom of
-Figure 2B). While the analyses in this paper involve exact computations, this
-approach quickly becomes infeasible as the number of species increases: with
+Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
+quickly becomes infeasible as the number of species increases: with
 $2^N$ possible assemblages of $N$ species the number of bookkeeping operations
@@
 [@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
-but these techniques are beyond the scope of this paper.
+but they are beyond the scope of this paper.
 
@@
 landscape with 100 sites by sampling exactly from a probability distribution
-defined by the interaction coefficients in that figure. Each of the methods
+defined by the interaction coefficients in that figure (Appendix A). Each of the methods
 described below (a Markov network, two correlation-based methods and a null modeling 
-approach) was then evaluated on its ability to correctly infer that all
-three species interactions were negative, despite the frequent co-occurrences
-of the two shrub species.
+approach) was then evaluated on its ability to correctly infer that 
+the two shrub species competed with one another, despite their frequent co-occurrence.
 
@@
 potentially-interacting species on landscapes composed of 20, 100, 500, or
-2500 observed communities (24 replicate simulations for each combination).
+2500 observed communities (24 replicate simulations for each combination; Appendix B).
 These simulated data sets span the range from small, single-observer data
 sets to large collaborative efforts such as the North American Breeding Bird
-Survey. As described in Appendix 1, I randomly drew the "true" coefficient
+Survey. As described in Appendix B, I randomly drew the "true" coefficient
 values for each replicate so that most species pairs interacted negligibly, a
@@
 landscapes with varying numbers of species and sites via Markov chain Monte
-Carlo (Appendix 1). For half of the simulated landscapes, I treated each
+Carlo (Appendix C). For half of the simulated landscapes, I treated each
 species' $\alpha$ coefficient as a constant, as described above. For the other
@@
 environmental factors that varied from location to location across the
-landscape (Appendix 1). The latter set of simulated landscapes provide an
+landscape (Appendix B). The latter set of simulated landscapes provide an
 important test of the methods' ability to distinguish co-occurrence patterns
@@
 ***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
-The previous sections involved known values of $\alpha$ and $\beta$. In most
-real cases, however, the true values of these parameters are not known, and
-ecologists  will need to estimate them from data. When the number of species
-is reasonably small, one can compute exact maximum likelihood estimates for
-all of the $\alpha$ and $\beta$ coefficients by optimizing Equation 1
-(Appendix 2).  Doing so allows us to find the unique set of $\alpha$ and
+The previous sections involved known values of $\alpha$ and $\beta$. In practice,
+ecologists will often need to estimate these parameters from data instead. 
+When the number of species is reasonably small, one can compute exact maximum 
+likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
+Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
 $\beta$ coefficients that would be expected to produce exactly the observed
-occurrence frequencies and co-occurrence frequencies. Fully-observed Markov
+co-occurrence frequencies. Fully-observed Markov
 networks like the ones considered here have unimodal likelihood surfaces

### 2015-05-20 Harris_2015/models.Rmd
more manuscript

@@
 
+# Use the full names of the methods for the plot below
+summarized_results$method = factor(
+  summarized_results$method, 
+  levels = c("rosalia", "partial", "cov", "Null"),
+  labels = c("Markov network", "Partial covariance", "Covariance", "Null")
+)
+
+# Use the full names of the methods for the plot below
+summarized_results$env = factor(
+  summarized_results$env, 
+  levels = c("FALSE", "TRUE"),
+  labels = c("No environment", "Enironment")
+)
+
 head(summarized_results) # Show the first 6 rows
@@
   scale_shape_manual(values = c(21, 25, 15, 18)) + 
-  coord_cartesian(ylim = c(-0.1, 1.01)) + 
+  coord_cartesian(ylim = c(-0.12, 1.01)) + 
   ylab(expression(R^2)) + 
@@
 
+## Additional assessment
+
 How often does each method get the sign of the interaction correct?

### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
ready to submit, maybe?

@@
 [@gotelli_empirical_2009]. More generally, it can be difficult to reason about
-the complex web of direct and indirect interactions that contribute to the
-structure of real assemblages, especially when these interactions occur
+the complex web of direct and indirect interactions in real assemblages,
+especially when these interactions occur
 against a background of other ecological processes such as dispersal and
@@
 against such a "null" background, however, rules out the possibility that the
-overall association between two species could be driven by an outside force,
-as shown in Figure 1. Even though the two shrub species in this example compete 
+overall association between two species could be driven by an outside force.
+For example, even though the two shrub species in Figure 1 compete 
 with one another for resources at a mechanistic level, they end up clustering
 together on the landscape because they both grow best in areas that are not overshadowed
-by trees. Thus, even significant deviations from independence cannot---by 
-themselves---tell us much about species' direct effects on one another.
+by trees. If this sort of effect is common, then significant deviations 
+from independence will not---by themselves---provide convincing evidence of
+species' direct effects on one another.
 
@@
 [@murphy_machine_2012], ensuring that this procedure will always converge on
-the global maximum.
+the global maximum. The optimization procedures described here and in Appendix C are 
+implemented in the *rosalia* package [@harris_rosalia_2015] for the 
+R programming language [@r_core_team_r_2015].
 
-For the analyses in this paper, I made a small modification to this maximum
-likelihood procedure. Given the large number of parameters associated with
-some of the networks to be estimated, I regularized the likelihood using a
-logistic prior distribution [@gelman_weakly_2008] with a scale of 1 on the
-$\alpha$ and $\beta$ terms.
-
-The optimization procedures described here and in Appendix 2 are implemented
-in the *rosalia* package^[source code available for download from
-https://github.com/davharris/rosalia] for the R programming language
-[@r_core_team_r_2015].
+For analyses with 5 or more species, I made a small modification to the maximum
+likelihood procedure described above. Given the large number of parameters 
+associated with some of the networks to be estimated, I regularized the 
+likelihood using a logistic prior distribution [@gelman_weakly_2008] with
+a scale of 1 on the $\alpha$ and $\beta$ terms.
 
@@
 used several other techniques for inferring the sign and strength of marginal
-associations between pairs of species.
+associations between pairs of species (Appendix B).
 
@@
 The third method, described in @gotelli_empirical_2009, involved simulating
-possible landscapes from a null model that retained the row and column sums of
+possible landscapes from a null model that retains the row and column sums of
 the original matrix [@strong_ecological_1984]. Using the default options in the
@@
 results, I changed its value to 32.5, which was the value of the next largest
-$Z$-score in the null model's results.]. Then, for each combination of species
-richness, landscape size, method, and presence/absence of environmental
-filtering, I calculated the mean squared error associated with the rescaled
-estimate across all 12 replicates. Finally, I calculated an $R^2$ value for
-each landscape by comparing this squared error estimate with the squared error
-associated with the null hypothesis that all interactions were zero.
+$Z$-score in the null model's results.]. For each estimate of a species pair's
+interactions, I used this regression to calculate the squared error associated
+with method that produced it.  Finally, I averaged these squared errors for
+each combination of species richness, landscape size, statistical method, and 
+presence/absence of environmental filtering across all 12 replicates; the mean
+squared errors associated with these subsets of the data determined the proportion
+of variance explained by each method under different conditions.
 
@@
 indirect pathway via the tree species and each correctly identified the direct
-negative interaction between the shrubs 92% of the time.
+negative interaction between the shrubs 94% of the time.
 
 ***Larger landscapes.***
-The Markov network consistently outperformed all of the other techniques that
-I evaluated (Figure 3). Of the remaining methods, partial covariances tended
-to be most accurate, especially when 100 or more communities had been observed
-on the landscape. $Z$-scores from the null model had the least predictive
-utility overall, as indicated by $R^2$ values. In general, all the models'
-accuracies declined when environmental filters were added, but the declines
-were smaller when environmentally-induced correlations were spread out among a
-larger number of pairwise interactions.
+The accuracy of the four evaluated methods varied substantially, depending
+on the parameters that produced the simulated communities (Figure 3).
+In general, however, there was a consistent ordering: the Markov network
+explained 54% of the variance overall, followed by partial covariances (32%),
+sample covariances (21%), and $Z$ scores from the null model (17%).  
+The models' accuracies tended to decline when environmental filters were added, 
+particularly when the number of species was small and the effects could not be 
+diluted among many pairwise interactions.
 
@@
 The results presented above are very promising, as they show that Markov
-networks can recover much of the variation in species' pairwise interaction
-strengths from observational data, even when direct interactions are largely
-overwhelmed by indirect effects (e.g. Figure 1) or environmental effects
-(lower panels of Figure 3). For cases where it is infeasible to fit a Markov
-network, these results also indicate that partial covariances---which can be
-computed straightforwardly by linear regression---can often provide an
-accurate approximation.
+networks can recover species' pairwise interactions from observational  data,
+even when direct interactions are largely overwhelmed by indirect effects
+(e.g. Figure 1) or environmental effects (lower panels of Figure 3). For cases
+where it is infeasible to fit a Markov network, these results also indicate
+that partial covariances---which can be computed straightforwardly by linear
+regression---can often provide an accurate approximation.
 
@@
 natural history or ecological experiments, which could substantially reduce
-the probability and magnitude of error. The rosalia package has built-in
-mechanisms for incorporating this kind of information, if it can be expressed
-as a prior probability distribution or a penalty on the likelihood.
+the probability and magnitude of error. The rosalia package [@harris_rosalia_2015]
+has built-in mechanisms for incorporating this kind of information, if it can 
+be expressed as a prior probability distribution or a penalty on the likelihood.
 
@@
 [@harris_generating_2015] can fit *approximate* Markov networks to large
-species assemblages (>100 species) while simultaneously modeling each species'
+species assemblages (>300 species) while simultaneously modeling each species'
 response to the abiotic environment with complex, nonlinear functions. This
@@
 networks---has been extremely well-studied in statistical physics for nearly a
-century, and the models' properties, capabilities, and limits are well-understood
-in a huge range of applications, from spatial lattices
-[@gelfand_modelling_2005] to haphazard neural connections in the retina and
-brain [@schneidman_weak_2006] to complex behavioral interactions among human
-beings [@lee_statistical_2013]. Modeling species interactions using the same framework
-would thus allow ecologists to tap into an enormous set of existing discoveries and
-techniques for dealing with indirect effects, stability, and alternative
-stable states.
+century, and the models' properties, capabilities, and limits are well-
+understood in a huge range of applications, from spatial modeling
+[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
+human behavior [@lee_statistical_2013; salakhutdinov_restricted_2007].
+Modeling species interactions using the same framework would thus allow
+ecologists to tap into an enormous set of existing discoveries and techniques
+for dealing with indirect effects, stability, and alternative stable states.
 
-Finally, the null models commonly used by ecologists for drawing inferences
-about pairwise species interactions do not seem to distinguish between direct
-and indirect interactions any better than simple correlation coefficients.
-Null and neutral models can be very useful for clarifying our thinking about the numerical
-consequences of species' richness and abundance patterns
-[@harris_occupancy_2011; @xiao_strong_2015], but it is less clear whether deviations
-from the null hypothesis can clearly implicate specific biological processes like 
-competition.
+These results also have important implications for the continued use of fixed-fixed
+null models in ecology.  The small simulated landscapes described by
+Figure 1 show that test statistics based on marginal co-occurrence (such as
+C-scores) will not always have a straightforward relationship with the
+underlying ecological processes.  Moreover, the larger communities analyzed in
+Figure 3 often fell so far outside the null distribution^[Nearly 20% of the
+species pairs fell outside the 99.99994% confidence intervals implied by their
+$Z$ scores ($|Z| > 5$), and about 10% had uncorrected p-values below R's
+default numerical precision of $2 \times 10^{-16}$.] that it probably makes 
+more sense to reject whole model rather than to assign blame for the discrepancy to
+any one species pair. On average, the pairwise $Z$ scores from the null
+model provided less information about direct species interactions than correlation
+coefficients did. Researchers using null modeling approaches may
+be able to predict twice as much of the variance in species' "true" interaction
+strengths using partial covariances from linear regression, or triple them 
+using a Markov network.
+
+Null and neutral models can be very useful for clarifying our thinking about
+the numerical consequences of species' richness and abundance patterns
+[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a null model
+must be interpreted with care [@roughgarden_competition_1983]. In complex
+networks of ecological interactions---and even in small networks with three
+species---it may simply not be possible to implicate individual species pairs
+or specific ecological processes like competition by rejecting a general-
+purpose null. Direct estimates of species' conditional associations may be the
+only way to make these inferences reliably.
 
@@
 with 5, 10, or 20 species arrayed across varying numbers of sampled locations
-when environmental filtering was absent (top row) or present (bottom  row).
+when environmental filtering was absent (top row) or present (bottom row). A
+negative $R^2$ values implies that the squared error associated with 
+the corresponding subset of the predictions was larger than the error one would
+get from assuming that all coefficients equalled zero.

### 2015-05-20 Harris_2015/manuscript-materials/draft.tex
ready to submit, maybe?

@@
 can be difficult to reason about the complex web of direct and indirect
-interactions that contribute to the structure of real assemblages,
-especially when these interactions occur against a background of other
-ecological processes such as dispersal and environmental filtering
-(Connor et al. 2013). For this reason, it isn't always clear what kinds
-of patterns would even constitute evidence of competition, as opposed to
-some other biological process or random sampling error (Lewin 1983,
-Roughgarden 1983).
+interactions in real assemblages, especially when these interactions
+occur against a background of other ecological processes such as
+dispersal and environmental filtering (Connor et al. 2013). For this
+reason, it isn't always clear what kinds of patterns would even
+constitute evidence of competition, as opposed to some other biological
+process or random sampling error (Lewin 1983, Roughgarden 1983).
 
@@
 possibility that the overall association between two species could be
-driven by an outside force, as shown in Figure 1. Even though the two
-shrub species in this example compete with one another for resources at
-a mechanistic level, they end up clustering together on the landscape
+driven by an outside force. For example, even though the two shrub
+species in Figure 1 compete with one another for resources at a
+mechanistic level, they end up clustering together on the landscape
 because they both grow best in areas that are not overshadowed by trees.
-Due to fundamental constraints on the structure of valid covariance
-matrices (Brown et al. 2010), this kind of positive correlation is
-almost inevitable, even when all the interactions are negative. For this
-reason, knowing that the two shrubs co-occur more than expected by
-chance under some null model (e.g.~that their correlation in the data
-set is positive) can not---by itself---tell us much about their direct
-effects on one another.
+If this sort of effect is common, then significant deviations from
+independence will not---by themselves---provide convincing evidence of
+species' direct effects on one another.
 
@@
 sites}, there will be a deficit of co-occurrences, and \emph{among
-unshaded sites}, there will also be such a deficit. These
-\emph{conditional} (``all-else-equal'') relationships can be estimated
-with partial correlations (Albrecht and Gotelli 2001), although this
-approach is best-suited for Gaussian variables. For non-Gaussian data, a
-more general approach is needed.
+unshaded sites}, there will also be such a deficit.
 
@@
 understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them from observational data. Markov
+interactions, and for detecting them with observational data. Markov
 networks, which generalize partial correlations to non-Gaussian data,
@@
 explored some related approaches in the 1980's (Whittam and
-Siegel-Causey 1981), computational limitations had previously forced
-researchers to rely on severe approximations that produced
-unintelligible results (e.g. ``probabilities'' greater than one; Gilpin
-and Diamond 1982). Now that it is computationally feasible to fit these
-models exactly, the approach has become worth a second look.
+Siegel-Causey 1981), computational limitations had previously imposed
+severe approximations that produced unintelligible results (e.g.
+``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
+it is computationally feasible to fit these models exactly, the approach
+has become worth a second look.
 
@@
 conditional relationship is the partial correlation, which indicates the
-portion of the correlation between two species that remains after
+portion of the sample correlation between two species that remains after
 controlling for other variables in the data set (Albrecht and Gotelli
-2001), as opposed to the simple correlation, which does not control for
-other variables. The example with the shrubs and trees in Figure 1 shows
-how the two correlation measures can have opposite signs, and suggests
-that the partial correlation is more relevant for drawing inferences
-about species interactions (e.g.~competition). To the extent that our
-observations are not multivariate Gaussian (e.g. presence-absence data),
-partial covariances do not describe variables' conditional relationships
-exactly, and a more general approach is needed. Markov networks allow
-researchers to generalize conditional relationships to non-Gaussian
-networks, much as generalized linear models allow researchers to extend
-regression to non-Gaussian response variables (Lee and Hastie 2012).
+2001). The example with the shrubs and trees in Figure 1 shows how the
+two correlation measures can have opposite signs, and suggests that the
+partial correlation is more relevant for drawing inferences about
+species interactions (e.g.~competition). Markov networks extend this
+approach to non-Gaussian data, much as generalized linear models do for
+linear regression (Lee and Hastie 2012).
 
@@
 
-A major feature of Markov networks is the fact that the conditional
+A major benefit of Markov networks is the fact that the conditional
 relationships between species can be read directly off the matrix of
@@
 ensures that the probabilities of all possible assemblages that could be
-produced by the model sum to one (bottom of Figure 2B). While the
-analyses in this paper involve exact computations, this approach quickly
-becomes infeasible as the number of species increases: with \(2^N\)
-possible assemblages of \(N\) species the number of bookkeeping
-operations required for exact inference quickly spirals exponentially
-into the billions. Numerous techniques are available for working with
-Markov networks that keep the computations tractable, either through
-analytic approximations (Lee and Hastie 2012) or Monte Carlo sampling
-(Salakhutdinov 2008), but these techniques are beyond the scope of this
-paper.
+produced by the model sum to one (bottom of Figure 2B). Calculating
+\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
+infeasible as the number of species increases: with \(2^N\) possible
+assemblages of \(N\) species the number of bookkeeping operations
+required for exact inference quickly spirals exponentially into the
+billions. Numerous techniques are available for working with Markov
+networks that keep the computations tractable, either through analytic
+approximations (Lee and Hastie 2012) or Monte Carlo sampling
+(Salakhutdinov 2008), but they are beyond the scope of this paper.
 
@@
 landscape with 100 sites by sampling exactly from a probability
-distribution defined by the interaction coefficients in that figure.
-Each of the methods described below (a Markov network, two
+distribution defined by the interaction coefficients in that figure
+(Appendix A). Each of the methods described below (a Markov network, two
 correlation-based methods and a null modeling approach) was then
-evaluated on its ability to correctly infer that all three species
-interactions were negative, despite the frequent co-occurrences of the
-two shrub species.
+evaluated on its ability to correctly infer that the two shrub species
+competed with one another, despite their frequent co-occurrence.
 
@@
 or 2500 observed communities (24 replicate simulations for each
-combination). These simulated data sets span the range from small,
-single-observer data sets to large collaborative efforts such as the
-North American Breeding Bird Survey. As described in Appendix 1, I
+combination; Appendix B). These simulated data sets span the range from
+small, single-observer data sets to large collaborative efforts such as
+the North American Breeding Bird Survey. As described in Appendix B, I
 randomly drew the ``true'' coefficient values for each replicate so that
@@
 I then used Gibbs sampling to randomly generate landscapes with varying
-numbers of species and sites via Markov chain Monte Carlo (Appendix 1).
+numbers of species and sites via Markov chain Monte Carlo (Appendix C).
 For half of the simulated landscapes, I treated each species' \(\alpha\)
@@
 environmental factors that varied from location to location across the
-landscape (Appendix 1). The latter set of simulated landscapes provide
+landscape (Appendix B). The latter set of simulated landscapes provide
 an important test of the methods' ability to distinguish co-occurrence
@@
 presence-absence data.}} The previous sections involved known values of
-\(\alpha\) and \(\beta\). In most real cases, however, the true values
-of these parameters are not known, and ecologists will need to estimate
-them from data. When the number of species is reasonably small, one can
-compute exact maximum likelihood estimates for all of the \(\alpha\) and
-\(\beta\) coefficients by optimizing Equation 1 (Appendix 2). Doing so
-allows us to find the unique set of \(\alpha\) and \(\beta\)
-coefficients that would be expected to produce exactly the observed
-occurrence frequencies and co-occurrence frequencies. Fully-observed
-Markov networks like the ones considered here have unimodal likelihood
-surfaces (Murphy 2012), ensuring that this procedure will always
-converge on the global maximum.
+\(\alpha\) and \(\beta\). In practice, ecologists will often need to
+estimate these parameters from data instead. When the number of species
+is reasonably small, one can compute exact maximum likelihood estimates
+for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
+Equation 1 (Appendix C). Doing so allows us to find the unique set of
+\(\alpha\) and \(\beta\) coefficients that would be expected to produce
+exactly the observed co-occurrence frequencies. Fully-observed Markov
+networks like the ones considered here have unimodal likelihood surfaces
+(Murphy 2012), ensuring that this procedure will always converge on the
+global maximum. The optimization procedures described here and in
+Appendix C are implemented in the \emph{rosalia} package
+({\textbf{???}}) for the R programming language (R Core Team 2015).
 
-For the analyses in this paper, I made a small modification to this
-maximum likelihood procedure. Given the large number of parameters
-associated with some of the networks to be estimated, I regularized the
-likelihood using a logistic prior distribution (Gelman et al. 2008) with
-a scale of 1 on the \(\alpha\) and \(\beta\) terms.
-
-The optimization procedures described here and in Appendix 2 are
-implemented in the \emph{rosalia} package\footnote{source code available
-  for download from https://github.com/davharris/rosalia} for the R
-programming language (R Core Team 2015).
+For analyses with 5 or more species, I made a small modification to the
+maximum likelihood procedure described above. Given the large number of
+parameters associated with some of the networks to be estimated, I
+regularized the likelihood using a logistic prior distribution (Gelman
+et al. 2008) with a scale of 1 on the \(\alpha\) and \(\beta\) terms.
 
@@
 several other techniques for inferring the sign and strength of marginal
-associations between pairs of species.
+associations between pairs of species (Appendix B).
 
@@
 The third method, described in Gotelli and Ulrich (2009), involved
-simulating possible landscapes from a null model that retained the row
+simulating possible landscapes from a null model that retains the row
 and column sums of the original matrix (Strong et al. 1984). Using the
@@
   its value to 32.5, which was the value of the next largest \(Z\)-score
-  in the null model's results.}. Then, for each combination of species
-richness, landscape size, method, and presence/absence of environmental
-filtering, I calculated the mean squared error associated with the
-rescaled estimate across all 12 replicates. Finally, I calculated an
-\(R^2\) value for each landscape by comparing this squared error
-estimate with the squared error associated with the null hypothesis that
-all interactions were zero.
+  in the null model's results.}. For each estimate of a species pair's
+interactions, I used this regression to calculate the squared error
+associated with method that produced it. Finally, I averaged these
+squared errors for each combination of species richness, landscape size,
+statistical method, and presence/absence of environmental filtering
+across all 12 replicates; the mean squared errors associated with these
+subsets of the data determined the proportion of variance explained by
+each method under different conditions.
 
@@
 correctly identified the direct negative interaction between the shrubs
-92\% of the time.
+94\% of the time.
 
-\textbf{\emph{Larger landscapes.}} The Markov network consistently
-outperformed all of the other techniques that I evaluated (Figure 3). Of
-the remaining methods, partial covariances tended to be most accurate,
-especially when 100 or more communities had been observed on the
-landscape. \(Z\)-scores from the null model had the least predictive
-utility overall, as indicated by \(R^2\) values. In general, all the
-models' accuracies declined when environmental filters were added, but
-the declines were smaller when environmentally-induced correlations were
-spread out among a larger number of pairwise interactions.
+\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
+methods varied substantially, depending on the parameters that produced
+the simulated communities (Figure 3). In general, however, there was a
+consistent ordering: the Markov network explained 54\% of the variance
+overall, followed by partial covariances (32\%), sample covariances
+(21\%), and \(Z\) scores from the null model (17\%).\\The models'
+accuracies tended to decline when environmental filters were added,
+particularly when the number of species was small and the effects could
+not be diluted among many pairwise interactions.
 
@@
 The results presented above are very promising, as they show that Markov
-networks can recover much of the variation in species' pairwise
-interaction strengths from observational data, even when direct
-interactions are largely overwhelmed by indirect effects (e.g.~Figure 1)
-or environmental effects (lower panels of Figure 3). For cases where it
-is infeasible to fit a Markov network, these results also indicate that
-partial covariances---which can be computed straightforwardly by linear
-regression---can often provide an accurate approximation.
+networks can recover species' pairwise interactions from observational
+data, even when direct interactions are largely overwhelmed by indirect
+effects (e.g.~Figure 1) or environmental effects (lower panels of Figure
+3). For cases where it is infeasible to fit a Markov network, these
+results also indicate that partial covariances---which can be computed
+straightforwardly by linear regression---can often provide an accurate
+approximation.
 
@@
 which could substantially reduce the probability and magnitude of error.
-The rosalia package has built-in mechanisms for incorporating this kind
-of information, if it can be expressed as a prior probability
-distribution or a penalty on the likelihood.
+The rosalia package ({\textbf{???}}) has built-in mechanisms for
+incorporating this kind of information, if it can be expressed as a
+prior probability distribution or a penalty on the likelihood.
 
@@
 species distribution modeling (Harris 2015) can fit \emph{approximate}
-Markov networks to large species assemblages (\textgreater{}100 species)
+Markov networks to large species assemblages (\textgreater{}300 species)
 while simultaneously modeling each species' response to the abiotic
@@
 nearly a century, and the models' properties, capabilities, and limits
-are well-understood in a huge range of applications, from spatial
-lattices (Gelfand et al. 2005) to haphazard neural connections in the
-retina and brain (Schneidman et al. 2006) to complex behavioral
-interactions among human beings (Lee et al. 2013). Modeling species
-interactions using the same framework would thus allow ecologists to tap
-into an enormous set of existing discoveries and techniques for dealing
-with indirect effects, stability, and alternative stable states.
+are well- understood in a huge range of applications, from spatial
+modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
+to models of human behavior {[}Lee et al. (2013);
+salakhutdinov\_restricted\_2007{]}. Modeling species interactions using
+the same framework would thus allow ecologists to tap into an enormous
+set of existing discoveries and techniques for dealing with indirect
+effects, stability, and alternative stable states.
 
-Finally, the null models commonly used by ecologists for drawing
-inferences about pairwise species interactions do not seem to
-distinguish between direct and indirect interactions any better than
-simple correlation coefficients. Null and neutral models can be very
-useful for clarifying our thinking about the numerical consequences of
-species' richness and abundance patterns (Harris et al. 2011, Xiao et
-al. 2015), but it is less clear whether deviations from the null
-hypothesis can clearly implicate specific biological processes like
-competition.
+These results also have important implications for the continued use of
+fixed-fixed null models in ecology. The small simulated landscapes
+described by Figure 1 show that test statistics based on marginal
+co-occurrence (such as C-scores) will not always have a straightforward
+relationship with the underlying ecological processes. Moreover, the
+larger communities analyzed in Figure 3 often fell so far outside the
+null distribution\footnote{Nearly 20\% of the species pairs fell outside
+  the 99.99994\% confidence intervals implied by their \(Z\) scores
+  (\(|Z| > 5\)), and about 10\% had uncorrected p-values below R's
+  default numerical precision of \(2 \times 10^{-16}\).} that it
+probably makes more sense to reject whole model rather than to assign
+blame for the discrepancy to any one species pair. On average, the
+pairwise \(Z\) scores from the null model provided less information
+about direct species interactions than correlation coefficients did.
+Researchers using the null modeling approach may be able to predict
+twice as much of the variance in species' ``true'' interaction strengths
+using partial covariances from linear regression or triple them using a
+Markov network.
+
+Null and neutral models can be very useful for clarifying our thinking
+about the numerical consequences of species' richness and abundance
+patterns (Harris et al. 2011, Xiao et al. 2015), but deviations from a
+null model must be interpreted with care (Roughgarden 1983). In complex
+networks of ecological interactions---and even in small networks with
+three species---it may simply not be possible to implicate individual
+species pairs or specific ecological processes like competition by
+rejecting a general- purpose null. Direct estimates of species'
+conditional associations may be the only way to make these inferences
+reliably.
 
@@
 across varying numbers of sampled locations when environmental filtering
-was absent (top row) or present (bottom row).
+was absent (top row) or present (bottom row). A negative \(R^2\) values
+implies that the squared error associated with the corresponding subset
+of the predictions was larger than the error one would get from assuming
+that all coefficients equalled zero.
 
@@
 pollen associations across the late Quaternary. Ecography 37:1095--1108.
-
-Brown, I. H., E. Bedrick, S. Morgan, I.-L. E. C. Ernest, and I. F.
-Kelly. 2010. Constraints on Negative Relationships: Mathematical Causes
-and Ecological. The Nature of Scientific Evidence: Statistical,
-Philosophical, and Empirical Considerations:298.
 

### 2015-05-20 Harris_2015/models.Rmd
ready to submit, maybe?

@@
 ```{r, cache = TRUE}
-rosalia_results = fit_model("rosalia", grep("^[^2]", dir("fakedata/matrices"), value = TRUE))
+rosalia_results = fit_model("rosalia", dir("fakedata/matrices"))
 ```
@@
     function(x){
-      residuals = residuals(
-        lm(
+      linear_model = lm(
         truth ~ predicted + 0, 
@@
       )
-      )
+      cat(x$method[[1]])
+      print(summary(linear_model))
+      cat("\n", "\n", "\n")
+      residuals = residuals(linear_model)
       cbind(x, residuals = residuals)
@@
 ```
+
+What percentage of absolute $Z$ scores exceed 5?
+
+```{r}
+mean(abs(pairs_results$predicted) > 5)
+```
+
+What percentage of absolute $Z$ scores produce p-values below `.Machine$double.eps`?
+
+```{r}
+.Machine$double.eps
+
+p_values = pnorm(1 - abs(pairs_results$predicted))
+mean(p_values < .Machine$double.eps)
+```

### 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
version bump, manuscript tweaks, final figure

@@
 [@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
-human behavior [@lee_statistical_2013; salakhutdinov_restricted_2007].
+human behavior [@lee_statistical_2013; @salakhutdinov_restricted_2007].
 Modeling species interactions using the same framework would thus allow

### 2015-05-20 Harris_2015/manuscript-materials/draft.tex
version bump, manuscript tweaks, final figure

@@
 about direct species interactions than correlation coefficients did.
-Researchers using the null modeling approach may be able to predict
-twice as much of the variance in species' ``true'' interaction strengths
-using partial covariances from linear regression or triple them using a
+Researchers using null modeling approaches may be able to predict twice
+as much of the variance in species' ``true'' interaction strengths using
+partial covariances from linear regression, or triple them using a
 Markov network.

### 2015-05-21 Harris_2015/manuscript-materials/Appendices.Rmd
I think this one's actually ready to submit.

@@
-# Appendices
-
-## 1. Simlulated community structures
-
-I randomly assigned each species pair an interaction magnitude, $\beta_{ij}$, 
-drawn from an exponential distribution with rate parameter 1.  I also randomly
-assigned three quarters of the interaction coefficients to be negative, and the
-remaining quarter of the coefficients to be positive.
-
-For the simulated landscapes where the abiotic environment was constant across
-locations, each species' $\alpha$ coefficient was drawn from a normal
-distribution with mean -1 and standard deviation 1.  For the remaining
-landscapes, two environmental variables, $x_1$ and $x_2$ were sampled from
-independent standard normals with different values for each location. Each
-species' response to these two environmental variables was a linear function
-of these two environmental variables, with coefficients drawn from normal
-distributions with mean 0 and standard deviation 2.  In this way,  species'
-$\alpha$ coefficients (and thus their occurrence probabilities) depended on
-external environmental factors.
-
-Once the "true" coefficients had been generated for each site on the
-landscape, I generated possible landscapes using Gibbs sampling.  In each
-round of Gibbs sampling, I cycled through all the species, randomly updating
-each one's presence/absence vector in response to its conditional occurrence
-probability:
-
-$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$
-
-where the logistic function is $\frac{1}{1+e^{-x}}$ [@murphy_machine_2012]. After 1000 rounds of
-sampling, I continued this procedure until I obtained a landscape matrix where
-all of the species occurred at least once. I treated the resulting matrices as
-"observed" data for analysis.
-
 ## 2. Log-likelihood gradient

### 2015-05-21 Harris_2015/manuscript-materials/draft.Rmd
I think this one's actually ready to submit.

--- 2015-05-20 Harris_2015/manuscript-materials/draft.Rmd
@@
 landscapes with varying numbers of species and sites via Markov chain Monte
-Carlo (Appendix C). For half of the simulated landscapes, I treated each
+Carlo (Appendix B). For half of the simulated landscapes, I treated each
 species' $\alpha$ coefficient as a constant, as described above. For the other
@@
 likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing 
-Equation 1 (Appendix C).  Doing so allows us to find the unique set of $\alpha$ and
-$\beta$ coefficients that would be expected to produce exactly the observed
-co-occurrence frequencies. Fully-observed Markov
+$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
 networks like the ones considered here have unimodal likelihood surfaces
 [@murphy_machine_2012], ensuring that this procedure will always converge on
-the global maximum. The optimization procedures described here and in Appendix C are 
-implemented in the *rosalia* package [@harris_rosalia_2015] for the 
-R programming language [@r_core_team_r_2015].
+the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
+coefficients that would be expected to produce exactly the observed co-occurrence 
+frequencies. For the analyses in this paper, I used the *rosalia* package 
+[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
+define the objective function and gradient as R code.  The rosalia package then 
+uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
+and $\beta$.
 
@@
 [@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
-human behavior [@lee_statistical_2013; salakhutdinov_restricted_2007].
+human behavior [@lee_statistical_2013].
 Modeling species interactions using the same framework would thus allow

### 2015-05-21 Harris_2015/manuscript-materials/draft.tex
I think this one's actually ready to submit.

@@
 I then used Gibbs sampling to randomly generate landscapes with varying
-numbers of species and sites via Markov chain Monte Carlo (Appendix C).
+numbers of species and sites via Markov chain Monte Carlo (Appendix B).
 For half of the simulated landscapes, I treated each species' \(\alpha\)
@@
 for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
-Equation 1 (Appendix C). Doing so allows us to find the unique set of
-\(\alpha\) and \(\beta\) coefficients that would be expected to produce
-exactly the observed co-occurrence frequencies. Fully-observed Markov
-networks like the ones considered here have unimodal likelihood surfaces
-(Murphy 2012), ensuring that this procedure will always converge on the
-global maximum. The optimization procedures described here and in
-Appendix C are implemented in the \emph{rosalia} package
-({\textbf{???}}) for the R programming language (R Core Team 2015).
+\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
+ones considered here have unimodal likelihood surfaces (Murphy 2012),
+ensuring that this procedure will always converge on the global maximum.
+This maximum is the unique combination of \(\alpha\) and \(\beta\)
+coefficients that would be expected to produce exactly the observed
+co-occurrence frequencies. For the analyses in this paper, I used the
+\emph{rosalia} package (Harris 2015a) for the R programming language (R
+Core Team 2015) to define the objective function and gradient as R code.
+The rosalia package then uses the \texttt{BFGS} method in R's
+\texttt{optim} function to find the best values for \(\alpha\) and
+\(\beta\).
 
@@
 which could substantially reduce the probability and magnitude of error.
-The rosalia package ({\textbf{???}}) has built-in mechanisms for
+The rosalia package (Harris 2015a) has built-in mechanisms for
 incorporating this kind of information, if it can be expressed as a
@@
 likelihood exactly. For example, the mistnet software package for joint
-species distribution modeling (Harris 2015) can fit \emph{approximate}
+species distribution modeling (Harris 2015b) can fit \emph{approximate}
 Markov networks to large species assemblages (\textgreater{}300 species)
@@
 modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
-to models of human behavior {[}Lee et al. (2013);
-salakhutdinov\_restricted\_2007{]}. Modeling species interactions using
-the same framework would thus allow ecologists to tap into an enormous
-set of existing discoveries and techniques for dealing with indirect
-effects, stability, and alternative stable states.
+to models of human behavior (Lee et al. 2013). Modeling species
+interactions using the same framework would thus allow ecologists to tap
+into an enormous set of existing discoveries and techniques for dealing
+with indirect effects, stability, and alternative stable states.
 
@@
 \textbf{\emph{References:}}
-
-\textbf{\emph{Figure captions}}
-
-\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, as indicated
-by their negative coefficient (circled), but this effect is
-substantially weaker. \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network allows us to correctly identify the negative
-interaction between these two species (circled).
-
-\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
-species. The depicted abiotic environment favors the occurrence of both
-species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
-these two species implies that they co-occur less than expected under
-independence. \textbf{B.} Relative probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. The exponent
-includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
-not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
-we can find the expected frequencies of all possible co-occurrence
-patterns between the two species of interest. \textbf{D.} If \(\beta\)
-equaled zero (e.g.~if the species no longer competed for the same
-resources), then the reduction in competition would allow each species
-to increase its occurrence rate and the deficit of co-occurrences would
-be eliminated.
-
-\textbf{\emph{Figure 3.}} Proportion of variance in interaction
-coefficients explained by each method with 5, 10, or 20 species arrayed
-across varying numbers of sampled locations when environmental filtering
-was absent (top row) or present (bottom row). A negative \(R^2\) values
-implies that the squared error associated with the corresponding subset
-of the predictions was larger than the error one would get from assuming
-that all coefficients equalled zero.
 
@@
 
-Harris, D. J. 2015. Generating realistic assemblages with a Joint
+Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
+networks. R package version 0.1.0. Zenodo.
+http://dx.doi.org/10.5281/zenodo.17808.
+
+Harris, D. J. 2015b. Generating realistic assemblages with a Joint
 Species Distribution Model. Methods in Ecology and Evolution.
@@
 
-Lewin, R. 1983. Santa Rosalia Was a Goat Ecologists have for two decades
-made assumptions about the importance of competition in community
-organization; that idea is now under vigorous attack. Science
-221:636--639.
+Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
+
+\textbf{\emph{Figure captions}}
+
+\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
+species. The tree (top) tends not to co-occur with either of the two
+shrub species, as indicated by the strongly negative coefficient linking
+them. The two shrub species also compete with one another, as indicated
+by their negative coefficient (circled), but this effect is
+substantially weaker. \textbf{B.} In spite of the competitive
+interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors
+positively correlated (circled). \textbf{C.} Controlling for the tree
+species' presence with a conditional method such as a partial covariance
+or a Markov network allows us to correctly identify the negative
+interaction between these two species (circled).
+
+\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
+species. The depicted abiotic environment favors the occurrence of both
+species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
+these two species implies that they co-occur less than expected under
+independence. \textbf{B.} Relative probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. The exponent
+includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
+not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
+\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
+\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
+constant \(Z\), ensures that the four relative probabilities sum to 1.
+In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
+we can find the expected frequencies of all possible co-occurrence
+patterns between the two species of interest. \textbf{D.} If \(\beta\)
+equaled zero (e.g.~if the species no longer competed for the same
+resources), then the reduction in competition would allow each species
+to increase its occurrence rate and the deficit of co-occurrences would
+be eliminated.
+
+\textbf{\emph{Figure 3.}} Proportion of variance in interaction
+coefficients explained by each method with 5, 10, or 20 species arrayed
+across varying numbers of sampled locations when environmental filtering
+was absent (top row) or present (bottom row). A negative \(R^2\) values
+implies that the squared error associated with the corresponding subset
+of the predictions was larger than the error one would get from assuming
+that all coefficients equalled zero.
+
 \end{flushleft}

### 2015-05-21 Harris_2015/models.Rmd
version bump, manuscript tweaks, final figure
I think this one's actually ready to submit.

@@
 ---
-# Simulating landscapes from known coefficients
+# Analyses for Figure 3
+
+## Simulating landscapes from known coefficients
 
@@
 I used the following function to generate "true" parameters with the method above, 
-simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo,
-and save the results to a "fakedata" folder.
+simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.
+
+I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
+In each round of Gibbs sampling, I cycled through all the species, randomly
+updating each one's presence/absence vector in response to its conditional
+occurrence probability:
+
+$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$
+
+where the logistic function is $\frac{1}{1+e^{-x}}$.
 
@@
 
-# Model fitting
+## Model fitting
 
@@
 
-## R-compatible methods
+### R-compatible methods
 
@@
 
-### Rosalia
+***Rosalia***
 
@@
 
-### Partial covariance
+***Partial covariance***
 
@@
 
-### Sample covariance
+***Sample covariance***
 
@@
 
-## Analyzing simulated landscapes with the first three models
+### Analyzing simulated landscapes with the first three models
 

### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
more tweaks

@@
 each species pair against the model's estimate, re-weighting the pairs so that
-each landscape contributed equally to the rescaled estimate^[The null model
-generated one $Z$-score outlier greater than 1000, which dominated the
-regression and squared error analyses. To reduce its influence on these
-results, I changed its value to 32.5, which was the value of the next largest
-$Z$-score in the null model's results.]. For each estimate of a species pair's
+each landscape contributed equally to the rescaled estimate. For each 
+estimate of a species pair's
 interactions, I used this regression to calculate the squared error associated
@@
 In general, however, there was a consistent ordering: the Markov network
-explained 54% of the variance overall, followed by partial covariances (32%),
-sample covariances (21%), and $Z$ scores from the null model (17%).  
-The models' accuracies tended to decline when environmental filters were added, 
-particularly when the number of species was small and the effects could not be 
-diluted among many pairwise interactions.
+explained 54% of the variance in the "true" parameters, followed by 
+partial covariances (33%), and sample covariances (22%).
+
+The null model scores initially explained 12% of the variation. After manually
+reducing the value of one especially strong outlier ($Z=1004$, implying
+$p<10^{-1000000}$), this increased to 17%.  Figure 3 reflects the adjusted
+version of the results.
 
@@
 even when direct interactions are largely overwhelmed by indirect effects
-(e.g. Figure 1) or environmental effects (lower panels of Figure 3). For cases
-where it is infeasible to fit a Markov network, these results also indicate
+(Figure 1) or environmental effects (lower panels of Figure 3). For cases
+where fitting a Markov network is infeasible, these results also indicate
 that partial covariances---which can be computed straightforwardly by linear
-regression---can often provide an accurate approximation.
-
-Apart from the environmental filters, the simulated landscapes presented here
-represent the best-case scenario for these methods.  Future research should
-thus examine these models' performance characteristics when the "true"
+regression---can often provide a surprisingly accurate approximation.
+On the other hand, inferences may be much less reliable in cases where the "true"
 interaction matrices include guild structure or trophic levels, which could
 make the $\beta$ coefficients much more difficult to infer (particularly for
-linear approximations like the partial covariance approach;
-@loh_structure_2013). On the other hand, ecologists may often have prior 
-information about the nature of real species' interaction patterns from 
-natural history or ecological experiments, which could substantially reduce
-the probability and magnitude of error. The rosalia package [@harris_rosalia_2015]
-has built-in mechanisms for incorporating this kind of information, if it can 
-be expressed as a prior probability distribution or a penalty on the likelihood.
+linear methods; @loh_structure_2013). In these cases, it may be necessary to
+include prior information about the nature of species' interactions from 
+natural history, experiments, or trait data. The rosalia package 
+[@harris_rosalia_2015] has built-in mechanisms for incorporating 
+this kind of information, if it can be expressed as a prior probability 
+distribution on $\beta$ or a penalty on the likelihood.
 
-Additionally, it is important to note that, while partial correlations and
-Markov networks both prevent us from mistaking marginal associations for
-conditional ones, they cannot tell us the underlying biological mechanism.
-Real species co-occurrence patterns will depend on a number of factors---especially
-in taxa that emigrate in response to other species---and the
-$\beta$ coefficients in Markov networks have to reduce this to a single
-number. Thus, experiments and natural history knowledge will generally be
-required to pin down the exact nature of the interaction (e.g. who outcompetes
-whom).
+Ecologists will also need experiments and natural history knowledge to pin
+down the exact nature of the interactions identified by the network (e.g. whether
+or not negative interactions are consumptive, or which species in a positively-associated
+pair is facilitating the other), particularly when real pairs of species can reciprocally 
+influence one another's occurrence probabilities in multiple ways simultaneously; the
+$\beta$ coefficients in Markov networks have to reduce this complexity to a single
+number. In short, partial correlations and Markov networks both prevent us 
+from mistaking marginal associations for conditional ones, but they can't tell us the 
+underlying biological mechanisms at work.
 
-Despite these limitations, the results with environmental filtering seem to
-indicate that the method can be very robust.  Additionally, the fact that
-Markov networks provide a likelihood function to optimize makes them highly
+Despite these limitations, Markov networks have enormous potential to improve
+ecological inferences.  For example, Markov networks provide a simple answer
+to the question of how competition should affect a species' overall
+prevalence, which was a major flash point for the null model debates in the
+1980’s [@strong_ecological_1984]. Equation 1 can be used to calculate the
+expected prevalence of a species in the absence of biotic influences
+[$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's effect on
+prevalence in a Markov network can thus be calculated by subtracting this value
+and the observed prevalence (cf Figure 2D).
+
+Markov networks---particularly the Ising model for binary networks---have been studied 
+in statistical physics for nearly a century [@cipra_introduction_1987], 
+and the models' properties, 
+capabilities, and limits are well-understood in a huge range of applications, 
+from spatial statistics [@gelfand_modelling_2005] to neuroscience 
+[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
+Modeling species interactions using the same framework would thus allow
+ecologists to tap into an enormous set of existing discoveries and techniques
+for dealing with indirect effects, stability, and alternative stable states
+(phase transitions).
+
+This modeling approach is also highly
 extensible, even when it is inconvenient to compute the likelihood exactly.
@@
 species assemblages (>300 species) while simultaneously modeling each species'
-response to the abiotic environment with complex, nonlinear functions. This
-sort of approach, which combines multiple ecological processes, could help
-ecologists to disentangle different factors behind the co-occurrence patterns
-we observe in nature. Numerous other extensions are possible: similar networks
-can be fit with continuous variables, count data, or both
+nonlinear response to the abiotic environment. Combining multiple ecological 
+processes into a common model could help ecologists to disentangle different 
+factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013]. 
+Numerous other extensions are possible: similar networks
+can be fit with a mix of discrete and continuous variables, for example
 [@lee_learning_2012]. There are even methods [@whittam_species_1981;
@@
 in an interaction matrix to vary as a function of the abiotic environment or
-of third-party species that could tip the balance between facilitation and
-exploitation [@bruno_inclusion_2003]. Fully exploring these possibilities will
-require more research into the various available approximations to the log-likelihood
-and to its gradient, in order to balance efficiency, accuracy, and
-the ability to generate confidence limits for statistical inference.
+of third-party species that tip the balance between facilitation and
+exploitation [@bruno_inclusion_2003].
 
-By providing precise quantitative expectations about the results of species
-interactions, Markov networks have the potential for addressing long-standing
-ecological questions. For example, Markov networks can provide a precise
-answer to the question of how competition affects species' overall prevalence,
-which was a major flash point for the null model debates in the 1980’s
-[@strong_ecological_1984]. From Equation 1, one can derive the expected
-prevalence of a species in the absence of biotic influences ($\frac{1}{1 +
-e^{-\alpha}}$). Any significant difference between this value and the observed
-prevalence can be attributed to the $\beta$ coefficients linking this species
-to its facilitators and competitors (cf Figure 2D).
-
-This paper only scratches the surface of what Markov networks can do for
-ecology. This family of models---particularly the Ising model for binary
-networks---has been extremely well-studied in statistical physics for nearly a
-century, and the models' properties, capabilities, and limits are well-
-understood in a huge range of applications, from spatial modeling
-[@gelfand_modelling_2005] to neuroscience [@schneidman_weak_2006] to models of
-human behavior [@lee_statistical_2013].
-Modeling species interactions using the same framework would thus allow
-ecologists to tap into an enormous set of existing discoveries and techniques
-for dealing with indirect effects, stability, and alternative stable states.
-
-These results also have important implications for the continued use of fixed-fixed
-null models in ecology.  The small simulated landscapes described by
-Figure 1 show that test statistics based on marginal co-occurrence (such as
-C-scores) will not always have a straightforward relationship with the
-underlying ecological processes.  Moreover, the larger communities analyzed in
-Figure 3 often fell so far outside the null distribution^[Nearly 20% of the
-species pairs fell outside the 99.99994% confidence intervals implied by their
-$Z$ scores ($|Z| > 5$), and about 10% had uncorrected p-values below R's
-default numerical precision of $2 \times 10^{-16}$.] that it probably makes 
-more sense to reject whole model rather than to assign blame for the discrepancy to
-any one species pair. On average, the pairwise $Z$ scores from the null
-model provided less information about direct species interactions than correlation
-coefficients did. Researchers using null modeling approaches may
-be able to predict twice as much of the variance in species' "true" interaction
-strengths using partial covariances from linear regression, or triple them 
-using a Markov network.
+Finally, the results presented here have important  implications for ecologists' 
+continued use of null models to draw inferences about species interactions. The
+small simulated data sets described by Figure 1 show that test statistics
+based on marginal co-occurrence (such as C-scores) will not always have a
+straightforward relationship with the underlying ecological processes. More
+generally, deviations from the null model generally provided less information
+about direct species interactions than correlation coefficients did.
+Scientists currently using null modeling approaches in their research may be
+able to capture twice as much of the variance in species' true interaction
+strengths using partial covariances from linear regression instead,  or three
+times as much using a Markov network.
 

### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
more tweaks

@@
 \setlength{\parindent}{0pt}
-\setlength{\parskip}{3pt}
+\setlength{\parskip}{5pt}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 re-weighting the pairs so that each landscape contributed equally to the
-rescaled estimate\footnote{The null model generated one \(Z\)-score
-  outlier greater than 1000, which dominated the regression and squared
-  error analyses. To reduce its influence on these results, I changed
-  its value to 32.5, which was the value of the next largest \(Z\)-score
-  in the null model's results.}. For each estimate of a species pair's
-interactions, I used this regression to calculate the squared error
-associated with method that produced it. Finally, I averaged these
-squared errors for each combination of species richness, landscape size,
-statistical method, and presence/absence of environmental filtering
-across all 12 replicates; the mean squared errors associated with these
-subsets of the data determined the proportion of variance explained by
-each method under different conditions.
+rescaled estimate. For each estimate of a species pair's interactions, I
+used this regression to calculate the squared error associated with
+method that produced it. Finally, I averaged these squared errors for
+each combination of species richness, landscape size, statistical
+method, and presence/absence of environmental filtering across all 12
+replicates; the mean squared errors associated with these subsets of the
+data determined the proportion of variance explained by each method
+under different conditions.
 
@@
 consistent ordering: the Markov network explained 54\% of the variance
-overall, followed by partial covariances (32\%), sample covariances
-(21\%), and \(Z\) scores from the null model (17\%).\\The models'
-accuracies tended to decline when environmental filters were added,
-particularly when the number of species was small and the effects could
-not be diluted among many pairwise interactions.
+in the ``true'' parameters, followed by partial covariances (33\%), and
+sample covariances (22\%).
+
+The null model scores initially explained 12\% of the variation. After
+manually reducing the value of one especially strong outlier
+(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%.
+Figure 3 reflects the adjusted version of the results.
 
@@
 data, even when direct interactions are largely overwhelmed by indirect
-effects (e.g.~Figure 1) or environmental effects (lower panels of Figure
-3). For cases where it is infeasible to fit a Markov network, these
-results also indicate that partial covariances---which can be computed
-straightforwardly by linear regression---can often provide an accurate
-approximation.
+effects (Figure 1) or environmental effects (lower panels of Figure 3).
+For cases where fitting a Markov network is infeasible, these results
+also indicate that partial covariances---which can be computed
+straightforwardly by linear regression---can often provide a
+surprisingly accurate approximation. On the other hand, inferences may
+be much less reliable in cases where the ``true'' interaction matrices
+include guild structure or trophic levels, which could make the
+\(\beta\) coefficients much more difficult to infer (particularly for
+linear methods; Loh and Wainwright (2013)). In these cases, it may be
+necessary to include prior information about the nature of species'
+interactions from natural history, experiments, or trait data. The
+rosalia package (Harris 2015a) has built-in mechanisms for incorporating
+this kind of information, if it can be expressed as a prior probability
+distribution on \(\beta\) or a penalty on the likelihood.
 
-Apart from the environmental filters, the simulated landscapes presented
-here represent the best-case scenario for these methods. Future research
-should thus examine these models' performance characteristics when the
-``true'' interaction matrices include guild structure or trophic levels,
-which could make the \(\beta\) coefficients much more difficult to infer
-(particularly for linear approximations like the partial covariance
-approach; Loh and Wainwright (2013)). On the other hand, ecologists may
-often have prior information about the nature of real species'
-interaction patterns from natural history or ecological experiments,
-which could substantially reduce the probability and magnitude of error.
-The rosalia package (Harris 2015a) has built-in mechanisms for
-incorporating this kind of information, if it can be expressed as a
-prior probability distribution or a penalty on the likelihood.
+Ecologists will also need experiments and natural history knowledge to
+pin down the exact nature of the interactions identified by the network
+(e.g.~whether or not negative interactions are consumptive, or which
+species in a positively-associated pair is facilitating the other),
+particularly when real pairs of species can reciprocally influence one
+another's occurrence probabilities in multiple ways simultaneously; the
+\(\beta\) coefficients in Markov networks have to reduce this complexity
+to a single number. In short, partial correlations and Markov networks
+both prevent us from mistaking marginal associations for conditional
+ones, but they can't tell us the underlying biological mechanisms at
+work.
 
-Additionally, it is important to note that, while partial correlations
-and Markov networks both prevent us from mistaking marginal associations
-for conditional ones, they cannot tell us the underlying biological
-mechanism. Real species co-occurrence patterns will depend on a number
-of factors---especially in taxa that emigrate in response to other
-species---and the \(\beta\) coefficients in Markov networks have to
-reduce this to a single number. Thus, experiments and natural history
-knowledge will generally be required to pin down the exact nature of the
-interaction (e.g.~who outcompetes whom).
+Despite these limitations, Markov networks have enormous potential to
+improve ecological inferences. For example, Markov networks provide a
+simple answer to the question of how competition should affect a
+species' overall prevalence, which was a major flash point for the null
+model debates in the 1980's (Strong et al. 1984). Equation 1 can be used
+to calculate the expected prevalence of a species in the absence of
+biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and Hastie 2012).
+Competition's effect on prevalence in a Markov network can thus be
+calculated by subtracting this value and the observed prevalence (cf
+Figure 2D).
 
-Despite these limitations, the results with environmental filtering seem
-to indicate that the method can be very robust. Additionally, the fact
-that Markov networks provide a likelihood function to optimize makes
-them highly extensible, even when it is inconvenient to compute the
-likelihood exactly. For example, the mistnet software package for joint
-species distribution modeling (Harris 2015b) can fit \emph{approximate}
-Markov networks to large species assemblages (\textgreater{}300 species)
-while simultaneously modeling each species' response to the abiotic
-environment with complex, nonlinear functions. This sort of approach,
-which combines multiple ecological processes, could help ecologists to
-disentangle different factors behind the co-occurrence patterns we
-observe in nature. Numerous other extensions are possible: similar
-networks can be fit with continuous variables, count data, or both (Lee
-and Hastie 2012). There are even methods (Whittam and Siegel-Causey
-1981, Tjelmeland and Besag 1998) that would allow the coefficient
-linking two species in an interaction matrix to vary as a function of
-the abiotic environment or of third-party species that could tip the
-balance between facilitation and exploitation (Bruno et al. 2003). Fully
-exploring these possibilities will require more research into the
-various available approximations to the log-likelihood and to its
-gradient, in order to balance efficiency, accuracy, and the ability to
-generate confidence limits for statistical inference.
+Markov networks---particularly the Ising model for binary
+networks---have been studied in statistical physics for nearly a century
+(Cipra 1987), and the models' properties, capabilities, and limits are
+well-understood in a huge range of applications, from spatial statistics
+(Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006) to models
+of human behavior (Lee et al. 2013). Modeling species interactions using
+the same framework would thus allow ecologists to tap into an enormous
+set of existing discoveries and techniques for dealing with indirect
+effects, stability, and alternative stable states (phase transitions).
 
-By providing precise quantitative expectations about the results of
-species interactions, Markov networks have the potential for addressing
-long-standing ecological questions. For example, Markov networks can
-provide a precise answer to the question of how competition affects
-species' overall prevalence, which was a major flash point for the null
-model debates in the 1980's (Strong et al. 1984). From Equation 1, one
-can derive the expected prevalence of a species in the absence of biotic
-influences (\(\frac{1}{1 + e^{-\alpha}}\)). Any significant difference
-between this value and the observed prevalence can be attributed to the
-\(\beta\) coefficients linking this species to its facilitators and
-competitors (cf Figure 2D).
+This modeling approach is also highly extensible, even when it is
+inconvenient to compute the likelihood exactly. For example, the mistnet
+software package for joint species distribution modeling (Harris 2015b)
+can fit \emph{approximate} Markov networks to large species assemblages
+(\textgreater{}300 species) while simultaneously modeling each species'
+nonlinear response to the abiotic environment. Combining multiple
+ecological processes into a common model could help ecologists to
+disentangle different factors that can confound simpler co-occurrence
+analyses (cf Connor et al. 2013). Numerous other extensions are
+possible: similar networks can be fit with a mix of discrete and
+continuous variables, for example (Lee and Hastie 2012). There are even
+methods (Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that
+would allow the coefficient linking two species in an interaction matrix
+to vary as a function of the abiotic environment or of third-party
+species that tip the balance between facilitation and exploitation
+(Bruno et al. 2003).
 
-This paper only scratches the surface of what Markov networks can do for
-ecology. This family of models---particularly the Ising model for binary
-networks---has been extremely well-studied in statistical physics for
-nearly a century, and the models' properties, capabilities, and limits
-are well- understood in a huge range of applications, from spatial
-modeling (Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006)
-to models of human behavior (Lee et al. 2013). Modeling species
-interactions using the same framework would thus allow ecologists to tap
-into an enormous set of existing discoveries and techniques for dealing
-with indirect effects, stability, and alternative stable states.
-
-These results also have important implications for the continued use of
-fixed-fixed null models in ecology. The small simulated landscapes
-described by Figure 1 show that test statistics based on marginal
-co-occurrence (such as C-scores) will not always have a straightforward
-relationship with the underlying ecological processes. Moreover, the
-larger communities analyzed in Figure 3 often fell so far outside the
-null distribution\footnote{Nearly 20\% of the species pairs fell outside
-  the 99.99994\% confidence intervals implied by their \(Z\) scores
-  (\(|Z| > 5\)), and about 10\% had uncorrected p-values below R's
-  default numerical precision of \(2 \times 10^{-16}\).} that it
-probably makes more sense to reject whole model rather than to assign
-blame for the discrepancy to any one species pair. On average, the
-pairwise \(Z\) scores from the null model provided less information
-about direct species interactions than correlation coefficients did.
-Researchers using null modeling approaches may be able to predict twice
-as much of the variance in species' ``true'' interaction strengths using
-partial covariances from linear regression, or triple them using a
-Markov network.
+Finally, the results presented here have important implications for
+ecologists' continued use of null models to draw inferences about
+species interactions. The small simulated data sets described by Figure
+1 show that test statistics based on marginal co-occurrence (such as
+C-scores) will not always have a straightforward relationship with the
+underlying ecological processes. More generally, deviations from the
+null model generally provided less information about direct species
+interactions than correlation coefficients did. Scientists currently
+using null modeling approaches in their research may be able to capture
+twice as much of the variance in species' true interaction strengths
+using partial covariances from linear regression instead, or three times
+as much using a Markov network.
 
@@
 \textbf{\emph{References:}}
+
+\textbf{\emph{Figure captions}}
+
+\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
+species. The tree (top) tends not to co-occur with either of the two
+shrub species, as indicated by the strongly negative coefficient linking
+them. The two shrub species also compete with one another, as indicated
+by their negative coefficient (circled), but this effect is
+substantially weaker. \textbf{B.} In spite of the competitive
+interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors
+positively correlated (circled). \textbf{C.} Controlling for the tree
+species' presence with a conditional method such as a partial covariance
+or a Markov network allows us to correctly identify the negative
+interaction between these two species (circled).
+
+\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
+species. The depicted abiotic environment favors the occurrence of both
+species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
+these two species implies that they co-occur less than expected under
+independence. \textbf{B.} Relative probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. The exponent
+includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
+not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
+\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
+\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
+constant \(Z\), ensures that the four relative probabilities sum to 1.
+In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
+we can find the expected frequencies of all possible co-occurrence
+patterns between the two species of interest. \textbf{D.} If \(\beta\)
+equaled zero (e.g.~if the species no longer competed for the same
+resources), then the reduction in competition would allow each species
+to increase its occurrence rate and the deficit of co-occurrences would
+be eliminated.
+
+\textbf{\emph{Figure 3.}} Proportion of variance in interaction
+coefficients explained by each method with 5, 10, or 20 species arrayed
+across varying numbers of sampled locations when environmental filtering
+was absent (top row) or present (bottom row). A negative \(R^2\) values
+implies that the squared error associated with the corresponding subset
+of the predictions was larger than the error one would get from assuming
+that all coefficients equalled zero.
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
-
-\textbf{\emph{Figure captions}}
-
-\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, as indicated
-by their negative coefficient (circled), but this effect is
-substantially weaker. \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network allows us to correctly identify the negative
-interaction between these two species (circled).
-
-\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
-species. The depicted abiotic environment favors the occurrence of both
-species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
-these two species implies that they co-occur less than expected under
-independence. \textbf{B.} Relative probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. The exponent
-includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
-not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
-we can find the expected frequencies of all possible co-occurrence
-patterns between the two species of interest. \textbf{D.} If \(\beta\)
-equaled zero (e.g.~if the species no longer competed for the same
-resources), then the reduction in competition would allow each species
-to increase its occurrence rate and the deficit of co-occurrences would
-be eliminated.
-
-\textbf{\emph{Figure 3.}} Proportion of variance in interaction
-coefficients explained by each method with 5, 10, or 20 species arrayed
-across varying numbers of sampled locations when environmental filtering
-was absent (top row) or present (bottom row). A negative \(R^2\) values
-implies that the squared error associated with the corresponding subset
-of the predictions was larger than the error one would get from assuming
-that all coefficients equalled zero.
-
 \end{flushleft}

### 2015-05-22 Harris_2015/models.Rmd
more tweaks

@@
 
-As noted in the manuscript, I manually altered one Z-score produced by the Pairs software because
-it was such a strong outlier that it completely dominated the regression model described below.
+As noted in the manuscript, the `pairs_results` include one outlier ($Z>1000$) that dominates the 
+subsequent analyses.
+```{r}
+head(sort(pairs_results$predicted, decreasing = TRUE))
+```
+
+With the outlier left as-is, the R-squared value is:
+
+```{r}
+summary(
+  lm(
+    truth ~ predicted + 0, 
+    data = pairs_results, 
+    weights = 1/choose(n_spp, 2),
+    y = TRUE
+)
+)$r.squared
+```
+
+For the analyses below, I adjusted the outlier to take on a somewhat more reasonable value.
 
 ```{r, cache = TRUE}
-# The largest predicted Z-score was over 1000; the next largest was about 32
-head(sort(pairs_results$predicted, decreasing = TRUE))
-pairs_results$predicted[which.max(pairs_results$predicted)] = 32.47
+pairs_results$predicted[which.max(pairs_results$predicted)] = 32.47 # Second largest result
 ```
@@
 
-```{r, cache = TRUE}
+```{r, cache = FALSE}
 
@@
       )
-      cat(x$method[[1]])
-      print(summary(linear_model))
-      cat("\n", "\n", "\n")
       residuals = residuals(linear_model)
@@
 
-```{r, cache = TRUE}
+```{r, message=FALSE}
 library(dplyr)
+```
 
+```{r, cache = FALSE}
 summarized_results = results %>% group_by_("n_spp", "n_loc", "method", "env") %>% 
@@
 
-## Additional assessment
-
-How often does each method get the sign of the interaction correct?
-```{r, cache = TRUE}
-sapply(
-  list(partial = partial_results, pairs = pairs_results, rosalia = rosalia_results, cov = cov_results), 
-  function(x){
-    signif(100 * mean(sign(x$predicted) == sign(x$truth)), 3)
-  }
-)
+```{r}
+summarized_results %>% group_by_("method") %>% 
+  summarize_(~round(mean(r2) * 100))
 ```
-
-Compare with the percent correct one would obtain "by chance" if one knew that 75%
-of the true interactions were negative
-```{r, cache = TRUE}
-signif(
-   100 *mean(rbinom(10000, size = 1, prob = .75) == rbinom(10000, size = 1, prob = .75)),
-  3
-)
-```
-
-What percentage of absolute $Z$ scores exceed 5?
-
-```{r}
-mean(abs(pairs_results$predicted) > 5)
-```
-
-What percentage of absolute $Z$ scores produce p-values below `.Machine$double.eps`?
-
-```{r}
-.Machine$double.eps
-
-p_values = pnorm(1 - abs(pairs_results$predicted))
-mean(p_values < .Machine$double.eps)
-```

### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
done

@@
 
-Ecologists' intense interest in drawing inferences about species interactions
----especially competition---from presence-absence data has a long history
-[@macarthur_population_1958; @diamond_island_1975; @connor_checkered_2013]. If
-nontrophic species interactions are important drivers of community assembly patterns,
-then we should expect to see their influence in our data sets. Despite decades
-of work and several major controversies, however [@lewin_santa_1983;
-@strong_ecological_1984; @gotelli_swap_2003], existing methods for detecting
-competition's effects on community structure are unreliable
-[@gotelli_empirical_2009]. More generally, it can be difficult to reason about
-the complex web of direct and indirect interactions in real assemblages,
-especially when these interactions occur
-against a background of other ecological processes such as dispersal and
-environmental filtering [@connor_checkered_2013]. For this reason, it isn't
-always clear what kinds of patterns would even constitute evidence of
-competition, as opposed to some other biological process or random sampling
-error [@roughgarden_competition_1983; @lewin_santa_1983].
+If nontrophic species interactions, such as competition, are important drivers
+of community assembly, then ecologists might expect to see their influence in
+our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
+decades of work and several major controversies, however [@lewin_santa_1983;
+@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
+methods for detecting competition's effects on community structure are
+unreliable [@gotelli_empirical_2009], and thus important ecological processes
+remain poorly understood. More generally, it can be difficult to reason about
+the complex web of direct and indirect species interactions in real
+assemblages, especially when these interactions occur against a background of
+other ecological processes such as dispersal and environmental filtering
+[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
+patterns would even constitute evidence of competition, as opposed to some
+other biological process or random sampling error
+[@roughgarden_competition_1983; @lewin_santa_1983].
 
 Most existing methods in this field compare the frequency with which two
-putative competitors are observed to co-occur against the frequency that would
+putative competitors are observed to co-occur, versus the frequency that would
 be expected if *all* species on the landscape were independent
@@
 overall association between two species could be driven by an outside force.
-For example, even though the two shrub species in Figure 1 compete 
-with one another for resources at a mechanistic level, they end up clustering
-together on the landscape because they both grow best in areas that are not overshadowed
-by trees. If this sort of effect is common, then significant deviations 
-from independence will not---by themselves---provide convincing evidence of
-species' direct effects on one another.
+For example, even though the two shrub species in Figure 1 compete with one
+another for resources at a mechanistic level, they end up clustering together
+on the landscape because they both grow best in areas that are not
+overshadowed by trees. If this sort of effect is common, then significant
+deviations from independence will not generally provide convincing
+evidence of species' direct effects on one another.
 
@@
 
-In this paper, I introduce Markov networks [undirected graphical models
-also known as Markov random fields; @murphy_machine_2012] as a framework for
+In this paper, I introduce Markov networks [undirected graphical models also
+known as Markov random fields; @murphy_machine_2012] as a framework for
 understanding the landscape-level consequences of pairwise species
 interactions, and for detecting them with observational data. Markov networks,
-which generalize partial correlations to non-Gaussian data, have been used in
-many scientific fields to model associations between various kinds of
-"particles". For example, a well-studied network called the Ising model has played
-an important role in our understanding of physics [where nearby particles tend
-to align magnetically with one another; @cipra_introduction_1987]. In spatial
+which generalize partial correlations to non-Gaussian data
+[@lee_learning_2012; @loh_structure_2013], have been used in many scientific
+fields for decades to model associations between various kinds of "particles".
+For example, a well-studied network named the Ising model has played an
+important role in our understanding of physics [where nearby particles tend to
+align magnetically with one another; @cipra_introduction_1987]. In spatial
 contexts, these models have been used to describe interactions between
@@
 @fort_statistical_2013, I suggest that ecologists could similarly treat
-species as the interacting particles in this modeling framework. Doing so
+species as the interacting particles in the same modeling framework. Doing so
 would allow ecologists to simulate and study the landscape-level consequences
@@
 [@whittam_species_1981], computational limitations had previously imposed
-severe approximations that produced unintelligible
-results [e.g. "probabilities" greater than one; @gilpin_factors_1982]. 
-Now that it is computationally feasible to fit these models
-exactly, the approach has become worth a second look.
+severe approximations that produced unintelligible results [e.g.
+"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
+computationally feasible to fit these models exactly, the approach has become
+worth a second look.
 
 The rest of the paper proceeds as follows. First, I discuss how Markov networks
-work and how they can be used to simulate landscape-level data and to predict
+work and how they can be used to simulate landscape-level data or to predict
 the direct and indirect consequences of possible interaction matrices. Then,
@@
 
-Here, $\alpha_{i}$ is the amount that the presence of species $i$ contributes
+Here, $\alpha_{i}$ is an intercept term determining the amount that 
+the presence of species $i$ contributes
 to the log-probability of $\vec{y}$; it directly controls the prevalence of
 species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
-species $i$ and species $j$ contributes to the log-probability, and controls
-how often the two species will be found together (Figure 2A, Figure 2B).
+species $i$ and species $j$ contributes to the log-probability; it controls
+the probability that the two species will be found together (Figure 2A, Figure 2B).
 $\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
@@
 quickly becomes infeasible as the number of species increases: with
-$2^N$ possible assemblages of $N$ species the number of bookkeeping operations
-required for exact inference quickly spirals exponentially into the billions.
+$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
+required for exact inference spirals exponentially into the billions and beyond.
 Numerous techniques are available for working with Markov networks that keep
-the computations tractable, either through analytic approximations
+the computations tractable, e.g. via analytic approximations
 [@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
@@
 
-The first set of simulated landscapes included the three competing
-species shown in Figure 1. For each of 1000 replicates, I generated a
-landscape with 100 sites by sampling exactly from a probability distribution
-defined by the interaction coefficients in that figure (Appendix A). Each of the methods
-described below (a Markov network, two correlation-based methods and a null modeling 
-approach) was then evaluated on its ability to correctly infer that 
-the two shrub species competed with one another, despite their frequent co-occurrence.
+The first set of simulated landscapes included the three competing species
+shown in Figure 1. For each of 1000 replicates, I generated a landscape with
+100 sites by sampling exactly from a probability distribution defined by the
+interaction coefficients in that figure (Appendix A). Each of the methods
+described below (a Markov network, two correlation-based methods and a null
+model) was then evaluated on its ability to correctly infer that the two
+shrub species competed with one another, despite their frequent co-occurrence.
 
@@
 common than facilitation. I then used Gibbs sampling to randomly generate
-landscapes with varying numbers of species and sites via Markov chain Monte
-Carlo (Appendix B). For half of the simulated landscapes, I treated each
+replicate landscapes with varying numbers of species and sites (Appendix B). 
+For half of the simulated landscapes, I treated each
 species' $\alpha$ coefficient as a constant, as described above. For the other
@@
 important test of the methods' ability to distinguish co-occurrence patterns
-that were generated from pairwise interactions among the observed species from
+that were generated from pairwise biotic interactions from
 those that were generated by external forces like abiotic environmental
@@
 
-The first two interaction measures were the simple and partial covariances
-between each pair of species' data vectors on the landscape
+The first two interaction measures were the sample covariances and the 
+partial covariances between each pair of species' data vectors on the landscape
 [@albrecht_spatial_2001]. Because partial covariances are undefined for
@@
 *non*-co-occurrences between two species). The software then calculated a
-$Z$-statistic for each species pair using this null distribution. After
+$Z$ statistic for each species pair using this null distribution. After
 multiplying this statistic by $-1$ so that positive values corresponded to
@@
 ***Method evaluation.***
-For the first simulated landscape (three species), I kept the evaluation
-simple and qualitative: any method that reliably determined that the two shrub
+I evaluated each method qualitatively on the simulated landscapes based on
+Figure 1: any method that reliably determined that the two shrub
 species were negatively associated passed; other methods failed.
@@
 each species pair against the model's estimate, re-weighting the pairs so that
-each landscape contributed equally to the rescaled estimate. For each 
-estimate of a species pair's
-interactions, I used this regression to calculate the squared error associated
-with method that produced it.  Finally, I averaged these squared errors for
-each combination of species richness, landscape size, statistical method, and 
-presence/absence of environmental filtering across all 12 replicates; the mean
-squared errors associated with these subsets of the data determined the proportion
-of variance explained by each method under different conditions.
+each landscape contributed equally to the rescaled estimate. For each  method,
+I calculated the squared deviations between each species pair's rescaled
+estimate and the "true" interactions used to generate the assemblages.
+Finally, I  averaged these squared errors for each combination of species
+richness, landscape size, statistical method, and presence/absence of
+environmental filtering across all 12 replicates; the mean squared errors
+associated with these subsets of the data determined the proportion of
+variance explained by each method under different conditions.
 
@@
 due to indirect effects of the dominant tree species. As a result, the
-covariance method falsely reported positive associations 94% of the time
+covariance method falsely reported positive associations on 94% of simulated landscapes,
 and the randomization-based null model falsely reported such associations 100%
@@
 on the parameters that produced the simulated communities (Figure 3).
-In general, however, there was a consistent ordering: the Markov network
+In general, however, there was a consistent ordering: overall, the Markov network
 explained 54% of the variance in the "true" parameters, followed by 
@@
 
-The null model scores initially explained 12% of the variation. After manually
-reducing the value of one especially strong outlier ($Z=1004$, implying
-$p<10^{-1000000}$), this increased to 17%.  Figure 3 reflects the adjusted
-version of the results.
+The null model scores initially explained only 12% of the variation. After
+manually reducing the value of one especially strong outlier ($Z=1004$,
+implying $p<10^{-1000000}$), this increased to 17% (Appendix B).  Figure 3
+reflects the adjusted version of the results.
 
@@
 
-The results presented above are very promising, as they show that Markov
-networks can recover species' pairwise interactions from observational data,
-even when direct interactions are largely overwhelmed by indirect effects
-(Figure 1) or environmental effects (lower panels of Figure 3). For cases
-where fitting a Markov network is infeasible, these results also indicate
-that partial covariances---which can be computed straightforwardly by linear
-regression---can often provide a surprisingly accurate approximation.
-On the other hand, inferences may be much less reliable in cases where the "true"
-interaction matrices include guild structure or trophic levels, which could
-make the $\beta$ coefficients much more difficult to infer (particularly for
-linear methods; @loh_structure_2013). In these cases, it may be necessary to
-include prior information about the nature of species' interactions from 
-natural history, experiments, or trait data. The rosalia package 
-[@harris_rosalia_2015] has built-in mechanisms for incorporating 
-this kind of information, if it can be expressed as a prior probability 
-distribution on $\beta$ or a penalty on the likelihood.
+The results presented above show that Markov networks, unlike a common null
+modeling approach, can recover species' pairwise interactions from
+observational data. The Markov networks were successful even when direct
+interactions are largely overwhelmed by indirect effects (Figure 1) or
+environmental effects (lower panels of Figure 3). For cases where fitting a
+Markov network is computationally infeasible, these results also indicate that
+partial covariances---which can be computed straightforwardly by linear
+regression---can often provide a surprisingly useful approximation. The
+success of the partial correlation's success on simulated data may not 
+carry over to real data sets, however; @loh_structure_2013 suggest that the linear 
+approximations may be less reliable in cases where the true interaction matrix 
+contains more structure (e.g. guilds or trophic levels). On the other hand,
+if ecologists are familiar enough with the natural history of their study
+systems to describe this kind of structure as a prior distribution on the
+parameters or as a penalty on the likelihood, then real-world results might
+be even better than those shown in Figure 3.
 
-Ecologists will also need experiments and natural history knowledge to pin
-down the exact nature of the interactions identified by the network (e.g. whether
-or not negative interactions are consumptive, or which species in a positively-associated
-pair is facilitating the other), particularly when real pairs of species can reciprocally 
-influence one another's occurrence probabilities in multiple ways simultaneously; the
-$\beta$ coefficients in Markov networks have to reduce this complexity to a single
-number. In short, partial correlations and Markov networks both prevent us 
-from mistaking marginal associations for conditional ones, but they can't tell us the 
-underlying biological mechanisms at work.
+Ecologists will also need natural history to pin down the exact nature of the
+interactions identified by a network model (e.g. which species in a
+positively-associated pair is facilitating the other),  particularly when real
+pairs of species can reciprocally influence one another in multiple ways
+simultaneously [@bruno_inclusion_2003]; the $\beta$ coefficients in Markov
+networks have to reduce this complexity  to a single number. In short, partial
+correlations and Markov  networks both help prevent us from mistaking marginal
+associations for  conditional ones, but they can't tell us the underlying
+biological mechanisms at work.
 
@@
 prevalence, which was a major flash point for the null model debates in the
-1980’s [@strong_ecological_1984]. Equation 1 can be used to calculate the
-expected prevalence of a species in the absence of biotic influences
-[$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's effect on
-prevalence in a Markov network can thus be calculated by subtracting this value
-and the observed prevalence (cf Figure 2D).
+1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation 
+1 can be used to calculate the expected prevalence of a species in the 
+absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. 
+Competition's effect on prevalence in a Markov network can then be calculated 
+by subtracting this value and the observed prevalence (cf Figure 2D).
 
@@
 for dealing with indirect effects, stability, and alternative stable states
-(phase transitions).
+(i.e. phase transitions; @cipra_introduction_1987).
 
@@
 factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013]. 
-Numerous other extensions are possible: similar networks
+Numerous other extensions are possible: Markov networks
 can be fit with a mix of discrete and continuous variables, for example

### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
done

@@
 
-Ecologists' intense interest in drawing inferences about species
-interactions ---especially competition---from presence-absence data has
-a long history (MacArthur 1958, Diamond 1975, Connor et al. 2013). If
-nontrophic species interactions are important drivers of community
-assembly patterns, then we should expect to see their influence in our
-data sets. Despite decades of work and several major controversies,
-however (Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003),
+If nontrophic species interactions, such as competition, are important
+drivers of community assembly, then ecologists might expect to see their
+influence in our data sets (MacArthur 1958, Diamond 1975). Despite
+decades of work and several major controversies, however (Lewin 1983,
+Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
 existing methods for detecting competition's effects on community
-structure are unreliable (Gotelli and Ulrich 2009). More generally, it
-can be difficult to reason about the complex web of direct and indirect
+structure are unreliable (Gotelli and Ulrich 2009), and thus important
+ecological processes remain poorly understood. More generally, it can be
+difficult to reason about the complex web of direct and indirect species
 interactions in real assemblages, especially when these interactions
@@
 Most existing methods in this field compare the frequency with which two
-putative competitors are observed to co-occur against the frequency that
+putative competitors are observed to co-occur, versus the frequency that
 would be expected if \emph{all} species on the landscape were
@@
 If this sort of effect is common, then significant deviations from
-independence will not---by themselves---provide convincing evidence of
-species' direct effects on one another.
+independence will not generally provide convincing evidence of species'
+direct effects on one another.
 
@@
 interactions, and for detecting them with observational data. Markov
-networks, which generalize partial correlations to non-Gaussian data,
-have been used in many scientific fields to model associations between
-various kinds of ``particles''. For example, a well-studied network
-called the Ising model has played an important role in our understanding
-of physics (where nearby particles tend to align magnetically with one
-another; Cipra 1987). In spatial contexts, these models have been used
-to describe interactions between adjacent grid cells (Harris 1974,
-Gelfand et al. 2005). In neurobiology, they have helped researchers
-determine which neurons are connected to one another by modeling the
-structure in their firing patterns (Schneidman et al. 2006). Following
-recent work by Azaele et al. (2010) and Fort (2013), I suggest that
-ecologists could similarly treat species as the interacting particles in
-this modeling framework. Doing so would allow ecologists to simulate and
+networks, which generalize partial correlations to non-Gaussian data
+(Lee and Hastie 2012, Loh and Wainwright 2013), have been used in many
+scientific fields for decades to model associations between various
+kinds of ``particles''. For example, a well-studied network named the
+Ising model has played an important role in our understanding of physics
+(where nearby particles tend to align magnetically with one another;
+Cipra 1987). In spatial contexts, these models have been used to
+describe interactions between adjacent grid cells (Harris 1974, Gelfand
+et al. 2005). In neurobiology, they have helped researchers determine
+which neurons are connected to one another by modeling the structure in
+their firing patterns (Schneidman et al. 2006). Following recent work by
+Azaele et al. (2010) and Fort (2013), I suggest that ecologists could
+similarly treat species as the interacting particles in the same
+modeling framework. Doing so would allow ecologists to simulate and
 study the landscape-level consequences of arbitrary species interaction
@@
 networks work and how they can be used to simulate landscape-level data
-and to predict the direct and indirect consequences of possible
+or to predict the direct and indirect consequences of possible
 interaction matrices. Then, using simulated data sets where the ``true''
@@
 
-Here, \(\alpha_{i}\) is the amount that the presence of species \(i\)
-contributes to the log-probability of \(\vec{y}\); it directly controls
-the prevalence of species \(i\). Similarly, \(\beta_{ij}\) is the amount
-that the co-occurrence of species \(i\) and species \(j\) contributes to
-the log-probability, and controls how often the two species will be
-found together (Figure 2A, Figure 2B). \(\beta\) thus acts as an analog
-of the partial covariance, but for non-Gaussian networks. Because the
-relative probability of a presence-absence vector increases when
+Here, \(\alpha_{i}\) is an intercept term determining the amount that
+the presence of species \(i\) contributes to the log-probability of
+\(\vec{y}\); it directly controls the prevalence of species \(i\).
+Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
+species \(i\) and species \(j\) contributes to the log-probability; it
+controls the probability that the two species will be found together
+(Figure 2A, Figure 2B). \(\beta\) thus acts as an analog of the partial
+covariance, but for non-Gaussian networks. Because the relative
+probability of a presence-absence vector increases when
 positively-associated species co-occur and decreases when
@@
 infeasible as the number of species increases: with \(2^N\) possible
-assemblages of \(N\) species the number of bookkeeping operations
-required for exact inference quickly spirals exponentially into the
-billions. Numerous techniques are available for working with Markov
-networks that keep the computations tractable, either through analytic
+assemblages of \(N\) species, the number of bookkeeping operations
+required for exact inference spirals exponentially into the billions and
+beyond. Numerous techniques are available for working with Markov
+networks that keep the computations tractable, e.g.~via analytic
 approximations (Lee and Hastie 2012) or Monte Carlo sampling
@@
 (Appendix A). Each of the methods described below (a Markov network, two
-correlation-based methods and a null modeling approach) was then
-evaluated on its ability to correctly infer that the two shrub species
-competed with one another, despite their frequent co-occurrence.
+correlation-based methods and a null model) was then evaluated on its
+ability to correctly infer that the two shrub species competed with one
+another, despite their frequent co-occurrence.
 
@@
 strongly, and competition was three times more common than facilitation.
-I then used Gibbs sampling to randomly generate landscapes with varying
-numbers of species and sites via Markov chain Monte Carlo (Appendix B).
-For half of the simulated landscapes, I treated each species' \(\alpha\)
-coefficient as a constant, as described above. For the other half, I
-treated the \(\alpha\) coefficients as linear functions of two abiotic
-environmental factors that varied from location to location across the
-landscape (Appendix B). The latter set of simulated landscapes provide
-an important test of the methods' ability to distinguish co-occurrence
-patterns that were generated from pairwise interactions among the
-observed species from those that were generated by external forces like
-abiotic environmental filtering. This task was made especially difficult
-because---as with most analyses of presence-absence data for
-co-occurrence patterns---the inference procedure did not have access to
-any information about the environmental or spatial variables that helped
-shape the landscape (cf Connor et al. 2013, Blois et al. 2014).
+I then used Gibbs sampling to randomly generate replicate landscapes
+with varying numbers of species and sites (Appendix B). For half of the
+simulated landscapes, I treated each species' \(\alpha\) coefficient as
+a constant, as described above. For the other half, I treated the
+\(\alpha\) coefficients as linear functions of two abiotic environmental
+factors that varied from location to location across the landscape
+(Appendix B). The latter set of simulated landscapes provide an
+important test of the methods' ability to distinguish co-occurrence
+patterns that were generated from pairwise biotic interactions from
+those that were generated by external forces like abiotic environmental
+filtering. This task was made especially difficult because---as with
+most analyses of presence-absence data for co-occurrence patterns---the
+inference procedure did not have access to any information about the
+environmental or spatial variables that helped shape the landscape (cf
+Connor et al. 2013, Blois et al. 2014).
 
@@
 
-The first two interaction measures were the simple and partial
-covariances between each pair of species' data vectors on the landscape
-(Albrecht and Gotelli 2001). Because partial covariances are undefined
-for landscapes with perfectly-correlated species pairs, I used a
-regularized estimate based on ridge regression {[}Wieringen and Peeters
-(2014); i.e.~linear regression with a Gaussian prior{]}. For these
-analyses, I set the ridge parameter to 0.2 divided by the number of
-sites on the landscape.
+The first two interaction measures were the sample covariances and the
+partial covariances between each pair of species' data vectors on the
+landscape (Albrecht and Gotelli 2001). Because partial covariances are
+undefined for landscapes with perfectly-correlated species pairs, I used
+a regularized estimate based on ridge regression {[}Wieringen and
+Peeters (2014); i.e.~linear regression with a Gaussian prior{]}. For
+these analyses, I set the ridge parameter to 0.2 divided by the number
+of sites on the landscape.
 
@@
 statistic describing the number of \emph{non}-co-occurrences between two
-species). The software then calculated a \(Z\)-statistic for each
+species). The software then calculated a \(Z\) statistic for each
 species pair using this null distribution. After multiplying this
@@
 
-\textbf{\emph{Method evaluation.}} For the first simulated landscape
-(three species), I kept the evaluation simple and qualitative: any
-method that reliably determined that the two shrub species were
-negatively associated passed; other methods failed.
+\textbf{\emph{Method evaluation.}} I evaluated each method qualitatively
+on the simulated landscapes based on Figure 1: any method that reliably
+determined that the two shrub species were negatively associated passed;
+other methods failed.
 
@@
 re-weighting the pairs so that each landscape contributed equally to the
-rescaled estimate. For each estimate of a species pair's interactions, I
-used this regression to calculate the squared error associated with
-method that produced it. Finally, I averaged these squared errors for
-each combination of species richness, landscape size, statistical
-method, and presence/absence of environmental filtering across all 12
-replicates; the mean squared errors associated with these subsets of the
-data determined the proportion of variance explained by each method
-under different conditions.
+rescaled estimate. For each method, I calculated the squared deviations
+between each species pair's rescaled estimate and the ``true''
+interactions used to generate the assemblages. Finally, I averaged these
+squared errors for each combination of species richness, landscape size,
+statistical method, and presence/absence of environmental filtering
+across all 12 replicates; the mean squared errors associated with these
+subsets of the data determined the proportion of variance explained by
+each method under different conditions.
 
@@
 of the dominant tree species. As a result, the covariance method falsely
-reported positive associations 94\% of the time and the
+reported positive associations on 94\% of simulated landscapes, and the
 randomization-based null model falsely reported such associations 100\%
@@
 the simulated communities (Figure 3). In general, however, there was a
-consistent ordering: the Markov network explained 54\% of the variance
-in the ``true'' parameters, followed by partial covariances (33\%), and
-sample covariances (22\%).
+consistent ordering: overall, the Markov network explained 54\% of the
+variance in the ``true'' parameters, followed by partial covariances
+(33\%), and sample covariances (22\%).
 
-The null model scores initially explained 12\% of the variation. After
-manually reducing the value of one especially strong outlier
-(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%.
-Figure 3 reflects the adjusted version of the results.
+The null model scores initially explained only 12\% of the variation.
+After manually reducing the value of one especially strong outlier
+(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%
+(Appendix B). Figure 3 reflects the adjusted version of the results.
 
@@
 
-The results presented above are very promising, as they show that Markov
-networks can recover species' pairwise interactions from observational
-data, even when direct interactions are largely overwhelmed by indirect
-effects (Figure 1) or environmental effects (lower panels of Figure 3).
-For cases where fitting a Markov network is infeasible, these results
+The results presented above show that Markov networks, unlike a common
+null modeling approach, can recover species' pairwise interactions from
+observational data. The Markov networks were successful even when direct
+interactions are largely overwhelmed by indirect effects (Figure 1) or
+environmental effects (lower panels of Figure 3). For cases where
+fitting a Markov network is computationally infeasible, these results
 also indicate that partial covariances---which can be computed
 straightforwardly by linear regression---can often provide a
-surprisingly accurate approximation. On the other hand, inferences may
-be much less reliable in cases where the ``true'' interaction matrices
-include guild structure or trophic levels, which could make the
-\(\beta\) coefficients much more difficult to infer (particularly for
-linear methods; Loh and Wainwright (2013)). In these cases, it may be
-necessary to include prior information about the nature of species'
-interactions from natural history, experiments, or trait data. The
-rosalia package (Harris 2015a) has built-in mechanisms for incorporating
-this kind of information, if it can be expressed as a prior probability
-distribution on \(\beta\) or a penalty on the likelihood.
+surprisingly useful approximation. The success of the partial
+correlation's success on simulated data may not carry over to real data
+sets, however; Loh and Wainwright (2013) suggest that the linear
+approximations may be less reliable in cases where the true interaction
+matrix contains more structure (e.g.~guilds or trophic levels). On the
+other hand, if ecologists are familiar enough with the natural history
+of their study systems to describe this kind of structure as a prior
+distribution on the parameters or as a penalty on the likelihood, then
+real-world results might be even better than those shown in Figure 3.
 
-Ecologists will also need experiments and natural history knowledge to
-pin down the exact nature of the interactions identified by the network
-(e.g.~whether or not negative interactions are consumptive, or which
-species in a positively-associated pair is facilitating the other),
-particularly when real pairs of species can reciprocally influence one
-another's occurrence probabilities in multiple ways simultaneously; the
-\(\beta\) coefficients in Markov networks have to reduce this complexity
-to a single number. In short, partial correlations and Markov networks
-both prevent us from mistaking marginal associations for conditional
+Ecologists will also need natural history to pin down the exact nature
+of the interactions identified by a network model (e.g.~which species in
+a positively-associated pair is facilitating the other), particularly
+when real pairs of species can reciprocally influence one another in
+multiple ways simultaneously (Bruno et al. 2003); the \(\beta\)
+coefficients in Markov networks have to reduce this complexity to a
+single number. In short, partial correlations and Markov networks both
+help prevent us from mistaking marginal associations for conditional
 ones, but they can't tell us the underlying biological mechanisms at
@@
 species' overall prevalence, which was a major flash point for the null
-model debates in the 1980's (Strong et al. 1984). Equation 1 can be used
-to calculate the expected prevalence of a species in the absence of
-biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and Hastie 2012).
-Competition's effect on prevalence in a Markov network can thus be
-calculated by subtracting this value and the observed prevalence (cf
-Figure 2D).
+model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
+Equation 1 can be used to calculate the expected prevalence of a species
+in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
+and Hastie 2012). Competition's effect on prevalence in a Markov network
+can then be calculated by subtracting this value and the observed
+prevalence (cf Figure 2D).
 
@@
 set of existing discoveries and techniques for dealing with indirect
-effects, stability, and alternative stable states (phase transitions).
+effects, stability, and alternative stable states (i.e.~phase
+transitions; Cipra (1987)).
 
@@
 analyses (cf Connor et al. 2013). Numerous other extensions are
-possible: similar networks can be fit with a mix of discrete and
+possible: Markov networks can be fit with a mix of discrete and
 continuous variables, for example (Lee and Hastie 2012). There are even

### 2015-05-22 Harris_2015/manuscript-materials/draft.Rmd
actually resubmitted

@@
 assemblages, especially when these interactions occur against a background of
-other ecological processes such as dispersal and environmental filtering
+other ecological processes such as dispersal limitation and environmental filtering
 [@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
@@
 
-While the competition between the two shrubs in the previous example does not
+While competition between the two shrubs in the previous example does not
 leave the commonly-expected pattern in community structure (negative
@@
 The rest of the paper proceeds as follows. First, I discuss how Markov networks
-work and how they can be used to simulate landscape-level data or to predict
+operate and how they can be used to simulate landscape-level data or to predict
 the direct and indirect consequences of possible interaction matrices. Then,
@@
 Equation 1 into an absolute probability entails scaling by a *partition
-function*, $Z(\alpha, \beta)$, which ensures that the probabilities of all
+function*, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
 possible assemblages that could be produced by the model sum to one (bottom of
@@
 ***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
-The previous sections involved known values of $\alpha$ and $\beta$. In practice,
-ecologists will often need to estimate these parameters from data instead. 
+In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
+however, ecologists will often need to estimate these parameters from co-occurrence data. 
 When the number of species is reasonably small, one can compute exact maximum 
@@
 
-The first two interaction measures were the sample covariances and the 
+The first two alternative interaction measures were the sample covariances and the 
 partial covariances between each pair of species' data vectors on the landscape
@@
 
-The third method, described in @gotelli_empirical_2009, involved simulating
+The third alternative method, described in @gotelli_empirical_2009, involved simulating
 possible landscapes from a null model that retains the row and column sums of
@@
 ***Method evaluation.*** 
-I evaluated each method qualitatively on the simulated landscapes based on
-Figure 1: any method that reliably determined that the two shrub
-species were negatively associated passed; other methods failed.
+For the simulated landscapes based on Figure 1, method evaluation was fairly
+qualitative: any method whose test statistic for the two shrubs indicated a
+negative relationship passed; other methods failed.
 
@@
 regression through the origin so that they all had a consistent
-interpretation. In each case, I regressed the "true" $\beta$ coefficient for
+interpretation. In each method, I regressed the "true" $\beta$ coefficient for
 each species pair against the model's estimate, re-weighting the pairs so that
-each landscape contributed equally to the rescaled estimate. For each  method,
-I calculated the squared deviations between each species pair's rescaled
-estimate and the "true" interactions used to generate the assemblages.
-Finally, I  averaged these squared errors for each combination of species
-richness, landscape size, statistical method, and presence/absence of
-environmental filtering across all 12 replicates; the mean squared errors
-associated with these subsets of the data determined the proportion of
-variance explained by each method under different conditions.
+each landscape contributed equally to the rescaled estimate. These regressions
+yielded squared errors for each of the 23,140 simulated species pairs, across
+all conditions and replicates. I then partitioned these errors into groups
+defined by the properties of the simulated landscapes (i.e. species richness,
+number of observed communities, and the presence/absence of environmental
+filtering). Within each partition, the mean squared error was used to
+calculate the proportion of variance explained (compared with a baseline model
+that assumed all interaction strengths to be zero).
 
@@
 As shown in Figure 1, the marginal relationship between the two shrub species
-was positive---despite their competition for space at a mechanistic level---
-due to indirect effects of the dominant tree species. As a result, the
-covariance method falsely reported positive associations on 94% of simulated landscapes,
-and the randomization-based null model falsely reported such associations 100%
-of the time. The two methods for evaluating conditional relationships (Markov
-networks and partial covariances), however, successfully controlled for the
-indirect pathway via the tree species and each correctly identified the direct
-negative interaction between the shrubs 94% of the time.
+was positive---despite their competition for space at a mechanistic level---due
+to indirect effects of the dominant tree species. As a result, the
+covariance method falsely reported positive associations for 94% of the simulated
+landscapes, and the randomization-based null model falsely reported such
+associations 100% of the time. The two methods for evaluating conditional
+relationships (Markov networks and partial covariances), however, successfully
+controlled for the indirect pathway via the tree species and each correctly
+identified the direct negative interaction between the shrubs 94% of the time.
 
 ***Larger landscapes.*** 
-The accuracy of the four evaluated methods varied substantially, depending
-on the parameters that produced the simulated communities (Figure 3).
-In general, however, there was a consistent ordering: overall, the Markov network
-explained 54% of the variance in the "true" parameters, followed by 
-partial covariances (33%), and sample covariances (22%).
-
-The null model scores initially explained only 12% of the variation. After
-manually reducing the value of one especially strong outlier ($Z=1004$,
-implying $p<10^{-1000000}$), this increased to 17% (Appendix B).  Figure 3
-reflects the adjusted version of the results.
+The accuracy of the four evaluated methods varied substantially, depending on
+the parameters that produced the simulated communities (Figure 3). In general,
+however, there was a consistent ordering: overall, the Markov network
+explained 54% of the "true" parameters' squared deviations from zero, followed
+by partial covariances (33%), and sample covariances (22%). The null model
+scores initially explained only 12%. After manually reducing the value of one
+especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
+increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
+results.
 
@@
 
-The results presented above show that Markov networks, unlike a common null
-modeling approach, can recover species' pairwise interactions from
-observational data. The Markov networks were successful even when direct
-interactions are largely overwhelmed by indirect effects (Figure 1) or
-environmental effects (lower panels of Figure 3). For cases where fitting a
-Markov network is computationally infeasible, these results also indicate that
-partial covariances---which can be computed straightforwardly by linear
+The results presented above show that Markov networks can reliably recover 
+species' pairwise interactions from observational data, even for cases where
+a common null modeling technique reliably fails. Specifically, Markov networks 
+were successful even when direct interactions were largely overwhelmed by indirect 
+effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases 
+where fitting a Markov network is computationally infeasible, these results also 
+indicate that partial covariances---which can be computed straightforwardly by linear
 regression---can often provide a surprisingly useful approximation. The
-success of the partial correlation's success on simulated data may not 
-carry over to real data sets, however; @loh_structure_2013 suggest that the linear 
-approximations may be less reliable in cases where the true interaction matrix 
+partial correlations' success on simulated data may not 
+carry over to real data sets, however; @loh_structure_2013 show that the linear 
+approximations can be less reliable in cases where the true interaction matrix 
 contains more structure (e.g. guilds or trophic levels). On the other hand,
@@
 systems to describe this kind of structure as a prior distribution on the
-parameters or as a penalty on the likelihood, then real-world results might
+parameters or as a penalty on the likelihood, then this information could
+reduce the effective degres of freedom to estimate and real-world results might
 be even better than those shown in Figure 3.
@@
 
-Finally, the results presented here have important implications for ecologists' 
-continued use of null models to draw inferences about species interactions. The
-small simulated data sets described by Figure 1 show that test statistics
-based on marginal co-occurrence (such as C-scores) will not always have a
-straightforward relationship with the underlying ecological processes. More
-generally, deviations from the null model generally provided less information
-about direct species interactions than correlation coefficients did.
-Scientists currently using null modeling approaches in their research may be
-able to capture twice as much of the variance in species' true interaction
-strengths using partial covariances from linear regression instead,  or three
-times as much using a Markov network.
-
-Null and neutral models can be very useful for clarifying our thinking about
-the numerical consequences of species' richness and abundance patterns
-[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a null model
-must be interpreted with care [@roughgarden_competition_1983]. In complex
-networks of ecological interactions---and even in small networks with three
-species---it may simply not be possible to implicate individual species pairs
-or specific ecological processes like competition by rejecting a general-
-purpose null. Direct estimates of species' conditional associations may be the
-only way to make these inferences reliably.
+Finally, the results presented here have important implications for
+ecologists' continued use of null models to draw inferences about species
+interactions. Null and neutral models can be very useful for clarifying our
+thinking about the numerical consequences of species' richness and abundance
+patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
+null model must be interpreted with care [@roughgarden_competition_1983]. In
+complex networks of ecological interactions (and even in small networks with
+three species), it may simply not be possible to implicate individual species
+pairs or specific ecological processes like competition by rejecting a
+general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
+coefficients directly seems like a much more promising approach: to the extent
+that the models' relative performance on real data sets is similar to the
+range of results shown in Figure 3, scientists in this field could easily
+double their explanatory power by switching from null models to linear
+regression and partial covariances, or triple it by switching to a Markov
+network.
 

### 2015-05-22 Harris_2015/manuscript-materials/draft.tex
actually resubmitted

@@
 \setlength{\parindent}{0pt}
-\setlength{\parskip}{5pt}
+\setlength{\parskip}{6pt}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 occur against a background of other ecological processes such as
-dispersal and environmental filtering (Connor et al. 2013). For this
-reason, it isn't always clear what kinds of patterns would even
+dispersal limitation and environmental filtering (Connor et al. 2013).
+For this reason, it isn't always clear what kinds of patterns would even
 constitute evidence of competition, as opposed to some other biological
@@
 
-While the competition between the two shrubs in the previous example
-does not leave the commonly-expected pattern in community structure
-(negative association at the landscape level), it nevertheless does
-leave a signal in the data (Figure 1C). Specifically, \emph{among shaded
-sites}, there will be a deficit of co-occurrences, and \emph{among
-unshaded sites}, there will also be such a deficit.
+While competition between the two shrubs in the previous example does
+not leave the commonly-expected pattern in community structure (negative
+association at the landscape level), it nevertheless does leave a signal
+in the data (Figure 1C). Specifically, \emph{among shaded sites}, there
+will be a deficit of co-occurrences, and \emph{among unshaded sites},
+there will also be such a deficit.
 
@@
 The rest of the paper proceeds as follows. First, I discuss how Markov
-networks work and how they can be used to simulate landscape-level data
-or to predict the direct and indirect consequences of possible
+networks operate and how they can be used to simulate landscape-level
+data or to predict the direct and indirect consequences of possible
 interaction matrices. Then, using simulated data sets where the ``true''
@@
 probability given by Equation 1 into an absolute probability entails
-scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which
-ensures that the probabilities of all possible assemblages that could be
-produced by the model sum to one (bottom of Figure 2B). Calculating
-\(Z(\alpha, \beta)\) exactly, as is done in this paper, quickly becomes
-infeasible as the number of species increases: with \(2^N\) possible
-assemblages of \(N\) species, the number of bookkeeping operations
-required for exact inference spirals exponentially into the billions and
-beyond. Numerous techniques are available for working with Markov
-networks that keep the computations tractable, e.g.~via analytic
+scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which is
+defined so that that the probabilities of all possible assemblages that
+could be produced by the model sum to one (bottom of Figure 2B).
+Calculating \(Z(\alpha, \beta)\) exactly, as is done in this paper,
+quickly becomes infeasible as the number of species increases: with
+\(2^N\) possible assemblages of \(N\) species, the number of bookkeeping
+operations required for exact inference spirals exponentially into the
+billions and beyond. Numerous techniques are available for working with
+Markov networks that keep the computations tractable, e.g.~via analytic
 approximations (Lee and Hastie 2012) or Monte Carlo sampling
@@
 \textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
-presence-absence data.}} The previous sections involved known values of
-\(\alpha\) and \(\beta\). In practice, ecologists will often need to
-estimate these parameters from data instead. When the number of species
-is reasonably small, one can compute exact maximum likelihood estimates
-for all of the \(\alpha\) and \(\beta\) coefficients by optimizing
-\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
-ones considered here have unimodal likelihood surfaces (Murphy 2012),
-ensuring that this procedure will always converge on the global maximum.
-This maximum is the unique combination of \(\alpha\) and \(\beta\)
-coefficients that would be expected to produce exactly the observed
-co-occurrence frequencies. For the analyses in this paper, I used the
-\emph{rosalia} package (Harris 2015a) for the R programming language (R
-Core Team 2015) to define the objective function and gradient as R code.
-The rosalia package then uses the \texttt{BFGS} method in R's
-\texttt{optim} function to find the best values for \(\alpha\) and
-\(\beta\).
+presence-absence data.}} In the previous two sections, the values of
+\(\alpha\) and \(\beta\) were known. In practice, however, ecologists
+will often need to estimate these parameters from co-occurrence data.
+When the number of species is reasonably small, one can compute exact
+maximum likelihood estimates for all of the \(\alpha\) and \(\beta\)
+coefficients by optimizing \(p(\vec{y}; \alpha, \beta)\). Fully-observed
+Markov networks like the ones considered here have unimodal likelihood
+surfaces (Murphy 2012), ensuring that this procedure will always
+converge on the global maximum. This maximum is the unique combination
+of \(\alpha\) and \(\beta\) coefficients that would be expected to
+produce exactly the observed co-occurrence frequencies. For the analyses
+in this paper, I used the \emph{rosalia} package (Harris 2015a) for the
+R programming language (R Core Team 2015) to define the objective
+function and gradient as R code. The rosalia package then uses the
+\texttt{BFGS} method in R's \texttt{optim} function to find the best
+values for \(\alpha\) and \(\beta\).
 
@@
 
-The first two interaction measures were the sample covariances and the
-partial covariances between each pair of species' data vectors on the
-landscape (Albrecht and Gotelli 2001). Because partial covariances are
-undefined for landscapes with perfectly-correlated species pairs, I used
-a regularized estimate based on ridge regression {[}Wieringen and
-Peeters (2014); i.e.~linear regression with a Gaussian prior{]}. For
-these analyses, I set the ridge parameter to 0.2 divided by the number
-of sites on the landscape.
+The first two alternative interaction measures were the sample
+covariances and the partial covariances between each pair of species'
+data vectors on the landscape (Albrecht and Gotelli 2001). Because
+partial covariances are undefined for landscapes with
+perfectly-correlated species pairs, I used a regularized estimate based
+on ridge regression {[}Wieringen and Peeters (2014); i.e.~linear
+regression with a Gaussian prior{]}. For these analyses, I set the ridge
+parameter to 0.2 divided by the number of sites on the landscape.
 
-The third method, described in Gotelli and Ulrich (2009), involved
-simulating possible landscapes from a null model that retains the row
-and column sums of the original matrix (Strong et al. 1984). Using the
-default options in the Pairs software described in Gotelli and Ulrich
-(2009), I simulated the null distribution of scaled C-scores (a test
-statistic describing the number of \emph{non}-co-occurrences between two
-species). The software then calculated a \(Z\) statistic for each
-species pair using this null distribution. After multiplying this
+The third alternative method, described in Gotelli and Ulrich (2009),
+involved simulating possible landscapes from a null model that retains
+the row and column sums of the original matrix (Strong et al. 1984).
+Using the default options in the Pairs software described in Gotelli and
+Ulrich (2009), I simulated the null distribution of scaled C-scores (a
+test statistic describing the number of \emph{non}-co-occurrences
+between two species). The software then calculated a \(Z\) statistic for
+each species pair using this null distribution. After multiplying this
 statistic by \(-1\) so that positive values corresponded to facilitation
@@
 
-\textbf{\emph{Method evaluation.}} I evaluated each method qualitatively
-on the simulated landscapes based on Figure 1: any method that reliably
-determined that the two shrub species were negatively associated passed;
-other methods failed.
+\textbf{\emph{Method evaluation.}} For the simulated landscapes based on
+Figure 1, method evaluation was fairly qualitative: any method whose
+test statistic for the two shrubs indicated a negative relationship
+passed; other methods failed.
 
@@
 linear regression through the origin so that they all had a consistent
-interpretation. In each case, I regressed the ``true'' \(\beta\)
+interpretation. In each method, I regressed the ``true'' \(\beta\)
 coefficient for each species pair against the model's estimate,
 re-weighting the pairs so that each landscape contributed equally to the
-rescaled estimate. For each method, I calculated the squared deviations
-between each species pair's rescaled estimate and the ``true''
-interactions used to generate the assemblages. Finally, I averaged these
-squared errors for each combination of species richness, landscape size,
-statistical method, and presence/absence of environmental filtering
-across all 12 replicates; the mean squared errors associated with these
-subsets of the data determined the proportion of variance explained by
-each method under different conditions.
+rescaled estimate. These regressions yielded squared errors for each of
+the 23,140 simulated species pairs, across all conditions and
+replicates. I then partitioned these errors into groups defined by the
+properties of the simulated landscapes (i.e.~species richness, number of
+observed communities, and the presence/absence of environmental
+filtering). Within each partition, the mean squared error was used to
+calculate the proportion of variance explained (compared with a baseline
+model that assumed all interaction strengths to be zero).
 
@@
 relationship between the two shrub species was positive---despite their
-competition for space at a mechanistic level--- due to indirect effects
+competition for space at a mechanistic level---due to indirect effects
 of the dominant tree species. As a result, the covariance method falsely
-reported positive associations on 94\% of simulated landscapes, and the
-randomization-based null model falsely reported such associations 100\%
-of the time. The two methods for evaluating conditional relationships
-(Markov networks and partial covariances), however, successfully
-controlled for the indirect pathway via the tree species and each
-correctly identified the direct negative interaction between the shrubs
-94\% of the time.
+reported positive associations for 94\% of the simulated landscapes, and
+the randomization-based null model falsely reported such associations
+100\% of the time. The two methods for evaluating conditional
+relationships (Markov networks and partial covariances), however,
+successfully controlled for the indirect pathway via the tree species
+and each correctly identified the direct negative interaction between
+the shrubs 94\% of the time.
 
@@
 consistent ordering: overall, the Markov network explained 54\% of the
-variance in the ``true'' parameters, followed by partial covariances
-(33\%), and sample covariances (22\%).
-
-The null model scores initially explained only 12\% of the variation.
-After manually reducing the value of one especially strong outlier
-(\(Z=1004\), implying \(p<10^{-1000000}\)), this increased to 17\%
-(Appendix B). Figure 3 reflects the adjusted version of the results.
+``true'' parameters' squared deviations from zero, followed by partial
+covariances (33\%), and sample covariances (22\%). The null model scores
+initially explained only 12\%. After manually reducing the value of one
+especially strong outlier (\(Z=1004\), implying \(p<10^{-1000000}\)),
+this increased to 17\% (Appendix B). Figure 3 reflects the adjusted
+version of the results.
 
@@
 
-The results presented above show that Markov networks, unlike a common
-null modeling approach, can recover species' pairwise interactions from
-observational data. The Markov networks were successful even when direct
-interactions are largely overwhelmed by indirect effects (Figure 1) or
+The results presented above show that Markov networks can reliably
+recover species' pairwise interactions from observational data, even for
+cases where a common null modeling technique reliably fails.
+Specifically, Markov networks were successful even when direct
+interactions were largely overwhelmed by indirect effects (Figure 1) or
 environmental effects (lower panels of Figure 3). For cases where
@@
 straightforwardly by linear regression---can often provide a
-surprisingly useful approximation. The success of the partial
-correlation's success on simulated data may not carry over to real data
-sets, however; Loh and Wainwright (2013) suggest that the linear
-approximations may be less reliable in cases where the true interaction
-matrix contains more structure (e.g.~guilds or trophic levels). On the
-other hand, if ecologists are familiar enough with the natural history
-of their study systems to describe this kind of structure as a prior
-distribution on the parameters or as a penalty on the likelihood, then
-real-world results might be even better than those shown in Figure 3.
+surprisingly useful approximation. The partial correlations' success on
+simulated data may not carry over to real data sets, however; Loh and
+Wainwright (2013) show that the linear approximations can be less
+reliable in cases where the true interaction matrix contains more
+structure (e.g.~guilds or trophic levels). On the other hand, if
+ecologists are familiar enough with the natural history of their study
+systems to describe this kind of structure as a prior distribution on
+the parameters or as a penalty on the likelihood, then this information
+could reduce the effective degres of freedom to estimate and real-world
+results might be even better than those shown in Figure 3.
 
@@
 ecologists' continued use of null models to draw inferences about
-species interactions. The small simulated data sets described by Figure
-1 show that test statistics based on marginal co-occurrence (such as
-C-scores) will not always have a straightforward relationship with the
-underlying ecological processes. More generally, deviations from the
-null model generally provided less information about direct species
-interactions than correlation coefficients did. Scientists currently
-using null modeling approaches in their research may be able to capture
-twice as much of the variance in species' true interaction strengths
-using partial covariances from linear regression instead, or three times
-as much using a Markov network.
-
-Null and neutral models can be very useful for clarifying our thinking
-about the numerical consequences of species' richness and abundance
-patterns (Harris et al. 2011, Xiao et al. 2015), but deviations from a
-null model must be interpreted with care (Roughgarden 1983). In complex
-networks of ecological interactions---and even in small networks with
-three species---it may simply not be possible to implicate individual
-species pairs or specific ecological processes like competition by
-rejecting a general- purpose null. Direct estimates of species'
-conditional associations may be the only way to make these inferences
-reliably.
+species interactions. Null and neutral models can be very useful for
+clarifying our thinking about the numerical consequences of species'
+richness and abundance patterns (Harris et al. 2011, Xiao et al. 2015),
+but deviations from a null model must be interpreted with care
+(Roughgarden 1983). In complex networks of ecological interactions (and
+even in small networks with three species), it may simply not be
+possible to implicate individual species pairs or specific ecological
+processes like competition by rejecting a general-purpose null (Gotelli
+and Ulrich 2009). Estimating pairwise coefficients directly seems like a
+much more promising approach: to the extent that the models' relative
+performance on real data sets is similar to the range of results shown
+in Figure 3, scientists in this field could easily double their
+explanatory power by switching from null models to linear regression and
+partial covariances, or triple it by switching to a Markov network.
 

### 2015-06-13 stochastic/draft.Rmd
add a script for stochastic approximation

```{r}
library(rosalia)
library(mistnet)
library(progress)
library(beepr)

beeping = TRUE
```

```{r}
`%plus%` = mistnet:::`%plus%`
logistic = binomial()$linkinv
rbern = function(p){rbinom(length(p), size = 1, prob = p)}
```

```{r}
set.seed(1)
n_spp = 250
n_loc = 2500
n_env = 3
n_gibbs = 1000

true_beta_vec = rnorm(
  choose(n_spp, 2), 
  sd = abs(rnorm(choose(n_spp, 2))) / 5
)

true_beta = matrix(0, nrow = n_spp, ncol = n_spp)
true_beta[upper.tri(true_beta)] = true_beta_vec
true_beta = true_beta + t(true_beta)

true_intercepts = rnorm(n_spp)

true_env = matrix(rnorm(n_loc * n_env, 0), nrow = n_loc)

true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 0.5), nrow = n_env)

true_alpha = true_env %*% true_alpha_env
```

```{r, cache = TRUE}
y = matrix(0.5, nrow = n_loc, ncol = n_spp)

pb <- progress_bar$new(
  format = "  simulating landscape [:bar] :percent eta: :eta",
  total = n_gibbs,
  clear = FALSE
)

for(i in 1:n_gibbs){
  pb$tick()
  for(j in sample.int(n_spp)){
    y[,j] = rbern(logistic(true_alpha[ , j] + y %*% true_beta[ , j]))
  }
}

if(beeping){beep()}
```

```{r}
# Calculate sufficient statistics of the data
y_stats = crossprod(y)
y_env_stats = t(true_env) %*% y
```

```{r}
y_sim = matrix(0.5, nrow = nrow(y), ncol = ncol(y))
env = true_env

alpha_env = delta_alpha_env = matrix(0, nrow = n_env, ncol = n_spp)
alpha_species = delta_alpha_species = qlogis(colMeans(y))
beta = delta_beta = matrix(0, nrow = n_spp, ncol = n_spp)
alpha = matrix(0, nrow = n_spp, ncol = n_spp) # no delta alpha to initialize
                                              # b/c alpha not optimized directly
```

```{r}
initial_learning_rate = 0.1
momentum = 0.9

pb <- progress_bar$new(
  format = "  Fitting [:bar] :percent eta: :eta",
  total = n_gibbs,
  clear = FALSE
)

for(i in 1:n_gibbs){
  pb$tick()

  ##############################
  # Gibbs sampling for predicted species composition
  ##############################

  # Update alpha
  alpha = env %*% alpha_env %plus% alpha_species

  # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
  for(j in sample.int(n_spp)){
    y_sim[,j] = rbern(logistic(alpha[ , j] + y_sim %*% beta[ , j]))
  }

  ##############################
  # Stochastic approximation for updating alpha and beta
  ##############################

  # Update learning rate
  learning_rate = initial_learning_rate * 1000 / (999 + i)

  # Calculate sufficient statistics
  y_sim_stats = crossprod(y_sim)
  y_sim_env_stats = t(env) %*% y_sim

  # Calculate the gradient with respect to alpha and beta
  delta_stats = y_stats - y_sim_stats
  beta_grad = (delta_stats - 10 * beta) / 10 / n_loc # Subtract beta as weight decay

  alpha_species_grad = diag(beta_grad) / n_loc
  diag(beta_grad) = 0
  alpha_env_grad = (y_env_stats - y_sim_env_stats - alpha_env)  / n_loc

  # Calculate parameter updates
  delta_beta = beta_grad * learning_rate + momentum * delta_beta
  delta_alpha_species = alpha_species_grad * learning_rate # no momentum
  delta_alpha_env = alpha_env_grad * learning_rate + momentum  * delta_alpha_env

  beta = beta + delta_beta
  alpha_species = alpha_species + delta_alpha_species
  alpha_env = alpha_env + delta_alpha_env

  if(i %% 10 == 0){
    plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
         pch = 16, cex = .5, col = "#00000050",
         xlab = "Estimated interaction strength",
         ylab = "\"True\" interaction strength",
         main = paste(
           "R^2 =", 
           signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
          )
    )
    abline(v = 0, h = 0)
  }
}
```

```{r}
pdf("poster-beta.pdf", width = 4, height = 4)
plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
     pch = 16, cex = .4, col = "#00000020",
     xlab = "Estimated interaction strength",
     ylab = "\"True\" interaction strength",
     main = "Recoverd pairwise interactions",
     las = 1,
     cex.axis = 0.75
    )
text(
  -0.7, 
  1,
  paste(
    "R^2 =", 
    signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
  )
)

dev.off()
```

```{r}
pdf("poster-alpha.pdf", width = 4, height = 4)
plot(alpha_env, true_alpha_env,
     pch = 16, cex = .4, col = "#00000090",
     xlab = "Estimated environmental coefficient",
     ylab = "\"True\" environmental coefficient",
     main = "Recovered environmental coefficients",
     las = 1,
     cex.axis = 0.75
)
text(
  -1, 
  1,
  paste(
    "R^2 =", 
    signif(cor(c(alpha_env), c(true_alpha_env))^2, 2)
  )
)
dev.off()

```

### 2015-06-16 stochastic/stochastic-recovery.Rmd
save stochastic recovery scrip

--- 2015-06-13 stochastic/draft.Rmd
@@
+# Estimating pairwise interactions by stochastic maximum likelihood
+
 ```{r}
@@
 ```{r}
-pdf("poster-beta.pdf", width = 4, height = 4)
+pdf("poster-beta.pdf", width = 4, height = 5)
 plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
@@
      ylab = "\"True\" interaction strength",
-     main = "Recoverd pairwise interactions",
+     main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
      las = 1,
@@
 ```{r}
-pdf("poster-alpha.pdf", width = 4, height = 4)
+pdf("poster-alpha.pdf", width = 4, height = 5)
 plot(alpha_env, true_alpha_env,
@@
      ylab = "\"True\" environmental coefficient",
-     main = "Recovered environmental coefficients",
+     main = paste0("Recoverd environmental coefficients", "\nN = ", length(alpha_env)),
      las = 1,

### 2015-06-21 stochastic/stochastic-recovery.Rmd
stochastic recovery

@@
 set.seed(1)
-n_spp = 250
-n_loc = 2500
+n_spp = 100
+n_loc = 1000
 n_env = 3
@@
   delta_stats = y_stats - y_sim_stats
-  beta_grad = (delta_stats - 10 * beta) / 10 / n_loc # Subtract beta as weight decay
+  beta_grad = (delta_stats - beta) / 10 / n_loc # Subtract beta as weight decay
 
@@
 dev.off()
-
 ```

### 2015-06-23 stochastic/chapter-3.Rmd
update stochastic approximation code/text

# Introduction

To the extent that species interactions are important for community assembly, ecologists generally expect them to leave a signature on species' co-occurrence patterns.  But using that signature to infer the underlying species interactions has been much harder.  Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @strong_ecological_1984; @hastings_can_1987]. [[**The next two sentences are unnecessarily confrontational and should probably be removed**]] *Despite these difficulties, ecologists have considered the underlying questions so important that, for the last few decades, they have continued to address them with whatever methods were available, even when these methods are known to perform poorly. For example, the False Discovery Rates above 95% demonstrated in @gotelli_empirical_2009 have not prevented wide adoption of method proposed in that paper*

During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among *thousands* of genes from their co-expression levels (e.g. @friedman_sparse_2008). The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013], and most of the test statistics in common use are actually minor variations on the sample correlation [*evidence to be added to Chapter 2*]. As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1, *to be added*). While some narrowly-focused methods have been proposed for incorporating a small number of specific factors into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many factors acting simultaneously.

In other fields, scientists have a number of procedures for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables.  The most familiar of these is the partial correlation. Rather than describing the *overall* relationship between two variables across *all* conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.  @harris_estimating_2015 showed that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from co-occurrence data, and that Markov networks (undirected graphical models also known as Markov random fields), which do not assume normally-distributed observations do even better (cf @azaele_inferring_2010 and @fort_statistical_2013). A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network) can be estimated by maximum likelihood.

Unfortunately, Markov networks have an intractable likelihood function whose computational complexity more than doubles each time a new species is added to the model. Thus, while Harris (2015) was able to use off-the-shelf optimization techniques to maximize the likelihood for networks of 20 species, extending the same approach to networks of 50-100 species would be completely infeasible. If ecologists want the model to account for abiotic variation among sites on the landscape, then these expensive computations need to be repeated independently for every site, increasing the computational burden even further.

In this paper, I present a different way of optimizing the likelihood, called *stochastic approximation* [@robbins_stochastic_1951; @salakhutdinov_efficient_2012], which replaces the intractable computations with tractable Monte Carlo estimates of the same quantities. Despite the sampling error introduced by this substitution, stochastic approximation provides strong guarantees for eventual convergence to the maximum likelihood estimate. This change in approach makes it feasible to estimate the interactions among hundreds of species from observational data while also accounting for possible responses to abiotic factors.

# Methods

## Markov networks

As discussed in @azaele_inferring_2010, @fort_statistical_2013, and @harris_estimating_2015, Markov networks such as the Ising model [@cipra_introduction_1987] can be used to describe community structure as follows.  Each species is represented by a binary random variable, describing whether it is present or absent.  A species’ conditional probability of presence under a given set of biotic and abiotic conditions depends on its coefficients. The coefficients linking species to one another describe their conditional associations: *all else equal*, a species will occur less often in the presence of its competitors and other species with which it is negatively associated, and more often in the presence of its mutualists.  These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences.  For example, hypothetical assemblages including many pairs of competitively exclusive species at the same site will tend to be very improbable under the model.

Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.  However, it is generally straightforward to simulate examples of assemblages that are consistent with graphical models such as these, using Monte Carlo techniques like Gibbs sampling [@salakhutdinov_efficient_2012]. Gibbs sampling is especially convenient for these models because it only requires the conditional probabilities, which are straightforward to compute (Harris 2015; Appendix).

In each of the scenarios below, the “observed” landscapes were simulated using Gibbs sampling from a pre-specified set of “true” parameters.

## Simulated landscapes with known interactions

In order to evaluate partial correlations' ability to recover the "true" parameters that generated a presence-absence matrix of interest, I first needed to generate such matrices from known processes.

To demonstrate that the stochastic approximation method worked, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.  These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@harris_rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.  For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described below.

I then simulated a large landscape with 250 species and 2500 sites, representing a large data set roughly the size of the North American Breeding Bird Survey.  Each site on the landscape was represented by 5 environmental variables, which were drawn independently from Gaussian distributions with mean 0 and standard deviation 2.

Each species was assigned two two sets of coefficients.  The coefficients determining species’ responses to the environment were each drawn from standard normal distributions.  The coefficients describing species’ pairwise interactions were drawn from a mixture of two zero-mean Gaussian distributions. 90% of the interaction coefficients were drawn from a distribution with standard deviation 0.2, while the remainder were drawn from a distribution with standard deviation 1.0.

The coefficients described above are sufficient to define each species' conditional occurrence probability (i.e. its probability of occurrence against any given backdrop of biotic and abiotic features).  I used these conditional probabilities to produce examples of communities that were consistent with the competition parameters and with the local environmental variables via Markov chain Monte Carlo (specifically, `r n_gibbs` rounds of Gibbs sampling; see code in the Appendix). I then used the methods from the next section to attempt to infer the underlying parameters from the simulated data.

## Coefficient estimation with stochastic approximation

A Markov network describes a probability distribution over possible assemblages, and this distribution can be summarized without loss of information by its *sufficient statistics*.  For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [[*Appendix, not yet written*]].  One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the average sufficient statistics produced by samples from the model [[ref?]]. Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], any iterative procedure that can reduce this discrepancy quickly enough will converge to the global maximum likelihood estimate.

The `rosalia` package reduces the discrepancy between observed and predicted sufficient statistics by calculating it exactly, averaging over *all* possible presence-absence combinations, but this approach is intractable for larger networks. Stochastic approximation [refs] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.  The procedure iterates through the following three steps as many times as needed (5000 for these analyses):

1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.

The specific procedure used here had the following implementation details:

The simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates. While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this difficulty does not prevent convergence to the maximum likelihood estimate [@younes_convergence_1999; @salakhutdinov_efficient_2012].

The approximate likelihood gradients in Step 2 match the ones from the Appendix of @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations. These gradients were augmented with a momentum term equal to 0.9 [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 1.0 (for environmental responses) or 0.1 (for pairwise interactions); the pairwise interaction terms were regularized more heavily to improve identifiability and reduce overfitting for this large set of 31,125 coefficients.

The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 500/(500 - t - 1)$, which satisfies the criteria in @powell_approximate_2007. $\alpha_0$ was 1.0 for species’ intercepts and environmental responses, and one tenth as large for their pairwise interaction coefficients.

# Results

![](convergence.pdf)

The squared deviations between the exact estimates produced by the `rosalia` package and the ones produced by stochastic approximation quickly decayed to negligible levels [[Figure XA]], indicating that the stochastic approximation procedure was implemented correctly and worked as the mathematical theory predicts.

![](estimates.pdf)

For the larger landscape, I found that the stochastic approximation approach achieved reasonably good performance after less than a minute of optimization [Figure XB]. With additional time, was able to recover most of the variance in species’ pairwise interactions from observational data ($R^2 = 0.66$; Figure YA), and nearly all of the variance in their responses to environmental variables ($R^2 = 0.95$, not shown).

While the maximum likelihood estimates produced by the Markov network worked well, the sample correlation, explained a negligible fraction of the variance in species’ pairwise interactions ($R^2 = 0.03$, Figure YB).  To the extent that ecologists’ other test statistics for co-occurrence are similar to the sample correlation [[*evidence to be added to chapter 2*]], they can be expected to perform equally poorly. As shown previously, the partial correlation was a much better indicator of species’ true interactions than the sample correlation ($R^2 = 0.44$ Figure YC).

# Discussion

For decades, ecologists have relied on minor variations on the correlation coefficient as their test statistics for inferring species interactions from observational data @harris_estimating_2015. As shown here and in @harris_estimating_2015, however, these inferences can only be made reliably by methods that control for other species in the network, such as partial correlations or Markov networks.  The biggest computational problem with large Markov networks is the impossibility of averaging over all their possible states, but these results demonstrate that this step can be avoided.  Stochastic approximation is able to recover the same estimates as exact methods while scaling gracefully to networks with hundreds of species.

The largest downside of using stochastic approximation instead of the exact methods introduced in @harris_estimating_2015 is the difficulty in generating confidence intervals.  The `rosalia` package can produce confidence intervals based on the Hessian matrix, but it is not feasible to calculate this matrix for large networks.  It may turn out that the best way to generate confidence intervals for large networks is by repeatedly fitting the model to bootstrapped samples of the data. Alternatively, ecologists may be able to take advantage of the advanced Monte Carlo techniques introduced in @murray_mcmc_2012 to sample from the posterior distribution over possible parameter values, which could be used to generate credible intervals for the parameter estimates. Other advanced Monte Carlo methods [e.g. @salakhutdinov_learning_2009] may also speed up convergence to the maximum likelihood estimate in cases where serial autocorrelation in the Gibbs sampler is too strong for effectively sampling the space of possible landscapes.

As the number of potentially-interacting species increases, the number of adjustable parameters increases even faster, so overfitting becomes a major concern. Fortunately, a number of good regularizers have been proposed.  Of these, some of the most interesting options include $L_1$ regularization, which ensures that many of the estimated coefficients are exactly zero [[refs]].  This sparsity matches ecologists’ intuition that many species from different guilds are unlikely to interact much at all [[refs]], and also produces computational benefits for model estimation. Of course, the best regularizers will take advantage of prior ecological knowledge (e.g. from field experiments, natural history, or trait data) to provide information about individual pairwise interactions, rather than about their overall distribution.

Ecologists could expand beyond simple Markov networks in a number of ways that would improve their ability to address a wider range of questions.  This paper demonstrated that it is possible to simultaneously estimate species’ responses to the abiotic and environment and to one another, but many other extensions are possible. For example, ecologists could condition the model on variables whose values have not been measured (e.g. partially-observed Markov networks, or the approximate networks in the `mistnet` package). This would allow ecologists to account for measurement error and for other ecologically-important factors that can be difficult to measure. Ecologists should also explore higher-order networks, where one species’ presence can affect the relationship between two other species.

# References

### 2015-06-23 stochastic/stochastic-recovery.Rmd
update stochastic approximation code/text

@@
 set.seed(1)
-n_spp = 100
-n_loc = 1000
-n_env = 3
+n_spp = 250
+n_loc = 2500
+n_env = 5
 n_gibbs = 1000
+
+pair_type = sample.int(
+  2, 
+  size = choose(n_spp, 2), 
+  replace = TRUE,
+  prob = c(.1, .9)
+)
+
+means = numeric(length(pair_type))
+means[pair_type == 1] = 0
+means[pair_type == 2] = 0
+
+sds = numeric(length(pair_type))
+sds[pair_type == 1] = 1.0
+sds[pair_type == 2] = 0.2
 
@@
   choose(n_spp, 2),
-  sd = abs(rnorm(choose(n_spp, 2))) / 5
+  mean = means, 
+  sd = sds
 )
+
+plot(density(true_beta_vec))
 
@@
 
-true_env = matrix(rnorm(n_loc * n_env, 0), nrow = n_loc)
+true_env = matrix(rnorm(n_loc * n_env, 0, 2), nrow = n_loc)
 
-true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 0.5), nrow = n_env)
+true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 1), nrow = n_env)
 
-true_alpha = true_env %*% true_alpha_env
+true_alpha_species = rnorm(n_spp, -2)
+
+true_alpha = true_env %*% true_alpha_env %plus% true_alpha_species
 ```
@@
 ```{r}
-initial_learning_rate = 0.1
+# Are there lurking factors of two because I use the whole symmetric matrix?
+beta_prior = rosalia::make_logistic_prior(scale = 0.1)$log_grad
+alpha_env_prior = rosalia::make_logistic_prior(scale = 1)$log_grad 
+alpha_species_prior = rosalia::make_logistic_prior(scale = 1)$log_grad
+```
+
+```{r}
+initial_learning_rate = 1
 momentum = 0.9
+maxit = 5E3
 
@@
   format = "  Fitting [:bar] :percent eta: :eta",
-  total = n_gibbs,
+  total = maxit,
   clear = FALSE
@@
 
-for(i in 1:n_gibbs){
+mses = numeric(maxit)
+times = integer(maxit)
+
+for(i in 1:maxit){
   pb$tick()
@@
   # Update learning rate
-  learning_rate = initial_learning_rate * 1000 / (999 + i)
+  learning_rate = initial_learning_rate * 500 / (499 + i)
 
@@
   # Calculate the gradient with respect to alpha and beta
-  delta_stats = y_stats - y_sim_stats
-  beta_grad = (delta_stats - beta) / 10 / n_loc # Subtract beta as weight decay
+  stats_difference = y_stats - y_sim_stats
+  beta_grad = (stats_difference + beta_prior(beta)) / n_loc
 
-  alpha_species_grad = diag(beta_grad) / n_loc
+  alpha_species_grad = (diag(stats_difference) + alpha_species_prior(alpha_species)) / n_loc
   diag(beta_grad) = 0
-  alpha_env_grad = (y_env_stats - y_sim_env_stats - alpha_env)  / n_loc
+  y_env_difference = y_env_stats - y_sim_env_stats
+  alpha_env_grad = (y_env_difference + alpha_env_prior(alpha_env))  / n_loc
 
   # Calculate parameter updates
-  delta_beta = beta_grad * learning_rate + momentum * delta_beta
-  delta_alpha_species = alpha_species_grad * learning_rate # no momentum
-  delta_alpha_env = alpha_env_grad * learning_rate + momentum  * delta_alpha_env
+  delta_beta = beta_grad * learning_rate / 10 + 
+    momentum * delta_beta
+  delta_alpha_species = alpha_species_grad * learning_rate + 
+    momentum  * delta_alpha_species
+  delta_alpha_env = alpha_env_grad * learning_rate + 
+    momentum  * delta_alpha_env
 
@@
 
-  if(i %% 10 == 0){
+  mses[i] = mean((true_beta - beta)^2)
+  times[i] = as.integer(Sys.time())
+
+  if(i%%10 == 0){
     plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
@@
     )
-    abline(v = 0, h = 0)
+    abline(v = 0, h = 0, a = 0, b = 1)
   }
@@
      main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
-     las = 1,
-     cex.axis = 0.75
+     las = 1
     )
@@
 ```
+
+```{r}
+covs =cov(y)
+cor(covs[upper.tri(covs)], true_beta[upper.tri(true_beta)])^2
+```
+
+```{r}
+library(rags2ridges)
+inv = -ridgeS(cov(y), 5 / nrow(y))
+cor(inv[upper.tri(inv)], true_beta[upper.tri(true_beta)])^2
+```
+
+```{r}
+library(cowplot)
+library(ggplot2)
+out = plot_grid(
+  ggplot(NULL, aes(x = beta[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
+    stat_binhex() + 
+    xlab("Estimated coefficient value") + 
+    ylab("\"True\" coefficient value") + 
+    stat_hline(yintercept = 0, size = 1/8) + 
+    stat_vline(xintercept = 0, size = 1/8) + 
+    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
+    stat_smooth(method = "lm", se = FALSE, size = 1),
+
+  ggplot(NULL, aes(x = inv[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
+    stat_binhex() + 
+    xlab("Estimated coefficient value") + 
+    ylab("\"True\" coefficient value") + 
+    stat_hline(yintercept = 0, size = 1/8) + 
+    stat_vline(xintercept = 0, size = 1/8) + 
+    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
+    stat_smooth(method = "lm", se = FALSE, size = 1),
+
+  ggplot(NULL, aes(x = covs[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
+    stat_binhex() + 
+    xlab("Estimated coefficient value") + 
+    ylab("\"True\" coefficient value") + 
+    stat_hline(yintercept = 0, size = 1/8) + 
+    stat_vline(xintercept = 0, size = 1/8) + 
+    scale_fill_gradient(low = "#DDDDDD", high = "black", trans = "log10") +
+    stat_smooth(method = "lm", se = FALSE, size = 1),
+  nrow = 1,
+  labels = c("A. Network estimates", "B. Partial covariance", "C. Covariance")
+)
+save_plot("estimates.pdf", out, base_aspect_ratio = 3)
+
+```
+
+```{r}
+pdf("chapter-3/convergence.pdf", width = 10, height = 7)
+par(mfrow = c(1, 2))
+par(mar = c(5, 5, 4, 2) + .1)
+plot(function(x){exp(-x * 10) * .15}, type = "l", axes = FALSE, ylab = "mean square deviation from MLE", xlab = "time", main = "A.
+Convergence to known MLE\n(small network)", yaxs = "i");axis(1, labels = FALSE);axis(2, seq(0, .16, .02))
+text(.5, .1, "(illustration of\nexpected results)")
+plot(mses, type = "l", axes = FALSE, xlab = "time (units to be added)", ylab = "mean squared deviation\nfrom \"true\" values",ylim = c(0, .15), main = "B.
+convergence to unknown MLE\n(large network)", yaxs = "i");axis(2);axis(1, labels = FALSE)
+dev.off()
+```

### 2015-06-29 R/stochastic.R
stochastic stuff

@@ -1,0 +1,82 @@
+logistic = binomial()$linkinv
+
+rbern = function(p){rbinom(length(p), size = 1, prob = p)}
+
+#' @importFrom progress progress_bar
+f = function(
+  x, 
+  y, 
+  maxit, 
+  intercept_prior, 
+  alpha_x_prior, 
+  beta_prior, 
+  initial_learning_rate, 
+  rate_decay
+){
+  n_nodes = ncol(y)
+  n_samples = nrow(y)
+  
+  # Calculate sufficient statistics for the observed data
+  y_stats = crossprod(y)
+  xy_stats = t(x) %*% y
+  
+  alpha_x = delta_alpha_x = matrix(0, nrow = n_env, ncol = n_spp)
+  intercept = delta_intercept = qlogis(colMeans(y))
+  beta = delta_beta = matrix(0, nrow = n_nodes, ncol = n_nodes)
+  alpha = matrix(0, nrow = n_nodes, ncol = n_nodes) # no delta alpha to initialize
+  # b/c alpha not optimized directly
+  
+  pb = progress_bar$new(
+    format = "  Fitting [:bar] :percent eta: :eta",
+    total = maxit,
+    clear = FALSE
+  )
+  
+  # Initialize simulated matrix
+  y_sim = matrix(0.5, nrow = nrow(y), ncol = ncol(y))
+  
+  for(i in 1:maxit){
+    
+    #### Update learning rates ####
+    learning_rate = initial_learning_rate * rate_decay / (rate_decay + i)
+    
+    #### Gibbs sampling ####
+    # Update alpha
+    alpha = x %*% alpha_x %plus% intercept
+    
+    # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
+    for(j in sample.int(n_nodes)){
+      y_sim[,j] = rbern(logistic(alpha[ , j] + y_sim %*% beta[ , j]))
+    }
+  }
+  
+  #### Stochastic gradients ####
+  # Calculate sufficient statistics
+  y_sim_stats = crossprod(y_sim)
+  xy_sim_stats = t(x) %*% y_sim
+  
+  # Calculate the gradient with respect to alpha and beta
+  stats_difference = y_stats - y_sim_stats
+  intercept_grad = (diag(stats_difference) + alpha_prior$log_grad(alpha_x)) / n_samples
+  beta_grad = (stats_difference + beta_prior$log_grad(beta)) / n_samples
+  diag(beta_grad) = 0
+  
[…]

### 2015-07-01 Harris_2015/manuscript-materials/draft.Rmd
more dissertation stuff

@@
 
-Inferring species interactions from observational data is one of the most
+Estimating species interactions from observational data is one of the most
 controversial tasks in community ecology. One difficulty is that a single
@@
 ***Conditional relationships and Markov networks.***
-Ecologists are often interested in inferring direct interactions between
+Ecologists are often interested in learning about direct interactions between
 species, controlling for the indirect influence of other species. In
@@
 $2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
-required for exact inference spirals exponentially into the billions and beyond.
+required for exact estimation spirals exponentially into the billions and beyond.
 Numerous techniques are available for working with Markov networks that keep
@@
 ***Simulations.***
-In order to compare different methods for drawing inferences from
-observational data, I simulated two sets of landscapes using known
+In order to compare different methods, I simulated two sets of landscapes using known
 parameters.
@@
 
-***Other inference techniques for comparison.***
+***Comparison with existing techniques.***
 After fitting Markov networks to the simulated landscapes described above, I
-used several other techniques for inferring the sign and strength of marginal
+used several other techniques for determining the sign and strength of marginal
 associations between pairs of species (Appendix B).

### 2015-07-01 stochastic/chapter-3.Rmd
stochastic stuff
more dissertation stuff

@@
+# Abstract
+
+Species interactions are believed to play an important role in community structure, but detecting their influence in observed communities has stymied ecologists for decades.
+In recent years, several papers have suggested that ecologists could solve this long-standing problem by fitting Markov networks (undirected graphical models also known as Markov random fields) to the data by maximum likelihood or similar methods.
+This approach allows ecologists to control for extraneous factors when estimating the direct interactions between a given pair of species, and previous work has shown that this kind of control is required for reliable inferences.
+Unfortunately, the likelihood function for Markov networks quickly becomes intractable once more than about 20 species are included in the model, due to the exponentially large number of possible combinations of presences and absences.
+Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and attempt to explain *all* of the co-occurrence patterns in ecological data sets with species interactions.
+In this paper, I introduce *stochastic approximation* as a method for fitting these models to large ecological networks where hundreds of species and multiple abiotic factors can all affect a given species’ occurrence probability.
+This approach, which converges to the maximum likelihood estimate with probability one, greatly expands the range of data sets we can analyze and the number of ecological questions we can address with these methods.
+
 # Introduction
@@
 Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @strong_ecological_1984; @hastings_can_1987].
-[[**The next two sentences are unnecessarily confrontational and should probably be removed**]] *Despite these difficulties, ecologists have considered the underlying questions so important that, for the last few decades, they have continued to address them with whatever methods were available, even when these methods are known to perform poorly.
-For example, the False Discovery Rates above 95% demonstrated in @gotelli_empirical_2009 have not prevented wide adoption of method proposed in that paper*
+As documented below, ecologists have, for the most part, not solved these problems in the ensuing decades.
 
@@
 @friedman_sparse_2008).
-The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013], and most of the test statistics in common use are actually minor variations on the sample correlation [*evidence to be added to Chapter 2*].
-As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1, *to be added*).
-While some narrowly-focused methods have been proposed for incorporating a small number of specific factors into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many factors acting simultaneously.
+The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients [@pollock_understanding_2014; see @faisal_inferring_2010 for a recent exception].
+As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1).
+When these factors are not accounted for, co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
+While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.
 
-In other fields, scientists have a number of procedures for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables.
+A number of procedures have been developed for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables, i.e.
+for estimating the *conditional* relationship between them.
 The most familiar of these is the partial correlation.
 Rather than describing the *overall* relationship between two variables across *all* conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.
-@harris_estimating_2015 showed that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from co-occurrence data, and that Markov networks (undirected graphical models also known as Markov random fields), which do not assume normally-distributed observations do even better (cf @azaele_inferring_2010 and @fort_statistical_2013).
-A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network) can be estimated by maximum likelihood.
+@harris_estimating_2015 found that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
 
+Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @fort_statistical_2013; @harris_estimating_2015).
+A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network).
 Unfortunately, Markov networks have an intractable likelihood function whose computational complexity more than doubles each time a new species is added to the model.
-Thus, while Harris (2015) was able to use off-the-shelf optimization techniques to maximize the likelihood for networks of 20 species, extending the same approach to networks of 50-100 species would be completely infeasible.
+Thus, while @azaele_inferring_2010 and @harris_estimating_2015 were able to optimize an exact objective function to maximize the constrained likelihood for networks of up to 20 species, extending these approaches to networks of 50-100 species would be completely infeasible.
 If ecologists want the model to account for abiotic variation among sites on the landscape, then these expensive computations need to be repeated independently for every site, increasing the computational burden even further.
@@
 Despite the sampling error introduced by this substitution, stochastic approximation provides strong guarantees for eventual convergence to the maximum likelihood estimate.
-This change in approach makes it feasible to estimate the interactions among hundreds of species from observational data while also accounting for possible responses to abiotic factors.
+This change in approach makes it feasible to estimate the interactions among hundreds of species from observational data while also accounting for possible responses to abiotic factors, possibly giving ecologists some power to distinguish between species pairs that co-occur due to shared environmental tolerances from those that co-occur due to direct interactions like mutualism.
 
@@
 The coefficients linking species to one another describe their conditional associations: *all else equal*, a species will occur less often in the presence of its competitors and other species with which it is negatively associated, and more often in the presence of its mutualists.
-These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences.
+
+[[math for conditional distributions]]
+
+These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences [[Appendix?]].
 For example, hypothetical assemblages including many pairs of competitively exclusive species at the same site will tend to be very improbable under the model.
-
 Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.
@@
 
-In each of the scenarios below, the “observed” landscapes were simulated using Gibbs sampling from a pre-specified set of “true” parameters.
+In each of the scenarios below, the “observed” landscapes were simulated using Gibbs sampling from a pre-specified set of “true” parameters, then stochastic approximation was used to recover these parameters from the simulated presence-absence data.
+
+## Coefficient estimation with stochastic approximation
+
+A Markov network describes a probability distribution over possible assemblages, and this distribution can be summarized without loss of information by its *sufficient statistics*.
+For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [[*Appendix, not yet written*]].
+One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the average sufficient statistics produced by samples from the model [[ref?]].
+Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by climbing the likelihood surface.
+At this point, any discrepancies between the sufficient statistics of the model and of the data will have been eliminated.
+
+In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates this different exactly, averaging over *all* possible presence-absence combinations.
+This approach is intractable for larger networks, as the number of possibilities grows exponentially with the size of the network.
+Stochastic approximation [refs] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees [refs].
+The procedure iterates through the following three steps as many times as needed (5000 for these analyses):
+
+1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
+2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
+3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.
+
+The specific procedure used here had the following implementation details:
+
+The simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates.
+While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this merely slows convergence to the maximum likelihood estimate rather than preventing it altogether [@younes_convergence_1999; @salakhutdinov_efficient_2012].
+
+The approximate likelihood gradients in Step 2 match the ones from the Appendix of @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
+These gradients were augmented with a momentum term equal to 0.9 [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 1.0 (for environmental responses) or 0.1 (for pairwise interactions); the pairwise interaction terms were regularized more heavily to improve identifiability and reduce overfitting for this large set of 31,125 coefficients.
+
+The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 500/(500 - t - 1)$, which satisfies the criteria in @powell_approximate_2007.
+$\alpha_0$ was 1.0 for species’ intercepts and environmental responses.
+Because of the enormous number of pairwise interaction coefficients, I set $\alpha_0$ to 0.1 for the $\beta$ parameters.
 
@@
 
-In order to evaluate partial correlations' ability to recover the "true" parameters that generated a presence-absence matrix of interest, I first needed to generate such matrices from known processes.
+In order to assess the models' ability to recover the "true" parameters that generated a presence-absence matrix of interest, I first needed to generate such matrices from known processes.
 
-To demonstrate that the stochastic approximation method worked, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.
-These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@harris_rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
-For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described below.
+To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.
+These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@alia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
+For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.
 
@@
 
-## Coefficient estimation with stochastic approximation
+## Comparison with directed models
 
-A Markov network describes a probability distribution over possible assemblages, and this distribution can be summarized without loss of information by its *sufficient statistics*.
-For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [[*Appendix, not yet written*]].
-One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the average sufficient statistics produced by samples from the model [[ref?]].
-Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], any iterative procedure that can reduce this discrepancy quickly enough will converge to the global maximum likelihood estimate.
-
-The `rosalia` package reduces the discrepancy between observed and predicted sufficient statistics by calculating it exactly, averaging over *all* possible presence-absence combinations, but this approach is intractable for larger networks.
-Stochastic approximation [refs] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.
-The procedure iterates through the following three steps as many times as needed (5000 for these analyses):
-
-1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
-2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
-3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.
-
-The specific procedure used here had the following implementation details:
-
-The simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates.
-While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this difficulty does not prevent convergence to the maximum likelihood estimate [@younes_convergence_1999; @salakhutdinov_efficient_2012].
-
-The approximate likelihood gradients in Step 2 match the ones from the Appendix of @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
-These gradients were augmented with a momentum term equal to 0.9 [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 1.0 (for environmental responses) or 0.1 (for pairwise interactions); the pairwise interaction terms were regularized more heavily to improve identifiability and reduce overfitting for this large set of 31,125 coefficients.
-
-The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 500/(500 - t - 1)$, which satisfies the criteria in @powell_approximate_2007.
-$\alpha_0$ was 1.0 for species’ intercepts and environmental responses, and one tenth as large for their pairwise interaction coefficients.
+[[In response to Chapter 2, reviewers suggested comparisons with two other methods based on directed acyclic graphs.
+The first one has a latent multivariate Gaussian, and relies on the correlation structure of that Gaussian to help determine which species interact.
+The second one is based on sparse directed graphs among the species in the network.
+Should I address one or both of these approaches in this chapter as well?]]
 
@@
 
-The squared deviations between the exact estimates produced by the `rosalia` package and the ones produced by stochastic approximation quickly decayed to negligible levels [[Figure XA]], indicating that the stochastic approximation procedure was implemented correctly and worked as the mathematical theory predicts.
+For the smaller communities, the squared deviations between the exact estimates produced by the `rosalia` package and the ones produced by stochastic approximation quickly decayed to negligible levels [[Figure 2A]], indicating that the stochastic approximation procedure was implemented correctly and worked as the mathematical theory predicts.
 
@@
 
-For the larger landscape, I found that the stochastic approximation approach achieved reasonably good performance after less than a minute of optimization [Figure XB].
-With additional time, was able to recover most of the variance in species’ pairwise interactions from observational data ($R^2 = 0.66$; Figure YA), and nearly all of the variance in their responses to environmental variables ($R^2 = 0.95$, not shown).
+For the larger landscape, I found that the stochastic approximation approach achieved reasonably good performance after less than a minute of optimization [Figure 2B].
+With additional time, was able to recover most of the variance in species’ pairwise interactions from observational data ($R^2 = 0.66$; Figure 3A), and nearly all of the variance in their responses to environmental variables ($R^2 = 0.95$, not shown).
 
-While the maximum likelihood estimates produced by the Markov network worked well, the sample correlation, explained a negligible fraction of the variance in species’ pairwise interactions ($R^2 = 0.03$, Figure YB).
+While the maximum likelihood estimates produced by the Markov network worked well, the sample correlation, explained a negligible fraction of the variance in species’ pairwise interactions ($R^2 = 0.03$, Figure 3B).
 To the extent that ecologists’ other test statistics for co-occurrence are similar to the sample correlation [[*evidence to be added to chapter 2*]], they can be expected to perform equally poorly.
-As shown previously, the partial correlation was a much better indicator of species’ true interactions than the sample correlation ($R^2 = 0.44$ Figure YC).
+As shown previously, the partial correlation was a much better indicator of species’ true interactions than the sample correlation ($R^2 = 0.44$ Figure 3C).
 
@@
 
-For decades, ecologists have relied on minor variations on the correlation coefficient as their test statistics for inferring species interactions from observational data @harris_estimating_2015.
+For decades, ecologists have relied on poor test statistics for inferring species interactions from observational data [@harris_estimating_2015].
 As shown here and in @harris_estimating_2015, however, these inferences can only be made reliably by methods that control for other species in the network, such as partial correlations or Markov networks.
-The biggest computational problem with large Markov networks is the impossibility of averaging over all their possible states, but these results demonstrate that this step can be avoided.
-Stochastic approximation is able to recover the same estimates as exact methods while scaling gracefully to networks with hundreds of species.
+The biggest computational problem with large Markov networks is the impossibility of averaging over all their possible states, but these results demonstrate that this step can be avoided in ecological analyses.
+Stochastic approximation is able to recover the same estimates as exact methods for smaller networks, while scaling gracefully to networks with hundreds of species.
 

### 2015-07-07 stochastic/chapter-3.Rmd
stochastic

@@
 As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1).
-When these factors are not accounted for, co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
+When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
 While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.
+
+![A hypothetical network of three species, two of which depend on an abiotic factor (annual rainfall).
+**A.** The “true” network involves the caterpillar exploiting the two tree species as well as competition between the two trees.
+The two trees also benefit from rainfall.
+**B.** Although the two trees directly reduce one another’s occurrence probabilities via competition, they tend to occur in the same areas (with sufficient water and without too many herbivores).
+An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate.
+**C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates.
+Note that, with observational data, the bi-directional effects of species interactions (pairs of arrows between species) need to be collapsed to a single number describing the conditional relationship between species (one double-headed arrow).](figure-1.pdf)
 
@@
 
-[[math for conditional distributions]]
+[[add math for conditional distributions; same as in chapter 2]]
 
 These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences [[Appendix?]].
-For example, hypothetical assemblages including many pairs of competitively exclusive species at the same site will tend to be very improbable under the model.
+
+[[Add math for joint distribution; same as in chapter 2]]
+
 Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.
@@
 In each of the scenarios below, the “observed” landscapes were simulated using Gibbs sampling from a pre-specified set of “true” parameters, then stochastic approximation was used to recover these parameters from the simulated presence-absence data.
+
+## Conditioning the network on the abiotic environment
+
+The Markov networks fit in @harris_estimating_2015 treated the $\alpha_i$ terms from Equations [[1 and 2]] as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.
+In this paper, I extend the model so that the local abiotic conditions can also affect species occurrence probabilities directly.
+Specifically, the $\alpha_i$ terms in these analyses are linear combinations of the local environmental conditions, represented by $x_1$ through $x_5$.
+By estimating the coefficients associated with each of these $x$ variables for each species, we can account for their responses to the abiotic environment as well as to one another.
+In other fields, a model like this one, where a Markov network’s parameters depend on external factors is called a conditional random field [refs]; these models have been used in a variety of contexts, from language models to image processing [refs].
 
@@
 To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.
-These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@alia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
+These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
 For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.
@@
 
-![](convergence.pdf)
+![Stochastic approximation decreases the model’s error over time.
+**A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation reaches the global optimum very quickly.
+**B.** With larger networks, where the “true” parameter values that generated the network are known but the maximum likelihood estimates for the observed landscape are not, stochastic approximation initially drives the error down very quickly, and then slowly approaches the apparent optimum.](convergence.pdf)
 
@@
 
-![](estimates.pdf)
+![“True” network parameters for all 31,125 pairwise interactions versus the estimates from the three methods presented.
+Blue lines represent linear regressions through each cloud of points.](estimates.pdf)
 
@@
 
+Richer forms of data, e.g.
+from observational time series or controlled experiments, could help ecologists recover more information about species interactions than is possible with the kinds of observational data analyzed here: in particular, these richer forms of data will be necessary to answer questions about which species in a positively associated pair is facilitating the other or which species in a negatively associated pair is excluding the other.
+
 Ecologists could expand beyond simple Markov networks in a number of ways that would improve their ability to address a wider range of questions.

### 2015-07-07 stochastic/dynamic_simulation.R

@@ -1,0 +1,84 @@
+n_spp = 15
+K = 20
+
+maxit = 1E7
+thin = 4E3
+
+seed_rain = rep(.1, n_spp)
+growth = rep(1, n_spp)
+
+competition_vector = ifelse(
+  rbinom(choose(n_spp, 2), size = 1, prob = .5),
+  rexp(choose(n_spp, 2), 5),
+  rexp(choose(n_spp, 2), 1/5)
+)
+
+competition = matrix(0, nrow = n_spp, ncol = n_spp)
+competition[upper.tri(competition)] = competition_vector
+competition = competition + t(competition)
+diag(competition) = 1
+
+pops = matrix(nrow = n_spp, ncol = maxit / thin)
+
+population = rep(ceiling(K / n_spp / mean(competition)), n_spp)
+
+for(i in 1:maxit){
+  local_growth = growth * population
+  birthrates = seed_rain + local_growth
+  deathrates = local_growth * (population %*% competition) / K
+  
+  rates = c(birthrates, deathrates)
+  
+  event = sample.int(length(rates), 1, prob = rates)
+  if(event <= length(birthrates)){
+    # Event from first batch, i.e. birth.  Increase the population by one
+    index = event
+    population[index] = population[index] + 1
+  }else{
+    # Event from the second batch, i.e. death. If population is nonzero,
+    # decrease it by one for the corresponding species
+    index = event - n_spp
+    if(population[index] > 0){
+      population[index] = population[index] - 1
+    }
+  }
+  if(i%%thin == 0){
+    pops[ , i %/% thin] = population
+  }
+}
+
+matplot(t(pops), type = "l", lty = 1)
+
+
+library(corpcor)
+pcor = pcor.shrink(t(pops))
+class(pcor) = "matrix"
+lattice::levelplot(pcor^2)
+
+plot(pcor[upper.tri(pcor)], -competition_vector)
+
+library(rosalia)
+fit = rosalia(
+  t(pops > 0) * 1, 
+  hessian = TRUE, 
+  maxit = 1E3
+)
+
+mles = fit$beta[upper.tri(fit$beta)]
+ses = sqrt(diag(solve(fit$opt$hessian)))[-(1:n_spp)]
+is_significant = ifelse(mles > 0, mles - 2 * ses > 0, mles + 2 * ses < 0)
+
+
[…]

### 2015-07-15 Harris_2015/manuscript-materials/draft.Rmd
more

@@
 
+\setlength{\parindent}{1cm}
+
+\noindent  
 ***Introduction***
@@
 methods for detecting competition's effects on community structure are
-unreliable [@gotelli_empirical_2009], and thus important ecological processes
-remain poorly understood. More generally, it can be difficult to reason about
-the complex web of direct and indirect species interactions in real
-assemblages, especially when these interactions occur against a background of
-other ecological processes such as dispersal limitation and environmental filtering
-[@connor_checkered_2013]. For this reason, it isn't always clear what kinds of
-patterns would even constitute evidence of competition, as opposed to some
-other biological process or random sampling error
-[@roughgarden_competition_1983; @lewin_santa_1983].
+unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
+another can become lost in the complex web of direct and indirect interactions
+in real assemblages. For example, the competitive interaction between the
+two shrub species in Figure 1A can become obscured by their shared tendency
+to occur in unshaded areas (Figure 1B). If this sort of effect is common, then
+the vast majority of ecologists' methods, which rely on test statistics describing
+the total number of co-occurrences or overall correlation between putatively
+interacting species [@diamond_island_1975; @strong_ecological_1984;
+@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
+will not generally provide much evidence regarding species' direct effects on
+one another.
 
-Most existing methods in this field compare the frequency with which two
-putative competitors are observed to co-occur, versus the frequency that would
-be expected if *all* species on the landscape were independent
-[@strong_ecological_1984; @gotelli_empirical_2009]. Examining a species pair
-against such a "null" background, however, rules out the possibility that the
-overall association between two species could be driven by an outside force.
-For example, even though the two shrub species in Figure 1 compete with one
-another for resources at a mechanistic level, they end up clustering together
-on the landscape because they both grow best in areas that are not
-overshadowed by trees. If this sort of effect is common, then significant
-deviations from independence will not generally provide convincing
-evidence of species' direct effects on one another.
+While competition doesn't reliably leave the expected pattern in community
+structure (a low co-occurrence rate at the landscape level), it nevertheless does
+leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
+there will be a deficit of co-occurrences between the competing shrub species,
+and *among unshaded sites*, there will also be such a deficit. More generally,
+we can obtain much better estimates of the association between two species from
+their conditional relationships (i.e. by controlling for other species in the
+network) than we could get from their overall co-occurrence rates.
 
-While competition between the two shrubs in the previous example does not
-leave the commonly-expected pattern in community structure (negative
-association at the landscape level), it nevertheless does leave a signal in
-the data (Figure 1C). Specifically, *among shaded sites*, there will be a
-deficit of co-occurrences, and *among unshaded sites*, there will also be such
-a deficit.
-
-In this paper, I introduce Markov networks [undirected graphical models also
-known as Markov random fields; @murphy_machine_2012] as a framework for
-understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them with observational data. Markov networks,
-which generalize partial correlations to non-Gaussian data
-[@lee_learning_2012; @loh_structure_2013], have been used in many scientific
-fields for decades to model associations between various kinds of "particles".
-For example, a well-studied network named the Ising model has played an
+In this paper, following recent work by @azaele_inferring_2010 and
+@fort_statistical_2013, I show that Markov networks [undirected graphical models
+also known as Markov random fields; @murphy_machine_2012] can provide a framework
+for understanding the landscape-level consequences of pairwise species
+interactions, and for detecting them from observational data. Markov networks
+have been used in many scientific fields for decades to model associations
+between various kinds of "particles". For example, a well-studied network named the Ising model has played an
 important role in our understanding of physics [where nearby particles tend to
@@
 connected to one another by modeling the structure in their firing patterns
-[@schneidman_weak_2006]. Following recent work by @azaele_inferring_2010 and
-@fort_statistical_2013, I suggest that ecologists could similarly treat
-species as the interacting particles in the same modeling framework. Doing so
+[@schneidman_weak_2006].
+
+Doing so
 would allow ecologists to simulate and study the landscape-level consequences
@@
 
-***Methods***
+\noindent  ***Methods***
 
+\noindent  
 ***Conditional relationships and Markov networks.***
@@
 
+\noindent  
 ***Simulations.***
@@
 
+\noindent  
 ***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
@@
 
+\noindent  
 ***Comparison with existing techniques.***
@@
 
+\noindent  
 ***Method evaluation.***
@@
 
+\noindent  
 ***Results***
 
+\noindent  
 ***Three species.***
@@
 
+\noindent  
 ***Larger landscapes.***
@@
 
+\noindent  
 ***Discussion***
 
+\noindent  
 The results presented above show that Markov networks can reliably recover
@@
 
+\noindent  
 ***Acknowledgements:***
@@
 this work.
+
+\setlength{\parindent}{0cm}
 

### 2015-07-15 stochastic/dynamic_simulation.R

@@ -1,16 +1,18 @@
+set.seed(1)
+
 n_spp = 15
-K = 20
+K = 15
 
 maxit = 1E7
-thin = 4E3
+thin = 2E4
 
 seed_rain = rep(.1, n_spp)
 growth = rep(1, n_spp)
 
 competition_vector = ifelse(
   rbinom(choose(n_spp, 2), size = 1, prob = .5),
-  rexp(choose(n_spp, 2), 5),
-  rexp(choose(n_spp, 2), 1/5)
+  rexp(choose(n_spp, 2), 2),
+  rexp(choose(n_spp, 2), 1/2)
 )
 
 competition = matrix(0, nrow = n_spp, ncol = n_spp)
@@ -50,35 +52,54 @@ for(i in 1:maxit){
 matplot(t(pops), type = "l", lty = 1)
 
 
-library(corpcor)
-pcor = pcor.shrink(t(pops))
-class(pcor) = "matrix"
-lattice::levelplot(pcor^2)
-
-plot(pcor[upper.tri(pcor)], -competition_vector)
-
 library(rosalia)
 fit = rosalia(
   t(pops > 0) * 1, 
   hessian = TRUE, 
-  maxit = 1E3
+  maxit = 1E3,
+  prior = make_logistic_prior(location = 0, scale = 2)
+)
+
+
+
+library(BayesComm)
+bc = BC(Y = t(pops > 0), model = "community", its = 1000, thin = 10)
+
+
+# Markov network performance
+cor(fit$beta[upper.tri(fit$beta)], -competition_vector, method = "spearman")
+
+# Correlation performance
+cor(
+  cor(t(pops > 0))[upper.tri(cor(t(pops > 0)))], 
+  -competition_vector, method = "spearman"
+)
+
+# Partial correlation performance
+cor( 
+  -solve(cor(t(pops > 0)))[upper.tri(cor(t(pops > 0)))], 
+  -competition_vector, 
+  method = "spearman"
+)
+
+# BayesComm performance
+cor( 
+  colMeans(bc$trace$R), 
+  -competition_vector, 
+  method = "spearman"
 )
 
 mles = fit$beta[upper.tri(fit$beta)]
 ses = sqrt(diag(solve(fit$opt$hessian)))[-(1:n_spp)]
 is_significant = ifelse(mles > 0, mles - 2 * ses > 0, mles + 2 * ses < 0)
 
-
[…]

### 2015-07-16 stochastic/dynamic_simulation.R
added mutualisms to dynamic simulation

@@ -1,33 +1,55 @@
 set.seed(1)
 
-n_spp = 15
-K = 15
+library(progress)
 
-maxit = 1E7
-thin = 2E4
+n_spp = 20
+K = n_spp / 3
+
+maxit = 1E6
+thin = 2E3
 
 seed_rain = rep(.1, n_spp)
 growth = rep(1, n_spp)
 
-competition_vector = ifelse(
-  rbinom(choose(n_spp, 2), size = 1, prob = .5),
-  rexp(choose(n_spp, 2), 2),
-  rexp(choose(n_spp, 2), 1/2)
+interaction_vector = rexp(choose(n_spp, 2), 1/2)
+
+interaction_signs = ifelse(
+  rbinom(choose(n_spp, 2), size = 1, prob = .75),
+  -1, 
+  1
 )
+
+signed_interaction_vector = interaction_vector * interaction_signs
+competition_vector = interaction_vector * (interaction_signs == -1)
+mutualism_vector = interaction_vector * (interaction_signs == +1)
+
+mutualism_scaler = function(x){
+  (1 + exp(-4 * x / K / n_spp)) / 2
+}
 
 competition = matrix(0, nrow = n_spp, ncol = n_spp)
 competition[upper.tri(competition)] = competition_vector
 competition = competition + t(competition)
-diag(competition) = 1
+
+mutualism = matrix(0, nrow = n_spp, ncol = n_spp)
+mutualism[upper.tri(mutualism)] = mutualism_vector
+mutualism = mutualism + t(mutualism)
+
+
 
 pops = matrix(nrow = n_spp, ncol = maxit / thin)
 
 population = rep(ceiling(K / n_spp / mean(competition)), n_spp)
 
+pb = progress_bar$new(total = maxit / thin)
+
 for(i in 1:maxit){
   local_growth = growth * population
+  interactions = population * mutualism_scaler(population %*% mutualism) + 
+    population %*% competition
+  
   birthrates = seed_rain + local_growth
-  deathrates = local_growth * (population %*% competition) / K
+  deathrates = local_growth * interactions / K
   
   rates = c(birthrates, deathrates)
   
@@ -46,11 +68,12 @@ for(i in 1:maxit){
   }
   if(i%%thin == 0){
     pops[ , i %/% thin] = population
+    pb$tick()
   }
 }
 
 matplot(t(pops), type = "l", lty = 1)
-
[…]

### 2015-07-21 Harris_2015/manuscript-materials/draft.Rmd
new analyses

@@
 \noindent  
-***Simulations.***
+***Simulated landscapes.***
 In order to compare different methods, I simulated two sets of landscapes using known
@@
 
-I also simulated a second set of landscapes with five, ten, or twenty
-potentially-interacting species on landscapes composed of 20, 100, 500, or
-2500 observed communities (24 replicate simulations for each combination; Appendix B).
-These simulated data sets span the range from small, single-observer data
-sets to large collaborative efforts such as the North American Breeding Bird
-Survey. As described in Appendix B, I randomly drew the "true" coefficient
-values for each replicate so that most species pairs interacted negligibly, a
-few pairs interacted very strongly, and competition was three times more
-common than facilitation. I then used Gibbs sampling to randomly generate
-replicate landscapes with varying numbers of species and sites (Appendix B).
-For half of the simulated landscapes, I treated each
-species' $\alpha$ coefficient as a constant, as described above. For the other
-half, I treated the $\alpha$ coefficients as linear functions of two abiotic
-environmental factors that varied from location to location across the
-landscape (Appendix B). The latter set of simulated landscapes provide an
-important test of the methods' ability to distinguish co-occurrence patterns
-that were generated from pairwise biotic interactions from
-those that were generated by external forces like abiotic environmental
-filtering. This task was made especially difficult because---as with most
-analyses of presence-absence data for co-occurrence patterns---the inference
-procedure did not have access to any information about the environmental or
-spatial variables that helped shape the landscape [cf @connor_checkered_2013;
-@blois_framework_2014].
+I also simulated a second set of landscapes using a stochastic community
+model based on generalized Lotka-Volterra dynamics, as described in the
+Appendix. In these simulations, each species pair was randomly assigned to
+either compete for a portion of the available carrying capacity (negative
+interaction) or to act as mutualists (positive interaction).  In these
+simulations, mutualism operated by mitigating the effects of intraspecific
+competition on each partner's death rate.  For these analyses, I simulated
+50 replicate landscapes, each with 25, 200, or 1600 sites and 20 species.
 

### 2015-07-21 Harris_2015/simulation.Rmd
new analyses

```{r}
library(progress)
set.seed(1)
```

```{r}

simulate_communities = function(n_spp = 20,
         K = n_spp / 3,
         maxit = 5E6,
         thin = 1E4,
         seed_rain = rep(.1, n_spp),
         growth = rep(1, n_spp),
         interaction_strength = 1/2,
         prob_negative = 0.75,
         mutualism_maximum = 0.5
){

  # Intraspecific competition can be mitigated by mutualists, up to some
  # maximal amount.  This function rescales the mutualism effect so that
  # it doesn't turn off intraspecific competition entirely.
  mutualism_scaler = function(x, mutualism_maximum){
    (1 + exp(-4 * x / K / n_spp)) * (1 - mutualism_maximum)
  }

  # Randomly generate a vector of interaction strengths (one for each 
  # pair of species)
  interaction_vector = rexp(choose(n_spp, 2), interaction_strength)

  # Randomly mark interactions as positive or negative
  interaction_signs = ifelse(
    rbinom(choose(n_spp, 2), size = 1, prob = prob_negative),
    -1, 
    1
  )

  # Negative interactions are competition; positive ones are mutualisms
  competition_vector = interaction_vector * (interaction_signs == -1)
  mutualism_vector = interaction_vector * (interaction_signs == +1)

  # Fill in the competition and mutualism matrices:
  # Start with a matrix of zeros
  competition = matrix(0, nrow = n_spp, ncol = n_spp)
  # Fill in the upper triangle with the appropriate interaction vector
  competition[upper.tri(competition)] = competition_vector
  # Symmetrize the matrix by adding it to its transpose
  competition = competition + t(competition)

  # Note that intraspecific competition (which would go along the diagonal)
  # is handled separately, and the diagonal is left with zeros.

  mutualism = matrix(0, nrow = n_spp, ncol = n_spp)
  mutualism[upper.tri(mutualism)] = mutualism_vector
  mutualism = mutualism + t(mutualism)

  # Create an empty matrix for storing population sizes
  pops = matrix(nrow = n_spp, ncol = maxit / thin)

  # Initialize the population with a fixed number of individuals per species,
  # depending on the number of species, carrying capacity, and strength of
  # interspecific competition.
  population = rep(ceiling(K / n_spp / mean(competition)), n_spp)

  for(i in 1:maxit){

    # The local growth rate is proportional to population size
    local_growth = growth * population

    # Mutualists act by mitigating intraspecific competition
    mutualist_mitigation = mutualism_scaler(
      population %*% mutualism, 
      mutualism_maximum
    )

    # Add up the two types of interactions affecting death rates
    interactions = population * mutualist_mitigation + population %*% competition

    # Germinating seeds can come from local growth or from outside the system
    birthrates = seed_rain + local_growth

    # Death rates depend on population size, interactions, and carrying capacity
    deathrates = local_growth * interactions / K

    # Generate a vector of vital rates
    rates = c(birthrates, deathrates)

    # Randomly select a birth/death event to occur. Probability
    # of occurring is proportional to the corresponding rate
    event = sample.int(length(rates), 1, prob = rates)

    if(event <= length(birthrates)){
      # Event from first batch, i.e. birth.  Increase the population by one
      index = event
      population[index] = population[index] + 1
    }else{
      # Event from the second batch, i.e. death. If population is nonzero,
      # decrease it by one for the corresponding species
      index = event - n_spp
      if(population[index] > 0){
        population[index] = population[index] - 1
      }
    }

    if(i%%thin == 0){
      # Save the results in the `pops` matrix
      pops[ , i %/% thin] = population
    }
  }

  # Return the "observed" presence-absence vector
  # along with the"true" interaction signs and strengths.
  # The output is transposed because all fitting methods but Pairs
  # require it in that format.
  # Multiplying by one turns TRUEs into 1s and FALSEs into 0s
  list(
    observed = t(pops > 0) * 1,
    truth = interaction_vector * interaction_signs
  )
}
```

```{r}
thin = 5E3
n_spp = 20
n_reps = 50

for(n_sites in c(25, 200, 1600)){
  message(n_sites)

  pb = progress_bar$new(total = n_reps)
  for(rep in 1:n_reps){
    id = paste(n_spp, n_sites, rep, sep = "-")

    results = simulate_communities(
      maxit = n_sites * thin, 
      thin = thin, 
      n_spp = n_spp
    )

    pb$tick()

    saveRDS(results, file = paste0("fakedata/", id, ".rds"))

    # Gotelli and Ulrich's Pairs software needs a different format.
    x = results$observed

    # It rejects empty sites, so I remove them here
    x_subset = x[rowSums(x) != 0, colSums(x) != 0]

    # It also expects the data matrices to be transposed
    # relative to the other methods
    write.table(
      t(x_subset), 
      file = paste0("fakedata/", id, "-transposed.txt"), 
      quote = FALSE
    )
  }
}
```

### 2015-07-21 Harris_2015/recovery.R

@@ -1,0 +1,115 @@
+library(dplyr)
+library(corpcor)
+library(rosalia)
+library(arm)
+library(BayesComm)
+
+set.seed(1)
+
+f = function(filename){
+  # Multiplying by one is necessary to prevent silly errors
+  # regarding TRUE/FALSE versus 1/0
+  raw_obs = readRDS(filename)[["observed"]] * 1
+  
+  # Identify species that are never present (or never absent) so they 
+  # can be dropped
+  species_is_variable = diag(var(raw_obs)) > 0
+  pair_is_variable = tcrossprod(species_is_variable) > 0
+  
+  x = raw_obs[ , species_is_variable]
+  truth = readRDS(filename)[["truth"]][pair_is_variable[upper.tri(pair_is_variable)]]
+  
+  p_corr = pcor.shrink(x)
+  
+  splitname = strsplit(filename, "/|-|\\.")[[1]]
+  n_sites = as.integer(splitname[[3]])
+  rep = as.integer(splitname[[4]])
+  
+  corr = cor(x)
+  
+  coef_matrix = matrix(0, ncol(x), ncol(x))
+  for(i in 1:ncol(x)){
+    if(var(x[,i]) > 0){
+      coefs = coef(bayesglm(x[,i] ~ x[ , -i], family = binomial))[-1]
+      coef_matrix[i, -i] = coefs
+    }
+  }
+  coef_matrix = (coef_matrix + t(coef_matrix)) / 2
+  
+  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_cauchy_prior(scale = 1))
+  
+  bc = BC(Y = x, model = "community", its = 1000)
+  
+  data.frame(
+    truth = truth, 
+    n_sites = n_sites,
+    rep = rep,
+    `partial correlation` = p_corr[upper.tri(p_corr)],
+    correlation = corr[upper.tri(corr)],
+    `Markov network` = rosie$beta[upper.tri(rosie$beta)],
+    GLM = coef_matrix[upper.tri(coef_matrix)],
+    `BayesComm` = colMeans(bc$trace$R)
+  )
+}
+
+
+
+files = dir("fakedata", pattern = "\\.rds$", full.names = TRUE)
+
+z = lapply(files, f) %>% bind_rows %>% as.data.frame
+colnames(z) = gsub("\\.", " ", colnames(z))
+
+library(ggplot2)
[…]

### 2015-07-21 R/exact.R

@@ -26,6 +26,33 @@ make_logistic_prior = function(location = 0, scale, ...){
   )
 }
 
+#' @export
+make_cauchy_prior = function(location = 0, scale, ...){
+  structure(
+    list(
+      log_d = function(x){dcauchy(x, location = location, scale = scale, log = TRUE)},
+      log_grad = function(x){
+        (-2 * x + 2 * location)/(scale^2 + (x - location)^2)
+      }
+    ),
+    class = "prior"
+  )
+}
+
+#' @export
+make_t_prior = function(df){
+  structure(
+    list(
+      log_d = function(x){dt(x, df, log = TRUE)},
+      log_grad = function(x){
+        -(x * (1 + df)) / (x^2 + df)
+      }
+    ),
+    class = "prior"
+  )
+}
+
+
 
 #' Make a flat prior for maximum likelihood estimation
 #' 

### 2015-07-21 inst/tests/tests.R

@@ -219,3 +219,31 @@ test_that("logistic prior works", {
   )
 })
 
+
+test_that("cauchy prior works", {
+  x = c(pi, log(7))
+  eps = 1E-6
+  s = c(exp(2), sqrt(12))
+  
+  diff = dcauchy(x + eps, scale = s, log = TRUE) - 
+    dcauchy(x - eps, scale = s, log = TRUE)
+  
+  expect_equal(
+    make_cauchy_prior(scale = s)$log_grad(x),
+    (diff) / 2 / eps
+  )
+})
+
+test_that("t prior works", {
+  x = c(pi, log(7))
+  eps = 1E-6
+  df = c(exp(2), sqrt(12))
+  
+  diff = dt(x + eps, df = df, log = TRUE) - 
+    dt(x - eps, df = df, log = TRUE)
+  
+  expect_equal(
+    make_t_prior(df = df)$log_grad(x),
+    (diff) / 2 / eps
+  )
+})

### 2015-07-22 Harris_2015/manuscript-materials/draft.Rmd
draft

@@
 connected to one another by modeling the structure in their firing patterns
-[@schneidman_weak_2006].
-
-Doing so
-would allow ecologists to simulate and study the landscape-level consequences
-of arbitrary species interaction matrices, even when our observations are not
-Gaussian. While ecologists explored some related approaches in the 1980's
+[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
 [@whittam_species_1981], computational limitations had previously imposed
@@
 "probabilities" greater than one; @gilpin_factors_1982].  Now that it is
-computationally feasible to fit these models exactly, the approach has become
+computationally feasible to fit these models exactly, this kind of approach has become
 worth a second look.
@@
 absolute probability estimates. Turning the relative probability given by
-Equation 1 into an absolute probability entails scaling by a *partition
-function*, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
+Equation 1 into an absolute probability entails scaling by a normalizing constant, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
 possible assemblages that could be produced by the model sum to one (bottom of
@@
 $2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
-required for exact estimation spirals exponentially into the billions and beyond.
+required spirals exponentially into the billions and beyond.
 Numerous techniques are available for working with Markov networks that keep
@@
 but they are beyond the scope of this paper.
+
+\noindent  
+***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
+In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
+however, ecologists will often need to estimate these parameters from co-occurrence data.
+When the number of species is reasonably small, one can compute exact maximum
+likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
+$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
+networks like the ones considered here have unimodal likelihood surfaces
+[@murphy_machine_2012], ensuring that this procedure will always converge on
+the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
+coefficients that would be expected to produce exactly the observed co-occurrence
+frequencies. For the analyses in this paper, I used the *rosalia* package
+[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
+define the objective function and gradient as R code.  The rosalia package then
+uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
+and $\beta$.
 
@@
 \noindent  
-***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
-In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
-however, ecologists will often need to estimate these parameters from co-occurrence data.
-When the number of species is reasonably small, one can compute exact maximum
-likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
-$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
-networks like the ones considered here have unimodal likelihood surfaces
-[@murphy_machine_2012], ensuring that this procedure will always converge on
-the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
-coefficients that would be expected to produce exactly the observed co-occurrence
-frequencies. For the analyses in this paper, I used the *rosalia* package
-[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
-define the objective function and gradient as R code.  The rosalia package then
-uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
-and $\beta$.
+***Recovering species interactions from simulated data.***
+I used Markov networks and  several other techniques for determining the sign and strength
+of the associations between pairs of species (Appendix B).
 
-For analyses with 5 or more species, I made a small modification to the maximum
-likelihood procedure described above. Given the large number of parameters
-associated with some of the networks to be estimated, I regularized the
-likelihood using a logistic prior distribution [@gelman_weakly_2008] with
-a scale of 1 on the $\alpha$ and $\beta$ terms.
+For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
+above. Given the large number of parameters associated with some of the networks to
+be estimated, I regularized the likelihood using a Cauchy prior distribution
+[@gelman_weakly_2008] with a scale of 1 on the $\alpha$ and $\beta$ terms.
 
-\noindent  
-***Comparison with existing techniques.***
-After fitting Markov networks to the simulated landscapes described above, I
-used several other techniques for determining the sign and strength of marginal
-associations between pairs of species (Appendix B).
+I also evaluated five alternative methods from the existing literature to calibrate the
+performance of the Markov network as a method for estimating species interactions.
+The first two alternative interaction measures were the sample correlations and the
+partial correlations between each pair of species' data vectors on the landscape
+[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations are undefined
+for landscapes with perfectly-correlated species pairs, I used a regularized
+estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
+`pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
+data, the partial correlation can be thought of as an approximation to the Markov network,
+and is much easier to compute.
 
-The first two alternative interaction measures were the sample covariances and the
-partial covariances between each pair of species' data vectors on the landscape
-[@albrecht_spatial_2001]. Because partial covariances are undefined for
-landscapes with perfectly-correlated species pairs, I used a regularized
-estimate based on ridge regression [@van_wieringen_ridge_2014; i.e. linear
-regression with a Gaussian prior]. For these analyses, I set the ridge
-parameter to 0.2 divided by the number of sites on the landscape.
+The third alternative, generalized linear models, can also be thought of as a computationally
+efficient approximation to the Markov network [@REF].  Following @faisal_inferring_2010,
+I fit regularized logistic regression models for each species, using the other species on the
+landscape as predictors. [[Observational presence-absence data does not have enough degrees
+of freedom to estimate this many parameters, so I symmetrized the pairwise correlations by
+averaging the coefficient for predicting species $i$’s status from species $j$ and vice versa.]]
 
-The third alternative method, described in @gotelli_empirical_2009, involved simulating
+The fourth alternative method, described in @gotelli_empirical_2009, involved simulating
 possible landscapes from a null model that retains the row and column sums of
@@
 
+Finally, I used the latent correlation matrix estimated by the BayesComm package [@REF]
+to evaluate the claim from recent papers that the correlation coefficients estimated by “joint
+species distribution models” provide an accurate assessment of species’ pairwise interactions
+[@pollock_understanding_2014; see also @harris_generating_2015].
+
 \noindent  
-***Method evaluation.***
+***Evaluating model performance.***
 For the simulated landscapes based on Figure 1, method evaluation was fairly
 qualitative: any method whose test statistic for the two shrubs indicated a
-negative relationship passed; other methods failed.
+negative relationship passed; other methods failed. For methods that provide
+confidence intervals or other inferential statistics, I reported them below.
 
-For the larger landscapes, I rescaled the four methods' estimates using linear
-regression through the origin so that they all had a consistent
-interpretation. In each method, I regressed the "true" $\beta$ coefficient for
-each species pair against the model's estimate, re-weighting the pairs so that
-each landscape contributed equally to the rescaled estimate. These regressions
-yielded squared errors for each of the 23,140 simulated species pairs, across
-all conditions and replicates. I then partitioned these errors into groups
-defined by the properties of the simulated landscapes (i.e. species richness,
-number of observed communities, and the presence/absence of environmental
-filtering). Within each partition, the mean squared error was used to
-calculate the proportion of variance explained (compared with a baseline model
-that assumed all interaction strengths to be zero).
+Because the different methods mostly describe species interactions on different
+scales (e.g. correlations versus $Z$ scores versus regression coefficients), I used
+linear regression through the origin to rescale the different estimates produced
+by each method so that they had a consistent interpretation. After rescaling each
+method’s estimates, I calculated squared errors for each of the [[N]] species pairs
+across all the simulated data sets and used these values to calculate the proportion of variance
+explained for landscapes with 25, 200, or 1600 sampled locations (compared with a baseline
+model that assumed all interaction strengths to be zero).
 

### 2015-07-22 Harris_2015/manuscript-materials/draft.Rmd
draft

@@
 their conditional relationships (i.e. by controlling for other species in the
-network) than we could get from their overall co-occurrence rates.
+network) than we could get from their overall co-occurrence rates. This kind of
+control is difficult to obtain from null models, which begin with the assumptions
+that all the pairwise interactions are zero and thus don’t need to be controlled for.
 
@@
 interactions, and for detecting them from observational data. Markov networks
-have been used in many scientific fields for decades to model associations
-between various kinds of "particles". For example, a well-studied network named the Ising model has played an
-important role in our understanding of physics [where nearby particles tend to
-align magnetically with one another; @cipra_introduction_1987]. In spatial
-contexts, these models have been used to describe interactions between
+have been used in many scientific fields for decades in similar contexts.
+For example, a well-studied network named the Ising model has played an important
+role in our understanding of physics [where nearby particles tend to align
+magnetically with one another; @cipra_introduction_1987].
+In spatial contexts, these models have been used to describe interactions between
 adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
-neurobiology, they have helped researchers determine which neurons are
-connected to one another by modeling the structure in their firing patterns
+neurobiology, they have helped researchers identify neural connections from firing patterns
 [@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
@@
 "probabilities" greater than one; @gilpin_factors_1982].  Now that it is
-computationally feasible to fit these models exactly, this kind of approach has become
+computationally feasible to fit these models exactly, this kind of approach is
 worth a second look.
@@
 absolute probability estimates. Turning the relative probability given by
-Equation 1 into an absolute probability entails scaling by a normalizing constant, $Z(\alpha, \beta)$, which is defined so that that the probabilities of all
+Equation 1 into an absolute probability entails scaling by a normalizing constant,
+$Z(\alpha, \beta)$, which is defined so that that the probabilities of all
 possible assemblages that could be produced by the model sum to one (bottom of
@@
 [@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
-define the objective function and gradient as R code.  The rosalia package then
-uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
-and $\beta$.
+define the objective function and gradient as R code.  The rosalia package uses
+standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.
 
@@
 were successful even when direct interactions were largely overwhelmed by indirect
-effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases
-where fitting a Markov network is computationally infeasible, these results also
-indicate that partial covariances---which can be computed straightforwardly by linear
-regression---can often provide a surprisingly useful approximation. The
-partial correlations' success on simulated data may not
-carry over to real data sets, however; @loh_structure_2013 show that the linear
+effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
+these results also indicate that partial covariances and generalized linear models (the two
+methods that estimated conditional relationships rather than marginal ones) can both
+provide useful approximations. The partial correlations' success on simulated
+data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
 approximations can be less reliable in cases where the true interaction matrix
-contains more structure (e.g. guilds or trophic levels). On the other hand,
-if ecologists are familiar enough with the natural history of their study
-systems to describe this kind of structure as a prior distribution on the
-parameters or as a penalty on the likelihood, then this information could
-reduce the effective degres of freedom to estimate and real-world results might
-be even better than those shown in Figure 3.
+contains more structure (e.g. guilds or trophic levels).
 
-Ecologists will also need natural history to pin down the exact nature of the
-interactions identified by a network model (e.g. which species in a
-positively-associated pair is facilitating the other),  particularly when real
-pairs of species can reciprocally influence one another in multiple ways
-simultaneously [@bruno_inclusion_2003]; the $\beta$ coefficients in Markov
-networks have to reduce this complexity  to a single number. In short, partial
-correlations and Markov  networks both help prevent us from mistaking marginal
-associations for conditional ones, but they can't tell us the underlying
-biological mechanisms at work.
+It’s important to note that none of these methods can identify the exact nature of the
+pairwise interactions (e.g. which species in a positively-associated pair is facilitating the other),
+particularly when real pairs of species can reciprocally influence one another in multiple ways
+simultaneously [@bruno_inclusion_2003]; with compositional data, there is only enough
+information to provide a single number describing each species pair.  On the other hand, if
+ecologists are familiar enough with the natural history of their study
+systems that they can augment the likelihood function with a prior distribution or a suitable
+penalty, then this information could reduce the effective degrees of freedom to estimate and
+real-world results might be even better than those shown in Figure 3.
 
@@
 to the question of how competition should affect a species' overall
-prevalence, which was a major flash point for the null model debates in the
+prevalence, which was a major flashpoint for the null model debates in the
 1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
@@
 Competition's effect on prevalence in a Markov network can then be calculated
-by subtracting this value and the observed prevalence (cf Figure 2D).
+by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
+BayesComm (which had the worst performance of any method evaluated here), none of
+the other methods could produce this inference because they do not make predictions
+about species composition. [[*back this up with more evidence or remove it*]]
 
@@
 and the models' properties,
-capabilities, and limits are well-understood in a huge range of applications,
-from spatial statistics [@gelfand_modelling_2005] to neuroscience
-[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
+capabilities, and limits are well-understood in a huge range of applications.
 Modeling species interactions using the same framework would thus allow
 ecologists to tap into an enormous set of existing discoveries and techniques
-for dealing with indirect effects, stability, and alternative stable states
-(i.e. phase transitions; @cipra_introduction_1987).
-
-This modeling approach is also highly
-extensible, even when it is inconvenient to compute the likelihood exactly.
-For example, the mistnet software package for joint species distribution modeling
-[@harris_generating_2015] can fit *approximate* Markov networks to large
-species assemblages (>300 species) while simultaneously modeling each species'
-nonlinear response to the abiotic environment. Combining multiple ecological
-processes into a common model could help ecologists to disentangle different
-factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013].
+for dealing with indirect effects, stability, and alternative stable states.
 Numerous other extensions are possible: Markov networks
@@
 Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
-Golding, and N. J. Gotelli provided useful feedback on an earlier draft of
-this work.
+Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
+of this work.
 

### 2015-07-22 Harris_2015/simulation.Rmd
simulation

@@
 ```
+
+$$r_e + r_i n_i (1 - \frac{n_i f_i(\vec{n}) + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.
+
+When no propagules enter from outside the system ($r_e=0$) and mutualisms do
+not occur ($f_i(\vec{n})$ = 1), this reduces to a multi-species Lotka-Volterra
+system
+
+$$r_i n_i (1 - \frac{n_i + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.
+
+where $r_i$ is the [[intrinsic rate of increase]] of species $i$, $n_i$ is its
+abundance, and $\alpha_{ij}$ describes the strength of its competition with
+species $j$.  In the more general form above, the function $f_i(\vec{n})$ 
+acts to reduce the strength of intraspecific competition in the presence of
+mutualists.  For these analyses, I used
+
+$$ $$
+
+[[demographic events happen with probability proportional to corresponding birth/death rate]].  Because these events' probabilities only depend on the 
+current state of the system, they form a Markov chain that explores the 
+space of possible species assemblages under the population dynamic model.  For
+each set of competition and mutualism coefficients, I collected one sample 
+after every `r thin`th demographic event.  For the subsequent analysis, 
+each species was considered "present" in a sample if its abundance was one
+or higher, and "absent" if its abundance in that sample was zero.
 

### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
more

@@
 direct interactions between species pairs even when indirect interactions or
-abiotic environmental effects largely overpowered them. A linear
-approximation, based on partial covariances, also performed well as long as
-the number of sampled locations exceeded the number of species in the data.
-Indirect effects reliably caused a common null modeling approach to produce
-incorrect inferences, however.
+abiotic environmental effects largely overpowered them. Two computationally
+efficient approximations, based on linear and generalized linear models, also performed
+well. Indirect effects reliably caused a common null modeling approach and a Joint
+Species Distribution Modeling approach to produce incorrect inferences, however.
 
@@
 network) than we could get from their overall co-occurrence rates. This kind of
-control is difficult to obtain from null models, which begin with the assumptions
+precision is difficult to obtain from null models, which begin with the assumptions
 that all the pairwise interactions are zero and thus don’t need to be controlled for.
+Nevertheless, null models have dominated this field for more than three decades
+[@diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009].
 
@@
 interactions, and for detecting them from observational data. Markov networks
-have been used in many scientific fields for decades in similar contexts.
-For example, a well-studied network named the Ising model has played an important
-role in our understanding of physics [where nearby particles tend to align
-magnetically with one another; @cipra_introduction_1987].
-In spatial contexts, these models have been used to describe interactions between
-adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
-neurobiology, they have helped researchers identify neural connections from firing patterns
-[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
-[@whittam_species_1981], computational limitations had previously imposed
+have been used in many scientific fields for decades in similar contexts, from physics
+[where nearby particles interact magnetically with one another; @cipra_introduction_1987]
+to spatial statistics [where adjacent grid cells interact with one another; @harris_contact_1974;
+@gelfand_modelling_2005]. While community ecologists explored some related approaches in
+the 1980's [@whittam_species_1981], computational limitations had previously imposed
 severe approximations that produced unintelligible results [e.g.
@@
 ("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
-The most familiar conditional relationship
-is the partial correlation, which indicates the portion of the sample correlation
-between two species that remains after controlling for other variables in the
-data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
-Figure 1 shows how the two correlation measures can have opposite signs,
-and suggests that the partial correlation is more relevant
+The example with the shrubs and trees in
+Figure 1 shows how the two measures can have opposite signs,
+and suggests that conditional relationships are more relevant
 for drawing inferences about species interactions (e.g. competition).
-Markov networks extend this approach to non-Gaussian data, much as
-generalized linear models do for linear regression [@lee_learning_2012].
 
@@
 the probability that the two species will be found together (Figure 2A, Figure 2B).
-$\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
-networks. Because the relative probability of a presence-absence vector
+$\beta$ can be interpreted as an analog of the partial covariance, but for non-Gaussian
+networks [@loh_structure_2013], or as similar to a coefficient in a logistic regression model
+[@lee_learning_2012]. Because the relative probability of a presence-absence vector
 increases when positively-associated species co-occur and decreases when
@@
 Equation 1 into an absolute probability entails scaling by a normalizing constant,
-$Z(\alpha, \beta)$, which is defined so that that the probabilities of all
+$Z$, which is defined so that that the probabilities of all
 possible assemblages that could be produced by the model sum to one (bottom of
-Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
+Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
 quickly becomes infeasible as the number of species increases: with
@@
 [@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
-but they are beyond the scope of this paper.
+but they are beyond the scope of this paper.  Alternatively, as noted below, some
+common linear and generalized linear methods can be used as computationally efficient
+approximations to the full Markov network.
 
@@
 coefficients that would be expected to produce exactly the observed co-occurrence
-frequencies. For the analyses in this paper, I used the *rosalia* package
+frequencies (i.e. it matches the sufficient statistics of the data to the sufficient statistics of the
+model distribution). For the analyses in this paper, I used the *rosalia* package
 [@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
@@
 competition on each partner's death rate.  For these analyses, I simulated
-50 replicate landscapes, each with 25, 200, or 1600 sites and 20 species.
+landscapes with up to 20 species and 25, 200, or 1600 sites (50 replicates per
+landscape size; see Appendix).
 
@@
 above. Given the large number of parameters associated with some of the networks to
-be estimated, I regularized the likelihood using a Cauchy prior distribution
-[@gelman_weakly_2008] with a scale of 1 on the $\alpha$ and $\beta$ terms.
+be estimated, I regularized the likelihood using a logistic prior distribution
+with a scale of 1 on the $\alpha$ and $\beta$ terms.  This convex prior distribution has a
+similar shape to the Laplace distribution used in LASSO regularization (especially in the tails),
+but does not force the estimates to be exactly zero.
 
@@
 `pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
-data, the partial correlation can be thought of as an approximation to the Markov network,
-and is much easier to compute.
+data, the partial correlation can be thought of as a computationally efficient approximation to the
+full Markov network model [@loh_structure_2013].
 
 The third alternative, generalized linear models, can also be thought of as a computationally
-efficient approximation to the Markov network [@REF].  Following @faisal_inferring_2010,
-I fit regularized logistic regression models for each species, using the other species on the
-landscape as predictors. [[Observational presence-absence data does not have enough degrees
-of freedom to estimate this many parameters, so I symmetrized the pairwise correlations by
-averaging the coefficient for predicting species $i$’s status from species $j$ and vice versa.]]
+efficient approximation to the Markov network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized logistic regression models for each species, using
+the other species on the landscape as predictors. [[Observational presence-absence data does
+not have enough degrees of freedom to estimate this many parameters, so I symmetrized the
+pairwise correlations by averaging the coefficient for predicting species $i$’s status from
+species $j$ and vice versa.]]
 
@@
 by each method so that they had a consistent interpretation. After rescaling each
-method’s estimates, I calculated squared errors for each of the [[N]] species pairs
-across all the simulated data sets and used these values to calculate the proportion of variance
-explained for landscapes with 25, 200, or 1600 sampled locations (compared with a baseline
+method’s estimates, I calculated squared errors between the scaled interaction estimates and
+“true” interaction values for each of the [[N]] species pairs
+across all the simulated data sets. These squared errors determined the proportion of variance
+explained for different combinations of method and landscape size (compared with a baseline
 model that assumed all interaction strengths to be zero).
@@
 \noindent  
-***Larger landscapes.***
+***Twenty species.***
 The accuracy of the four evaluated methods varied substantially, depending on
 the parameters that produced the simulated communities (Figure 3). In general,
-however, there was a consistent ordering: overall, the Markov network
-explained 54% of the "true" parameters' squared deviations from zero, followed
-by partial covariances (33%), and sample covariances (22%). The null model
-scores initially explained only 12%. After manually reducing the value of one
-especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
-increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
-results.
+however, there was a consistent ordering…
+
+The three methods that estimated marginal relationships among species and did
+not explicitly control for indirect species interactions (the null model, the joint species
+distribution model, and the sample correlation coefficients) differed from one another
+in their $R^2$ values, but their estimates were surprisingly similar to one another.
+In particular, the correlation coefficients were extremely similar to the estimates
+from the other two methods (Spearman’s rho = X for the joint species distribution
+model and Y for the null model).
 
@@
 approximations can be less reliable in cases where the true interaction matrix
-contains more structure (e.g. guilds or trophic levels).
+contains more structure (e.g. guilds or trophic levels). Similarly, the approximation involved
+in using separate generalized linear models for each species can occasionally lead to catastrophic
+overfitting with small-to-moderate sample sizes [@lee_learning_2012].
 
@@
 range of results shown in Figure 3, scientists in this field could easily
-double their explanatory power by switching from null models to linear
-regression and partial covariances, or triple it by switching to a Markov
-network.
+double their explanatory power by switching from null models to partial correlations or
+generalized linear models, or triple it by switching to a Markov network. Regardless
+of the specific methods ecologists ultimately choose, the most important consideration
+is clearly that we need to control for indirect effects and estimate conditional relationships
+between species, rather than marginal ones.
 

### 2015-07-23 Harris_2015/recovery.R

@@ -36,7 +36,7 @@ f = function(filename){
   }
   coef_matrix = (coef_matrix + t(coef_matrix)) / 2
   
-  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_cauchy_prior(scale = 1))
+  rosie = rosalia(x, maxit = 200, trace = 0)
   
   bc = BC(Y = x, model = "community", its = 1000)
   

### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
BC_inv

@@
 
-***Abstract***
+**Abstract**
 
@@
 \noindent  
-***Introduction***
+**Introduction**
 
@@
 interacting species [@diamond_island_1975; @strong_ecological_1984;
-@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
-will not generally provide much evidence regarding species' direct effects on
-one another.
+@gotelli_empirical_2009; @veech_probabilistic_2013;
+@pollock_understanding_2014] will not generally provide much
+evidence regarding species' direct effects on one another.
 
@@
 In this paper, following recent work by @azaele_inferring_2010 and
-@fort_statistical_2013, I show that Markov networks [undirected graphical models
-also known as Markov random fields; @murphy_machine_2012] can provide a framework
-for understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them from observational data. Markov networks
-have been used in many scientific fields for decades in similar contexts, from physics
-[where nearby particles interact magnetically with one another; @cipra_introduction_1987]
-to spatial statistics [where adjacent grid cells interact with one another; @harris_contact_1974;
-@gelfand_modelling_2005]. While community ecologists explored some related approaches in
-the 1980's [@whittam_species_1981], computational limitations had previously imposed
-severe approximations that produced unintelligible results [e.g.
-"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
-computationally feasible to fit these models exactly, this kind of approach is
-worth a second look.
+@fort_statistical_2013, I show that Markov networks [undirected
+graphical models also known as Markov random fields;
+@murphy_machine_2012] can provide a framework for understanding
+the landscape-level consequences of pairwise species interactions, and
+ for detecting them from observed presence-absence matrices. Markov
+networks have been used in many scientific fields for decades in similar
+contexts, from physics [where nearby particles interact magnetically
+with one another; @cipra_introduction_1987] to spatial statistics [where
+adjacent grid cells interact with one another; @harris_contact_1974;
+@gelfand_modelling_2005]. While community ecologists explored
+some related approaches in the 1980's [@whittam_species_1981],
+computational limitations had previously imposed severe approximations
+that produced unintelligible results [e.g. "probabilities" greater than one;
+@gilpin_factors_1982].  Now that it is computationally feasible to fit these
+models exactly, this kind of approach is worth a second look.
 
-The rest of the paper proceeds as follows. First, I discuss how Markov networks
-operate and how they can be used to simulate landscape-level data or to predict
-the direct and indirect consequences of possible interaction matrices. Then,
-using simulated data sets where the "true" ecological structure is known, I
-compare this approach with several existing methods for detecting species
-interactions. Finally, I discuss opportunities for extending the approach
-presented here to larger problems in community ecology.
+The rest of the paper proceeds as follows. First, I discuss how Markov
+networks operate and how they can be used to simulate landscape-level
+data or to predict the direct and indirect consequences of possible
+interaction matrices. Then, using simulated data sets where the "true"
+ecological structure is known, I compare this approach with several
+existing methods for detecting species interactions. Finally, I discuss
+opportunities for extending the approach presented here to larger
+problems in community ecology.
 
-\noindent  ***Methods***
+\noindent  **Methods**
 
 \noindent  
-***Conditional relationships and Markov networks.***
+**Conditional relationships and Markov networks.**
 Ecologists are often interested in learning about direct interactions between
@@
 
-Markov networks give a probability value for every possible combination of
-presences and absences in communities. For example, given a network with
-binary outcomes (i.e. 0 for absence and 1 for presence), the relative
-probability of observing a given presence-absence vector, $\vec{y}$, is given
-by
+Markov networks define species interactions in conditional terms, and use these
+conditional relationships to determine whether a given combination of co-occurring
+species is consistent with their interactions.  Given a set of binary outcomes
+for each species (i.e. 0 for absence and 1 for presence), the Markov network
+defines the relative probability of observing a given presence-absence vector,
+$\vec{y}$, as
 
@@
 the probability that the two species will be found together (Figure 2A, Figure 2B).
-$\beta$ can be interpreted as an analog of the partial covariance, but for non-Gaussian
-networks [@loh_structure_2013], or as similar to a coefficient in a logistic regression model
-[@lee_learning_2012]. Because the relative probability of a presence-absence vector
+Because the relative probability of a presence-absence vector
 increases when positively-associated species co-occur and decreases when
@@
 either species increase by a factor of $e^2$ when its partner is present
-[@murphy_machine_2012]. Of course, if all else is *not* equal (e.g. Figure 1,
+[@murphy_machine_2012]. In this way, the $\beta$ coefficients behave
+very similarly to the coefficients in a logistic regression model [@lee_learning_2012].
+
+Of course, if all else is *not* equal (e.g. Figure 1,
 where the presence of one competitor is associated with release from another
@@
 Equation 1 into an absolute probability entails scaling by a normalizing constant,
-$Z$, which is defined so that that the probabilities of all
-possible assemblages that could be produced by the model sum to one (bottom of
+$Z$, so that that the probabilities of all possible assemblages will sum to one (bottom of
 Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
@@
 required spirals exponentially into the billions and beyond.
-Numerous techniques are available for working with Markov networks that keep
-the computations tractable, e.g. via analytic approximations
-[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
-but they are beyond the scope of this paper.  Alternatively, as noted below, some
-common linear and generalized linear methods can be used as computationally efficient
-approximations to the full Markov network.
+Numerous techniques are available for keeping the calculations tractable,
+e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
+sampling [@salakhutdinov_learning_2008], but they are beyond
+the scope of this paper.  Alternatively, as noted below, some common linear
+and generalized linear methods can also be used as computationally efficient
+approximations to the full network.
 
 \noindent  
-***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
+**Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.**
 In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
-however, ecologists will often need to estimate these parameters from co-occurrence data.
+however, ecologists will often need to estimate these parameters from co-occurrence data
+(i.e. from a matrix of ones and zeros indicating which species are present at which sites).
 When the number of species is reasonably small, one can compute exact maximum
-likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
-$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
+likelihood estimates for all of the $\alpha$ and $\beta$ coefficients given a
+presence-absence matrix by optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
 networks like the ones considered here have unimodal likelihood surfaces
 [@murphy_machine_2012], ensuring that this procedure will always converge on
-the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
+the global maximum. This maximum represents the unique combination of $\alpha$ and $\beta$
 coefficients that would be expected to produce exactly the observed co-occurrence
-frequencies (i.e. it matches the sufficient statistics of the data to the sufficient statistics of the
-model distribution). For the analyses in this paper, I used the *rosalia* package
+frequencies (i.e. it matches the sufficient statistics of the model distribution to the
+sufficient statistics of the data). For the analyses in this paper, I used the *rosalia* package
 [@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
-define the objective function and gradient as R code.  The rosalia package uses
-standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.
+define the objective function and gradient in R code and then find the optimum.
 
 \noindent  
-***Simulated landscapes.***
+**Simulated landscapes.**
 In order to compare different methods, I simulated two sets of landscapes using known
@@
 interaction coefficients in that figure (Appendix A). Each of the methods
-described below (a Markov network, two correlation-based methods and a null
-model) was then evaluated on its ability to correctly infer that the two
+described below was then evaluated on its ability to correctly infer that the two
 shrub species competed with one another, despite their frequent co-occurrence.
@@
 \noindent  
-***Recovering species interactions from simulated data.***
-I used Markov networks and  several other techniques for determining the sign and strength
-of the associations between pairs of species (Appendix B).
+**Recovering species interactions from simulated data.**
+I used Markov networks and  several other techniques for
+determining the sign and strength of the associations between
+pairs of species (Appendix B).
 
-For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
-above. Given the large number of parameters associated with some of the networks to
-be estimated, I regularized the likelihood using a logistic prior distribution
-with a scale of 1 on the $\alpha$ and $\beta$ terms.  This convex prior distribution has a
-similar shape to the Laplace distribution used in LASSO regularization (especially in the tails),
+For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
+as described above. Given the large number of parameters associated with
+some of the networks to  be estimated, I regularized the likelihood
+using a logistic prior distribution with a scale of 2 on the $\alpha$ and
+$\beta$ terms.  This convex prior distribution has a similar shape to the
+Laplace distribution used in LASSO regularization (especially in the tails),
 but does not force the estimates to be exactly zero.
@@
 partial correlations between each pair of species' data vectors on the landscape
-[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations are undefined
-for landscapes with perfectly-correlated species pairs, I used a regularized
+[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
+are undefined for landscapes with perfectly-correlated species pairs, I used a regularized
 estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
@@
 
-Finally, I used the latent correlation matrix estimated by the BayesComm package [@REF]
-to evaluate the claim from recent papers that the correlation coefficients estimated by “joint
-species distribution models” provide an accurate assessment of species’ pairwise interactions
-[@pollock_understanding_2014; see also @harris_generating_2015].
+The last two methods used the latent correlation matrix estimated by the BayesComm
+package [@REF] to evaluate the claim from recent papers that the correlation
+coefficients estimated by “joint species distribution models” provide an accurate
+assessment of species’ pairwise interactions [@pollock_understanding_2014;
+see also @harris_generating_2015].  I reported both the average correlation
+between each pair of species’ latent variables across 1000 Monte Carlo
+samples, as well as the average partial correlation.
 
 \noindent  
-***Evaluating model performance.***
+**Evaluating model performance.**
 For the simulated landscapes based on Figure 1, method evaluation was fairly
@@
 negative relationship passed; other methods failed. For methods that provide
-confidence intervals or other inferential statistics, I reported them below.
+confidence intervals or other inferential statistics (Appendix), I also reported how
+often they rejected the null hypothesis of no interaction.
 
@@
 \noindent  
-***Results***
+**Results**
 
 \noindent  
-***Three species.***
+**Three species.**
 As shown in Figure 1, the marginal relationship between the two shrub species
@@
 landscapes, and the randomization-based null model falsely reported such
-associations 100% of the time. The two methods for evaluating conditional
+associations 100% of the time. The [[two]] methods for evaluating conditional
 relationships (Markov networks and partial covariances), however, successfully
@@
 \noindent  
-***Twenty species.***
+**Twenty species.**
 The accuracy of the four evaluated methods varied substantially, depending on
@@
 not explicitly control for indirect species interactions (the null model, the joint species
-distribution model, and the sample correlation coefficients) differed from one another
-in their $R^2$ values, but their estimates were surprisingly similar to one another.
-In particular, the correlation coefficients were extremely similar to the estimates
-from the other two methods (Spearman’s rho = X for the joint species distribution
-model and Y for the null model).
+distribution model’s correlation coefficients, and the sample correlation coefficients)
+differed from one another in their $R^2$ values, but their estimates were surprisingly
+similar to one another. In particular, the correlation coefficients were extremely similar
+to the estimates  from the other two methods (Spearman’s rho = X for the joint species distribution model’s correlation matrix and Y for the null model).
 
 \noindent  
-***Discussion***
+**Discussion**
 
@@
 were successful even when direct interactions were largely overwhelmed by indirect
-effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
-these results also indicate that partial covariances and generalized linear models (the two
-methods that estimated conditional relationships rather than marginal ones) can both
-provide useful approximations. The partial correlations' success on simulated
-data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
-approximations can be less reliable in cases where the true interaction matrix
-contains more structure (e.g. guilds or trophic levels). Similarly, the approximation involved
-in using separate generalized linear models for each species can occasionally lead to catastrophic
+effects (Figure 1). For cases where fitting a Markov network is computationally
+infeasible, these results also indicate that partial covariances and generalized linear
+ models (the two methods that estimated conditional relationships rather than
+marginal ones) can both provide useful approximations. The partial correlations'
+success on simulated data may not carry over to real data sets, however;
+@loh_structure_2013 show that the linear approximations can be less reliable
+in cases where the true interaction matrix contains more structure (e.g. guilds
+or trophic levels). Similarly, the approximation involved in using separate
+generalized linear models for each species can occasionally lead to catastrophic
 overfitting with small-to-moderate sample sizes [@lee_learning_2012].
+For these reasons, it will usually be best to fit a Markov network
+rather than one of the alternative methods when one’s computational resources
+allow it.
 
@@
 Despite these limitations, Markov networks have enormous potential to improve
-ecological inferences. For example, Markov networks provide a simple answer
+ecological understanding. For example, Markov networks provide a simple answer
 to the question of how competition should affect a species' overall
@@
 BayesComm (which had the worst performance of any method evaluated here), none of
-the other methods could produce this inference because they do not make predictions
+the other methods could produce this insight because they do not make predictions
 about species composition. [[*back this up with more evidence or remove it*]]
@@
 pairs or specific ecological processes like competition by rejecting a
-general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
-coefficients directly seems like a much more promising approach: to the extent
-that the models' relative performance on real data sets is similar to the
-range of results shown in Figure 3, scientists in this field could easily
+general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s estimates
+were extremely similar to what could be obtained from a simple correlation matrix,
+which raises questions about how much additional value the null model adds.
+Estimating pairwise coefficients directly seems like a much more promising approach:
+to the extent that the models' relative performance on real data sets is similar to the
+range of results shown in Figure 3, scientists in this field could [[easily
 double their explanatory power by switching from null models to partial correlations or
-generalized linear models, or triple it by switching to a Markov network. Regardless
+generalized linear models, or triple it by switching to a Markov network]]. Regardless
 of the specific methods ecologists ultimately choose, the most important consideration
@@
 \noindent  
-***Acknowledgements:***
+**Acknowledgements:**
 This research was funded by a Graduate Research Fellowship from the US
@@
 
-***References:***
+**References:**
 
-***Figure captions***
+**Figure captions**
 
-***Figure 1.***
+**Figure 1.**
 **A.** A small network of three competing species. The tree (top) tends not to
@@
 
-***Figure 2.***
+**Figure 2.**
 **A.** A small Markov network with two species.  The depicted abiotic
@@
 
-***Figure 3.***
+**Figure 3.**
 Proportion of variance in interaction coefficients explained by each method
-with 5, 10, or 20 species arrayed across varying numbers of sampled locations
-when environmental filtering was absent (top row) or present (bottom row). A
-negative $R^2$ values implies that the squared error associated with
-the corresponding subset of the predictions was larger than the error one would
-get from assuming that all coefficients equalled zero.
+across varying numbers of sampled locations.

### 2015-07-23 Harris_2015/recovery.R

@@ -36,9 +36,20 @@ f = function(filename){
   }
   coef_matrix = (coef_matrix + t(coef_matrix)) / 2
   
-  rosie = rosalia(x, maxit = 200, trace = 0)
+  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2))
   
   bc = BC(Y = x, model = "community", its = 1000)
+  
+  bc_inv = 0
+  
+  for(i in 1:nrow(bc$trace$R)){
+    Sigma = matrix(0, nrow = ncol(x), ncol = ncol(x))
+    Sigma[upper.tri(Sigma)] <- bc$trace$R[i, ]  # Fill in upper triangle
+    Sigma <- Sigma + t(Sigma)                   # Fill in lower triangle
+    diag(Sigma) <- 1  # Diagonal equals 1 in multivariate probit model
+    
+    bc_inv = bc_inv + cor2pcor(Sigma) / nrow(bc$trace$R)
+  }
   
   data.frame(
     truth = truth, 
@@ -48,7 +59,8 @@ f = function(filename){
     correlation = corr[upper.tri(corr)],
     `Markov network` = rosie$beta[upper.tri(rosie$beta)],
     GLM = coef_matrix[upper.tri(coef_matrix)],
-    `BayesComm` = colMeans(bc$trace$R)
+    `BayesComm` = colMeans(bc$trace$R),
+    bc_inv = bc_inv[upper.tri(bc_inv)]
   )
 }
 
@@ -69,6 +81,9 @@ ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) +
   scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
   stat_smooth(method = "lm", se = FALSE, size = 1) +
   theme_bw()
+
+
+
 
 resids = sapply(
   colnames(z)[-(1:3)],

### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
TeX

@@
 
-**Abstract**
-
-Estimating species interactions from observational data is one of the most
+**Abstract:** Estimating species interactions from observational data is one of the most
 controversial tasks in community ecology. One difficulty is that a single
@@
 abiotic environmental effects largely overpowered them. Two computationally
-efficient approximations, based on linear and generalized linear models, also performed
-well. Indirect effects reliably caused a common null modeling approach and a Joint
-Species Distribution Modeling approach to produce incorrect inferences, however.
+efficient approximations, based on linear and generalized linear models, also
+performed well. Indirect effects reliably caused a common null modeling approach
+to produce incorrect inferences, however.
 
@@
 \setlength{\parindent}{1cm}
+\setlength{\parskip}{6pt}
 
-\noindent  
-**Introduction**
+### Introduction
 
@@
 
-\noindent  **Methods**
+### Methods
 
-\noindent  
-**Conditional relationships and Markov networks.**
+#### Conditional relationships and Markov networks.
 Ecologists are often interested in learning about direct interactions between
@@
 
-$$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$$
+\centering
+$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
+\raggedright
 
@@
 
-\noindent  
-**Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.**
+#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
 In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
@@
 
-\noindent  
-**Simulated landscapes.**
-In order to compare different methods, I simulated two sets of landscapes using known
-parameters.
+#### Simulated landscapes.
+In order to compare different methods, I simulated two sets of landscapes using
+known parameters.
 
@@
 
-\noindent  
-**Recovering species interactions from simulated data.**
+#### Recovering species interactions from simulated data.
 I used Markov networks and  several other techniques for
@@
 
-I also evaluated five alternative methods from the existing literature to calibrate the
-performance of the Markov network as a method for estimating species interactions.
-The first two alternative interaction measures were the sample correlations and the
-partial correlations between each pair of species' data vectors on the landscape
-[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
-are undefined for landscapes with perfectly-correlated species pairs, I used a regularized
-estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
-`pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
-data, the partial correlation can be thought of as a computationally efficient approximation to the
-full Markov network model [@loh_structure_2013].
+I also evaluated five alternative methods from the existing literature to
+calibrate the performance of the Markov network as a method for estimating
+species interactions. The first two alternative interaction measures were the
+sample correlations and the partial correlations between each pair of species'
+data vectors on the landscape [@albrecht_spatial_2001; @faisal_inferring_2010].
+Because partial correlations are undefined for landscapes with
+perfectly-correlated species pairs, I used a regularized estimate based on
+James-Stein shrinkage, as implemented in the corpcor package’s `pcor.shrink`
+function with the default settings [@REF]. In the context of non-Gaussian
+data, the partial correlation can be thought of as a computationally efficient
+approximation to the full Markov network model [@loh_structure_2013].
 
-The third alternative, generalized linear models, can also be thought of as a computationally
-efficient approximation to the Markov network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized logistic regression models for each species, using
-the other species on the landscape as predictors. [[Observational presence-absence data does
-not have enough degrees of freedom to estimate this many parameters, so I symmetrized the
-pairwise correlations by averaging the coefficient for predicting species $i$’s status from
-species $j$ and vice versa.]]
+The third alternative, generalized linear models (GLMs), can also be
+thought of as a computationally efficient approximation to the Markov
+network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
+regularized logistic regression models for each species, using
+the other species on the landscape as predictors. [[Observational
+presence-absence data does not have enough degrees of freedom to estimate this
+many parameters, so I symmetrized the pairwise correlations by averaging the
+coefficient for predicting species $i$’s status from species $j$ and vice
+versa.]]
 
-The fourth alternative method, described in @gotelli_empirical_2009, involved simulating
-possible landscapes from a null model that retains the row and column sums of
-the original matrix [@strong_ecological_1984]. Using the default options in the
-Pairs software described in @gotelli_empirical_2009, I simulated  the null
+The fourth alternative method, described in @gotelli_empirical_2009, involved
+simulating possible landscapes from a null model that retains the row and column
+sums of the original matrix [@strong_ecological_1984]. Using the default options
+in the Pairs software described in @gotelli_empirical_2009, I simulated the null
 distribution of scaled C-scores (a test statistic describing the number of
@@
 
-The last two methods used the latent correlation matrix estimated by the BayesComm
-package [@REF] to evaluate the claim from recent papers that the correlation
-coefficients estimated by “joint species distribution models” provide an accurate
-assessment of species’ pairwise interactions [@pollock_understanding_2014;
-see also @harris_generating_2015].  I reported both the average correlation
-between each pair of species’ latent variables across 1000 Monte Carlo
-samples, as well as the average partial correlation.
+The last two methods used the latent correlation matrix estimated by the
+BayesComm package [@REF] to evaluate the claim from recent papers that the
+correlation coefficients estimated by “joint species distribution models”
+provide an accurate assessment of species’ pairwise interactions
+[@pollock_understanding_2014; see also @harris_generating_2015].  I reported
+both the average correlation between each pair of species’ latent variables
+across 1000 Monte Carlo samples, as well as the average *partial* correlation.
 
-\noindent  
-**Evaluating model performance.**
+#### Evaluating model performance.
 For the simulated landscapes based on Figure 1, method evaluation was fairly
@@
 negative relationship passed; other methods failed. For methods that provide
-confidence intervals or other inferential statistics (Appendix), I also reported how
-often they rejected the null hypothesis of no interaction.
+confidence intervals or other inferential statistics (Appendix), I also reported
+how often they rejected the null hypothesis of no interaction.
 
@@
 by each method so that they had a consistent interpretation. After rescaling each
-method’s estimates, I calculated squared errors between the scaled interaction estimates and
-“true” interaction values for each of the [[N]] species pairs
-across all the simulated data sets. These squared errors determined the proportion of variance
-explained for different combinations of method and landscape size (compared with a baseline
-model that assumed all interaction strengths to be zero).
+method’s estimates, I calculated squared errors between the scaled interaction
+estimates and “true” interaction values for each of the [[N]] species pairs
+across all the simulated data sets. These squared errors determined the
+proportion of variance explained for different combinations of method and
+landscape size (compared with a baseline model that assumed all interaction
+strengths to be zero).
 
-\noindent  
-**Results**
+### Results
 
-\noindent  
-**Three species.**
+#### Three species.
 As shown in Figure 1, the marginal relationship between the two shrub species
@@
 landscapes, and the randomization-based null model falsely reported such
-associations 100% of the time. The [[two]] methods for evaluating conditional
-relationships (Markov networks and partial covariances), however, successfully
-controlled for the indirect pathway via the tree species and each correctly
-identified the direct negative interaction between the shrubs 94% of the time.
+associations 100% of the time. The methods for evaluating conditional
+relationships (Markov networks, GLMs, and both types of partial covariance
+estimates), however, successfully controlled for the indirect pathway via
+the tree species and each correctly identified the direct negative interaction
+between the shrubs 94% of the time.
 
-\noindent  
-**Twenty species.**
+#### Twenty species.
 The accuracy of the four evaluated methods varied substantially, depending on
@@
 The three methods that estimated marginal relationships among species and did
-not explicitly control for indirect species interactions (the null model, the joint species
-distribution model’s correlation coefficients, and the sample correlation coefficients)
-differed from one another in their $R^2$ values, but their estimates were surprisingly
-similar to one another. In particular, the correlation coefficients were extremely similar
-to the estimates  from the other two methods (Spearman’s rho = X for the joint species distribution model’s correlation matrix and Y for the null model).
+not explicitly control for indirect species interactions (the null model, the
+joint species distribution model’s correlation coefficients, and the sample
+correlation coefficients) differed from one another in their $R^2$ values,
+but their estimates were surprisingly similar to one another. In particular, the
+correlation coefficients were extremely similar to the estimates  from the other
+two methods (Spearman’s rho = X for the joint species distribution model’s
+correlation matrix and Y for the null model).
 
-\noindent  
-**Discussion**
+### Discussion
 
-\noindent  
 The results presented above show that Markov networks can reliably recover
@@
 
-It’s important to note that none of these methods can identify the exact nature of the
-pairwise interactions (e.g. which species in a positively-associated pair is facilitating the other),
-particularly when real pairs of species can reciprocally influence one another in multiple ways
-simultaneously [@bruno_inclusion_2003]; with compositional data, there is only enough
-information to provide a single number describing each species pair.  On the other hand, if
-ecologists are familiar enough with the natural history of their study
-systems that they can augment the likelihood function with a prior distribution or a suitable
-penalty, then this information could reduce the effective degrees of freedom to estimate and
-real-world results might be even better than those shown in Figure 3.
+It’s important to note that none of these methods can identify the exact nature
+of the pairwise interactions (e.g. which species in a positively-associated
+  pair is facilitating the other), particularly when real pairs of species can
+reciprocally influence one another in multiple ways simultaneously
+[@bruno_inclusion_2003]; with compositional data, there is only enough
+information to provide a single number describing each species pair.  
+On the other hand, if ecologists are familiar enough with the natural history
+of their study systems that they can augment the likelihood function with a
+prior distribution or a suitable penalty, then this information could reduce the
+effective degrees of freedom to estimate and real-world results might be even
+better than those shown in Figure 3.
 
@@
 Competition's effect on prevalence in a Markov network can then be calculated
-by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
-BayesComm (which had the worst performance of any method evaluated here), none of
-the other methods could produce this insight because they do not make predictions
-about species composition. [[*back this up with more evidence or remove it*]]
+by subtracting this value and the observed prevalence (cf Figure 2D).
 
@@
 
-\noindent  
-**Acknowledgements:**
+#### Acknowledgements:
 This research was funded by a Graduate Research Fellowship from the US
@@
 \setlength{\parindent}{0cm}
+\setlength{\parskip}{4pt}
 

### 2015-07-23 Harris_2015/manuscript-materials/draft.tex
TeX

@@
 \documentclass[12pt,]{article}
+\usepackage[compact]{titlesec}
 \usepackage{lmodern}
@@
 \setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt}
+\setlength{\parskip}{4pt}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 
-\textbf{\emph{Abstract}}
-
-Inferring species interactions from observational data is one of the
-most controversial tasks in community ecology. One difficulty is that a
-single pairwise interaction can ripple through an ecological network and
-produce surprising indirect consequences. For example, two competing
-species would ordinarily correlate negatively in space, but this effect
-can be reversed in the presence of a third species that is capable of
-outcompeting both of them when it is present. Here, I apply models from
-statistical physics, called Markov networks or Markov random fields,
-that can predict the direct and indirect consequences of any possible
-species interaction matrix. Interactions in these models can be
-estimated from observational data via maximum likelihood. Using
-simulated landscapes with known pairwise interaction strengths, I
-evaluated Markov networks and several existing approaches. The Markov
-networks consistently outperformed other methods, correctly isolating
-direct interactions between species pairs even when indirect
+\textbf{Abstract:} Estimating species interactions from observational
+data is one of the most controversial tasks in community ecology. One
+difficulty is that a single pairwise interaction can ripple through an
+ecological network and produce surprising indirect consequences. For
+example, two competing species would ordinarily correlate negatively in
+space, but this effect can be reversed in the presence of a third
+species that is capable of outcompeting both of them when it is present.
+Here, I apply models from statistical physics, called Markov networks or
+Markov random fields, that can predict the direct and indirect
+consequences of any possible species interaction matrix. Interactions in
+these models can be estimated from observational data via maximum
+likelihood. Using simulated landscapes with known pairwise interaction
+strengths, I evaluated Markov networks and several existing approaches.
+The Markov networks consistently outperformed other methods, correctly
+isolating direct interactions between species pairs even when indirect
 interactions or abiotic environmental effects largely overpowered them.
-A linear approximation, based on partial covariances, also performed
-well as long as the number of sampled locations exceeded the number of
-species in the data. Indirect effects reliably caused a common null
-modeling approach to produce incorrect inferences, however.
+Two computationally efficient approximations, based on linear and
+generalized linear models, also performed well. Indirect effects
+reliably caused a common null modeling approach to produce incorrect
+inferences, however.
 
@@
 
-\textbf{\emph{Introduction}}
+\setlength{\parindent}{1cm} \setlength{\parskip}{6pt}
+
+\subsubsection{Introduction}\label{introduction}
 
@@
 existing methods for detecting competition's effects on community
-structure are unreliable (Gotelli and Ulrich 2009), and thus important
-ecological processes remain poorly understood. More generally, it can be
-difficult to reason about the complex web of direct and indirect species
-interactions in real assemblages, especially when these interactions
-occur against a background of other ecological processes such as
-dispersal limitation and environmental filtering (Connor et al. 2013).
-For this reason, it isn't always clear what kinds of patterns would even
-constitute evidence of competition, as opposed to some other biological
-process or random sampling error (Lewin 1983, Roughgarden 1983).
+structure are unreliable (Gotelli and Ulrich 2009). In particular,
+species' effects on one another can become lost in the complex web of
+direct and indirect interactions in real assemblages. For example, the
+competitive interaction between the two shrub species in Figure 1A can
+become obscured by their shared tendency to occur in unshaded areas
+(Figure 1B). If this sort of effect is common, then the vast majority of
+ecologists' methods, which rely on test statistics describing the total
+number of co-occurrences or overall correlation between putatively
+interacting species (Diamond 1975, Strong et al. 1984, Gotelli and
+Ulrich 2009, Veech 2013, Pollock et al. 2014) will not generally provide
+much evidence regarding species' direct effects on one another.
 
-Most existing methods in this field compare the frequency with which two
-putative competitors are observed to co-occur, versus the frequency that
-would be expected if \emph{all} species on the landscape were
-independent (Strong et al. 1984, Gotelli and Ulrich 2009). Examining a
-species pair against such a ``null'' background, however, rules out the
-possibility that the overall association between two species could be
-driven by an outside force. For example, even though the two shrub
-species in Figure 1 compete with one another for resources at a
-mechanistic level, they end up clustering together on the landscape
-because they both grow best in areas that are not overshadowed by trees.
-If this sort of effect is common, then significant deviations from
-independence will not generally provide convincing evidence of species'
-direct effects on one another.
+While competition doesn't reliably leave the expected pattern in
+community structure (a low co-occurrence rate at the landscape level),
+it nevertheless does leave a signal in the data (Figure 1C).
+Specifically, \emph{among shaded sites}, there will be a deficit of
+co-occurrences between the competing shrub species, and \emph{among
+unshaded sites}, there will also be such a deficit. More generally, we
+can obtain much better estimates of the association between two species
+from their conditional relationships (i.e.~by controlling for other
+species in the network) than we could get from their overall
+co-occurrence rates. This kind of precision is difficult to obtain from
+null models, which begin with the assumptions that all the pairwise
+interactions are zero and thus don't need to be controlled for.
+Nevertheless, null models have dominated this field for more than three
+decades (Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009).
 
-While competition between the two shrubs in the previous example does
-not leave the commonly-expected pattern in community structure (negative
-association at the landscape level), it nevertheless does leave a signal
-in the data (Figure 1C). Specifically, \emph{among shaded sites}, there
-will be a deficit of co-occurrences, and \emph{among unshaded sites},
-there will also be such a deficit.
-
-In this paper, I introduce Markov networks (undirected graphical models
-also known as Markov random fields; Murphy 2012) as a framework for
+In this paper, following recent work by Azaele et al. (2010) and Fort
+(2013), I show that Markov networks (undirected graphical models also
+known as Markov random fields; Murphy 2012) can provide a framework for
 understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them with observational data. Markov
-networks, which generalize partial correlations to non-Gaussian data
-(Lee and Hastie 2012, Loh and Wainwright 2013), have been used in many
-scientific fields for decades to model associations between various
-kinds of ``particles''. For example, a well-studied network named the
-Ising model has played an important role in our understanding of physics
-(where nearby particles tend to align magnetically with one another;
-Cipra 1987). In spatial contexts, these models have been used to
-describe interactions between adjacent grid cells (Harris 1974, Gelfand
-et al. 2005). In neurobiology, they have helped researchers determine
-which neurons are connected to one another by modeling the structure in
-their firing patterns (Schneidman et al. 2006). Following recent work by
-Azaele et al. (2010) and Fort (2013), I suggest that ecologists could
-similarly treat species as the interacting particles in the same
-modeling framework. Doing so would allow ecologists to simulate and
-study the landscape-level consequences of arbitrary species interaction
-matrices, even when our observations are not Gaussian. While ecologists
-explored some related approaches in the 1980's (Whittam and
-Siegel-Causey 1981), computational limitations had previously imposed
-severe approximations that produced unintelligible results (e.g.
-``probabilities'' greater than one; Gilpin and Diamond 1982). Now that
-it is computationally feasible to fit these models exactly, the approach
-has become worth a second look.
+interactions, and for detecting them from observed presence-absence
+matrices. Markov networks have been used in many scientific fields for
+decades in similar contexts, from physics (where nearby particles
+interact magnetically with one another; Cipra 1987) to spatial
+statistics (where adjacent grid cells interact with one another; Harris
+1974, Gelfand et al. 2005). While community ecologists explored some
+related approaches in the 1980's (Whittam and Siegel-Causey 1981),
+computational limitations had previously imposed severe approximations
+that produced unintelligible results (e.g. ``probabilities'' greater
+than one; Gilpin and Diamond 1982). Now that it is computationally
+feasible to fit these models exactly, this kind of approach is worth a
+second look.
 
@@
 
-\textbf{\emph{Methods}}
+\subsubsection{Methods}\label{methods}
 
-\textbf{\emph{Conditional relationships and Markov networks.}}
-Ecologists are often interested in inferring direct interactions between
-species, controlling for the indirect influence of other species. In
-statistical terms, this implies that ecologists want to estimate
-\emph{conditional} (``all-else-equal'') relationships, rather than
-\emph{marginal} (``overall'') relationships. The most familiar
-conditional relationship is the partial correlation, which indicates the
-portion of the sample correlation between two species that remains after
-controlling for other variables in the data set (Albrecht and Gotelli
-2001). The example with the shrubs and trees in Figure 1 shows how the
-two correlation measures can have opposite signs, and suggests that the
-partial correlation is more relevant for drawing inferences about
-species interactions (e.g.~competition). Markov networks extend this
-approach to non-Gaussian data, much as generalized linear models do for
-linear regression (Lee and Hastie 2012).
+\paragraph{Conditional relationships and Markov
+networks.}\label{conditional-relationships-and-markov-networks.}
 
-Markov networks give a probability value for every possible combination
-of presences and absences in communities. For example, given a network
-with binary outcomes (i.e.~0 for absence and 1 for presence), the
-relative probability of observing a given presence-absence vector,
-\(\vec{y}\), is given by
+Ecologists are often interested in learning about direct interactions
+between species, controlling for the indirect influence of other
+species. In statistical terms, this implies that ecologists want to
+estimate \emph{conditional} (``all-else-equal'') relationships, rather
+than \emph{marginal} (``overall'') relationships. The example with the
+shrubs and trees in Figure 1 shows how the two measures can have
+opposite signs, and suggests that conditional relationships are more
+relevant for drawing inferences about species interactions
+(e.g.~competition).
 
-\[p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\]
+Markov networks define species interactions in conditional terms, and
+use these conditional relationships to determine whether a given
+combination of co-occurring species is consistent with their
+interactions. Given a set of binary outcomes for each species (i.e.~0
+for absence and 1 for presence), the Markov network defines the relative
+probability of observing a given presence-absence vector, \(\vec{y}\),
+as
+
+\centering
+\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)
+\raggedright
 
@@
 controls the probability that the two species will be found together
-(Figure 2A, Figure 2B). \(\beta\) thus acts as an analog of the partial
-covariance, but for non-Gaussian networks. Because the relative
-probability of a presence-absence vector increases when
-positively-associated species co-occur and decreases when
-negatively-associated species co-occur, the model tends to produce
-assemblages that have many pairs of positively-associated species and
-relatively few pairs of negatively-associated species (exactly as an
-ecologist might expect).
+(Figure 2A, Figure 2B). Because the relative probability of a
+presence-absence vector increases when positively-associated species
+co-occur and decreases when negatively-associated species co-occur, the
+model tends to produce assemblages that have many pairs of
+positively-associated species and relatively few pairs of
+negatively-associated species (exactly as an ecologist might expect).
 
@@
 odds of observing either species increase by a factor of \(e^2\) when
-its partner is present (Murphy 2012). Of course, if all else is
-\emph{not} equal (e.g.~Figure 1, where the presence of one competitor is
-associated with release from another competitor), then species' marginal
-association rates can differ from this expectation. For this reason, it
-is important to consider how coefficients' effects propagate through the
-network, as discussed below.
+its partner is present (Murphy 2012). In this way, the \(\beta\)
+coefficients behave very similarly to the coefficients in a logistic
+regression model (Lee and Hastie 2012).
+
+Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
+presence of one competitor is associated with release from another
+competitor), then species' marginal association rates can differ from
+this expectation. For this reason, it is important to consider how
+coefficients' effects propagate through the network, as discussed below.
 
@@
 probability given by Equation 1 into an absolute probability entails
-scaling by a \emph{partition function}, \(Z(\alpha, \beta)\), which is
-defined so that that the probabilities of all possible assemblages that
-could be produced by the model sum to one (bottom of Figure 2B).
-Calculating \(Z(\alpha, \beta)\) exactly, as is done in this paper,
-quickly becomes infeasible as the number of species increases: with
-\(2^N\) possible assemblages of \(N\) species, the number of bookkeeping
-operations required for exact inference spirals exponentially into the
-billions and beyond. Numerous techniques are available for working with
-Markov networks that keep the computations tractable, e.g.~via analytic
-approximations (Lee and Hastie 2012) or Monte Carlo sampling
-(Salakhutdinov 2008), but they are beyond the scope of this paper.
+scaling by a normalizing constant, \(Z\), so that that the probabilities
+of all possible assemblages will sum to one (bottom of Figure 2B).
+Calculating \(Z\) exactly, as is done in these analyses, quickly becomes
+infeasible as the number of species increases: with \(2^N\) possible
+assemblages of \(N\) species, the number of bookkeeping operations
+required spirals exponentially into the billions and beyond. Numerous
+techniques are available for keeping the calculations tractable,
+e.g.~via analytic approximations (Lee and Hastie 2012) or Monte Carlo
+sampling (Salakhutdinov 2008), but they are beyond the scope of this
+paper. Alternatively, as noted below, some common linear and generalized
+linear methods can also be used as computationally efficient
+approximations to the full network.
 
-\textbf{\emph{Simulations.}} In order to compare different methods for
-drawing inferences from observational data, I simulated two sets of
+\paragraph{\texorpdfstring{Estimating \(\alpha\) and \(\beta\)
+coefficients from presence-absence
+data.}{Estimating \textbackslash{}alpha and \textbackslash{}beta coefficients from presence-absence data.}}\label{estimating-alpha-and-beta-coefficients-from-presence-absence-data.}
+
+In the previous two sections, the values of \(\alpha\) and \(\beta\)
+were known. In practice, however, ecologists will often need to estimate
+these parameters from co-occurrence data (i.e.~from a matrix of ones and
+zeros indicating which species are present at which sites). When the
+number of species is reasonably small, one can compute exact maximum
+likelihood estimates for all of the \(\alpha\) and \(\beta\)
+coefficients given a presence-absence matrix by optimizing
+\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
+ones considered here have unimodal likelihood surfaces (Murphy 2012),
+ensuring that this procedure will always converge on the global maximum.
+This maximum represents the unique combination of \(\alpha\) and
+\(\beta\) coefficients that would be expected to produce exactly the
+observed co-occurrence frequencies (i.e.~it matches the sufficient
+statistics of the model distribution to the sufficient statistics of the
+data). For the analyses in this paper, I used the \emph{rosalia} package
+(Harris 2015a) for the R programming language (R Core Team 2015) to
+define the objective function and gradient in R code and then find the
+optimum.
+
+\paragraph{Simulated landscapes.}\label{simulated-landscapes.}
+
+In order to compare different methods, I simulated two sets of
 landscapes using known parameters.
@@
 distribution defined by the interaction coefficients in that figure
-(Appendix A). Each of the methods described below (a Markov network, two
-correlation-based methods and a null model) was then evaluated on its
-ability to correctly infer that the two shrub species competed with one
-another, despite their frequent co-occurrence.
+(Appendix A). Each of the methods described below was then evaluated on
+its ability to correctly infer that the two shrub species competed with
+one another, despite their frequent co-occurrence.
 
-I also simulated a second set of landscapes with five, ten, or twenty
-potentially-interacting species on landscapes composed of 20, 100, 500,
-or 2500 observed communities (24 replicate simulations for each
-combination; Appendix B). These simulated data sets span the range from
-small, single-observer data sets to large collaborative efforts such as
-the North American Breeding Bird Survey. As described in Appendix B, I
-randomly drew the ``true'' coefficient values for each replicate so that
-most species pairs interacted negligibly, a few pairs interacted very
-strongly, and competition was three times more common than facilitation.
-I then used Gibbs sampling to randomly generate replicate landscapes
-with varying numbers of species and sites (Appendix B). For half of the
-simulated landscapes, I treated each species' \(\alpha\) coefficient as
-a constant, as described above. For the other half, I treated the
-\(\alpha\) coefficients as linear functions of two abiotic environmental
-factors that varied from location to location across the landscape
-(Appendix B). The latter set of simulated landscapes provide an
-important test of the methods' ability to distinguish co-occurrence
-patterns that were generated from pairwise biotic interactions from
-those that were generated by external forces like abiotic environmental
-filtering. This task was made especially difficult because---as with
-most analyses of presence-absence data for co-occurrence patterns---the
-inference procedure did not have access to any information about the
-environmental or spatial variables that helped shape the landscape (cf
-Connor et al. 2013, Blois et al. 2014).
+I also simulated a second set of landscapes using a stochastic community
+model based on generalized Lotka-Volterra dynamics, as described in the
+Appendix. In these simulations, each species pair was randomly assigned
+to either compete for a portion of the available carrying capacity
+(negative interaction) or to act as mutualists (positive interaction).
+In these simulations, mutualism operated by mitigating the effects of
+intraspecific competition on each partner's death rate. For these
+analyses, I simulated landscapes with up to 20 species and 25, 200, or
+1600 sites (50 replicates per landscape size; see Appendix).
 
-\textbf{\emph{Inferring \(\alpha\) and \(\beta\) coefficients from
-presence-absence data.}} In the previous two sections, the values of
-\(\alpha\) and \(\beta\) were known. In practice, however, ecologists
-will often need to estimate these parameters from co-occurrence data.
-When the number of species is reasonably small, one can compute exact
-maximum likelihood estimates for all of the \(\alpha\) and \(\beta\)
-coefficients by optimizing \(p(\vec{y}; \alpha, \beta)\). Fully-observed
-Markov networks like the ones considered here have unimodal likelihood
-surfaces (Murphy 2012), ensuring that this procedure will always
-converge on the global maximum. This maximum is the unique combination
-of \(\alpha\) and \(\beta\) coefficients that would be expected to
-produce exactly the observed co-occurrence frequencies. For the analyses
-in this paper, I used the \emph{rosalia} package (Harris 2015a) for the
-R programming language (R Core Team 2015) to define the objective
-function and gradient as R code. The rosalia package then uses the
-\texttt{BFGS} method in R's \texttt{optim} function to find the best
-values for \(\alpha\) and \(\beta\).
+\paragraph{Recovering species interactions from simulated
+data.}\label{recovering-species-interactions-from-simulated-data.}
 
-For analyses with 5 or more species, I made a small modification to the
-maximum likelihood procedure described above. Given the large number of
-parameters associated with some of the networks to be estimated, I
-regularized the likelihood using a logistic prior distribution (Gelman
-et al. 2008) with a scale of 1 on the \(\alpha\) and \(\beta\) terms.
+I used Markov networks and several other techniques for determining the
+sign and strength of the associations between pairs of species (Appendix
+B).
 
-\textbf{\emph{Other inference techniques for comparison.}} After fitting
-Markov networks to the simulated landscapes described above, I used
-several other techniques for inferring the sign and strength of marginal
-associations between pairs of species (Appendix B).
+For the Markov networks, I used the rosalia package (Harris 2015a), as
+described above. Given the large number of parameters associated with
+some of the networks to be estimated, I regularized the likelihood using
+a logistic prior distribution with a scale of 2 on the \(\alpha\) and
+\(\beta\) terms. This convex prior distribution has a similar shape to
+the Laplace distribution used in LASSO regularization (especially in the
+tails), but does not force the estimates to be exactly zero.
 
-The first two alternative interaction measures were the sample
-covariances and the partial covariances between each pair of species'
-data vectors on the landscape (Albrecht and Gotelli 2001). Because
-partial covariances are undefined for landscapes with
-perfectly-correlated species pairs, I used a regularized estimate based
-on ridge regression {[}Wieringen and Peeters (2014); i.e.~linear
-regression with a Gaussian prior{]}. For these analyses, I set the ridge
-parameter to 0.2 divided by the number of sites on the landscape.
+I also evaluated five alternative methods from the existing literature
+to calibrate the performance of the Markov network as a method for
+estimating species interactions. The first two alternative interaction
+measures were the sample correlations and the partial correlations
+between each pair of species' data vectors on the landscape (Albrecht
+and Gotelli 2001, Faisal et al. 2010). Because partial correlations are
+undefined for landscapes with perfectly-correlated species pairs, I used
+a regularized estimate based on James-Stein shrinkage, as implemented in
+the corpcor package's \texttt{pcor.shrink} function with the default
+settings ({\textbf{???}}). In the context of non-Gaussian data, the
+partial correlation can be thought of as a computationally efficient
+approximation to the full Markov network model (Loh and Wainwright
+2013).
 
-The third alternative method, described in Gotelli and Ulrich (2009),
+The third alternative, generalized linear models (GLMs), can also be
+thought of as a computationally efficient approximation to the Markov
+network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
+regularized logistic regression models for each species, using the other
+species on the landscape as predictors. {[}{[}Observational
+presence-absence data does not have enough degrees of freedom to
+estimate this many parameters, so I symmetrized the pairwise
+correlations by averaging the coefficient for predicting species \(i\)'s
+status from species \(j\) and vice versa.{]}{]}
+
+The fourth alternative method, described in Gotelli and Ulrich (2009),
 involved simulating possible landscapes from a null model that retains
@@
 
-\textbf{\emph{Method evaluation.}} For the simulated landscapes based on
-Figure 1, method evaluation was fairly qualitative: any method whose
-test statistic for the two shrubs indicated a negative relationship
-passed; other methods failed.
+The last two methods used the latent correlation matrix estimated by the
+BayesComm package ({\textbf{???}}) to evaluate the claim from recent
+papers that the correlation coefficients estimated by ``joint species
+distribution models'' provide an accurate assessment of species'
+pairwise interactions (Pollock et al. 2014, see also Harris 2015b). I
+reported both the average correlation between each pair of species'
+latent variables across 1000 Monte Carlo samples, as well as the average
+\emph{partial} correlation.
 
-For the larger landscapes, I rescaled the four methods' estimates using
-linear regression through the origin so that they all had a consistent
-interpretation. In each method, I regressed the ``true'' \(\beta\)
-coefficient for each species pair against the model's estimate,
-re-weighting the pairs so that each landscape contributed equally to the
-rescaled estimate. These regressions yielded squared errors for each of
-the 23,140 simulated species pairs, across all conditions and
-replicates. I then partitioned these errors into groups defined by the
-properties of the simulated landscapes (i.e.~species richness, number of
-observed communities, and the presence/absence of environmental
-filtering). Within each partition, the mean squared error was used to
-calculate the proportion of variance explained (compared with a baseline
-model that assumed all interaction strengths to be zero).
+\paragraph{Evaluating model
+performance.}\label{evaluating-model-performance.}
 
-\textbf{\emph{Results}}
+For the simulated landscapes based on Figure 1, method evaluation was
+fairly qualitative: any method whose test statistic for the two shrubs
+indicated a negative relationship passed; other methods failed. For
+methods that provide confidence intervals or other inferential
+statistics (Appendix), I also reported how often they rejected the null
+hypothesis of no interaction.
 
-\textbf{\emph{Three species.}} As shown in Figure 1, the marginal
-relationship between the two shrub species was positive---despite their
-competition for space at a mechanistic level---due to indirect effects
-of the dominant tree species. As a result, the covariance method falsely
-reported positive associations for 94\% of the simulated landscapes, and
-the randomization-based null model falsely reported such associations
-100\% of the time. The two methods for evaluating conditional
-relationships (Markov networks and partial covariances), however,
-successfully controlled for the indirect pathway via the tree species
-and each correctly identified the direct negative interaction between
-the shrubs 94\% of the time.
+Because the different methods mostly describe species interactions on
+different scales (e.g.~correlations versus \(Z\) scores versus
+regression coefficients), I used linear regression through the origin to
+rescale the different estimates produced by each method so that they had
+a consistent interpretation. After rescaling each method's estimates, I
+calculated squared errors between the scaled interaction estimates and
+``true'' interaction values for each of the {[}{[}N{]}{]} species pairs
+across all the simulated data sets. These squared errors determined the
+proportion of variance explained for different combinations of method
+and landscape size (compared with a baseline model that assumed all
+interaction strengths to be zero).
 
-\textbf{\emph{Larger landscapes.}} The accuracy of the four evaluated
-methods varied substantially, depending on the parameters that produced
-the simulated communities (Figure 3). In general, however, there was a
-consistent ordering: overall, the Markov network explained 54\% of the
-``true'' parameters' squared deviations from zero, followed by partial
-covariances (33\%), and sample covariances (22\%). The null model scores
-initially explained only 12\%. After manually reducing the value of one
-especially strong outlier (\(Z=1004\), implying \(p<10^{-1000000}\)),
-this increased to 17\% (Appendix B). Figure 3 reflects the adjusted
-version of the results.
+\subsubsection{Results}\label{results}
 
-\textbf{\emph{Discussion}}
+\paragraph{Three species.}\label{three-species.}
+
+As shown in Figure 1, the marginal relationship between the two shrub
+species was positive---despite their competition for space at a
+mechanistic level---due to indirect effects of the dominant tree
+species. As a result, the covariance method falsely reported positive
+associations for 94\% of the simulated landscapes, and the
+randomization-based null model falsely reported such associations 100\%
+of the time. The methods for evaluating conditional relationships
+(Markov networks, GLMs, and both types of partial covariance estimates),
+however, successfully controlled for the indirect pathway via the tree
+species and each correctly identified the direct negative interaction
+between the shrubs 94\% of the time.
+
+\paragraph{Twenty species.}\label{twenty-species.}
+
+The accuracy of the four evaluated methods varied substantially,
+depending on the parameters that produced the simulated communities
+(Figure 3). In general, however, there was a consistent ordering\ldots{}
+
+The three methods that estimated marginal relationships among species
+and did not explicitly control for indirect species interactions (the
+null model, the joint species distribution model's correlation
+coefficients, and the sample correlation coefficients) differed from one
+another in their \(R^2\) values, but their estimates were surprisingly
+similar to one another. In particular, the correlation coefficients were
+extremely similar to the estimates from the other two methods
+(Spearman's rho = X for the joint species distribution model's
+correlation matrix and Y for the null model).
+
+\subsubsection{Discussion}\label{discussion}
 
@@
 Specifically, Markov networks were successful even when direct
-interactions were largely overwhelmed by indirect effects (Figure 1) or
-environmental effects (lower panels of Figure 3). For cases where
-fitting a Markov network is computationally infeasible, these results
-also indicate that partial covariances---which can be computed
-straightforwardly by linear regression---can often provide a
-surprisingly useful approximation. The partial correlations' success on
-simulated data may not carry over to real data sets, however; Loh and
-Wainwright (2013) show that the linear approximations can be less
-reliable in cases where the true interaction matrix contains more
-structure (e.g.~guilds or trophic levels). On the other hand, if
-ecologists are familiar enough with the natural history of their study
-systems to describe this kind of structure as a prior distribution on
-the parameters or as a penalty on the likelihood, then this information
-could reduce the effective degres of freedom to estimate and real-world
-results might be even better than those shown in Figure 3.
+interactions were largely overwhelmed by indirect effects (Figure 1).
+For cases where fitting a Markov network is computationally infeasible,
+these results also indicate that partial covariances and generalized
+linear models (the two methods that estimated conditional relationships
+rather than marginal ones) can both provide useful approximations. The
+partial correlations' success on simulated data may not carry over to
+real data sets, however; Loh and Wainwright (2013) show that the linear
+approximations can be less reliable in cases where the true interaction
+matrix contains more structure (e.g.~guilds or trophic levels).
+Similarly, the approximation involved in using separate generalized
+linear models for each species can occasionally lead to catastrophic
+overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
+For these reasons, it will usually be best to fit a Markov network
+rather than one of the alternative methods when one's computational
+resources allow it.
 
-Ecologists will also need natural history to pin down the exact nature
-of the interactions identified by a network model (e.g.~which species in
-a positively-associated pair is facilitating the other), particularly
-when real pairs of species can reciprocally influence one another in
-multiple ways simultaneously (Bruno et al. 2003); the \(\beta\)
-coefficients in Markov networks have to reduce this complexity to a
-single number. In short, partial correlations and Markov networks both
-help prevent us from mistaking marginal associations for conditional
-ones, but they can't tell us the underlying biological mechanisms at
-work.
+It's important to note that none of these methods can identify the exact
+nature of the pairwise interactions (e.g.~which species in a
+positively-associated pair is facilitating the other), particularly when
+real pairs of species can reciprocally influence one another in multiple
+ways simultaneously (Bruno et al. 2003); with compositional data, there
+is only enough information to provide a single number describing each
+species pair.\\On the other hand, if ecologists are familiar enough with
+the natural history of their study systems that they can augment the
+likelihood function with a prior distribution or a suitable penalty,
+then this information could reduce the effective degrees of freedom to
+estimate and real-world results might be even better than those shown in
+Figure 3.
 
 Despite these limitations, Markov networks have enormous potential to
-improve ecological inferences. For example, Markov networks provide a
+improve ecological understanding. For example, Markov networks provide a
 simple answer to the question of how competition should affect a
-species' overall prevalence, which was a major flash point for the null
+species' overall prevalence, which was a major flashpoint for the null
 model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
@@
 (Cipra 1987), and the models' properties, capabilities, and limits are
-well-understood in a huge range of applications, from spatial statistics
-(Gelfand et al. 2005) to neuroscience (Schneidman et al. 2006) to models
-of human behavior (Lee et al. 2013). Modeling species interactions using
-the same framework would thus allow ecologists to tap into an enormous
-set of existing discoveries and techniques for dealing with indirect
-effects, stability, and alternative stable states (i.e.~phase
-transitions; Cipra (1987)).
-
-This modeling approach is also highly extensible, even when it is
-inconvenient to compute the likelihood exactly. For example, the mistnet
-software package for joint species distribution modeling (Harris 2015b)
-can fit \emph{approximate} Markov networks to large species assemblages
-(\textgreater{}300 species) while simultaneously modeling each species'
-nonlinear response to the abiotic environment. Combining multiple
-ecological processes into a common model could help ecologists to
-disentangle different factors that can confound simpler co-occurrence
-analyses (cf Connor et al. 2013). Numerous other extensions are
-possible: Markov networks can be fit with a mix of discrete and
-continuous variables, for example (Lee and Hastie 2012). There are even
-methods (Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that
-would allow the coefficient linking two species in an interaction matrix
-to vary as a function of the abiotic environment or of third-party
-species that tip the balance between facilitation and exploitation
-(Bruno et al. 2003).
+well-understood in a huge range of applications. Modeling species
+interactions using the same framework would thus allow ecologists to tap
+into an enormous set of existing discoveries and techniques for dealing
+with indirect effects, stability, and alternative stable states.
+Numerous other extensions are possible: Markov networks can be fit with
+a mix of discrete and continuous variables, for example (Lee and Hastie
+2012). There are even methods (Whittam and Siegel-Causey 1981,
+Tjelmeland and Besag 1998) that would allow the coefficient linking two
+species in an interaction matrix to vary as a function of the abiotic
+environment or of third-party species that tip the balance between
+facilitation and exploitation (Bruno et al. 2003).
 
@@
 processes like competition by rejecting a general-purpose null (Gotelli
-and Ulrich 2009). Estimating pairwise coefficients directly seems like a
-much more promising approach: to the extent that the models' relative
-performance on real data sets is similar to the range of results shown
-in Figure 3, scientists in this field could easily double their
-explanatory power by switching from null models to linear regression and
-partial covariances, or triple it by switching to a Markov network.
+and Ulrich 2009). Moreover, the null model's estimates were extremely
+similar to what could be obtained from a simple correlation matrix,
+which raises questions about how much additional value the null model
+adds. Estimating pairwise coefficients directly seems like a much more
+promising approach: to the extent that the models' relative performance
+on real data sets is similar to the range of results shown in Figure 3,
+scientists in this field could {[}{[}easily double their explanatory
+power by switching from null models to partial correlations or
+generalized linear models, or triple it by switching to a Markov
+network{]}{]}. Regardless of the specific methods ecologists ultimately
+choose, the most important consideration is clearly that we need to
+control for indirect effects and estimate conditional relationships
+between species, rather than marginal ones.
 
-\textbf{\emph{Acknowledgements:}} This research was funded by a Graduate
-Research Fellowship from the US National Science Foundation and
-benefited greatly from discussions with A. Sih, M. L. Baskett, R.
-McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A.
-K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding,
-and N. J. Gotelli provided useful feedback on an earlier draft of this
-work.
+\paragraph{Acknowledgements:}\label{acknowledgements}
 
-\textbf{\emph{References:}}
+This research was funded by a Graduate Research Fellowship from the US
+National Science Foundation and benefited greatly from discussions with
+A. Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C.
+S. Tysor. Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li,
+D. L. Miller, N. Golding, N. J. Gotelli, and C. F. Dormann provided
+useful feedback on earlier drafts of this work.
+
+\setlength{\parindent}{0cm} \setlength{\parskip}{4pt}
+
+\textbf{References:}
+
+\textbf{Figure captions}
+
+\textbf{Figure 1.} \textbf{A.} A small network of three competing
+species. The tree (top) tends not to co-occur with either of the two
+shrub species, as indicated by the strongly negative coefficient linking
+them. The two shrub species also compete with one another, as indicated
+by their negative coefficient (circled), but this effect is
+substantially weaker. \textbf{B.} In spite of the competitive
+interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors
+positively correlated (circled). \textbf{C.} Controlling for the tree
+species' presence with a conditional method such as a partial covariance
+or a Markov network allows us to correctly identify the negative
+interaction between these two species (circled).
+
+\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
+The depicted abiotic environment favors the occurrence of both species
+(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
+negative \(\beta\) coefficient linking these two species implies that
+they co-occur less than expected under independence. \textbf{B.}
+Relative probabilities of all four possible presence-absence
+combinations for Species 1 and Species 2. The exponent includes
+\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
+it is absent (\(y_1 = 0\)). Similarly, the exponent includes
+\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
+\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
+constant \(Z\), ensures that the four relative probabilities sum to 1.
+In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
+we can find the expected frequencies of all possible co-occurrence
+patterns between the two species of interest. \textbf{D.} If \(\beta\)
+equaled zero (e.g.~if the species no longer competed for the same
+resources), then the reduction in competition would allow each species
+to increase its occurrence rate and the deficit of co-occurrences would
+be eliminated.
+
+\textbf{Figure 3.} Proportion of variance in interaction coefficients
+explained by each method across varying numbers of sampled locations.
 
@@
 of theoretical biology 262:323--329.
-
-Blois, J. L., N. J. Gotelli, A. K. Behrensmeyer, J. T. Faith, S. K.
-Lyons, J. W. Williams, K. L. Amatangelo, A. Bercovici, A. Du, J. T.
-Eronen, and others. 2014. A framework for evaluating the influence of
-climate, dispersal limitation, and biotic interactions using fossil
-pollen associations across the late Quaternary. Ecography 37:1095--1108.
 
@@
 
+Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
+Inferring species interaction networks from species abundance data: A
+comparative evaluation of various statistical and machine learning
+methods. Ecological Informatics 5:451--464.
+
 Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
@@
 C (Applied Statistics) 54:1--20.
-
-Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
-Informative Default Prior Distribution for Logistic and Other Regression
-Models. The Annals of Applied Statistics 2:1360--1383.
 
@@
 
-Lee, E. D., C. P. Broedersz, and W. Bialek. 2013. Statistical mechanics
-of the US Supreme Court. arXiv:1306.5004 {[}cond-mat, physics:physics,
-q-bio{]}.
-
 Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
@@
 
+Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K.
+M. Parris, P. A. Vesk, and M. A. McCarthy. 2014. Understanding
+co-occurrence by modelling species simultaneously with a Joint Species
+Distribution Model (JSDM). Methods in Ecology and Evolution:n/a--n/a.
+
 R Core Team. 2015. R: A Language and Environment for Statistical
@@
 
-Schneidman, E., M. J. Berry, R. Segev, and W. Bialek. 2006. Weak
-pairwise correlations imply strongly correlated network states in a
-neural population. Nature 440:1007--1012.
-
 Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.
@@
 
+Veech, J. A. 2013. A probabilistic model for analysing species
+co-occurrence. Global Ecology and Biogeography 22:252--260.
+
 Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
@@
 
-Wieringen, W. N. van, and C. F. Peeters. 2014. Ridge Estimation of
-Inverse Covariance Matrices from High-Dimensional Data. arXiv preprint
-arXiv:1403.0904.
-
 Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
-
-\textbf{\emph{Figure captions}}
-
-\textbf{\emph{Figure 1.}} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, as indicated
-by their negative coefficient (circled), but this effect is
-substantially weaker. \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network allows us to correctly identify the negative
-interaction between these two species (circled).
-
-\textbf{\emph{Figure 2.}} \textbf{A.} A small Markov network with two
-species. The depicted abiotic environment favors the occurrence of both
-species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
-these two species implies that they co-occur less than expected under
-independence. \textbf{B.} Relative probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. The exponent
-includes \(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but
-not when it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
-we can find the expected frequencies of all possible co-occurrence
-patterns between the two species of interest. \textbf{D.} If \(\beta\)
-equaled zero (e.g.~if the species no longer competed for the same
-resources), then the reduction in competition would allow each species
-to increase its occurrence rate and the deficit of co-occurrences would
-be eliminated.
-
-\textbf{\emph{Figure 3.}} Proportion of variance in interaction
-coefficients explained by each method with 5, 10, or 20 species arrayed
-across varying numbers of sampled locations when environmental filtering
-was absent (top row) or present (bottom row). A negative \(R^2\) values
-implies that the squared error associated with the corresponding subset
-of the predictions was larger than the error one would get from assuming
-that all coefficients equalled zero.
-
 \end{flushleft}

### 2015-07-23 Harris_2015/recovery.R

@@ -40,7 +40,7 @@ f = function(filename){
   
   bc = BC(Y = x, model = "community", its = 1000)
   
-  bc_inv = 0
+  `partial BayesComm` = 0
   
   for(i in 1:nrow(bc$trace$R)){
     Sigma = matrix(0, nrow = ncol(x), ncol = ncol(x))
@@ -48,7 +48,7 @@ f = function(filename){
     Sigma <- Sigma + t(Sigma)                   # Fill in lower triangle
     diag(Sigma) <- 1  # Diagonal equals 1 in multivariate probit model
     
-    bc_inv = bc_inv + cor2pcor(Sigma) / nrow(bc$trace$R)
+    `partial BayesComm` = `partial BayesComm` + cor2pcor(Sigma) / nrow(bc$trace$R)
   }
   
   data.frame(
@@ -60,7 +60,7 @@ f = function(filename){
     `Markov network` = rosie$beta[upper.tri(rosie$beta)],
     GLM = coef_matrix[upper.tri(coef_matrix)],
     `BayesComm` = colMeans(bc$trace$R),
-    bc_inv = bc_inv[upper.tri(bc_inv)]
+    `partial BayesComm` = `partial BayesComm`[upper.tri(`partial BayesComm`)]
   )
 }
 

### 2015-07-28 Harris_2015/manuscript-materials/draft.Rmd
stuff

@@
 \centering
+
 $p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
+
 \raggedright
@@
 [@murphy_machine_2012]. In this way, the $\beta$ coefficients behave
-very similarly to the coefficients in a logistic regression model [@lee_learning_2012].
+very similarly to the coefficients in a logistic regression model
+[@lee_learning_2012].
 
@@
 #### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
-In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
-however, ecologists will often need to estimate these parameters from co-occurrence data
-(i.e. from a matrix of ones and zeros indicating which species are present at which sites).
-When the number of species is reasonably small, one can compute exact maximum
-likelihood estimates for all of the $\alpha$ and $\beta$ coefficients given a
-presence-absence matrix by optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
-networks like the ones considered here have unimodal likelihood surfaces
-[@murphy_machine_2012], ensuring that this procedure will always converge on
-the global maximum. This maximum represents the unique combination of $\alpha$ and $\beta$
-coefficients that would be expected to produce exactly the observed co-occurrence
-frequencies (i.e. it matches the sufficient statistics of the model distribution to the
-sufficient statistics of the data). For the analyses in this paper, I used the *rosalia* package
-[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
-define the objective function and gradient in R code and then find the optimum.
+In the previous two sections, the values of $\alpha$ and $\beta$ were known. In
+practice, however, ecologists will often need to estimate these parameters from
+co-occurrence data (i.e. from a matrix of ones and zeros indicating which
+species are present at which sites). When the number of species is reasonably
+small, one can compute exact maximum likelihood estimates for all of the
+$\alpha$ and $\beta$ coefficients given a presence-absence matrix by optimizing
+$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the ones
+considered here have unimodal likelihood surfaces [@murphy_machine_2012],
+ensuring that this procedure will always converge on the global maximum. This
+maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
+that would be expected to produce exactly the observed co-occurrence frequencies
+(i.e. it matches the sufficient statistics of the model distribution to the
+sufficient statistics of the data). For the analyses in this paper, I used the
+*rosalia* package [@harris_rosalia_2015] for the R programming language
+[@r_core_team_r_2015] to define the objective function and gradient in R code
+and then find the optimum.
 
@@
 *non*-co-occurrences between two species). The software then calculated a
-$Z$ statistic for each species pair using this null distribution. After
-multiplying this statistic by $-1$ so that positive values corresponded to
-facilitation and negative values corresponded to competition, I used it as
-another estimate of species interactions.
+$Z$ statistic for each species pair using this null distribution, which I used
+as its estimator of species interactions.
 

### 2015-07-28 Harris_2015/recovery.R
recovery code

@@ -6,6 +6,21 @@ library(BayesComm)
 
 set.seed(1)
 
+
+# Load in the results from the `pairs` program
+pairs_txt = readLines("fakedata/Pairs.txt")
+
+# Find areas of the data file that correspond
+# to species pairs' results
+beginnings = grep("Sp1", pairs_txt) + 1
+ends = c(
+  grep("^[^ ]", pairs_txt)[-1],
+  length(pairs_txt)
+) - 1
+# The above code breaks on the very last line of the file
+ends[length(ends)] = ends[length(ends)] + 1
+
+
 f = function(filename){
   # Multiplying by one is necessary to prevent silly errors
   # regarding TRUE/FALSE versus 1/0
@@ -19,14 +34,18 @@ f = function(filename){
   x = raw_obs[ , species_is_variable]
   truth = readRDS(filename)[["truth"]][pair_is_variable[upper.tri(pair_is_variable)]]
   
-  p_corr = pcor.shrink(x)
-  
   splitname = strsplit(filename, "/|-|\\.")[[1]]
   n_sites = as.integer(splitname[[3]])
   rep = as.integer(splitname[[4]])
   
+  
+  ####### Partial correlations
+  p_corr = pcor.shrink(x)
+  
+  ####### Correlations
   corr = cor(x)
   
+  ####### GLM
   coef_matrix = matrix(0, ncol(x), ncol(x))
   for(i in 1:ncol(x)){
     if(var(x[,i]) > 0){
@@ -36,8 +55,10 @@ f = function(filename){
   }
   coef_matrix = (coef_matrix + t(coef_matrix)) / 2
   
+  ####### Markov network
   rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2))
   
+  ####### BayesComm and partial BayesComm
   bc = BC(Y = x, model = "community", its = 1000)
   
   `partial BayesComm` = 0
@@ -51,16 +72,60 @@ f = function(filename){
     `partial BayesComm` = `partial BayesComm` + cor2pcor(Sigma) / nrow(bc$trace$R)
   }
   
+  ####### Pairs
+  # Find the line where the current data set is mentioned in
+  # pairs.txt
+  filename_line = grep(
+    paste0(
+      gsub("fakedata/(.*)\\.rds", "\\1", filename),
+      "-"
[…]

### 2015-07-30 Harris_2015/manuscript-materials/draft.Rmd
graphs

@@
 
-**Abstract:** Estimating species interactions from observational data is one of the most
-controversial tasks in community ecology. One difficulty is that a single
-pairwise interaction can ripple through an ecological network and produce
+**Abstract:** Estimating species interactions from observational data is one of
+the most controversial tasks in community ecology. One difficulty is that a
+single pairwise interaction can ripple through an ecological network and produce
 surprising indirect consequences. For example, two competing species would
@@
 to occur in unshaded areas (Figure 1B). If this sort of effect is common, then
-the vast majority of ecologists' methods, which rely on test statistics describing
-the total number of co-occurrences or overall correlation between putatively
-interacting species [@diamond_island_1975; @strong_ecological_1984;
-@gotelli_empirical_2009; @veech_probabilistic_2013;
-@pollock_understanding_2014] will not generally provide much
-evidence regarding species' direct effects on one another.
+the vast majority of ecologists' methods, which rely on test statistics
+describing the total number of co-occurrences or overall correlation between
+putatively interacting species [@diamond_island_1975; @strong_ecological_1984;
+@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
+will not generally provide much evidence regarding species' direct effects on
+one another.
 
 While competition doesn't reliably leave the expected pattern in community
-structure (a low co-occurrence rate at the landscape level), it nevertheless does
-leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
+structure (a low co-occurrence rate at the landscape level), it nevertheless
+does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
 there will be a deficit of co-occurrences between the competing shrub species,
@@
 network) than we could get from their overall co-occurrence rates. This kind of
-precision is difficult to obtain from null models, which begin with the assumptions
-that all the pairwise interactions are zero and thus don’t need to be controlled for.
-Nevertheless, null models have dominated this field for more than three decades
-[@diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009].
+precision is difficult to obtain from null models, which begin with the
+assumptions that all the pairwise interactions are zero and thus don’t need to
+be controlled for.  Nevertheless, null models have dominated this field for more
+than three decades [@diamond_island_1975; @strong_ecological_1984;
+@gotelli_empirical_2009].
 
@@
 #### Twenty species.
-The accuracy of the four evaluated methods varied substantially, depending on
-the parameters that produced the simulated communities (Figure 3). In general,
-however, there was a consistent ordering…
+The accuracy of the evaluated methods varied substantially, depending on
+the parameters that produced the simulated communities (Figure 3a). In general,
+however, there was a consistent ordering.  The Markov network explained the
+largest portion of the variance (34.6% overall), followed
+by the generalized linear models (29.9%), partial correlations from the raw
+presence-absence data (27.5%), and partial correlations from BayesComm, the joint
+species distribution model (26.1).  The three methods that did not attempt to
+control for indirect interactions all explained less than 20% of the variance.
+Of these, the sample correlation matrix based on the raw data performed the best
+(18.6%), followed by the null model (15.4) and BayesComm's correlation matrix
+(11.6).
 
-The three methods that estimated marginal relationships among species and did
-not explicitly control for indirect species interactions (the null model, the
-joint species distribution model’s correlation coefficients, and the sample
-correlation coefficients) differed from one another in their $R^2$ values,
-but their estimates were surprisingly similar to one another. In particular, the
-correlation coefficients were extremely similar to the estimates  from the other
-two methods (Spearman’s rho = X for the joint species distribution model’s
-correlation matrix and Y for the null model).
+Although these last three methods had very different $R^2$ values, there was a
+very close mapping between their estimates (after controlling for the size of
+the simulated landscapes; Figure 3b).  This suggests that the correlation
+matrices from joint species distribution models and especially the
+effect sizes from null models contain essentially the same information that
+could be obtained more easily by calculating correlation coefficients between
+species' presence-absence vectors. This, in turn, demonstrates that these
+methods do not effectively control for any indirect effects in the network.
 
@@
 effects (Figure 1). For cases where fitting a Markov network is computationally
-infeasible, these results also indicate that partial covariances and generalized linear
- models (the two methods that estimated conditional relationships rather than
-marginal ones) can both provide useful approximations. The partial correlations'
-success on simulated data may not carry over to real data sets, however;
-@loh_structure_2013 show that the linear approximations can be less reliable
-in cases where the true interaction matrix contains more structure (e.g. guilds
-or trophic levels). Similarly, the approximation involved in using separate
-generalized linear models for each species can occasionally lead to catastrophic
-overfitting with small-to-moderate sample sizes [@lee_learning_2012].
-For these reasons, it will usually be best to fit a Markov network
-rather than one of the alternative methods when one’s computational resources
-allow it.
+infeasible, these results also indicate that partial covariances and generalized
+linear models (the two methods that estimated conditional relationships rather
+than marginal ones) can both provide useful approximations. The partial
+correlations' success on simulated data may not carry over to real data sets,
+however; @loh_structure_2013 show that the linear approximations can be less
+reliable in cases where the true interaction matrix contains more structure
+(e.g. guilds or trophic levels). Similarly, the approximation involved in using
+separate generalized linear models for each species can occasionally lead to
+catastrophic overfitting with small-to-moderate sample sizes
+[@lee_learning_2012]. For these reasons, it will usually be best to fit a Markov
+network rather than one of the alternative methods when one’s computational
+resources allow it.
 
@@
 
-Markov networks---particularly the Ising model for binary networks---have been studied
-in statistical physics for nearly a century [@cipra_introduction_1987],
-and the models' properties,
-capabilities, and limits are well-understood in a huge range of applications.
-Modeling species interactions using the same framework would thus allow
-ecologists to tap into an enormous set of existing discoveries and techniques
-for dealing with indirect effects, stability, and alternative stable states.
-Numerous other extensions are possible: Markov networks
-can be fit with a mix of discrete and continuous variables, for example
+Markov networks---particularly the Ising model for binary networks---have been
+studied in statistical physics for nearly a century [@cipra_introduction_1987],
+and the models' properties, capabilities, and limits are well-understood in a
+huge range of applications. Modeling species interactions using the same
+framework would thus allow ecologists to tap into an enormous set of existing
+discoveries and techniques for dealing with indirect effects, stability, and
+alternative stable states. Numerous other extensions are possible: Markov
+networks can be fit with a mix of discrete and continuous variables, for example
 [@lee_learning_2012]. There are even methods [@whittam_species_1981;
@@
 pairs or specific ecological processes like competition by rejecting a
-general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s estimates
-were extremely similar to what could be obtained from a simple correlation matrix,
-which raises questions about how much additional value the null model adds.
-Estimating pairwise coefficients directly seems like a much more promising approach:
-to the extent that the models' relative performance on real data sets is similar to the
-range of results shown in Figure 3, scientists in this field could [[easily
-double their explanatory power by switching from null models to partial correlations or
-generalized linear models, or triple it by switching to a Markov network]]. Regardless
-of the specific methods ecologists ultimately choose, the most important consideration
-is clearly that we need to control for indirect effects and estimate conditional relationships
+general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s
+estimates were extremely similar to what could be obtained from a simple
+correlation matrix, which raises questions about how much additional value the
+null model adds. Estimating pairwise coefficients directly seems like a much
+more promising approach: to the extent that the models' relative performance on
+real data sets is similar to the range of results shown in Figure 3, scientists
+in this field could [[easily double their explanatory power by switching from
+null models to partial correlations or generalized linear models, or triple it
+by switching to a Markov network]]. Regardless of the specific methods
+ecologists ultimately choose, the most important consideration is clearly that
+we need to control for indirect effects and estimate conditional relationships
 between species, rather than marginal ones.
@@
 Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
-Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
-of this work.
+Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier
+drafts of this work.
 
@@
 **Figure 3.**
-Proportion of variance in interaction coefficients explained by each method
+**A.** Proportion of variance in interaction coefficients explained by each method
 across varying numbers of sampled locations.
+**B.** The $Z$-scores produced by the null model (i.e. the Pairs software) for
+each pair of species are very predictable from the correlation between the
+presence-absence vectors of those same species and from the number of sites on
+the landscape.

### 2015-07-30 Harris_2015/manuscript-materials/draft.Rmd
appeasing the editor

@@
 
+<!-- exp(qlogis(.975, scale = 2)) -->
+
 For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
-as described above. Given the large number of parameters associated with
-some of the networks to  be estimated, I regularized the likelihood
-using a logistic prior distribution with a scale of 2 on the $\alpha$ and
-$\beta$ terms.  This convex prior distribution has a similar shape to the
-Laplace distribution used in LASSO regularization (especially in the tails),
-but does not force the estimates to be exactly zero.
+as described above. For the analyses with 20 species, I added a very weak
+logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
+ensure that the model estimates were always finite. The bias introduced by this
+prior should be small: the 95% credible interval on $\beta$ indicates that one
+species' presence should not generally affect the odds of observing a different
+species by more than a factor of 1500 (which is not much of a constraint). The
+logistic distribution was chosen because it is convex and has a similar shape to
+the Laplace distribution used in LASSO regularization (especially in the tails),
+but it is differentiable everywhere and does not force any estimates to be
+exactly zero.
 
@@
 data, the partial correlation can be thought of as a computationally efficient
-approximation to the full Markov network model [@loh_structure_2013].
+approximation to the full Markov network model [@loh_structure_2013]. This sort
+of model is very common for drawing inferences about interaction networks in
+bioinformatics contexts [@REF].
 
@@
 
-Because the different methods mostly describe species interactions on different
-scales (e.g. correlations versus $Z$ scores versus regression coefficients), I used
-linear regression through the origin to rescale the different estimates produced
-by each method so that they had a consistent interpretation. After rescaling each
-method’s estimates, I calculated squared errors between the scaled interaction
-estimates and “true” interaction values for each of the [[N]] species pairs
-across all the simulated data sets. These squared errors determined the
-proportion of variance explained for different combinations of method and
-landscape size (compared with a baseline model that assumed all interaction
-strengths to be zero).
+For the larger landscapes, I evaluated the relationship between each method's
+estimates and the "true" interaction strengths among all of the species that
+were observed to occur across the simulated landscapes. This kind of evaluation
+is important because ecologists need a consistent way to know how strong species
+interactions are regardless of which species were present in a particular data
+set or how many observations were taken. Because the different methods mostly
+describe species interactions on different scales (e.g. correlations versus $Z$
+scores versus regression coefficients), I used linear regression through the
+origin to rescale the different estimates produced by each method so that they
+had a consistent interpretation. After rescaling each method’s estimates, I
+calculated squared errors between the scaled interaction estimates and “true”
+interaction values for each of the [[N]] species pairs across all the simulated
+data sets. These squared errors determined the proportion of variance explained
+for different combinations of method and landscape size (compared with a
+baseline model that assumed all interaction strengths to be zero).
 

### 2015-07-30 Harris_2015/manuscript-materials/draft.Rmd
tighten up early methods

@@
 statistical terms, this implies that ecologists want to estimate *conditional*
-("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
-The example with the shrubs and trees in
-Figure 1 shows how the two measures can have opposite signs,
-and suggests that conditional relationships are more relevant
-for drawing inferences about species interactions (e.g. competition).
+("all-else-equal") relationships, rather than *marginal* ("overall")
+relationships. The example with the shrubs and trees in Figure 1 shows how the
+two measures can have opposite signs, and suggests that conditional
+relationships are more closely related to the underlying ecological interactions.
 
-Markov networks define species interactions in conditional terms, and use these
-conditional relationships to determine whether a given combination of co-occurring
-species is consistent with their interactions.  Given a set of binary outcomes
-for each species (i.e. 0 for absence and 1 for presence), the Markov network
-defines the relative probability of observing a given presence-absence vector,
-$\vec{y}$, as
+Markov networks provide a method for translating back and forth between the
+conditional relationships among species and the kinds of species assemblages
+that these relationships produce. In this section, I show how a set of
+conditional relationships can be used to determine the kinds of species
+assemblages that can occur. Methods for estimating conditional relationships
+from presence-absence data are discussed in the next section.
+
+Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
+presence), a Markov network defines the relative probability of observing a
+given presence-absence vector, $\vec{y}$, as
 
@@
 
-Here, $\alpha_{i}$ is an intercept term determining the amount that
-the presence of species $i$ contributes
-to the log-probability of $\vec{y}$; it directly controls the prevalence of
-species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
-species $i$ and species $j$ contributes to the log-probability; it controls
-the probability that the two species will be found together (Figure 2A, Figure 2B).
-Because the relative probability of a presence-absence vector
-increases when positively-associated species co-occur and decreases when
-negatively-associated species co-occur, the model tends to produce assemblages
-that have many pairs of positively-associated species and relatively few pairs
-of negatively-associated species (exactly as an ecologist might expect).
-
-A major benefit of Markov networks is the fact that the conditional
-relationships between species can be read directly off the matrix of $\beta$
-coefficients [@murphy_machine_2012]. For example, if the coefficient linking
-two mutualist species is $+2$, then---all else equal---the odds of observing
-either species increase by a factor of $e^2$ when its partner is present
-[@murphy_machine_2012]. In this way, the $\beta$ coefficients behave
-very similarly to the coefficients in a logistic regression model
-[@lee_learning_2012].
+Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
+of species $i$ contributes to the log-probability of $\vec{y}$; it directly
+controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
+that the co-occurrence of species $i$ and species $j$ contributes to the
+log-probability; it controls the conditional relationship between two species,
+i.e. the probability that they will be found together, after controlling for the
+other species in the network (Figure 2A, Figure 2B). Because the relative
+probability of a presence-absence vector increases when positively-associated
+species co-occur and decreases when negatively-associated species co-occur, the
+model tends to produce assemblages that have many pairs of positively-associated
+species and relatively few pairs of negatively-associated species (exactly as an
+ecologist might expect).
 
@@
 competitor), then species' marginal association rates can differ from this
-expectation. For this reason, it is important to consider how coefficients'
-effects propagate through the network, as discussed below.
-
-Estimating the marginal relationships predicted by a Markov network is more
-difficult than estimating conditional relationships, because doing so requires
-absolute probability estimates. Turning the relative probability given by
-Equation 1 into an absolute probability entails scaling by a normalizing constant,
-$Z$, so that that the probabilities of all possible assemblages will sum to one (bottom of
-Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
-quickly becomes infeasible as the number of species increases: with
-$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
-required spirals exponentially into the billions and beyond.
-Numerous techniques are available for keeping the calculations tractable,
+expectation. Predicting the marginal relationships between species from their
+conditional interactions entails summing over the different possible assemblages
+(Figure 2B), and the number of possible assemblages can be intractably large.
+Numerous techniques are available for keeping the calculations feasible,
 e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
@@
 #### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
-In the previous two sections, the values of $\alpha$ and $\beta$ were known. In
-practice, however, ecologists will often need to estimate these parameters from
-co-occurrence data (i.e. from a matrix of ones and zeros indicating which
-species are present at which sites). When the number of species is reasonably
-small, one can compute exact maximum likelihood estimates for all of the
-$\alpha$ and $\beta$ coefficients given a presence-absence matrix by optimizing
-$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the ones
-considered here have unimodal likelihood surfaces [@murphy_machine_2012],
+In the previous section, the values of $\alpha$ and $\beta$ were known and the
+goal was to make predictions about possible species assemblages. In
+practice, however, ecologists will often need to estimate the parameters from
+an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
+indicating which species are present at which sites). When the number of species
+is reasonably small, one can compute exact maximum likelihood estimates for all
+of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
+optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
+ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
 ensuring that this procedure will always converge on the global maximum. This
@@
 interaction coefficients in that figure (Appendix A). Each of the methods
-described below was then evaluated on its ability to correctly infer that the two
-shrub species competed with one another, despite their frequent co-occurrence.
+described below was then evaluated on its ability to correctly infer that the
+two shrub species competed with one another, despite their frequent
+co-occurrence.
 
-I also simulated a second set of landscapes using a stochastic community
-model based on generalized Lotka-Volterra dynamics, as described in the
-Appendix. In these simulations, each species pair was randomly assigned to
-either compete for a portion of the available carrying capacity (negative
-interaction) or to act as mutualists (positive interaction).  In these
-simulations, mutualism operated by mitigating the effects of intraspecific
-competition on each partner's death rate.  For these analyses, I simulated
-landscapes with up to 20 species and 25, 200, or 1600 sites (50 replicates per
-landscape size; see Appendix).
+I also simulated a second set of landscapes using a stochastic community model
+based on generalized Lotka-Volterra dynamics, as described in the Appendix. In
+these simulations, each species pair was randomly assigned to either compete for
+a portion of the available carrying capacity (negative interaction) or to act as
+mutualists (positive interaction).  In these simulations, mutualism operated by
+mitigating the effects of intraspecific competition on each partner's death
+rate.  For these analyses, I simulated landscapes with up to 20 species and 25,
+200, or 1600 sites (50 replicates per landscape size; see Appendix).
 

### 2015-07-31 Harris_2015/manuscript-materials/draft.Rmd
more

@@
 \setlength{\parindent}{1cm}
-\setlength{\parskip}{6pt}
+\setlength{\parskip}{4pt}
 
@@
 
-#### Conditional relationships and Markov networks.
-Ecologists are often interested in learning about direct interactions between
-species, controlling for the indirect influence of other species. In
-statistical terms, this implies that ecologists want to estimate *conditional*
-("all-else-equal") relationships, rather than *marginal* ("overall")
-relationships. The example with the shrubs and trees in Figure 1 shows how the
-two measures can have opposite signs, and suggests that conditional
-relationships are more closely related to the underlying ecological interactions.
-
+#### Markov networks.
 Markov networks provide a method for translating back and forth between the
-conditional relationships among species and the kinds of species assemblages
-that these relationships produce. In this section, I show how a set of
-conditional relationships can be used to determine the kinds of species
+conditional relationships among species (Figure 1C) and the kinds of species
+assemblages that these relationships produce. In this section, I show how a set
+of conditional relationships can be used to determine the kinds of species
 assemblages that can occur. Methods for estimating conditional relationships
@@
 \raggedright
+\setlength{\parindent}{1cm}
 
@@
 
-Of course, if all else is *not* equal (e.g. Figure 1,
-where the presence of one competitor is associated with release from another
-competitor), then species' marginal association rates can differ from this
-expectation. Predicting the marginal relationships between species from their
-conditional interactions entails summing over the different possible assemblages
-(Figure 2B), and the number of possible assemblages can be intractably large.
-Numerous techniques are available for keeping the calculations feasible,
-e.g. via analytic approximations [@lee_learning_2012] or Monte Carlo
-sampling [@salakhutdinov_learning_2008], but they are beyond
-the scope of this paper.  Alternatively, as noted below, some common linear
-and generalized linear methods can also be used as computationally efficient
-approximations to the full network.
+Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
+competitor is associated with release from another competitor), then species'
+marginal association rates can differ from this expectation. Predicting the
+marginal relationships between species from their conditional interactions
+entails summing over the different possible assemblages (Figure 2B), and the
+number of possible assemblages can be intractably large, though some methods
+beyond the scope of this paper can be employed to keep the calculations feasible
+[@lee_learning_2012; @salakhutdinov_learning_2008]. Alternatively, as noted
+below, some common linear and generalized linear methods can also be used as
+computationally efficient approximations to the full network
+[@lee_learning_2012; @loh_structure_2013].
 
@@
 ensure that the model estimates were always finite. The bias introduced by this
-prior should be small: the 95% credible interval on $\beta$ indicates that one
-species' presence should not generally affect the odds of observing a different
-species by more than a factor of 1500 (which is not much of a constraint). The
-logistic distribution was chosen because it is convex and has a similar shape to
-the Laplace distribution used in LASSO regularization (especially in the tails),
-but it is differentiable everywhere and does not force any estimates to be
-exactly zero.
+prior should be small: the 95% credible interval on $\beta$ only requires that
+one species' effect on the odds of observing a different species to be less than
+a factor of 1500 (which is not much of a constraint). The logistic distribution
+was chosen because it is convex and has a similar shape to the Laplace
+distribution used in LASSO regularization (especially in the tails), but unlike
+the Laplace distribution it is differentiable everywhere and does not force any
+estimates to be exactly zero.
 
-I also evaluated five alternative methods from the existing literature to
-calibrate the performance of the Markov network as a method for estimating
-species interactions. The first two alternative interaction measures were the
-sample correlations and the partial correlations between each pair of species'
-data vectors on the landscape [@albrecht_spatial_2001; @faisal_inferring_2010].
-Because partial correlations are undefined for landscapes with
-perfectly-correlated species pairs, I used a regularized estimate based on
-James-Stein shrinkage, as implemented in the corpcor package’s `pcor.shrink`
-function with the default settings [@REF]. In the context of non-Gaussian
-data, the partial correlation can be thought of as a computationally efficient
-approximation to the full Markov network model [@loh_structure_2013]. This sort
-of model is very common for drawing inferences about interaction networks in
-bioinformatics contexts [@REF].
+I also evaluated five alternative methods from the existing literature (plus a
+novel variant of the fifth method) to calibrate the performance of the Markov
+network as a method for estimating species interactions. The first two
+alternative interaction measures were the sample correlations and the partial
+correlations between each pair of species' data vectors on the landscape
+[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
+are undefined for landscapes with perfectly-correlated species pairs, I used a
+regularized estimate based on James-Stein shrinkage, as implemented in the
+corpcor package’s `pcor.shrink` function with the default settings
+[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
+correlation can be thought of as a computationally efficient approximation to
+the full Markov network model [@loh_structure_2013]. This sort of model is very
+common for estimating relationships among genes and gene products [@REF].
 
@@
 network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
-regularized logistic regression models for each species, using
-the other species on the landscape as predictors. [[Observational
+regularized logistic regression models [@gelman_weakly_2008] for each species,
+using the other species on the landscape as predictors. [[Observational
 presence-absence data does not have enough degrees of freedom to estimate this
@@
 The last two methods used the latent correlation matrix estimated by the
-BayesComm package [@REF] to evaluate the claim from recent papers that the
-correlation coefficients estimated by “joint species distribution models”
-provide an accurate assessment of species’ pairwise interactions
-[@pollock_understanding_2014; see also @harris_generating_2015].  I reported
-both the average correlation between each pair of species’ latent variables
-across 1000 Monte Carlo samples, as well as the average *partial* correlation.
+BayesComm package [@golding_bayescomm_2015] to evaluate the claim from recent
+papers that the correlation coefficients estimated by "joint species
+distribution models" provide an accurate assessment of species’ pairwise
+interactions [@pollock_understanding_2014; see also @harris_generating_2015].  I
+used both the average correlation between each pair of species’ latent
+variables across 1000 Monte Carlo samples, as well as the average *partial*
+correlation, as BayesComm's two estimators of species interactions.
 
@@
 calculated squared errors between the scaled interaction estimates and “true”
-interaction values for each of the [[N]] species pairs across all the simulated
-data sets. These squared errors determined the proportion of variance explained
-for different combinations of method and landscape size (compared with a
-baseline model that assumed all interaction strengths to be zero).
+interaction values across all the simulated data sets. These squared errors
+determined the proportion of variance explained for different combinations of
+method and landscape size (compared with a baseline model that assumed all
+interaction strengths to be zero).
 
@@
 the parameters that produced the simulated communities (Figure 3a). In general,
-however, there was a consistent ordering.  The Markov network explained the
-largest portion of the variance (34.6% overall), followed
+however, there was a consistent ordering, with the four methods that controlled
+for indirect effects of species interactions performing the best: the Markov
+network explained the largest portion of the variance (34.6% overall), followed
 by the generalized linear models (29.9%), partial correlations from the raw
-presence-absence data (27.5%), and partial correlations from BayesComm, the joint
-species distribution model (26.1).  The three methods that did not attempt to
-control for indirect interactions all explained less than 20% of the variance.
-Of these, the sample correlation matrix based on the raw data performed the best
-(18.6%), followed by the null model (15.4) and BayesComm's correlation matrix
-(11.6).
+presence-absence data (27.5%), and partial correlations from BayesComm, the
+joint species distribution model (26.1%).  The benefit of choosing the full
+Markov network over the other three methods was largest on the smaller
+landscapes, which are also the ones that are most representative of typical
+analyses in this field [[ref]].
 
-Although these last three methods had very different $R^2$ values, there was a
-very close mapping between their estimates (after controlling for the size of
-the simulated landscapes; Figure 3b).  This suggests that the correlation
-matrices from joint species distribution models and especially the
-effect sizes from null models contain essentially the same information that
-could be obtained more easily by calculating correlation coefficients between
-species' presence-absence vectors. This, in turn, demonstrates that these
-methods do not effectively control for any indirect effects in the network.
+The three methods that did not attempt to control for indirect interactions all
+explained less than 20% of the variance. Of these, the sample correlation matrix
+based on the raw data performed the best (18.6%), followed by the null model
+(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
+methods had different $R^2$ values, there was a close mapping between
+their estimates (after controlling for the size of the simulated landscapes;
+Figure 3b).  This suggests that the correlation matrices from joint species
+distribution models and especially the effect sizes from null models only
+contain noisy versions of the information that could be obtained more easily by
+calculating correlation coefficients between species' presence-absence vectors.
 
@@
 [@bruno_inclusion_2003]; with compositional data, there is only enough
-information to provide a single number describing each species pair.  
-On the other hand, if ecologists are familiar enough with the natural history
-of their study systems that they can augment the likelihood function with a
-prior distribution or a suitable penalty, then this information could reduce the
-effective degrees of freedom to estimate and real-world results might be even
-better than those shown in Figure 3.
+information to provide a single number describing each species pair. To estimate
+asymmetric interactions, such as commensalism or predation, ecologists would
+need other kinds of data, as from time series, behavioral observations,
+manipulative experiments, or natural history. These other sources of information
+could also be used to augment the likelihood function with an informative prior
+distribution, which could lead to better results on some real data sets than
+was shown in Figure 3a.
 
@@
 discoveries and techniques for dealing with indirect effects, stability, and
-alternative stable states. Numerous other extensions are possible: Markov
-networks can be fit with a mix of discrete and continuous variables, for example
-[@lee_learning_2012]. There are even methods [@whittam_species_1981;
-@tjelmeland_markov_1998] that would allow the coefficient linking two species
-in an interaction matrix to vary as a function of the abiotic environment or
-of third-party species that tip the balance between facilitation and
-exploitation [@bruno_inclusion_2003].
+alternative stable states. Numerous other extensions are possible: for example,
+Markov networks can be fit with a mix of discrete and continuous variables,
+and likely states of the interaction network can be modeled as a function of
+the local abiotic environment [@lee_learning_2012]. There are even methods
+[@whittam_species_1981; @tjelmeland_markov_1998] that would allow one species
+to affect the sign or strength of the relationship between two other species,
+tipping the balance between facilitation and exploitation
+[@bruno_inclusion_2003].
 
@@
 patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
-null model must be interpreted with care [@roughgarden_competition_1983]. In
-complex networks of ecological interactions (and even in small networks with
-three species), it may simply not be possible to implicate individual species
-pairs or specific ecological processes like competition by rejecting a
-general-purpose null [@gotelli_empirical_2009]. Moreover, the null model’s
-estimates were extremely similar to what could be obtained from a simple
-correlation matrix, which raises questions about how much additional value the
-null model adds. Estimating pairwise coefficients directly seems like a much
-more promising approach: to the extent that the models' relative performance on
-real data sets is similar to the range of results shown in Figure 3, scientists
-in this field could [[easily double their explanatory power by switching from
-null models to partial correlations or generalized linear models, or triple it
-by switching to a Markov network]]. Regardless of the specific methods
-ecologists ultimately choose, the most important consideration is clearly that
-we need to control for indirect effects and estimate conditional relationships
-between species, rather than marginal ones.
+particular null model must be interpreted with care
+[@roughgarden_competition_1983]. In complex ecological networks (and even in
+small networks with three species), it may simply not be possible to implicate
+individual species pairs or specific ecological processes like competition by
+rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
+test statistic is effectively just a correlation coefficient (Figure 3b).
+Estimating pairwise coefficients directly seems like a much more promising
+approach: to the extent that the models' relative performance on real data sets
+is similar to the range of results shown in Figure 3a, scientists in this field
+could often double their explanatory power by switching from null models to
+Markov networks (or increase it by nearly as much with linear or generalized
+linear approximations). Regardless of the specific methods ecologists ultimately
+choose, the most important consideration is clearly that we need to control for
+indirect effects and estimate conditional relationships between species, rather
+than marginal ones.
 
@@
 Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
-Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier
-drafts of this work.
+Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers also provided
+useful feedback.
 
 \setlength{\parindent}{0cm}
-\setlength{\parskip}{4pt}
 
@@
 these two species implies that they co-occur less than expected under
-independence.  **B.** Relative probabilities of all four possible presence-absence
-combinations for Species 1 and Species 2. The exponent includes
+independence.  **B.** Relative probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. The exponent includes
 $\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
@@
 probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
-probabilities, we can find the expected frequencies of all possible co-occurrence
- patterns between the two species of interest. **D.** If $\beta$
+probabilities, we can find the expected frequencies of all possible
+co-occurrence patterns between the two species of interest. **D.** If $\beta$
 equaled zero (e.g. if the species no longer competed for the same resources),
@@
 **Figure 3.**
-**A.** Proportion of variance in interaction coefficients explained by each method
-across varying numbers of sampled locations.
+**A.** Proportion of variance in interaction coefficients explained by each
+method across varying numbers of sampled locations.
 **B.** The $Z$-scores produced by the null model (i.e. the Pairs software) for

### 2015-07-31 Harris_2015/recovery.R

@@ -183,8 +183,11 @@ library(dichromat)
 library(RColorBrewer)
 colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]
 
-pdf("manuscript-materials/figures/performance.pdf", height = 4, width = 5)
-par(mar = c(5, 4, 1, 0) + .1)
+pdf("manuscript-materials/figures/performance.pdf", height = 8, width = 5)
+par(mfrow = c(2, 1))
+par(mar = c(5, 4, 3, 0) + .1)
+
+
 matplot(
   sizes, 
   results, 
@@ -195,17 +198,22 @@ matplot(
   pch = c(15, 1, 17, 0, 16, 2, 3), 
   lty = 1,
   log = "x",
-  xlab = "Number of sites (log scale)",
+  xlab = "",
   axes = FALSE,
   xaxs = "i",
   yaxs = "i",
   col = colors,
   lwd = 1.25,
-  cex = 0.9
+  cex = 0.9,
+  bty = "l"
 )
 axis(1, c(1, 25, 200, 1600))
 axis(2, seq(0, 1, .1), las = 1)
+
 mtext(expression(R^2), side = 2, line = 3, las = 1)
+mtext("Number of sites (log scale)", side = 1, line = 2.25, at = 200)
+mtext("A.", line = 1.25, at = 25, cex = 1.5)
+
 heights = results[nrow(results), ]
 
 heights$Pairs = heights$Pairs - .01
@@ -218,10 +226,56 @@ heights$GLM = heights$GLM - .01
 
 
 text(1625, heights, colnames(results), pos = 4, cex = 0.75, col = colors)
+
+
+
+# silly code to make the next graph line up prettily with the previous one
+full_range = log10(c(25 * .9, 1600 * 8))
+base_range = log10(c(25, 1600))
+
+base_scaled = 2 * base_range / (base_range[2] - base_range[1])
+full_scaled = 2 * full_range / (base_range[2] - base_range[1])
+
+colors2 = brewer.pal(10, "Paired")[c(2, 4, 10)]
+plot(
+  z$correlation, 
+  z$Pairs, 
+  col = colors2[factor(z$n_sites)],
+  las = 1,
+  xlab = "",
+  ylab = expression(italic(Z)-score),
+  bty = "l",
+  xaxs = "i",
+  axes = FALSE,
[…]

### 2015-08-01 Harris_2015/manuscript-materials/draft.Rmd
things

@@
-*Running head:* Species interactions in Markov networks
+\chead{\textit{Species interactions in Markov networks}}
 
-**Title: Estimating species interactions from observational data with Markov
-networks**
+**Title:** Estimating species interactions from observational data with Markov
+networks
 
@@
 
-\setlength{\parindent}{1cm}
-\setlength{\parskip}{4pt}
+\setlength{\parindent}{1em}
+\setlength{\parskip}{3pt}
 
@@
 be controlled for.  Nevertheless, null models have dominated this field for more
-than three decades [@diamond_island_1975; @strong_ecological_1984;
-@gotelli_empirical_2009].
+than three decades [@strong_ecological_1984; @gotelli_empirical_2009].
 
@@
 given presence-absence vector, $\vec{y}$, as
-
-\centering
-
 $p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
-
-\raggedright
-\setlength{\parindent}{1cm}
 
@@
 *rosalia* package [@harris_rosalia_2015] for the R programming language
-[@r_core_team_r_2015] to define the objective function and gradient in R code
-and then find the optimum.
+[@r_core_team_r_2015] to optimize the Markov network parameters.
 
@@
 the full Markov network model [@loh_structure_2013]. This sort of model is very
-common for estimating relationships among genes and gene products [@REF].
+common for estimating relationships among genes and gene products
+[@friedman_sparse_2008].
 
@@
 regularized logistic regression models [@gelman_weakly_2008] for each species,
-using the other species on the landscape as predictors. [[Observational
-presence-absence data does not have enough degrees of freedom to estimate this
-many parameters, so I symmetrized the pairwise correlations by averaging the
-coefficient for predicting species $i$’s status from species $j$ and vice
-versa.]]
+using the other species on the landscape as predictors. To avoid the
+identifiability problems associated with directed cyclic graphs
+[@schmidt_modeling_2012], I then symmetrized the pairwise relationships within
+species pairs via averaging.
 
@@
 simulating possible landscapes from a null model that retains the row and column
-sums of the original matrix [@strong_ecological_1984]. Using the default options
-in the Pairs software described in @gotelli_empirical_2009, I simulated the null
-distribution of scaled C-scores (a test statistic describing the number of
-*non*-co-occurrences between two species). The software then calculated a
-$Z$ statistic for each species pair using this null distribution, which I used
-as its estimator of species interactions.
+sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
+computed by the Pairs software described in @gotelli_empirical_2009 as my
+null model-based estimator of species interactions.
 
@@
 qualitative: any method whose test statistic for the two shrubs indicated a
-negative relationship passed; other methods failed. For methods that provide
-confidence intervals or other inferential statistics (Appendix), I also reported
-how often they rejected the null hypothesis of no interaction.
+negative relationship passed; other methods failed (Appendix). For the null
+model (Pairs), I calculated statistical significance using its $Z$-score. For
+the Markov network, I used the Hessian matrix to generate approximate confidence
+intervals and noted whether these intervals included zero.
 
@@
 was positive---despite their competition for space at a mechanistic level---due
-to indirect effects of the dominant tree species. As a result, the
-covariance method falsely reported positive associations for 94% of the simulated
-landscapes, and the randomization-based null model falsely reported such
-associations 100% of the time. The methods for evaluating conditional
-relationships (Markov networks, GLMs, and both types of partial covariance
-estimates), however, successfully controlled for the indirect pathway via
-the tree species and each correctly identified the direct negative interaction
-between the shrubs 94% of the time.
+to indirect effects of the dominant tree species. As a result, the correlation
+between these species was positive in 94% of replicates, and the
+randomization-based null model falsely reported positive associations 100% of
+the time. Worse, more than 98% of these false conclusions were statistically
+significant. The partial correlation and Markov network estimates, on the other
+hand, each correctly identified the direct negative interaction between the
+shrubs 94% of the time (although the confidence intervals overlapped zero in
+most of the replicates).
 
@@
 The accuracy of the evaluated methods varied substantially, depending on
-the parameters that produced the simulated communities (Figure 3a). In general,
+the parameters that produced the simulated communities (Figure 3A). In general,
 however, there was a consistent ordering, with the four methods that controlled
@@
 landscapes, which are also the ones that are most representative of typical
-analyses in this field [[ref]].
+analyses in this field [@gotelli_empirical_2009].
 
@@
 their estimates (after controlling for the size of the simulated landscapes;
-Figure 3b).  This suggests that the correlation matrices from joint species
+Figure 3B).  This suggests that the correlation matrices from joint species
 distribution models and especially the effect sizes from null models only
@@
 distribution, which could lead to better results on some real data sets than
-was shown in Figure 3a.
+was shown in Figure 3A.
 
@@
 rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
-test statistic is effectively just a correlation coefficient (Figure 3b).
+test statistic is effectively just a correlation coefficient (Figure 3B).
 Estimating pairwise coefficients directly seems like a much more promising
 approach: to the extent that the models' relative performance on real data sets
-is similar to the range of results shown in Figure 3a, scientists in this field
+is similar to the range of results shown in Figure 3A, scientists in this field
 could often double their explanatory power by switching from null models to
@@
 the landscape.
+
+\setlength{\parskip}{2pt}

### 2015-08-01 Harris_2015/simulation.Rmd
things

@@
 set.seed(1)
+
+thin = 5E3
+n_spp = 20
+n_reps = 50
 ```
@@
 
-$$ $$
+(1 + exp(0.2 * x / K / n_spp)) / 2
+
+$$f(\vec{n} = $$
 
@@
 ```{r}
-
 simulate_communities = function(n_spp = 20,
@@
 
-```{r}
-thin = 5E3
-n_spp = 20
-n_reps = 50
-
+```{r, eval=FALSE}
 for(n_sites in c(25, 200, 1600)){

### 2015-08-01 Harris_2015/three_species.Rmd
ready to submit, maybe?
things

@@
 
-```{r}
+```{r, message = FALSE}
 library(dplyr)
@@
 library(knitr)
-library(rosalia)
+library(corpcor)
 ```
@@
 
-Once a landscape has been simulated, test whether `rosalia`, the sample covariance, and the partial 
-covariance got the sign of the shrub-shrub interaction correct.
-```{r}
+Once a landscape has been simulated, test whether `rosalia`, the sample correlation, and the partial 
+correlation got the sign of the shrub-shrub interaction correct.
+
+```{r, message=FALSE, cache = TRUE}
 evaluation = sapply(
@@
 
+    rosie = rosalia(x, trace = 0, prior = make_flat_prior(), hessian = TRUE)
+
+    se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
+    rosalia_confident_positive = pnorm(0, mean = rosie$beta[2, 3], sd = se) < .025
+    rosalia_confident_negative = pnorm(0, mean = rosie$beta[2, 3], sd = se) > .975
+
     c(
-      cov = cov(x)[2, 3] < 0,
-      partial = -solve(cov(x))[2, 3] < 0,
-      rosalia = rosalia(x, trace = 0, hessian = FALSE)$par[[3]] < 0
+      cor = cor(x)[2, 3] < 0,
+      partial = cor2pcor(cor(x))[2, 3] < 0,
+      rosalia = rosie$beta[2,3] < 0,
+      rosalia_confident_positive = rosalia_confident_positive,
+      rosalia_confident_negative = rosalia_confident_negative
     )
@@
 
-After running the "Pairs" software on transposed versions of the files simulated above, 
-we can import its results to see how the method performed.
+After running the "Pairs" software on transposed versions of the files simulated above, we can import its results to see how the method performed.
 
@@
 # Find the right rows of the text file, split them on whitespace, pull out
-# the column identified above, and find its sign
-pairs_evaluation = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
+# the column identified above, and save the Z-score
+pairs_z = Pairs.txt[grep("(Var2.*Var1)|(Var1.*Var2)", Pairs.txt)] %>% 
     strsplit(" +") %>%
-    sapply(. %>% extract(z_score_column) %>% as.numeric) %>%
-    is_greater_than(0) # Z>0 implies negative interactions because the Z-score
-                       # relates to the distribution of C-scores, which describe
-                       # the frequency with which species *do not* co-occur
+    sapply(. %>% extract(z_score_column) %>% as.numeric)
+
+# Z>0 implies negative interactions because the Z-score relates to the
+# distribution of C-scores, which describe the frequency with which species *do
+# not* co-occur. It has the opposite sign of metrics based on when species *do*
+# co-occur.
+pairs_evaluation = pairs_z > 0
 ```
 
-Percentage of correct sign estimates:
-```{r, eval=FALSE}
-mean(pairs_evaluation) * 100
-
-rowMeans(evaluation) * 100
-```
-
-```{r, echo = FALSE}
-out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation) * 100)
+```{r}
+out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation)[1:3] * 100)
 
@@
 ```
+
+```{r}
+# percentage of Pairs runs that found statistically significant mutualism at the
+# p<.05 level (two-tailed test)
+mean(pnorm(pairs_z) < 0.025) * 100
+
+# percentage of Pairs runs that found statistically significant competition at the
+# p<.05 level (two-tailed test)
+pairs_significant_competition = pnorm(pairs_z) > 0.975
+
+# percentage of rosalia runs whose 95% confidence intervals did not include 0
+rowMeans(evaluation)[c("rosalia_confident_negative", "rosalia_confident_positive")] * 100
+```

### 2015-08-01 Harris_2015/manuscript-materials/boot.R

@@ -1,0 +1,22 @@
+library(rosalia)
+
+n_boot = 50
+
+d = readRDS("fakedata/20-200-1.rds")$observed * 1
+
+full_fit = rosalia(d, prior = make_logistic_prior(scale = 2), hessian = TRUE)
+
+
+
+boots = replicate(
+  n_boot,
+  {
+    boot_sample = d[sample.int(nrow(d), replace = TRUE), ]
+    boot_fit = rosalia(
+      boot_sample, 
+      prior = make_logistic_prior(scale = 2), 
+      parlist = full_fit$opt$par
+    )
+    boot_fit$beta[upper.tri(boot_fit$beta)]
+  }
+)

### 2015-08-02 Harris_2015/manuscript-materials/draft.Rmd
x

@@
-\chead{\textit{Species interactions in Markov networks}}
+\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
+\renewcommand{\headrulewidth}{0pt}
 
@@
 of community assembly, then ecologists might expect to see their influence in
-our data sets [@macarthur_population_1958; @diamond_island_1975]. Despite
-decades of work and several major controversies, however [@lewin_santa_1983;
-@strong_ecological_1984; @gotelli_swap_2003; @connor_checkered_2013], existing
-methods for detecting competition's effects on community structure are
-unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
-another can become lost in the complex web of direct and indirect interactions
-in real assemblages. For example, the competitive interaction between the
-two shrub species in Figure 1A can become obscured by their shared tendency
-to occur in unshaded areas (Figure 1B). If this sort of effect is common, then
-the vast majority of ecologists' methods, which rely on test statistics
-describing the total number of co-occurrences or overall correlation between
-putatively interacting species [@diamond_island_1975; @strong_ecological_1984;
-@gotelli_empirical_2009; @veech_probabilistic_2013; @pollock_understanding_2014]
-will not generally provide much evidence regarding species' direct effects on
-one another.
+species composition data [@macarthur_population_1958; @diamond_island_1975].
+Despite decades of work and several major controversies, however
+[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
+@connor_checkered_2013], existing methods for detecting competition's effects
+on community structure are unreliable [@gotelli_empirical_2009]. In particular,
+species' effects on one another can become lost in the complex web of direct and
+indirect interactions in real assemblages. For example, the competitive
+interaction between the two shrub species in Figure 1A can become obscured by
+their shared tendency to occur in unshaded areas (Figure 1B). If this sort of
+effect is common, then the vast majority of ecologists' methods, which rely on
+test statistics describing the total number of co-occurrences or overall
+correlation between putatively interacting species [@diamond_island_1975;
+@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
+@pollock_understanding_2014] will not generally provide much evidence regarding
+species' direct effects on one another. The goal of this paper is to resolve
+this long-standing problem in community ecology.
 
 While competition doesn't reliably leave the expected pattern in community
-structure (a low co-occurrence rate at the landscape level), it nevertheless
+structure (low co-occurrence rates at the landscape level), it nevertheless
 does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
@@
 In this paper, following recent work by @azaele_inferring_2010 and
-@fort_statistical_2013, I show that Markov networks [undirected
-graphical models also known as Markov random fields;
-@murphy_machine_2012] can provide a framework for understanding
-the landscape-level consequences of pairwise species interactions, and
- for detecting them from observed presence-absence matrices. Markov
-networks have been used in many scientific fields for decades in similar
-contexts, from physics [where nearby particles interact magnetically
-with one another; @cipra_introduction_1987] to spatial statistics [where
-adjacent grid cells interact with one another; @harris_contact_1974;
-@gelfand_modelling_2005]. While community ecologists explored
-some related approaches in the 1980's [@whittam_species_1981],
-computational limitations had previously imposed severe approximations
-that produced unintelligible results [e.g. "probabilities" greater than one;
-@gilpin_factors_1982].  Now that it is computationally feasible to fit these
-models exactly, this kind of approach is worth a second look.
+@fort_statistical_2013, I show that Markov networks [undirected graphical models
+also known as Markov random fields; @murphy_machine_2012] can provide a
+framework for understanding the landscape-level consequences of pairwise species
+interactions, and  for detecting them from observed presence-absence matrices.
+Markov networks have been used in many scientific fields for decades in similar
+contexts, from physics [where nearby particles interact magnetically with one
+another; @cipra_introduction_1987] to spatial statistics [where adjacent grid
+cells interact with one another; @harris_contact_1974; @gelfand_modelling_2005].
+While community ecologists explored some related approaches in the 1980's
+[@whittam_species_1981], computational limitations had previously imposed severe
+approximations that led to unintelligible results [e.g. "probabilities" greater
+than one; @gilpin_factors_1982].
 
-The rest of the paper proceeds as follows. First, I discuss how Markov
-networks operate and how they can be used to simulate landscape-level
-data or to predict the direct and indirect consequences of possible
+The rest of the paper proceeds as follows. First, I discuss how Markov networks
+operate and how they can be used to simulate landscape-level data or to make
+exact predictions regarding the direct and indirect consequences of possible
 interaction matrices. Then, using simulated data sets where the "true"
-ecological structure is known, I compare this approach with several
-existing methods for detecting species interactions. Finally, I discuss
-opportunities for extending the approach presented here to larger
-problems in community ecology.
+ecological structure is known, I compare this approach with several existing
+methods for detecting species interactions. Finally, I discuss opportunities for
+extending the approach presented here to other problems in community ecology,
+such as disentangling the effects of biotic and abiotic interactions on species
+composition.
 
@@
 #### Markov networks.
-Markov networks provide a method for translating back and forth between the
+Markov networks provide a framework for translating back and forth between the
 conditional relationships among species (Figure 1C) and the kinds of species
 assemblages that these relationships produce. In this section, I show how a set
-of conditional relationships can be used to determine the kinds of species
-assemblages that can occur. Methods for estimating conditional relationships
-from presence-absence data are discussed in the next section.
+of conditional relationships can be used to determine the how groups of species
+can co-occur. Methods for estimating conditional relationships from
+presence-absence data are discussed in the next section.
 
-Given a set of binary outcomes for each species (i.e. 0 for absence and 1 for
-presence), a Markov network defines the relative probability of observing a
-given presence-absence vector, $\vec{y}$, as
+A Markov network defines the relative probability of observing a given vector of
+species-level presences (1s) and absences (0s), $\vec{y}$, as
+
+\centering
+
 $p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
+
+\raggedright
+\setlength{\parindent}{1em}
 
@@
 i.e. the probability that they will be found together, after controlling for the
-other species in the network (Figure 2A, Figure 2B). Because the relative
-probability of a presence-absence vector increases when positively-associated
-species co-occur and decreases when negatively-associated species co-occur, the
-model tends to produce assemblages that have many pairs of positively-associated
-species and relatively few pairs of negatively-associated species (exactly as an
-ecologist might expect).
+other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
+might have a value of $+2$ for two mutualists, indicating that the odds of
+observing one species are $e^2$ times higher its partner is present than when
+it is absent. Because the relative probability of a presence-absence vector
+increases when positively-associated species co-occur and decreases when
+negatively-associated species co-occur, the model tends---all else equal---to
+produce assemblages that have many pairs of positively-associated species and
+relatively few pairs of negatively-associated species (exactly as an ecologist
+might expect).
 
@@
 competitor is associated with release from another competitor), then species'
-marginal association rates can differ from this expectation. Predicting the
+marginal association rates can differ from this expectation. Determining the
 marginal relationships between species from their conditional interactions
-entails summing over the different possible assemblages (Figure 2B), and the
-number of possible assemblages can be intractably large, though some methods
-beyond the scope of this paper can be employed to keep the calculations feasible
-[@lee_learning_2012; @salakhutdinov_learning_2008]. Alternatively, as noted
-below, some common linear and generalized linear methods can also be used as
-computationally efficient approximations to the full network
+entails summing over the different possible assemblages (Figure 2B). This
+becomes intractable when the number of possible assemblages is large, though
+several methods beyond the scope of this paper can be employed to keep the
+calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
+Alternatively, as noted below, some common linear and generalized linear methods
+can also be used as computationally efficient approximations to the full network
 [@lee_learning_2012; @loh_structure_2013].
@@
 that would be expected to produce exactly the observed co-occurrence frequencies
-(i.e. it matches the sufficient statistics of the model distribution to the
-sufficient statistics of the data). For the analyses in this paper, I used the
-*rosalia* package [@harris_rosalia_2015] for the R programming language
+on average (i.e. it matches the sufficient statistics of the model distribution
+to the sufficient statistics of the data). For the analyses in this paper, I
+used the *rosalia* package [@harris_rosalia_2015] for the R programming language
 [@r_core_team_r_2015] to optimize the Markov network parameters.
@@
 a portion of the available carrying capacity (negative interaction) or to act as
-mutualists (positive interaction).  In these simulations, mutualism operated by
-mitigating the effects of intraspecific competition on each partner's death
-rate.  For these analyses, I simulated landscapes with up to 20 species and 25,
-200, or 1600 sites (50 replicates per landscape size; see Appendix).
+mutualists (positive interaction).  Here, mutualisms operate by mitigating the
+effects of intraspecific competition on each partner's death rate.  For these
+analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
+sites (50 replicates per landscape size; see Appendix).
 
 #### Recovering species interactions from simulated data.
-I used Markov networks and  several other techniques for
-determining the sign and strength of the associations between
-pairs of species (Appendix B).
+I used Markov networks and several other techniques for determining the sign and
+strength of the associations between pairs of species (Appendix B).
 
@@
 the Laplace distribution it is differentiable everywhere and does not force any
-estimates to be exactly zero.
+estimates to be exactly zero. To confirm that this procedure produced reliable
+estimates, I evaluated it on 50 bootstrap replicates (Appendix).
 
 I also evaluated five alternative methods from the existing literature (plus a
-novel variant of the fifth method) to calibrate the performance of the Markov
-network as a method for estimating species interactions. The first two
-alternative interaction measures were the sample correlations and the partial
-correlations between each pair of species' data vectors on the landscape
-[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations
-are undefined for landscapes with perfectly-correlated species pairs, I used a
-regularized estimate based on James-Stein shrinkage, as implemented in the
-corpcor package’s `pcor.shrink` function with the default settings
-[@schafer_corpcor_2014]. In the context of non-Gaussian data, the partial
-correlation can be thought of as a computationally efficient approximation to
-the full Markov network model [@loh_structure_2013]. This sort of model is very
-common for estimating relationships among genes and gene products
-[@friedman_sparse_2008].
+novel variant of the fifth method). The first two alternative interaction
+measures were the sample correlations and the partial correlations between each
+pair of species' data vectors on the landscape [@albrecht_spatial_2001;
+@faisal_inferring_2010]. Because partial correlations are undefined for
+landscapes with perfectly-correlated species pairs, I used a regularized
+estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
+`pcor.shrink` function with the default settings [@schafer_corpcor_2014]. In the
+context of non-Gaussian data, the partial correlation can be thought of as a
+computationally efficient approximation to the full Markov network model
+[@loh_structure_2013]. This sort of model is very common for estimating
+relationships among genes and gene products [@friedman_sparse_2008].
 
@@
 identifiability problems associated with directed cyclic graphs
-[@schmidt_modeling_2012], I then symmetrized the pairwise relationships within
-species pairs via averaging.
+[@schmidt_modeling_2012], I then symmetrized the relationships within species
+pairs via averaging.
 
 The fourth alternative method, described in @gotelli_empirical_2009, involved
-simulating possible landscapes from a null model that retains the row and column
+simulating new landscapes from a null model that retains the row and column
 sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
-computed by the Pairs software described in @gotelli_empirical_2009 as my
-null model-based estimator of species interactions.
+computed by the Pairs software described in @gotelli_empirical_2009 as my null
+model-based estimator of species interactions.
 
-The last two methods used the latent correlation matrix estimated by the
-BayesComm package [@golding_bayescomm_2015] to evaluate the claim from recent
-papers that the correlation coefficients estimated by "joint species
-distribution models" provide an accurate assessment of species’ pairwise
-interactions [@pollock_understanding_2014; see also @harris_generating_2015].  I
-used both the average correlation between each pair of species’ latent
-variables across 1000 Monte Carlo samples, as well as the average *partial*
-correlation, as BayesComm's two estimators of species interactions.
+The last two estimators used the latent correlation matrix estimated by the
+BayesComm package [@golding_bayescomm_2015] in order to evaluate the untested
+claim from recent papers that the correlation coefficients estimated by "joint
+species distribution models" provide an accurate assessment of species’ pairwise
+interactions [@pollock_understanding_2014; see also @harris_generating_2015].  
+In addition to using the posterior mean of the correlation coefficient
+[@pollock_understanding_2014], I also used the posterior mean of the *partial*
+correlation, which might provide a better estimate of species' conditional
+associations.
 
@@
 For the larger landscapes, I evaluated the relationship between each method's
-estimates and the "true" interaction strengths among all of the species that
-were observed to occur across the simulated landscapes. This kind of evaluation
-is important because ecologists need a consistent way to know how strong species
-interactions are regardless of which species were present in a particular data
-set or how many observations were taken. Because the different methods mostly
-describe species interactions on different scales (e.g. correlations versus $Z$
-scores versus regression coefficients), I used linear regression through the
-origin to rescale the different estimates produced by each method so that they
-had a consistent interpretation. After rescaling each method’s estimates, I
-calculated squared errors between the scaled interaction estimates and “true”
-interaction values across all the simulated data sets. These squared errors
-determined the proportion of variance explained for different combinations of
-method and landscape size (compared with a baseline model that assumed all
-interaction strengths to be zero).
+estimates and the "true" interaction strengths among all of the species pairs
+from the simulated landscapes. This kind of evaluation is important because
+ecologists need a consistent way to know how strong species interactions are
+regardless of which species were present in a particular data set or how many
+observations were taken. Because the different methods mostly describe species
+interactions on different scales (e.g. correlations versus $Z$ scores versus
+regression coefficients), I used linear regression through the origin to rescale
+the different estimates produced by each method so that they had a consistent
+interpretation. After rescaling each method’s estimates, I calculated squared
+errors between the scaled interaction estimates and “true” interaction values
+across all the simulated data sets. These squared errors determined the
+proportion of variance explained for different combinations of method and
+landscape size (compared with a baseline model that assumed all interaction
+strengths to be zero).
 
@@
 randomization-based null model falsely reported positive associations 100% of
-the time. Worse, more than 98% of these false conclusions were statistically
-significant. The partial correlation and Markov network estimates, on the other
-hand, each correctly identified the direct negative interaction between the
-shrubs 94% of the time (although the confidence intervals overlapped zero in
-most of the replicates).
+the time. Worse, more than 98% of false conclusions from the null model were
+statistically significant. The partial correlation and Markov network estimates,
+on the other hand, each correctly isolated the direct negative interaction
+between the shrubs from their positive indirect interaction 94% of the time
+(although the confidence intervals overlapped zero in most replicates).
 
@@
 for indirect effects of species interactions performing the best: the Markov
-network explained the largest portion of the variance (34.6% overall), followed
-by the generalized linear models (29.9%), partial correlations from the raw
-presence-absence data (27.5%), and partial correlations from BayesComm, the
-joint species distribution model (26.1%).  The benefit of choosing the full
+network explained the largest portion of the variance (35% overall), followed
+by the generalized linear models (30%), partial correlations from the raw
+presence-absence data (28%), and partial correlations from BayesComm, the
+joint species distribution model (26%).  The benefit of choosing the full
 Markov network over the other three methods was largest on the smaller
@@
 explained less than 20% of the variance. Of these, the sample correlation matrix
-based on the raw data performed the best (18.6%), followed by the null model
-(15.4%) and BayesComm's correlation matrix (11.6%). Although these last three
-methods had different $R^2$ values, there was a close mapping between
+based on the raw data performed the best (19%), followed by the null model
+(15%) and BayesComm's correlation matrix (11%). Although these last three
+methods had different $R^2$ values, there was a close mapping among
 their estimates (after controlling for the size of the simulated landscapes;
@@
 distribution models and especially the effect sizes from null models only
-contain noisy versions of the information that could be obtained more easily by
-calculating correlation coefficients between species' presence-absence vectors.
+contain noisy versions of the same information that could be obtained more
+easily and interpretably by calculating correlation coefficients between
+species' presence-absence vectors.
+
+The bootstrapped results confirmed that the rosalia package's estimates of
+species' conditional relationships were reasonably insensitive to sampling
+variation (Appendix), and that the above ranking of the different methods was
+robust (Appendix).
 
@@
 a common null modeling technique reliably fails. Specifically, Markov networks
-were successful even when direct interactions were largely overwhelmed by indirect
-effects (Figure 1). For cases where fitting a Markov network is computationally
-infeasible, these results also indicate that partial covariances and generalized
-linear models (the two methods that estimated conditional relationships rather
-than marginal ones) can both provide useful approximations. The partial
-correlations' success on simulated data may not carry over to real data sets,
-however; @loh_structure_2013 show that the linear approximations can be less
-reliable in cases where the true interaction matrix contains more structure
-(e.g. guilds or trophic levels). Similarly, the approximation involved in using
-separate generalized linear models for each species can occasionally lead to
-catastrophic overfitting with small-to-moderate sample sizes
+were successful even when direct interactions were largely overwhelmed by
+indirect effects (Figure 1). For cases where fitting a Markov network is
+computationally infeasible, these results also indicate that partial covariances
+and generalized linear models (the two methods that estimated conditional
+relationships rather than marginal ones) can both provide useful approximations.
+The partial correlations' success on simulated data may not carry over to real
+data sets, however; @loh_structure_2013 show that the linear approximations can
+be less reliable in cases where the true interaction matrix contains more
+structure (e.g. guilds or trophic levels). Similarly, the approximation involved
+in using separate generalized linear models for each species can occasionally
+lead to catastrophic overfitting with small-to-moderate sample sizes
 [@lee_learning_2012]. For these reasons, it will usually be best to fit a Markov
@@
 It’s important to note that none of these methods can identify the exact nature
-of the pairwise interactions (e.g. which species in a positively-associated
-pair is facilitating the other), particularly when real pairs of species can
-reciprocally influence one another in multiple ways simultaneously
-[@bruno_inclusion_2003]; with compositional data, there is only enough
-information to provide a single number describing each species pair. To estimate
-asymmetric interactions, such as commensalism or predation, ecologists would
-need other kinds of data, as from time series, behavioral observations,
-manipulative experiments, or natural history. These other sources of information
-could also be used to augment the likelihood function with an informative prior
-distribution, which could lead to better results on some real data sets than
-was shown in Figure 3A.
+of the pairwise interactions [e.g. which species in a positively-associated
+pair is facilitating the other; @schmidt_modeling_2012], particularly when real
+pairs of species can reciprocally influence one another in multiple ways
+simultaneously [@bruno_inclusion_2003]; with compositional data, there is only
+enough information to provide a single number describing each species pair.
+To estimate asymmetric interactions, such as commensalism or predation,
+ecologists would need other kinds of data, as from time series, behavioral
+observations, manipulative experiments, or natural history. These other sources
+of information could also be used to augment the likelihood function with an
+informative prior distribution, which could lead to better results on some real
+data sets than was shown in Figure 3A.
 
-Despite these limitations, Markov networks have enormous potential to improve
-ecological understanding. For example, Markov networks provide a simple answer
-to the question of how competition should affect a species' overall
-prevalence, which was a major flashpoint for the null model debates in the
-1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
-1 can be used to calculate the expected prevalence of a species in the
-absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
-Competition's effect on prevalence in a Markov network can then be calculated
-by subtracting this value and the observed prevalence (cf Figure 2D).
+Despite their limitations, Markov networks have enormous potential to improve
+ecological understanding. In particular, they are less vulnerable than some
+of the most commonly-used methods to mistakenly identifying positive species
+interactions between competing species. They also provide a simple answer to the
+question of how competition should affect a species' overall prevalence, which
+was a major flashpoint for the null model debates in the 1980’s
+[@roughgarden_competition_1983; @strong_ecological_1984]. Equation 1 can be used
+to calculate the expected prevalence of a species in the absence of biotic
+influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's
+effect on prevalence in a Markov network can then be calculated by subtracting
+this value and the observed prevalence (cf Figure 2D). This kind of insight
+would have been difficult to obtain without a generative model that makes
+predictions about the consequences of species interactions; null models (which
+presume *a priori* that interactions do not exist) have no way to make such
+predictions.
 
@@
 and the models' properties, capabilities, and limits are well-understood in a
-huge range of applications. Modeling species interactions using the same
-framework would thus allow ecologists to tap into an enormous set of existing
+huge range of applications. Using the same framework for species interactions
+would thus allow ecologists to tap into an enormous set of existing
 discoveries and techniques for dealing with indirect effects, stability, and
@@
 and likely states of the interaction network can be modeled as a function of
-the local abiotic environment [@lee_learning_2012]. There are even methods
-[@whittam_species_1981; @tjelmeland_markov_1998] that would allow one species
-to affect the sign or strength of the relationship between two other species,
-tipping the balance between facilitation and exploitation
+the local abiotic environment [@lee_learning_2012], which would provide a
+rigorous and straightforward way to incorporate whole networks of biotic
+interactions into species distribution models. Estimating each species'
+conditional relationship with both biotic and abiotic factors in the same model
+should lead to a better understanding of both kinds of ecological process. There
+are even methods [@whittam_species_1981; @tjelmeland_markov_1998] that would
+allow one species to affect the sign or strength of the relationship between
+two other species, tipping the balance between facilitation and exploitation
 [@bruno_inclusion_2003].
@@
 particular null model must be interpreted with care
-[@roughgarden_competition_1983]. In complex ecological networks (and even in
-small networks with three species), it may simply not be possible to implicate
-individual species pairs or specific ecological processes like competition by
-rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
-test statistic is effectively just a correlation coefficient (Figure 3B).
-Estimating pairwise coefficients directly seems like a much more promising
-approach: to the extent that the models' relative performance on real data sets
-is similar to the range of results shown in Figure 3A, scientists in this field
-could often double their explanatory power by switching from null models to
-Markov networks (or increase it by nearly as much with linear or generalized
-linear approximations). Regardless of the specific methods ecologists ultimately
-choose, the most important consideration is clearly that we need to control for
-indirect effects and estimate conditional relationships between species, rather
-than marginal ones.
+[@roughgarden_competition_1983]. Even in small networks with three species, it
+may simply not be possible to implicate individual species pairs or specific
+ecological processes like competition by rejecting a general-purpose null
+[@gotelli_empirical_2009], especially when the test statistic is effectively
+just a correlation coefficient (Figure 3B). Simultaneous estimation of multiple
+ecological parameters seems like a much more promising approach: to the extent
+that the models' relative performance on real data sets is similar to the range
+of results shown in Figure 3A, scientists in this field could often double their
+explanatory power by switching from null models to Markov networks (or increase
+it by nearly as much with linear or generalized linear approximations).
+Regardless of the specific methods ecologists ultimately choose, the most
+important consideration is clearly that ecologists need to control for indirect
+effects and estimate conditional relationships between species, rather than
+marginal ones.
 
@@
 Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
-Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers also provided
-useful feedback.
+Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided
+very helpful feedback on the text.
 
@@
 negative coefficient linking them. The two shrub species also compete with one
-another, as indicated by their negative coefficient (circled), but this effect
-is substantially weaker. **B.** In spite of the competitive interactions
-between the two shrub species, their shared tendency to occur in locations
-without trees makes their occurrence vectors positively correlated (circled).
-**C.** Controlling for the tree species' presence with a conditional method
-such as a partial covariance or a Markov network allows us to correctly
-identify the negative interaction between these two species (circled).
+another, but more weakly (circled coefficient). **B.** In spite of the
+competitive interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors positively
+correlated (circled). **C.** Controlling for the tree species' presence with a
+conditional method such as a partial covariance or a Markov network leads to
+correct identification of the negative shrub-shrub interaction (circled).
 
 **Figure 2.**
-**A.** A small Markov network with two species.  The depicted abiotic
-environment favors the occurrence of both species ($\alpha >0$), particularly
-species 2 ($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking
-these two species implies that they co-occur less than expected under
-independence.  **B.** Relative probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. The exponent includes
-$\alpha_1$ whenever Species 1 is present ($y_1 = 1$), but not when it is
-absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
-species $2$ is present ($y_2 = 1$), and $\beta$ only when both are present
-($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four relative
-probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** Using the
-probabilities, we can find the expected frequencies of all possible
-co-occurrence patterns between the two species of interest. **D.** If $\beta$
-equaled zero (e.g. if the species no longer competed for the same resources),
-then the reduction in competition would allow each species to increase its
-occurrence rate and the deficit of co-occurrences would be eliminated.
+**A.** A small Markov network with two species.  The abiotic environment favors
+the occurrence of both species ($\alpha >0$), particularly species 2
+($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
+species implies that they co-occur less than expected under independence.
+**B.** Relative probabilities of all four possible presence-absence combinations
+for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
+is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
+exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
+$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
+ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
+about 18.5.  **C.** Using the probabilities, we can find the expected
+frequencies of all possible co-occurrence patterns between the two species of
+interest. **D.** If $\beta$ equaled zero (e.g. if the species no longer competed
+for the same resources), then the reduction in competition would allow each
+species to increase its occurrence rate and the co-occurrence deficit would
+be eliminated.
 
@@
 the landscape.
-
-\setlength{\parskip}{2pt}

### 2015-08-02 Harris_2015/recovery.Rmd
other things

---
output: 
  pdf_document:
    keep_tex: true
title: Appendix [[X]]":"" estimating species interactions
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

This document describes how the different models were fit to the simulated data 
from Appendix [[XX]] and how each model's performance was evaluated.[^1]

[^1]: The PDF version of this document has been manually altered to omit 150
lines of output from the `corpcor` package of the form "`## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326`"

Initialization:

```{r, cache = TRUE, message = FALSE}
library(dplyr)        # For manipulating data structures
library(corpcor)      # For regularized partial covariances
library(rosalia)      # For Markov networks
library(arm)          # For regularized logistic regression
library(BayesComm)    # For joint species distribution modeling
library(RColorBrewer) # For color palette
set.seed(1)
```

Load in the results from the `pairs` program, run outside of R with
the following options:

  * Batch mode
  * Sequential swap ("s")
  * Printing all pairs ("y")
  * C-score co-occurrence measure ("c")
  * Default confidence limits (0.05)
  * Default iterations (100)
  * Maximum of 20 species

```{r, cache = TRUE}
pairs_txt = readLines("fakedata/Pairs.txt")

# Find areas of the data file that correspond
# to species pairs' results
beginnings = grep("Sp1", pairs_txt) + 1
ends = c(
  grep("^[^ ]", pairs_txt)[-1],
  length(pairs_txt)
) - 1

# The above code fails on the very last line of the file
ends[length(ends)] = ends[length(ends)] + 1
```

A function to import the data file and run each method on it:

```{r, cache = TRUE}
fit_all = function(filename){
  ######## Import ########

  # Multiplying by one is necessary to prevent silly errors
  # regarding TRUE/FALSE versus 1/0
  raw_obs = readRDS(filename)[["observed"]] * 1

  # Identify species that are never present (or never absent) so they 
  # can be dropped
  species_is_variable = diag(var(raw_obs)) > 0
  pair_is_variable = tcrossprod(species_is_variable) > 0

  x = raw_obs[ , species_is_variable]
  truth = readRDS(filename)[["truth"]][pair_is_variable[upper.tri(pair_is_variable)]]

  splitname = strsplit(filename, "/|-|\\.")[[1]]
  n_sites = as.integer(splitname[[3]])
  rep = as.integer(splitname[[4]])

  ######## Partial correlations ########
  p_corr = pcor.shrink(x)

  ######## Correlations ########
  corr = cor(x)

  ######## GLM ########
  coef_matrix = matrix(0, ncol(x), ncol(x))
  for(i in 1:ncol(x)){
    if(var(x[,i]) > 0){
      coefs = coef(bayesglm(x[,i] ~ x[ , -i], family = binomial))[-1]
      coef_matrix[i, -i] = coefs
    }
  }
  coef_matrix = (coef_matrix + t(coef_matrix)) / 2

  ######## Markov network ########
  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2))

  ######## BayesComm and partial BayesComm ########
  bc = BC(Y = x, model = "community", its = 10)

  `partial BayesComm` = 0

  for(i in 1:nrow(bc$trace$R)){
    Sigma = matrix(0, nrow = ncol(x), ncol = ncol(x))
    Sigma[upper.tri(Sigma)] <- bc$trace$R[i, ]  # Fill in upper triangle
    Sigma <- Sigma + t(Sigma)                   # Fill in lower triangle
    diag(Sigma) <- 1  # Diagonal equals 1 in multivariate probit model

    `partial BayesComm` = `partial BayesComm` + cor2pcor(Sigma) / nrow(bc$trace$R)
  }

  ######## Pairs ########
  # Find the line where the current data set is mentioned in
  # pairs.txt
  filename_line = grep(
    paste0(
      gsub("fakedata/(.*)\\.rds", "\\1", filename),
      "-"
    ),
    pairs_txt
  )

  # Which chunk of the data file corresponds to this file?
  chunk = min(which(beginnings > filename_line))

  # Split the chunk on whitespace.
  splitted = strsplit(pairs_txt[beginnings[chunk]:ends[chunk]], " +")

  # Pull out the species numbers and their Z-scores
  pairs_results = lapply(
    splitted,
    function(x){
      spp = sort(as.integer(x[3:4]))
      data.frame(
        sp1 = spp[1], 
        sp2 = spp[2], 
        z = as.numeric(x[14])
      )
    }
  ) %>% 
    bind_rows %>%
    mutate(spp = paste(sp1, sp2, sep = "-"))

  # Re-order the pairs_results to match the other methods
  m = matrix(NA, ncol(x), ncol(x))
  new_order = match(
    paste(row(m)[upper.tri(m)], col(m)[upper.tri(m)], sep = "-"),
    pairs_results$spp
  )
  ordered_pairs_results = pairs_results[new_order, ]

  ######## Output ########
  data.frame(
    truth = truth, 
    n_sites = n_sites,
    rep = rep,
    sp1 = ordered_pairs_results$sp1,
    sp2 = ordered_pairs_results$sp2,
    `partial correlation` = p_corr[upper.tri(p_corr)],
    correlation = corr[upper.tri(corr)],
    `Markov network` = rosie$beta[upper.tri(rosie$beta)],
    GLM = coef_matrix[upper.tri(coef_matrix)],
    `BayesComm` = colMeans(bc$trace$R),
    `partial BayesComm` = `partial BayesComm`[upper.tri(`partial BayesComm`)],
    Pairs = ordered_pairs_results$z
  )
}
```

Run the above function on all the files:

```{r, cache = TRUE}

# Find all the .rds files in the fakedata folder
files = dir("fakedata", pattern = "\\.rds$", full.names = TRUE)

# Run all the analyses on all the files
z = lapply(files, fit_all) %>% bind_rows %>% as.data.frame

# Fix a formatting issue with the column names
colnames(z) = gsub("\\.", " ", colnames(z))
```

Summarize the results of all 7 methods across the simulated landscapes:

```{r, cache = TRUE}

# Calculate residuals for each method
resids = sapply(
  colnames(z)[-(1:5)],
  function(i){resid(lm(z$truth ~ z[,i] + 0))}
)

sizes = sort(unique(z$n_sites))

# Compute proportion of variance explained, compared with a null that
# assumes all species interactions are 0.
results = as.data.frame(
  t(
    sapply(
      sizes,
      function(n){
        total_ss = mean(resid(lm(truth ~ 0, data = z))[z$n_sites == n]^2)
        1 - colMeans(resids[z$n_sites == n , ]^2) / total_ss
      }
    )
  )
)
# Sort the results in decreasing order
results = results[order(colMeans(results), decreasing = TRUE)]
```

Plot the results:

```{r, cache = TRUE, message=FALSE}
colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]

# Set up the plotting canvas
pdf("manuscript-materials/figures/performance.pdf", height = 8, width = 5)
par(mfrow = c(2, 1))
par(mar = c(5, 4, 3, 0) + .1)

# Plot the model performance
matplot(
  sizes, 
  results, 
  type = "o", 
  ylab = "", 
  xlim = c(25 * .9, 1600 * 8),
  ylim = c(0, .5 + 1E-9),
  pch = c(15, 1, 17, 0, 16, 2, 3), 
  lty = 1,
  log = "x",
  xlab = "",
  axes = FALSE,
  xaxs = "i",
  yaxs = "i",
  col = colors,
  lwd = 1.25,
  cex = 0.9,
  bty = "l"
)
axis(1, c(1, 25, 200, 1600))
axis(2, seq(0, 1, .1), las = 1)

mtext(expression(R^2), side = 2, line = 3, las = 1)
mtext("Number of sites (log scale)", side = 1, line = 2.25, at = 200)
mtext("A.", line = 1.25, at = 25, cex = 1.5)

# Determine how high on the y-axis each method label should be so they
# don't overlap.
heights = results[nrow(results), ]

heights$Pairs = heights$Pairs - .01
heights$correlation = heights$correlation + .01

heights$`partial correlation` = heights$`partial correlation` + .01
heights$`partial BayesComm` = heights$`partial BayesComm` - .01

heights$GLM = heights$GLM - .01

text(1625, heights, colnames(results), pos = 4, cex = 0.75, col = colors)

######

# silly code to make the next graph line up prettily with the previous one
full_range = log10(c(25 * .9, 1600 * 8))
base_range = log10(c(25, 1600))

base_scaled = 2 * base_range / (base_range[2] - base_range[1])
full_scaled = 2 * full_range / (base_range[2] - base_range[1])

# New colors for plot B
colors2 = brewer.pal(10, "Paired")[c(2, 4, 10)]
plot(
  z$correlation, 
  z$Pairs, 
  col = colors2[factor(z$n_sites)],
  las = 1,
  xlab = "",
  ylab = expression(italic(Z)-score),
  bty = "l",
  xaxs = "i",
  axes = FALSE,
  xlim = full_scaled - base_scaled[2] + 1
)
axis(1, seq(-2, 1, .5))
axis(2, seq(-50, 50, 10), las = 1)
mtext("Correlation coefficient", side = 1, line = 2.25, at = 0)
mtext("B.", line = 1.25, at = -1, cex = 1.5)

legend(
  x = 1.05, 
  y = max(z$Pairs),
  pch = 1,
  lty = 0,
  lwd = 2,
  col = colors2[1:length(unique(z$n_sites))], 
  legend = levels(factor(z$n_sites)),
  title = "Number of sites",
  bty = "n",
  cex = 0.75,
  pt.cex = 1
)
segments(-1000, 0, 1.04, 0, col = "#00000025", lwd = 1.25)
segments(0, -1000, 0, 1000, col = "#00000025", lwd = 1.25)

dev.off()
```

![Figure 3](manuscript-materials/figures/performance.pdf)

Summarize the results:

```{r, cache = TRUE}

# R-squareds computed across *all* landscape types
total_ss = mean(resid(lm(truth ~ 0, data = z))^2)
round(100 * sort(1 - colMeans(resids^2) / total_ss), 1)

# Rank correlations among the methods that estimate marginal relationships
round(cor(z[ , c(7, 10, 12)], method = "spearman")[,1], 2)
```

Plot the true interactions versus estimated interactions for two methods:

```{r, cache = TRUE}
library(ggplot2)

# ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) + 
#   stat_binhex(bins = 100) + 
#   xlab("Estimated coefficient value") + 
#   ylab("\"True\" coefficient value") + 
#   stat_hline(yintercept = 0, size = 1/8) + 
#   stat_vline(xintercept = 0, size = 1/8) + 
#   scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
#   stat_smooth(method = "lm", se = FALSE, size = 1) +
#   theme_bw()

ggplot(z[z$n_sites >0, ], aes(x = -`Pairs`, y = truth)) + 
  stat_binhex(bins = 100) + 
  xlab("Estimated coefficient value") + 
  ylab("\"True\" coefficient value") + 
  stat_hline(yintercept = 0, size = 1/8) + 
  stat_vline(xintercept = 0, size = 1/8) + 
  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
  stat_smooth(method = "lm", se = FALSE, size = 1) +
  theme_bw()
```

Bootstrap resample from each of the three landscape size classes, generating
new sets of 150 landscapes. For each one, calculate the proportion of variance
explained (compared to a null baseline that assumes all species pairs' 
interaction strengths are zero).

```{r, cache = TRUE}
rep = 12
n = 200
boots = replicate(500,
          {
            boot = lapply(
              sizes, 
              function(n){lapply(
                sample.int(max(z$rep), replace = TRUE),
                function(rep){
                  z[z$rep == rep & z$n_sites == n, ]
                }
              ) %>% bind_rows}
            ) %>% bind_rows

            resids = sapply(
              colnames(boot)[-(1:5)],
              function(i){resid(lm(boot$truth ~ boot[[i]] + 0))}
            )
            total_ss = mean(resid(lm(truth ~ 0, data = boot))^2)
            1 - colMeans(resids^2) / total_ss
          }
)
```

Summarize the bootstrap results:

```{r, cache = TRUE}
for(compared in c("partial correlation", "correlation", "GLM", "BayesComm", "partial BayesComm", "Pairs")){
  CI = round(
    quantile(
      boots["Markov network", ] / boots[compared, ], 
      c(.025, .975)), 
    2
  )

  message("R-squared is " , CI[1], "-", CI[2], " times higher than from ", compared, " (95% bootstrap CI)")
}
```

### 2015-08-02 Harris_2015/simulation.Rmd
other things

--- 2015-07-21 Harris_2015/simulation.Rmd
@@
+---
+output: pdf_document
+title: Appendix [[X]]: simulating species interactions
+subtitle: \"Estimating species interactions from observational data with Markov networks\"
+author: David J. Harris
+fontsize: 11pt
+---
+
+This code simulates from a stochastic demographic model, then saves the results
+for subsequent analysis.
+
 ```{r}
-library(progress)
+# Initialize the random number generator for reproducibility
 set.seed(1)
+
+thin = 5E3  # Save every 5000th sample
+n_spp = 20  # 20 species
+n_reps = 50 # 50 replicates for each landscape size
 ```
 
+In the large population limit, the number of individuals of species $i$, $n_i$,
+evolves over time based on the following differential equation, based on 
+multi-species Lotka-Volterra dynamics.  The basic multi-species model is
+
+\begin{equation}
+\frac{\mathrm{d}n_i}{\mathrm{d}t} =  
+    r_i n_i (1 - \frac{n_i + \sum{\alpha_{ij}n_j}}{K}),
+\end{equation}
+
+where $r_i$ is the per-capita growth rate of species $i$, $n_i$ is its
+abundance, $\alpha_{ij}$ describes the strength of its competition with
+species $j$, and $K$ is the carrying capacity.  These simulations use a
+more general form, which includes propagules from outside the system ($r_e$)
+and a separate mechanism for mutualisms ($f_i(\vec{n})$).
+
+\begin{equation}
+\frac{\mathrm{d}n_i}{\mathrm{d}t} =  
+   r_e + r_i n_i (1 - \frac{n_i f_i(\vec{n}) + \sum{\alpha_{ij}n_j}}{K}).
+\end{equation}
+
+Setting $r_e=0$ and $f_i(\vec{n}) = 1$ recovers the original Lotka-Volterra
+dynamics. The function $f_i(\vec{n})$ acts to reduce the strength of
+intraspecific competition in the presence of mutualists.  For these simulations,
+I used
+
+\begin{equation}
+f_i(\vec{n}) = \frac{1}{2}(1 + \mathrm{exp}(-\frac{\sum_j{\gamma_{ij} n_j}}{5K}),
+\end{equation}
+
+where $\gamma_{ij}$ is the strength of the mutualism between species $i$ and
+species $j$.  The nonlinearity ensures that mutualisms can never eliminate more
+than half of the effects of intraspecific competition and destabilize the system
+too strongly. In these simulations, each species pair acted either as competitors
+$\alpha_{ij} > 0$ or as mutualists $\gamma_{ij} > 0$, but not both. Note that
+$\alpha_{ii}$ and $\gamma_{ii}$ were always zero (i.e. intraspecific competition
+is modeled outside of the $\alpha$ term and species cannot be their own
+mutualists).
+
+Equation 2 can be decomposed into a birth rate, $r_e + r_i n_i$, and a death 
+rate, $r_i n_i (\frac{n_i f_i(\vec{n}) + \sum{\alpha_{ij}n_j}}{K})$.  If we
+treat births and deaths as discrete events (e.g. if we leave the large population
+regime), then the probability that the next event will be a birth (or death) of
+species $i$ is directly proportional to the corresponding demographic rate.
+
+Because these events' probabilities only depend on the  current state of the 
+system, they form a Markov chain that explores the  space of possible species 
+assemblages under the population dynamic model.  For each set of competition and
+mutualism coefficients, I simulated births and deaths in this chain, collecting
+one sample after every `r thin`th demographic event.  For the subsequent analysis,
+each species was considered "present" in a  sample if one or more individuals was
+present, and "absent" if its abundance in that sample was zero.
+
+The following function randomly generates a set of $\alpha$ and $\gamma$ values,
+then simulates `maxit` demographic events with those parameters, retaining every
+`thin`=`r thin` samples.
+
 ```{r}
-
 simulate_communities = function(n_spp = 20,
@@
   mutualism_scaler = function(x, mutualism_maximum){
-    (1 + exp(-4 * x / K / n_spp)) * (1 - mutualism_maximum)
+    (1 + exp(- x / 5 / K)) * (1 - mutualism_maximum)
   }
@@
 
-```{r}
-thin = 5E3
-n_spp = 20
-n_reps = 50
+I called the above code `r n_reps` times for each landscape size, saving the
+results in a folder called `fakedata` in two formats (one for Gotelli and 
+Ulrich's Pairs software, and one for the other methods).
 
+```{r, eval=FALSE}
 for(n_sites in c(25, 200, 1600)){
@@
 
-  pb = progress_bar$new(total = n_reps)
   for(rep in 1:n_reps){
@@
     )
-
-    pb$tick()
 

### 2015-08-04 Harris_2015/manuscript-materials/draft.Rmd
other things
almost done

@@
 \renewcommand{\headrulewidth}{0pt}
+
+\setlength{\parskip}{0pt}
 
@@
 
+### Introduction
+
 \setlength{\parindent}{1em}
 \setlength{\parskip}{3pt}
-
-### Introduction
 
@@
 interaction between the two shrub species in Figure 1A can become obscured by
-their shared tendency to occur in unshaded areas (Figure 1B). If this sort of
-effect is common, then the vast majority of ecologists' methods, which rely on
-test statistics describing the total number of co-occurrences or overall
-correlation between putatively interacting species [@diamond_island_1975;
-@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
-@pollock_understanding_2014] will not generally provide much evidence regarding
-species' direct effects on one another. The goal of this paper is to resolve
-this long-standing problem in community ecology.
+their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
+have long known that indirect effects can overwhelm direct ones at the landscape
+level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
+inferenes from observational data do not control for these effects
+[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
+@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
+indirect interactions like those in Figure 1 are generally important
+[@dodson_complementary_1970], existing methods will thus not generally provide
+much evidence regarding species' direct effects on one another. The goal of this
+paper is to resolve this long-standing problem.
 
@@
 
-In this paper, following recent work by @azaele_inferring_2010 and
-@fort_statistical_2013, I show that Markov networks [undirected graphical models
-also known as Markov random fields; @murphy_machine_2012] can provide a
-framework for understanding the landscape-level consequences of pairwise species
-interactions, and  for detecting them from observed presence-absence matrices.
-Markov networks have been used in many scientific fields for decades in similar
-contexts, from physics [where nearby particles interact magnetically with one
-another; @cipra_introduction_1987] to spatial statistics [where adjacent grid
-cells interact with one another; @harris_contact_1974; @gelfand_modelling_2005].
-While community ecologists explored some related approaches in the 1980's
-[@whittam_species_1981], computational limitations had previously imposed severe
-approximations that led to unintelligible results [e.g. "probabilities" greater
-than one; @gilpin_factors_1982].
+Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
+paper shows that Markov networks [undirected graphical models also known as
+Markov random fields; @murphy_machine_2012] can provide a framework for
+understanding the landscape-level consequences of pairwise species interactions,
+and for detecting them from observed presence-absence matrices. Markov networks
+have been used in many scientific fields for decades in similar contexts, from
+physics [where nearby particles interact magnetically; @cipra_introduction_1987]
+to spatial statistics [where adjacent grid cells have correlated values;
+@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
+explored some related approaches in the 1980's [@whittam_species_1981],
+they used severe approximations that led to unintelligible results
+[e.g. "probabilities" greater than one; @gilpin_factors_1982].
 
-The rest of the paper proceeds as follows. First, I discuss how Markov networks
-operate and how they can be used to simulate landscape-level data or to make
-exact predictions regarding the direct and indirect consequences of possible
-interaction matrices. Then, using simulated data sets where the "true"
-ecological structure is known, I compare this approach with several existing
-methods for detecting species interactions. Finally, I discuss opportunities for
-extending the approach presented here to other problems in community ecology,
-such as disentangling the effects of biotic and abiotic interactions on species
-composition.
+Below, I introduce Markov networks and show how they can be used to simulate
+landscape-level data or to make exact predictions about the direct and
+indirect consequences of possible interaction matrices. Then, using simulated
+data sets where the "true" interactions are known, I compare this approach with
+several existing methods. Finally, I discuss opportunities for extending the
+approach presented here to other problems in community ecology, e.g. quantifying
+the overall effect of species interactions on occurrence rates
+[@roughgarden_competition_1983] and disentangling the effects of biotic versus
+abiotic interactions on species composition [@kissling_towards_2012;
+@pollock_understanding_2014].
 
@@
 conditional relationships among species (Figure 1C) and the kinds of species
-assemblages that these relationships produce. In this section, I show how a set
-of conditional relationships can be used to determine the how groups of species
-can co-occur. Methods for estimating conditional relationships from
-presence-absence data are discussed in the next section.
+assemblages that these relationships produce. Here, I show how a set of
+conditional relationships can be used to determine the how groups of species
+can co-occur. Methods for estimating conditional relationships from data are
+discussed in the next section.
 
@@
 might have a value of $+2$ for two mutualists, indicating that the odds of
-observing one species are $e^2$ times higher its partner is present than when
-it is absent. Because the relative probability of a presence-absence vector
-increases when positively-associated species co-occur and decreases when
-negatively-associated species co-occur, the model tends---all else equal---to
-produce assemblages that have many pairs of positively-associated species and
-relatively few pairs of negatively-associated species (exactly as an ecologist
-might expect).
+observing one species are $e^2$ times higher in sites where its partner is
+present than in comparable sites where its partner is absent. Because the
+relative probability of a presence-absence vector increases when
+positively-associated species co-occur and decreases when negatively-associated
+species co-occur, the model tends---all else equal---to produce assemblages that
+have many pairs of positively-associated species and relatively few pairs of
+negatively-associated species (exactly as an ecologist might expect).
 
@@
 that would be expected to produce exactly the observed co-occurrence frequencies
-on average (i.e. it matches the sufficient statistics of the model distribution
-to the sufficient statistics of the data). For the analyses in this paper, I
-used the *rosalia* package [@harris_rosalia_2015] for the R programming language
-[@r_core_team_r_2015] to optimize the Markov network parameters.
+on average [i.e. it matches the sufficient statistics of the model distribution
+to the sufficient statistics of the data; @murphy_machine_2012]. I used the
+rosalia package [@harris_rosalia_2015] for the R programming language
+[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
+named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
+miraculous healing powers played an important rhetorical role in the null model
+debates of the 1970's and 1980's [@lewin_santa_1983].
 
@@
 In order to compare different methods, I simulated two sets of landscapes using
-known parameters.
-
-The first set of simulated landscapes included the three competing species
-shown in Figure 1. For each of 1000 replicates, I generated a landscape with
-100 sites by sampling exactly from a probability distribution defined by the
-interaction coefficients in that figure (Appendix A). Each of the methods
-described below was then evaluated on its ability to correctly infer that the
-two shrub species competed with one another, despite their frequent
-co-occurrence.
+known parameters. The first set included the three competing species shown in
+Figure 1. For each of 1000 replicates, I generated a landscape with 100 sites by
+sampling from a probability distribution defined by the figure's interaction
+coefficients (Appendix 1). Each of the methods described below was then
+evaluated on its ability to correctly infer that the two shrub species competed
+with one another, despite their frequent co-occurrence.
 
 I also simulated a second set of landscapes using a stochastic community model
-based on generalized Lotka-Volterra dynamics, as described in the Appendix. In
+based on generalized Lotka-Volterra dynamics, as described in Appendix 2. In
 these simulations, each species pair was randomly assigned to either compete for
@@
 analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
-sites (50 replicates per landscape size; see Appendix).
+sites (50 replicates per landscape size; see Appendix 2).
 
 #### Recovering species interactions from simulated data.
-I used Markov networks and several other techniques for determining the sign and
-strength of the associations between pairs of species (Appendix B).
+I compared seven techniques for determining the sign and strength of the
+associations between pairs of species from simulated data (Appendix 3).
 
@@
 
-For the Markov networks, I used the rosalia package [@harris_rosalia_2015],
-as described above. For the analyses with 20 species, I added a very weak
-logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2 to
-ensure that the model estimates were always finite. The bias introduced by this
-prior should be small: the 95% credible interval on $\beta$ only requires that
-one species' effect on the odds of observing a different species to be less than
-a factor of 1500 (which is not much of a constraint). The logistic distribution
-was chosen because it is convex and has a similar shape to the Laplace
-distribution used in LASSO regularization (especially in the tails), but unlike
-the Laplace distribution it is differentiable everywhere and does not force any
-estimates to be exactly zero. To confirm that this procedure produced reliable
-estimates, I evaluated it on 50 bootstrap replicates (Appendix).
+First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
+models,  as described above. For the analyses with 20 species, I added a very
+weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
+to ensure that the model estimates were always finite. The bias introduced by
+this prior should be small: the 95% credible interval on $\beta$ only requires
+that one species' effect on the odds of observing a different species to be less
+than a factor of 1500 (which is not much of a constraint). The logistic
+distribution was chosen because it is convex and has a similar shape to the
+Laplace distribution used in LASSO regularization (especially in the tails), but
+unlike the Laplace distribution it is differentiable everywhere and does not
+force any estimates to be exactly zero. To confirm that this procedure produced
+stable estimates, I compared its estimates on 50 bootstrap replicates
+(Appendix 4).
 
@@
 pair of species' data vectors on the landscape [@albrecht_spatial_2001;
-@faisal_inferring_2010]. Because partial correlations are undefined for
+@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
+correlation can be thought of as a computationally efficient approximation to
+the full Markov network model [@loh_structure_2013]. This sort of model is very
+common for estimating relationships among genes and gene products
+[@friedman_sparse_2008]. Because partial correlations are undefined for
 landscapes with perfectly-correlated species pairs, I used a regularized
 estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
-`pcor.shrink` function with the default settings [@schafer_corpcor_2014]. In the
-context of non-Gaussian data, the partial correlation can be thought of as a
-computationally efficient approximation to the full Markov network model
-[@loh_structure_2013]. This sort of model is very common for estimating
-relationships among genes and gene products [@friedman_sparse_2008].
+`pcor.shrink` function with the default settings [@schafer_corpcor_2014].
 
@@
 The last two estimators used the latent correlation matrix estimated by the
-BayesComm package [@golding_bayescomm_2015] in order to evaluate the untested
-claim from recent papers that the correlation coefficients estimated by "joint
-species distribution models" provide an accurate assessment of species’ pairwise
-interactions [@pollock_understanding_2014; see also @harris_generating_2015].  
-In addition to using the posterior mean of the correlation coefficient
-[@pollock_understanding_2014], I also used the posterior mean of the *partial*
-correlation, which might provide a better estimate of species' conditional
-associations.
+BayesComm package [@golding_bayescomm_2015] in order to evaluate the recent
+claim that the correlation coefficients estimated by "joint species distribution
+models" provide an accurate assessment of species’ pairwise interactions
+[@pollock_understanding_2014; see also @harris_generating_2015]. In addition to
+using the posterior mean correlation [@pollock_understanding_2014], I also used
+the posterior mean *partial* correlation, which might be able to control for
+indirect effects.
 
 #### Evaluating model performance.
-For the simulated landscapes based on Figure 1, method evaluation was fairly
-qualitative: any method whose test statistic for the two shrubs indicated a
-negative relationship passed; other methods failed (Appendix). For the null
-model (Pairs), I calculated statistical significance using its $Z$-score. For
-the Markov network, I used the Hessian matrix to generate approximate confidence
-intervals and noted whether these intervals included zero.
+For the simulated landscapes based on Figure 1, I assessed whether each method's
+test statistic indicated a positive or negative relationship between the two
+shrubs (Appendix 1). For the null model (Pairs), I calculated statistical
+significance using its $Z$-score. For the Markov network, I used the Hessian
+matrix to generate approximate confidence intervals and noted whether these
+intervals included zero.
 
-For the larger landscapes, I evaluated the relationship between each method's
-estimates and the "true" interaction strengths among all of the species pairs
-from the simulated landscapes. This kind of evaluation is important because
-ecologists need a consistent way to know how strong species interactions are
-regardless of which species were present in a particular data set or how many
-observations were taken. Because the different methods mostly describe species
-interactions on different scales (e.g. correlations versus $Z$ scores versus
-regression coefficients), I used linear regression through the origin to rescale
-the different estimates produced by each method so that they had a consistent
-interpretation. After rescaling each method’s estimates, I calculated squared
-errors between the scaled interaction estimates and “true” interaction values
-across all the simulated data sets. These squared errors determined the
-proportion of variance explained for different combinations of method and
-landscape size (compared with a baseline model that assumed all interaction
-strengths to be zero).
+I then evaluated the relationship between each method's estimates and the "true"
+interaction strengths among all of the species pairs from the larger simulated
+landscapes. This determined which the methods provide a consistent way to know
+how strong species interactions are---regardless of which species were present
+in a particular data set or how many observations were taken. Because the
+different methods mostly describe species interactions on different scales
+(e.g. correlations versus $Z$ scores versus regression coefficients), I used
+linear regression through the origin to rescale the different estimates produced
+by each method so that they had a consistent interpretation. After rescaling
+each method’s estimates, I calculated squared errors between the scaled
+interaction estimates and “true” interaction values across all the simulated
+data sets. These squared errors determined the proportion of variance explained
+for different combinations of method and landscape size (compared with a
+null model that assumed all interaction strengths to be zero).
 
@@
 randomization-based null model falsely reported positive associations 100% of
-the time. Worse, more than 98% of false conclusions from the null model were
-statistically significant. The partial correlation and Markov network estimates,
-on the other hand, each correctly isolated the direct negative interaction
-between the shrubs from their positive indirect interaction 94% of the time
-(although the confidence intervals overlapped zero in most replicates).
+the time. Worse, more than 98% of these false conclusions were statistically
+significant. The partial correlation and Markov network estimates, on the other
+hand, each correctly isolated the direct negative interaction between the shrubs
+from their positive indirect interaction 94% of the time (although the
+confidence intervals overlapped zero in most replicates).
 
 #### Twenty species.
-The accuracy of the evaluated methods varied substantially, depending on
-the parameters that produced the simulated communities (Figure 3A). In general,
-however, there was a consistent ordering, with the four methods that controlled
-for indirect effects of species interactions performing the best: the Markov
-network explained the largest portion of the variance (35% overall), followed
-by the generalized linear models (30%), partial correlations from the raw
-presence-absence data (28%), and partial correlations from BayesComm, the
-joint species distribution model (26%).  The benefit of choosing the full
-Markov network over the other three methods was largest on the smaller
-landscapes, which are also the ones that are most representative of typical
-analyses in this field [@gotelli_empirical_2009].
+Despite some variability across contexts (Figure 3A), the four methods that
+controlled for indirect effects clearly performed the best: the Markov network
+explained the largest portion of the variance in the "true" interaction
+coefficients (35% overall), followed by the generalized linear models (30%),
+partial correlations from the raw presence-absence data (28%), and partial
+correlations from BayesComm, the joint species distribution model (26%).  The
+benefit of choosing the full Markov network over the other three methods was
+largest on the smaller landscapes, which are also the ones that are most
+representative of typical analyses in this field [@gotelli_empirical_2009].
 
@@
 methods had different $R^2$ values, there was a close mapping among
-their estimates (after controlling for the size of the simulated landscapes;
-Figure 3B).  This suggests that the correlation matrices from joint species
-distribution models and especially the effect sizes from null models only
-contain noisy versions of the same information that could be obtained more
-easily and interpretably by calculating correlation coefficients between
-species' presence-absence vectors.
+their estimates (especially after controlling for the size of the simulated
+landscapes; Figure 3B).  This suggests that the effect sizes from the null model
+(and, to a lesser extent, the correlation matrices from joint species
+distribution models) only contain noisy versions of the same information that
+could be obtained more easily and interpretably by calculating correlation
+coefficients between species' presence-absence vectors.
 
-The bootstrapped results confirmed that the rosalia package's estimates of
-species' conditional relationships were reasonably insensitive to sampling
-variation (Appendix), and that the above ranking of the different methods was
-robust (Appendix).
+Bootstrap resampling indicated that the above ranking of the different methods
+was robust (Appendix 3). In particular, the 95% confidence interval of the
+bootstrap distribution indicated that the Markov network's overall $R^2$ value
+was between 14 and 18 percent higher than the second-most effective
+method (generalized linear models) and between 2.12 and 2.38 times higher than
+could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
+landscape also confirmed that the rosalia package's estimates of species'
+conditional relationships were reasonably robust to sampling variation
+(Appendix 4).
 
@@
 of the most commonly-used methods to mistakenly identifying positive species
-interactions between competing species. They also provide a simple answer to the
-question of how competition should affect a species' overall prevalence, which
-was a major flashpoint for the null model debates in the 1980’s
-[@roughgarden_competition_1983; @strong_ecological_1984]. Equation 1 can be used
-to calculate the expected prevalence of a species in the absence of biotic
-influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012]. Competition's
-effect on prevalence in a Markov network can then be calculated by subtracting
-this value and the observed prevalence (cf Figure 2D). This kind of insight
-would have been difficult to obtain without a generative model that makes
-predictions about the consequences of species interactions; null models (which
-presume *a priori* that interactions do not exist) have no way to make such
-predictions.
+interactions between competing species, and can make precise statements about
+the conditions whre indirect interactions will overwhelm direct ones. They also
+provide a simple answer to the question of how competition should affect a
+species' overall prevalence, which was a major flashpoint for the null model
+debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
+Equation 1 can be used to calculate the expected prevalence of a species in the
+absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
+Competition's effect on prevalence in a Markov network can then be calculated by
+subtracting this value from the observed prevalence (cf Figure 2D). This kind of
+insight would have been difficult to obtain without a generative model that
+makes predictions about the consequences of species interactions; null models
+(which presume *a priori* that interactions do not exist) have no way to make
+such predictions.
 
@@
 alternative stable states. Numerous other extensions are possible: for example,
-Markov networks can be fit with a mix of discrete and continuous variables,
-and likely states of the interaction network can be modeled as a function of
+the states of the interaction network can be modeled as a function of
 the local abiotic environment [@lee_learning_2012], which would provide a
-rigorous and straightforward way to incorporate whole networks of biotic
-interactions into species distribution models. Estimating each species'
-conditional relationship with both biotic and abiotic factors in the same model
-should lead to a better understanding of both kinds of ecological process. There
-are even methods [@whittam_species_1981; @tjelmeland_markov_1998] that would
-allow one species to affect the sign or strength of the relationship between
-two other species, tipping the balance between facilitation and exploitation
-[@bruno_inclusion_2003].
+rigorous and straightforward approach to the difficult and important task of
+incorporating whole networks of biotic interactions into species distribution
+models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
+better understanding of the interplay between biotic and abiotic effects on
+community structure. There are even methods [@whittam_species_1981;
+@tjelmeland_markov_1998] that would allow one species to affect the sign or
+strength of the relationship between two other species, tipping the balance
+between facilitation and exploitation [@bruno_inclusion_2003].
 
 Finally, the results presented here have important implications for
-ecologists' continued use of null models to draw inferences about species
-interactions. Null and neutral models can be very useful for clarifying our
-thinking about the numerical consequences of species' richness and abundance
-patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
-particular null model must be interpreted with care
-[@roughgarden_competition_1983]. Even in small networks with three species, it
-may simply not be possible to implicate individual species pairs or specific
-ecological processes like competition by rejecting a general-purpose null
-[@gotelli_empirical_2009], especially when the test statistic is effectively
-just a correlation coefficient (Figure 3B). Simultaneous estimation of multiple
-ecological parameters seems like a much more promising approach: to the extent
-that the models' relative performance on real data sets is similar to the range
-of results shown in Figure 3A, scientists in this field could often double their
-explanatory power by switching from null models to Markov networks (or increase
-it by nearly as much with linear or generalized linear approximations).
-Regardless of the specific methods ecologists ultimately choose, the most
-important consideration is clearly that ecologists need to control for indirect
-effects and estimate conditional relationships between species, rather than
-marginal ones.
+ecologists' continued use of null models for studying species interactions. Null
+and neutral models can be useful for clarifying our thinking about the numerical
+consequences of species' richness and abundance patterns
+[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
+null model must be interpreted with care [@roughgarden_competition_1983]. Even
+in small networks with three species, it may simply not be possible to implicate
+individual species pairs or specific ecological processes like competition by
+rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
+test statistic is effectively just a correlation coefficient (Figure 3B).
+Simultaneous estimation of multiple ecological parameters seems like a much more
+promising approach: to the extent that the models' relative performance on real
+data sets is similar to the range of results shown in Figure 3A, scientists in
+this field could often double their explanatory power by switching from null
+models to Markov networks (or increase it substantially by with linear or
+generalized linear approximations). Regardless of the methods ecologists
+ultimately choose, controlling for indirect effects could clearly improve our
+understanding of species' direct effects on one another and on community
+structure.
 
@@
 ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
-about 18.5.  **C.** Using the probabilities, we can find the expected
-frequencies of all possible co-occurrence patterns between the two species of
-interest. **D.** If $\beta$ equaled zero (e.g. if the species no longer competed
-for the same resources), then the reduction in competition would allow each
-species to increase its occurrence rate and the co-occurrence deficit would
-be eliminated.
+about 18.5.  **C.** We can find the expected frequencies of all possible
+co-occurrence patterns between the two species of interest. **D.** If
+$\beta_{12}$ equaled zero (e.g. if the species no longer competed for the same
+resources), then the reduction in competition would allow each species to
+increase its occurrence rate and the co-occurrence deficit would be eliminated.
 
@@
 **A.** Proportion of variance in interaction coefficients explained by each
-method across varying numbers of sampled locations.
-**B.** The $Z$-scores produced by the null model (i.e. the Pairs software) for
-each pair of species are very predictable from the correlation between the
-presence-absence vectors of those same species and from the number of sites on
-the landscape.
+versus number of sampled locations.
+**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
+species are very predictable from the correlation between the presence-absence
+vectors of those same species and from the number of sites on the landscape.
+
+\setlength{\parskip}{0pt}
+\setlength{\parindent}{-1em}
+\setlength{\leftskip}{1em}

### 2015-08-04 Harris_2015/manuscript-materials/draft.tex
almost done

@@
 \usepackage[compact]{titlesec}
+\usepackage{fancyhdr}
+\usepackage{hanging}
+\pagestyle{fancy}
 \usepackage{lmodern}
@@
 \usepackage{geometry}
+\usepackage{ifthen}
 \geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
@@
 \begin{flushleft}
-\emph{Running head:} Species interactions in Markov networks
+\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
+\renewcommand{\headrulewidth}{0pt}
 
-\textbf{Title: Estimating species interactions from observational data
-with Markov networks}
+\setlength{\parskip}{0pt}
+
+\textbf{Title:} Estimating species interactions from observational data
+with Markov networks
 
@@
 
-\setlength{\parindent}{1cm} \setlength{\parskip}{6pt}
+\subsubsection{Introduction}\label{introduction}
 
-\subsubsection{Introduction}\label{introduction}
+\setlength{\parindent}{1em} \setlength{\parskip}{3pt}
 
@@
 drivers of community assembly, then ecologists might expect to see their
-influence in our data sets (MacArthur 1958, Diamond 1975). Despite
-decades of work and several major controversies, however (Lewin 1983,
-Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al. 2013),
-existing methods for detecting competition's effects on community
+influence in species composition data (MacArthur 1958, Diamond 1975).
+Despite decades of work and several major controversies, however (Lewin
+1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al.
+2013), existing methods for detecting competition's effects on community
 structure are unreliable (Gotelli and Ulrich 2009). In particular,
@@
 become obscured by their shared tendency to occur in unshaded areas
-(Figure 1B). If this sort of effect is common, then the vast majority of
-ecologists' methods, which rely on test statistics describing the total
-number of co-occurrences or overall correlation between putatively
-interacting species (Diamond 1975, Strong et al. 1984, Gotelli and
-Ulrich 2009, Veech 2013, Pollock et al. 2014) will not generally provide
-much evidence regarding species' direct effects on one another.
+(Figure 1B). While ecologists have long known that indirect effects can
+overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
+the vast majority of our methods for drawing inferenes from
+observational data do not control for these effects (e.g. Diamond 1975,
+Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
+2014). To the extent that indirect interactions like those in Figure 1
+are generally important (Dodson 1970), existing methods will thus not
+generally provide much evidence regarding species' direct effects on one
+another. The goal of this paper is to resolve this long-standing
+problem.
 
 While competition doesn't reliably leave the expected pattern in
-community structure (a low co-occurrence rate at the landscape level),
-it nevertheless does leave a signal in the data (Figure 1C).
-Specifically, \emph{among shaded sites}, there will be a deficit of
-co-occurrences between the competing shrub species, and \emph{among
-unshaded sites}, there will also be such a deficit. More generally, we
-can obtain much better estimates of the association between two species
-from their conditional relationships (i.e.~by controlling for other
-species in the network) than we could get from their overall
-co-occurrence rates. This kind of precision is difficult to obtain from
-null models, which begin with the assumptions that all the pairwise
-interactions are zero and thus don't need to be controlled for.
-Nevertheless, null models have dominated this field for more than three
-decades (Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009).
+community structure (low co-occurrence rates at the landscape level), it
+nevertheless does leave a signal in the data (Figure 1C). Specifically,
+\emph{among shaded sites}, there will be a deficit of co-occurrences
+between the competing shrub species, and \emph{among unshaded sites},
+there will also be such a deficit. More generally, we can obtain much
+better estimates of the association between two species from their
+conditional relationships (i.e.~by controlling for other species in the
+network) than we could get from their overall co-occurrence rates. This
+kind of precision is difficult to obtain from null models, which begin
+with the assumptions that all the pairwise interactions are zero and
+thus don't need to be controlled for. Nevertheless, null models have
+dominated this field for more than three decades (Strong et al. 1984,
+Gotelli and Ulrich 2009).
 
-In this paper, following recent work by Azaele et al. (2010) and Fort
-(2013), I show that Markov networks (undirected graphical models also
-known as Markov random fields; Murphy 2012) can provide a framework for
+Following recent work by Azaele et al. (2010) and Fort (2013), this
+paper shows that Markov networks (undirected graphical models also known
+as Markov random fields; Murphy 2012) can provide a framework for
 understanding the landscape-level consequences of pairwise species
@@
 decades in similar contexts, from physics (where nearby particles
-interact magnetically with one another; Cipra 1987) to spatial
-statistics (where adjacent grid cells interact with one another; Harris
-1974, Gelfand et al. 2005). While community ecologists explored some
-related approaches in the 1980's (Whittam and Siegel-Causey 1981),
-computational limitations had previously imposed severe approximations
-that produced unintelligible results (e.g. ``probabilities'' greater
-than one; Gilpin and Diamond 1982). Now that it is computationally
-feasible to fit these models exactly, this kind of approach is worth a
-second look.
+interact magnetically; Cipra 1987) to spatial statistics (where adjacent
+grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
+While community ecologists explored some related approaches in the
+1980's (Whittam and Siegel-Causey 1981), they used severe approximations
+that led to unintelligible results (e.g. ``probabilities'' greater than
+one; Gilpin and Diamond 1982).
 
-The rest of the paper proceeds as follows. First, I discuss how Markov
-networks operate and how they can be used to simulate landscape-level
-data or to predict the direct and indirect consequences of possible
-interaction matrices. Then, using simulated data sets where the ``true''
-ecological structure is known, I compare this approach with several
-existing methods for detecting species interactions. Finally, I discuss
-opportunities for extending the approach presented here to larger
-problems in community ecology.
+Below, I introduce Markov networks and show how they can be used to
+simulate landscape-level data or to make exact predictions about the
+direct and indirect consequences of possible interaction matrices. Then,
+using simulated data sets where the ``true'' interactions are known, I
+compare this approach with several existing methods. Finally, I discuss
+opportunities for extending the approach presented here to other
+problems in community ecology, e.g.~quantifying the overall effect of
+species interactions on occurrence rates (Roughgarden 1983) and
+disentangling the effects of biotic versus abiotic interactions on
+species composition (Kissling et al. 2012, Pollock et al. 2014).
 
@@
 
-\paragraph{Conditional relationships and Markov
-networks.}\label{conditional-relationships-and-markov-networks.}
+\paragraph{Markov networks.}\label{markov-networks.}
 
-Ecologists are often interested in learning about direct interactions
-between species, controlling for the indirect influence of other
-species. In statistical terms, this implies that ecologists want to
-estimate \emph{conditional} (``all-else-equal'') relationships, rather
-than \emph{marginal} (``overall'') relationships. The example with the
-shrubs and trees in Figure 1 shows how the two measures can have
-opposite signs, and suggests that conditional relationships are more
-relevant for drawing inferences about species interactions
-(e.g.~competition).
+Markov networks provide a framework for translating back and forth
+between the conditional relationships among species (Figure 1C) and the
+kinds of species assemblages that these relationships produce. Here, I
+show how a set of conditional relationships can be used to determine the
+how groups of species can co-occur. Methods for estimating conditional
+relationships from data are discussed in the next section.
 
-Markov networks define species interactions in conditional terms, and
-use these conditional relationships to determine whether a given
-combination of co-occurring species is consistent with their
-interactions. Given a set of binary outcomes for each species (i.e.~0
-for absence and 1 for presence), the Markov network defines the relative
-probability of observing a given presence-absence vector, \(\vec{y}\),
+A Markov network defines the relative probability of observing a given
+vector of species-level presences (1s) and absences (0s), \(\vec{y}\),
 as
@@
 \centering
+
 \(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)
+
 \raggedright
+\setlength{\parindent}{1em}
 
@@
 species \(i\) and species \(j\) contributes to the log-probability; it
-controls the probability that the two species will be found together
-(Figure 2A, Figure 2B). Because the relative probability of a
+controls the conditional relationship between two species, i.e.~the
+probability that they will be found together, after controlling for the
+other species in the network (Figure 2A, Figure 2B). For example,
+\(\beta_{ij}\) might have a value of \(+2\) for two mutualists,
+indicating that the odds of observing one species are \(e^2\) times
+higher in sites where its partner is present than in comparable sites
+where its partner is absent. Because the relative probability of a
 presence-absence vector increases when positively-associated species
 co-occur and decreases when negatively-associated species co-occur, the
-model tends to produce assemblages that have many pairs of
-positively-associated species and relatively few pairs of
+model tends---all else equal---to produce assemblages that have many
+pairs of positively-associated species and relatively few pairs of
 negatively-associated species (exactly as an ecologist might expect).
-
-A major benefit of Markov networks is the fact that the conditional
-relationships between species can be read directly off the matrix of
-\(\beta\) coefficients (Murphy 2012). For example, if the coefficient
-linking two mutualist species is \(+2\), then---all else equal---the
-odds of observing either species increase by a factor of \(e^2\) when
-its partner is present (Murphy 2012). In this way, the \(\beta\)
-coefficients behave very similarly to the coefficients in a logistic
-regression model (Lee and Hastie 2012).
 
@@
 competitor), then species' marginal association rates can differ from
-this expectation. For this reason, it is important to consider how
-coefficients' effects propagate through the network, as discussed below.
-
-Estimating the marginal relationships predicted by a Markov network is
-more difficult than estimating conditional relationships, because doing
-so requires absolute probability estimates. Turning the relative
-probability given by Equation 1 into an absolute probability entails
-scaling by a normalizing constant, \(Z\), so that that the probabilities
-of all possible assemblages will sum to one (bottom of Figure 2B).
-Calculating \(Z\) exactly, as is done in these analyses, quickly becomes
-infeasible as the number of species increases: with \(2^N\) possible
-assemblages of \(N\) species, the number of bookkeeping operations
-required spirals exponentially into the billions and beyond. Numerous
-techniques are available for keeping the calculations tractable,
-e.g.~via analytic approximations (Lee and Hastie 2012) or Monte Carlo
-sampling (Salakhutdinov 2008), but they are beyond the scope of this
-paper. Alternatively, as noted below, some common linear and generalized
-linear methods can also be used as computationally efficient
-approximations to the full network.
+this expectation. Determining the marginal relationships between species
+from their conditional interactions entails summing over the different
+possible assemblages (Figure 2B). This becomes intractable when the
+number of possible assemblages is large, though several methods beyond
+the scope of this paper can be employed to keep the calculations
+feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
+noted below, some common linear and generalized linear methods can also
+be used as computationally efficient approximations to the full network
+(Lee and Hastie 2012, Loh and Wainwright 2013).
 
@@
 
-In the previous two sections, the values of \(\alpha\) and \(\beta\)
-were known. In practice, however, ecologists will often need to estimate
-these parameters from co-occurrence data (i.e.~from a matrix of ones and
-zeros indicating which species are present at which sites). When the
-number of species is reasonably small, one can compute exact maximum
-likelihood estimates for all of the \(\alpha\) and \(\beta\)
-coefficients given a presence-absence matrix by optimizing
+In the previous section, the values of \(\alpha\) and \(\beta\) were
+known and the goal was to make predictions about possible species
+assemblages. In practice, however, ecologists will often need to
+estimate the parameters from an observed co-occurrence matrix (i.e.~from
+a matrix of ones and zeros indicating which species are present at which
+sites). When the number of species is reasonably small, one can compute
+exact maximum likelihood estimates for all of the \(\alpha\) and
+\(\beta\) coefficients given a presence-absence matrix by optimizing
 \(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
@@
 \(\beta\) coefficients that would be expected to produce exactly the
-observed co-occurrence frequencies (i.e.~it matches the sufficient
-statistics of the model distribution to the sufficient statistics of the
-data). For the analyses in this paper, I used the \emph{rosalia} package
-(Harris 2015a) for the R programming language (R Core Team 2015) to
-define the objective function and gradient in R code and then find the
-optimum.
+observed co-occurrence frequencies on average (i.e.~it matches the
+sufficient statistics of the model distribution to the sufficient
+statistics of the data; Murphy 2012). I used the rosalia package (Harris
+2015a) for the R programming language (R Core Team 2015) to optimize the
+Markov network parameters. The package was named after Santa Rosalia,
+the patron saint of biodiversity, whose supposedly miraculous healing
+powers played an important rhetorical role in the null model debates of
+the 1970's and 1980's (Lewin 1983).
 
@@
 In order to compare different methods, I simulated two sets of
-landscapes using known parameters.
-
-The first set of simulated landscapes included the three competing
-species shown in Figure 1. For each of 1000 replicates, I generated a
-landscape with 100 sites by sampling exactly from a probability
-distribution defined by the interaction coefficients in that figure
-(Appendix A). Each of the methods described below was then evaluated on
-its ability to correctly infer that the two shrub species competed with
-one another, despite their frequent co-occurrence.
+landscapes using known parameters. The first set included the three
+competing species shown in Figure 1. For each of 1000 replicates, I
+generated a landscape with 100 sites by sampling from a probability
+distribution defined by the figure's interaction coefficients (Appendix
+1). Each of the methods described below was then evaluated on its
+ability to correctly infer that the two shrub species competed with one
+another, despite their frequent co-occurrence.
 
 I also simulated a second set of landscapes using a stochastic community
-model based on generalized Lotka-Volterra dynamics, as described in the
-Appendix. In these simulations, each species pair was randomly assigned
-to either compete for a portion of the available carrying capacity
-(negative interaction) or to act as mutualists (positive interaction).
-In these simulations, mutualism operated by mitigating the effects of
+model based on generalized Lotka-Volterra dynamics, as described in
+Appendix 2. In these simulations, each species pair was randomly
+assigned to either compete for a portion of the available carrying
+capacity (negative interaction) or to act as mutualists (positive
+interaction). Here, mutualisms operate by mitigating the effects of
 intraspecific competition on each partner's death rate. For these
 analyses, I simulated landscapes with up to 20 species and 25, 200, or
-1600 sites (50 replicates per landscape size; see Appendix).
+1600 sites (50 replicates per landscape size; see Appendix 2).
 
@@
 
-I used Markov networks and several other techniques for determining the
-sign and strength of the associations between pairs of species (Appendix
-B).
+I compared seven techniques for determining the sign and strength of the
+associations between pairs of species from simulated data (Appendix 3).
 
-For the Markov networks, I used the rosalia package (Harris 2015a), as
-described above. Given the large number of parameters associated with
-some of the networks to be estimated, I regularized the likelihood using
-a logistic prior distribution with a scale of 2 on the \(\alpha\) and
-\(\beta\) terms. This convex prior distribution has a similar shape to
-the Laplace distribution used in LASSO regularization (especially in the
-tails), but does not force the estimates to be exactly zero.
+First, I used the rosalia package (Harris 2015a) to fit Markov newtork
+models, as described above. For the analyses with 20 species, I added a
+very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
+terms with scale 2 to ensure that the model estimates were always
+finite. The bias introduced by this prior should be small: the 95\%
+credible interval on \(\beta\) only requires that one species' effect on
+the odds of observing a different species to be less than a factor of
+1500 (which is not much of a constraint). The logistic distribution was
+chosen because it is convex and has a similar shape to the Laplace
+distribution used in LASSO regularization (especially in the tails), but
+unlike the Laplace distribution it is differentiable everywhere and does
+not force any estimates to be exactly zero. To confirm that this
+procedure produced stable estimates, I compared its estimates on 50
+bootstrap replicates (Appendix 4).
 
 I also evaluated five alternative methods from the existing literature
-to calibrate the performance of the Markov network as a method for
-estimating species interactions. The first two alternative interaction
-measures were the sample correlations and the partial correlations
-between each pair of species' data vectors on the landscape (Albrecht
-and Gotelli 2001, Faisal et al. 2010). Because partial correlations are
-undefined for landscapes with perfectly-correlated species pairs, I used
-a regularized estimate based on James-Stein shrinkage, as implemented in
-the corpcor package's \texttt{pcor.shrink} function with the default
-settings ({\textbf{???}}). In the context of non-Gaussian data, the
-partial correlation can be thought of as a computationally efficient
-approximation to the full Markov network model (Loh and Wainwright
-2013).
+(plus a novel variant of the fifth method). The first two alternative
+interaction measures were the sample correlations and the partial
+correlations between each pair of species' data vectors on the landscape
+(Albrecht and Gotelli 2001, Faisal et al. 2010). In the context of
+non-Gaussian data, the partial correlation can be thought of as a
+computationally efficient approximation to the full Markov network model
+(Loh and Wainwright 2013). This sort of model is very common for
+estimating relationships among genes and gene products (Friedman et al.
+2008). Because partial correlations are undefined for landscapes with
+perfectly-correlated species pairs, I used a regularized estimate based
+on James-Stein shrinkage, as implemented in the corpcor package's
+\texttt{pcor.shrink} function with the default settings (Schäfer et al.
+2014).
 
@@
 network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
-regularized logistic regression models for each species, using the other
-species on the landscape as predictors. {[}{[}Observational
-presence-absence data does not have enough degrees of freedom to
-estimate this many parameters, so I symmetrized the pairwise
-correlations by averaging the coefficient for predicting species \(i\)'s
-status from species \(j\) and vice versa.{]}{]}
+regularized logistic regression models (Gelman et al. 2008) for each
+species, using the other species on the landscape as predictors. To
+avoid the identifiability problems associated with directed cyclic
+graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
+within species pairs via averaging.
 
 The fourth alternative method, described in Gotelli and Ulrich (2009),
-involved simulating possible landscapes from a null model that retains
-the row and column sums of the original matrix (Strong et al. 1984).
-Using the default options in the Pairs software described in Gotelli and
-Ulrich (2009), I simulated the null distribution of scaled C-scores (a
-test statistic describing the number of \emph{non}-co-occurrences
-between two species). The software then calculated a \(Z\) statistic for
-each species pair using this null distribution. After multiplying this
-statistic by \(-1\) so that positive values corresponded to facilitation
-and negative values corresponded to competition, I used it as another
-estimate of species interactions.
+involved simulating new landscapes from a null model that retains the
+row and column sums of the original matrix (Strong et al. 1984). I used
+the \(Z\)-scores computed by the Pairs software described in Gotelli and
+Ulrich (2009) as my null model-based estimator of species interactions.
 
-The last two methods used the latent correlation matrix estimated by the
-BayesComm package ({\textbf{???}}) to evaluate the claim from recent
-papers that the correlation coefficients estimated by ``joint species
-distribution models'' provide an accurate assessment of species'
-pairwise interactions (Pollock et al. 2014, see also Harris 2015b). I
-reported both the average correlation between each pair of species'
-latent variables across 1000 Monte Carlo samples, as well as the average
-\emph{partial} correlation.
+The last two estimators used the latent correlation matrix estimated by
+the BayesComm package (Golding and Harris 2015) in order to evaluate the
+recent claim that the correlation coefficients estimated by ``joint
+species distribution models'' provide an accurate assessment of species'
+pairwise interactions (Pollock et al. 2014, see also Harris 2015b). In
+addition to using the posterior mean correlation (Pollock et al. 2014),
+I also used the posterior mean \emph{partial} correlation, which might
+be able to control for indirect effects.
 
@@
 
-For the simulated landscapes based on Figure 1, method evaluation was
-fairly qualitative: any method whose test statistic for the two shrubs
-indicated a negative relationship passed; other methods failed. For
-methods that provide confidence intervals or other inferential
-statistics (Appendix), I also reported how often they rejected the null
-hypothesis of no interaction.
+For the simulated landscapes based on Figure 1, I assessed whether each
+method's test statistic indicated a positive or negative relationship
+between the two shrubs (Appendix 1). For the null model (Pairs), I
+calculated statistical significance using its \(Z\)-score. For the
+Markov network, I used the Hessian matrix to generate approximate
+confidence intervals and noted whether these intervals included zero.
 
-Because the different methods mostly describe species interactions on
-different scales (e.g.~correlations versus \(Z\) scores versus
-regression coefficients), I used linear regression through the origin to
-rescale the different estimates produced by each method so that they had
-a consistent interpretation. After rescaling each method's estimates, I
-calculated squared errors between the scaled interaction estimates and
-``true'' interaction values for each of the {[}{[}N{]}{]} species pairs
-across all the simulated data sets. These squared errors determined the
-proportion of variance explained for different combinations of method
-and landscape size (compared with a baseline model that assumed all
-interaction strengths to be zero).
+I then evaluated the relationship between each method's estimates and
+the ``true'' interaction strengths among all of the species pairs from
+the larger simulated landscapes. This determined which the methods
+provide a consistent way to know how strong species interactions
+are---regardless of which species were present in a particular data set
+or how many observations were taken. Because the different methods
+mostly describe species interactions on different scales
+(e.g.~correlations versus \(Z\) scores versus regression coefficients),
+I used linear regression through the origin to rescale the different
+estimates produced by each method so that they had a consistent
+interpretation. After rescaling each method's estimates, I calculated
+squared errors between the scaled interaction estimates and ``true''
+interaction values across all the simulated data sets. These squared
+errors determined the proportion of variance explained for different
+combinations of method and landscape size (compared with a null model
+that assumed all interaction strengths to be zero).
 
@@
 mechanistic level---due to indirect effects of the dominant tree
-species. As a result, the covariance method falsely reported positive
-associations for 94\% of the simulated landscapes, and the
-randomization-based null model falsely reported such associations 100\%
-of the time. The methods for evaluating conditional relationships
-(Markov networks, GLMs, and both types of partial covariance estimates),
-however, successfully controlled for the indirect pathway via the tree
-species and each correctly identified the direct negative interaction
-between the shrubs 94\% of the time.
+species. As a result, the correlation between these species was positive
+in 94\% of replicates, and the randomization-based null model falsely
+reported positive associations 100\% of the time. Worse, more than 98\%
+of these false conclusions were statistically significant. The partial
+correlation and Markov network estimates, on the other hand, each
+correctly isolated the direct negative interaction between the shrubs
+from their positive indirect interaction 94\% of the time (although the
+confidence intervals overlapped zero in most replicates).
 
@@
 
-The accuracy of the four evaluated methods varied substantially,
-depending on the parameters that produced the simulated communities
-(Figure 3). In general, however, there was a consistent ordering\ldots{}
+Despite some variability across contexts (Figure 3A), the four methods
+that controlled for indirect effects clearly performed the best: the
+Markov network explained the largest portion of the variance in the
+``true'' interaction coefficients (35\% overall), followed by the
+generalized linear models (30\%), partial correlations from the raw
+presence-absence data (28\%), and partial correlations from BayesComm,
+the joint species distribution model (26\%). The benefit of choosing the
+full Markov network over the other three methods was largest on the
+smaller landscapes, which are also the ones that are most representative
+of typical analyses in this field (Gotelli and Ulrich 2009).
 
-The three methods that estimated marginal relationships among species
-and did not explicitly control for indirect species interactions (the
-null model, the joint species distribution model's correlation
-coefficients, and the sample correlation coefficients) differed from one
-another in their \(R^2\) values, but their estimates were surprisingly
-similar to one another. In particular, the correlation coefficients were
-extremely similar to the estimates from the other two methods
-(Spearman's rho = X for the joint species distribution model's
-correlation matrix and Y for the null model).
+The three methods that did not attempt to control for indirect
+interactions all explained less than 20\% of the variance. Of these, the
+sample correlation matrix based on the raw data performed the best
+(19\%), followed by the null model (15\%) and BayesComm's correlation
+matrix (11\%). Although these last three methods had different \(R^2\)
+values, there was a close mapping among their estimates (especially
+after controlling for the size of the simulated landscapes; Figure 3B).
+This suggests that the effect sizes from the null model (and, to a
+lesser extent, the correlation matrices from joint species distribution
+models) only contain noisy versions of the same information that could
+be obtained more easily and interpretably by calculating correlation
+coefficients between species' presence-absence vectors.
+
+Bootstrap resampling indicated that the above ranking of the different
+methods was robust (Appendix 3). In particular, the 95\% confidence
+interval of the bootstrap distribution indicated that the Markov
+network's overall \(R^2\) value was between 14 and 18 percent higher
+than the second-most effective method (generalized linear models) and
+between 2.12 and 2.38 times higher than could be achieved by the null
+model (Pairs). Bootstrap resampling of a 200-site landscape also
+confirmed that the rosalia package's estimates of species' conditional
+relationships were reasonably robust to sampling variation (Appendix 4).
 
@@
 nature of the pairwise interactions (e.g.~which species in a
-positively-associated pair is facilitating the other), particularly when
-real pairs of species can reciprocally influence one another in multiple
-ways simultaneously (Bruno et al. 2003); with compositional data, there
-is only enough information to provide a single number describing each
-species pair.\\On the other hand, if ecologists are familiar enough with
-the natural history of their study systems that they can augment the
-likelihood function with a prior distribution or a suitable penalty,
-then this information could reduce the effective degrees of freedom to
-estimate and real-world results might be even better than those shown in
-Figure 3.
+positively-associated pair is facilitating the other; Schmidt and Murphy
+2012), particularly when real pairs of species can reciprocally
+influence one another in multiple ways simultaneously (Bruno et al.
+2003); with compositional data, there is only enough information to
+provide a single number describing each species pair. To estimate
+asymmetric interactions, such as commensalism or predation, ecologists
+would need other kinds of data, as from time series, behavioral
+observations, manipulative experiments, or natural history. These other
+sources of information could also be used to augment the likelihood
+function with an informative prior distribution, which could lead to
+better results on some real data sets than was shown in Figure 3A.
 
-Despite these limitations, Markov networks have enormous potential to
-improve ecological understanding. For example, Markov networks provide a
-simple answer to the question of how competition should affect a
-species' overall prevalence, which was a major flashpoint for the null
-model debates in the 1980's (Roughgarden 1983, Strong et al. 1984).
-Equation 1 can be used to calculate the expected prevalence of a species
-in the absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee
-and Hastie 2012). Competition's effect on prevalence in a Markov network
-can then be calculated by subtracting this value and the observed
-prevalence (cf Figure 2D).
+Despite their limitations, Markov networks have enormous potential to
+improve ecological understanding. In particular, they are less
+vulnerable than some of the most commonly-used methods to mistakenly
+identifying positive species interactions between competing species, and
+can make precise statements about the conditions whre indirect
+interactions will overwhelm direct ones. They also provide a simple
+answer to the question of how competition should affect a species'
+overall prevalence, which was a major flashpoint for the null model
+debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
+can be used to calculate the expected prevalence of a species in the
+absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and
+Hastie 2012). Competition's effect on prevalence in a Markov network can
+then be calculated by subtracting this value from the observed
+prevalence (cf Figure 2D). This kind of insight would have been
+difficult to obtain without a generative model that makes predictions
+about the consequences of species interactions; null models (which
+presume \emph{a priori} that interactions do not exist) have no way to
+make such predictions.
 
@@
 (Cipra 1987), and the models' properties, capabilities, and limits are
-well-understood in a huge range of applications. Modeling species
-interactions using the same framework would thus allow ecologists to tap
+well-understood in a huge range of applications. Using the same
+framework for species interactions would thus allow ecologists to tap
 into an enormous set of existing discoveries and techniques for dealing
 with indirect effects, stability, and alternative stable states.
-Numerous other extensions are possible: Markov networks can be fit with
-a mix of discrete and continuous variables, for example (Lee and Hastie
-2012). There are even methods (Whittam and Siegel-Causey 1981,
-Tjelmeland and Besag 1998) that would allow the coefficient linking two
-species in an interaction matrix to vary as a function of the abiotic
-environment or of third-party species that tip the balance between
-facilitation and exploitation (Bruno et al. 2003).
+Numerous other extensions are possible: for example, the states of the
+interaction network can be modeled as a function of the local abiotic
+environment (Lee and Hastie 2012), which would provide a rigorous and
+straightforward approach to the difficult and important task of
+incorporating whole networks of biotic interactions into species
+distribution models (Kissling et al. 2012, Pollock et al. 2014), leading
+to a better understanding of the interplay between biotic and abiotic
+effects on community structure. There are even methods (Whittam and
+Siegel-Causey 1981, Tjelmeland and Besag 1998) that would allow one
+species to affect the sign or strength of the relationship between two
+other species, tipping the balance between facilitation and exploitation
+(Bruno et al. 2003).
 
 Finally, the results presented here have important implications for
-ecologists' continued use of null models to draw inferences about
-species interactions. Null and neutral models can be very useful for
-clarifying our thinking about the numerical consequences of species'
-richness and abundance patterns (Harris et al. 2011, Xiao et al. 2015),
-but deviations from a null model must be interpreted with care
-(Roughgarden 1983). In complex networks of ecological interactions (and
-even in small networks with three species), it may simply not be
-possible to implicate individual species pairs or specific ecological
-processes like competition by rejecting a general-purpose null (Gotelli
-and Ulrich 2009). Moreover, the null model's estimates were extremely
-similar to what could be obtained from a simple correlation matrix,
-which raises questions about how much additional value the null model
-adds. Estimating pairwise coefficients directly seems like a much more
+ecologists' continued use of null models for studying species
+interactions. Null and neutral models can be useful for clarifying our
+thinking about the numerical consequences of species' richness and
+abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
+deviations from a particular null model must be interpreted with care
+(Roughgarden 1983). Even in small networks with three species, it may
+simply not be possible to implicate individual species pairs or specific
+ecological processes like competition by rejecting a general-purpose
+null (Gotelli and Ulrich 2009), especially when the test statistic is
+effectively just a correlation coefficient (Figure 3B). Simultaneous
+estimation of multiple ecological parameters seems like a much more
 promising approach: to the extent that the models' relative performance
-on real data sets is similar to the range of results shown in Figure 3,
-scientists in this field could {[}{[}easily double their explanatory
-power by switching from null models to partial correlations or
-generalized linear models, or triple it by switching to a Markov
-network{]}{]}. Regardless of the specific methods ecologists ultimately
-choose, the most important consideration is clearly that we need to
-control for indirect effects and estimate conditional relationships
-between species, rather than marginal ones.
+on real data sets is similar to the range of results shown in Figure 3A,
+scientists in this field could often double their explanatory power by
+switching from null models to Markov networks (or increase it
+substantially by with linear or generalized linear approximations).
+Regardless of the methods ecologists ultimately choose, controlling for
+indirect effects could clearly improve our understanding of species'
+direct effects on one another and on community structure.
 
@@
 S. Tysor. Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li,
-D. L. Miller, N. Golding, N. J. Gotelli, and C. F. Dormann provided
-useful feedback on earlier drafts of this work.
+D. L. Miller, N. Golding, N. J. Gotelli, C. F. Dormann, and two
+anonymous reviewers provided very helpful feedback on the text.
 
-\setlength{\parindent}{0cm} \setlength{\parskip}{4pt}
+\setlength{\parindent}{0cm}
 
@@
 shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, as indicated
-by their negative coefficient (circled), but this effect is
-substantially weaker. \textbf{B.} In spite of the competitive
+them. The two shrub species also compete with one another, but more
+weakly (circled coefficient). \textbf{B.} In spite of the competitive
 interactions between the two shrub species, their shared tendency to
@@
 species' presence with a conditional method such as a partial covariance
-or a Markov network allows us to correctly identify the negative
-interaction between these two species (circled).
+or a Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled).
 
 \textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
-The depicted abiotic environment favors the occurrence of both species
+The abiotic environment favors the occurrence of both species
 (\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
@@
 constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} Using the probabilities,
-we can find the expected frequencies of all possible co-occurrence
-patterns between the two species of interest. \textbf{D.} If \(\beta\)
-equaled zero (e.g.~if the species no longer competed for the same
-resources), then the reduction in competition would allow each species
-to increase its occurrence rate and the deficit of co-occurrences would
-be eliminated.
+In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
+the species no longer competed for the same resources), then the
+reduction in competition would allow each species to increase its
+occurrence rate and the co-occurrence deficit would be eliminated.
 
-\textbf{Figure 3.} Proportion of variance in interaction coefficients
-explained by each method across varying numbers of sampled locations.
+\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
+coefficients explained by each versus number of sampled locations.
+\textbf{B.} The \(Z\)-scores produced by the null model (``Pairs'') for
+each pair of species are very predictable from the correlation between
+the presence-absence vectors of those same species and from the number
+of sites on the landscape.
+
+\setlength{\parskip}{0pt} \setlength{\parindent}{-1em}
+\setlength{\leftskip}{1em}
 
@@
 
+Dodson, S. I. 1970. COMPLEMENTARY FEEDING NICHES SUSTAINED BY
+SIZE-SELECTIVE PREDATION. Limnology and Oceanography 15:131--137.
+
 Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
@@
 
+Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
+covariance estimation with the graphical lasso. Biostatistics
+9:432--441.
+
 Gelfand, A. E., A. M. Schmidt, S. Wu, J. A. Silander, A. Latimer, and A.
@@
 
+Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
+Informative Default Prior Distribution for Logistic and Other Regression
+Models. The Annals of Applied Statistics 2:1360--1383.
+
 Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
@@
 52:75--84.
+
+Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian Community
+Ecology Analysis.
 
@@
 
+Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
+J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
+Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
+Towards novel approaches to modelling biotic interactions in
+multispecies assemblages at large spatial extents. Journal of
+Biogeography 39:2163--2178.
+
 Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
 Models.
+
+Levine, S. H. 1976. Competitive Interactions in Ecosystems. The American
+Naturalist 110:903--910.
 
@@
 
+Schäfer, J., R. Opgen-Rhein, V. Zuber, M. Ahdesmäki, A. P. D. Silva, and
+K. Strimmer. 2014. Corpcor: Efficient Estimation of Covariance and
+(Partial) Correlation.
+
+Schmidt, M., and K. Murphy. 2012. Modeling Discrete Interventional Data
+using Directed Cyclic Graphical Models. arXiv preprint arXiv:1205.2617.
+
 Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.

### 2015-08-04 Harris_2015/recovery.Rmd
almost done

@@
     keep_tex: true
-title: Appendix [[X]]":"" estimating species interactions
+title: "Appendix 3: Estimating species interactions"
 subtitle: \"Estimating species interactions from observational data with Markov networks\"
@@
 This document describes how the different models were fit to the simulated data 
-from Appendix [[XX]] and how each model's performance was evaluated.[^1]
+from Appendix 2 and how each model's performance was evaluated.[^1]
 
@@
   ######## BayesComm and partial BayesComm ########
-  bc = BC(Y = x, model = "community", its = 10)
+  bc = BC(Y = x, model = "community", its = 1000)
 
@@
 
-# ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) + 
-#   stat_binhex(bins = 100) + 
-#   xlab("Estimated coefficient value") + 
-#   ylab("\"True\" coefficient value") + 
-#   stat_hline(yintercept = 0, size = 1/8) + 
-#   stat_vline(xintercept = 0, size = 1/8) + 
-#   scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
-#   stat_smooth(method = "lm", se = FALSE, size = 1) +
-#   theme_bw()
-
-ggplot(z[z$n_sites >0, ], aes(x = -`Pairs`, y = truth)) + 
+ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) + 
   stat_binhex(bins = 100) + 
@@
   scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
-  stat_smooth(method = "lm", se = FALSE, size = 1) +
-  theme_bw()
+  theme_bw() + 
+  ggtitle("Markov network") + 
+  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$`Markov network` + 0)))
+
+ggplot(z[z$n_sites >0, ], aes(x = `Pairs`, y = truth)) + 
+  stat_binhex(bins = 100) + 
+  xlab("Estimated coefficient value") + 
+  ylab("\"True\" coefficient value") + 
+  stat_hline(yintercept = 0, size = 1/8) + 
+  stat_vline(xintercept = 0, size = 1/8) + 
+  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
+  theme_bw() + 
+  ggtitle("Null model (\"Pairs\")") + 
+  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$Pairs + 0)))
 ```

### 2015-08-04 Harris_2015/recovery.tex
almost done

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={David J. Harris},
            pdftitle={Appendix 3: Estimating species interactions},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Appendix 3: Estimating species interactions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
\subtitle{"Estimating species interactions from observational data with Markov
networks"}
  \author{David J. Harris}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

\begin{document}

\maketitle

This document describes how the different models were fit to the
simulated data from Appendix 2 and how each model's performance was
evaluated.\footnote{The PDF version of this document has been manually
  altered to omit 150 lines of output from the \texttt{corpcor} package
  of the form
  ``\texttt{\#\#\ Estimating\ optimal\ shrinkage\ intensity\ lambda\ (correlation\ matrix):\ 0.0326}''}

Initialization:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(dplyr)        }\CommentTok{# For manipulating data structures}
\KeywordTok{library}\NormalTok{(corpcor)      }\CommentTok{# For regularized partial covariances}
\KeywordTok{library}\NormalTok{(rosalia)      }\CommentTok{# For Markov networks}
\KeywordTok{library}\NormalTok{(arm)          }\CommentTok{# For regularized logistic regression}
\KeywordTok{library}\NormalTok{(BayesComm)    }\CommentTok{# For joint species distribution modeling}
\KeywordTok{library}\NormalTok{(RColorBrewer) }\CommentTok{# For color palette}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Load in the results from the \texttt{pairs} program, run outside of R
with the following options:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Batch mode
\item
  Sequential swap (``s'')
\item
  Printing all pairs (``y'')
\item
  C-score co-occurrence measure (``c'')
\item
  Default confidence limits (0.05)
\item
  Default iterations (100)
\item
  Maximum of 20 species
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pairs_txt =}\StringTok{ }\KeywordTok{readLines}\NormalTok{(}\StringTok{"fakedata/Pairs.txt"}\NormalTok{)}

\CommentTok{# Find areas of the data file that correspond}
\CommentTok{# to species pairs' results}
\NormalTok{beginnings =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}\StringTok{"Sp1"}\NormalTok{, pairs_txt) +}\StringTok{ }\DecValTok{1}
\NormalTok{ends =}\StringTok{ }\KeywordTok{c}\NormalTok{(}
  \KeywordTok{grep}\NormalTok{(}\StringTok{"^[^ ]"}\NormalTok{, pairs_txt)[-}\DecValTok{1}\NormalTok{],}
  \KeywordTok{length}\NormalTok{(pairs_txt)}
\NormalTok{) -}\StringTok{ }\DecValTok{1}

\CommentTok{# The above code fails on the very last line of the file}
\NormalTok{ends[}\KeywordTok{length}\NormalTok{(ends)] =}\StringTok{ }\NormalTok{ends[}\KeywordTok{length}\NormalTok{(ends)] +}\StringTok{ }\DecValTok{1}
\end{Highlighting}
\end{Shaded}

A function to import the data file and run each method on it:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit_all =}\StringTok{ }\NormalTok{function(filename)\{}
  \NormalTok{######## Import ########}

  \CommentTok{# Multiplying by one is necessary to prevent silly errors}
  \CommentTok{# regarding TRUE/FALSE versus 1/0}
  \NormalTok{raw_obs =}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(filename)[[}\StringTok{"observed"}\NormalTok{]] *}\StringTok{ }\DecValTok{1}

  \CommentTok{# Identify species that are never present (or never absent) so they }
  \CommentTok{# can be dropped}
  \NormalTok{species_is_variable =}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\KeywordTok{var}\NormalTok{(raw_obs)) >}\StringTok{ }\DecValTok{0}
  \NormalTok{pair_is_variable =}\StringTok{ }\KeywordTok{tcrossprod}\NormalTok{(species_is_variable) >}\StringTok{ }\DecValTok{0}

  \NormalTok{x =}\StringTok{ }\NormalTok{raw_obs[ , species_is_variable]}
  \NormalTok{truth =}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(filename)[[}\StringTok{"truth"}\NormalTok{]][pair_is_variable[}\KeywordTok{upper.tri}\NormalTok{(pair_is_variable)]]}

  \NormalTok{splitname =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(filename, }\StringTok{"/|-|}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{."}\NormalTok{)[[}\DecValTok{1}\NormalTok{]]}
  \NormalTok{n_sites =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(splitname[[}\DecValTok{3}\NormalTok{]])}
  \NormalTok{rep =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(splitname[[}\DecValTok{4}\NormalTok{]])}

  \NormalTok{######## Partial correlations ########}
  \NormalTok{p_corr =}\StringTok{ }\KeywordTok{pcor.shrink}\NormalTok{(x)}

  \NormalTok{######## Correlations ########}
  \NormalTok{corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(x)}

  \NormalTok{######## GLM ########}
  \NormalTok{coef_matrix =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{ncol}\NormalTok{(x), }\KeywordTok{ncol}\NormalTok{(x))}
  \NormalTok{for(i in }\DecValTok{1}\NormalTok{:}\KeywordTok{ncol}\NormalTok{(x))\{}
    \NormalTok{if(}\KeywordTok{var}\NormalTok{(x[,i]) >}\StringTok{ }\DecValTok{0}\NormalTok{)\{}
      \NormalTok{coefs =}\StringTok{ }\KeywordTok{coef}\NormalTok{(}\KeywordTok{bayesglm}\NormalTok{(x[,i] ~}\StringTok{ }\NormalTok{x[ , -i], }\DataTypeTok{family =} \NormalTok{binomial))[-}\DecValTok{1}\NormalTok{]}
      \NormalTok{coef_matrix[i, -i] =}\StringTok{ }\NormalTok{coefs}
    \NormalTok{\}}
  \NormalTok{\}}
  \NormalTok{coef_matrix =}\StringTok{ }\NormalTok{(coef_matrix +}\StringTok{ }\KeywordTok{t}\NormalTok{(coef_matrix)) /}\StringTok{ }\DecValTok{2}

  \NormalTok{######## Markov network ########}
  \NormalTok{rosie =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{maxit =} \DecValTok{200}\NormalTok{, }\DataTypeTok{trace =} \DecValTok{0}\NormalTok{, }\DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{))}

  \NormalTok{######## BayesComm and partial BayesComm ########}
  \NormalTok{bc =}\StringTok{ }\KeywordTok{BC}\NormalTok{(}\DataTypeTok{Y =} \NormalTok{x, }\DataTypeTok{model =} \StringTok{"community"}\NormalTok{, }\DataTypeTok{its =} \DecValTok{1000}\NormalTok{)}

  \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\DecValTok{0}

  \NormalTok{for(i in }\DecValTok{1}\NormalTok{:}\KeywordTok{nrow}\NormalTok{(bc$trace$R))\{}
    \NormalTok{Sigma =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{nrow =} \KeywordTok{ncol}\NormalTok{(x), }\DataTypeTok{ncol =} \KeywordTok{ncol}\NormalTok{(x))}
    \NormalTok{Sigma[}\KeywordTok{upper.tri}\NormalTok{(Sigma)] <-}\StringTok{ }\NormalTok{bc$trace$R[i, ]  }\CommentTok{# Fill in upper triangle}
    \NormalTok{Sigma <-}\StringTok{ }\NormalTok{Sigma +}\StringTok{ }\KeywordTok{t}\NormalTok{(Sigma)                   }\CommentTok{# Fill in lower triangle}
    \KeywordTok{diag}\NormalTok{(Sigma) <-}\StringTok{ }\DecValTok{1}  \CommentTok{# Diagonal equals 1 in multivariate probit model}

    \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ `}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{+}\StringTok{ }\KeywordTok{cor2pcor}\NormalTok{(Sigma) /}\StringTok{ }\KeywordTok{nrow}\NormalTok{(bc$trace$R)}
  \NormalTok{\}}

  \NormalTok{######## Pairs ########}
  \CommentTok{# Find the line where the current data set is mentioned in}
  \CommentTok{# pairs.txt}
  \NormalTok{filename_line =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}
    \KeywordTok{paste0}\NormalTok{(}
      \KeywordTok{gsub}\NormalTok{(}\StringTok{"fakedata/(.*)}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.rds"}\NormalTok{, }\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{1"}\NormalTok{, filename),}
      \StringTok{"-"}
    \NormalTok{),}
    \NormalTok{pairs_txt}
  \NormalTok{)}

  \CommentTok{# Which chunk of the data file corresponds to this file?}
  \NormalTok{chunk =}\StringTok{ }\KeywordTok{min}\NormalTok{(}\KeywordTok{which}\NormalTok{(beginnings >}\StringTok{ }\NormalTok{filename_line))}

  \CommentTok{# Split the chunk on whitespace.}
  \NormalTok{splitted =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(pairs_txt[beginnings[chunk]:ends[chunk]], }\StringTok{" +"}\NormalTok{)}

  \CommentTok{# Pull out the species numbers and their Z-scores}
  \NormalTok{pairs_results =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
    \NormalTok{splitted,}
    \NormalTok{function(x)\{}
      \NormalTok{spp =}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(x[}\DecValTok{3}\NormalTok{:}\DecValTok{4}\NormalTok{]))}
      \KeywordTok{data.frame}\NormalTok{(}
        \DataTypeTok{sp1 =} \NormalTok{spp[}\DecValTok{1}\NormalTok{], }
        \DataTypeTok{sp2 =} \NormalTok{spp[}\DecValTok{2}\NormalTok{], }
        \DataTypeTok{z =} \KeywordTok{as.numeric}\NormalTok{(x[}\DecValTok{14}\NormalTok{])}
      \NormalTok{)}
    \NormalTok{\}}
  \NormalTok{) %>%}\StringTok{ }
\StringTok{    }\NormalTok{bind_rows %>%}
\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{spp =} \KeywordTok{paste}\NormalTok{(sp1, sp2, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{))}

  \CommentTok{# Re-order the pairs_results to match the other methods}
  \NormalTok{m =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\OtherTok{NA}\NormalTok{, }\KeywordTok{ncol}\NormalTok{(x), }\KeywordTok{ncol}\NormalTok{(x))}
  \NormalTok{new_order =}\StringTok{ }\KeywordTok{match}\NormalTok{(}
    \KeywordTok{paste}\NormalTok{(}\KeywordTok{row}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)], }\KeywordTok{col}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)], }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{),}
    \NormalTok{pairs_results$spp}
  \NormalTok{)}
  \NormalTok{ordered_pairs_results =}\StringTok{ }\NormalTok{pairs_results[new_order, ]}

  \NormalTok{######## Output ########}
  \KeywordTok{data.frame}\NormalTok{(}
    \DataTypeTok{truth =} \NormalTok{truth, }
    \DataTypeTok{n_sites =} \NormalTok{n_sites,}
    \DataTypeTok{rep =} \NormalTok{rep,}
    \DataTypeTok{sp1 =} \NormalTok{ordered_pairs_results$sp1,}
    \DataTypeTok{sp2 =} \NormalTok{ordered_pairs_results$sp2,}
    \StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{p_corr[}\KeywordTok{upper.tri}\NormalTok{(p_corr)],}
    \DataTypeTok{correlation =} \NormalTok{corr[}\KeywordTok{upper.tri}\NormalTok{(corr)],}
    \StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{rosie$beta[}\KeywordTok{upper.tri}\NormalTok{(rosie$beta)],}
    \DataTypeTok{GLM =} \NormalTok{coef_matrix[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)],}
    \StringTok{`}\DataTypeTok{BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(bc$trace$R),}
    \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ `}\DataTypeTok{partial BayesComm}\StringTok{`}\NormalTok{[}\KeywordTok{upper.tri}\NormalTok{(}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`}\NormalTok{)],}
    \DataTypeTok{Pairs =} \NormalTok{ordered_pairs_results$z}
  \NormalTok{)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Run the above function on all the files:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Find all the .rds files in the fakedata folder}
\NormalTok{files =}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"fakedata"}\NormalTok{, }\DataTypeTok{pattern =} \StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.rds$"}\NormalTok{, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{)}

\CommentTok{# Run all the analyses on all the files}
\NormalTok{z =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(files, fit_all) %>%}\StringTok{ }\NormalTok{bind_rows %>%}\StringTok{ }\NormalTok{as.data.frame}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0302 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0189 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0172 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0461 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0297 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0165 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0315 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.021 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0126 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0942 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0302 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0281 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0199 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0156 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0256 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0323 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0293 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0253 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0283 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0165 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0125 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0221 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0426 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0403 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0353 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0224 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0238 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0149 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0132 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0169 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0364 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0322 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0377 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.009 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0213 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.018 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.034 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0255 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0321 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0147 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0161 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0253 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0121 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0166 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0235 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0237 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0358 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0173 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0532 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1276 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.311 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3631 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2586 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1348 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1437 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3609 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0889 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1302 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0517 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1508 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1163 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1135 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1597 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2075 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1392 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.459 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1726 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1095 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1174 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1909 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1924 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2261 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1249 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.159 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1499 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1224 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1935 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2215 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1248 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1337 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0892 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1546 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0723 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1481 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1154 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1386 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1158 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1681 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1362 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2259 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1116 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.12 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1533 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1226 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1985 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3044 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1416 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1534 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3426 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7236 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5589 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4883 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6616 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7645 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.494 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7008 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3284 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6008 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6168 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4011 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3676 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2673 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6011 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6675 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6813 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4592 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6586 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6237 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3866 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4468 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.767 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6031 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.851 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6081 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5805 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5058 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4585 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6108 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4147 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5198 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4652 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6374 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4167 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6822 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3729 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5584 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5641 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7974 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7246 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5429 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6172 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6817 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5112 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6673 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7543 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.755 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7918 
## 
## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7204
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Fix a formatting issue with the column names}
\KeywordTok{colnames}\NormalTok{(z) =}\StringTok{ }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{."}\NormalTok{, }\StringTok{" "}\NormalTok{, }\KeywordTok{colnames}\NormalTok{(z))}
\end{Highlighting}
\end{Shaded}

Summarize the results of all 7 methods across the simulated landscapes:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Calculate residuals for each method}
\NormalTok{resids =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
  \KeywordTok{colnames}\NormalTok{(z)[-(}\DecValTok{1}\NormalTok{:}\DecValTok{5}\NormalTok{)],}
  \NormalTok{function(i)\{}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z[,i] +}\StringTok{ }\DecValTok{0}\NormalTok{))\}}
\NormalTok{)}

\NormalTok{sizes =}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\KeywordTok{unique}\NormalTok{(z$n_sites))}

\CommentTok{# Compute proportion of variance explained, compared with a null that}
\CommentTok{# assumes all species interactions are 0.}
\NormalTok{results =}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}
  \KeywordTok{t}\NormalTok{(}
    \KeywordTok{sapply}\NormalTok{(}
      \NormalTok{sizes,}
      \NormalTok{function(n)\{}
        \NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{z))[z$n_sites ==}\StringTok{ }\NormalTok{n]^}\DecValTok{2}\NormalTok{)}
        \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids[z$n_sites ==}\StringTok{ }\NormalTok{n , ]^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss}
      \NormalTok{\}}
    \NormalTok{)}
  \NormalTok{)}
\NormalTok{)}
\CommentTok{# Sort the results in decreasing order}
\NormalTok{results =}\StringTok{ }\NormalTok{results[}\KeywordTok{order}\NormalTok{(}\KeywordTok{colMeans}\NormalTok{(results), }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

Plot the results:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{colors =}\StringTok{ }\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{8}\NormalTok{, }\StringTok{"Dark2"}\NormalTok{)[}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{8}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{7}\NormalTok{)]}

\CommentTok{# Set up the plotting canvas}
\KeywordTok{pdf}\NormalTok{(}\StringTok{"manuscript-materials/figures/performance.pdf"}\NormalTok{, }\DataTypeTok{height =} \DecValTok{8}\NormalTok{, }\DataTypeTok{width =} \DecValTok{5}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar =} \KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{0}\NormalTok{) +}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{)}

\CommentTok{# Plot the model performance}
\KeywordTok{matplot}\NormalTok{(}
  \NormalTok{sizes, }
  \NormalTok{results, }
  \DataTypeTok{type =} \StringTok{"o"}\NormalTok{, }
  \DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }
  \DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\DecValTok{25} \NormalTok{*}\StringTok{ }\NormalTok{.}\DecValTok{9}\NormalTok{, }\DecValTok{1600} \NormalTok{*}\StringTok{ }\DecValTok{8}\NormalTok{),}
  \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{5} \NormalTok{+}\StringTok{ }\FloatTok{1E-9}\NormalTok{),}
  \DataTypeTok{pch =} \KeywordTok{c}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{16}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{), }
  \DataTypeTok{lty =} \DecValTok{1}\NormalTok{,}
  \DataTypeTok{log =} \StringTok{"x"}\NormalTok{,}
  \DataTypeTok{xlab =} \StringTok{""}\NormalTok{,}
  \DataTypeTok{axes =} \OtherTok{FALSE}\NormalTok{,}
  \DataTypeTok{xaxs =} \StringTok{"i"}\NormalTok{,}
  \DataTypeTok{yaxs =} \StringTok{"i"}\NormalTok{,}
  \DataTypeTok{col =} \NormalTok{colors,}
  \DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{,}
  \DataTypeTok{cex =} \FloatTok{0.9}\NormalTok{,}
  \DataTypeTok{bty =} \StringTok{"l"}
\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{25}\NormalTok{, }\DecValTok{200}\NormalTok{, }\DecValTok{1600}\NormalTok{))}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{), }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}

\KeywordTok{mtext}\NormalTok{(}\KeywordTok{expression}\NormalTok{(R^}\DecValTok{2}\NormalTok{), }\DataTypeTok{side =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \DecValTok{3}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"Number of sites (log scale)"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{1}\NormalTok{, }\DataTypeTok{line =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{200}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"A."}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{25}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{1.5}\NormalTok{)}

\CommentTok{# Determine how high on the y-axis each method label should be so they}
\CommentTok{# don't overlap.}
\NormalTok{heights =}\StringTok{ }\NormalTok{results[}\KeywordTok{nrow}\NormalTok{(results), ]}

\NormalTok{heights$Pairs =}\StringTok{ }\NormalTok{heights$Pairs -}\StringTok{ }\NormalTok{.}\DecValTok{01}
\NormalTok{heights$correlation =}\StringTok{ }\NormalTok{heights$correlation +}\StringTok{ }\NormalTok{.}\DecValTok{01}

\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{+}\StringTok{ }\NormalTok{.}\DecValTok{01}
\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{-}\StringTok{ }\NormalTok{.}\DecValTok{01}

\NormalTok{heights$GLM =}\StringTok{ }\NormalTok{heights$GLM -}\StringTok{ }\NormalTok{.}\DecValTok{01}

\KeywordTok{text}\NormalTok{(}\DecValTok{1625}\NormalTok{, heights, }\KeywordTok{colnames}\NormalTok{(results), }\DataTypeTok{pos =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.75}\NormalTok{, }\DataTypeTok{col =} \NormalTok{colors)}

\NormalTok{######}

\CommentTok{# silly code to make the next graph line up prettily with the previous one}
\NormalTok{full_range =}\StringTok{ }\KeywordTok{log10}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{25} \NormalTok{*}\StringTok{ }\NormalTok{.}\DecValTok{9}\NormalTok{, }\DecValTok{1600} \NormalTok{*}\StringTok{ }\DecValTok{8}\NormalTok{))}
\NormalTok{base_range =}\StringTok{ }\KeywordTok{log10}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DecValTok{1600}\NormalTok{))}

\NormalTok{base_scaled =}\StringTok{ }\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{base_range /}\StringTok{ }\NormalTok{(base_range[}\DecValTok{2}\NormalTok{] -}\StringTok{ }\NormalTok{base_range[}\DecValTok{1}\NormalTok{])}
\NormalTok{full_scaled =}\StringTok{ }\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{full_range /}\StringTok{ }\NormalTok{(base_range[}\DecValTok{2}\NormalTok{] -}\StringTok{ }\NormalTok{base_range[}\DecValTok{1}\NormalTok{])}

\CommentTok{# New colors for plot B}
\NormalTok{colors2 =}\StringTok{ }\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{10}\NormalTok{, }\StringTok{"Paired"}\NormalTok{)[}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{10}\NormalTok{)]}
\KeywordTok{plot}\NormalTok{(}
  \NormalTok{z$correlation, }
  \NormalTok{z$Pairs, }
  \DataTypeTok{col =} \NormalTok{colors2[}\KeywordTok{factor}\NormalTok{(z$n_sites)],}
  \DataTypeTok{las =} \DecValTok{1}\NormalTok{,}
  \DataTypeTok{xlab =} \StringTok{""}\NormalTok{,}
  \DataTypeTok{ylab =} \KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(Z)-score),}
  \DataTypeTok{bty =} \StringTok{"l"}\NormalTok{,}
  \DataTypeTok{xaxs =} \StringTok{"i"}\NormalTok{,}
  \DataTypeTok{axes =} \OtherTok{FALSE}\NormalTok{,}
  \DataTypeTok{xlim =} \NormalTok{full_scaled -}\StringTok{ }\NormalTok{base_scaled[}\DecValTok{2}\NormalTok{] +}\StringTok{ }\DecValTok{1}
\NormalTok{)}
\KeywordTok{axis}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{seq}\NormalTok{(-}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, .}\DecValTok{5}\NormalTok{))}
\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\KeywordTok{seq}\NormalTok{(-}\DecValTok{50}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DecValTok{10}\NormalTok{), }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"Correlation coefficient"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{1}\NormalTok{, }\DataTypeTok{line =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{0}\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"B."}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.25}\NormalTok{, }\DataTypeTok{at =} \NormalTok{-}\DecValTok{1}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{1.5}\NormalTok{)}

\KeywordTok{legend}\NormalTok{(}
  \DataTypeTok{x =} \FloatTok{1.05}\NormalTok{, }
  \DataTypeTok{y =} \KeywordTok{max}\NormalTok{(z$Pairs),}
  \DataTypeTok{pch =} \DecValTok{1}\NormalTok{,}
  \DataTypeTok{lty =} \DecValTok{0}\NormalTok{,}
  \DataTypeTok{lwd =} \DecValTok{2}\NormalTok{,}
  \DataTypeTok{col =} \NormalTok{colors2[}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(}\KeywordTok{unique}\NormalTok{(z$n_sites))], }
  \DataTypeTok{legend =} \KeywordTok{levels}\NormalTok{(}\KeywordTok{factor}\NormalTok{(z$n_sites)),}
  \DataTypeTok{title =} \StringTok{"Number of sites"}\NormalTok{,}
  \DataTypeTok{bty =} \StringTok{"n"}\NormalTok{,}
  \DataTypeTok{cex =} \FloatTok{0.75}\NormalTok{,}
  \DataTypeTok{pt.cex =} \DecValTok{1}
\NormalTok{)}
\KeywordTok{segments}\NormalTok{(-}\DecValTok{1000}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{1.04}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DataTypeTok{col =} \StringTok{"#00000025"}\NormalTok{, }\DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{)}
\KeywordTok{segments}\NormalTok{(}\DecValTok{0}\NormalTok{, -}\DecValTok{1000}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1000}\NormalTok{, }\DataTypeTok{col =} \StringTok{"#00000025"}\NormalTok{, }\DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{)}

\KeywordTok{dev.off}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\includegraphics{manuscript-materials/figures/performance.pdf}

Summarize the results:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# R-squareds computed across *all* landscape types}
\NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{z))^}\DecValTok{2}\NormalTok{)}
\KeywordTok{round}\NormalTok{(}\DecValTok{100} \NormalTok{*}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss), }\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           BayesComm               Pairs   partial BayesComm 
##                12.7                15.4                18.5 
##         correlation partial correlation                 GLM 
##                18.6                27.5                29.9 
##      Markov network 
##                34.6
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Rank correlations among the methods that estimate marginal relationships}
\KeywordTok{round}\NormalTok{(}\KeywordTok{cor}\NormalTok{(z[ , }\KeywordTok{c}\NormalTok{(}\DecValTok{7}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{)], }\DataTypeTok{method =} \StringTok{"spearman"}\NormalTok{)[,}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## correlation   BayesComm       Pairs 
##        1.00        0.79       -0.92
\end{verbatim}

Plot the true interactions versus estimated interactions for two
methods:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(ggplot2)}

\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Markov network}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Markov network"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\includegraphics{recovery_files/figure-latex/unnamed-chunk-8-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$Pairs +}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\includegraphics{recovery_files/figure-latex/unnamed-chunk-8-2.pdf}

Bootstrap resample from each of the three landscape size classes,
generating new sets of 150 landscapes. For each one, calculate the
proportion of variance explained (compared to a null baseline that
assumes all species pairs' interaction strengths are zero).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rep =}\StringTok{ }\DecValTok{12}
\NormalTok{n =}\StringTok{ }\DecValTok{200}
\NormalTok{boots =}\StringTok{ }\KeywordTok{replicate}\NormalTok{(}\DecValTok{500}\NormalTok{,}
          \NormalTok{\{}
            \NormalTok{boot =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
              \NormalTok{sizes, }
              \NormalTok{function(n)\{}\KeywordTok{lapply}\NormalTok{(}
                \KeywordTok{sample.int}\NormalTok{(}\KeywordTok{max}\NormalTok{(z$rep), }\DataTypeTok{replace =} \OtherTok{TRUE}\NormalTok{),}
                \NormalTok{function(rep)\{}
                  \NormalTok{z[z$rep ==}\StringTok{ }\NormalTok{rep &}\StringTok{ }\NormalTok{z$n_sites ==}\StringTok{ }\NormalTok{n, ]}
                \NormalTok{\}}
              \NormalTok{) %>%}\StringTok{ }\NormalTok{bind_rows\}}
            \NormalTok{) %>%}\StringTok{ }\NormalTok{bind_rows}

            \NormalTok{resids =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
              \KeywordTok{colnames}\NormalTok{(boot)[-(}\DecValTok{1}\NormalTok{:}\DecValTok{5}\NormalTok{)],}
              \NormalTok{function(i)\{}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(boot$truth ~}\StringTok{ }\NormalTok{boot[[i]] +}\StringTok{ }\DecValTok{0}\NormalTok{))\}}
            \NormalTok{)}
            \NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{boot))^}\DecValTok{2}\NormalTok{)}
            \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss}
          \NormalTok{\}}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Summarize the bootstrap results:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{for(compared in }\KeywordTok{c}\NormalTok{(}\StringTok{"partial correlation"}\NormalTok{, }\StringTok{"correlation"}\NormalTok{, }\StringTok{"GLM"}\NormalTok{, }\StringTok{"BayesComm"}\NormalTok{, }\StringTok{"partial BayesComm"}\NormalTok{, }\StringTok{"Pairs"}\NormalTok{))\{}
  \NormalTok{CI =}\StringTok{ }\KeywordTok{round}\NormalTok{(}
    \KeywordTok{quantile}\NormalTok{(}
      \NormalTok{boots[}\StringTok{"Markov network"}\NormalTok{, ] /}\StringTok{ }\NormalTok{boots[compared, ], }
      \KeywordTok{c}\NormalTok{(.}\DecValTok{025}\NormalTok{, .}\DecValTok{975}\NormalTok{)), }
    \DecValTok{2}
  \NormalTok{)}

  \KeywordTok{message}\NormalTok{(}\StringTok{"R-squared is "} \NormalTok{, CI[}\DecValTok{1}\NormalTok{], }\StringTok{"-"}\NormalTok{, CI[}\DecValTok{2}\NormalTok{], }\StringTok{" times higher than from "}\NormalTok{, compared, }\StringTok{" (95% bootstrap CI)"}\NormalTok{)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
## R-squared is 1.77-1.95 times higher than from correlation (95% bootstrap CI)
## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
## R-squared is 2.54-2.9 times higher than from BayesComm (95% bootstrap CI)
## R-squared is 1.77-1.97 times higher than from partial BayesComm (95% bootstrap CI)
## R-squared is 2.12-2.38 times higher than from Pairs (95% bootstrap CI)
\end{verbatim}

\end{document}

### 2015-08-04 Harris_2015/boot.R

@@ -1,0 +1,42 @@
+library(rosalia)
+
+set.seed(1)
+
+n_boot = 50
+
+# Load a simulated data set with 20 species at 200 sites.
+# Multiply by one to avoid problems with FALSE/TRUE vs 0/1
+d = readRDS("fakedata/20-200-1.rds")$observed * 1
+
+# Fit a Markov network to the full data set
+full_fit = rosalia(d, prior = make_logistic_prior(scale = 2), hessian = TRUE,
+                   maxit = 200)
+
+
+# Repeat the analysis on each bootstrap replicate
+boots = replicate(
+  n_boot,
+  {
+    boot_sample = d[sample.int(nrow(d), replace = TRUE), ]
+    
+    # Initializing the optimization at full_fit$opt$par should save 
+    # time and shouldn't change the final outcome because the prior
+    # and likelihood are both convex
+    boot_fit = rosalia(
+      boot_sample, 
+      prior = make_logistic_prior(scale = 2), 
+      parlist = relist(full_fit$opt$par),
+      trace = 0
+    )
+    
+    # Output the upper triangle of the fitted model
+    boot_fit$beta[upper.tri(boot_fit$beta)]
+  }
+)
+
+# Correlations among the bootstrapped estimates
+# (upper triangle avoids the diagonal, which is always 1)
+boot_cors = cor(boots)[upper.tri(cor(boots))]
+
+# Mean squared correlation among replicates = 0.494
+mean(boot_cors^2)

### 2015-08-04 Harris_2015/manuscript-materials/boot.R

@@ -1,22 +1,0 @@
-library(rosalia)
-
-n_boot = 50
-
-d = readRDS("fakedata/20-200-1.rds")$observed * 1
-
-full_fit = rosalia(d, prior = make_logistic_prior(scale = 2), hessian = TRUE)
-
-
-
-boots = replicate(
-  n_boot,
-  {
-    boot_sample = d[sample.int(nrow(d), replace = TRUE), ]
-    boot_fit = rosalia(
-      boot_sample, 
-      prior = make_logistic_prior(scale = 2), 
-      parlist = full_fit$opt$par
-    )
-    boot_fit$beta[upper.tri(boot_fit$beta)]
-  }
-)

### 2015-08-04 Harris_2015/recovery.R

@@ -1,13 +1,22 @@
-library(dplyr)
-library(corpcor)
-library(rosalia)
-library(arm)
-library(BayesComm)
+library(dplyr)        # For manipulating data structures
+library(corpcor)      # For regularized partial covariances
+library(rosalia)      # For Markov networks
+library(arm)          # For regularized logistic regression
+library(BayesComm)    # For joint species distribution modeling
+library(RColorBrewer) # For color palette
 
-set.seed(1)
+set.seed(1) # set seed for reproducibility
 
 
-# Load in the results from the `pairs` program
+# Load in the results from the `pairs` program, run outside of R with
+# the following options:
+#   Batch mode
+#   Sequential swap ("s")
+#   Printing all pairs ("y")
+#   C-score co-occurrence measure ("c")
+#   Default confidence limits (0.05)
+#   Default iterations (100)
+#   Maximum of 20 species
 pairs_txt = readLines("fakedata/Pairs.txt")
 
 # Find areas of the data file that correspond
@@ -20,7 +29,7 @@ ends = c(
 # The above code breaks on the very last line of the file
 ends[length(ends)] = ends[length(ends)] + 1
 
-
+# Import the data file and run each method on it
 f = function(filename){
   # Multiplying by one is necessary to prevent silly errors
   # regarding TRUE/FALSE versus 1/0
@@ -130,42 +139,26 @@ f = function(filename){
 }
 
 
-
+# Find all the .rds files in the fakedata folder
 files = dir("fakedata", pattern = "\\.rds$", full.names = TRUE)
 
+# Run all the analyses on all the files
 z = lapply(files, f) %>% bind_rows %>% as.data.frame
+
+# Fix a formatting issue with the column names
 colnames(z) = gsub("\\.", " ", colnames(z))
 
-library(ggplot2)
-ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) + 
-  stat_binhex(bins = 100) + 
-  xlab("Estimated coefficient value") + 
-  ylab("\"True\" coefficient value") + 
-  stat_hline(yintercept = 0, size = 1/8) + 
-  stat_vline(xintercept = 0, size = 1/8) + 
[…]

### 2015-08-05 Harris_2015/manuscript-materials/draft.Rmd
submitted to Ecology

@@
 
-\setlength{\parskip}{0pt}
+\setlength{\parskip}{3pt}
 
@@
 single pairwise interaction can ripple through an ecological network and produce
-surprising indirect consequences. For example, two competing species would
-ordinarily correlate negatively in space, but this effect can be reversed in
-the presence of a third species that is capable of outcompeting both of them
-when it is present. Here, I apply models from statistical physics, called
-Markov networks or Markov random fields, that can predict the direct and
-indirect consequences of any possible species interaction matrix. Interactions
-in these models can be estimated from observational data via maximum
-likelihood. Using simulated landscapes with known pairwise interaction
-strengths, I evaluated Markov networks and several existing approaches. The
+surprising indirect consequences. For example, the negative correlation between
+two competing species can be reversed in the presence of a third species that is
+capable of outcompeting both of them. Here, I apply models from statistical
+physics, called Markov networks or Markov random fields, that can predict
+the direct and indirect consequences of any possible species interaction matrix.
+Interactions in these models can be estimated from observational data via
+maximum likelihood. Using simulated landscapes with known pairwise interaction
+strengths, I evaluated Markov networks and six existing approaches. The
 Markov networks consistently outperformed other methods, correctly isolating
-direct interactions between species pairs even when indirect interactions or
-abiotic environmental effects largely overpowered them. Two computationally
-efficient approximations, based on linear and generalized linear models, also
-performed well. Indirect effects reliably caused a common null modeling approach
-to produce incorrect inferences, however.
+direct interactions between species pairs even when indirect interactions
+largely overpowered them. Two computationally efficient approximations, based on
+linear and generalized linear models, also performed well. Indirect effects
+reliably caused a common null modeling approach to produce incorrect inferences,
+however.
 
@@
 
-\setlength{\parindent}{1em}
-\setlength{\parskip}{3pt}
+\setlength{\parindent}{1cm}
 
-If nontrophic species interactions, such as competition, are important drivers
-of community assembly, then ecologists might expect to see their influence in
+To the extent that nontrophic species interactions (such as competition) affect
+community assembly, ecologists might expect to find signatures of these interactions in
 species composition data [@macarthur_population_1958; @diamond_island_1975].
@@
 
-While competition doesn't reliably leave the expected pattern in community
-structure (low co-occurrence rates at the landscape level), it nevertheless
-does leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
-there will be a deficit of co-occurrences between the competing shrub species,
-and *among unshaded sites*, there will also be such a deficit. More generally,
+While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
+level (as most of our methods assume), it nevertheless does leave a signal in
+the data (Figure 1C). Specifically, after partitioning the data set into shaded
+sites and unshaded sites, there will be co-occurrence deficits in each subset
+that might not be apparent at the landscape level. More generally,
 we can obtain much better estimates of the association between two species from
@@
 precision is difficult to obtain from null models, which begin with the
-assumptions that all the pairwise interactions are zero and thus don’t need to
+assumption that all the pairwise interactions are zero and thus don’t need to
 be controlled for.  Nevertheless, null models have dominated this field for more
@@
 and for detecting them from observed presence-absence matrices. Markov networks
-have been used in many scientific fields for decades in similar contexts, from
+have been used in many scientific fields in similar contexts for decades, from
 physics [where nearby particles interact magnetically; @cipra_introduction_1987]
@@
 assemblages that these relationships produce. Here, I show how a set of
-conditional relationships can be used to determine the how groups of species
+conditional relationships can be used to determine how groups of species
 can co-occur. Methods for estimating conditional relationships from data are
@@
 \raggedright
-\setlength{\parindent}{1em}
+\setlength{\parindent}{1cm}
 
@@
 that would be expected to produce exactly the observed co-occurrence frequencies
-on average [i.e. it matches the sufficient statistics of the model distribution
+on average [i.e. maximizing the likelihood matches the sufficient statistics of the model distribution
 to the sufficient statistics of the data; @murphy_machine_2012]. I used the
@@
 #### Recovering species interactions from simulated data.
+<!-- exp(qlogis(.975, scale = 2)) -->
+
 I compared seven techniques for determining the sign and strength of the
 associations between pairs of species from simulated data (Appendix 3).
-
-<!-- exp(qlogis(.975, scale = 2)) -->
-
 First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
@@
 
-I also evaluated five alternative methods from the existing literature (plus a
-novel variant of the fifth method). The first two alternative interaction
-measures were the sample correlations and the partial correlations between each
-pair of species' data vectors on the landscape [@albrecht_spatial_2001;
+I also evaluated six alternative methods: five from the existing literature,
+plus a novel combination of two of these methods. The first alternative interaction
+metric was the sample correlation between species' presence-absence vectors,
+which summarizes their marginal association.  Next, I used partial correlations,
+which summarize species' conditional relationships [@albrecht_spatial_2001;
 @faisal_inferring_2010]. In the context of non-Gaussian data, the partial
@@
 
-The fourth alternative method, described in @gotelli_empirical_2009, involved
+The next method, described in @gotelli_empirical_2009, involved
 simulating new landscapes from a null model that retains the row and column
@@
 interaction strengths among all of the species pairs from the larger simulated
-landscapes. This determined which the methods provide a consistent way to know
+landscapes. This determined which of the methods provide a consistent way to know
 how strong species interactions are---regardless of which species were present
@@
 data sets. These squared errors determined the proportion of variance explained
-for different combinations of method and landscape size (compared with a
+for different combinations of model type and landscape size (compared with a
 null model that assumed all interaction strengths to be zero).
@@
 landscape also confirmed that the rosalia package's estimates of species'
-conditional relationships were reasonably robust to sampling variation
-(Appendix 4).
+conditional relationships were robust to sampling variation for reasonably-sized
+landscapes (Appendix 4).
 
@@
 interactions between competing species, and can make precise statements about
-the conditions whre indirect interactions will overwhelm direct ones. They also
+the conditions where indirect interactions will overwhelm direct ones. They also
 provide a simple answer to the question of how competition should affect a
@@
 Equation 1 can be used to calculate the expected prevalence of a species in the
-absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
+absence of biotic influences [$\frac{e^\alpha}{1 + e^{\alpha}}$; @lee_learning_2012].
 Competition's effect on prevalence in a Markov network can then be calculated by
@@
 Markov networks---particularly the Ising model for binary networks---have been
-studied in statistical physics for nearly a century [@cipra_introduction_1987],
-and the models' properties, capabilities, and limits are well-understood in a
+studied for nearly a century [@cipra_introduction_1987], and the models'
+properties, capabilities, and limits are well-understood in a
 huge range of applications. Using the same framework for species interactions
@@
 this field could often double their explanatory power by switching from null
-models to Markov networks (or increase it substantially by with linear or
+models to Markov networks (or increase it substantially with linear or
 generalized linear approximations). Regardless of the methods ecologists
@@
 **A.** Proportion of variance in interaction coefficients explained by each
-versus number of sampled locations.
+method versus number of sampled locations.
 **B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
-species are very predictable from the correlation between the presence-absence
+species can be predicted using the correlation between the presence-absence
 vectors of those same species and from the number of sites on the landscape.

### 2015-08-05 Harris_2015/manuscript-materials/draft.tex
submitted to Ecology

@@
 ecological network and produce surprising indirect consequences. For
-example, two competing species would ordinarily correlate negatively in
-space, but this effect can be reversed in the presence of a third
-species that is capable of outcompeting both of them when it is present.
-Here, I apply models from statistical physics, called Markov networks or
-Markov random fields, that can predict the direct and indirect
-consequences of any possible species interaction matrix. Interactions in
-these models can be estimated from observational data via maximum
-likelihood. Using simulated landscapes with known pairwise interaction
-strengths, I evaluated Markov networks and several existing approaches.
-The Markov networks consistently outperformed other methods, correctly
-isolating direct interactions between species pairs even when indirect
-interactions or abiotic environmental effects largely overpowered them.
-Two computationally efficient approximations, based on linear and
-generalized linear models, also performed well. Indirect effects
-reliably caused a common null modeling approach to produce incorrect
-inferences, however.
+example, the negative correlation between two competing species can be
+reversed in the presence of a third species that is capable of
+outcompeting both of them. Here, I apply models from statistical
+physics, called Markov networks or Markov random fields, that can
+predict the direct and indirect consequences of any possible species
+interaction matrix. Interactions in these models can be estimated from
+observational data via maximum likelihood. Using simulated landscapes
+with known pairwise interaction strengths, I evaluated Markov networks
+and six existing approaches. The Markov networks consistently
+outperformed other methods, correctly isolating direct interactions
+between species pairs even when indirect interactions largely
+overpowered them. Two computationally efficient approximations, based on
+linear and generalized linear models, also performed well. Indirect
+effects reliably caused a common null modeling approach to produce
+incorrect inferences, however.
 
@@
 
-\setlength{\parindent}{1em} \setlength{\parskip}{3pt}
+\setlength{\parindent}{1cm}
 
-If nontrophic species interactions, such as competition, are important
-drivers of community assembly, then ecologists might expect to see their
-influence in species composition data (MacArthur 1958, Diamond 1975).
-Despite decades of work and several major controversies, however (Lewin
-1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et al.
-2013), existing methods for detecting competition's effects on community
-structure are unreliable (Gotelli and Ulrich 2009). In particular,
-species' effects on one another can become lost in the complex web of
-direct and indirect interactions in real assemblages. For example, the
-competitive interaction between the two shrub species in Figure 1A can
-become obscured by their shared tendency to occur in unshaded areas
-(Figure 1B). While ecologists have long known that indirect effects can
-overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
-the vast majority of our methods for drawing inferenes from
-observational data do not control for these effects (e.g. Diamond 1975,
-Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
-2014). To the extent that indirect interactions like those in Figure 1
-are generally important (Dodson 1970), existing methods will thus not
-generally provide much evidence regarding species' direct effects on one
-another. The goal of this paper is to resolve this long-standing
-problem.
+To the extent that nontrophic species interactions (such as competition)
+affect community assembly, ecologists might expect to find signatures of
+these interactions in species composition data (MacArthur 1958, Diamond
+1975). Despite decades of work and several major controversies, however
+(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
+al. 2013), existing methods for detecting competition's effects on
+community structure are unreliable (Gotelli and Ulrich 2009). In
+particular, species' effects on one another can become lost in the
+complex web of direct and indirect interactions in real assemblages. For
+example, the competitive interaction between the two shrub species in
+Figure 1A can become obscured by their shared tendency to occur in
+unshaded areas (Figure 1B). While ecologists have long known that
+indirect effects can overwhelm direct ones at the landscape level
+(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
+inferenes from observational data do not control for these effects (e.g.
+Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
+Pollock et al. 2014). To the extent that indirect interactions like
+those in Figure 1 are generally important (Dodson 1970), existing
+methods will thus not generally provide much evidence regarding species'
+direct effects on one another. The goal of this paper is to resolve this
+long-standing problem.
 
-While competition doesn't reliably leave the expected pattern in
-community structure (low co-occurrence rates at the landscape level), it
-nevertheless does leave a signal in the data (Figure 1C). Specifically,
-\emph{among shaded sites}, there will be a deficit of co-occurrences
-between the competing shrub species, and \emph{among unshaded sites},
-there will also be such a deficit. More generally, we can obtain much
-better estimates of the association between two species from their
-conditional relationships (i.e.~by controlling for other species in the
-network) than we could get from their overall co-occurrence rates. This
-kind of precision is difficult to obtain from null models, which begin
-with the assumptions that all the pairwise interactions are zero and
-thus don't need to be controlled for. Nevertheless, null models have
-dominated this field for more than three decades (Strong et al. 1984,
-Gotelli and Ulrich 2009).
+While competition doesn't reliably reduce co-occurrence rates at the
+whole-landscape level (as most of our methods assume), it nevertheless
+does leave a signal in the data (Figure 1C). Specifically, after
+partitioning the data set into shaded sites and unshaded sites, there
+will be co-occurrence deficits in each subset that might not be apparent
+at the landscape level. More generally, we can obtain much better
+estimates of the association between two species from their conditional
+relationships (i.e.~by controlling for other species in the network)
+than we could get from their overall co-occurrence rates. This kind of
+precision is difficult to obtain from null models, which begin with the
+assumption that all the pairwise interactions are zero and thus don't
+need to be controlled for. Nevertheless, null models have dominated this
+field for more than three decades (Strong et al. 1984, Gotelli and
+Ulrich 2009).
 
@@
 interactions, and for detecting them from observed presence-absence
-matrices. Markov networks have been used in many scientific fields for
-decades in similar contexts, from physics (where nearby particles
+matrices. Markov networks have been used in many scientific fields in
+similar contexts for decades, from physics (where nearby particles
 interact magnetically; Cipra 1987) to spatial statistics (where adjacent
@@
 kinds of species assemblages that these relationships produce. Here, I
-show how a set of conditional relationships can be used to determine the
-how groups of species can co-occur. Methods for estimating conditional
+show how a set of conditional relationships can be used to determine how
+groups of species can co-occur. Methods for estimating conditional
 relationships from data are discussed in the next section.
@@
 \(\beta\) coefficients that would be expected to produce exactly the
-observed co-occurrence frequencies on average (i.e.~it matches the
-sufficient statistics of the model distribution to the sufficient
-statistics of the data; Murphy 2012). I used the rosalia package (Harris
-2015a) for the R programming language (R Core Team 2015) to optimize the
-Markov network parameters. The package was named after Santa Rosalia,
-the patron saint of biodiversity, whose supposedly miraculous healing
-powers played an important rhetorical role in the null model debates of
-the 1970's and 1980's (Lewin 1983).
+observed co-occurrence frequencies on average (i.e.~maximizing the
+likelihood matches the sufficient statistics of the model distribution
+to the sufficient statistics of the data; Murphy 2012). I used the
+rosalia package (Harris 2015a) for the R programming language (R Core
+Team 2015) to optimize the Markov network parameters. The package was
+named after Santa Rosalia, the patron saint of biodiversity, whose
+supposedly miraculous healing powers played an important rhetorical role
+in the null model debates of the 1970's and 1980's (Lewin 1983).
 
@@
 associations between pairs of species from simulated data (Appendix 3).
-
 First, I used the rosalia package (Harris 2015a) to fit Markov newtork
@@
 
-I also evaluated five alternative methods from the existing literature
-(plus a novel variant of the fifth method). The first two alternative
-interaction measures were the sample correlations and the partial
-correlations between each pair of species' data vectors on the landscape
-(Albrecht and Gotelli 2001, Faisal et al. 2010). In the context of
-non-Gaussian data, the partial correlation can be thought of as a
-computationally efficient approximation to the full Markov network model
-(Loh and Wainwright 2013). This sort of model is very common for
-estimating relationships among genes and gene products (Friedman et al.
-2008). Because partial correlations are undefined for landscapes with
-perfectly-correlated species pairs, I used a regularized estimate based
-on James-Stein shrinkage, as implemented in the corpcor package's
-\texttt{pcor.shrink} function with the default settings (Schäfer et al.
-2014).
+I also evaluated six alternative methods: five from the existing
+literature, plus a novel combination of two of these methods. The first
+alternative interaction metric was the sample correlation between
+species' presence-absence vectors, which summarizes their marginal
+association. Next, I used partial correlations, which summarize species'
+conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
+2010). In the context of non-Gaussian data, the partial correlation can
+be thought of as a computationally efficient approximation to the full
+Markov network model (Loh and Wainwright 2013). This sort of model is
+very common for estimating relationships among genes and gene products
+(Friedman et al. 2008). Because partial correlations are undefined for
+landscapes with perfectly-correlated species pairs, I used a regularized
+estimate based on James-Stein shrinkage, as implemented in the corpcor
+package's \texttt{pcor.shrink} function with the default settings
+(Schäfer et al. 2014).
 
@@
 
-The fourth alternative method, described in Gotelli and Ulrich (2009),
-involved simulating new landscapes from a null model that retains the
-row and column sums of the original matrix (Strong et al. 1984). I used
-the \(Z\)-scores computed by the Pairs software described in Gotelli and
+The next method, described in Gotelli and Ulrich (2009), involved
+simulating new landscapes from a null model that retains the row and
+column sums of the original matrix (Strong et al. 1984). I used the
+\(Z\)-scores computed by the Pairs software described in Gotelli and
 Ulrich (2009) as my null model-based estimator of species interactions.
@@
 the ``true'' interaction strengths among all of the species pairs from
-the larger simulated landscapes. This determined which the methods
+the larger simulated landscapes. This determined which of the methods
 provide a consistent way to know how strong species interactions
@@
 errors determined the proportion of variance explained for different
-combinations of method and landscape size (compared with a null model
-that assumed all interaction strengths to be zero).
+combinations of model type and landscape size (compared with a null
+model that assumed all interaction strengths to be zero).
 
@@
 confirmed that the rosalia package's estimates of species' conditional
-relationships were reasonably robust to sampling variation (Appendix 4).
+relationships were robust to sampling variation for reasonably-sized
+landscapes (Appendix 4).
 
@@
 identifying positive species interactions between competing species, and
-can make precise statements about the conditions whre indirect
+can make precise statements about the conditions where indirect
 interactions will overwhelm direct ones. They also provide a simple
@@
 can be used to calculate the expected prevalence of a species in the
-absence of biotic influences (\(\frac{1}{1 + e^{-\alpha}}\); Lee and
-Hastie 2012). Competition's effect on prevalence in a Markov network can
-then be calculated by subtracting this value from the observed
+absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
+and Hastie 2012). Competition's effect on prevalence in a Markov network
+can then be calculated by subtracting this value from the observed
 prevalence (cf Figure 2D). This kind of insight would have been
@@
 Markov networks---particularly the Ising model for binary
-networks---have been studied in statistical physics for nearly a century
-(Cipra 1987), and the models' properties, capabilities, and limits are
-well-understood in a huge range of applications. Using the same
-framework for species interactions would thus allow ecologists to tap
-into an enormous set of existing discoveries and techniques for dealing
-with indirect effects, stability, and alternative stable states.
-Numerous other extensions are possible: for example, the states of the
-interaction network can be modeled as a function of the local abiotic
-environment (Lee and Hastie 2012), which would provide a rigorous and
-straightforward approach to the difficult and important task of
-incorporating whole networks of biotic interactions into species
-distribution models (Kissling et al. 2012, Pollock et al. 2014), leading
-to a better understanding of the interplay between biotic and abiotic
-effects on community structure. There are even methods (Whittam and
-Siegel-Causey 1981, Tjelmeland and Besag 1998) that would allow one
-species to affect the sign or strength of the relationship between two
-other species, tipping the balance between facilitation and exploitation
-(Bruno et al. 2003).
+networks---have been studied for nearly a century (Cipra 1987), and the
+models' properties, capabilities, and limits are well-understood in a
+huge range of applications. Using the same framework for species
+interactions would thus allow ecologists to tap into an enormous set of
+existing discoveries and techniques for dealing with indirect effects,
+stability, and alternative stable states. Numerous other extensions are
+possible: for example, the states of the interaction network can be
+modeled as a function of the local abiotic environment (Lee and Hastie
+2012), which would provide a rigorous and straightforward approach to
+the difficult and important task of incorporating whole networks of
+biotic interactions into species distribution models (Kissling et al.
+2012, Pollock et al. 2014), leading to a better understanding of the
+interplay between biotic and abiotic effects on community structure.
+There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
+Besag 1998) that would allow one species to affect the sign or strength
+of the relationship between two other species, tipping the balance
+between facilitation and exploitation (Bruno et al. 2003).
 
@@
 switching from null models to Markov networks (or increase it
-substantially by with linear or generalized linear approximations).
+substantially with linear or generalized linear approximations).
 Regardless of the methods ecologists ultimately choose, controlling for
@@
 \textbf{References:}
-
-\textbf{Figure captions}
-
-\textbf{Figure 1.} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, but more
-weakly (circled coefficient). \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled).
-
-\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
-The abiotic environment favors the occurrence of both species
-(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
-negative \(\beta\) coefficient linking these two species implies that
-they co-occur less than expected under independence. \textbf{B.}
-Relative probabilities of all four possible presence-absence
-combinations for Species 1 and Species 2. The exponent includes
-\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
-it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
-the species no longer competed for the same resources), then the
-reduction in competition would allow each species to increase its
-occurrence rate and the co-occurrence deficit would be eliminated.
-
-\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
-coefficients explained by each versus number of sampled locations.
-\textbf{B.} The \(Z\)-scores produced by the null model (``Pairs'') for
-each pair of species are very predictable from the correlation between
-the presence-absence vectors of those same species and from the number
-of sites on the landscape.
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
+
+\textbf{Figure captions}
+
+\setlength{\parskip}{3pt} \setlength{\parindent}{0em}
+\setlength{\leftskip}{0em}
+
+\textbf{Figure 1.} \textbf{A.} A small network of three competing
+species. The tree (top) tends not to co-occur with either of the two
+shrub species, as indicated by the strongly negative coefficient linking
+them. The two shrub species also compete with one another, but more
+weakly (circled coefficient). \textbf{B.} In spite of the competitive
+interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors
+positively correlated (circled). \textbf{C.} Controlling for the tree
+species' presence with a conditional method such as a partial covariance
+or a Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled).
+
+\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
+The abiotic environment favors the occurrence of both species
+(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
+negative \(\beta\) coefficient linking these two species implies that
+they co-occur less than expected under independence. \textbf{B.}
+Relative probabilities of all four possible presence-absence
+combinations for Species 1 and Species 2. The exponent includes
+\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
+it is absent (\(y_1 = 0\)). Similarly, the exponent includes
+\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
+\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
+constant \(Z\), ensures that the four relative probabilities sum to 1.
+In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
+the species no longer competed for the same resources), then the
+reduction in competition would allow each species to increase its
+occurrence rate and the co-occurrence deficit would be eliminated.
+
+\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
+coefficients explained by each method versus number of sampled
+locations. \textbf{B.} The \(Z\)-scores produced by the null model
+(``Pairs'') for each pair of species can be predicted using the
+correlation between the presence-absence vectors of those same species
+and from the number of sites on the landscape.
+
 \end{flushleft}

### 2015-08-05 Harris_2015/recovery.tex
submitted to Ecology

@@
 
-\begin{verbatim}
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0302 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0189 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0172 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0461 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0297 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0165 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0315 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.021 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0126 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0942 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0302 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0281 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0199 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0156 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0256 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0323 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0293 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0253 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0283 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0165 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0125 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0221 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0426 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0403 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0353 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0224 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0238 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0149 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0132 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0169 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0364 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0322 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0377 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.009 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0213 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.018 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.034 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0255 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0321 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0147 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0161 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0253 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0121 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0166 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0235 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0237 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0358 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0173 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0532 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1276 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.311 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3631 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2586 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1348 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1437 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3609 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0889 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1302 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0517 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1508 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1163 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1135 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1597 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2075 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1392 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.459 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1726 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1095 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1174 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1909 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1924 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2261 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1249 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.159 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1499 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1224 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1935 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2215 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1248 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1337 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0892 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1546 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0723 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1481 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1154 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1386 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1158 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1681 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1362 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2259 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1116 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.12 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1533 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1226 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1985 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3044 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1416 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.1534 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3426 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7236 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5589 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4883 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6616 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7645 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.494 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7008 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3284 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6008 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6168 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4011 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3676 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.2673 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6011 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6675 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6813 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4592 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6586 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6237 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3866 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4468 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.767 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6031 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.851 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6081 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5805 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5058 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4585 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6108 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4147 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5198 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4652 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6374 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.4167 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6822 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.3729 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5584 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5641 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7974 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7246 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5429 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6172 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6817 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.5112 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.6673 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7543 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.755 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7918 
-## 
-## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.7204
-\end{verbatim}
-
 \begin{Shaded}
@@
 \begin{verbatim}
-##           BayesComm               Pairs   partial BayesComm 
-##                12.7                15.4                18.5 
-##         correlation partial correlation                 GLM 
-##                18.6                27.5                29.9 
+##           BayesComm               Pairs         correlation 
+##                11.6                15.4                18.6 
+##   partial BayesComm partial correlation                 GLM 
+##                26.1                27.5                29.9 
 ##      Markov network 
@@
 ## correlation   BayesComm       Pairs 
-##        1.00        0.79       -0.92
+##        1.00        0.85       -0.92
 \end{verbatim}
@@
 ## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
-## R-squared is 1.77-1.95 times higher than from correlation (95% bootstrap CI)
+## R-squared is 1.79-1.95 times higher than from correlation (95% bootstrap CI)
 ## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
-## R-squared is 2.54-2.9 times higher than from BayesComm (95% bootstrap CI)
-## R-squared is 1.77-1.97 times higher than from partial BayesComm (95% bootstrap CI)
-## R-squared is 2.12-2.38 times higher than from Pairs (95% bootstrap CI)
+## R-squared is 2.72-3.28 times higher than from BayesComm (95% bootstrap CI)
+## R-squared is 1.29-1.37 times higher than from partial BayesComm (95% bootstrap CI)
+## R-squared is 2.12-2.37 times higher than from Pairs (95% bootstrap CI)
 \end{verbatim}

### 2015-08-05 Harris_2015/boot.R

@@ -1,3 +1,5 @@
+# Appendix 4: bootstrap estimates of model stability
+
 library(rosalia)
 
 set.seed(1)

### 2015-08-05 Harris_2015/manuscript-materials/pdf.sh

@@ -3,7 +3,7 @@ sed -i "" "s/:_/_/g" My\ Library.bib
 
 pandoc -o draft.tex --template=mytemplate.latex --bibliography=My\ Library.bib --csl=ecology.csl draft.Rmd
 
-pandoc -o draft.pdf --template=mytemplate.latex --bibliography=My\ Library.bib --csl=ecology.csl draft.Rmd
+pdflatex draft.tex
 
 pandoc -o Appendices.pdf --bibliography=My\ Library.bib --csl=ecology.csl Appendices.Rmd
 

### 2015-08-20 stochastic/chapter-3.Rmd
chapter 3 stuff

@@
+---
+output: pdf_document
+title: "Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation"
+---
+
 # Abstract
 
-Species interactions are believed to play an important role in community structure, but detecting their influence in observed communities has stymied ecologists for decades.
+Species interactions are believed to play an important role in community structure, but detecting their influence in the co-occurrence patterns of observed communities has stymied ecologists for decades.
 In recent years, several papers have suggested that ecologists could solve this long-standing problem by fitting Markov networks (undirected graphical models also known as Markov random fields) to the data by maximum likelihood or similar methods.
-This approach allows ecologists to control for extraneous factors when estimating the direct interactions between a given pair of species, and previous work has shown that this kind of control is required for reliable inferences.
+This approach allows ecologists to control for extraneous factors when estimating the direct interactions between a given pair of species, and previous work has shown that this kind of control is required for reliable inferences about interactions' sign and strength.
 Unfortunately, the likelihood function for Markov networks quickly becomes intractable once more than about 20 species are included in the model, due to the exponentially large number of possible combinations of presences and absences.
-Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and attempt to explain *all* of the co-occurrence patterns in ecological data sets with species interactions.
-In this paper, I introduce *stochastic approximation* as a method for fitting these models to large ecological networks where hundreds of species and multiple abiotic factors can all affect a given species’ occurrence probability.
+Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions.
+In this paper, I introduce "stochastic approximation" as a method for fitting these models to large ecological networks where hundreds of species and multiple abiotic factors can all affect a given species’ occurrence probability.
 This approach, which converges to the maximum likelihood estimate with probability one, greatly expands the range of data sets we can analyze and the number of ecological questions we can address with these methods.
@@
 To the extent that species interactions are important for community assembly, ecologists generally expect them to leave a signature on species' co-occurrence patterns.
-But using that signature to infer the underlying species interactions has been much harder.
-Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @strong_ecological_1984; @hastings_can_1987].
+But using that signature to infer the underlying species interactions from observational data has been much harder.
+Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @gilpin_factors_1982; @strong_ecological_1984; @hastings_can_1987].
 As documented below, ecologists have, for the most part, not solved these problems in the ensuing decades.
+Resolving these issues would solve a longstanding issue in community ecology and make a significant contribution to other fields, such as species distribution modeling [@kissling_towards_2012].
 
-During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among *thousands* of genes from their co-expression levels (e.g.
+During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among thousands of genes from their co-expression levels (e.g.
 @friedman_sparse_2008).
-The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients [@pollock_understanding_2014; see @faisal_inferring_2010 for a recent exception].
-As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (Figure 1).
+The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al.
+2014; see Faisal et al.
+2010 and Harris 2015 for two recent exceptions).
+As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (e.g.
+abiotic influences and indirect biotic effects; Figure 1).
 When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
@@
 A number of procedures have been developed for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables, i.e.
-for estimating the *conditional* relationship between them.
+for estimating the conditional relationship between them.
 The most familiar of these is the partial correlation.
-Rather than describing the *overall* relationship between two variables across *all* conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.
+Rather than describing the overall relationship between two variables across all conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.
 @harris_estimating_2015 found that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
 
-Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @fort_statistical_2013; @harris_estimating_2015).
+Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @harris_estimating_2015).
 A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network).
-Unfortunately, Markov networks have an intractable likelihood function whose computational complexity more than doubles each time a new species is added to the model.
-Thus, while @azaele_inferring_2010 and @harris_estimating_2015 were able to optimize an exact objective function to maximize the constrained likelihood for networks of up to 20 species, extending these approaches to networks of 50-100 species would be completely infeasible.
-If ecologists want the model to account for abiotic variation among sites on the landscape, then these expensive computations need to be repeated independently for every site, increasing the computational burden even further.
+This approach is optimal in the sense that it produces a model that matches the observed properties of the data (i.e.
+the occurrence and co-occurrence rates), and that it does so with the fewest possible parameters or constraints [i.e.
+the Markov network has the most information entropy of any possible model satisfying its constraints; azaele_inferring_2010].
+As shown in @harris_estimating_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.
 
-In this paper, I present a different way of optimizing the likelihood, called *stochastic approximation* [@robbins_stochastic_1951; @salakhutdinov_efficient_2012], which replaces the intractable computations with tractable Monte Carlo estimates of the same quantities.
-Despite the sampling error introduced by this substitution, stochastic approximation provides strong guarantees for eventual convergence to the maximum likelihood estimate.
-This change in approach makes it feasible to estimate the interactions among hundreds of species from observational data while also accounting for possible responses to abiotic factors, possibly giving ecologists some power to distinguish between species pairs that co-occur due to shared environmental tolerances from those that co-occur due to direct interactions like mutualism.
+Unfortunately, Markov networks have an intractable likelihood function whose computational difficulty more than doubles each time a new species is added to the model.
+This exponential growth in the likelihood function's complexity means that the methods that worked for @azaele_inferring_2010 and @harris_estimating_2015 with networks of 20 species would be completely infeasible for networks with 50 species, requiring over a billion times more computational effort ($2^{50}/2^{20} \approx 10^{9}$).
+Extending the method to account for abiotic variation among sites on the landscape would require repeating these expensive computations independently for every site, increasing the computational burden even further.
+If ecologists want to apply these methods to larger problems, a different model-fitting algorithm would be needed.
+
+In this paper, I present a different way of optimizing the likelihood, called "stochastic approximation" [@robbins_stochastic_1951; @salakhutdinov_efficient_2012], which replaces the intractable computations with tractable Monte Carlo estimates of the same quantities.
+Despite the sampling error introduced by this substitution, stochastic approximation provides strong guarantees for eventual convergence to the maximum likelihood estimate [@younes_convergence_1999; @salakhutdinov_efficient_2012].
+This change in approach makes it feasible to estimate the interactions among hundreds of species from observational data while also accounting for possible responses to abiotic factors (such as those used in species distribution models), giving ecologists some power to distinguish between species pairs that co-occur due to shared environmental tolerances from those that co-occur due to direct interactions like mutualism.
 
@@
 
-As discussed in @azaele_inferring_2010, @fort_statistical_2013, and @harris_estimating_2015, Markov networks such as the Ising model [@cipra_introduction_1987] can be used to describe community structure as follows.
-Each species is represented by a binary random variable, describing whether it is present or absent.
+As discussed in @azaele_inferring_2010 and @harris_estimating_2015, Markov networks such as the Ising model [@cipra_introduction_1987] can be used to describe community structure as follows.
+Each species is represented by a binary random variable, describing whether it is present (1) or absent (0).
 A species’ conditional probability of presence under a given set of biotic and abiotic conditions depends on its coefficients.
-The coefficients linking species to one another describe their conditional associations: *all else equal*, a species will occur less often in the presence of its competitors and other species with which it is negatively associated, and more often in the presence of its mutualists.
+The coefficients linking species to one another describe their conditional associations: all else equal, a species will occur less often in the presence of its competitors and other species with which it is negatively associated, and more often in the presence of its mutualists:
 
-[[add math for conditional distributions; same as in chapter 2]]
+$$ p(y_i | \vec{y}_{j \neq i}) = \mathrm{logistic}\Big(\alpha_i + \sum_{j \neq i}{\beta_{ij}y_j}\Big), $$
 
-These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences [[Appendix?]].
+where $y_i$ is 1 when species $i$ is present and 0 when it is absent, $\alpha_i$ is an intercept term, $\beta_ij$ represents the conditional relationship between species $i$ and species $j$, and $\mathrm{logistic}(x) = e^x / (1 + e^x)$.
+These conditional probabilities can be combined to calculate the probability of observing any given combination of presences and absences, up to a multiplicative constant [@lee_learning_2012; @murphy_machine_2012]:
 
-[[Add math for joint distribution; same as in chapter 2]]
+$$p(\vec{y}) \propto \mathrm{exp}\Big({\sum_{i}{\alpha_i y_i} + \sum_{j \neq i}}{\beta_{ij} y_i y_j}\Big).$$
 
-Unfortunately, the extremely large number of possible assemblages makes this joint probability intractable when the number of species is large.
+Unfortunately, the exponentially large number of possible assemblages makes the normalizing constant for this joint probability distribution intractable when the number of species is larger than 20 or so.
 However, it is generally straightforward to simulate examples of assemblages that are consistent with graphical models such as these, using Monte Carlo techniques like Gibbs sampling [@salakhutdinov_efficient_2012].
@@
 
-The Markov networks fit in @harris_estimating_2015 treated the $\alpha_i$ terms from Equations [[1 and 2]] as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.
+The Markov networks fit in @harris_estimating_2015 treated the $\alpha_i$ terms from Equations 1 and 2 as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.
 In this paper, I extend the model so that the local abiotic conditions can also affect species occurrence probabilities directly.
@@
 By estimating the coefficients associated with each of these $x$ variables for each species, we can account for their responses to the abiotic environment as well as to one another.
-In other fields, a model like this one, where a Markov network’s parameters depend on external factors is called a conditional random field [refs]; these models have been used in a variety of contexts, from language models to image processing [refs].
+In other fields, a model like this one, where a Markov network’s parameters depend on external factors is called a conditional random field [@lee_learning_2012]; these models have been used in a variety of contexts, from language models to image processing [@murphy_machine_2012].
 
@@
 
-A Markov network describes a probability distribution over possible assemblages, and this distribution can be summarized without loss of information by its *sufficient statistics*.
-For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [[*Appendix, not yet written*]].
-One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the average sufficient statistics produced by samples from the model [[ref?]].
+A Markov network describes a probability distribution over possible assemblages.
+Since the model is a member of the exponential family it can be summarized without loss of information by its "sufficient statistics".
+For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [@azaele_inferring_2010; @lee_learning_2012; @murphy_machine_2012].
+One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the expected sufficient statistics under the model [@bickel_mathematical_1977].
 Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by climbing the likelihood surface.
@@
 
-In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates this different exactly, averaging over *all* possible presence-absence combinations.
-This approach is intractable for larger networks, as the number of possibilities grows exponentially with the size of the network.
-Stochastic approximation [refs] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees [refs].
-The procedure iterates through the following three steps as many times as needed (5000 for these analyses):
+In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates each value exactly, averaging over all possible presence-absence combinations.
+This approach is infeasible for larger networks, as the number of possibilities grows exponentially with the size of the network.
+Stochastic approximation [@robbins_stochastic_1951; @salakhutdinov_efficient_2012] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.
+The procedure iterates through the following three steps as many times as needed (50,000 for these analyses):
 
 1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
+
 2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
+
 3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.
@@
 
-The approximate likelihood gradients in Step 2 match the ones from the Appendix of @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
-These gradients were augmented with a momentum term equal to 0.9 [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 1.0 (for environmental responses) or 0.1 (for pairwise interactions); the pairwise interaction terms were regularized more heavily to improve identifiability and reduce overfitting for this large set of 31,125 coefficients.
+The approximate likelihood gradients in Step 2 match the ones from @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
+These gradients were augmented with a momentum term [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 2.0 (for environmental responses) or 0.5 (for pairwise interactions).
 
-The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 500/(500 - t - 1)$, which satisfies the criteria in @powell_approximate_2007.
-$\alpha_0$ was 1.0 for species’ intercepts and environmental responses.
+The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 1000/(999 + t)$, which satisfies the criteria in @powell_approximate_2007.
+The scaling factor, $\alpha_0$, was 1.0 for species’ intercepts and environmental responses.
 Because of the enormous number of pairwise interaction coefficients, I set $\alpha_0$ to 0.1 for the $\beta$ parameters.
@@
 
-To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated 10 landscapes with 20 species and 500 sites, using the Gibbs sampling approach described in @harris_estimating_2015.
-These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
-For each of these simulated landscapes, I then fit the same model using both the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.
+To demonstrate that my stochastic approximation implementation could converge to the global optimum, I first simulated a landscape with 20 species and 500 sites, using Gibbs sampling.
+These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@harris_rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
+For each of these simulated landscapes, I then fit the same Markov network model to the simulated data using the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.
 
 I then simulated a large landscape with 250 species and 2500 sites, representing a large data set roughly the size of the North American Breeding Bird Survey.
-Each site on the landscape was represented by 5 environmental variables, which were drawn independently from Gaussian distributions with mean 0 and standard deviation 2.
+Each site on the landscape was represented by 5 environmental variables, which were drawn independently from Gaussian distributions with mean 0 and standard deviation 2.5.
 
@@
 The coefficients determining species’ responses to the environment were each drawn from standard normal distributions.
-The coefficients describing species’ pairwise interactions were drawn from a mixture of two zero-mean Gaussian distributions.
-90% of the interaction coefficients were drawn from a distribution with standard deviation 0.2, while the remainder were drawn from a distribution with standard deviation 1.0.
+The coefficients describing species’ pairwise interactions were drawn from a mixture of normals (see Appendix) so that most interactions were weak and negative, but a few strong positive and interactions also occurred.
 
@@
 its probability of occurrence against any given backdrop of biotic and abiotic features).
-I used these conditional probabilities to produce examples of communities that were consistent with the competition parameters and with the local environmental variables via Markov chain Monte Carlo (specifically, `r n_gibbs` rounds of Gibbs sampling; see code in the Appendix).
+I used these conditional probabilities to produce examples of communities that were consistent with the competition parameters and with the local environmental variables via Markov chain Monte Carlo (specifically, 1000 rounds of Gibbs sampling; see code in the Appendix).
 I then used the methods from the next section to attempt to infer the underlying parameters from the simulated data.
-
-## Comparison with directed models
-
-[[In response to Chapter 2, reviewers suggested comparisons with two other methods based on directed acyclic graphs.
-The first one has a latent multivariate Gaussian, and relies on the correlation structure of that Gaussian to help determine which species interact.
-The second one is based on sparse directed graphs among the species in the network.
-Should I address one or both of these approaches in this chapter as well?]]
 
@@
 
-![Stochastic approximation decreases the model’s error over time.
+![Stochastic approximation improves the model over time.
 **A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation reaches the global optimum very quickly.
-**B.** With larger networks, where the “true” parameter values that generated the network are known but the maximum likelihood estimates for the observed landscape are not, stochastic approximation initially drives the error down very quickly, and then slowly approaches the apparent optimum.](convergence.pdf)
+Within a few minutes, the squared deviation between the approximate and exact estimates disappears.
+**B.** With larger networks, where the maximum likelihood estimate cannot be calculated exactly, stochastic approximation converges to an apparent optimum that explains most (but not all) of the variation in the "true" parameter values.](convergence.pdf)
 
-For the smaller communities, the squared deviations between the exact estimates produced by the `rosalia` package and the ones produced by stochastic approximation quickly decayed to negligible levels [[Figure 2A]], indicating that the stochastic approximation procedure was implemented correctly and worked as the mathematical theory predicts.
+For the smaller communities, the squared deviations between the exact estimates produced by the `rosalia` package and the ones produced by stochastic approximation quickly decayed to negligible levels (Figure 2A), indicating that the stochastic approximation procedure was implemented correctly and worked as the mathematical theory predicts.
 
-![“True” network parameters for all 31,125 pairwise interactions versus the estimates from the three methods presented.
-Blue lines represent linear regressions through each cloud of points.](estimates.pdf)
+!["True" versus estimated network parameters for all 250 species' responses to the other 249 species in the data set (**A.** $R^2 = .71$) and to the 5 abiotic variables (**B.** $R^2 = .95$).
+](estimates.pdf)
 
-For the larger landscape, I found that the stochastic approximation approach achieved reasonably good performance after less than a minute of optimization [Figure 2B].
-With additional time, was able to recover most of the variance in species’ pairwise interactions from observational data ($R^2 = 0.66$; Figure 3A), and nearly all of the variance in their responses to environmental variables ($R^2 = 0.95$, not shown).
-
-While the maximum likelihood estimates produced by the Markov network worked well, the sample correlation, explained a negligible fraction of the variance in species’ pairwise interactions ($R^2 = 0.03$, Figure 3B).
-To the extent that ecologists’ other test statistics for co-occurrence are similar to the sample correlation [[*evidence to be added to chapter 2*]], they can be expected to perform equally poorly.
-As shown previously, the partial correlation was a much better indicator of species’ true interactions than the sample correlation ($R^2 = 0.44$ Figure 3C).
+For the larger landscape, I found that the stochastic approximation approach achieved reasonably good performance after ten minutes of optimization (Figure 2B).
+After 50,000 iterations (about 5.4 hours on my laptop), the model was able to recover more than two thirds of the variance in species’ pairwise interactions($R^2 = 0.71$; Figure 3A), and nearly all of the variance in their responses to environmental variables ($R^2 = 0.95$; Figure 3B).
 
@@
 Fortunately, a number of good regularizers have been proposed.
-Of these, some of the most interesting options include $L_1$ regularization, which ensures that many of the estimated coefficients are exactly zero [[refs]].
-This sparsity matches ecologists’ intuition that many species from different guilds are unlikely to interact much at all [[refs]], and also produces computational benefits for model estimation.
-Of course, the best regularizers will take advantage of prior ecological knowledge (e.g.
+Of these, some of the most interesting options include $L_1$ regularization, which ensures that many of the estimated coefficients are exactly zero [@lee_learning_2012].
+This sparsity matches ecologists’ intuition that many species from different guilds are unlikely to interact much at all [@faisal_inferring_2010], and also produces computational benefits for model estimation.
+Of course, the best regularizers will take advantage of specific ecological knowledge (e.g.
 from field experiments, natural history, or trait data) to provide information about individual pairwise interactions, rather than about their overall distribution.
 
-Richer forms of data, e.g.
-from observational time series or controlled experiments, could help ecologists recover more information about species interactions than is possible with the kinds of observational data analyzed here: in particular, these richer forms of data will be necessary to answer questions about which species in a positively associated pair is facilitating the other or which species in a negatively associated pair is excluding the other.
+These richer sources of information will be especially important for cases where ecologists expect that the interactions between species are asymmetric (e.g.
+where only one species is affected by an interaction or where one species benefits at a cost to the other).
+With snapshot observations of species' spatial associations, as discussed here, interactions must be reduced to a single number describing the net association, as in undirected models such as the Markov networks presented here [@schmidt_modeling_2012].
+With richer data types, however, ecologists could more easily learn about asymmetric interactions such as facilitation, predation, and parasitism.
 
-Ecologists could expand beyond simple Markov networks in a number of ways that would improve their ability to address a wider range of questions.
+Even without better data, ecologists have a number of options for expanding beyond simple Markov networks in a number of ways that would improve their ability to address a wider range of questions.
 This paper demonstrated that it is possible to simultaneously estimate species’ responses to the abiotic and environment and to one another, but many other extensions are possible.
 For example, ecologists could condition the model on variables whose values have not been measured (e.g.
-partially-observed Markov networks, or the approximate networks in the `mistnet` package).
+partially-observed Markov networks, or the approximate networks in the `mistnet` package; cf.
+@pollock_understanding_2014).
 This would allow ecologists to account for measurement error and for other ecologically-important factors that can be difficult to measure.
-Ecologists should also explore higher-order networks, where one species’ presence can affect the relationship between two other species.
+Ecologists should also explore higher-order networks, where one species’ presence can affect the relationship between two other species [@whittam_species_1981; @tjelmeland_markov_1998].
 

### 2015-08-20 stochastic/stochastic-recovery.Rmd
chapter 3 stuff

--- 2015-06-16 stochastic/stochastic-recovery.Rmd
@@
 n_loc = 2500
-n_env = 3
-n_gibbs = 1000
+n_env = 5
+n_gibbs = 5000
 
-true_beta_vec = rnorm(
-  choose(n_spp, 2), 
-  sd = abs(rnorm(choose(n_spp, 2))) / 5
+type_frequencies = rmultinom(1, size = choose(n_spp, 2), prob = c(.2, .5, .3))
+
+true_beta_vec = sample(
+  c(
+    rnorm(type_frequencies[[1]], 0, 2), 
+    rnorm(type_frequencies[[2]], 0, .2), 
+    rnorm(type_frequencies[[3]], -.5, .5)
   )
+)
+
+plot(density(true_beta_vec))
 
@@
 
-true_intercepts = rnorm(n_spp)
+true_env = matrix(rnorm(n_loc * n_env, 0, 2.5), nrow = n_loc)
 
-true_env = matrix(rnorm(n_loc * n_env, 0), nrow = n_loc)
+true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 1), nrow = n_env)
 
-true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 0.5), nrow = n_env)
+true_alpha_species = rnorm(n_spp, 2)
 
-true_alpha = true_env %*% true_alpha_env
+true_alpha = true_env %*% true_alpha_env %plus% true_alpha_species
 ```
@@
 alpha_env = delta_alpha_env = matrix(0, nrow = n_env, ncol = n_spp)
-alpha_species = delta_alpha_species = qlogis(colMeans(y))
+alpha_species = qlogis((colSums(y) + 1) / (nrow(y) + 2))
+delta_alpha_species = rep(0, n_spp)
 beta = delta_beta = matrix(0, nrow = n_spp, ncol = n_spp)
@@
 ```{r}
-initial_learning_rate = 0.1
-momentum = 0.9
+# Are there lurking factors of two because I use the whole symmetric matrix?
+beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
+alpha_env_prior = rosalia::make_logistic_prior(scale = 2)$log_grad 
+alpha_species_prior = rosalia::make_logistic_prior(scale = 2)$log_grad
+```
+
+```{r}
+initial_learning_rate = 1
+maxit = 50000
+start_time = as.integer(Sys.time())
 
@@
 
-for(i in 1:n_gibbs){
+r2s = numeric(maxit)
+times = integer(maxit)
+
+for(i in 1:maxit){
   pb$tick()
@@
 
-  # Update learning rate
-  learning_rate = initial_learning_rate * 1000 / (999 + i)
+  # Update learning rate and momentum
+  learning_rate = initial_learning_rate * 1000 / (998 + 1 + i)
+  momentum = .9 * (1 - 1/(.1 * i + 2))
 
@@
   # Calculate the gradient with respect to alpha and beta
-  delta_stats = y_stats - y_sim_stats
-  beta_grad = (delta_stats - 10 * beta) / 10 / n_loc # Subtract beta as weight decay
+  stats_difference = y_stats - y_sim_stats
+  beta_grad = (stats_difference + beta_prior(beta)) / n_loc
 
-  alpha_species_grad = diag(beta_grad) / n_loc
+  alpha_species_grad = (diag(stats_difference) + alpha_species_prior(alpha_species)) / n_loc
   diag(beta_grad) = 0
-  alpha_env_grad = (y_env_stats - y_sim_env_stats - alpha_env)  / n_loc
+  y_env_difference = y_env_stats - y_sim_env_stats
+  alpha_env_grad = (y_env_difference + alpha_env_prior(alpha_env))  / n_loc
 
   # Calculate parameter updates
-  delta_beta = beta_grad * learning_rate + momentum * delta_beta
-  delta_alpha_species = alpha_species_grad * learning_rate # no momentum
-  delta_alpha_env = alpha_env_grad * learning_rate + momentum  * delta_alpha_env
+  delta_beta = beta_grad * learning_rate / 10 + 
+    momentum * delta_beta
+  delta_alpha_species = alpha_species_grad * learning_rate + 
+    momentum  * delta_alpha_species
+  delta_alpha_env = alpha_env_grad * learning_rate + 
+    momentum  * delta_alpha_env
 
@@
 
-  if(i %% 10 == 0){
-    plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
-         pch = 16, cex = .5, col = "#00000050",
-         xlab = "Estimated interaction strength",
-         ylab = "\"True\" interaction strength",
-         main = paste(
-           "R^2 =", 
-           signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
-          )
-    )
-    abline(v = 0, h = 0)
-  }
+  r2s[i] = cor(true_beta[upper.tri(true_beta)], beta[upper.tri(beta)])^2
+  times[i] = as.integer(Sys.time()) - start_time
 }
@@
 ```{r}
-pdf("poster-beta.pdf", width = 4, height = 5)
-plot(beta[upper.tri(beta)], true_beta[upper.tri(beta)],
-     pch = 16, cex = .4, col = "#00000020",
-     xlab = "Estimated interaction strength",
-     ylab = "\"True\" interaction strength",
-     main = paste0("Recoverd pairwise interactions", "\nN = ", choose(n_spp, 2)),
-     las = 1,
-     cex.axis = 0.75
+library(cowplot)
+library(ggplot2)
+out = plot_grid(
+  ggplot(NULL, aes(x = beta[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
+    stat_binhex() + 
+    xlab("Estimated coefficient value") + 
+    ylab("\"True\" coefficient value") + 
+    stat_hline(yintercept = 0, size = 1/8) + 
+    stat_vline(xintercept = 0, size = 1/8) + 
+    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "log10") +
+    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
+    coord_equal(),  
+  ggplot(NULL, aes(x = c(alpha_env), y = c(true_alpha_env))) + 
+    stat_binhex() + 
+    xlab("Estimated coefficient value") + 
+    ylab("\"True\" coefficient value") + 
+    stat_hline(yintercept = 0, size = 1/8) + 
+    stat_vline(xintercept = 0, size = 1/8) + 
+    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "log10") +
+    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
+    coord_equal(),  
+  nrow = 1,
+  labels = c("A. Pairwise biotic coefficients (n = 31125)", "B. Abiotic coefficients (n = 1250)")
 )
-text(
-  -0.7, 
-  1,
-  paste(
-    "R^2 =", 
-    signif(cor(beta[upper.tri(beta)], true_beta[upper.tri(beta)])^2, 2)
-  )
-)
+save_plot("stochastic/estimates.pdf", out, base_aspect_ratio = 2, base_height = 6)
 
-dev.off()
+```
+
+## Small landscape
+
+```{r}
+small_n_spp = 20
+small_n_loc = 500
+small_true_beta = true_beta[1:small_n_spp, 1:small_n_spp]
+small_true_alpha = true_alpha_species - 2
+small_y = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)
+
+for(i in 1:n_gibbs){
+  for(j in sample.int(small_n_spp)){
+    small_y[,j] = rbern(logistic(small_true_alpha[j] + small_y %*% small_true_beta[ , j]))
+  }
+}
+
 ```
@@
 ```{r}
-pdf("poster-alpha.pdf", width = 4, height = 5)
-plot(alpha_env, true_alpha_env,
-     pch = 16, cex = .4, col = "#00000090",
-     xlab = "Estimated environmental coefficient",
-     ylab = "\"True\" environmental coefficient",
-     main = paste0("Recoverd environmental coefficients", "\nN = ", length(alpha_env)),
-     las = 1,
-     cex.axis = 0.75
-)
-text(
-  -1, 
-  1,
-  paste(
-    "R^2 =", 
-    signif(cor(c(alpha_env), c(true_alpha_env))^2, 2)
-  )
-)
+exact_time = system.time({
+  exact = rosalia(small_y, prior = make_logistic_prior(scale = 1), maxit = 500)
+})
+
+mle_beta = exact$beta
+```
+
+```{r}
+maxit_small = 50000
+start_time_small = as.numeric(Sys.time())
+
+beta_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad
+alpha_species_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad
+
+alpha_small = qlogis((colSums(small_y) + 1) / (nrow(small_y) + 2))
+delta_alpha_small = rep(0, small_n_spp)
+beta_small = delta_beta_small = matrix(0, nrow = small_n_spp, ncol = small_n_spp)
+
+y_sim_small = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)
+
+y_stats = crossprod(small_y)
+
+mses = numeric(maxit_small)
+small_times = numeric(maxit_small)
+
+for(i in 1:maxit_small){
+  ##############################
+  # Gibbs sampling for predicted species composition
+  ##############################
+
+  # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
+  for(j in sample.int(small_n_spp)){
+    y_sim_small[,j] = rbern(logistic(alpha_small[j] + y_sim_small %*% beta_small[ , j]))
+  }
+
+  ##############################
+  # Stochastic approximation for updating alpha and beta
+  ##############################
+
+  # Update learning rate and momentum
+  learning_rate = initial_learning_rate * 1000 / (998 + 1 + i)
+  momentum = .9 * (1 - 1/(.1 * i + 2))
+
+  # Calculate sufficient statistics
+  y_sim_stats = crossprod(y_sim_small)
+
+  # Calculate the gradient with respect to alpha and beta
+  stats_difference = y_stats - y_sim_stats
+  beta_grad = (stats_difference + beta_prior_small(beta_small)) / small_n_loc
+
+  alpha_species_grad = (diag(stats_difference) +
+                          alpha_species_prior_small(alpha_small)) / small_n_loc
+  diag(beta_grad) = 0
+
+  # Calculate parameter updates
+  delta_beta = beta_grad * learning_rate + 
+    momentum * delta_beta_small
+  delta_alpha_species = alpha_species_grad * learning_rate + 
+    momentum  * delta_alpha_small
+
+  beta_small = beta_small + delta_beta
+  alpha_small = alpha_small + delta_alpha_species
+
+  mses[i] = mean((beta_small[upper.tri(beta_small)] - exact$beta[upper.tri(exact$beta)])^2)
+  small_times[i] = as.numeric(Sys.time()) - start_time_small
+}
+```
+
+## Plotting results
+
+```{r}
+pdf("stochastic/convergence.pdf", width = 10, height = 7)
+par(mfrow = c(1, 2), las = 1)
+
+plot(small_times, mses, type = "l", ylab = "mean square deviation from MLE\n(log scale)", xlab = "time (seconds)", main = "A.
+Convergence to known MLE\n(small network)", yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, col = "blueviolet", xlim = range(c(-.1, small_times)))
+
+plot(c(0, times / 60 / 60), c(0, r2s), type = "l", xlab = "time (hours)", ylab = "R^2",ylim = c(0, max(r2s) * 1.04), main = "B.
+Apparent convergence to unknown MLE\n(large network)", yaxs = "i", xaxs = "i", lwd = 2, xlim = range(c(-.05, times / 60 / 60)), bty = "l", col = "blueviolet")
+abline(h = max(r2s), col = "#00000080", lty = 2, lwd = 2)
 dev.off()
-
 ```

### 2015-10-15 Harris_2015/manuscript-materials/draft.Rmd
clean up directory

@@
 paper is to resolve this long-standing problem.
+
+![**Figure 1.** **A.** A small network of three competing species. The tree (top) tends not to
+co-occur with either of the two shrub species, as indicated by the strongly
+negative coefficient linking them. The two shrub species also compete with one
+another, but more weakly (circled coefficient). **B.** In spite of the
+competitive interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors positively
+correlated (circled). **C.** Controlling for the tree species' presence with a
+conditional method such as a partial covariance or a Markov network leads to
+correct identification of the negative shrub-shrub interaction (circled).](figures/Figure_1.pdf)
 
@@
 negatively-associated species (exactly as an ecologist might expect).
+
+![**Figure 2.**
+**A.** A small Markov network with two species.  The abiotic environment favors
+the occurrence of both species ($\alpha >0$), particularly species 2
+($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
+species implies that they co-occur less than expected under independence.
+**B.** Relative probabilities of all four possible presence-absence combinations
+for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
+is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
+exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
+$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
+ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
+about 18.5.  **C.** We can find the expected frequencies of all possible
+co-occurrence patterns between the two species of interest. **D.** If
+$\beta_{12}$ equaled zero (e.g. if the species no longer competed for the same
+resources), then the reduction in competition would allow each species to
+increase its occurrence rate and the co-occurrence deficit would be eliminated.](figures/Figure_2.pdf)
 
@@
 
+![**Figure 3.**
+**A.** Proportion of variance in interaction coefficients explained by each
+method versus number of sampled locations.
+**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
+species can be predicted using the correlation between the presence-absence
+vectors of those same species and from the number of sites on the landscape.](figures/performance.pdf)
+
 The three methods that did not attempt to control for indirect interactions all
@@
 
-**Figure captions**
-
-**Figure 1.**
-**A.** A small network of three competing species. The tree (top) tends not to
-co-occur with either of the two shrub species, as indicated by the strongly
-negative coefficient linking them. The two shrub species also compete with one
-another, but more weakly (circled coefficient). **B.** In spite of the
-competitive interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors positively
-correlated (circled). **C.** Controlling for the tree species' presence with a
-conditional method such as a partial covariance or a Markov network leads to
-correct identification of the negative shrub-shrub interaction (circled).
-
-**Figure 2.**
-**A.** A small Markov network with two species.  The abiotic environment favors
-the occurrence of both species ($\alpha >0$), particularly species 2
-($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
-species implies that they co-occur less than expected under independence.
-**B.** Relative probabilities of all four possible presence-absence combinations
-for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
-is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
-exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
-$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
-ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
-about 18.5.  **C.** We can find the expected frequencies of all possible
-co-occurrence patterns between the two species of interest. **D.** If
-$\beta_{12}$ equaled zero (e.g. if the species no longer competed for the same
-resources), then the reduction in competition would allow each species to
-increase its occurrence rate and the co-occurrence deficit would be eliminated.
-
-**Figure 3.**
-**A.** Proportion of variance in interaction coefficients explained by each
-method versus number of sampled locations.
-**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
-species can be predicted using the correlation between the presence-absence
-vectors of those same species and from the number of sites on the landscape.
-
 \setlength{\parskip}{0pt}

### 2015-10-15 Harris_2015/manuscript-materials/draft.tex
clean up directory

@@
-\documentclass[12pt,]{article}
-\usepackage[compact]{titlesec}
-\usepackage{fancyhdr}
-\usepackage{hanging}
-\pagestyle{fancy}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{setspace}
-\usepackage{lineno}
-\linenumbers
-\usepackage{setspace}
-\usepackage{geometry}
-\usepackage{ifthen}
-\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{4pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{0}
-
-\date{}
-
-\begin{document}
-
-\begin{spacing}{1.9}
-\begin{flushleft}
 \chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
@@
 affect community assembly, ecologists might expect to find signatures of
-these interactions in species composition data (MacArthur 1958, Diamond
-1975). Despite decades of work and several major controversies, however
-(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
-al. 2013), existing methods for detecting competition's effects on
-community structure are unreliable (Gotelli and Ulrich 2009). In
-particular, species' effects on one another can become lost in the
-complex web of direct and indirect interactions in real assemblages. For
-example, the competitive interaction between the two shrub species in
-Figure 1A can become obscured by their shared tendency to occur in
-unshaded areas (Figure 1B). While ecologists have long known that
-indirect effects can overwhelm direct ones at the landscape level
-(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
-inferenes from observational data do not control for these effects (e.g.
-Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
-Pollock et al. 2014). To the extent that indirect interactions like
-those in Figure 1 are generally important (Dodson 1970), existing
-methods will thus not generally provide much evidence regarding species'
-direct effects on one another. The goal of this paper is to resolve this
+these interactions in species composition data
+\citep{macarthur_population_1958, diamond_island_1975}. Despite decades
+of work and several major controversies, however
+\citep{lewin_santa_1983, strong_ecological_1984, gotelli_swap_2003, connor_checkered_2013},
+existing methods for detecting competition's effects on community
+structure are unreliable \citep{gotelli_empirical_2009}. In particular,
+species' effects on one another can become lost in the complex web of
+direct and indirect interactions in real assemblages. For example, the
+competitive interaction between the two shrub species in Figure 1A can
+become obscured by their shared tendency to occur in unshaded areas
+(Figure 1B). While ecologists have long known that indirect effects can
+overwhelm direct ones at the landscape level
+\citep{dodson_complementary_1970, levine_competitive_1976}, the vast
+majority of our methods for drawing inferenes from observational data do
+not control for these effects
+\citep[e.g.][]{diamond_island_1975, strong_ecological_1984, gotelli_empirical_2009, veech_probabilistic_2013, pollock_understanding_2014}.
+To the extent that indirect interactions like those in Figure 1 are
+generally important \citep{dodson_complementary_1970}, existing methods
+will thus not generally provide much evidence regarding species' direct
+effects on one another. The goal of this paper is to resolve this
 long-standing problem.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/Figure_1.pdf}
+\caption{\textbf{Figure 1.} \textbf{A.} A small network of three
+competing species. The tree (top) tends not to co-occur with either of
+the two shrub species, as indicated by the strongly negative coefficient
+linking them. The two shrub species also compete with one another, but
+more weakly (circled coefficient). \textbf{B.} In spite of the
+competitive interactions between the two shrub species, their shared
+tendency to occur in locations without trees makes their occurrence
+vectors positively correlated (circled). \textbf{C.} Controlling for the
+tree species' presence with a conditional method such as a partial
+covariance or a Markov network leads to correct identification of the
+negative shrub-shrub interaction (circled).}
+\end{figure}
 
@@
 need to be controlled for. Nevertheless, null models have dominated this
-field for more than three decades (Strong et al. 1984, Gotelli and
-Ulrich 2009).
+field for more than three decades
+\citep{strong_ecological_1984, gotelli_empirical_2009}.
 
-Following recent work by Azaele et al. (2010) and Fort (2013), this
-paper shows that Markov networks (undirected graphical models also known
-as Markov random fields; Murphy 2012) can provide a framework for
+Following recent work by \citet{azaele_inferring_2010} and
+\citet{fort_statistical_2013}, this paper shows that Markov networks
+\citep[undirected graphical models also known as Markov random
+fields;][]{murphy_machine_2012} can provide a framework for
 understanding the landscape-level consequences of pairwise species
@@
 matrices. Markov networks have been used in many scientific fields in
-similar contexts for decades, from physics (where nearby particles
-interact magnetically; Cipra 1987) to spatial statistics (where adjacent
-grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
-While community ecologists explored some related approaches in the
-1980's (Whittam and Siegel-Causey 1981), they used severe approximations
-that led to unintelligible results (e.g. ``probabilities'' greater than
-one; Gilpin and Diamond 1982).
+similar contexts for decades, from physics \citep[where nearby particles
+interact magnetically;][]{cipra_introduction_1987} to spatial statistics
+\citep[where adjacent grid cells have correlated
+values;][]{harris_contact_1974, gelfand_modelling_2005}. While community
+ecologists explored some related approaches in the 1980's
+\citep{whittam_species_1981}, they used severe approximations that led
+to unintelligible results \citep[e.g. ``probabilities'' greater than
+one;][]{gilpin_factors_1982}.
 
@@
 problems in community ecology, e.g.~quantifying the overall effect of
-species interactions on occurrence rates (Roughgarden 1983) and
-disentangling the effects of biotic versus abiotic interactions on
-species composition (Kissling et al. 2012, Pollock et al. 2014).
+species interactions on occurrence rates
+\citep{roughgarden_competition_1983} and disentangling the effects of
+biotic versus abiotic interactions on species composition
+\citep{kissling_towards_2012, pollock_understanding_2014}.
 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/Figure_2.pdf}
+\caption{\textbf{Figure 2.} \textbf{A.} A small Markov network with two
+species. The abiotic environment favors the occurrence of both species
+(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
+negative \(\beta\) coefficient linking these two species implies that
+they co-occur less than expected under independence. \textbf{B.}
+Relative probabilities of all four possible presence-absence
+combinations for Species 1 and Species 2. The exponent includes
+\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
+it is absent (\(y_1 = 0\)). Similarly, the exponent includes
+\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
+\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
+constant \(Z\), ensures that the four relative probabilities sum to 1.
+In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
+the species no longer competed for the same resources), then the
+reduction in competition would allow each species to increase its
+occurrence rate and the co-occurrence deficit would be eliminated.}
+\end{figure}
+
 Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
@@
 the scope of this paper can be employed to keep the calculations
-feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
-noted below, some common linear and generalized linear methods can also
-be used as computationally efficient approximations to the full network
-(Lee and Hastie 2012, Loh and Wainwright 2013).
+feasible \citep{lee_learning_2012, salakhutdinov_learning_2008}.
+Alternatively, as noted below, some common linear and generalized linear
+methods can also be used as computationally efficient approximations to
+the full network \citep{lee_learning_2012, loh_structure_2013}.
 
@@
 \(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
-ones considered here have unimodal likelihood surfaces (Murphy 2012),
-ensuring that this procedure will always converge on the global maximum.
-This maximum represents the unique combination of \(\alpha\) and
-\(\beta\) coefficients that would be expected to produce exactly the
-observed co-occurrence frequencies on average (i.e.~maximizing the
-likelihood matches the sufficient statistics of the model distribution
-to the sufficient statistics of the data; Murphy 2012). I used the
-rosalia package (Harris 2015a) for the R programming language (R Core
-Team 2015) to optimize the Markov network parameters. The package was
-named after Santa Rosalia, the patron saint of biodiversity, whose
-supposedly miraculous healing powers played an important rhetorical role
-in the null model debates of the 1970's and 1980's (Lewin 1983).
+ones considered here have unimodal likelihood surfaces
+\citep{murphy_machine_2012}, ensuring that this procedure will always
+converge on the global maximum. This maximum represents the unique
+combination of \(\alpha\) and \(\beta\) coefficients that would be
+expected to produce exactly the observed co-occurrence frequencies on
+average \citep[i.e.~maximizing the likelihood matches the sufficient
+statistics of the model distribution to the sufficient statistics of the
+data;][]{murphy_machine_2012}. I used the rosalia package
+\citep{harris_rosalia_2015} for the R programming language
+\citep{r_core_team_r_2015} to optimize the Markov network parameters.
+The package was named after Santa Rosalia, the patron saint of
+biodiversity, whose supposedly miraculous healing powers played an
+important rhetorical role in the null model debates of the 1970's and
+1980's \citep{lewin_santa_1983}.
 
@@
 associations between pairs of species from simulated data (Appendix 3).
-First, I used the rosalia package (Harris 2015a) to fit Markov newtork
-models, as described above. For the analyses with 20 species, I added a
-very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
-terms with scale 2 to ensure that the model estimates were always
-finite. The bias introduced by this prior should be small: the 95\%
-credible interval on \(\beta\) only requires that one species' effect on
-the odds of observing a different species to be less than a factor of
-1500 (which is not much of a constraint). The logistic distribution was
-chosen because it is convex and has a similar shape to the Laplace
-distribution used in LASSO regularization (especially in the tails), but
-unlike the Laplace distribution it is differentiable everywhere and does
-not force any estimates to be exactly zero. To confirm that this
-procedure produced stable estimates, I compared its estimates on 50
-bootstrap replicates (Appendix 4).
+First, I used the rosalia package \citep{harris_rosalia_2015} to fit
+Markov newtork models, as described above. For the analyses with 20
+species, I added a very weak logistic prior distribution on the
+\(\alpha\) and \(\beta\) terms with scale 2 to ensure that the model
+estimates were always finite. The bias introduced by this prior should
+be small: the 95\% credible interval on \(\beta\) only requires that one
+species' effect on the odds of observing a different species to be less
+than a factor of 1500 (which is not much of a constraint). The logistic
+distribution was chosen because it is convex and has a similar shape to
+the Laplace distribution used in LASSO regularization (especially in the
+tails), but unlike the Laplace distribution it is differentiable
+everywhere and does not force any estimates to be exactly zero. To
+confirm that this procedure produced stable estimates, I compared its
+estimates on 50 bootstrap replicates (Appendix 4).
 
@@
 association. Next, I used partial correlations, which summarize species'
-conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
-2010). In the context of non-Gaussian data, the partial correlation can
-be thought of as a computationally efficient approximation to the full
-Markov network model (Loh and Wainwright 2013). This sort of model is
-very common for estimating relationships among genes and gene products
-(Friedman et al. 2008). Because partial correlations are undefined for
-landscapes with perfectly-correlated species pairs, I used a regularized
-estimate based on James-Stein shrinkage, as implemented in the corpcor
-package's \texttt{pcor.shrink} function with the default settings
-(Schäfer et al. 2014).
+conditional relationships
+\citep{albrecht_spatial_2001, faisal_inferring_2010}. In the context of
+non-Gaussian data, the partial correlation can be thought of as a
+computationally efficient approximation to the full Markov network model
+\citep{loh_structure_2013}. This sort of model is very common for
+estimating relationships among genes and gene products
+\citep{friedman_sparse_2008}. Because partial correlations are undefined
+for landscapes with perfectly-correlated species pairs, I used a
+regularized estimate based on James-Stein shrinkage, as implemented in
+the corpcor package's \texttt{pcor.shrink} function with the default
+settings \citep{schafer_corpcor_2014}.
 
@@
 thought of as a computationally efficient approximation to the Markov
-network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
-regularized logistic regression models (Gelman et al. 2008) for each
-species, using the other species on the landscape as predictors. To
-avoid the identifiability problems associated with directed cyclic
-graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
+network \citep{lee_learning_2012}. Following
+\citet{faisal_inferring_2010}, I fit regularized logistic regression
+models \citep{gelman_weakly_2008} for each species, using the other
+species on the landscape as predictors. To avoid the identifiability
+problems associated with directed cyclic graphs
+\citep{schmidt_modeling_2012}, I then symmetrized the relationships
 within species pairs via averaging.
 
-The next method, described in Gotelli and Ulrich (2009), involved
+The next method, described in \citet{gotelli_empirical_2009}, involved
 simulating new landscapes from a null model that retains the row and
-column sums of the original matrix (Strong et al. 1984). I used the
-\(Z\)-scores computed by the Pairs software described in Gotelli and
-Ulrich (2009) as my null model-based estimator of species interactions.
+column sums of the original matrix \citep{strong_ecological_1984}. I
+used the \(Z\)-scores computed by the Pairs software described in
+\citet{gotelli_empirical_2009} as my null model-based estimator of
+species interactions.
 
 The last two estimators used the latent correlation matrix estimated by
-the BayesComm package (Golding and Harris 2015) in order to evaluate the
-recent claim that the correlation coefficients estimated by ``joint
-species distribution models'' provide an accurate assessment of species'
-pairwise interactions (Pollock et al. 2014, see also Harris 2015b). In
-addition to using the posterior mean correlation (Pollock et al. 2014),
-I also used the posterior mean \emph{partial} correlation, which might
-be able to control for indirect effects.
+the BayesComm package \citep{golding_bayescomm_2015} in order to
+evaluate the recent claim that the correlation coefficients estimated by
+``joint species distribution models'' provide an accurate assessment of
+species' pairwise interactions
+\citetext{\citealp{pollock_understanding_2014}; \citealp[see
+also][]{harris_generating_2015}}. In addition to using the posterior
+mean correlation \citep{pollock_understanding_2014}, I also used the
+posterior mean \emph{partial} correlation, which might be able to
+control for indirect effects.
 
@@
 smaller landscapes, which are also the ones that are most representative
-of typical analyses in this field (Gotelli and Ulrich 2009).
+of typical analyses in this field \citep{gotelli_empirical_2009}.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/performance.pdf}
+\caption{\textbf{Figure 3.} \textbf{A.} Proportion of variance in
+interaction coefficients explained by each method versus number of
+sampled locations. \textbf{B.} The \(Z\)-scores produced by the null
+model (``Pairs'') for each pair of species can be predicted using the
+correlation between the presence-absence vectors of those same species
+and from the number of sites on the landscape.}
+\end{figure}
 
@@
 partial correlations' success on simulated data may not carry over to
-real data sets, however; Loh and Wainwright (2013) show that the linear
+real data sets, however; \citet{loh_structure_2013} show that the linear
 approximations can be less reliable in cases where the true interaction
@@
 linear models for each species can occasionally lead to catastrophic
-overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
-For these reasons, it will usually be best to fit a Markov network
-rather than one of the alternative methods when one's computational
-resources allow it.
+overfitting with small-to-moderate sample sizes
+\citep{lee_learning_2012}. For these reasons, it will usually be best to
+fit a Markov network rather than one of the alternative methods when
+one's computational resources allow it.
 
 It's important to note that none of these methods can identify the exact
-nature of the pairwise interactions (e.g.~which species in a
-positively-associated pair is facilitating the other; Schmidt and Murphy
-2012), particularly when real pairs of species can reciprocally
-influence one another in multiple ways simultaneously (Bruno et al.
-2003); with compositional data, there is only enough information to
-provide a single number describing each species pair. To estimate
-asymmetric interactions, such as commensalism or predation, ecologists
-would need other kinds of data, as from time series, behavioral
-observations, manipulative experiments, or natural history. These other
-sources of information could also be used to augment the likelihood
-function with an informative prior distribution, which could lead to
-better results on some real data sets than was shown in Figure 3A.
+nature of the pairwise interactions \citep[e.g.~which species in a
+positively-associated pair is facilitating the
+other;][]{schmidt_modeling_2012}, particularly when real pairs of
+species can reciprocally influence one another in multiple ways
+simultaneously \citep{bruno_inclusion_2003}; with compositional data,
+there is only enough information to provide a single number describing
+each species pair. To estimate asymmetric interactions, such as
+commensalism or predation, ecologists would need other kinds of data, as
+from time series, behavioral observations, manipulative experiments, or
+natural history. These other sources of information could also be used
+to augment the likelihood function with an informative prior
+distribution, which could lead to better results on some real data sets
+than was shown in Figure 3A.
 
@@
 overall prevalence, which was a major flashpoint for the null model
-debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
+debates in the 1980's
+\citep{roughgarden_competition_1983, strong_ecological_1984}. Equation 1
 can be used to calculate the expected prevalence of a species in the
-absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
-and Hastie 2012). Competition's effect on prevalence in a Markov network
-can then be calculated by subtracting this value from the observed
-prevalence (cf Figure 2D). This kind of insight would have been
-difficult to obtain without a generative model that makes predictions
-about the consequences of species interactions; null models (which
-presume \emph{a priori} that interactions do not exist) have no way to
-make such predictions.
+absence of biotic influences
+\citep[\(\frac{e^\alpha}{1 + e^{\alpha}}\);][]{lee_learning_2012}.
+Competition's effect on prevalence in a Markov network can then be
+calculated by subtracting this value from the observed prevalence (cf
+Figure 2D). This kind of insight would have been difficult to obtain
+without a generative model that makes predictions about the consequences
+of species interactions; null models (which presume \emph{a priori} that
+interactions do not exist) have no way to make such predictions.
 
 Markov networks---particularly the Ising model for binary
-networks---have been studied for nearly a century (Cipra 1987), and the
-models' properties, capabilities, and limits are well-understood in a
-huge range of applications. Using the same framework for species
-interactions would thus allow ecologists to tap into an enormous set of
-existing discoveries and techniques for dealing with indirect effects,
-stability, and alternative stable states. Numerous other extensions are
-possible: for example, the states of the interaction network can be
-modeled as a function of the local abiotic environment (Lee and Hastie
-2012), which would provide a rigorous and straightforward approach to
-the difficult and important task of incorporating whole networks of
-biotic interactions into species distribution models (Kissling et al.
-2012, Pollock et al. 2014), leading to a better understanding of the
-interplay between biotic and abiotic effects on community structure.
-There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
-Besag 1998) that would allow one species to affect the sign or strength
-of the relationship between two other species, tipping the balance
-between facilitation and exploitation (Bruno et al. 2003).
+networks---have been studied for nearly a century
+\citep{cipra_introduction_1987}, and the models' properties,
+capabilities, and limits are well-understood in a huge range of
+applications. Using the same framework for species interactions would
+thus allow ecologists to tap into an enormous set of existing
+discoveries and techniques for dealing with indirect effects, stability,
+and alternative stable states. Numerous other extensions are possible:
+for example, the states of the interaction network can be modeled as a
+function of the local abiotic environment \citep{lee_learning_2012},
+which would provide a rigorous and straightforward approach to the
+difficult and important task of incorporating whole networks of biotic
+interactions into species distribution models
+\citep{kissling_towards_2012, pollock_understanding_2014}, leading to a
+better understanding of the interplay between biotic and abiotic effects
+on community structure. There are even methods
+\citep{whittam_species_1981, tjelmeland_markov_1998} that would allow
+one species to affect the sign or strength of the relationship between
+two other species, tipping the balance between facilitation and
+exploitation \citep{bruno_inclusion_2003}.
 
@@
 thinking about the numerical consequences of species' richness and
-abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
+abundance patterns \citep{harris_occupancy_2011, xiao_strong_2015}, but
 deviations from a particular null model must be interpreted with care
-(Roughgarden 1983). Even in small networks with three species, it may
-simply not be possible to implicate individual species pairs or specific
-ecological processes like competition by rejecting a general-purpose
-null (Gotelli and Ulrich 2009), especially when the test statistic is
-effectively just a correlation coefficient (Figure 3B). Simultaneous
-estimation of multiple ecological parameters seems like a much more
-promising approach: to the extent that the models' relative performance
-on real data sets is similar to the range of results shown in Figure 3A,
-scientists in this field could often double their explanatory power by
-switching from null models to Markov networks (or increase it
-substantially with linear or generalized linear approximations).
-Regardless of the methods ecologists ultimately choose, controlling for
-indirect effects could clearly improve our understanding of species'
-direct effects on one another and on community structure.
+\citep{roughgarden_competition_1983}. Even in small networks with three
+species, it may simply not be possible to implicate individual species
+pairs or specific ecological processes like competition by rejecting a
+general-purpose null \citep{gotelli_empirical_2009}, especially when the
+test statistic is effectively just a correlation coefficient (Figure
+3B). Simultaneous estimation of multiple ecological parameters seems
+like a much more promising approach: to the extent that the models'
+relative performance on real data sets is similar to the range of
+results shown in Figure 3A, scientists in this field could often double
+their explanatory power by switching from null models to Markov networks
+(or increase it substantially with linear or generalized linear
+approximations). Regardless of the methods ecologists ultimately choose,
+controlling for indirect effects could clearly improve our understanding
+of species' direct effects on one another and on community structure.
 
@@
 \setlength{\leftskip}{1em}
-
-Albrecht, M., and N. J. Gotelli. 2001. Spatial and temporal niche
-partitioning in grassland ants. Oecologia 126:134--141.
-
-Azaele, S., R. Muneepeerakul, A. Rinaldo, and I. Rodriguez-Iturbe. 2010.
-Inferring plant ecosystem organization from species occurrences. Journal
-of theoretical biology 262:323--329.
-
-Bruno, J. F., J. J. Stachowicz, and M. D. Bertness. 2003. Inclusion of
-facilitation into ecological theory. Trends in Ecology \& Evolution
-18:119--125.
-
-Cipra, B. A. 1987. An introduction to the Ising model. American
-Mathematical Monthly 94:937--959.
-
-Connor, E. F., M. D. Collins, and D. Simberloff. 2013. The checkered
-history of checkerboard distributions. Ecology 94:2403--2414.
-
-Diamond, J. M. 1975. The island dilemma: Lessons of modern biogeographic
-studies for the design of natural reserves. Biological conservation
-7:129--146.
-
-Dodson, S. I. 1970. COMPLEMENTARY FEEDING NICHES SUSTAINED BY
-SIZE-SELECTIVE PREDATION. Limnology and Oceanography 15:131--137.
-
-Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
-Inferring species interaction networks from species abundance data: A
-comparative evaluation of various statistical and machine learning
-methods. Ecological Informatics 5:451--464.
-
-Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
-Selected Problems in Ecology. Entropy 15:5237--5276.
-
-Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
-covariance estimation with the graphical lasso. Biostatistics
-9:432--441.
-
-Gelfand, A. E., A. M. Schmidt, S. Wu, J. A. Silander, A. Latimer, and A.
-G. Rebelo. 2005. Modelling species diversity through species level
-hierarchical modelling. Journal of the Royal Statistical Society: Series
-C (Applied Statistics) 54:1--20.
-
-Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
-Informative Default Prior Distribution for Logistic and Other Regression
-Models. The Annals of Applied Statistics 2:1360--1383.
-
-Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
-non-randomness in species Co-occurrences on Islands. Oecologia
-52:75--84.
-
-Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian Community
-Ecology Analysis.
-
-Gotelli, N. J., and G. L. Entsminger. 2003. Swap algorithms in null
-model analysis. Ecology:532--535.
-
-Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
-tool to identify non-random species associations. Oecologia
-162:463--477.
-
-Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
-networks. R package version 0.1.0. Zenodo.
-http://dx.doi.org/10.5281/zenodo.17808.
-
-Harris, D. J. 2015b. Generating realistic assemblages with a Joint
-Species Distribution Model. Methods in Ecology and Evolution.
-
-Harris, D. J., K. G. Smith, and P. J. Hanly. 2011. Occupancy is
-nine-tenths of the law: Occupancy rates determine the homogenizing and
-differentiating effects of exotic species. The American naturalist
-177:535.
-
-Harris, T. E. 1974. Contact Interactions on a Lattice. The Annals of
-Probability 2:969--988.
-
-Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
-J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
-Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
-Towards novel approaches to modelling biotic interactions in
-multispecies assemblages at large spatial extents. Journal of
-Biogeography 39:2163--2178.
-
-Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
-Models.
-
-Levine, S. H. 1976. Competitive Interactions in Ecosystems. The American
-Naturalist 110:903--910.
-
-Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.
-
-Loh, P.-L., and M. J. Wainwright. 2013. Structure estimation for
-discrete graphical models: Generalized covariance matrices and their
-inverses. The Annals of Statistics 41:3022--3049.
-
-MacArthur, R. H. 1958. Population ecology of some warblers of
-northeastern coniferous forests. Ecology 39:599--619.
-
-Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
-MIT Press.
-
-Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K.
-M. Parris, P. A. Vesk, and M. A. McCarthy. 2014. Understanding
-co-occurrence by modelling species simultaneously with a Joint Species
-Distribution Model (JSDM). Methods in Ecology and Evolution:n/a--n/a.
-
-R Core Team. 2015. R: A Language and Environment for Statistical
-Computing. R Foundation for Statistical Computing, Vienna, Austria.
-
-Roughgarden, J. 1983. Competition and Theory in Community Ecology. The
-American Naturalist 122:583--601.
-
-Salakhutdinov, R. 2008. Learning and evaluating Boltzmann machines.
-Technical Report UTML TR 2008-002, Department of Computer Science,
-University of Toronto, Dept. of Computer Science, University of Toronto.
-
-Schäfer, J., R. Opgen-Rhein, V. Zuber, M. Ahdesmäki, A. P. D. Silva, and
-K. Strimmer. 2014. Corpcor: Efficient Estimation of Covariance and
-(Partial) Correlation.
-
-Schmidt, M., and K. Murphy. 2012. Modeling Discrete Interventional Data
-using Directed Cyclic Graphical Models. arXiv preprint arXiv:1205.2617.
-
-Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.
-Ecological communities: Conceptual issues and the evidence. Princeton
-University Press.
-
-Tjelmeland, H., and J. Besag. 1998. Markov Random Fields with
-Higher-order Interactions. Scandinavian Journal of Statistics
-25:415--433.
-
-Veech, J. A. 2013. A probabilistic model for analysing species
-co-occurrence. Global Ecology and Biogeography 22:252--260.
-
-Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
-Community Structure in Alaskan Seabird Colonies. Ecology 62:1515--1524.
-
-Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
-Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
-
-\textbf{Figure captions}
-
-\setlength{\parskip}{3pt} \setlength{\parindent}{0em}
-\setlength{\leftskip}{0em}
-
-\textbf{Figure 1.} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, but more
-weakly (circled coefficient). \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled).
-
-\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
-The abiotic environment favors the occurrence of both species
-(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
-negative \(\beta\) coefficient linking these two species implies that
-they co-occur less than expected under independence. \textbf{B.}
-Relative probabilities of all four possible presence-absence
-combinations for Species 1 and Species 2. The exponent includes
-\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
-it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
-the species no longer competed for the same resources), then the
-reduction in competition would allow each species to increase its
-occurrence rate and the co-occurrence deficit would be eliminated.
-
-\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
-coefficients explained by each method versus number of sampled
-locations. \textbf{B.} The \(Z\)-scores produced by the null model
-(``Pairs'') for each pair of species can be predicted using the
-correlation between the presence-absence vectors of those same species
-and from the number of sites on the landscape.
-
-\end{flushleft}
-\end{spacing}
-
-\end{document}

### 2015-10-15 Harris_2015/recovery.Rmd
clean up directory

@@
 
-```{r, cache = TRUE, message = FALSE}
+```{r, cache = TRUE, message = FALSE, eval = FALSE}
 library(dplyr)        # For manipulating data structures
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 pairs_txt = readLines("fakedata/Pairs.txt")
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 fit_all = function(filename){
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 
@@
 
-```{r, cache = TRUE, message=FALSE}
+```{r, cache = TRUE, message=FALSE, eval = FALSE}
 colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 library(ggplot2)
@@
 
-ggplot(z[z$n_sites >0, ], aes(x = `Pairs`, y = truth)) + 
+z$`-Pairs` = -z$Pairs
+ggplot(z[z$n_sites >0, ], aes(x = `-Pairs`, y = truth)) + 
   stat_binhex(bins = 100) + 
@@
   ggtitle("Null model (\"Pairs\")") + 
-  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$Pairs + 0)))
+  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$`-Pairs` + 0)))
 ```
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 rep = 12
@@
 
-```{r, cache = TRUE}
+```{r, cache = TRUE, eval = FALSE}
 for(compared in c("partial correlation", "correlation", "GLM", "BayesComm", "partial BayesComm", "Pairs")){
@@
 ```
+
+```{r, eval = FALSE}
+save(results, sizes, file = "graph_data.Rdata")
+```

### 2015-10-15 Harris_2015/recovery.tex
clean up directory

--- 2015-08-04 Harris_2015/recovery.tex
@@
 \begin{Highlighting}[]
-\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
+\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{-z$Pairs}
+\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{-Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
 \StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
@@
 \StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$Pairs +}\StringTok{ }\DecValTok{0}\NormalTok{)))}
+\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
 \end{Highlighting}
@@
 ## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
-## R-squared is 1.77-1.95 times higher than from correlation (95% bootstrap CI)
+## R-squared is 1.79-1.95 times higher than from correlation (95% bootstrap CI)
 ## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
-## R-squared is 2.54-2.9 times higher than from BayesComm (95% bootstrap CI)
-## R-squared is 1.77-1.97 times higher than from partial BayesComm (95% bootstrap CI)
-## R-squared is 2.12-2.38 times higher than from Pairs (95% bootstrap CI)
+## R-squared is 2.72-3.28 times higher than from BayesComm (95% bootstrap CI)
+## R-squared is 1.29-1.37 times higher than from partial BayesComm (95% bootstrap CI)
+## R-squared is 2.12-2.37 times higher than from Pairs (95% bootstrap CI)
 \end{verbatim}
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{save}\NormalTok{(results, sizes, }\DataTypeTok{file =} \StringTok{"graph_data.Rdata"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
 \end{document}

### 2015-10-15 Harris_2015/simulation.Rmd
clean up directory

@@
 output: pdf_document
-title: Appendix [[X]]: simulating species interactions
+title: Appendix 2: simulating species interactions
 subtitle: \"Estimating species interactions from observational data with Markov networks\"

### 2015-10-15 Harris_2015/three_species.Rmd
submitted to Ecology
clean up directory

@@
-# Evaluation of the three species in Figure 1
+---
+output: pdf_document
+title: "Appendix 1: Evaluation of the three species in Figure 1"
+---
 
@@
 
-    rosie = rosalia(x, trace = 0, prior = make_flat_prior(), hessian = TRUE)
-
-    se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
-    rosalia_confident_positive = pnorm(0, mean = rosie$beta[2, 3], sd = se) < .025
-    rosalia_confident_negative = pnorm(0, mean = rosie$beta[2, 3], sd = se) > .975
+    rosie = rosalia(x, trace = 0, prior = make_logistic_prior(scale = 2), hessian = TRUE)
 
@@
       partial = cor2pcor(cor(x))[2, 3] < 0,
-      rosalia = rosie$beta[2,3] < 0,
-      rosalia_confident_positive = rosalia_confident_positive,
-      rosalia_confident_negative = rosalia_confident_negative
+      rosalia = rosie$beta[2,3],
+      rosalia_se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
     )
@@
 
+```{r}
+upper_ci_below_zero = evaluation["rosalia", ] + 
+  qnorm(.975) * evaluation["rosalia_se.beta3", ] < 0
+lower_ci_above_zero = evaluation["rosalia", ] + 
+  qnorm(.025) * evaluation["rosalia_se.beta3", ] > 0
+```
+
+Evaluate the frequentist coverage of the confidence intervals (CIs).
+If the coverage is good, then approximately 95% of the confidence intervals generated on independent data sets should contain the true value (i.e., in 95% of cases, the upper CI should be above the true value of $-1$ and the lower CI should be below the true value).
+
+```{r}
+upper_ci_above_true = evaluation["rosalia", ] + 
+  qnorm(.975) * evaluation["rosalia_se.beta3", ] > -1
+lower_ci_below_true = evaluation["rosalia", ] + 
+  qnorm(.025) * evaluation["rosalia_se.beta3", ] < -1
+
+# Coverage is close to 95%
+100 * mean(upper_ci_above_true & lower_ci_below_true)
+
+# According to this binomial test, the confidence intervals include the
+# true value a bit too often to be consistent with 95% coverage (i.e. there
+# is some evidence that the confidence intervals are too conservative).
+binom.test(x = sum(upper_ci_above_true & lower_ci_below_true), 
+           n = length(upper_ci_above_true), 
+           p = 0.95,
+           conf.level = 0.95
+)
+
+```
+
 After running the "Pairs" software on transposed versions of the files simulated above, we can import its results to see how the method performed.
 
-```{r}
+```{r, eval = FALSE}
 Pairs.txt = readLines("fakedata/three/Pairs.txt")
@@
 
-```{r}
+```{r, eval = FALSE}
 out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation)[1:3] * 100)
@@
 
-```{r}
+```{r, eval = FALSE}
 # percentage of Pairs runs that found statistically significant mutualism at the

### 2015-10-15 stochastic/chapter-3.Rmd
clean up directory

@@
 Species interactions are believed to play an important role in community structure, but detecting their influence in the co-occurrence patterns of observed communities has stymied ecologists for decades.
-In recent years, several papers have suggested that ecologists could solve this long-standing problem by fitting Markov networks (undirected graphical models also known as Markov random fields) to the data by maximum likelihood or similar methods.
-This approach allows ecologists to control for extraneous factors when estimating the direct interactions between a given pair of species, and previous work has shown that this kind of control is required for reliable inferences about interactions' sign and strength.
-Unfortunately, the likelihood function for Markov networks quickly becomes intractable once more than about 20 species are included in the model, due to the exponentially large number of possible combinations of presences and absences.
-Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions.
-In this paper, I introduce "stochastic approximation" as a method for fitting these models to large ecological networks where hundreds of species and multiple abiotic factors can all affect a given species’ occurrence probability.
-This approach, which converges to the maximum likelihood estimate with probability one, greatly expands the range of data sets we can analyze and the number of ecological questions we can address with these methods.
+While Markov networks (undirected graphical models also known as Markov random fields) represent a promising approach to this long-standing problem by isolating direct interactions from indirect ones, the methods ecologists have suggested for fitting these models are limited to small communities with about 20 species or fewer.
+Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and thus attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions.
+In this paper, I introduce stochastic approximation as an alternative method for fitting these models that addresses both of these problems, making it feasible to use Markov networksin cases where each species responds to multiple abiotic factors and hundreds of competitors or mutualists.
+While stochastic approximation does introduce some sampling noise during the optimization process, it still converges to the maximum likelihood estimate for species' interaction coefficients with probability one.
 
@@
 To the extent that species interactions are important for community assembly, ecologists generally expect them to leave a signature on species' co-occurrence patterns.
-But using that signature to infer the underlying species interactions from observational data has been much harder.
+Using that signature to infer the underlying species interactions from observational data has been much harder, however.
 Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @gilpin_factors_1982; @strong_ecological_1984; @hastings_can_1987].
@@
 @friedman_sparse_2008).
-The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al.
+The discrepancy is due to the fact that ecologists have chosen to focus, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al.
 2014; see Faisal et al.
@@
 When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
-While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.
+While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still lack a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.
 
@@
 **B.** Although the two trees directly reduce one another’s occurrence probabilities via competition, they tend to occur in the same areas (with sufficient water and without too many herbivores).
-An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate.
-**C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates.
+An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate (circled "+" sign).
+**C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates (note that the circled interaction between the trees has the correct sign).
 Note that, with observational data, the bi-directional effects of species interactions (pairs of arrows between species) need to be collapsed to a single number describing the conditional relationship between species (one double-headed arrow).](figure-1.pdf)
@@
 Rather than describing the overall relationship between two variables across all conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.
-@harris_estimating_2015 found that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
+@harris_estimating_2015 found that the partial correlation can do extract accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
 
@@
 the occurrence and co-occurrence rates), and that it does so with the fewest possible parameters or constraints [i.e.
-the Markov network has the most information entropy of any possible model satisfying its constraints; azaele_inferring_2010].
+the Markov network has the most information entropy of any possible model satisfying its constraints; @azaele_inferring_2010].
 As shown in @harris_estimating_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.
@@
 For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [@azaele_inferring_2010; @lee_learning_2012; @murphy_machine_2012].
-One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the expected sufficient statistics under the model [@bickel_mathematical_1977].
-Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by climbing the likelihood surface.
-At this point, any discrepancies between the sufficient statistics of the model and of the data will have been eliminated.
+Finding the maximum likelihood estimate for the model parameters is equivalent to minimizing the discrepancy between the values of the data's sufficient statistics and the expected sufficient statistics under the model  [@bickel_mathematical_1977].
+Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by iteratively reducing the discrepancies between the sufficient statistics of the model and of the data to zero.
 
 In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates each value exactly, averaging over all possible presence-absence combinations.
-This approach is infeasible for larger networks, as the number of possibilities grows exponentially with the size of the network.
-Stochastic approximation [@robbins_stochastic_1951; @salakhutdinov_efficient_2012] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.
-The procedure iterates through the following three steps as many times as needed (50,000 for these analyses):
+Stochastic approximation [@robbins_stochastic_1951; @salakhutdinov_efficient_2012] instead estimates the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.
+The procedure iterates through the following three steps as many times as needed (50,000 for these analyses; see Appendix for annotated R code):
 
-1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
+1. Simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.
 
-2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
+2. Subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient
 
-3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.
+3. Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.
 
-The specific procedure used here had the following implementation details:
-
-The simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates.
+Here, the simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates.
 While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this merely slows convergence to the maximum likelihood estimate rather than preventing it altogether [@younes_convergence_1999; @salakhutdinov_efficient_2012].
@@
 These gradients were augmented with a momentum term [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 2.0 (for environmental responses) or 0.5 (for pairwise interactions).
-
-The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 1000/(999 + t)$, which satisfies the criteria in @powell_approximate_2007.
-The scaling factor, $\alpha_0$, was 1.0 for species’ intercepts and environmental responses.
-Because of the enormous number of pairwise interaction coefficients, I set $\alpha_0$ to 0.1 for the $\beta$ parameters.
+The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = 1000/(999 + t)$, which satisfies the criteria in @powell_approximate_2007.
 
@@
 
-Each species was assigned two two sets of coefficients.
+Each species was assigned two sets of coefficients.
 The coefficients determining species’ responses to the environment were each drawn from standard normal distributions.
@@
 ![Stochastic approximation improves the model over time.
-**A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation reaches the global optimum very quickly.
-Within a few minutes, the squared deviation between the approximate and exact estimates disappears.
+**A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation approaches the global optimum very quickly (note the log scale of the y axis).
+Within a few seconds, the mean squared deviation between the approximate and exact estimates drops to less than 0.01; in less than a minute, it drops below 0.0001.
+For comparison, the `rosalia` package took about six minutes to find the maximum likelihood estimate.
 **B.** With larger networks, where the maximum likelihood estimate cannot be calculated exactly, stochastic approximation converges to an apparent optimum that explains most (but not all) of the variation in the "true" parameter values.](convergence.pdf)

### 2015-10-15 stochastic/chapter-3.tex
clean up directory

\section{Abstract}\label{abstract}

Species interactions are believed to play an important role in community
structure, but detecting their influence in the co-occurrence patterns
of observed communities has stymied ecologists for decades. While Markov
networks (undirected graphical models also known as Markov random
fields) represent a promising approach to this long-standing problem by
isolating direct interactions from indirect ones, the methods ecologists
have suggested for fitting these models are limited to small communities
with about 20 species or fewer. Additionally, the methods that have been
proposed so far do not account for environmental heterogeneity, and thus
attempt to explain all of the co-occurrence patterns in ecological data
sets with species interactions. In this paper, I introduce stochastic
approximation as an alternative method for fitting these models that
addresses both of these problems, making it feasible to use Markov
networksin cases where each species responds to multiple abiotic factors
and hundreds of competitors or mutualists. While stochastic
approximation does introduce some sampling noise during the optimization
process, it still converges to the maximum likelihood estimate for
species' interaction coefficients with probability one.

\section{Introduction}\label{introduction}

To the extent that species interactions are important for community
assembly, ecologists generally expect them to leave a signature on
species' co-occurrence patterns. Using that signature to infer the
underlying species interactions from observational data has been much
harder, however. Disagreements about how to draw these inferences led to
an acrimonious, decade-long argument among community ecologists (Lewin
1983), where essentially all of the proposed statistical approaches were
criticized for having high error rates, a poor match with the underlying
ecological questions, computational infeasibility, or all three (Connor
and Simberloff 1979; Gilpin and Diamond 1982; Strong et al. 1984;
Hastings 1987). As documented below, ecologists have, for the most part,
not solved these problems in the ensuing decades. Resolving these issues
would solve a longstanding issue in community ecology and make a
significant contribution to other fields, such as species distribution
modeling (Kissling et al. 2012).

During the time that ecologists have struggled to identify the
interactions among dozens of species, bioinformaticians have largely
solved the analogous problem of estimating interactions among thousands
of genes from their co-expression levels (e.g. Friedman, Hastie, and
Tibshirani (2008)). The discrepancy is due to the fact that ecologists
have chosen to focus, almost exclusively, on overall co-occurrence rates
(Connor and Simberloff 1979; Gotelli and Ulrich 2009; Gilpin and Diamond
1982; Veech 2013) or on closely-related values such as correlation
coefficients (Pollock et al. 2014; see Faisal et al. 2010 and Harris
2015 for two recent exceptions). As ecologists know, however, the
correlation between two species will often reflect other factors beyond
their direct pairwise interactions (e.g.~abiotic influences and indirect
biotic effects; Figure 1). When these factors are not accounted for,
this reliance on overall co-occurrence rates can reliably lead to
incorrect inferences (D. J. Harris 2015). While some methods have been
proposed for incorporating a small number of specific factors (such as
geographic or environmental dissimilarity) into ecologists' null models
at a time (Lessard et al. 2011), we still lack a good way to scale these
approaches up for cases with many biotic and abiotic factors acting
simultaneously.

\begin{figure}[htbp]
\centering
\includegraphics{figure-1.pdf}
\caption{A hypothetical network of three species, two of which depend on
an abiotic factor (annual rainfall). \textbf{A.} The ``true'' network
involves the caterpillar exploiting the two tree species as well as
competition between the two trees. The two trees also benefit from
rainfall. \textbf{B.} Although the two trees directly reduce one
another's occurrence probabilities via competition, they tend to occur
in the same areas (with sufficient water and without too many
herbivores). An analysis that ignores these other factors could
mistakenly conclude that the two trees were mutualists because of their
high co-occurrence rate (circled ``+'' sign). \textbf{C.} The methods
presented here are designed to estimate the direct interactions between
each pair of species, after accounting for other species and abiotic
factors that could affect their co-occurrence rates (note that the
circled interaction between the trees has the correct sign). Note that,
with observational data, the bi-directional effects of species
interactions (pairs of arrows between species) need to be collapsed to a
single number describing the conditional relationship between species
(one double-headed arrow).}
\end{figure}

A number of procedures have been developed for controlling for the
influence of extraneous factors and focusing on the direct relationship
between a single pair of variables, i.e.~for estimating the conditional
relationship between them. The most familiar of these is the partial
correlation. Rather than describing the overall relationship between two
variables across all conditions, partial correlations (like regression
coefficients) describe the portion of the relationship that remains
after the other variables in the data have been accounted for. D. J.
Harris (2015) found that the partial correlation can do extract accurate
information about pairwise species interactions from binary
co-occurrence data in some circumstances, but the fact that this
approach assumes multivariate Gaussian data makes it less appealing.

Recently, several papers have suggested that ecologists could use Markov
networks (undirected graphical models also known as Markov random
fields) to estimate conditional relationships from binary data by
maximum entropy or maximum likelihood (Azaele et al. (2010); D. J.
Harris (2015)). A Markov network defines a probability distribution over
possible binary species assemblages, and its coefficients (including one
coefficient describing the conditional relationship between each pair of
species in the network). This approach is optimal in the sense that it
produces a model that matches the observed properties of the data
(i.e.~the occurrence and co-occurrence rates), and that it does so with
the fewest possible parameters or constraints (i.e.~the Markov network
has the most information entropy of any possible model satisfying its
constraints; Azaele et al. 2010). As shown in D. J. Harris (2015),
Markov networks do a better job of estimating species' influences on one
another better than a number of existing methods, particularly when
sample sizes are low.

Unfortunately, Markov networks have an intractable likelihood function
whose computational difficulty more than doubles each time a new species
is added to the model. This exponential growth in the likelihood
function's complexity means that the methods that worked for Azaele et
al. (2010) and D. J. Harris (2015) with networks of 20 species would be
completely infeasible for networks with 50 species, requiring over a
billion times more computational effort
(\(2^{50}/2^{20} \approx 10^{9}\)). Extending the method to account for
abiotic variation among sites on the landscape would require repeating
these expensive computations independently for every site, increasing
the computational burden even further. If ecologists want to apply these
methods to larger problems, a different model-fitting algorithm would be
needed.

In this paper, I present a different way of optimizing the likelihood,
called ``stochastic approximation'' (Robbins and Monro 1951;
Salakhutdinov and Hinton 2012), which replaces the intractable
computations with tractable Monte Carlo estimates of the same
quantities. Despite the sampling error introduced by this substitution,
stochastic approximation provides strong guarantees for eventual
convergence to the maximum likelihood estimate (Younes 1999;
Salakhutdinov and Hinton 2012). This change in approach makes it
feasible to estimate the interactions among hundreds of species from
observational data while also accounting for possible responses to
abiotic factors (such as those used in species distribution models),
giving ecologists some power to distinguish between species pairs that
co-occur due to shared environmental tolerances from those that co-occur
due to direct interactions like mutualism.

\section{Methods}\label{methods}

\subsection{Markov networks}\label{markov-networks}

As discussed in Azaele et al. (2010) and D. J. Harris (2015), Markov
networks such as the Ising model (Cipra 1987) can be used to describe
community structure as follows. Each species is represented by a binary
random variable, describing whether it is present (1) or absent (0). A
species' conditional probability of presence under a given set of biotic
and abiotic conditions depends on its coefficients. The coefficients
linking species to one another describe their conditional associations:
all else equal, a species will occur less often in the presence of its
competitors and other species with which it is negatively associated,
and more often in the presence of its mutualists:

\[ p(y_i | \vec{y}_{j \neq i}) = \mathrm{logistic}\Big(\alpha_i + \sum_{j \neq i}{\beta_{ij}y_j}\Big), \]

where \(y_i\) is 1 when species \(i\) is present and 0 when it is
absent, \(\alpha_i\) is an intercept term, \(\beta_ij\) represents the
conditional relationship between species \(i\) and species \(j\), and
\(\mathrm{logistic}(x) = e^x / (1 + e^x)\). These conditional
probabilities can be combined to calculate the probability of observing
any given combination of presences and absences, up to a multiplicative
constant (Lee and Hastie 2012; Murphy 2012):

\[p(\vec{y}) \propto \mathrm{exp}\Big({\sum_{i}{\alpha_i y_i} + \sum_{j \neq i}}{\beta_{ij} y_i y_j}\Big).\]

Unfortunately, the exponentially large number of possible assemblages
makes the normalizing constant for this joint probability distribution
intractable when the number of species is larger than 20 or so. However,
it is generally straightforward to simulate examples of assemblages that
are consistent with graphical models such as these, using Monte Carlo
techniques like Gibbs sampling (Salakhutdinov and Hinton 2012). Gibbs
sampling is especially convenient for these models because it only
requires the conditional probabilities, which are straightforward to
compute (Harris 2015; Appendix).

In each of the scenarios below, the ``observed'' landscapes were
simulated using Gibbs sampling from a pre-specified set of ``true''
parameters, then stochastic approximation was used to recover these
parameters from the simulated presence-absence data.

\subsection{Conditioning the network on the abiotic
environment}\label{conditioning-the-network-on-the-abiotic-environment}

The Markov networks fit in D. J. Harris (2015) treated the \(\alpha_i\)
terms from Equations 1 and 2 as constants, meaning that the models
expected each species to have the same conditional occurrence
probability in any location with a given set of facilitators and
competitors. In this paper, I extend the model so that the local abiotic
conditions can also affect species occurrence probabilities directly.
Specifically, the \(\alpha_i\) terms in these analyses are linear
combinations of the local environmental conditions, represented by
\(x_1\) through \(x_5\). By estimating the coefficients associated with
each of these \(x\) variables for each species, we can account for their
responses to the abiotic environment as well as to one another. In other
fields, a model like this one, where a Markov network's parameters
depend on external factors is called a conditional random field (Lee and
Hastie 2012); these models have been used in a variety of contexts, from
language models to image processing (Murphy 2012).

\subsection{Coefficient estimation with stochastic
approximation}\label{coefficient-estimation-with-stochastic-approximation}

A Markov network describes a probability distribution over possible
assemblages. Since the model is a member of the exponential family it
can be summarized without loss of information by its ``sufficient
statistics''. For the model presented here, the sufficient statistics
include: the number of occurrences for each species, the number of
co-occurrences between each species, and the cross-product between the
species and environment matrices (Azaele et al. 2010; Lee and Hastie
2012; Murphy 2012). Finding the maximum likelihood estimate for the
model parameters is equivalent to minimizing the discrepancy between the
values of the data's sufficient statistics and the expected sufficient
statistics under the model (Bickel and Doksum 1977). Because
fully-observed Markov networks, such as the ones analyzed here, have
unimodal likelihood functions (Murphy 2012), it is possible to find the
global optimum by iteratively reducing the discrepancies between the
sufficient statistics of the model and of the data to zero.

In order to reduce the discrepancy between the observed and predicted
sufficient statistics, the \texttt{rosalia} calculates each value
exactly, averaging over all possible presence-absence combinations.
Stochastic approximation (Robbins and Monro 1951; Salakhutdinov and
Hinton 2012) instead estimates the expected values of the sufficient
statistics by averaging over a more manageable number of simulated
assemblages during each model-fitting iteration, while still retaining
maximum likelihood convergence guarantees. The procedure iterates
through the following three steps as many times as needed (50,000 for
these analyses; see Appendix for annotated R code):

1: simulate a set of assemblages from the current model parameters and
calculate sufficient statistics for the sample.

2: subtract the simulated sufficient statistics from the observed ones
to calculate the approximate likelihood gradient

3: Adjust the model parameters to climb the approximate gradient, using
a schedule of step sizes that satisfies the criteria in Chapter 6 of
Powell (2007).

Here, the simulations in Step 1 used Gibbs sampling to generate examples
of landscapes based on the model's current parameter estimates. While
the simulated landscapes produced by Gibbs sampling are serially
autocorrelated, statisticians have shown that this merely slows
convergence to the maximum likelihood estimate rather than preventing it
altogether (Younes 1999; Salakhutdinov and Hinton 2012).

The approximate likelihood gradients in Step 2 match the ones from D. J.
Harris (2015), except that they are averaged over a set of Monte Carlo
samples rather than over all possible presence-absence combinations.
These gradients were augmented with a momentum term (Hinton 2012) and by
regularizers based on a logistic prior with location 0 and scale 2.0
(for environmental responses) or 0.5 (for pairwise interactions).

The step size parameter in Step 3, \(\alpha_t\), decreased after each
iteration according to a generalized harmonic sequence,
\(\alpha_t = \alpha_0 1000/(999 + t)\), which satisfies the criteria in
Powell (2007). The scaling factor, \(\alpha_0\), was 1.0 for species'
intercepts and environmental responses. Because of the enormous number
of pairwise interaction coefficients, I set \(\alpha_0\) to 0.1 for the
\(\beta\) parameters.

\subsection{Simulated landscapes with known
interactions}\label{simulated-landscapes-with-known-interactions}

In order to assess the models' ability to recover the ``true''
parameters that generated a presence-absence matrix of interest, I first
needed to generate such matrices from known processes.

To demonstrate that my stochastic approximation implementation could
converge to the global optimum, I first simulated a landscape with 20
species and 500 sites, using Gibbs sampling. These landscapes had few
enough species that the \texttt{rosalia} package for estimating Markov
networks (David J. Harris 2015) could find the penalized maximum
likelihood estimates for the Markov network in a reasonable amount of
time. For each of these simulated landscapes, I then fit the same Markov
network model to the simulated data using the exact approach from D. J.
Harris (2015) and the stochastic approximation method described above.

I then simulated a large landscape with 250 species and 2500 sites,
representing a large data set roughly the size of the North American
Breeding Bird Survey. Each site on the landscape was represented by 5
environmental variables, which were drawn independently from Gaussian
distributions with mean 0 and standard deviation 2.5.

Each species was assigned two sets of coefficients. The coefficients
determining species' responses to the environment were each drawn from
standard normal distributions. The coefficients describing species'
pairwise interactions were drawn from a mixture of normals (see
Appendix) so that most interactions were weak and negative, but a few
strong positive and interactions also occurred.

The coefficients described above are sufficient to define each species'
conditional occurrence probability (i.e.~its probability of occurrence
against any given backdrop of biotic and abiotic features). I used these
conditional probabilities to produce examples of communities that were
consistent with the competition parameters and with the local
environmental variables via Markov chain Monte Carlo (specifically, 1000
rounds of Gibbs sampling; see code in the Appendix). I then used the
methods from the next section to attempt to infer the underlying
parameters from the simulated data.

\section{Results}\label{results}

\begin{figure}[htbp]
\centering
\includegraphics{convergence.pdf}
\caption{Stochastic approximation improves the model over time.
\textbf{A.} With small networks where the exact maximum likelihood
estimates are known, stochastic approximation approaches the global
optimum very quickly (note the log scale of the y axis). Within a few
seconds, the mean squared deviation between the approximate and exact
estimates drops to less than 0.01; in less than a minute, it drops below
0.0001. For comparison, the \texttt{rosalia} package took about six
minutes to find the maximum likelihood estimate. \textbf{B.} With larger
networks, where the maximum likelihood estimate cannot be calculated
exactly, stochastic approximation converges to an apparent optimum that
explains most (but not all) of the variation in the ``true'' parameter
values.}
\end{figure}

For the smaller communities, the squared deviations between the exact
estimates produced by the \texttt{rosalia} package and the ones produced
by stochastic approximation quickly decayed to negligible levels (Figure
2A), indicating that the stochastic approximation procedure was
implemented correctly and worked as the mathematical theory predicts.

\begin{figure}[htbp]
\centering
\includegraphics{estimates.pdf}
\caption{``True'' versus estimated network parameters for all 250
species' responses to the other 249 species in the data set (\textbf{A.}
\(R^2 = .71\)) and to the 5 abiotic variables (\textbf{B.}
\(R^2 = .95\)).}
\end{figure}

For the larger landscape, I found that the stochastic approximation
approach achieved reasonably good performance after ten minutes of
optimization (Figure 2B). After 50,000 iterations (about 5.4 hours on my
laptop), the model was able to recover more than two thirds of the
variance in species' pairwise interactions(\(R^2 = 0.71\); Figure 3A),
and nearly all of the variance in their responses to environmental
variables (\(R^2 = 0.95\); Figure 3B).

\section{Discussion}\label{discussion}

For decades, ecologists have relied on poor test statistics for
inferring species interactions from observational data (D. J. Harris
2015). As shown here and in D. J. Harris (2015), however, these
inferences can only be made reliably by methods that control for other
species in the network, such as partial correlations or Markov networks.
The biggest computational problem with large Markov networks is the
impossibility of averaging over all their possible states, but these
results demonstrate that this step can be avoided in ecological
analyses. Stochastic approximation is able to recover the same estimates
as exact methods for smaller networks, while scaling gracefully to
networks with hundreds of species.

The largest downside of using stochastic approximation instead of the
exact methods introduced in D. J. Harris (2015) is the difficulty in
generating confidence intervals. The \texttt{rosalia} package can
produce confidence intervals based on the Hessian matrix, but it is not
feasible to calculate this matrix for large networks. It may turn out
that the best way to generate confidence intervals for large networks is
by repeatedly fitting the model to bootstrapped samples of the data.
Alternatively, ecologists may be able to take advantage of the advanced
Monte Carlo techniques introduced in Murray, Ghahramani, and MacKay
(2012) to sample from the posterior distribution over possible parameter
values, which could be used to generate credible intervals for the
parameter estimates. Other advanced Monte Carlo methods (e.g.
Salakhutdinov 2009) may also speed up convergence to the maximum
likelihood estimate in cases where serial autocorrelation in the Gibbs
sampler is too strong for effectively sampling the space of possible
landscapes.

As the number of potentially-interacting species increases, the number
of adjustable parameters increases even faster, so overfitting becomes a
major concern. Fortunately, a number of good regularizers have been
proposed. Of these, some of the most interesting options include \(L_1\)
regularization, which ensures that many of the estimated coefficients
are exactly zero (Lee and Hastie 2012). This sparsity matches
ecologists' intuition that many species from different guilds are
unlikely to interact much at all (Faisal et al. 2010), and also produces
computational benefits for model estimation. Of course, the best
regularizers will take advantage of specific ecological knowledge
(e.g.~from field experiments, natural history, or trait data) to provide
information about individual pairwise interactions, rather than about
their overall distribution.

These richer sources of information will be especially important for
cases where ecologists expect that the interactions between species are
asymmetric (e.g.~where only one species is affected by an interaction or
where one species benefits at a cost to the other). With snapshot
observations of species' spatial associations, as discussed here,
interactions must be reduced to a single number describing the net
association, as in undirected models such as the Markov networks
presented here (Schmidt and Murphy 2012). With richer data types,
however, ecologists could more easily learn about asymmetric
interactions such as facilitation, predation, and parasitism.

Even without better data, ecologists have a number of options for
expanding beyond simple Markov networks in a number of ways that would
improve their ability to address a wider range of questions. This paper
demonstrated that it is possible to simultaneously estimate species'
responses to the abiotic and environment and to one another, but many
other extensions are possible. For example, ecologists could condition
the model on variables whose values have not been measured
(e.g.~partially-observed Markov networks, or the approximate networks in
the \texttt{mistnet} package; cf. Pollock et al. (2014)). This would
allow ecologists to account for measurement error and for other
ecologically-important factors that can be difficult to measure.
Ecologists should also explore higher-order networks, where one species'
presence can affect the relationship between two other species (Whittam
and Siegel-Causey 1981; Tjelmeland and Besag 1998).

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

Azaele, S, R Muneepeerakul, A Rinaldo, and I Rodriguez-Iturbe. 2010.
``Inferring Plant Ecosystem Organization from Species Occurrences.''
\emph{Journal of Theoretical Biology} 262 (2): 323--29.

Bickel, PJ, and K Doksum. 1977. \emph{Mathematical Statistics: Basic
Ideas and Selected Topics}. San Francisco: Holden---Day.

Cipra, Barry A. 1987. ``An Introduction to the Ising Model.''
\emph{American Mathematical Monthly} 94 (10): 937--59.

Connor, Edward F, and Daniel Simberloff. 1979. ``The Assembly of Species
Communities: Chance or Competition?'' \emph{Ecology}, 1132--40.

Faisal, Ali, Frank Dondelinger, Dirk Husmeier, and Colin M. Beale. 2010.
``Inferring Species Interaction Networks from Species Abundance Data: A
Comparative Evaluation of Various Statistical and Machine Learning
Methods.'' \emph{Ecological Informatics} 5 (6): 451--64.
doi:\href{http://dx.doi.org/http://dx.doi.org/10.1016/j.ecoinf.2010.06.005}{http://dx.doi.org/10.1016/j.ecoinf.2010.06.005}.

Friedman, Jerome, Trevor Hastie, and Robert Tibshirani. 2008. ``Sparse
Inverse Covariance Estimation with the Graphical Lasso.''
\emph{Biostatistics} 9 (3): 432--41.
doi:\href{http://dx.doi.org/10.1093/biostatistics/kxm045}{10.1093/biostatistics/kxm045}.

Gilpin, Michael E., and Jared M. Diamond. 1982. ``Factors Contributing
to Non-Randomness in Species Co-Occurrences on Islands.''
\emph{Oecologia} 52 (1): 75--84.
doi:\href{http://dx.doi.org/10.1007/BF00349014}{10.1007/BF00349014}.

Gotelli, Nicholas J., and Werner Ulrich. 2009. ``The Empirical Bayes
Approach as a Tool to Identify Non-Random Species Associations.''
\emph{Oecologia} 162 (2): 463--77.
doi:\href{http://dx.doi.org/10.1007/s00442-009-1474-y}{10.1007/s00442-009-1474-y}.

Harris, D. J. 2015. ``Estimating Species Interactions from Observational
Data with Markov Networks.'' \emph{BioRxiv}.
doi:\href{http://dx.doi.org/10.1101/018861}{10.1101/018861}.

Harris, David J. 2015. \emph{Rosalia: Exact Inference for Small Binary
Markov Networks. R Package Version 0.1.0}. Zenodo.
http://dx.doi.org/10.5281/zenodo.17808.
\url{http://dx.doi.org/10.5281/zenodo.17808}.

Hastings, Alan. 1987. ``Can Competition Be Detected Using Species
Co-Occurrence Data?'' \emph{Ecology}, 117--23.

Hinton, Geoffrey E. 2012. ``A Practical Guide to Training Restricted
Boltzmann Machines.'' In \emph{Neural Networks: Tricks of the Trade},
599--619. Springer.

Kissling, W. D., Carsten F. Dormann, Jürgen Groeneveld, Thomas Hickler,
Ingolf Kühn, Greg J. McInerny, José M. Montoya, et al. 2012. ``Towards
Novel Approaches to Modelling Biotic Interactions in Multispecies
Assemblages at Large Spatial Extents.'' \emph{Journal of Biogeography}
39 (12): 2163--78.
doi:\href{http://dx.doi.org/10.1111/j.1365-2699.2011.02663.x}{10.1111/j.1365-2699.2011.02663.x}.

Lee, Jason D., and Trevor J. Hastie. 2012. ``Learning Mixed Graphical
Models.'' \emph{CERN Document Server}.
\url{http://cds.cern.ch/record/1451206}.

Lessard, Jean-Philippe, Michael K. Borregaard, James A. Fordyce, Carsten
Rahbek, Michael D. Weiser, Robert R. Dunn, and Nathan J. Sanders. 2011.
``Strong Influence of Regional Species Pools on Continent-Wide
Structuring of Local Communities.'' \emph{Proceedings of the Royal
Society of London B: Biological Sciences}.
doi:\href{http://dx.doi.org/10.1098/rspb.2011.0552}{10.1098/rspb.2011.0552}.

Lewin, Roger. 1983. ``Santa Rosalia Was a Goat.'' \emph{Science} 221
(4611): 636--39.

Murphy, Kevin P. 2012. \emph{Machine Learning: A Probabilistic
Perspective}. The MIT Press.

Murray, Iain, Zoubin Ghahramani, and David MacKay. 2012. ``MCMC for
Doubly-Intractable Distributions.'' \emph{ArXiv Preprint
ArXiv:1206.6848}.

Pollock, Laura J., Reid Tingley, William K. Morris, Nick Golding, Robert
B. O'Hara, Kirsten M. Parris, Peter A. Vesk, and Michael A. McCarthy.
2014. ``Understanding Co-Occurrence by Modelling Species Simultaneously
with a Joint Species Distribution Model (JSDM).'' \emph{Methods in
Ecology and Evolution}, n/a--/a.
doi:\href{http://dx.doi.org/10.1111/2041-210X.12180}{10.1111/2041-210X.12180}.

Powell, Warren B. 2007. \emph{Approximate Dynamic Programming: Solving
the Curses of Dimensionality}. John Wiley \& Sons.

Robbins, Herbert, and Sutton Monro. 1951. ``A Stochastic Approximation
Method.'' \emph{The Annals of Mathematical Statistics} 22 (3): 400--407.
doi:\href{http://dx.doi.org/10.1214/aoms/1177729586}{10.1214/aoms/1177729586}.

Salakhutdinov, Ruslan. 2009. ``Learning in Markov Random Fields Using
Tempered Transitions.'' \emph{Advances in Neural Information Processing
Systems} 22: 1598--1606.
\url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.163.9244\&rep=rep1\&type=pdf}.

Salakhutdinov, Ruslan, and Geoffrey Hinton. 2012. ``An Efficient
Learning Procedure for Deep Boltzmann Machines.'' \emph{Neural
Computation} 24 (8): 1967--2006.
\url{http://www.mitpressjournals.org/doi/pdf/10.1162/NECO_a_00311}.

Schmidt, Mark, and Kevin Murphy. 2012. ``Modeling Discrete
Interventional Data Using Directed Cyclic Graphical Models.''
\emph{ArXiv Preprint ArXiv:1205.2617}.

Strong, Donald R, Daniel Simberloff, Lawrence G Abele, and Anne B
Thistle. 1984. \emph{Ecological Communities: Conceptual Issues and the
Evidence}. Princeton University Press.

Tjelmeland, Håkon, and Julian Besag. 1998. ``Markov Random Fields with
Higher-Order Interactions.'' \emph{Scandinavian Journal of Statistics}
25 (3): 415--33.
doi:\href{http://dx.doi.org/10.1111/1467-9469.00113}{10.1111/1467-9469.00113}.

Veech, Joseph A. 2013. ``A Probabilistic Model for Analysing Species
Co-Occurrence.'' \emph{Global Ecology and Biogeography} 22 (2): 252--60.
doi:\href{http://dx.doi.org/10.1111/j.1466-8238.2012.00789.x}{10.1111/j.1466-8238.2012.00789.x}.

Whittam, Thomas S., and Douglas Siegel-Causey. 1981. ``Species
Interactions and Community Structure in Alaskan Seabird Colonies.''
\emph{Ecology} 62 (6): 1515--24.
doi:\href{http://dx.doi.org/10.2307/1941508}{10.2307/1941508}.

Younes, Laurent. 1999. ``On the Convergence of Markovian Stochastic
Algorithms with Rapidly Decreasing Ergodicity Rates.''
\emph{Stochastics: An International Journal of Probability and
Stochastic Processes} 65 (3-4): 177--228.

### 2015-10-15 stochastic/stochastic-recovery.Rmd
clean up directory

@@
+---
+output: pdf_document
+---
 # Estimating pairwise interactions by stochastic maximum likelihood
 
-```{r}
+```{r,eval=FALSE}
 library(rosalia)
 library(mistnet)
-library(progress)
-library(beepr)
 
-beeping = TRUE
-```
+# convenience function for adding intercepts to each column
+`%plus%` = mistnet:::`%plus%`
+logistic = binomial()$linkinv # logistic inverse link
 
-```{r}
-`%plus%` = mistnet:::`%plus%`
-logistic = binomial()$linkinv
+# Random bernoulli trial
 rbern = function(p){rbinom(length(p), size = 1, prob = p)}
@@
 
-```{r}
+```{r,eval=FALSE}
 set.seed(1)
-n_spp = 250
-n_loc = 2500
-n_env = 5
-n_gibbs = 5000
+n_spp = 250     # number of species
+n_loc = 2500    # number of locations
+n_env = 5       # number of environmental predictors
+n_gibbs = 5000  # number of Gibbs sampling iterations
 
-type_frequencies = rmultinom(1, size = choose(n_spp, 2), prob = c(.2, .5, .3))
+# What portion of the coefficients should come from each 
+# mixture component?
+type_frequencies = rmultinom(1, size = choose(n_spp, 2), 
+                             prob = c(.2, .5, .3))
 
+# Vector of "true" interaction strengths.
+# Three mixture components, shuffled by `sample`.
 true_beta_vec = sample(
@@
 
+# Visualize the distribution of interaction terms
 plot(density(true_beta_vec))
 
+# Make a symmetric matrix of "true" beta coefficients
 true_beta = matrix(0, nrow = n_spp, ncol = n_spp)
@@
 
+# Create a set of environmental predictors
 true_env = matrix(rnorm(n_loc * n_env, 0, 2.5), nrow = n_loc)
 
+# "true" responses of each species to each environmental variable
 true_alpha_env = matrix(rnorm(n_spp * n_env, 0, 1), nrow = n_env)
 
+# "true" intercepts for each species
 true_alpha_species = rnorm(n_spp, 2)
 
+# site-level intercept depends on species baselines 
+# environmental responses
 true_alpha = true_env %*% true_alpha_env %plus% true_alpha_species
@@
 
-```{r, cache = TRUE}
+```{r,eval=FALSE, cache = TRUE}
+# Initialize the `y` matrix.
+# This will hold the "observed" presence-absence data
 y = matrix(0.5, nrow = n_loc, ncol = n_spp)
 
-pb <- progress_bar$new(
-  format = "  simulating landscape [:bar] :percent eta: :eta",
-  total = n_gibbs,
-  clear = FALSE
-)
-
+# For each round of Gibbs sampling...
 for(i in 1:n_gibbs){
-  pb$tick()
+  # For each species (in random order)...
   for(j in sample.int(n_spp)){
+    # update its occurrence with samples from the 
+    # conditional distribution
     y[,j] = rbern(logistic(true_alpha[ , j] + y %*% true_beta[ , j]))
@@
 }
-
-if(beeping){beep()}
 ```
 
-```{r}
+```{r,eval=FALSE}
 # Calculate sufficient statistics of the data
@@
 
-```{r}
+```{r,eval=FALSE}
+# Initialize the simulated landscape for stochastic approximation
 y_sim = matrix(0.5, nrow = nrow(y), ncol = ncol(y))
+
+# In this example, the true state of the environment is 
+# known without error
 env = true_env
 
+# Initialize species' responses to environment at 0.
+# Also initialize the delta (change in parameter values from the
+# previous optimization iteration) to zero, since no optimization
+# has occurred yet
 alpha_env = delta_alpha_env = matrix(0, nrow = n_env, ncol = n_spp)
+
+# Initialize species' intercepts to match observed occurrence rates
+# plus a small amount of regularization
 alpha_species = qlogis((colSums(y) + 1) / (nrow(y) + 2))
+
+# Initialize the deltas for the intercepts to zero
 delta_alpha_species = rep(0, n_spp)
+
+# Initialize pairwise interactions and deltas to zero
 beta = delta_beta = matrix(0, nrow = n_spp, ncol = n_spp)
-alpha = matrix(0, nrow = n_spp, ncol = n_spp) # no delta alpha to initialize
-                                              # b/c alpha not optimized directly
+
+# overall alpha depends on alpha_species and alpha_env.
+# Will be filled in later, so can initialize it with zeros
+# no delta alpha to initialize b/c alpha not optimized directly
+alpha = matrix(0, nrow = n_spp, ncol = n_spp) 
 ```
 
-```{r}
-# Are there lurking factors of two because I use the whole symmetric matrix?
-beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
+```{r,eval=FALSE}
+# Very weak priors on alpha terms, somewhat stronger on beta terms
 alpha_env_prior = rosalia::make_logistic_prior(scale = 2)$log_grad 
 alpha_species_prior = rosalia::make_logistic_prior(scale = 2)$log_grad
+beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
 ```
 
-```{r}
-initial_learning_rate = 1
-maxit = 50000
+```{r,eval=FALSE}
+initial_learning_rate = 1 # step size at start of optimization
+maxit = 50000             # Number of rounds of optimization
 start_time = as.integer(Sys.time())
 
-pb <- progress_bar$new(
-  format = "  Fitting [:bar] :percent eta: :eta",
-  total = maxit,
-  clear = FALSE
-)
+# Record the R-squared values in this vector
+r2s = numeric(maxit)
 
-r2s = numeric(maxit)
+# Record the timing history in this vector
 times = integer(maxit)
@@
 for(i in 1:maxit){
-  pb$tick()
-
   ##############################
@@
 
-  # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
+  # Sample entries in y_sim from their conditional 
+  # distribution (Gibbs sampling)
   for(j in sample.int(n_spp)){
@@
 
-  # Calculate the gradient with respect to alpha and beta
+  # Calculate the gradient with respect to alpha and beta.
+  # Gradients are differences in sufficient statistics plus prior
+  # gradients, all divided by the number of locations
   stats_difference = y_stats - y_sim_stats
@@
 
-  alpha_species_grad = (diag(stats_difference) + alpha_species_prior(alpha_species)) / n_loc
-  diag(beta_grad) = 0
+  alpha_species_grad = (
+    diag(stats_difference) + 
+      alpha_species_prior(alpha_species)
+  ) / n_loc
+  diag(beta_grad) = 0 # beta_ii is 0 by convention
   y_env_difference = y_env_stats - y_sim_env_stats
-  alpha_env_grad = (y_env_difference + alpha_env_prior(alpha_env))  / n_loc
+  alpha_env_grad = (y_env_difference + 
+                      alpha_env_prior(alpha_env))  / n_loc
 
-  # Calculate parameter updates
-  delta_beta = beta_grad * learning_rate / 10 + 
-    momentum * delta_beta
+  # Calculate parameter updates: gradient times learning rate 
+  # plus momentum times delta
+  delta_beta = beta_grad * learning_rate + momentum * delta_beta
   delta_alpha_species = alpha_species_grad * learning_rate + 
@@
 
+  # Add the deltas to the previous parameter values
   beta = beta + delta_beta
@@
 
-  r2s[i] = cor(true_beta[upper.tri(true_beta)], beta[upper.tri(beta)])^2
+  # Record R-squared and timing
+  r2s[i] = cor(
+    true_beta[upper.tri(true_beta)], 
+    beta[upper.tri(beta)]
+  )^2
   times[i] = as.integer(Sys.time()) - start_time
@@
 
-```{r}
+```{r,eval=FALSE}
 library(cowplot)
@@
 out = plot_grid(
-  ggplot(NULL, aes(x = beta[upper.tri(beta)], y = true_beta[upper.tri(beta)])) + 
+  ggplot(NULL, aes(x = beta[upper.tri(beta)], 
+                   y = true_beta[upper.tri(beta)])) + 
     stat_binhex() + 
@@
     stat_vline(xintercept = 0, size = 1/8) + 
-    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "log10") +
+    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", 
+                        trans = "log10") +
     stat_abline(intercept = 0, slope = 1, size = 1/2) + 
@@
   nrow = 1,
-  labels = c("A. Pairwise biotic coefficients (n = 31125)", "B. Abiotic coefficients (n = 1250)")
+  labels = c(
+    "A. Pairwise biotic coefficients (n = 31125)", 
+    "B. Abiotic coefficients (n = 1250)"
   )
-save_plot("stochastic/estimates.pdf", out, base_aspect_ratio = 2, base_height = 6)
+)
+save_plot(
+  "stochastic/estimates.pdf", 
+  out, 
+  base_aspect_ratio = 2, 
+  base_height = 6
+)
 
@@
 
-```{r}
+```{r,eval=FALSE}
 small_n_spp = 20
@@
   for(j in sample.int(small_n_spp)){
-    small_y[,j] = rbern(logistic(small_true_alpha[j] + small_y %*% small_true_beta[ , j]))
+    small_y[,j] = rbern(logistic(small_true_alpha[j] + 
+                                   small_y %*% small_true_beta[ , j]))
   }
@@
 
-```{r}
+```{r,eval=FALSE}
 exact_time = system.time({
-  exact = rosalia(small_y, prior = make_logistic_prior(scale = 1), maxit = 500)
+  exact = rosalia(small_y, 
+                  prior = make_logistic_prior(scale = 1), maxit = 500)
 })
@@
 
-```{r}
+```{r,eval=FALSE}
 maxit_small = 50000
@@
 beta_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad
-alpha_species_prior_small = rosalia::make_logistic_prior(scale = 1)$log_grad
+alpha_species_prior_small = 
+  rosalia::make_logistic_prior(scale = 1)$log_grad
 
@@
 delta_alpha_small = rep(0, small_n_spp)
-beta_small = delta_beta_small = matrix(0, nrow = small_n_spp, ncol = small_n_spp)
+beta_small = delta_beta_small = 
+  matrix(0, nrow = small_n_spp, ncol = small_n_spp)
 
@@
 
-  # Sample entries in y_sim from their conditional distribution (Gibbs sampling)
+  # Sample entries in y_sim from their conditional distribution 
+  # (Gibbs sampling)
   for(j in sample.int(small_n_spp)){
-    y_sim_small[,j] = rbern(logistic(alpha_small[j] + y_sim_small %*% beta_small[ , j]))
+    y_sim_small[,j] = rbern(
+      logistic(alpha_small[j] + 
+                 y_sim_small %*% beta_small[ , j])
+    )
   }
@@
   stats_difference = y_stats - y_sim_stats
-  beta_grad = (stats_difference + beta_prior_small(beta_small)) / small_n_loc
+  beta_grad = (
+    stats_difference + beta_prior_small(beta_small)
+  ) / small_n_loc
 
-  alpha_species_grad = (diag(stats_difference) +
-                          alpha_species_prior_small(alpha_small)) / small_n_loc
+  alpha_species_grad = (
+    diag(stats_difference) +
+      alpha_species_prior_small(alpha_small)
+  ) / small_n_loc
   diag(beta_grad) = 0
@@
 
-  mses[i] = mean((beta_small[upper.tri(beta_small)] - exact$beta[upper.tri(exact$beta)])^2)
+  mses[i] = mean((beta_small[upper.tri(beta_small)] - 
+                    exact$beta[upper.tri(exact$beta)])^2)
   small_times[i] = as.numeric(Sys.time()) - start_time_small
@@
 
-```{r}
+```{r,eval=FALSE}
 pdf("stochastic/convergence.pdf", width = 10, height = 7)
@@
 
-plot(small_times, mses, type = "l", ylab = "mean square deviation from MLE\n(log scale)", xlab = "time (seconds)", main = "A.
-Convergence to known MLE\n(small network)", yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, col = "blueviolet", xlim = range(c(-.1, small_times)))
+plot(small_times, mses, type = "l", 
+     ylab = "mean square deviation from MLE\n(log scale)", 
+     xlab = "time (seconds)", 
+     main = "A. Convergence to known MLE\n(small network)", 
+     yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, 
+     col = "blueviolet", xlim = range(c(-.1, small_times)))
 
-plot(c(0, times / 60 / 60), c(0, r2s), type = "l", xlab = "time (hours)", ylab = "R^2",ylim = c(0, max(r2s) * 1.04), main = "B.
-Apparent convergence to unknown MLE\n(large network)", yaxs = "i", xaxs = "i", lwd = 2, xlim = range(c(-.05, times / 60 / 60)), bty = "l", col = "blueviolet")
+plot(c(0, times / 60 / 60), c(0, r2s), type = "l", 
+     xlab = "time (hours)", ylab = "R^2",
+     ylim = c(0, max(r2s) * 1.04), 
+     main = "B. Apparent convergence to unknown MLE\n(large network)", 
+     yaxs = "i", xaxs = "i", lwd = 2, 
+     xlim = range(c(-.05, times / 60 / 60)), 
+     bty = "l", col = "blueviolet")
 abline(h = max(r2s), col = "#00000080", lty = 2, lwd = 2)

### 2015-10-16 Harris_2015/dynamic_simulation.Rmd
new simulations

--- 2015-10-15 Harris_2015/simulation.Rmd
@@
 output: pdf_document
-title: Appendix 2: simulating species interactions
+title: Appendix 2: simulating population dynamics with species interactions
 subtitle: \"Estimating species interactions from observational data with Markov networks\"
@@
 
-In the large population limit, the number of individuals of species $i$, $n_i$,
-evolves over time based on the following differential equation, based on 
-multi-species Lotka-Volterra dynamics.  The basic multi-species model is
+According to the deterministic skeleton of this model, the number of 
+individuals of species $i$, $n_i$, evolves over time based on the following
+differential equation, based on multi-species Lotka-Volterra dynamics.  The
+basic multi-species model is
 
@@
 $\alpha_{ij} > 0$ or as mutualists $\gamma_{ij} > 0$, but not both. Note that
-$\alpha_{ii}$ and $\gamma_{ii}$ were always zero (i.e. intraspecific competition
+$\alpha_{ii}$ and $\gamma_{ii}$ were always zero (i.e. *intra*specific competition
 is modeled outside of the $\alpha$ term and species cannot be their own
@@
 
-```{r, eval=FALSE}
+```{r, eval=TRUE}
 for(n_sites in c(25, 200, 1600)){
@@
   for(rep in 1:n_reps){
-    id = paste(n_spp, n_sites, rep, sep = "-")
+    id = paste(n_spp, n_sites, paste0("pop", rep), sep = "-")
 
@@
 
-    saveRDS(results, file = paste0("fakedata/", id, ".rds"))
+    write.csv(
+      results$observed, 
+      file = paste0("fakedata/matrices/", id, ".csv")
+    )
+
+    write(
+      results$truth, 
+      file = paste0("fakedata/truths/", id, ".txt"), 
+      ncolumns = length(results$truth)
+    )
 
@@
       t(x_subset), 
-      file = paste0("fakedata/", id, "-transposed.txt"), 
+      file = paste0("fakedata/matrices/", id, "-transposed.txt"), 
       quote = FALSE

### 2015-10-16 Harris_2015/simulation.Rmd
new simulations

--- 2015-08-01 Harris_2015/simulation.Rmd
@@
-```{r}
-library(progress)
-set.seed(1)
+# Analyses for Figure 3
 
-thin = 5E3
-n_spp = 20
-n_reps = 50
+## Simulating landscapes from known coefficients
+
+The following function draws random coefficients for a Markov network of a pre-specified size.
+The probability distribution was chosen so that most species pairs interact weakly, a few
+act strongly, and that most of the interactions were negative rather than positive.
+
+```{r, cache=TRUE}
+make_coefficients = function(n_spp){
+  # Exponential distribution has lots of mass near 0 but has
+  # a long tail.
+  true_beta_magnitudes = rexp(choose(n_spp, 2))
+
+  # Species interactions are negative 3/4 of the time in 
+  # these simulations.
+  b = true_beta_magnitudes * sample(
+    c(-1, 1), 
+    size = length(true_beta_magnitudes), 
+    prob = c(.75, .25),
+    replace = TRUE
+  )
+
+  # Species' intercepts are normally distributed
+  a = rnorm(n_spp, -1)
+
+  # Return the simulated values.
+  # The rosalia function stores pairwise parameters in the upper
+  # triangle of an n-by-n matrix and stores the species intercepts
+  # along the diagonal, so these values are named accordingly.
+  c(alpha  = a, beta = b)
+}
 ```
 
-$$r_e + r_i n_i (1 - \frac{n_i f_i(\vec{n}) + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.
+I used the following function to generate "true" parameters with the method above, 
+simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.
 
-When no propagules enter from outside the system ($r_e=0$) and mutualisms do
-not occur ($f_i(\vec{n})$ = 1), this reduces to a multi-species Lotka-Volterra
-system
+I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
+In each round of Gibbs sampling, I cycled through all the species, randomly
+updating each one's presence/absence vector in response to its conditional
+occurrence probability:
 
-$$r_i n_i (1 - \frac{n_i + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.
+$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$
 
-where $r_i$ is the [[intrinsic rate of increase]] of species $i$, $n_i$ is its
-abundance, and $\alpha_{ij}$ describes the strength of its competition with
-species $j$.  In the more general form above, the function $f_i(\vec{n})$ 
-acts to reduce the strength of intraspecific competition in the presence of
-mutualists.  For these analyses, I used
+where the logistic function is $\frac{1}{1+e^{-x}}$.
 
-(1 + exp(0.2 * x / K / n_spp)) / 2
+```{r, cache = TRUE}
+simulate_data = function(n_spp, n_sites, rep_name, n_gibbs = 1000, n_env = 2, sd = 2){
 
-$$f(\vec{n} = $$
+  # Determine the "true" parameters for the simulated assemblage
+  par = make_coefficients(n_spp)
 
-[[demographic events happen with probability proportional to corresponding birth/death rate]].  Because these events' probabilities only depend on the 
-current state of the system, they form a Markov chain that explores the 
-space of possible species assemblages under the population dynamic model.  For
-each set of competition and mutualism coefficients, I collected one sample 
-after every `r thin`th demographic event.  For the subsequent analysis, 
-each species was considered "present" in a sample if its abundance was one
-or higher, and "absent" if its abundance in that sample was zero.
+  # "True" interaction strengths, to save for later
+  truth = par[-(1:n_spp)]
 
-```{r}
-simulate_communities = function(n_spp = 20,
-         K = n_spp / 3,
-         maxit = 5E6,
-         thin = 1E4,
-         seed_rain = rep(.1, n_spp),
-         growth = rep(1, n_spp),
-         interaction_strength = 1/2,
-         prob_negative = 0.75,
-         mutualism_maximum = 0.5
-){
+  # "True" intercepts, possibly adjusted below by environment
+  alpha = par[1:n_spp]
 
-  # Intraspecific competition can be mitigated by mutualists, up to some
-  # maximal amount.  This function rescales the mutualism effect so that
-  # it doesn't turn off intraspecific competition entirely.
-  mutualism_scaler = function(x, mutualism_maximum){
-    (1 + exp(-4 * x / K / n_spp)) * (1 - mutualism_maximum)
-  }
+  # Turn the interaction values into an n-by-n matrix
+  # Start with empty matrix; fill in upper triangle; 
+  # then fill in lower triangle with its transpose
+  beta = matrix(0, n_spp, n_spp)
+  beta[upper.tri(beta)] = truth
+  beta = beta + t(beta)
 
-  # Randomly generate a vector of interaction strengths (one for each 
-  # pair of species)
-  interaction_vector = rexp(choose(n_spp, 2), interaction_strength)
+  # Environmental states are normally distributed
+  env = matrix(rnorm(n_sites * n_env), ncol = n_env)
 
-  # Randomly mark interactions as positive or negative
-  interaction_signs = ifelse(
-    rbinom(choose(n_spp, 2), size = 1, prob = prob_negative),
-    -1, 
-    1
+  alpha_env = matrix(rnorm(n_spp * n_env, sd = sd), nrow = n_env)
+
+  # Simulate the landscape from known process with Gibbs sampling
+
+  # Landscape starts as if betas were all zero. Each species' occurrence probability
+  # depends on its alpha value and on the environment (assuming alpha_env is not zero).
+  x = matrix(
+    plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
+    nrow = n_sites, 
+    ncol = n_spp
   )
 
-  # Negative interactions are competition; positive ones are mutualisms
-  competition_vector = interaction_vector * (interaction_signs == -1)
-  mutualism_vector = interaction_vector * (interaction_signs == +1)
+  # Gibbs sampling
+  for(i in 1:n_gibbs){
 
-  # Fill in the competition and mutualism matrices:
-  # Start with a matrix of zeros
-  competition = matrix(0, nrow = n_spp, ncol = n_spp)
-  # Fill in the upper triangle with the appropriate interaction vector
-  competition[upper.tri(competition)] = competition_vector
-  # Symmetrize the matrix by adding it to its transpose
-  competition = competition + t(competition)
-
-  # Note that intraspecific competition (which would go along the diagonal)
-  # is handled separately, and the diagonal is left with zeros.
-
-  mutualism = matrix(0, nrow = n_spp, ncol = n_spp)
-  mutualism[upper.tri(mutualism)] = mutualism_vector
-  mutualism = mutualism + t(mutualism)
-
-  # Create an empty matrix for storing population sizes
-  pops = matrix(nrow = n_spp, ncol = maxit / thin)
-
-  # Initialize the population with a fixed number of individuals per species,
-  # depending on the number of species, carrying capacity, and strength of
-  # interspecific competition.
-  population = rep(ceiling(K / n_spp / mean(competition)), n_spp)
-
-  for(i in 1:maxit){
-
-    # The local growth rate is proportional to population size
-    local_growth = growth * population
-
-    # Mutualists act by mitigating intraspecific competition
-    mutualist_mitigation = mutualism_scaler(
-      population %*% mutualism, 
-      mutualism_maximum
+    # Each round of Gibbs sampling updates one species (column) across all sites
+    # according to its conditional probability (i.e. conditional on environment
+    # and the other species that are present).
+    for(j in 1:n_spp){
+      x[,j] = rbinom(
+        nrow(x),
+        size = 1,
+        prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
       )
-
-    # Add up the two types of interactions affecting death rates
-    interactions = population * mutualist_mitigation + population %*% competition
-
-    # Germinating seeds can come from local growth or from outside the system
-    birthrates = seed_rain + local_growth
-
-    # Death rates depend on population size, interactions, and carrying capacity
-    deathrates = local_growth * interactions / K
-
-    # Generate a vector of vital rates
-    rates = c(birthrates, deathrates)
-
-    # Randomly select a birth/death event to occur. Probability
-    # of occurring is proportional to the corresponding rate
-    event = sample.int(length(rates), 1, prob = rates)
-
-    if(event <= length(birthrates)){
-      # Event from first batch, i.e. birth.  Increase the population by one
-      index = event
-      population[index] = population[index] + 1
-    }else{
-      # Event from the second batch, i.e. death. If population is nonzero,
-      # decrease it by one for the corresponding species
-      index = event - n_spp
-      if(population[index] > 0){
-        population[index] = population[index] - 1
     }
@@
 
-    if(i%%thin == 0){
-      # Save the results in the `pops` matrix
-      pops[ , i %/% thin] = population
-    }
-  }
+  # Save the results in a "fake data" folder
 
-  # Return the "observed" presence-absence vector
-  # along with the"true" interaction signs and strengths.
-  # The output is transposed because all fitting methods but Pairs
-  # require it in that format.
-  # Multiplying by one turns TRUEs into 1s and FALSEs into 0s
-  list(
-    observed = t(pops > 0) * 1,
-    truth = interaction_vector * interaction_signs
+  file_stem = paste(n_spp, n_sites, rep_name, sep = "-")
+
+  # Save the matrix of presence/absence observations
+  write.csv(
+    x, 
+    file = paste0("fakedata/matrices/", file_stem, ".csv")
+  )
+
+  # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
+  x_subset = x[rowSums(x) != 0, colSums(x) != 0]
+
+  # Gotelli and Ulrich's Pairs method expects the data matrices to be transposed,
+  # So I save them separately
+  write.table(
+    t(x_subset), 
+    file = paste0("fakedata/matrices/", file_stem, "-transposed.txt"), 
+    quote = FALSE
+  )
+
+  # Save the "true" species interactions
+  write(
+    truth, 
+    file = paste0("fakedata/truths/", file_stem, ".txt"), 
+    ncolumns = length(truth)
   )
@@
 
-```{r, eval=FALSE}
+```{r}
+set.seed(1)
+n_spp = 20
+
 for(n_sites in c(25, 200, 1600)){
-  message(n_sites)
-
-  pb = progress_bar$new(total = n_reps)
-  for(rep in 1:n_reps){
-    id = paste(n_spp, n_sites, rep, sep = "-")
-
-    results = simulate_communities(
-      maxit = n_sites * thin, 
-      thin = thin, 
-      n_spp = n_spp
-    )
-
-    pb$tick()
-
-    saveRDS(results, file = paste0("fakedata/", id, ".rds"))
-
-    # Gotelli and Ulrich's Pairs software needs a different format.
-    x = results$observed
-
-    # It rejects empty sites, so I remove them here
-    x_subset = x[rowSums(x) != 0, colSums(x) != 0]
-
-    # It also expects the data matrices to be transposed
-    # relative to the other methods
-    write.table(
-      t(x_subset), 
-      file = paste0("fakedata/", id, "-transposed.txt"), 
-      quote = FALSE
+  for(type in c("env", "no_env")){
+    for(i in 1:50){
+      simulate_data(
+        n_spp = n_spp, 
+        n_sites = n_sites, 
+        rep_name = paste0(type, i),
+        sd = ifelse(type == "env", 2, 0)
       )
@@
   }
+}
+
 ```

### 2015-10-25 Harris_2015/manuscript-materials/draft.Rmd
various

--- 2015-08-05 Harris_2015/manuscript-materials/draft.Rmd
@@
 
-**Title:** Estimating species interactions from observational data with Markov
+**Title:** Estimating species interactions from co-occurrence data with Markov
 networks
@@
 
-**Abstract:** Estimating species interactions from observational data is one of
+**Abstract:** Estimating species interactions from co-occurrence data is one of
 the most controversial tasks in community ecology. One difficulty is that a
@@
 the direct and indirect consequences of any possible species interaction matrix.
-Interactions in these models can be estimated from observational data via
-maximum likelihood. Using simulated landscapes with known pairwise interaction
-strengths, I evaluated Markov networks and six existing approaches. The
-Markov networks consistently outperformed other methods, correctly isolating
+Interactions in these models can also be estimated from observed co-occurrence
+rates via maximum likelihood. Using simulated landscapes with known pairwise
+interaction strengths, I evaluated Markov networks and six existing approaches.
+The Markov networks consistently outperformed other methods, correctly isolating
 direct interactions between species pairs even when indirect interactions
 largely overpowered them. Two computationally efficient approximations, based on
-linear and generalized linear models, also performed well. Indirect effects
-reliably caused a common null modeling approach to produce incorrect inferences,
-however.
+controlling for indirect effects with linear or generalized linear models,
+also performed well. Indirect effects reliably caused a common null modeling
+approach to produce incorrect inferences, however.
 
@@
 Despite decades of work and several major controversies, however
-[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
+[@lewin_santa_1983; @strong_ecological_1984;
 @connor_checkered_2013], existing methods for detecting competition's effects
 on community structure are unreliable [@gotelli_empirical_2009]. In particular,
-species' effects on one another can become lost in the complex web of direct and
-indirect interactions in real assemblages. For example, the competitive
-interaction between the two shrub species in Figure 1A can become obscured by
-their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
-have long known that indirect effects can overwhelm direct ones at the landscape
-level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
-inferenes from observational data do not control for these effects
+species' effects on one another can become lost in the complex web of
+indirect effects. For example, the competitive interaction between the two shrub
+species in Figure 1A can become obscured by their shared tendency to occur in
+unshaded areas (Figure 1B). While ecologists have long known that indirect
+effects can overwhelm direct ones at the landscape level
+[@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our
+methods for drawing inferences from observational data do not control for these effects
 [e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@@
 While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
-level (as most of our methods assume), it nevertheless does leave a signal in
-the data (Figure 1C). Specifically, after partitioning the data set into shaded
+level (as most methods assume), it nevertheless does leave a signal in
+the data (Figure 1C). For example, after partitioning the data set into shaded
 sites and unshaded sites, there will be co-occurrence deficits in each subset
-that might not be apparent at the landscape level. More generally,
-we can obtain much better estimates of the association between two species from
-their conditional relationships (i.e. by controlling for other species in the
-network) than we could get from their overall co-occurrence rates. This kind of
-precision is difficult to obtain from null models, which begin with the
-assumption that all the pairwise interactions are zero and thus don’t need to
-be controlled for.  Nevertheless, null models have dominated this field for more
-than three decades [@strong_ecological_1984; @gotelli_empirical_2009].
+that might not be apparent at the whole-landscape level. More generally,
+controlling for other species in the network will often be important for obtaining
+reliable estimates of direct (conditional, or all-else-equal) effects.  This kind of precision is
+difficult to obtain from null models, which only simulate the most extreme
+possible scenario: that *all* direct and indirect interactions are exactly zero.
+Nevertheless, null models have dominated this  field for more than three decades
+[@strong_ecological_1984; @gotelli_empirical_2009].
 
-Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
-paper shows that Markov networks [undirected graphical models also known as
+Following @azaele_inferring_2010, this paper shows that Markov networks
+[undirected graphical models also known as
 Markov random fields; @murphy_machine_2012] can provide a framework for
@@
 Below, I introduce Markov networks and show how they can be used to simulate
-landscape-level data or to make exact predictions about the direct and
-indirect consequences of possible interaction matrices. Then, using simulated
-data sets where the "true" interactions are known, I compare this approach with
-several existing methods. Finally, I discuss opportunities for extending the
-approach presented here to other problems in community ecology, e.g. quantifying
-the overall effect of species interactions on occurrence rates
-[@roughgarden_competition_1983] and disentangling the effects of biotic versus
-abiotic interactions on species composition [@kissling_towards_2012;
-@pollock_understanding_2014].
+landscape-level data from an interaction matrix or to make exact predictions
+about the direct and indirect consequences of possible interaction matrices.
+Then, using simulated data sets where the "true" interactions are known,
+I compare this approach with several existing methods. Finally, I discuss
+opportunities for extending the approach presented here to other problems in
+community ecology, e.g. quantifying the overall effect of species interactions
+on occurrence rates [@roughgarden_competition_1983] and disentangling the
+effects of biotic versus abiotic interactions on species composition
+[@pollock_understanding_2014].
 
@@
 Markov networks provide a framework for translating back and forth between the
-conditional relationships among species (Figure 1C) and the kinds of species
-assemblages that these relationships produce. Here, I show how a set of
-conditional relationships can be used to determine how groups of species
+conditional (all-else-equal) relationships among species (Figure 1C) and the
+kinds of species assemblages that these relationships produce. Here, I show how
+a set of conditional relationships can be used to determine how groups of species
 can co-occur. Methods for estimating conditional relationships from data are
@@
 
-$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
+$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),$
 
@@
 
-Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
+\noindent where the second sum is over all $\binom{n}{2}$ species pairs. In this model,  $\alpha_{i}$ is
+an intercept term determining the amount that the presence
 of species $i$ contributes to the log-probability of $\vec{y}$; it directly
@@
 competitor is associated with release from another competitor), then species'
-marginal association rates can differ from this expectation. Determining the
+overall co-occurrence patterns can be more complicated. Determining the
 marginal relationships between species from their conditional interactions
@@
 practice, however, ecologists will often need to estimate the parameters from
-an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
-indicating which species are present at which sites). When the number of species
+an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
+indicating which species are present at each site on the landscape). When the number of species
 is reasonably small, one can compute exact maximum likelihood estimates for all
@@
 that would be expected to produce exactly the observed co-occurrence frequencies
-on average [i.e. maximizing the likelihood matches the sufficient statistics of the model distribution
-to the sufficient statistics of the data; @murphy_machine_2012]. I used the
-rosalia package [@harris_rosalia_2015] for the R programming language
-[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
-named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
-miraculous healing powers played an important rhetorical role in the null model
+on average [i.e. maximizing the likelihood ensures that the sufficient
+statistics of the model distribution match the sufficient statistics of the data;
+@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
+R programming language [@r_core_team_r_2015] to optimize the Markov network
+parameters. The package was named after Santa Rosalia, the patron saint of
+biodiversity, whose mythos played an important rhetorical role in the null model
 debates of the 1970's and 1980's [@lewin_santa_1983].
@@
 #### Simulated landscapes.
-In order to compare different methods, I simulated two sets of landscapes using
-known parameters. The first set included the three competing species shown in
-Figure 1. For each of 1000 replicates, I generated a landscape with 100 sites by
-sampling from a probability distribution defined by the figure's interaction
-coefficients (Appendix 1). Each of the methods described below was then
-evaluated on its ability to correctly infer that the two shrub species competed
-with one another, despite their frequent co-occurrence.
+I simulated several sets of landscapes using known parameters as a test bed for
+different statistical methods. The first set of landscapes included the three
+competing species shown in Figure 1. For each of 1000 replicates, I generated a
+landscape with 100 sites by sampling from a probability distribution defined by
+the figure's interaction coefficients (Appendix 1). Each of the methods
+described below was then evaluated on its ability to correctly infer that the
+two shrub species competed with one another, despite their frequent co-occurrence.
 
-I also simulated a second set of landscapes using a stochastic community model
-based on generalized Lotka-Volterra dynamics, as described in Appendix 2. In
-these simulations, each species pair was randomly assigned to either compete for
-a portion of the available carrying capacity (negative interaction) or to act as
-mutualists (positive interaction).  Here, mutualisms operate by mitigating the
-effects of intraspecific competition on each partner's death rate.  For these
-analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
-sites (50 replicates per landscape size; see Appendix 2).
+I then generated landscapes with up to 20 interacting species at 25, 200, or
+1600 sites using three increasingly complex models (50 replicates for each
+combination of size and model). As described in Appendix [[X]], I randomly drew
+the “true” coefficient values for each replicate landscape in these simulations
+so that most species pairs interacted negligibly, a few pairs interacted very
+strongly, and competition was three times more common than facilitation.
+
+The first set of 20-species landscapes, like the landscapes with three species,
+were generated directly from a Markov network to ensure that the model could
+recover the parameters used to generate the "observed" co-occurrence data.
+Then, I added two environmental factors that varied from location to location
+across the simulated landscapes, and simulated a new set of co-occurrence data
+so that species' $\alpha$ coefficients depended on these environmental variables.
+The latter set of simulated landscapes provide an important test of the methods’
+ability to distinguish co-occurrence patterns that were generated from pairwise
+biotic interactions from those that were generated by external forces like
+abiotic environmental filtering. This task was made especially difficult
+because---as with most analyses of presence-absence data for co-occurrence
+patterns---the inference procedure did not have access to any information about
+the environmental or spatial variables that helped shape the landscape [cf
+@connor_checkered_2013].  Finally, I simulated the population dynamics of 20
+interacting species, as described in Appendix [[X]]
 
 #### Recovering species interactions from simulated data.
-<!-- exp(qlogis(.975, scale = 2)) -->
 
@@
 associations between pairs of species from simulated data (Appendix 3).
-First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
-models,  as described above. For the analyses with 20 species, I added a very
-weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
-to ensure that the model estimates were always finite. The bias introduced by
-this prior should be small: the 95% credible interval on $\beta$ only requires
-that one species' effect on the odds of observing a different species to be less
-than a factor of 1500 (which is not much of a constraint). The logistic
-distribution was chosen because it is convex and has a similar shape to the
-Laplace distribution used in LASSO regularization (especially in the tails), but
-unlike the Laplace distribution it is differentiable everywhere and does not
-force any estimates to be exactly zero. To confirm that this procedure produced
-stable estimates, I compared its estimates on 50 bootstrap replicates
-(Appendix 4).
+First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
+models,  as described above. For the analyses with 20 species, I added a weak
+regularizer (equivalent to a logistic prior with scale 2) to ensure that the
+model estimates were always finite [[Appendix]].
 
@@
 which summarizes their marginal association.  Next, I used partial correlations,
-which summarize species' conditional relationships [@albrecht_spatial_2001;
-@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
-correlation can be thought of as a computationally efficient approximation to
-the full Markov network model [@loh_structure_2013]. This sort of model is very
-common for estimating relationships among genes and gene products
-[@friedman_sparse_2008]. Because partial correlations are undefined for
+which summarize species' conditional relationships. This approach is
+common in molecular biology [@friedman_sparse_2008], but less so in ecology
+[see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions].
+In the context of non-Gaussian data, the partial correlation can be thought of
+as a computationally efficient approximation to the full Markov network model
+[@loh_structure_2013]. Because partial correlations are undefined for
 landscapes with perfectly-correlated species pairs, I used a regularized
@@
 
-The third alternative, generalized linear models (GLMs), can also be
-thought of as a computationally efficient approximation to the Markov
-network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
-regularized logistic regression models [@gelman_weakly_2008] for each species,
-using the other species on the landscape as predictors. To avoid the
-identifiability problems associated with directed cyclic graphs
-[@schmidt_modeling_2012], I then symmetrized the relationships within species
-pairs via averaging.
+The third alternative, generalized linear models (GLMs), also provide a
+computationally efficient approximation to the Markov network
+[@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized
+logistic regression models [@gelman_weakly_2008] for each species, using the
+other species on the landscape as predictors. This produced two interaction
+estimates for each species pair; one for the effect of species $i$ on species
+$j$ and one for the reverse. These two coefficients are not identifiable from
+the data, however [@schmidt_modeling_2012], so I used their average as an
+overall measure of the overall relationship between species $i$ and species $j$.
 
@@
 interaction strengths among all of the species pairs from the larger simulated
-landscapes. This determined which of the methods provide a consistent way to know
-how strong species interactions are---regardless of which species were present
-in a particular data set or how many observations were taken. Because the
-different methods mostly describe species interactions on different scales
-(e.g. correlations versus $Z$ scores versus regression coefficients), I used
-linear regression through the origin to rescale the different estimates produced
-by each method so that they had a consistent interpretation. After rescaling
-each method’s estimates, I calculated squared errors between the scaled
-interaction estimates and “true” interaction values across all the simulated
-data sets. These squared errors determined the proportion of variance explained
+landscapes. This determined which of the methods provide a consistent way to
+know how strong species interactions are---regardless of which species were
+present in a particular data set or how many observations were taken. To ensure
+that the different test statistics (e.g. correlations versus $Z$ scores) were on
+a common scale, I rescaled them using linear regression through the origin. After
+rescaling each method’s estimates, I calculated squared errors between the scaled
+interaction estimates and “true” interaction values across the simulated data
+sets. These squared errors determined the proportion of variance explained
 for different combinations of model type and landscape size (compared with a
@@
 #### Twenty species.
-Despite some variability across contexts (Figure 3A), the four methods that
-controlled for indirect effects clearly performed the best: the Markov network
-explained the largest portion of the variance in the "true" interaction
-coefficients (35% overall), followed by the generalized linear models (30%),
-partial correlations from the raw presence-absence data (28%), and partial
-correlations from BayesComm, the joint species distribution model (26%).  The
-benefit of choosing the full Markov network over the other three methods was
-largest on the smaller landscapes, which are also the ones that are most
-representative of typical analyses in this field [@gotelli_empirical_2009].
+Despite some variability across contexts (Figure 3), the four methods that
+controlled for indirect effects (the Markov network, the generalized linear
+models, and the two partial correlation-based methods) clearly outperformed those
+that did not.  Of these, the Markov network consistently performed the best,
+especially on the smaller landscapes [which are also the ones that are most
+representative of typical analyses in this field; @gotelli_empirical_2009].
 
-The three methods that did not attempt to control for indirect interactions all
-explained less than 20% of the variance. Of these, the sample correlation matrix
-based on the raw data performed the best (19%), followed by the null model
-(15%) and BayesComm's correlation matrix (11%). Although these last three
-methods had different $R^2$ values, there was a close mapping among
-their estimates (especially after controlling for the size of the simulated
-landscapes; Figure 3B).  This suggests that the effect sizes from the null model
-(and, to a lesser extent, the correlation matrices from joint species
-distribution models) only contain noisy versions of the same information that
-could be obtained more easily and interpretably by calculating correlation
-coefficients between species' presence-absence vectors.
+Of the methods that did not control for indirect effects, Figure 3 shows that
+simple correlation coefficients provided a more reliable indicator of species'
+true interaction strengths than either the joint species distribution model
+(BayesComm) or the null model (Pairs). Interestingly, the estimates from these
+three approaches were tightly correlated [[Figure 3B]]. This suggests that the
+effect sizes from the null model (and, to a lesser extent, the correlation
+matrices from joint species distribution models) only contain noisy versions of
+the same information that could be obtained more easily and interpretably by
+calculating correlation coefficients between species' presence-absence vectors.
 
@@
 bootstrap distribution indicated that the Markov network's overall $R^2$ value
-was between 14 and 18 percent higher than the second-most effective
-method (generalized linear models) and between 2.12 and 2.38 times higher than
-could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
-landscape also confirmed that the rosalia package's estimates of species'
-conditional relationships were robust to sampling variation for reasonably-sized
-landscapes (Appendix 4).
+was between [[14 and 18]] percent higher than the second-most effective
+method (generalized linear models) and between [[2.12 and 2.38]] times higher than
+could be achieved by the null model (Pairs).
 
@@
 species' pairwise interactions from observational data, even for cases where
-a common null modeling technique reliably fails. Specifically, Markov networks
-were successful even when direct interactions were largely overwhelmed by
-indirect effects (Figure 1). For cases where fitting a Markov network is
+environmental heterogeneity and indirect interactions cause ecologists' typical
+null modeling approaches to reliably fail. For cases where fitting a Markov network is
 computationally infeasible, these results also indicate that partial covariances
-and generalized linear models (the two methods that estimated conditional
-relationships rather than marginal ones) can both provide useful approximations.
+and generalized linear models can both provide useful approximations.
 The partial correlations' success on simulated data may not carry over to real
@@
 incorporating whole networks of biotic interactions into species distribution
-models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
+models [@pollock_understanding_2014], leading to a
 better understanding of the interplay between biotic and abiotic effects on
@@
 
-**References:**
-
-**Figure captions**
-
-**Figure 1.**
-**A.** A small network of three competing species. The tree (top) tends not to
+**Figure 1: A.** A small network of three competing species. The tree (top) tends not to
 co-occur with either of the two shrub species, as indicated by the strongly
@@
 correlated (circled). **C.** Controlling for the tree species' presence with a
-conditional method such as a partial covariance or a Markov network leads to
-correct identification of the negative shrub-shrub interaction (circled).
+conditional (all-else-equal) approach such as a partial covariance or a
+Markov network leads to correct identification of the negative shrub-shrub
+interaction (circled).
 
-**Figure 2.**
-**A.** A small Markov network with two species.  The abiotic environment favors
-the occurrence of both species ($\alpha >0$), particularly species 2
-($\alpha_2 > \alpha_1$). The  negative $\beta$ coefficient linking these two
-species implies that they co-occur less than expected under independence.
-**B.** Relative probabilities of all four possible presence-absence combinations
-for Species 1 and Species 2. The exponent includes $\alpha_1$ whenever Species 1
-is present ($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the
-exponent includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
+**Figure 2: A.** A small Markov network with two species, defined by its $\alpha$ and
+$\beta$ values.  The abiotic environment favors the occurrence of each species
+($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
+$\beta$ coefficient linking these two species implies that they co-occur less
+than expected under independence. **B.** These coefficients determine the
+probabilities of all four possible presence-absence combinations for Species 1
+and Species 2. $\alpha_1$ is added to the exponent whenever Species 1 is present
+($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the exponent
+includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
 $\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
-ensures that the four relative probabilities sum to 1.  In this case, $Z$ is
-about 18.5.  **C.** We can find the expected frequencies of all possible
-co-occurrence patterns between the two species of interest. **D.** If
-$\beta_{12}$ equaled zero (e.g. if the species no longer competed for the same
-resources), then the reduction in competition would allow each species to
-increase its occurrence rate and the co-occurrence deficit would be eliminated.
+ensures that the four probabilities sum to 1.  In this case, $Z$ is
+about 18.5.  **C.** The expected frequencies of all possible
+co-occurrence patterns between the two species of interest, as calculated in the
+previous panel. **D.** If, as assumed in null model simulations, $\beta_{12}$
+equaled zero (e.g. if the species no longer competed for the same resources),
+then the reduction in competition would allow each species to increase its
+overall occurrence rate.
 
-**Figure 3.**
-**A.** Proportion of variance in interaction coefficients explained by each
+**Figure 3: A.** Proportion of variance in interaction coefficients explained by each
 method versus number of sampled locations.
@@
 
+**References:**
+
 \setlength{\parskip}{0pt}

### 2015-10-25 Harris_2015/manuscript-materials/draft.tex
various

--- 2015-08-05 Harris_2015/manuscript-materials/draft.tex
@@
 
-\textbf{Title:} Estimating species interactions from observational data
+\textbf{Title:} Estimating species interactions from co-occurrence data
 with Markov networks
@@
 
-\textbf{Abstract:} Estimating species interactions from observational
+\textbf{Abstract:} Estimating species interactions from co-occurrence
 data is one of the most controversial tasks in community ecology. One
@@
 predict the direct and indirect consequences of any possible species
-interaction matrix. Interactions in these models can be estimated from
-observational data via maximum likelihood. Using simulated landscapes
-with known pairwise interaction strengths, I evaluated Markov networks
-and six existing approaches. The Markov networks consistently
-outperformed other methods, correctly isolating direct interactions
-between species pairs even when indirect interactions largely
-overpowered them. Two computationally efficient approximations, based on
-linear and generalized linear models, also performed well. Indirect
-effects reliably caused a common null modeling approach to produce
-incorrect inferences, however.
+interaction matrix. Interactions in these models can also be estimated
+from observed co-occurrence rates via maximum likelihood. Using
+simulated landscapes with known pairwise interaction strengths, I
+evaluated Markov networks and six existing approaches. The Markov
+networks consistently outperformed other methods, correctly isolating
+direct interactions between species pairs even when indirect
+interactions largely overpowered them. Two computationally efficient
+approximations, based on controlling for indirect effects with linear or
+generalized linear models, also performed well. Indirect effects
+reliably caused a common null modeling approach to produce incorrect
+inferences, however.
 
@@
 1975). Despite decades of work and several major controversies, however
-(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
-al. 2013), existing methods for detecting competition's effects on
-community structure are unreliable (Gotelli and Ulrich 2009). In
-particular, species' effects on one another can become lost in the
-complex web of direct and indirect interactions in real assemblages. For
+(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
+for detecting competition's effects on community structure are
+unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
+one another can become lost in the complex web of indirect effects. For
 example, the competitive interaction between the two shrub species in
@@
 (Dodson 1970, Levine 1976), the vast majority of our methods for drawing
-inferenes from observational data do not control for these effects (e.g.
-Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
-Pollock et al. 2014). To the extent that indirect interactions like
-those in Figure 1 are generally important (Dodson 1970), existing
+inferences from observational data do not control for these effects
+(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
+2013, Pollock et al. 2014). To the extent that indirect interactions
+like those in Figure 1 are generally important (Dodson 1970), existing
 methods will thus not generally provide much evidence regarding species'
@@
 While competition doesn't reliably reduce co-occurrence rates at the
-whole-landscape level (as most of our methods assume), it nevertheless
-does leave a signal in the data (Figure 1C). Specifically, after
-partitioning the data set into shaded sites and unshaded sites, there
-will be co-occurrence deficits in each subset that might not be apparent
-at the landscape level. More generally, we can obtain much better
-estimates of the association between two species from their conditional
-relationships (i.e.~by controlling for other species in the network)
-than we could get from their overall co-occurrence rates. This kind of
-precision is difficult to obtain from null models, which begin with the
-assumption that all the pairwise interactions are zero and thus don't
-need to be controlled for. Nevertheless, null models have dominated this
-field for more than three decades (Strong et al. 1984, Gotelli and
+whole-landscape level (as most methods assume), it nevertheless does
+leave a signal in the data (Figure 1C). For example, after partitioning
+the data set into shaded sites and unshaded sites, there will be
+co-occurrence deficits in each subset that might not be apparent at the
+whole-landscape level. More generally, controlling for other species in
+the network will often be important for obtaining reliable estimates of
+direct (conditional, or all-else-equal) effects. This kind of precision
+is difficult to obtain from null models, which only simulate the most
+extreme possible scenario: that \emph{all} direct and indirect
+interactions are exactly zero. Nevertheless, null models have dominated
+this field for more than three decades (Strong et al. 1984, Gotelli and
 Ulrich 2009).
 
-Following recent work by Azaele et al. (2010) and Fort (2013), this
-paper shows that Markov networks (undirected graphical models also known
-as Markov random fields; Murphy 2012) can provide a framework for
-understanding the landscape-level consequences of pairwise species
-interactions, and for detecting them from observed presence-absence
-matrices. Markov networks have been used in many scientific fields in
-similar contexts for decades, from physics (where nearby particles
-interact magnetically; Cipra 1987) to spatial statistics (where adjacent
-grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
-While community ecologists explored some related approaches in the
-1980's (Whittam and Siegel-Causey 1981), they used severe approximations
-that led to unintelligible results (e.g. ``probabilities'' greater than
-one; Gilpin and Diamond 1982).
+Following Azaele et al. (2010), this paper shows that Markov networks
+(undirected graphical models also known as Markov random fields; Murphy
+2012) can provide a framework for understanding the landscape-level
+consequences of pairwise species interactions, and for detecting them
+from observed presence-absence matrices. Markov networks have been used
+in many scientific fields in similar contexts for decades, from physics
+(where nearby particles interact magnetically; Cipra 1987) to spatial
+statistics (where adjacent grid cells have correlated values; Harris
+1974, Gelfand et al. 2005). While community ecologists explored some
+related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
+used severe approximations that led to unintelligible results (e.g.
+``probabilities'' greater than one; Gilpin and Diamond 1982).
 
 Below, I introduce Markov networks and show how they can be used to
-simulate landscape-level data or to make exact predictions about the
-direct and indirect consequences of possible interaction matrices. Then,
-using simulated data sets where the ``true'' interactions are known, I
-compare this approach with several existing methods. Finally, I discuss
-opportunities for extending the approach presented here to other
-problems in community ecology, e.g.~quantifying the overall effect of
-species interactions on occurrence rates (Roughgarden 1983) and
-disentangling the effects of biotic versus abiotic interactions on
-species composition (Kissling et al. 2012, Pollock et al. 2014).
+simulate landscape-level data from an interaction matrix or to make
+exact predictions about the direct and indirect consequences of possible
+interaction matrices. Then, using simulated data sets where the ``true''
+interactions are known, I compare this approach with several existing
+methods. Finally, I discuss opportunities for extending the approach
+presented here to other problems in community ecology, e.g.~quantifying
+the overall effect of species interactions on occurrence rates
+(Roughgarden 1983) and disentangling the effects of biotic versus
+abiotic interactions on species composition (Pollock et al. 2014).
 
@@
 Markov networks provide a framework for translating back and forth
-between the conditional relationships among species (Figure 1C) and the
-kinds of species assemblages that these relationships produce. Here, I
-show how a set of conditional relationships can be used to determine how
-groups of species can co-occur. Methods for estimating conditional
-relationships from data are discussed in the next section.
+between the conditional (all-else-equal) relationships among species
+(Figure 1C) and the kinds of species assemblages that these
+relationships produce. Here, I show how a set of conditional
+relationships can be used to determine how groups of species can
+co-occur. Methods for estimating conditional relationships from data are
+discussed in the next section.
 
@@
 
-\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).\)
+\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),\)
 
@@
 
-Here, \(\alpha_{i}\) is an intercept term determining the amount that
-the presence of species \(i\) contributes to the log-probability of
-\(\vec{y}\); it directly controls the prevalence of species \(i\).
-Similarly, \(\beta_{ij}\) is the amount that the co-occurrence of
-species \(i\) and species \(j\) contributes to the log-probability; it
-controls the conditional relationship between two species, i.e.~the
-probability that they will be found together, after controlling for the
-other species in the network (Figure 2A, Figure 2B). For example,
-\(\beta_{ij}\) might have a value of \(+2\) for two mutualists,
-indicating that the odds of observing one species are \(e^2\) times
-higher in sites where its partner is present than in comparable sites
-where its partner is absent. Because the relative probability of a
-presence-absence vector increases when positively-associated species
-co-occur and decreases when negatively-associated species co-occur, the
-model tends---all else equal---to produce assemblages that have many
-pairs of positively-associated species and relatively few pairs of
+\noindent where the second sum is over all \(\binom{n}{2}\) species
+pairs. In this model, \(\alpha_{i}\) is an intercept term determining
+the amount that the presence of species \(i\) contributes to the
+log-probability of \(\vec{y}\); it directly controls the prevalence of
+species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
+co-occurrence of species \(i\) and species \(j\) contributes to the
+log-probability; it controls the conditional relationship between two
+species, i.e.~the probability that they will be found together, after
+controlling for the other species in the network (Figure 2A, Figure 2B).
+For example, \(\beta_{ij}\) might have a value of \(+2\) for two
+mutualists, indicating that the odds of observing one species are
+\(e^2\) times higher in sites where its partner is present than in
+comparable sites where its partner is absent. Because the relative
+probability of a presence-absence vector increases when
+positively-associated species co-occur and decreases when
+negatively-associated species co-occur, the model tends---all else
+equal---to produce assemblages that have many pairs of
+positively-associated species and relatively few pairs of
 negatively-associated species (exactly as an ecologist might expect).
@@
 presence of one competitor is associated with release from another
-competitor), then species' marginal association rates can differ from
-this expectation. Determining the marginal relationships between species
-from their conditional interactions entails summing over the different
+competitor), then species' overall co-occurrence patterns can be more
+complicated. Determining the marginal relationships between species from
+their conditional interactions entails summing over the different
 possible assemblages (Figure 2B). This becomes intractable when the
@@
 estimate the parameters from an observed co-occurrence matrix (i.e.~from
-a matrix of ones and zeros indicating which species are present at which
-sites). When the number of species is reasonably small, one can compute
-exact maximum likelihood estimates for all of the \(\alpha\) and
-\(\beta\) coefficients given a presence-absence matrix by optimizing
-\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
-ones considered here have unimodal likelihood surfaces (Murphy 2012),
-ensuring that this procedure will always converge on the global maximum.
-This maximum represents the unique combination of \(\alpha\) and
-\(\beta\) coefficients that would be expected to produce exactly the
-observed co-occurrence frequencies on average (i.e.~maximizing the
-likelihood matches the sufficient statistics of the model distribution
-to the sufficient statistics of the data; Murphy 2012). I used the
-rosalia package (Harris 2015a) for the R programming language (R Core
-Team 2015) to optimize the Markov network parameters. The package was
-named after Santa Rosalia, the patron saint of biodiversity, whose
-supposedly miraculous healing powers played an important rhetorical role
-in the null model debates of the 1970's and 1980's (Lewin 1983).
+a set of independent \(\vec{y}\) vectors indicating which species are
+present at each site on the landscape). When the number of species is
+reasonably small, one can compute exact maximum likelihood estimates for
+all of the \(\alpha\) and \(\beta\) coefficients given a
+presence-absence matrix by optimizing \(p(\vec{y}; \alpha, \beta)\).
+Fully-observed Markov networks like the ones considered here have
+unimodal likelihood surfaces (Murphy 2012), ensuring that this procedure
+will always converge on the global maximum. This maximum represents the
+unique combination of \(\alpha\) and \(\beta\) coefficients that would
+be expected to produce exactly the observed co-occurrence frequencies on
+average (i.e.~maximizing the likelihood ensures that the sufficient
+statistics of the model distribution match the sufficient statistics of
+the data; Murphy 2012). I used the rosalia package (Harris 2015a) for
+the R programming language (R Core Team 2015) to optimize the Markov
+network parameters. The package was named after Santa Rosalia, the
+patron saint of biodiversity, whose mythos played an important
+rhetorical role in the null model debates of the 1970's and 1980's
+(Lewin 1983).
 
@@
 
-In order to compare different methods, I simulated two sets of
-landscapes using known parameters. The first set included the three
-competing species shown in Figure 1. For each of 1000 replicates, I
-generated a landscape with 100 sites by sampling from a probability
-distribution defined by the figure's interaction coefficients (Appendix
-1). Each of the methods described below was then evaluated on its
-ability to correctly infer that the two shrub species competed with one
-another, despite their frequent co-occurrence.
+I simulated several sets of landscapes using known parameters as a test
+bed for different statistical methods. The first set of landscapes
+included the three competing species shown in Figure 1. For each of 1000
+replicates, I generated a landscape with 100 sites by sampling from a
+probability distribution defined by the figure's interaction
+coefficients (Appendix 1). Each of the methods described below was then
+evaluated on its ability to correctly infer that the two shrub species
+competed with one another, despite their frequent co-occurrence.
 
-I also simulated a second set of landscapes using a stochastic community
-model based on generalized Lotka-Volterra dynamics, as described in
-Appendix 2. In these simulations, each species pair was randomly
-assigned to either compete for a portion of the available carrying
-capacity (negative interaction) or to act as mutualists (positive
-interaction). Here, mutualisms operate by mitigating the effects of
-intraspecific competition on each partner's death rate. For these
-analyses, I simulated landscapes with up to 20 species and 25, 200, or
-1600 sites (50 replicates per landscape size; see Appendix 2).
+I then generated landscapes with up to 20 interacting species at 25,
+200, or 1600 sites using three increasingly complex models (50
+replicates for each combination of size and model). As described in
+Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
+for each replicate landscape in these simulations so that most species
+pairs interacted negligibly, a few pairs interacted very strongly, and
+competition was three times more common than facilitation.
+
+The first set of 20-species landscapes, like the landscapes with three
+species, were generated directly from a Markov network to ensure that
+the model could recover the parameters used to generate the ``observed''
+co-occurrence data. Then, I added two environmental factors that varied
+from location to location across the simulated landscapes, and simulated
+a new set of co-occurrence data so that species' \(\alpha\) coefficients
+depended on these environmental variables. The latter set of simulated
+landscapes provide an important test of the methods' ability to
+distinguish co-occurrence patterns that were generated from pairwise
+biotic interactions from those that were generated by external forces
+like abiotic environmental filtering. This task was made especially
+difficult because---as with most analyses of presence-absence data for
+co-occurrence patterns---the inference procedure did not have access to
+any information about the environmental or spatial variables that helped
+shape the landscape (cf Connor et al. 2013). Finally, I simulated the
+population dynamics of 20 interacting species, as described in Appendix
+{[}{[}X{]}{]}
 
@@
 associations between pairs of species from simulated data (Appendix 3).
-First, I used the rosalia package (Harris 2015a) to fit Markov newtork
+First, I used the rosalia package (Harris 2015a) to fit Markov network
 models, as described above. For the analyses with 20 species, I added a
-very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
-terms with scale 2 to ensure that the model estimates were always
-finite. The bias introduced by this prior should be small: the 95\%
-credible interval on \(\beta\) only requires that one species' effect on
-the odds of observing a different species to be less than a factor of
-1500 (which is not much of a constraint). The logistic distribution was
-chosen because it is convex and has a similar shape to the Laplace
-distribution used in LASSO regularization (especially in the tails), but
-unlike the Laplace distribution it is differentiable everywhere and does
-not force any estimates to be exactly zero. To confirm that this
-procedure produced stable estimates, I compared its estimates on 50
-bootstrap replicates (Appendix 4).
+weak regularizer (equivalent to a logistic prior with scale 2) to ensure
+that the model estimates were always finite {[}{[}Appendix{]}{]}.
 
@@
 association. Next, I used partial correlations, which summarize species'
-conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
-2010). In the context of non-Gaussian data, the partial correlation can
-be thought of as a computationally efficient approximation to the full
-Markov network model (Loh and Wainwright 2013). This sort of model is
-very common for estimating relationships among genes and gene products
-(Friedman et al. 2008). Because partial correlations are undefined for
-landscapes with perfectly-correlated species pairs, I used a regularized
-estimate based on James-Stein shrinkage, as implemented in the corpcor
-package's \texttt{pcor.shrink} function with the default settings
-(Schäfer et al. 2014).
+conditional relationships. This approach is common in molecular biology
+(Friedman et al. 2008), but less so in ecology (see Albrecht and Gotelli
+2001 @faisal\_inferring\_2010 for two exceptions). In the context of
+non-Gaussian data, the partial correlation can be thought of as a
+computationally efficient approximation to the full Markov network model
+(Loh and Wainwright 2013). Because partial correlations are undefined
+for landscapes with perfectly-correlated species pairs, I used a
+regularized estimate based on James-Stein shrinkage, as implemented in
+the corpcor package's \texttt{pcor.shrink} function with the default
+settings (Schäfer et al. 2014).
 
-The third alternative, generalized linear models (GLMs), can also be
-thought of as a computationally efficient approximation to the Markov
-network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
-regularized logistic regression models (Gelman et al. 2008) for each
-species, using the other species on the landscape as predictors. To
-avoid the identifiability problems associated with directed cyclic
-graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
-within species pairs via averaging.
+The third alternative, generalized linear models (GLMs), also provide a
+computationally efficient approximation to the Markov network (Lee and
+Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
+regression models (Gelman et al. 2008) for each species, using the other
+species on the landscape as predictors. This produced two interaction
+estimates for each species pair; one for the effect of species \(i\) on
+species \(j\) and one for the reverse. These two coefficients are not
+identifiable from the data, however (Schmidt and Murphy 2012), so I used
+their average as an overall measure of the overall relationship between
+species \(i\) and species \(j\).
 
@@
 are---regardless of which species were present in a particular data set
-or how many observations were taken. Because the different methods
-mostly describe species interactions on different scales
-(e.g.~correlations versus \(Z\) scores versus regression coefficients),
-I used linear regression through the origin to rescale the different
-estimates produced by each method so that they had a consistent
-interpretation. After rescaling each method's estimates, I calculated
-squared errors between the scaled interaction estimates and ``true''
-interaction values across all the simulated data sets. These squared
-errors determined the proportion of variance explained for different
-combinations of model type and landscape size (compared with a null
-model that assumed all interaction strengths to be zero).
+or how many observations were taken. To ensure that the different test
+statistics (e.g.~correlations versus \(Z\) scores) were on a common
+scale, I rescaled them using linear regression through the origin. After
+rescaling each method's estimates, I calculated squared errors between
+the scaled interaction estimates and ``true'' interaction values across
+the simulated data sets. These squared errors determined the proportion
+of variance explained for different combinations of model type and
+landscape size (compared with a null model that assumed all interaction
+strengths to be zero).
 
@@
 
-Despite some variability across contexts (Figure 3A), the four methods
-that controlled for indirect effects clearly performed the best: the
-Markov network explained the largest portion of the variance in the
-``true'' interaction coefficients (35\% overall), followed by the
-generalized linear models (30\%), partial correlations from the raw
-presence-absence data (28\%), and partial correlations from BayesComm,
-the joint species distribution model (26\%). The benefit of choosing the
-full Markov network over the other three methods was largest on the
-smaller landscapes, which are also the ones that are most representative
-of typical analyses in this field (Gotelli and Ulrich 2009).
+Despite some variability across contexts (Figure 3), the four methods
+that controlled for indirect effects (the Markov network, the
+generalized linear models, and the two partial correlation-based
+methods) clearly outperformed those that did not. Of these, the Markov
+network consistently performed the best, especially on the smaller
+landscapes (which are also the ones that are most representative of
+typical analyses in this field; Gotelli and Ulrich 2009).
 
-The three methods that did not attempt to control for indirect
-interactions all explained less than 20\% of the variance. Of these, the
-sample correlation matrix based on the raw data performed the best
-(19\%), followed by the null model (15\%) and BayesComm's correlation
-matrix (11\%). Although these last three methods had different \(R^2\)
-values, there was a close mapping among their estimates (especially
-after controlling for the size of the simulated landscapes; Figure 3B).
-This suggests that the effect sizes from the null model (and, to a
-lesser extent, the correlation matrices from joint species distribution
-models) only contain noisy versions of the same information that could
-be obtained more easily and interpretably by calculating correlation
-coefficients between species' presence-absence vectors.
+Of the methods that did not control for indirect effects, Figure 3 shows
+that simple correlation coefficients provided a more reliable indicator
+of species' true interaction strengths than either the joint species
+distribution model (BayesComm) or the null model (Pairs). Interestingly,
+the estimates from these three approaches were tightly correlated
+{[}{[}Figure 3B{]}{]}. This suggests that the effect sizes from the null
+model (and, to a lesser extent, the correlation matrices from joint
+species distribution models) only contain noisy versions of the same
+information that could be obtained more easily and interpretably by
+calculating correlation coefficients between species' presence-absence
+vectors.
 
@@
 interval of the bootstrap distribution indicated that the Markov
-network's overall \(R^2\) value was between 14 and 18 percent higher
-than the second-most effective method (generalized linear models) and
-between 2.12 and 2.38 times higher than could be achieved by the null
-model (Pairs). Bootstrap resampling of a 200-site landscape also
-confirmed that the rosalia package's estimates of species' conditional
-relationships were robust to sampling variation for reasonably-sized
-landscapes (Appendix 4).
+network's overall \(R^2\) value was between {[}{[}14 and 18{]}{]}
+percent higher than the second-most effective method (generalized linear
+models) and between {[}{[}2.12 and 2.38{]}{]} times higher than could be
+achieved by the null model (Pairs).
 
@@
 recover species' pairwise interactions from observational data, even for
-cases where a common null modeling technique reliably fails.
-Specifically, Markov networks were successful even when direct
-interactions were largely overwhelmed by indirect effects (Figure 1).
-For cases where fitting a Markov network is computationally infeasible,
-these results also indicate that partial covariances and generalized
-linear models (the two methods that estimated conditional relationships
-rather than marginal ones) can both provide useful approximations. The
-partial correlations' success on simulated data may not carry over to
-real data sets, however; Loh and Wainwright (2013) show that the linear
-approximations can be less reliable in cases where the true interaction
-matrix contains more structure (e.g.~guilds or trophic levels).
-Similarly, the approximation involved in using separate generalized
-linear models for each species can occasionally lead to catastrophic
-overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
-For these reasons, it will usually be best to fit a Markov network
-rather than one of the alternative methods when one's computational
-resources allow it.
+cases where environmental heterogeneity and indirect interactions cause
+ecologists' typical null modeling approaches to reliably fail. For cases
+where fitting a Markov network is computationally infeasible, these
+results also indicate that partial covariances and generalized linear
+models can both provide useful approximations. The partial correlations'
+success on simulated data may not carry over to real data sets, however;
+Loh and Wainwright (2013) show that the linear approximations can be
+less reliable in cases where the true interaction matrix contains more
+structure (e.g.~guilds or trophic levels). Similarly, the approximation
+involved in using separate generalized linear models for each species
+can occasionally lead to catastrophic overfitting with small-to-moderate
+sample sizes (Lee and Hastie 2012). For these reasons, it will usually
+be best to fit a Markov network rather than one of the alternative
+methods when one's computational resources allow it.
 
@@
 the difficult and important task of incorporating whole networks of
-biotic interactions into species distribution models (Kissling et al.
-2012, Pollock et al. 2014), leading to a better understanding of the
-interplay between biotic and abiotic effects on community structure.
-There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
-Besag 1998) that would allow one species to affect the sign or strength
-of the relationship between two other species, tipping the balance
-between facilitation and exploitation (Bruno et al. 2003).
+biotic interactions into species distribution models (Pollock et al.
+2014), leading to a better understanding of the interplay between biotic
+and abiotic effects on community structure. There are even methods
+(Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that would
+allow one species to affect the sign or strength of the relationship
+between two other species, tipping the balance between facilitation and
+exploitation (Bruno et al. 2003).
 
@@
 \setlength{\parindent}{0cm}
+
+\textbf{Figure 1: A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for the tree species'
+presence with a conditional (all-else-equal) approach such as a partial
+covariance or a Markov network leads to correct identification of the
+negative shrub-shrub interaction (circled).
+
+\textbf{Figure 2: A.} A small Markov network with two species, defined
+by its \(\alpha\) and \(\beta\) values. The abiotic environment favors
+the occurrence of each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
+these two species implies that they co-occur less than expected under
+independence. \textbf{B.} These coefficients determine the probabilities
+of all four possible presence-absence combinations for Species 1 and
+Species 2. \(\alpha_1\) is added to the exponent whenever Species 1 is
+present (\(y_1 = 1\)), but not when it is absent (\(y_1 = 0\)).
+Similarly, the exponent includes \(\alpha_2\) only when species \(2\) is
+present (\(y_2 = 1\)), and \(\beta\) only when both are present
+(\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that the four
+probabilities sum to 1. In this case, \(Z\) is about 18.5. \textbf{C.}
+The expected frequencies of all possible co-occurrence patterns between
+the two species of interest, as calculated in the previous panel.
+\textbf{D.} If, as assumed in null model simulations, \(\beta_{12}\)
+equaled zero (e.g.~if the species no longer competed for the same
+resources), then the reduction in competition would allow each species
+to increase its overall occurrence rate.
+
+\textbf{Figure 3: A.} Proportion of variance in interaction coefficients
+explained by each method versus number of sampled locations. \textbf{B.}
+The \(Z\)-scores produced by the null model (``Pairs'') for each pair of
+species can be predicted using the correlation between the
+presence-absence vectors of those same species and from the number of
+sites on the landscape.
 
@@
 
-Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
-Selected Problems in Ecology. Entropy 15:5237--5276.
-
 Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
@@
 
-Gotelli, N. J., and G. L. Entsminger. 2003. Swap algorithms in null
-model analysis. Ecology:532--535.
-
 Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
@@
 Probability 2:969--988.
-
-Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
-J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
-Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
-Towards novel approaches to modelling biotic interactions in
-multispecies assemblages at large spatial extents. Journal of
-Biogeography 39:2163--2178.
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
-
-\textbf{Figure captions}
-
-\setlength{\parskip}{3pt} \setlength{\parindent}{0em}
-\setlength{\leftskip}{0em}
-
-\textbf{Figure 1.} \textbf{A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, but more
-weakly (circled coefficient). \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional method such as a partial covariance
-or a Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled).
-
-\textbf{Figure 2.} \textbf{A.} A small Markov network with two species.
-The abiotic environment favors the occurrence of both species
-(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
-negative \(\beta\) coefficient linking these two species implies that
-they co-occur less than expected under independence. \textbf{B.}
-Relative probabilities of all four possible presence-absence
-combinations for Species 1 and Species 2. The exponent includes
-\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
-it is absent (\(y_1 = 0\)). Similarly, the exponent includes
-\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
-\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
-constant \(Z\), ensures that the four relative probabilities sum to 1.
-In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
-the species no longer competed for the same resources), then the
-reduction in competition would allow each species to increase its
-occurrence rate and the co-occurrence deficit would be eliminated.
-
-\textbf{Figure 3.} \textbf{A.} Proportion of variance in interaction
-coefficients explained by each method versus number of sampled
-locations. \textbf{B.} The \(Z\)-scores produced by the null model
-(``Pairs'') for each pair of species can be predicted using the
-correlation between the presence-absence vectors of those same species
-and from the number of sites on the landscape.
-
 \end{flushleft}

### 2015-10-25 Harris_2015/recovery.Rmd
various

--- 2015-08-04 Harris_2015/recovery.Rmd
@@
 
-```{r, cache = TRUE, message = FALSE}
+```{r, cache = FALSE, message = FALSE, eval = TRUE}
+
+mc.cores = 2
+
 library(dplyr)        # For manipulating data structures
@@
 library(BayesComm)    # For joint species distribution modeling
-library(RColorBrewer) # For color palette
+library(parallel)     # for mclapply
 set.seed(1)
@@
 
-```{r, cache = TRUE}
-pairs_txt = readLines("fakedata/Pairs.txt")
-
-# Find areas of the data file that correspond
-# to species pairs' results
-beginnings = grep("Sp1", pairs_txt) + 1
-ends = c(
-  grep("^[^ ]", pairs_txt)[-1],
-  length(pairs_txt)
-) - 1
-
-# The above code fails on the very last line of the file
-ends[length(ends)] = ends[length(ends)] + 1
-```
-
 A function to import the data file and run each method on it:
 
-```{r, cache = TRUE}
-fit_all = function(filename){
+```{r, cache = FALSE, eval = TRUE}
+fit_all = function(identifier){
   ######## Import ########
 
-  # Multiplying by one is necessary to prevent silly errors
-  # regarding TRUE/FALSE versus 1/0
-  raw_obs = readRDS(filename)[["observed"]] * 1
+  data_filename = paste0("fakedata/matrices/", identifier, ".csv")
+  truth_filename = paste0("fakedata/truths/", identifier, ".txt")
+
+  # first column is row numbers; drop it
+  raw_obs = as.matrix(read.csv(data_filename)[ , -1])
 
@@
   x = raw_obs[ , species_is_variable]
-  truth = readRDS(filename)[["truth"]][pair_is_variable[upper.tri(pair_is_variable)]]
+  truth = unlist(read.table(truth_filename))[pair_is_variable[upper.tri(pair_is_variable)]]
 
-  splitname = strsplit(filename, "/|-|\\.")[[1]]
-  n_sites = as.integer(splitname[[3]])
-  rep = as.integer(splitname[[4]])
+  splitname = strsplit(identifier, "/|-|\\.")[[1]]
+  n_spp = as.integer(splitname[[1]]) # simulated species, NOT present species
+  n_sites = as.integer(splitname[[2]])
+  rep_name = splitname[[3]]
+
+  # Species IDs
+  sp1 = combn(colnames(x), 2)[1, ]
+  sp2 = combn(colnames(x), 2)[2, ]
 
@@
   corr = cor(x)
-
   ######## GLM ########
@@
   ######## Markov network ########
-  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2))
+  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2), hessian = TRUE)
+
+  rosie_point = rosie$beta[upper.tri(rosie$beta)]
+  rosie_se = sqrt(diag(solve(rosie$opt$hessian)))[-(1:sum(species_is_variable))]
 
@@
 
-  `partial BayesComm` = 0
-
-  for(i in 1:nrow(bc$trace$R)){
+  bc_pcors = sapply(
+    1:nrow(bc$trace$R), 
+    function(i){
       Sigma = matrix(0, nrow = ncol(x), ncol = ncol(x))
@@
       diag(Sigma) <- 1  # Diagonal equals 1 in multivariate probit model
-
-    `partial BayesComm` = `partial BayesComm` + cor2pcor(Sigma) / nrow(bc$trace$R)
+      pcor = cor2pcor(Sigma)
+      pcor[upper.tri(pcor)]
     }
-
-  ######## Pairs ########
-  # Find the line where the current data set is mentioned in
-  # pairs.txt
-  filename_line = grep(
-    paste0(
-      gsub("fakedata/(.*)\\.rds", "\\1", filename),
-      "-"
-    ),
-    pairs_txt
   )
 
-  # Which chunk of the data file corresponds to this file?
-  chunk = min(which(beginnings > filename_line))
-
-  # Split the chunk on whitespace.
-  splitted = strsplit(pairs_txt[beginnings[chunk]:ends[chunk]], " +")
-
-  # Pull out the species numbers and their Z-scores
-  pairs_results = lapply(
-    splitted,
-    function(x){
-      spp = sort(as.integer(x[3:4]))
-      data.frame(
-        sp1 = spp[1], 
-        sp2 = spp[2], 
-        z = as.numeric(x[14])
+  bind_rows(
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "Markov network",
+      truth = truth,
+      estimate = rosie_point, 
+      lower = qnorm(.025, rosie_point, rosie_se), 
+      upper = qnorm(.975, rosie_point, rosie_se)
+    ),
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "GLM",
+      truth = truth,
+      estimate = coef_matrix[upper.tri(coef_matrix)], 
+      lower = NA, 
+      upper = NA
+    ),
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "correlation",
+      truth = truth,
+      estimate = corr[upper.tri(corr)], 
+      lower = NA, 
+      upper = NA
+    ),
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "partial correlation",
+      truth = truth,
+      estimate = p_corr[upper.tri(p_corr)], 
+      lower = NA, 
+      upper = NA
+    ),
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "partial BayesComm",
+      truth = truth,
+      estimate = rowMeans(bc_pcors),
+      lower = apply(bc_pcors, 1, quantile, .025),
+      upper = apply(bc_pcors, 1, quantile, .975)
+    ),
+    data_frame(
+      rep_name = rep_name,
+      n_sites = n_sites,
+      sp1 = sp1,
+      sp2 = sp2,
+      method = "BayesComm",
+      truth = truth,
+      estimate = colMeans(bc$trace$R),
+      lower = apply(bc$trace$R, 2, quantile, .025),
+      upper = apply(bc$trace$R, 2, quantile, .975)
     )
-    }
-  ) %>% 
-    bind_rows %>%
-    mutate(spp = paste(sp1, sp2, sep = "-"))
-
-  # Re-order the pairs_results to match the other methods
-  m = matrix(NA, ncol(x), ncol(x))
-  new_order = match(
-    paste(row(m)[upper.tri(m)], col(m)[upper.tri(m)], sep = "-"),
-    pairs_results$spp
-  )
-  ordered_pairs_results = pairs_results[new_order, ]
-
-  ######## Output ########
-  data.frame(
-    truth = truth, 
-    n_sites = n_sites,
-    rep = rep,
-    sp1 = ordered_pairs_results$sp1,
-    sp2 = ordered_pairs_results$sp2,
-    `partial correlation` = p_corr[upper.tri(p_corr)],
-    correlation = corr[upper.tri(corr)],
-    `Markov network` = rosie$beta[upper.tri(rosie$beta)],
-    GLM = coef_matrix[upper.tri(coef_matrix)],
-    `BayesComm` = colMeans(bc$trace$R),
-    `partial BayesComm` = `partial BayesComm`[upper.tri(`partial BayesComm`)],
-    Pairs = ordered_pairs_results$z
   )
@@
 
-```{r, cache = TRUE}
+```{r, cache = FALSE, eval = TRUE}
 
-# Find all the .rds files in the fakedata folder
-files = dir("fakedata", pattern = "\\.rds$", full.names = TRUE)
+# Find all the csv files in the fakedata/matrices folder,
+# then drop .csv
+identifiers = dir("fakedata/matrices", pattern = "\\.csv$") %>%
+  gsub("\\.csv$", "", .)
 
-# Run all the analyses on all the files
-z = lapply(files, fit_all) %>% bind_rows %>% as.data.frame
-
-# Fix a formatting issue with the column names
-colnames(z) = gsub("\\.", " ", colnames(z))
+mclapply(identifiers, fit_all, mc.cores = mc.cores, mc.preschedule = FALSE, mc.silent = TRUE) %>% 
+  bind_rows() %>% 
+  write.csv(file = "estimates.csv")
 ```
-
-Summarize the results of all 7 methods across the simulated landscapes:
-
-```{r, cache = TRUE}
-
-# Calculate residuals for each method
-resids = sapply(
-  colnames(z)[-(1:5)],
-  function(i){resid(lm(z$truth ~ z[,i] + 0))}
-)
-
-sizes = sort(unique(z$n_sites))
-
-# Compute proportion of variance explained, compared with a null that
-# assumes all species interactions are 0.
-results = as.data.frame(
-  t(
-    sapply(
-      sizes,
-      function(n){
-        total_ss = mean(resid(lm(truth ~ 0, data = z))[z$n_sites == n]^2)
-        1 - colMeans(resids[z$n_sites == n , ]^2) / total_ss
-      }
-    )
-  )
-)
-# Sort the results in decreasing order
-results = results[order(colMeans(results), decreasing = TRUE)]
-```
-
-Plot the results:
-
-```{r, cache = TRUE, message=FALSE}
-colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]
-
-# Set up the plotting canvas
-pdf("manuscript-materials/figures/performance.pdf", height = 8, width = 5)
-par(mfrow = c(2, 1))
-par(mar = c(5, 4, 3, 0) + .1)
-
-# Plot the model performance
-matplot(
-  sizes, 
-  results, 
-  type = "o", 
-  ylab = "", 
-  xlim = c(25 * .9, 1600 * 8),
-  ylim = c(0, .5 + 1E-9),
-  pch = c(15, 1, 17, 0, 16, 2, 3), 
-  lty = 1,
-  log = "x",
-  xlab = "",
-  axes = FALSE,
-  xaxs = "i",
-  yaxs = "i",
-  col = colors,
-  lwd = 1.25,
-  cex = 0.9,
-  bty = "l"
-)
-axis(1, c(1, 25, 200, 1600))
-axis(2, seq(0, 1, .1), las = 1)
-
-mtext(expression(R^2), side = 2, line = 3, las = 1)
-mtext("Number of sites (log scale)", side = 1, line = 2.25, at = 200)
-mtext("A.", line = 1.25, at = 25, cex = 1.5)
-
-# Determine how high on the y-axis each method label should be so they
-# don't overlap.
-heights = results[nrow(results), ]
-
-heights$Pairs = heights$Pairs - .01
-heights$correlation = heights$correlation + .01
-
-heights$`partial correlation` = heights$`partial correlation` + .01
-heights$`partial BayesComm` = heights$`partial BayesComm` - .01
-
-heights$GLM = heights$GLM - .01
-
-text(1625, heights, colnames(results), pos = 4, cex = 0.75, col = colors)
-
-######
-
-# silly code to make the next graph line up prettily with the previous one
-full_range = log10(c(25 * .9, 1600 * 8))
-base_range = log10(c(25, 1600))
-
-base_scaled = 2 * base_range / (base_range[2] - base_range[1])
-full_scaled = 2 * full_range / (base_range[2] - base_range[1])
-
-# New colors for plot B
-colors2 = brewer.pal(10, "Paired")[c(2, 4, 10)]
-plot(
-  z$correlation, 
-  z$Pairs, 
-  col = colors2[factor(z$n_sites)],
-  las = 1,
-  xlab = "",
-  ylab = expression(italic(Z)-score),
-  bty = "l",
-  xaxs = "i",
-  axes = FALSE,
-  xlim = full_scaled - base_scaled[2] + 1
-)
-axis(1, seq(-2, 1, .5))
-axis(2, seq(-50, 50, 10), las = 1)
-mtext("Correlation coefficient", side = 1, line = 2.25, at = 0)
-mtext("B.", line = 1.25, at = -1, cex = 1.5)
-
-legend(
-  x = 1.05, 
-  y = max(z$Pairs),
-  pch = 1,
-  lty = 0,
-  lwd = 2,
-  col = colors2[1:length(unique(z$n_sites))], 
-  legend = levels(factor(z$n_sites)),
-  title = "Number of sites",
-  bty = "n",
-  cex = 0.75,
-  pt.cex = 1
-)
-segments(-1000, 0, 1.04, 0, col = "#00000025", lwd = 1.25)
-segments(0, -1000, 0, 1000, col = "#00000025", lwd = 1.25)
-
-dev.off()
-```
-
-![Figure 3](manuscript-materials/figures/performance.pdf)
-
-Summarize the results:
-
-```{r, cache = TRUE}
-
-# R-squareds computed across *all* landscape types
-total_ss = mean(resid(lm(truth ~ 0, data = z))^2)
-round(100 * sort(1 - colMeans(resids^2) / total_ss), 1)
-
-# Rank correlations among the methods that estimate marginal relationships
-round(cor(z[ , c(7, 10, 12)], method = "spearman")[,1], 2)
-```
-
-Plot the true interactions versus estimated interactions for two methods:
-
-```{r, cache = TRUE}
-library(ggplot2)
-
-ggplot(z[z$n_sites >0, ], aes(x = `Markov network`, y = truth)) + 
-  stat_binhex(bins = 100) + 
-  xlab("Estimated coefficient value") + 
-  ylab("\"True\" coefficient value") + 
-  stat_hline(yintercept = 0, size = 1/8) + 
-  stat_vline(xintercept = 0, size = 1/8) + 
-  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
-  theme_bw() + 
-  ggtitle("Markov network") + 
-  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$`Markov network` + 0)))
-
-ggplot(z[z$n_sites >0, ], aes(x = `Pairs`, y = truth)) + 
-  stat_binhex(bins = 100) + 
-  xlab("Estimated coefficient value") + 
-  ylab("\"True\" coefficient value") + 
-  stat_hline(yintercept = 0, size = 1/8) + 
-  stat_vline(xintercept = 0, size = 1/8) + 
-  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
-  theme_bw() + 
-  ggtitle("Null model (\"Pairs\")") + 
-  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$Pairs + 0)))
-```
-
-Bootstrap resample from each of the three landscape size classes, generating
-new sets of 150 landscapes. For each one, calculate the proportion of variance
-explained (compared to a null baseline that assumes all species pairs' 
-interaction strengths are zero).
-
-```{r, cache = TRUE}
-rep = 12
-n = 200
-boots = replicate(500,
-          {
-            boot = lapply(
-              sizes, 
-              function(n){lapply(
-                sample.int(max(z$rep), replace = TRUE),
-                function(rep){
-                  z[z$rep == rep & z$n_sites == n, ]
-                }
-              ) %>% bind_rows}
-            ) %>% bind_rows
-
-            resids = sapply(
-              colnames(boot)[-(1:5)],
-              function(i){resid(lm(boot$truth ~ boot[[i]] + 0))}
-            )
-            total_ss = mean(resid(lm(truth ~ 0, data = boot))^2)
-            1 - colMeans(resids^2) / total_ss
-          }
-)
-```
-
-Summarize the bootstrap results:
-
-```{r, cache = TRUE}
-for(compared in c("partial correlation", "correlation", "GLM", "BayesComm", "partial BayesComm", "Pairs")){
-  CI = round(
-    quantile(
-      boots["Markov network", ] / boots[compared, ], 
-      c(.025, .975)), 
-    2
-  )
-
-  message("R-squared is " , CI[1], "-", CI[2], " times higher than from ", compared, " (95% bootstrap CI)")
-}
-```

### 2015-10-25 Harris_2015/recovery.tex
various

--- 2015-08-05 Harris_2015/recovery.tex
@@
 \NormalTok{z =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(files, fit_all) %>%}\StringTok{ }\NormalTok{bind_rows %>%}\StringTok{ }\NormalTok{as.data.frame}
-\end{Highlighting}
-\end{Shaded}
 
-\begin{Shaded}
-\begin{Highlighting}[]
 \CommentTok{# Fix a formatting issue with the column names}
@@
 
-\begin{verbatim}
-## pdf 
-##   2
-\end{verbatim}
-
 \includegraphics{manuscript-materials/figures/performance.pdf}
@@
 \KeywordTok{round}\NormalTok{(}\DecValTok{100} \NormalTok{*}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss), }\DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
-\begin{verbatim}
-##           BayesComm               Pairs         correlation 
-##                11.6                15.4                18.6 
-##   partial BayesComm partial correlation                 GLM 
-##                26.1                27.5                29.9 
-##      Markov network 
-##                34.6
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \CommentTok{# Rank correlations among the methods that estimate marginal relationships}
@@
 \end{Shaded}
-
-\begin{verbatim}
-## correlation   BayesComm       Pairs 
-##        1.00        0.85       -0.92
-\end{verbatim}
 
@@
 \StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
-\end{Highlighting}
-\end{Shaded}
 
-\includegraphics{recovery_files/figure-latex/unnamed-chunk-8-1.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
+\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{-z$Pairs}
+\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{-Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
 \StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
@@
 \StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$Pairs +}\StringTok{ }\DecValTok{0}\NormalTok{)))}
+\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
 \end{Highlighting}
 \end{Shaded}
-
-\includegraphics{recovery_files/figure-latex/unnamed-chunk-8-2.pdf}
 
@@
 
-\begin{verbatim}
-## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
-## R-squared is 1.79-1.95 times higher than from correlation (95% bootstrap CI)
-## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
-## R-squared is 2.72-3.28 times higher than from BayesComm (95% bootstrap CI)
-## R-squared is 1.29-1.37 times higher than from partial BayesComm (95% bootstrap CI)
-## R-squared is 2.12-2.37 times higher than from Pairs (95% bootstrap CI)
-\end{verbatim}
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{save}\NormalTok{(results, sizes, }\DataTypeTok{file =} \StringTok{"graph_data.Rdata"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
 

### 2015-10-25 Harris_2015/show-results.R

@@ -1,0 +1,90 @@
+library(dplyr)
+library(mgcv)
+library(ggplot2)
+
+
+
+x = read.csv("estimates.csv", stringsAsFactors = FALSE)
+x$dynamic_simulation = grepl("pop", x$rep_name)
+x$simulation_type = gsub("[0-9]", "", x$rep_name)
+
+resids = function(data){
+  resid(lm(truth ~ estimate + 0, data = data))
+}
+
+result_summary = x %>% 
+  group_by(method, simulation_type) %>%
+  do(data.frame(., resids = resids(.))) %>%
+  ungroup %>%
+  group_by(method, simulation_type, n_sites) %>%
+  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))
+
+result_summary$method = reorder(result_summary$method, -result_summary$r2)
+
+result_summary$simulation_type = reorder(result_summary$simulation_type, -result_summary$r2)
+
+
+result_summary %>% 
+  filter(simulation_type == "no_env") %>%
+  group_by(method) %>% 
+  summarise(round(mean(r2), 3))
+
+result_summary %>% 
+  filter(simulation_type == "env") %>%
+  group_by(method) %>% 
+  summarise(round(mean(r2), 3))
+
+result_summary %>% 
+  filter(simulation_type == "pop") %>%
+  group_by(method) %>% 
+  summarise(round(mean(r2), 3))
+
+
+
+pdf("manuscript-materials/figures/performance.pdf", width = 7.5, height = 3)
+ggplot(result_summary, aes(x = n_sites, y = r2, col = method)) + 
+  facet_grid(~simulation_type) + 
+  geom_line(size = .5) + 
+  geom_point(size = 2, fill = "white") + 
+  geom_hline(yintercept = 0, size = 1/2) +
+  geom_vline(xintercept = 0, size = 1) + 
+  #scale_shape_manual(values = c(21, 25, 15, 18)) + 
+  coord_cartesian(ylim = c(-.01, 0.76)) + 
+  ylab(expression(R^2)) + 
+  xlab("Number of sites (log scale)") + 
+  scale_x_log10(breaks = unique(x$n_sites), limits = range(x$n_sites)) +
+  theme_bw(base_size = 11) + 
+  theme(panel.margin = grid::unit(1.25, "lines")) + 
+  theme(panel.border = element_blank(), axis.line = element_blank()) + 
+  theme(
[…]

### 2015-10-26 Harris_2015/manuscript-materials/draft.Rmd
moar stuff

@@
 
-\setlength{\parskip}{3pt}
+\setlength{\parskip}{1pt}
 
@@
 Interactions in these models can also be estimated from observed co-occurrence
-rates via maximum likelihood. Using simulated landscapes with known pairwise
-interaction strengths, I evaluated Markov networks and six existing approaches.
-The Markov networks consistently outperformed other methods, correctly isolating
-direct interactions between species pairs even when indirect interactions
-largely overpowered them. Two computationally efficient approximations, based on
-controlling for indirect effects with linear or generalized linear models,
-also performed well. Indirect effects reliably caused a common null modeling
-approach to produce incorrect inferences, however.
+rates via maximum likelihood, controlling for indirect effects. Using simulated
+landscapes with known pairwise interaction strengths, I evaluated Markov
+networks and six existing approaches. The Markov networks consistently
+outperformed other methods, correctly isolating direct interactions between
+species pairs even when indirect interactions or abiotic factors largely
+overpowered them. Two computationally efficient approximations, based on
+controlling for indirect effects with linear or generalized linear models, also
+performed well. Indirect effects reliably caused a common null modeling approach
+to produce incorrect inferences, however.
 
@@
 To the extent that nontrophic species interactions (such as competition) affect
-community assembly, ecologists might expect to find signatures of these interactions in
-species composition data [@macarthur_population_1958; @diamond_island_1975].
-Despite decades of work and several major controversies, however
-[@lewin_santa_1983; @strong_ecological_1984;
-@connor_checkered_2013], existing methods for detecting competition's effects
-on community structure are unreliable [@gotelli_empirical_2009]. In particular,
-species' effects on one another can become lost in the complex web of
-indirect effects. For example, the competitive interaction between the two shrub
-species in Figure 1A can become obscured by their shared tendency to occur in
-unshaded areas (Figure 1B). While ecologists have long known that indirect
-effects can overwhelm direct ones at the landscape level
-[@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our
-methods for drawing inferences from observational data do not control for these effects
-[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
-@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
-indirect interactions like those in Figure 1 are generally important
-[@dodson_complementary_1970], existing methods will thus not generally provide
-much evidence regarding species' direct effects on one another. The goal of this
-paper is to resolve this long-standing problem.
+community assembly, ecologists might expect to find signatures of these
+interactions in species composition data [@macarthur_population_1958;
+@diamond_island_1975]. Despite decades of work and several major controversies,
+however [@lewin_santa_1983; @strong_ecological_1984; @connor_checkered_2013],
+existing methods for detecting competition's effects on community structure are
+unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
+another can become lost in the complex web of indirect effects. For example, the
+competitive interaction between the two shrub species in Figure 1A can become
+obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
+ecologists have long known that indirect effects can overwhelm direct ones at
+the landscape level [@dodson_complementary_1970; @levine_competitive_1976], the
+vast majority of our methods for drawing inferences from observational data do
+not control for these effects [e.g. @diamond_island_1975;
+@strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
+@pollock_understanding_2014]. To the extent that indirect interactions like
+those in Figure 1 are generally important [@dodson_complementary_1970], existing
+methods will thus not provide much evidence regarding species interactions.
 
-While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
-level (as most methods assume), it nevertheless does leave a signal in
-the data (Figure 1C). For example, after partitioning the data set into shaded
-sites and unshaded sites, there will be co-occurrence deficits in each subset
-that might not be apparent at the whole-landscape level. More generally,
-controlling for other species in the network will often be important for obtaining
-reliable estimates of direct (conditional, or all-else-equal) effects.  This kind of precision is
-difficult to obtain from null models, which only simulate the most extreme
-possible scenario: that *all* direct and indirect interactions are exactly zero.
-Nevertheless, null models have dominated this  field for more than three decades
-[@strong_ecological_1984; @gotelli_empirical_2009].
+While competition doesn't reliably reduce co-occurrence rates at the
+whole-landscape level (as most methods assume), it nevertheless does leave a
+signal in the data (Figure 1C). For example, after partitioning the data set
+into shaded sites and unshaded sites, there will be co-occurrence deficits in
+each subset that might not be apparent at the whole-landscape level. More
+generally, controlling for other species in the network will often be important
+for obtaining reliable estimates of direct (conditional, or all-else-equal)
+effects.  This kind of precision is difficult to obtain from null models, which
+only simulate the most extreme possible scenario: that *all* direct and indirect
+interactions are exactly zero. Nevertheless, null models have dominated this
+field for more than three decades [@strong_ecological_1984;
+@gotelli_empirical_2009].
 
 Following @azaele_inferring_2010, this paper shows that Markov networks
-[undirected graphical models also known as
-Markov random fields; @murphy_machine_2012] can provide a framework for
-understanding the landscape-level consequences of pairwise species interactions,
+[undirected graphical models also known as Markov random fields;
+@murphy_machine_2012] can provide a framework for understanding the
+landscape-level consequences of pairwise species interactions,
 and for detecting them from observed presence-absence matrices. Markov networks
@@
 
-#### Markov networks.
+**Markov networks.**
 Markov networks provide a framework for translating back and forth between the
@@
 
-$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),$
+$\displaystyle{p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),}$
 
@@
 
-\noindent where the second sum is over all $\binom{n}{2}$ species pairs. In this model,  $\alpha_{i}$ is
-an intercept term determining the amount that the presence
+\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$ species. In this
+model, $\alpha_{i}$ is an intercept term determining the amount that the presence
 of species $i$ contributes to the log-probability of $\vec{y}$; it directly
@@
 observing one species are $e^2$ times higher in sites where its partner is
-present than in comparable sites where its partner is absent. Because the
-relative probability of a presence-absence vector increases when
+present than in comparable sites where its partner is absent.
+
+Because the relative probability of a presence-absence vector increases when
 positively-associated species co-occur and decreases when negatively-associated
@@
 negatively-associated species (exactly as an ecologist might expect).
-
 Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
 competitor is associated with release from another competitor), then species'
-overall co-occurrence patterns can be more complicated. Determining the
-marginal relationships between species from their conditional interactions
-entails summing over the different possible assemblages (Figure 2B). This
-becomes intractable when the number of possible assemblages is large, though
-several methods beyond the scope of this paper can be employed to keep the
-calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
-Alternatively, as noted below, some common linear and generalized linear methods
-can also be used as computationally efficient approximations to the full network
-[@lee_learning_2012; @loh_structure_2013].
+overall co-occurrence patterns can be more complicated, and may require summing
+over the different possible assemblages (Figure 2B) to determine the overall
+outcomes.
 
-#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
+\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
 In the previous section, the values of $\alpha$ and $\beta$ were known and the
@@
 that would be expected to produce exactly the observed co-occurrence frequencies
-on average [i.e. maximizing the likelihood ensures that the sufficient
+on average [i.e., maximizing the likelihood ensures that the sufficient
 statistics of the model distribution match the sufficient statistics of the data;
@@
 R programming language [@r_core_team_r_2015] to optimize the Markov network
-parameters. The package was named after Santa Rosalia, the patron saint of
-biodiversity, whose mythos played an important rhetorical role in the null model
-debates of the 1970's and 1980's [@lewin_santa_1983].
+parameters.
 
-#### Simulated landscapes.
+\noindent \textbf{Simulated landscapes.}
 I simulated several sets of landscapes using known parameters as a test bed for
@@
 combination of size and model). As described in Appendix [[X]], I randomly drew
-the “true” coefficient values for each replicate landscape in these simulations
+the “true” coefficient values for each replicate landscape
 so that most species pairs interacted negligibly, a few pairs interacted very
@@
 across the simulated landscapes, and simulated a new set of co-occurrence data
-so that species' $\alpha$ coefficients depended on these environmental variables.
+so that species' $\alpha$ coefficients depended on the local environment.
 The latter set of simulated landscapes provide an important test of the methods’
@@
 the environmental or spatial variables that helped shape the landscape [cf
-@connor_checkered_2013].  Finally, I simulated the population dynamics of 20
-interacting species, as described in Appendix [[X]]
+@connor_checkered_2013].  I generated the final set of landscapes by simulating
+the population dynamics of 20 interacting species, as described in Appendix [[X]].
 
-#### Recovering species interactions from simulated data.
-
+\noindent \textbf{Recovering species interactions from simulated data.}
 I compared seven techniques for determining the sign and strength of the
@@
 which summarize species' conditional relationships. This approach is
-common in molecular biology [@friedman_sparse_2008], but less so in ecology
-[see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions].
+common in molecular biology [@friedman_sparse_2008], but is rare in ecology
+(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
 In the context of non-Gaussian data, the partial correlation can be thought of
@@
 
-#### Evaluating model performance.
+\noindent \textbf{Evaluating model performance.}
 For the simulated landscapes based on Figure 1, I assessed whether each method's
@@
 rescaling each method’s estimates, I calculated squared errors between the scaled
-interaction estimates and “true” interaction values across the simulated data
+interaction estimates and "true" interaction values across the simulated data
 sets. These squared errors determined the proportion of variance explained
@@
 
-#### Three species.
+\noindent \textbf{Three species.}
 As shown in Figure 1, the marginal relationship between the two shrub species
@@
 
-#### Twenty species.
+\noindent
+\textbf{Twenty species.}
 Despite some variability across contexts (Figure 3), the four methods that
 controlled for indirect effects (the Markov network, the generalized linear
-models, and the two partial correlation-based methods) clearly outperformed those
+models, and the two partial correlation-based methods) outperformed those
 that did not.  Of these, the Markov network consistently performed the best,
-especially on the smaller landscapes [which are also the ones that are most
+especially on the smaller landscapes [which are also the most
 representative of typical analyses in this field; @gotelli_empirical_2009].
+As shown in Appendix [[X]], the standard errors associated with the estimates
+in Figure 3 are all small (less than 0.01), so differences across methods should
+not be attributed to sampling error.
 
@@
 true interaction strengths than either the joint species distribution model
-(BayesComm) or the null model (Pairs). Interestingly, the estimates from these
-three approaches were tightly correlated [[Figure 3B]]. This suggests that the
+(BayesComm) or the null model (Pairs). The estimates from these
+three approaches were tightly correlated (Figure 3B), suggesting that the
 effect sizes from the null model (and, to a lesser extent, the correlation
@@
 
-Bootstrap resampling indicated that the above ranking of the different methods
-was robust (Appendix 3). In particular, the 95% confidence interval of the
-bootstrap distribution indicated that the Markov network's overall $R^2$ value
-was between [[14 and 18]] percent higher than the second-most effective
-method (generalized linear models) and between [[2.12 and 2.38]] times higher than
-could be achieved by the null model (Pairs).
+Whereas Figure 3 summarizes the accuracy of each model's point estimates, Figure
+4 compares the inferential value of the Markov network versus the null model.
+When the "true" $\beta$ coefficient is near zero, the Markov network's 95%
+confidence intervals usually include zero (but not quite 95% of the time), and
+it is rare for the model to confidently attribute the wrong sign to a
+coefficient (Figure 4A). The null model, by contrast, typically rejects the null
+hypothesis at the $p=0.05$ level regardless of the underlying parameters, and
+the sign of its $Z$-scores is not a reliable indicator of the interaction's
+actual sign.
 
@@
 environmental heterogeneity and indirect interactions cause ecologists' typical
-null modeling approaches to reliably fail. For cases where fitting a Markov network is
-computationally infeasible, these results also indicate that partial covariances
-and generalized linear models can both provide useful approximations.
+null modeling approaches to reliably fail. For cases where fitting a Markov
+network is computationally infeasible, these results also indicate that partial
+covariances and generalized linear models can both provide useful approximations.
 The partial correlations' success on simulated data may not carry over to real
@@
 studied for nearly a century [@cipra_introduction_1987], and the models'
-properties, capabilities, and limits are well-understood in a
-huge range of applications. Using the same framework for species interactions
-would thus allow ecologists to tap into an enormous set of existing
-discoveries and techniques for dealing with indirect effects, stability, and
-alternative stable states. Numerous other extensions are possible: for example,
-the states of the interaction network can be modeled as a function of
-the local abiotic environment [@lee_learning_2012], which would provide a
-rigorous and straightforward approach to the difficult and important task of
-incorporating whole networks of biotic interactions into species distribution
-models [@pollock_understanding_2014], leading to a
+properties, capabilities, and limits are well-understood. Using the same
+framework for species interactions would thus allow ecologists to tap into an
+enormous set of existing discoveries and techniques for dealing with indirect
+effects, stability, and alternative stable states. Numerous other extensions are
+possible: for example, the states of the interaction network can be modeled as a
+function of the local abiotic environment [@lee_learning_2012], which would
+provide a rigorous and straightforward approach to the difficult and important
+task of incorporating whole networks of biotic interactions into species
+distribution models [@pollock_understanding_2014], leading to a
 better understanding of the interplay between biotic and abiotic effects on
-community structure. There are even methods [@whittam_species_1981;
-@tjelmeland_markov_1998] that would allow one species to affect the sign or
-strength of the relationship between two other species, tipping the balance
-between facilitation and exploitation [@bruno_inclusion_2003].
+community structure. Alternatively, models could allow one species to affect the
+interactions between two other species [@tjelmeland_markov_1998], tipping the
+balance between facilitation and exploitation [@bruno_inclusion_2003].
 
@@
 rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
-test statistic is effectively just a correlation coefficient (Figure 3B).
-Simultaneous estimation of multiple ecological parameters seems like a much more
-promising approach: to the extent that the models' relative performance on real
-data sets is similar to the range of results shown in Figure 3A, scientists in
-this field could often double their explanatory power by switching from null
-models to Markov networks (or increase it substantially with linear or
-generalized linear approximations). Regardless of the methods ecologists
-ultimately choose, controlling for indirect effects could clearly improve our
-understanding of species' direct effects on one another and on community
-structure.
+test statistic is effectively just a correlation coefficient [[Figure 3B]].
+When a non-null backdrop is not controlled for, Type I error rates can
+skyrocket, the apparent sign of the interaction can change, and null models
+can lead to reliably incorrect inferences (Figure 1, Figure 4).
 
-#### Acknowledgements:
+Controlling for indirect effects via simultaneous estimation of multiple
+ecological parameters seems like a much more promising approach: to the extent
+that the models' relative performance on real data sets is similar to the range
+of results shown in Figure 3A, scientists in this field could more than double
+their explanatory power by switching from null models to Markov networks (or
+increase it nearly as much with linear or generalized linear approximations).
+Regardless of the methods ecologists ultimately choose, controlling for indirect
+effects could clearly improve our understanding of species' direct effects on
+one another and on community structure.
+
+\noindent \textbf{Acknowledgements:}
 This research was funded by a Graduate Research Fellowship from the US
@@
 
-**Figure 1: A.** A small network of three competing species. The tree (top) tends not to
+\noindent \textbf{Figure 1: A.} A small network of three competing species. The tree (top) tends not to
 co-occur with either of the two shrub species, as indicated by the strongly
@@
 
-**Figure 2: A.** A small Markov network with two species, defined by its $\alpha$ and
+\noindent \textbf{Figure 2: A.} A small Markov network with two species, defined by its $\alpha$ and
 $\beta$ values.  The abiotic environment favors the occurrence of each species
@@
 
-**Figure 3: A.** Proportion of variance in interaction coefficients explained by each
-method versus number of sampled locations.
+\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
+coefficients explained by each method versus number of sampled locations.
 **B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
@@
 
+\noindent \textbf{Figure 4: A.} Inferences from the Markov network's 95%
+confidence intervals on the "no-environment" simulated landscapes, as a function
+of the "true" coefficient values. **B.** Inferences from the $Z$-statistics
+estimated by the "Pairs" software for null modeling.
+
 **References:**
 
-\setlength{\parskip}{0pt}
 \setlength{\parindent}{-1em}

### 2015-10-26 Harris_2015/manuscript-materials/draft.tex
moar stuff

@@
 \usepackage{ifthen}
-\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+\geometry{verbose,letterpaper,tmargin=2.35cm, bmargin=2.35cm,lmargin=2.44cm,rmargin=2.44cm}
 \ifxetex
@@
 interaction matrix. Interactions in these models can also be estimated
-from observed co-occurrence rates via maximum likelihood. Using
-simulated landscapes with known pairwise interaction strengths, I
-evaluated Markov networks and six existing approaches. The Markov
-networks consistently outperformed other methods, correctly isolating
-direct interactions between species pairs even when indirect
-interactions largely overpowered them. Two computationally efficient
-approximations, based on controlling for indirect effects with linear or
-generalized linear models, also performed well. Indirect effects
-reliably caused a common null modeling approach to produce incorrect
-inferences, however.
+from observed co-occurrence rates via maximum likelihood, controlling
+for indirect effects. Using simulated landscapes with known pairwise
+interaction strengths, I evaluated Markov networks and six existing
+approaches. The Markov networks consistently outperformed other methods,
+correctly isolating direct interactions between species pairs even when
+indirect interactions or abiotic factors largely overpowered them. Two
+computationally efficient approximations, based on controlling for
+indirect effects with linear or generalized linear models, also
+performed well. Indirect effects reliably caused a common null modeling
+approach to produce incorrect inferences, however.
 
@@
 like those in Figure 1 are generally important (Dodson 1970), existing
-methods will thus not generally provide much evidence regarding species'
-direct effects on one another. The goal of this paper is to resolve this
-long-standing problem.
+methods will thus not provide much evidence regarding species
+interactions.
 
@@
 
-\paragraph{Markov networks.}\label{markov-networks.}
-
-Markov networks provide a framework for translating back and forth
-between the conditional (all-else-equal) relationships among species
-(Figure 1C) and the kinds of species assemblages that these
-relationships produce. Here, I show how a set of conditional
-relationships can be used to determine how groups of species can
-co-occur. Methods for estimating conditional relationships from data are
-discussed in the next section.
+\textbf{Markov networks.} Markov networks provide a framework for
+translating back and forth between the conditional (all-else-equal)
+relationships among species (Figure 1C) and the kinds of species
+assemblages that these relationships produce. Here, I show how a set of
+conditional relationships can be used to determine how groups of species
+can co-occur. Methods for estimating conditional relationships from data
+are discussed in the next section.
 
@@
 
-\(p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),\)
+\(\displaystyle{p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),}\)
 
@@
 
-\noindent where the second sum is over all \(\binom{n}{2}\) species
-pairs. In this model, \(\alpha_{i}\) is an intercept term determining
-the amount that the presence of species \(i\) contributes to the
-log-probability of \(\vec{y}\); it directly controls the prevalence of
-species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
+\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
+of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
+determining the amount that the presence of species \(i\) contributes to
+the log-probability of \(\vec{y}\); it directly controls the prevalence
+of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
 co-occurrence of species \(i\) and species \(j\) contributes to the
@@
 \(e^2\) times higher in sites where its partner is present than in
-comparable sites where its partner is absent. Because the relative
-probability of a presence-absence vector increases when
-positively-associated species co-occur and decreases when
+comparable sites where its partner is absent.
+
+Because the relative probability of a presence-absence vector increases
+when positively-associated species co-occur and decreases when
 negatively-associated species co-occur, the model tends---all else
@@
 positively-associated species and relatively few pairs of
-negatively-associated species (exactly as an ecologist might expect).
-
-Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
+negatively-associated species (exactly as an ecologist might expect). Of
+course, if all else is \emph{not} equal (e.g.~Figure 1, where the
 presence of one competitor is associated with release from another
 competitor), then species' overall co-occurrence patterns can be more
-complicated. Determining the marginal relationships between species from
-their conditional interactions entails summing over the different
-possible assemblages (Figure 2B). This becomes intractable when the
-number of possible assemblages is large, though several methods beyond
-the scope of this paper can be employed to keep the calculations
-feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
-noted below, some common linear and generalized linear methods can also
-be used as computationally efficient approximations to the full network
-(Lee and Hastie 2012, Loh and Wainwright 2013).
+complicated, and may require summing over the different possible
+assemblages (Figure 2B) to determine the overall outcomes.
 
-\paragraph{\texorpdfstring{Estimating \(\alpha\) and \(\beta\)
-coefficients from presence-absence
-data.}{Estimating \textbackslash{}alpha and \textbackslash{}beta coefficients from presence-absence data.}}\label{estimating-alpha-and-beta-coefficients-from-presence-absence-data.}
-
+\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
 In the previous section, the values of \(\alpha\) and \(\beta\) were
@@
 be expected to produce exactly the observed co-occurrence frequencies on
-average (i.e.~maximizing the likelihood ensures that the sufficient
+average (i.e., maximizing the likelihood ensures that the sufficient
 statistics of the model distribution match the sufficient statistics of
@@
 the R programming language (R Core Team 2015) to optimize the Markov
-network parameters. The package was named after Santa Rosalia, the
-patron saint of biodiversity, whose mythos played an important
-rhetorical role in the null model debates of the 1970's and 1980's
-(Lewin 1983).
+network parameters.
 
-\paragraph{Simulated landscapes.}\label{simulated-landscapes.}
-
-I simulated several sets of landscapes using known parameters as a test
-bed for different statistical methods. The first set of landscapes
-included the three competing species shown in Figure 1. For each of 1000
-replicates, I generated a landscape with 100 sites by sampling from a
-probability distribution defined by the figure's interaction
-coefficients (Appendix 1). Each of the methods described below was then
-evaluated on its ability to correctly infer that the two shrub species
-competed with one another, despite their frequent co-occurrence.
+\noindent \textbf{Simulated landscapes.} I simulated several sets of
+landscapes using known parameters as a test bed for different
+statistical methods. The first set of landscapes included the three
+competing species shown in Figure 1. For each of 1000 replicates, I
+generated a landscape with 100 sites by sampling from a probability
+distribution defined by the figure's interaction coefficients (Appendix
+1). Each of the methods described below was then evaluated on its
+ability to correctly infer that the two shrub species competed with one
+another, despite their frequent co-occurrence.
 
@@
 Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
-for each replicate landscape in these simulations so that most species
-pairs interacted negligibly, a few pairs interacted very strongly, and
-competition was three times more common than facilitation.
+for each replicate landscape so that most species pairs interacted
+negligibly, a few pairs interacted very strongly, and competition was
+three times more common than facilitation.
 
@@
 a new set of co-occurrence data so that species' \(\alpha\) coefficients
-depended on these environmental variables. The latter set of simulated
+depended on the local environment. The latter set of simulated
 landscapes provide an important test of the methods' ability to
@@
 any information about the environmental or spatial variables that helped
-shape the landscape (cf Connor et al. 2013). Finally, I simulated the
-population dynamics of 20 interacting species, as described in Appendix
-{[}{[}X{]}{]}
+shape the landscape (cf Connor et al. 2013). I generated the final set
+of landscapes by simulating the population dynamics of 20 interacting
+species, as described in Appendix {[}{[}X{]}{]}.
 
-\paragraph{Recovering species interactions from simulated
-data.}\label{recovering-species-interactions-from-simulated-data.}
-
+\noindent \textbf{Recovering species interactions from simulated data.}
 I compared seven techniques for determining the sign and strength of the
@@
 conditional relationships. This approach is common in molecular biology
-(Friedman et al. 2008), but less so in ecology (see Albrecht and Gotelli
-2001 @faisal\_inferring\_2010 for two exceptions). In the context of
+(Friedman et al. 2008), but is rare in ecology (see Albrecht and Gotelli
+(2001) and Faisal et al. (2010) for two exceptions). In the context of
 non-Gaussian data, the partial correlation can be thought of as a
@@
 
-\paragraph{Evaluating model
-performance.}\label{evaluating-model-performance.}
-
-For the simulated landscapes based on Figure 1, I assessed whether each
-method's test statistic indicated a positive or negative relationship
-between the two shrubs (Appendix 1). For the null model (Pairs), I
-calculated statistical significance using its \(Z\)-score. For the
-Markov network, I used the Hessian matrix to generate approximate
-confidence intervals and noted whether these intervals included zero.
+\noindent \textbf{Evaluating model performance.} For the simulated
+landscapes based on Figure 1, I assessed whether each method's test
+statistic indicated a positive or negative relationship between the two
+shrubs (Appendix 1). For the null model (Pairs), I calculated
+statistical significance using its \(Z\)-score. For the Markov network,
+I used the Hessian matrix to generate approximate confidence intervals
+and noted whether these intervals included zero.
 
@@
 
-\paragraph{Three species.}\label{three-species.}
+\noindent \textbf{Three species.} As shown in Figure 1, the marginal
+relationship between the two shrub species was positive---despite their
+competition for space at a mechanistic level---due to indirect effects
+of the dominant tree species. As a result, the correlation between these
+species was positive in 94\% of replicates, and the randomization-based
+null model falsely reported positive associations 100\% of the time.
+Worse, more than 98\% of these false conclusions were statistically
+significant. The partial correlation and Markov network estimates, on
+the other hand, each correctly isolated the direct negative interaction
+between the shrubs from their positive indirect interaction 94\% of the
+time (although the confidence intervals overlapped zero in most
+replicates).
 
-As shown in Figure 1, the marginal relationship between the two shrub
-species was positive---despite their competition for space at a
-mechanistic level---due to indirect effects of the dominant tree
-species. As a result, the correlation between these species was positive
-in 94\% of replicates, and the randomization-based null model falsely
-reported positive associations 100\% of the time. Worse, more than 98\%
-of these false conclusions were statistically significant. The partial
-correlation and Markov network estimates, on the other hand, each
-correctly isolated the direct negative interaction between the shrubs
-from their positive indirect interaction 94\% of the time (although the
-confidence intervals overlapped zero in most replicates).
-
-\paragraph{Twenty species.}\label{twenty-species.}
-
-Despite some variability across contexts (Figure 3), the four methods
-that controlled for indirect effects (the Markov network, the
-generalized linear models, and the two partial correlation-based
-methods) clearly outperformed those that did not. Of these, the Markov
-network consistently performed the best, especially on the smaller
-landscapes (which are also the ones that are most representative of
-typical analyses in this field; Gotelli and Ulrich 2009).
+\noindent
+\textbf{Twenty species.} Despite some variability across contexts
+(Figure 3), the four methods that controlled for indirect effects (the
+Markov network, the generalized linear models, and the two partial
+correlation-based methods) outperformed those that did not. Of these,
+the Markov network consistently performed the best, especially on the
+smaller landscapes (which are also the most representative of typical
+analyses in this field; Gotelli and Ulrich 2009). As shown in Appendix
+{[}{[}X{]}{]}, the standard errors associated with the estimates in
+Figure 3 are all small (less than 0.01), so differences across methods
+should not be attributed to sampling error.
 
@@
 of species' true interaction strengths than either the joint species
-distribution model (BayesComm) or the null model (Pairs). Interestingly,
-the estimates from these three approaches were tightly correlated
-{[}{[}Figure 3B{]}{]}. This suggests that the effect sizes from the null
-model (and, to a lesser extent, the correlation matrices from joint
-species distribution models) only contain noisy versions of the same
-information that could be obtained more easily and interpretably by
-calculating correlation coefficients between species' presence-absence
-vectors.
+distribution model (BayesComm) or the null model (Pairs). The estimates
+from these three approaches were tightly correlated (Figure 3B),
+suggesting that the effect sizes from the null model (and, to a lesser
+extent, the correlation matrices from joint species distribution models)
+only contain noisy versions of the same information that could be
+obtained more easily and interpretably by calculating correlation
+coefficients between species' presence-absence vectors.
 
-Bootstrap resampling indicated that the above ranking of the different
-methods was robust (Appendix 3). In particular, the 95\% confidence
-interval of the bootstrap distribution indicated that the Markov
-network's overall \(R^2\) value was between {[}{[}14 and 18{]}{]}
-percent higher than the second-most effective method (generalized linear
-models) and between {[}{[}2.12 and 2.38{]}{]} times higher than could be
-achieved by the null model (Pairs).
+Whereas Figure 3 summarizes the accuracy of each model's point
+estimates, Figure 4 compares the inferential value of the Markov network
+versus the null model. When the ``true'' \(\beta\) coefficient is near
+zero, the Markov network's 95\% confidence intervals usually include
+zero (but not quite 95\% of the time), and it is rare for the model to
+confidently attribute the wrong sign to a coefficient (Figure 4A). The
+null model, by contrast, typically rejects the null hypothesis at the
+\(p=0.05\) level regardless of the underlying parameters, and the sign
+of its \(Z\)-scores is not a reliable indicator of the interaction's
+actual sign.
 
@@
 networks---have been studied for nearly a century (Cipra 1987), and the
-models' properties, capabilities, and limits are well-understood in a
-huge range of applications. Using the same framework for species
-interactions would thus allow ecologists to tap into an enormous set of
-existing discoveries and techniques for dealing with indirect effects,
-stability, and alternative stable states. Numerous other extensions are
-possible: for example, the states of the interaction network can be
-modeled as a function of the local abiotic environment (Lee and Hastie
-2012), which would provide a rigorous and straightforward approach to
-the difficult and important task of incorporating whole networks of
-biotic interactions into species distribution models (Pollock et al.
-2014), leading to a better understanding of the interplay between biotic
-and abiotic effects on community structure. There are even methods
-(Whittam and Siegel-Causey 1981, Tjelmeland and Besag 1998) that would
-allow one species to affect the sign or strength of the relationship
-between two other species, tipping the balance between facilitation and
-exploitation (Bruno et al. 2003).
+models' properties, capabilities, and limits are well-understood. Using
+the same framework for species interactions would thus allow ecologists
+to tap into an enormous set of existing discoveries and techniques for
+dealing with indirect effects, stability, and alternative stable states.
+Numerous other extensions are possible: for example, the states of the
+interaction network can be modeled as a function of the local abiotic
+environment (Lee and Hastie 2012), which would provide a rigorous and
+straightforward approach to the difficult and important task of
+incorporating whole networks of biotic interactions into species
+distribution models (Pollock et al. 2014), leading to a better
+understanding of the interplay between biotic and abiotic effects on
+community structure. Alternatively, models could allow one species to
+affect the interactions between two other species (Tjelmeland and Besag
+1998), tipping the balance between facilitation and exploitation (Bruno
+et al. 2003).
 
@@
 null (Gotelli and Ulrich 2009), especially when the test statistic is
-effectively just a correlation coefficient (Figure 3B). Simultaneous
-estimation of multiple ecological parameters seems like a much more
-promising approach: to the extent that the models' relative performance
-on real data sets is similar to the range of results shown in Figure 3A,
-scientists in this field could often double their explanatory power by
-switching from null models to Markov networks (or increase it
-substantially with linear or generalized linear approximations).
-Regardless of the methods ecologists ultimately choose, controlling for
-indirect effects could clearly improve our understanding of species'
-direct effects on one another and on community structure.
+effectively just a correlation coefficient {[}{[}Figure 3B{]}{]}. When a
+non-null backdrop is not controlled for, Type I error rates can
+skyrocket, the apparent sign of the interaction can change, and null
+models can lead to reliably incorrect inferences (Figure 1, Figure 4).
 
-\paragraph{Acknowledgements:}\label{acknowledgements}
+Controlling for indirect effects via simultaneous estimation of multiple
+ecological parameters seems like a much more promising approach: to the
+extent that the models' relative performance on real data sets is
+similar to the range of results shown in Figure 3A, scientists in this
+field could more than double their explanatory power by switching from
+null models to Markov networks (or increase it nearly as much with
+linear or generalized linear approximations). Regardless of the methods
+ecologists ultimately choose, controlling for indirect effects could
+clearly improve our understanding of species' direct effects on one
+another and on community structure.
 
-This research was funded by a Graduate Research Fellowship from the US
-National Science Foundation and benefited greatly from discussions with
-A. Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C.
-S. Tysor. Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li,
-D. L. Miller, N. Golding, N. J. Gotelli, C. F. Dormann, and two
-anonymous reviewers provided very helpful feedback on the text.
+\noindent \textbf{Acknowledgements:} This research was funded by a
+Graduate Research Fellowship from the US National Science Foundation and
+benefited greatly from discussions with A. Sih, M. L. Baskett, R.
+McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A.
+K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding,
+N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided very
+helpful feedback on the text.
 
@@
 
-\textbf{Figure 1: A.} A small network of three competing species. The
-tree (top) tends not to co-occur with either of the two shrub species,
-as indicated by the strongly negative coefficient linking them. The two
-shrub species also compete with one another, but more weakly (circled
-coefficient). \textbf{B.} In spite of the competitive interactions
-between the two shrub species, their shared tendency to occur in
-locations without trees makes their occurrence vectors positively
-correlated (circled). \textbf{C.} Controlling for the tree species'
-presence with a conditional (all-else-equal) approach such as a partial
-covariance or a Markov network leads to correct identification of the
-negative shrub-shrub interaction (circled).
+\noindent \textbf{Figure 1: A.} A small network of three competing
+species. The tree (top) tends not to co-occur with either of the two
+shrub species, as indicated by the strongly negative coefficient linking
+them. The two shrub species also compete with one another, but more
+weakly (circled coefficient). \textbf{B.} In spite of the competitive
+interactions between the two shrub species, their shared tendency to
+occur in locations without trees makes their occurrence vectors
+positively correlated (circled). \textbf{C.} Controlling for the tree
+species' presence with a conditional (all-else-equal) approach such as a
+partial covariance or a Markov network leads to correct identification
+of the negative shrub-shrub interaction (circled).
 
-\textbf{Figure 2: A.} A small Markov network with two species, defined
-by its \(\alpha\) and \(\beta\) values. The abiotic environment favors
-the occurrence of each species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient linking
-these two species implies that they co-occur less than expected under
-independence. \textbf{B.} These coefficients determine the probabilities
-of all four possible presence-absence combinations for Species 1 and
-Species 2. \(\alpha_1\) is added to the exponent whenever Species 1 is
-present (\(y_1 = 1\)), but not when it is absent (\(y_1 = 0\)).
-Similarly, the exponent includes \(\alpha_2\) only when species \(2\) is
-present (\(y_2 = 1\)), and \(\beta\) only when both are present
-(\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that the four
-probabilities sum to 1. In this case, \(Z\) is about 18.5. \textbf{C.}
-The expected frequencies of all possible co-occurrence patterns between
-the two species of interest, as calculated in the previous panel.
-\textbf{D.} If, as assumed in null model simulations, \(\beta_{12}\)
-equaled zero (e.g.~if the species no longer competed for the same
-resources), then the reduction in competition would allow each species
-to increase its overall occurrence rate.
+\noindent \textbf{Figure 2: A.} A small Markov network with two species,
+defined by its \(\alpha\) and \(\beta\) values. The abiotic environment
+favors the occurrence of each species (\(\alpha >0\)), particularly
+species 2 (\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient
+linking these two species implies that they co-occur less than expected
+under independence. \textbf{B.} These coefficients determine the
+probabilities of all four possible presence-absence combinations for
+Species 1 and Species 2. \(\alpha_1\) is added to the exponent whenever
+Species 1 is present (\(y_1 = 1\)), but not when it is absent
+(\(y_1 = 0\)). Similarly, the exponent includes \(\alpha_2\) only when
+species \(2\) is present (\(y_2 = 1\)), and \(\beta\) only when both are
+present (\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that
+the four probabilities sum to 1. In this case, \(Z\) is about 18.5.
+\textbf{C.} The expected frequencies of all possible co-occurrence
+patterns between the two species of interest, as calculated in the
+previous panel. \textbf{D.} If, as assumed in null model simulations,
+\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
+the same resources), then the reduction in competition would allow each
+species to increase its overall occurrence rate.
 
-\textbf{Figure 3: A.} Proportion of variance in interaction coefficients
-explained by each method versus number of sampled locations. \textbf{B.}
-The \(Z\)-scores produced by the null model (``Pairs'') for each pair of
-species can be predicted using the correlation between the
-presence-absence vectors of those same species and from the number of
-sites on the landscape.
+\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
+coefficients explained by each method versus number of sampled
+locations. \textbf{B.} The \(Z\)-scores produced by the null model
+(``Pairs'') for each pair of species can be predicted using the
+correlation between the presence-absence vectors of those same species
+and from the number of sites on the landscape.
+
+\noindent \textbf{Figure 4: A.} Inferences from the Markov network's
+95\% confidence intervals on the ``no-environment'' simulated
+landscapes, as a function of the ``true'' coefficient values.
+\textbf{B.} Inferences from the \(Z\)-statistics estimated by the
+``Pairs'' software for null modeling.
 
@@
 
-\setlength{\parskip}{0pt} \setlength{\parindent}{-1em}
-\setlength{\leftskip}{1em}
+\setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
 
@@
 
-Salakhutdinov, R. 2008. Learning and evaluating Boltzmann machines.
-Technical Report UTML TR 2008-002, Department of Computer Science,
-University of Toronto, Dept. of Computer Science, University of Toronto.
-
 Schäfer, J., R. Opgen-Rhein, V. Zuber, M. Ahdesmäki, A. P. D. Silva, and

### 2015-10-26 Harris_2015/show-results.R

@@ -41,7 +41,7 @@ result_summary %>%
 
 
 
-pdf("manuscript-materials/figures/performance.pdf", width = 7.5, height = 3)
+pdf("manuscript-materials/figures/performance.pdf", width = 7.5, height = 2.5)
 ggplot(result_summary, aes(x = n_sites, y = r2, col = method)) + 
   facet_grid(~simulation_type) + 
   geom_line(size = .5) + 
@@ -71,6 +71,21 @@ ggplot(result_summary, aes(x = n_sites, y = r2, col = method)) +
 dev.off()
 
 
+####
+library(lme4)
+library(multcomp)
+
+landscape_estimates = x %>% 
+  group_by(method, simulation_type) %>%
+  do(data.frame(., resids = resids(.))) %>%
+  ungroup %>%
+  group_by(method, simulation_type, rep_name, n_sites) %>%
+  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))
+
+
+summary(lmer(r2 ~ method + n_sites + simulation_type + (1|rep_name), data = landscape_estimates))
+
+####
 
 
 xx = x[x$method == "Markov network" & grepl("^env[0-9]*$", x$rep_name), ]
@@ -84,7 +99,20 @@ pdf("manuscript-materials/figures/error_rates.pdf", width = 5, height = 5)
 plot(p_ns[order(xx$truth)] ~ sort(xx$truth), type = "l", lwd = 3, ylim = c(0, 1), yaxs = "i", bty = "l", ylab = "Proportion", las = 1, xlab = "\"True\" coefficient", xaxs = "i")
 lines(p_positive[order(xx$truth)] ~ sort(xx$truth), lwd = 3, col = 4)
 lines(p_negative[order(xx$truth)] ~ sort(xx$truth), lwd = 3, col = 2)
-text(0, .95, "Not\nsignificant")
+text(0, .95, "Not significant\nat the 0.05 level")
 text(5, .8, "Significantly\npositive", col = 4)
 text(-7, .35, "Significantly\nnegative", col = 2)
 dev.off()
+
+
+library(tidyr)
+z = x %>%
+  filter(method %in% c("correlation", "BayesComm") & simulation_type == "no_env") %>%
+  dplyr::select(-lower, -upper, -X) %>%
+  spread(method, estimate)
+
+ggplot(z, aes(x = correlation, y = BayesComm, col = factor(n_sites))) + 
+  geom_point()
+
+
+

### 2015-10-29 Harris_2015/manuscript-materials/draft.Rmd
almost done

@@
 it is rare for the model to confidently attribute the wrong sign to a
-coefficient (Figure 4A). The null model, by contrast, typically rejects the null
+coefficient (Figure 4A). The Pairs software, by contrast, typically rejects the null
 hypothesis at the $p=0.05$ level regardless of the underlying parameters, and
 the sign of its $Z$-scores is not a reliable indicator of the interaction's
-actual sign.
+actual sign (Figure 4B).
 

### 2015-10-29 Harris_2015/show-results.R

@@ -1,12 +1,103 @@
 library(dplyr)
 library(mgcv)
 library(ggplot2)
-
+library(tidyr)
 
 
 x = read.csv("estimates.csv", stringsAsFactors = FALSE)
 x$dynamic_simulation = grepl("pop", x$rep_name)
 x$simulation_type = gsub("[0-9]", "", x$rep_name)
+
+
+######## Pairs ########
+
+pairs_txt = readLines("fakedata/matrices/Pairs.txt")
+library(stringr)
+
+# Find areas of the data file that correspond
+# to species pairs' results
+beginnings = grep("Sp1", pairs_txt) + 1
+ends = c(
+  grep("^[^ ]", pairs_txt)[-1],
+  length(pairs_txt)
+) - 1
+
+file_list = readLines("manuscript-materials/fakedata_list.txt")[-1]
+partial_names = sapply(strsplit(grep("^>", pairs_txt, value = TRUE), " +"), function(x) x[[3]])
+filename_lines = grep("^>", pairs_txt)
+
+# File_list agrees with partial_names to the extent possible
+all(sapply(1:450, function(i)grepl(partial_names[i], file_list[i])))
+
+
+split_names = strsplit(file_list, "-")
+
+pairs_results = lapply(
+  1:length(file_list),
+  function(i){
+    
+    message(i)
+    
+    n_sites = as.integer(split_names[[i]][[2]])
+    rep_name = split_names[[i]][[3]]
+    
+    # Find the line where the current data set is mentioned in
+    # pairs.txt
+    filename_line = filename_lines[i]
+    
+    # Which chunk of the data file corresponds to this file?
+    chunk = min(which(beginnings > filename_line))
+    
+    # Split the chunk on whitespace.
+    splitted = strsplit(pairs_txt[beginnings[chunk]:ends[chunk]], " +")
+    
+    # Pull out the corresponding chunk of the "x" data frame, based on n_sites and rep_name
+    x_subset = x[x$n_sites == n_sites & x$rep_name == rep_name & x$method == "correlation", ]
+    
+    # Pull out the species numbers and their Z-scores, then join to x_subset
+    pairs_results = lapply(
+      splitted,
+      function(x){
+        # in the x data frame, species 1 is always a lower number than species 2
[…]

### 2015-10-31 Harris_2015/manuscript-materials/draft.Rmd
closer

--- 2015-10-26 Harris_2015/manuscript-materials/draft.Rmd
@@
 
-\setlength{\parindent}{1cm}
+\setlength{\parindent}{2em}
 
@@
 unreliable [@gotelli_empirical_2009]. In particular, species' effects on one
-another can become lost in the complex web of indirect effects. For example, the
-competitive interaction between the two shrub species in Figure 1A can become
+another can become lost in a web of indirect effects. For example, the
+competitive interaction between the two shrub species in Figure 1A is
 obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
@@
 those in Figure 1 are generally important [@dodson_complementary_1970], existing
-methods will thus not provide much evidence regarding species interactions.
+methods will not provide much evidence regarding species interactions.
 
@@
 signal in the data (Figure 1C). For example, after partitioning the data set
-into shaded sites and unshaded sites, there will be co-occurrence deficits in
-each subset that might not be apparent at the whole-landscape level. More
-generally, controlling for other species in the network will often be important
-for obtaining reliable estimates of direct (conditional, or all-else-equal)
-effects.  This kind of precision is difficult to obtain from null models, which
-only simulate the most extreme possible scenario: that *all* direct and indirect
-interactions are exactly zero. Nevertheless, null models have dominated this
-field for more than three decades [@strong_ecological_1984;
-@gotelli_empirical_2009].
+into shaded and unshaded sites, there will be co-occurrence deficits in
+each subset that wouldn't otherwise be apparent. More generally, controlling for
+other species in the network will often be important for obtaining reliable
+estimates of direct (conditional, or all-else-equal) effects.  This kind of
+precision is difficult to obtain from null models, which only simulate the most
+extreme possible scenario: that *all* direct and indirect interactions are
+exactly zero. Nevertheless, null models have dominated this field for more than
+three decades [@strong_ecological_1984; @gotelli_empirical_2009].
 
@@
 landscape-level consequences of pairwise species interactions,
-and for detecting them from observed presence-absence matrices. Markov networks
+and for estimating them from observed presence-absence matrices. Markov networks
 have been used in many scientific fields in similar contexts for decades, from
@@
 
-Below, I introduce Markov networks and show how they can be used to simulate
-landscape-level data from an interaction matrix or to make exact predictions
-about the direct and indirect consequences of possible interaction matrices.
-Then, using simulated data sets where the "true" interactions are known,
-I compare this approach with several existing methods. Finally, I discuss
-opportunities for extending the approach presented here to other problems in
-community ecology, e.g. quantifying the overall effect of species interactions
-on occurrence rates [@roughgarden_competition_1983] and disentangling the
-effects of biotic versus abiotic interactions on species composition
-[@pollock_understanding_2014].
+Below, I demonstrate Markov networks' ability to produce exact predictions
+about the direct and indirect consequences of an interaction matrix, and also to
+make inferences about the species interactions that contributed to an observed
+set of co-occurrences. Using simulated data sets where the "true" interactions
+are known, I compare this approach with several existing methods. Finally, I
+discuss opportunities for extending the approach presented here to other
+problems in community ecology, e.g. quantifying the overall effect of species
+interactions on occurrence rates [@roughgarden_competition_1983] and
+disentangling the effects of biotic versus abiotic interactions on species
+composition [@pollock_understanding_2014].
 
@@
 \raggedright
-\setlength{\parindent}{1cm}
+\setlength{\parindent}{2em}
 
-\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$ species. In this
-model, $\alpha_{i}$ is an intercept term determining the amount that the presence
-of species $i$ contributes to the log-probability of $\vec{y}$; it directly
-controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
-that the co-occurrence of species $i$ and species $j$ contributes to the
-log-probability; it controls the conditional relationship between two species,
-i.e. the probability that they will be found together, after controlling for the
-other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
-might have a value of $+2$ for two mutualists, indicating that the odds of
-observing one species are $e^2$ times higher in sites where its partner is
-present than in comparable sites where its partner is absent.
-
-Because the relative probability of a presence-absence vector increases when
-positively-associated species co-occur and decreases when negatively-associated
-species co-occur, the model tends---all else equal---to produce assemblages that
-have many pairs of positively-associated species and relatively few pairs of
-negatively-associated species (exactly as an ecologist might expect).
-Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
-competitor is associated with release from another competitor), then species'
-overall co-occurrence patterns can be more complicated, and may require summing
-over the different possible assemblages (Figure 2B) to determine the overall
-outcomes.
+\noindent where the second sum is over all $\frac{1}{2}n(n-1)$ pairs of $n$
+species. In this model, $\alpha_{i}$ is an intercept term determining the amount
+that the presence of species $i$ contributes to the log-probability of
+$\vec{y}$; it directly controls the prevalence of species $i$. Similarly,
+$\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
+contributes to the log-probability; it controls the conditional relationship
+between two species, i.e. the probability that they will be found together,
+after controlling for the other species in the network (Figure 2A, Figure 2B).
+For example, if $\beta_{ij} = -2$, then each species' odds of occurrence are
+$exp{2}$ times lower when the other one is present (as compared with otherwise
+comparable sites). Because the relative probability of a presence-absence vector
+increases when positively-associated species co-occur and decreases when
+negatively-associated species co-occur, the model tends---all else equal---to
+produce assemblages that have many pairs of positively-associated species and
+relatively few pairs of negatively-associated species (exactly as an ecologist
+might expect). Of course, if all else is *not* equal (e.g. Figure 1, where the
+presence of one competitor is associated with release from another competitor),
+then species' overall co-occurrence patterns can be more complicated, and may
+require summing over the different possible assemblages (Figure 2B) to determine
+the overall outcomes.
 
@@
 significance using its $Z$-score. For the Markov network, I used the Hessian
-matrix to generate approximate confidence intervals and noted whether these
-intervals included zero.
+matrix to generate approximate confidence intervals.
 
@@
 that the different test statistics (e.g. correlations versus $Z$ scores) were on
-a common scale, I rescaled them using linear regression through the origin. After
-rescaling each method’s estimates, I calculated squared errors between the scaled
-interaction estimates and "true" interaction values across the simulated data
-sets. These squared errors determined the proportion of variance explained
-for different combinations of model type and landscape size (compared with a
-null model that assumed all interaction strengths to be zero).
+a common scale, I rescaled them using linear regression through the origin. I
+then calculated the proportion of variance explained for different combinations
+of model type and landscape size (compared with a null model that assumed all
+interaction strengths to be zero).
 
@@
 As shown in Appendix [[X]], the standard errors associated with the estimates
-in Figure 3 are all small (less than 0.01), so differences across methods should
-not be attributed to sampling error.
+in Figure 3 are small (less than 0.01), so these differences should not be
+attributed to sampling error.
 
@@
 (BayesComm) or the null model (Pairs). The estimates from these
-three approaches were tightly correlated (Figure 3B), suggesting that the
-effect sizes from the null model (and, to a lesser extent, the correlation
-matrices from joint species distribution models) only contain noisy versions of
-the same information that could be obtained more easily and interpretably by
-calculating correlation coefficients between species' presence-absence vectors.
+three approaches were tightly correlated (Figure [[3B]]), suggesting that the
+null model only contains a noisy version of the same information that could be
+obtained more easily and interpretably with simple correlation coefficients.
 
-Whereas Figure 3 summarizes the accuracy of each model's point estimates, Figure
-4 compares the inferential value of the Markov network versus the null model.
-When the "true" $\beta$ coefficient is near zero, the Markov network's 95%
-confidence intervals usually include zero (but not quite 95% of the time), and
-it is rare for the model to confidently attribute the wrong sign to a
-coefficient (Figure 4A). The null model, by contrast, typically rejects the null
-hypothesis at the $p=0.05$ level regardless of the underlying parameters, and
-the sign of its $Z$-scores is not a reliable indicator of the interaction's
-actual sign.
+Finally, we can evaluate the models' statistical inferences.  Figure 4 shows,
+across a range of true interaction strengths, the probability that the null model
+or the Markov network will predict the wrong sign of the interaction with 95%
+confidence.  The null model makes such errors nearly 8 times as often as the
+Markov network, even though it only reject the null hypothesis twice as often
+overall [[Appendix]].  The Markov network's errors were also more concentrated
+around 0, so it never misclassified strong interactions like the null model did.
 
@@
 environmental heterogeneity and indirect interactions cause ecologists' typical
-null modeling approaches to reliably fail. For cases where fitting a Markov
-network is computationally infeasible, these results also indicate that partial
-covariances and generalized linear models can both provide useful approximations.
-The partial correlations' success on simulated data may not carry over to real
-data sets, however; @loh_structure_2013 show that the linear approximations can
-be less reliable in cases where the true interaction matrix contains more
-structure (e.g. guilds or trophic levels). Similarly, the approximation involved
-in using separate generalized linear models for each species can occasionally
-lead to catastrophic overfitting with small-to-moderate sample sizes
-[@lee_learning_2012]. For these reasons, it will usually be best to fit a Markov
-network rather than one of the alternative methods when one’s computational
-resources allow it.
+null modeling approaches to reliably fail. Partial covariances and generalized
+linear models can both provide computationally efficient approximations, but
+with somewhat lower accuracy. The difference in accuracy may be larger for real
+data sets than for the simulated landscapes in Figure 3, however; linear
+approximations to the Markov network make larger errors when the interaction
+matrix is structured [e.g. due to guilds or trophic levels;
+@loh_structure_2013]. Similarly, the separate generalized linear models
+for each species can severely overfit in some cases [@lee_learning_2012]. The
+full Markov network should thus be preferred to the approximations when
+it is computationally feasible.
 
-It’s important to note that none of these methods can identify the exact nature
-of the pairwise interactions [e.g. which species in a positively-associated
-pair is facilitating the other; @schmidt_modeling_2012], particularly when real
-pairs of species can reciprocally influence one another in multiple ways
-simultaneously [@bruno_inclusion_2003]; with compositional data, there is only
-enough information to provide a single number describing each species pair.
-To estimate asymmetric interactions, such as commensalism or predation,
-ecologists would need other kinds of data, as from time series, behavioral
-observations, manipulative experiments, or natural history. These other sources
-of information could also be used to augment the likelihood function with an
-informative prior distribution, which could lead to better results on some real
-data sets than was shown in Figure 3A.
+Compositional data only has enough degrees of freedom to estimate one interaction
+per species pair [@schmidt_modeling_2012], so none of these methods can identify
+the exact nature of the pairwise interactions [e.g. which species in a
+positively-associated pair is facilitating the other]. To estimate asymmetric
+interactions, such as commensalism or predation, ecologists could use time
+series, behavioral observations, manipulative experiments, or natural history.
+These other sources of information could also be used to augment the likelihood
+function with an informative prior distribution, reducing ecologists' eror and
+uncertainty.
 
@@
 insight would have been difficult to obtain without a generative model that
-makes predictions about the consequences of species interactions; null models
-(which presume *a priori* that interactions do not exist) have no way to make
-such predictions.
+predicts the consequences of species interactions (as opposed to null
+simulations that omit interactions entirely).
 
 Markov networks---particularly the Ising model for binary networks---have been
-studied for nearly a century [@cipra_introduction_1987], and the models'
-properties, capabilities, and limits are well-understood. Using the same
-framework for species interactions would thus allow ecologists to tap into an
-enormous set of existing discoveries and techniques for dealing with indirect
-effects, stability, and alternative stable states. Numerous other extensions are
-possible: for example, the states of the interaction network can be modeled as a
-function of the local abiotic environment [@lee_learning_2012], which would
-provide a rigorous and straightforward approach to the difficult and important
-task of incorporating whole networks of biotic interactions into species
-distribution models [@pollock_understanding_2014], leading to a
-better understanding of the interplay between biotic and abiotic effects on
-community structure. Alternatively, models could allow one species to affect the
-interactions between two other species [@tjelmeland_markov_1998], tipping the
-balance between facilitation and exploitation [@bruno_inclusion_2003].
+studied for nearly a century [@cipra_introduction_1987], and the models are
+well-understood. Using the same framework for species interactions would thus
+allow ecologists to tap into a vast set of existing discoveries and techniques
+for dealing with indirect effects, stability, and alternative stable states.
+Numerous other extensions are possible: for example, the states of the
+interaction network can be modeled as a function of the local abiotic
+environment [@lee_learning_2012], which would help incorporate networks of
+biotic interactions into species distribution models
+[@pollock_understanding_2014] and lead to a better understanding of the
+interplay between biotic and abiotic effects on community structure.
+Alternatively, models could allow one species to alter the relationship between
+two other species [@tjelmeland_markov_1998; cf @bruno_inclusion_2003].
 
@@
 ecologists' continued use of null models for studying species interactions. Null
-and neutral models can be useful for clarifying our thinking about the numerical
-consequences of species' richness and abundance patterns
+and neutral models can be useful for clarifying our thinking
 [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
@@
 in small networks with three species, it may simply not be possible to implicate
-individual species pairs or specific ecological processes like competition by
-rejecting a general-purpose null [@gotelli_empirical_2009], especially when the
-test statistic is effectively just a correlation coefficient [[Figure 3B]].
-When a non-null backdrop is not controlled for, Type I error rates can
-skyrocket, the apparent sign of the interaction can change, and null models
-can lead to reliably incorrect inferences (Figure 1, Figure 4).
+specific ecological processes like competition by rejecting a general-purpose
+null [@gotelli_empirical_2009], especially when the test statistic is
+effectively just a correlation coefficient [[Figure 3B]]. When a non-null
+backdrop is not controlled for, Type I error rates can skyrocket, the apparent
+sign of the interaction can change, and null models can lead to reliably
+incorrect inferences (Figure 1, Figure 4).
 
@@
 \noindent \textbf{Acknowledgements:}
-This research was funded by a Graduate Research Fellowship from the US
-National Science Foundation and benefited greatly from discussions with A.
-Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
-Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
-Golding, N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided
-very helpful feedback on the text.
+This work benefited greatly from discussions with A. Sih, M. L. Baskett, R.
+McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A. K.
+Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding, N. J.
+Gotelli, C. F. Dormann, and two anonymous reviewers provided very helpful
+feedback on the text. This research was partially supported by a Graduate
+Research Fellowship from the US National Science Foundation and by the Gordon
+and Betty Moore Foundation's Data-Driven Discovery Initiative through Grant
+GBMF4563 to E. P. White.
 
@@
 
-\noindent \textbf{Figure 2: A.} A small Markov network with two species, defined by its $\alpha$ and
+\noindent \textbf{Figure 2: A.} A small Markov network, defined by its $\alpha$ and
 $\beta$ values.  The abiotic environment favors the occurrence of each species
 ($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
-$\beta$ coefficient linking these two species implies that they co-occur less
-than expected under independence. **B.** These coefficients determine the
-probabilities of all four possible presence-absence combinations for Species 1
-and Species 2. $\alpha_1$ is added to the exponent whenever Species 1 is present
-($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the exponent
-includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
-$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
-ensures that the four probabilities sum to 1.  In this case, $Z$ is
-about 18.5.  **C.** The expected frequencies of all possible
-co-occurrence patterns between the two species of interest, as calculated in the
-previous panel. **D.** If, as assumed in null model simulations, $\beta_{12}$
-equaled zero (e.g. if the species no longer competed for the same resources),
-then the reduction in competition would allow each species to increase its
-overall occurrence rate.
+$\beta_12$ coefficient is consistent with competition between the two species.
+**B.** The coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
+to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
+absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
+species $2$ is present ($y_2 = 1$), and includes $\beta_12$ only when both are
+present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
+probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
+frequencies of all possible co-occurrence patterns between the two species of
+interest, as calculated in the previous panel. **D.** If $\beta_{12}$ equaled
+zero (e.g. if the species no longer competed for the same resources), then the
+each species' overall occurrence rates would increase.
 
-\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
+\noindent \textbf{Figure 3:} Proportion of variance in interaction
 coefficients explained by each method versus number of sampled locations.
-**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
-species can be predicted using the correlation between the presence-absence
-vectors of those same species and from the number of sites on the landscape.
 

### 2015-10-31 Harris_2015/manuscript-materials/draft.tex
closer

@@
 \usepackage{ifthen}
-\geometry{verbose,letterpaper,tmargin=2.35cm, bmargin=2.35cm,lmargin=2.44cm,rmargin=2.44cm}
+\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
 \ifxetex
@@
 unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
-one another can become lost in the complex web of indirect effects. For
-example, the competitive interaction between the two shrub species in
-Figure 1A can become obscured by their shared tendency to occur in
-unshaded areas (Figure 1B). While ecologists have long known that
-indirect effects can overwhelm direct ones at the landscape level
-(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
-inferences from observational data do not control for these effects
-(e.g. Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech
-2013, Pollock et al. 2014). To the extent that indirect interactions
-like those in Figure 1 are generally important (Dodson 1970), existing
-methods will thus not provide much evidence regarding species
-interactions.
+one another can become lost in a web of indirect effects. For example,
+the competitive interaction between the two shrub species in Figure 1A
+is obscured by their shared tendency to occur in unshaded areas (Figure
+1B). While ecologists have long known that indirect effects can
+overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
+the vast majority of our methods for drawing inferences from
+observational data do not control for these effects (e.g. Diamond 1975,
+Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
+2014). To the extent that indirect interactions like those in Figure 1
+are generally important (Dodson 1970), existing methods will not provide
+much evidence regarding species interactions.
 
@@
 leave a signal in the data (Figure 1C). For example, after partitioning
-the data set into shaded sites and unshaded sites, there will be
-co-occurrence deficits in each subset that might not be apparent at the
-whole-landscape level. More generally, controlling for other species in
-the network will often be important for obtaining reliable estimates of
-direct (conditional, or all-else-equal) effects. This kind of precision
-is difficult to obtain from null models, which only simulate the most
-extreme possible scenario: that \emph{all} direct and indirect
-interactions are exactly zero. Nevertheless, null models have dominated
-this field for more than three decades (Strong et al. 1984, Gotelli and
-Ulrich 2009).
+the data set into shaded and unshaded sites, there will be co-occurrence
+deficits in each subset that wouldn't otherwise be apparent. More
+generally, controlling for other species in the network will often be
+important for obtaining reliable estimates of direct (conditional, or
+all-else-equal) effects. This kind of precision is difficult to obtain
+from null models, which only simulate the most extreme possible
+scenario: that \emph{all} direct and indirect interactions are exactly
+zero. Nevertheless, null models have dominated this field for more than
+three decades (Strong et al. 1984, Gotelli and Ulrich 2009).
 
@@
 2012) can provide a framework for understanding the landscape-level
-consequences of pairwise species interactions, and for detecting them
+consequences of pairwise species interactions, and for estimating them
 from observed presence-absence matrices. Markov networks have been used
@@
 
-Below, I introduce Markov networks and show how they can be used to
-simulate landscape-level data from an interaction matrix or to make
-exact predictions about the direct and indirect consequences of possible
-interaction matrices. Then, using simulated data sets where the ``true''
-interactions are known, I compare this approach with several existing
-methods. Finally, I discuss opportunities for extending the approach
-presented here to other problems in community ecology, e.g.~quantifying
-the overall effect of species interactions on occurrence rates
-(Roughgarden 1983) and disentangling the effects of biotic versus
-abiotic interactions on species composition (Pollock et al. 2014).
+Below, I demonstrate Markov networks' ability to produce exact
+predictions about the direct and indirect consequences of an interaction
+matrix, and also to make inferences about the species interactions that
+contributed to an observed set of co-occurrences. Using simulated data
+sets where the ``true'' interactions are known, I compare this approach
+with several existing methods. Finally, I discuss opportunities for
+extending the approach presented here to other problems in community
+ecology, e.g.~quantifying the overall effect of species interactions on
+occurrence rates (Roughgarden 1983) and disentangling the effects of
+biotic versus abiotic interactions on species composition (Pollock et
+al. 2014).
 
@@
 \raggedright
-\setlength{\parindent}{1cm}
+\setlength{\parindent}{2em}
 
@@
 controlling for the other species in the network (Figure 2A, Figure 2B).
-For example, \(\beta_{ij}\) might have a value of \(+2\) for two
-mutualists, indicating that the odds of observing one species are
-\(e^2\) times higher in sites where its partner is present than in
-comparable sites where its partner is absent.
-
-Because the relative probability of a presence-absence vector increases
-when positively-associated species co-occur and decreases when
+For example, if \(\beta_{ij} = -2\), then each species' odds of
+occurrence are \(exp{2}\) times lower when the other one is present (as
+compared with otherwise comparable sites). Because the relative
+probability of a presence-absence vector increases when
+positively-associated species co-occur and decreases when
 negatively-associated species co-occur, the model tends---all else
@@
 statistical significance using its \(Z\)-score. For the Markov network,
-I used the Hessian matrix to generate approximate confidence intervals
-and noted whether these intervals included zero.
+I used the Hessian matrix to generate approximate confidence intervals.
 
@@
 statistics (e.g.~correlations versus \(Z\) scores) were on a common
-scale, I rescaled them using linear regression through the origin. After
-rescaling each method's estimates, I calculated squared errors between
-the scaled interaction estimates and ``true'' interaction values across
-the simulated data sets. These squared errors determined the proportion
-of variance explained for different combinations of model type and
-landscape size (compared with a null model that assumed all interaction
-strengths to be zero).
+scale, I rescaled them using linear regression through the origin. I
+then calculated the proportion of variance explained for different
+combinations of model type and landscape size (compared with a null
+model that assumed all interaction strengths to be zero).
 
@@
 {[}{[}X{]}{]}, the standard errors associated with the estimates in
-Figure 3 are all small (less than 0.01), so differences across methods
-should not be attributed to sampling error.
+Figure 3 are small (less than 0.01), so these differences should not be
+attributed to sampling error.
 
@@
 distribution model (BayesComm) or the null model (Pairs). The estimates
-from these three approaches were tightly correlated (Figure 3B),
-suggesting that the effect sizes from the null model (and, to a lesser
-extent, the correlation matrices from joint species distribution models)
-only contain noisy versions of the same information that could be
-obtained more easily and interpretably by calculating correlation
-coefficients between species' presence-absence vectors.
+from these three approaches were tightly correlated (Figure
+{[}{[}3B{]}{]}), suggesting that the null model only contains a noisy
+version of the same information that could be obtained more easily and
+interpretably with simple correlation coefficients.
 
-Whereas Figure 3 summarizes the accuracy of each model's point
-estimates, Figure 4 compares the inferential value of the Markov network
-versus the null model. When the ``true'' \(\beta\) coefficient is near
-zero, the Markov network's 95\% confidence intervals usually include
-zero (but not quite 95\% of the time), and it is rare for the model to
-confidently attribute the wrong sign to a coefficient (Figure 4A). The
-null model, by contrast, typically rejects the null hypothesis at the
-\(p=0.05\) level regardless of the underlying parameters, and the sign
-of its \(Z\)-scores is not a reliable indicator of the interaction's
-actual sign.
+Finally, we can evaluate the models' statistical inferences. Figure 4
+shows, across a range of true interaction strengths, the probability
+that the null model or the Markov network will predict the wrong sign of
+the interaction with 95\% confidence. The null model makes such errors
+nearly 8 times as often as the Markov network, even though it only
+reject the null hypothesis twice as often overall {[}{[}Appendix{]}{]}.
+The Markov network's errors were also more concentrated around 0, so it
+never misclassified strong interactions like the null model did.
 
@@
 cases where environmental heterogeneity and indirect interactions cause
-ecologists' typical null modeling approaches to reliably fail. For cases
-where fitting a Markov network is computationally infeasible, these
-results also indicate that partial covariances and generalized linear
-models can both provide useful approximations. The partial correlations'
-success on simulated data may not carry over to real data sets, however;
-Loh and Wainwright (2013) show that the linear approximations can be
-less reliable in cases where the true interaction matrix contains more
-structure (e.g.~guilds or trophic levels). Similarly, the approximation
-involved in using separate generalized linear models for each species
-can occasionally lead to catastrophic overfitting with small-to-moderate
-sample sizes (Lee and Hastie 2012). For these reasons, it will usually
-be best to fit a Markov network rather than one of the alternative
-methods when one's computational resources allow it.
+ecologists' typical null modeling approaches to reliably fail. Partial
+covariances and generalized linear models can both provide
+computationally efficient approximations, but with somewhat lower
+accuracy. The difference in accuracy may be larger for real data sets
+than for the simulated landscapes in Figure 3, however; linear
+approximations to the Markov network make larger errors when the
+interaction matrix is structured (e.g.~due to guilds or trophic levels;
+Loh and Wainwright 2013). Similarly, the separate generalized linear
+models for each species can severely overfit in some cases (Lee and
+Hastie 2012). The full Markov network should thus be preferred to the
+approximations when it is computationally feasible.
 
-It's important to note that none of these methods can identify the exact
-nature of the pairwise interactions (e.g.~which species in a
-positively-associated pair is facilitating the other; Schmidt and Murphy
-2012), particularly when real pairs of species can reciprocally
-influence one another in multiple ways simultaneously (Bruno et al.
-2003); with compositional data, there is only enough information to
-provide a single number describing each species pair. To estimate
-asymmetric interactions, such as commensalism or predation, ecologists
-would need other kinds of data, as from time series, behavioral
-observations, manipulative experiments, or natural history. These other
-sources of information could also be used to augment the likelihood
-function with an informative prior distribution, which could lead to
-better results on some real data sets than was shown in Figure 3A.
+Compositional data only has enough degrees of freedom to estimate one
+interaction per species pair (Schmidt and Murphy 2012), so none of these
+methods can identify the exact nature of the pairwise interactions
+{[}e.g.~which species in a positively-associated pair is facilitating
+the other{]}. To estimate asymmetric interactions, such as commensalism
+or predation, ecologists could use time series, behavioral observations,
+manipulative experiments, or natural history. These other sources of
+information could also be used to augment the likelihood function with
+an informative prior distribution, reducing ecologists' eror and
+uncertainty.
 
@@
 prevalence (cf Figure 2D). This kind of insight would have been
-difficult to obtain without a generative model that makes predictions
-about the consequences of species interactions; null models (which
-presume \emph{a priori} that interactions do not exist) have no way to
-make such predictions.
+difficult to obtain without a generative model that predicts the
+consequences of species interactions (as opposed to null simulations
+that omit interactions entirely).
 
@@
 networks---have been studied for nearly a century (Cipra 1987), and the
-models' properties, capabilities, and limits are well-understood. Using
-the same framework for species interactions would thus allow ecologists
-to tap into an enormous set of existing discoveries and techniques for
-dealing with indirect effects, stability, and alternative stable states.
-Numerous other extensions are possible: for example, the states of the
-interaction network can be modeled as a function of the local abiotic
-environment (Lee and Hastie 2012), which would provide a rigorous and
-straightforward approach to the difficult and important task of
-incorporating whole networks of biotic interactions into species
-distribution models (Pollock et al. 2014), leading to a better
+models are well-understood. Using the same framework for species
+interactions would thus allow ecologists to tap into a vast set of
+existing discoveries and techniques for dealing with indirect effects,
+stability, and alternative stable states. Numerous other extensions are
+possible: for example, the states of the interaction network can be
+modeled as a function of the local abiotic environment (Lee and Hastie
+2012), which would help incorporate networks of biotic interactions into
+species distribution models (Pollock et al. 2014) and lead to a better
 understanding of the interplay between biotic and abiotic effects on
 community structure. Alternatively, models could allow one species to
-affect the interactions between two other species (Tjelmeland and Besag
-1998), tipping the balance between facilitation and exploitation (Bruno
-et al. 2003).
+alter the relationship between two other species (Tjelmeland and Besag
+1998, cf Bruno et al. 2003).
 
@@
 interactions. Null and neutral models can be useful for clarifying our
-thinking about the numerical consequences of species' richness and
-abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
-deviations from a particular null model must be interpreted with care
-(Roughgarden 1983). Even in small networks with three species, it may
-simply not be possible to implicate individual species pairs or specific
-ecological processes like competition by rejecting a general-purpose
-null (Gotelli and Ulrich 2009), especially when the test statistic is
-effectively just a correlation coefficient {[}{[}Figure 3B{]}{]}. When a
-non-null backdrop is not controlled for, Type I error rates can
-skyrocket, the apparent sign of the interaction can change, and null
-models can lead to reliably incorrect inferences (Figure 1, Figure 4).
+thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
+particular null model must be interpreted with care (Roughgarden 1983).
+Even in small networks with three species, it may simply not be possible
+to implicate specific ecological processes like competition by rejecting
+a general-purpose null (Gotelli and Ulrich 2009), especially when the
+test statistic is effectively just a correlation coefficient
+{[}{[}Figure 3B{]}{]}. When a non-null backdrop is not controlled for,
+Type I error rates can skyrocket, the apparent sign of the interaction
+can change, and null models can lead to reliably incorrect inferences
+(Figure 1, Figure 4).
 
@@
 
-\noindent \textbf{Acknowledgements:} This research was funded by a
-Graduate Research Fellowship from the US National Science Foundation and
-benefited greatly from discussions with A. Sih, M. L. Baskett, R.
-McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor. Additionally, A.
-K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N. Golding,
-N. J. Gotelli, C. F. Dormann, and two anonymous reviewers provided very
-helpful feedback on the text.
+\noindent \textbf{Acknowledgements:} This work benefited greatly from
+discussions with A. Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A.
+C. Perry, and C. S. Tysor. Additionally, A. K. Barner, E. Baldridge, E.
+P. White, D. Li, D. L. Miller, N. Golding, N. J. Gotelli, C. F. Dormann,
+and two anonymous reviewers provided very helpful feedback on the text.
+This research was partially supported by a Graduate Research Fellowship
+from the US National Science Foundation and by the Gordon and Betty
+Moore Foundation's Data-Driven Discovery Initiative through Grant
+GBMF4563 to E. P. White.
 
@@
 
-\noindent \textbf{Figure 2: A.} A small Markov network with two species,
-defined by its \(\alpha\) and \(\beta\) values. The abiotic environment
-favors the occurrence of each species (\(\alpha >0\)), particularly
-species 2 (\(\alpha_2 > \alpha_1\)). The negative \(\beta\) coefficient
-linking these two species implies that they co-occur less than expected
-under independence. \textbf{B.} These coefficients determine the
-probabilities of all four possible presence-absence combinations for
-Species 1 and Species 2. \(\alpha_1\) is added to the exponent whenever
-Species 1 is present (\(y_1 = 1\)), but not when it is absent
-(\(y_1 = 0\)). Similarly, the exponent includes \(\alpha_2\) only when
-species \(2\) is present (\(y_2 = 1\)), and \(\beta\) only when both are
-present (\(y_1y_2 = 1\)). The normalizing constant \(Z\), ensures that
-the four probabilities sum to 1. In this case, \(Z\) is about 18.5.
-\textbf{C.} The expected frequencies of all possible co-occurrence
-patterns between the two species of interest, as calculated in the
-previous panel. \textbf{D.} If, as assumed in null model simulations,
+\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
+\(\alpha\) and \(\beta\) values. The abiotic environment favors the
+occurrence of each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_12\) coefficient is
+consistent with competition between the two species. \textbf{B.} The
+coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
+is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
+but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
+includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
+and includes \(\beta_12\) only when both are present (\(y_1y_2 = 1\)).
+The normalizing constant \(Z\), ensures that the four probabilities sum
+to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest, as calculated in the previous panel. \textbf{D.} If
 \(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
-the same resources), then the reduction in competition would allow each
-species to increase its overall occurrence rate.
+the same resources), then the each species' overall occurrence rates
+would increase.
 
-\noindent \textbf{Figure 3: A.} Proportion of variance in interaction
+\noindent \textbf{Figure 3:} Proportion of variance in interaction
 coefficients explained by each method versus number of sampled
-locations. \textbf{B.} The \(Z\)-scores produced by the null model
-(``Pairs'') for each pair of species can be predicted using the
-correlation between the presence-absence vectors of those same species
-and from the number of sites on the landscape.
+locations.
 

### 2015-10-31 Harris_2015/show-results.R

@@ -174,60 +174,124 @@ landscape_estimates = x %>%
   summarise(r2 = 1 - sum(resids^2) / sum(truth^2))
 
 
-summary(lmer(r2 ~ method + n_sites + simulation_type + (1|rep_name), data = landscape_estimates))
-
 ####
 
+summary(lmer(r2 ~ method + n_sites + simulation_type + (1|rep_name), data = landscape_estimates))
 
-xx = x[x$method == "Markov network" & grepl("^env[0-9]*$", x$rep_name), ]
-
-p_positive = predict(gam(I(lower > 0) ~ s(truth), family = binomial, data = xx), type = "response")
-p_negative = predict(gam(I(upper < 0) ~ s(truth), family = binomial, data = xx), type = "response")
-p_ns = predict(gam(I(lower < 0 & upper > 0) ~ s(truth), family = binomial, data = xx), type = "response")
-
-xx2 = x[x$method == "null" & grepl("^env[0-9]*$", x$rep_name), ]
-
-p_positive2 = predict(gam(I(estimate < pnorm(.025)) ~ s(truth), family = binomial, data = xx2), type = "response")
-p_negative2 = predict(gam(I(estimate > pnorm(.975)) ~ s(truth), family = binomial, data = xx2), type = "response")
-p_ns2 = predict(gam(I(abs(estimate) < pnorm(.975)) ~ s(truth), family = binomial, data = xx2), type = "response")
+pairs_summary = x[x$method == "null" & !grepl("pop", x$rep_name), ]
+markov_summary = x[x$method == "Markov network" & !grepl("pop", x$rep_name), ]
 
 
+# Pairs's Z score is based on C-scores, which are positive when species are 
+# disaggregated.  So significantly positive interactions == negative estimates
+pairs_summary$sig_pos = pairs_summary$estimate < qnorm(.025)
+pairs_summary$sig_neg = pairs_summary$estimate > qnorm(.975)
 
-pdf("manuscript-materials/figures/error_rates.pdf", width = 10, height = 5)
-par(mfrow = c(1, 2))
-plot(p_ns[order(xx$truth)] ~ sort(xx$truth), type = "l", lwd = 3, ylim = c(0, 1), yaxs = "i", bty = "l", ylab = "Proportion", las = 1, xlab = "\"True\" coefficient", xaxs = "i")
[…]

### 2015-11-01 Harris_2015/manuscript-materials/draft.Rmd
and another thing

@@
 
-### Introduction
+\noindent\textbf{Introduction}
 
@@
 
+\noindent
 To the extent that nontrophic species interactions (such as competition) affect
@@
 
-### Methods
+\noindent\textbf{Methods}
 
-**Markov networks.**
+\noindent
+\textbf{Markov networks.}
 Markov networks provide a framework for translating back and forth between the
@@
 
-### Results
+\noindent
+\textbf{Results}
 
@@
 models, and the two partial correlation-based methods) outperformed those
-that did not.  Of these, the Markov network consistently performed the best,
-especially on the smaller landscapes [which are also the most
-representative of typical analyses in this field; @gotelli_empirical_2009].
-As shown in Appendix [[X]], the standard errors associated with the estimates
-in Figure 3 are small (less than 0.01), so these differences should not be
-attributed to sampling error.
+that did not.  Of these, the Markov network consistently performed the best.
+As anticipated by @lee_learning_2012, the Markov network produced very
+similar estimates to the generalized linear models (Figure 4A), but with
+higher accuracy---especially with smaller data sets.  As reviewed in
+@gotelli_empirical_2009, most analyses in this field in ecology involve
+fewer than 50 sites, so this is the most important regime for model
+performance. As shown in Appendix [[X]], the standard errors associated
+with the estimates in Figure 3 are small (less than 0.01), so these
+differences should not be attributed to sampling error.
 
@@
 true interaction strengths than either the joint species distribution model
-(BayesComm) or the null model (Pairs). The estimates from these
-three approaches were tightly correlated (Figure [[3B]]), suggesting that the
-null model only contains a noisy version of the same information that could be
-obtained more easily and interpretably with simple correlation coefficients.
+(BayesComm) or the null model (Pairs). The estimates from these three approaches
+were tightly correlated, after controlling for the size of the landscape,
+suggesting that the null model only contains a noisy version of the
+same information that could be obtained more easily and interpretably with
+simple correlation coefficients (Figure 4B).
 
-Finally, we can evaluate the models' statistical inferences.  Figure 4 shows,
-across a range of true interaction strengths, the probability that the null model
-or the Markov network will predict the wrong sign of the interaction with 95%
-confidence.  The null model makes such errors nearly 8 times as often as the
-Markov network, even though it only reject the null hypothesis twice as often
-overall [[Appendix]].  The Markov network's errors were also more concentrated
-around 0, so it never misclassified strong interactions like the null model did.
+Finally, we can evaluate the models' statistical inferences (focusing on the
+first two simulation types, for which the true interaction rates are easiest to
+interpret).  The Markov network's Type I error rate (defined here as its
+probability of rejecting the null hypothesis for a pair of species where
+$|\beta_{ij}|<0.1$ depended on the simulation type: 0.03 for simulations that
+matched the model's assumptions, versus 0.15 for simulations that included
+environmental heterogeneity [[Appendix]].  For the null model, these rates were
+0.31 and 0.50, respectively. Figure 4C shows, across a range of true interaction
+strengths, the probability that the null model or the Markov network will
+predict the wrong sign of the interaction with 95% confidence.  The null model
+makes such errors nearly 8 times as often as the Markov network, even though it
+only reject the null hypothesis twice as often overall [[Appendix]].  The Markov
+network's errors were also more concentrated around 0, so it never misclassified
+strong interactions like the null model did.
 
-### Discussion
+\noindent
+\textbf{Discussion}
 
+\noindent
 The results presented above show that Markov networks can reliably recover
@@
 linear models can both provide computationally efficient approximations, but
-with somewhat lower accuracy. The difference in accuracy may be larger for real
-data sets than for the simulated landscapes in Figure 3, however; linear
-approximations to the Markov network make larger errors when the interaction
-matrix is structured [e.g. due to guilds or trophic levels;
+with somewhat lower accuracy, especially for typically-sized data sets with small
+numbers of sites [@gotelli_empirical_2009]. The difference in accuracy may be
+larger for real data sets than for the simulated landscapes in Figure 3,
+however; linear approximations to the Markov network make larger errors when
+the interaction matrix is structured [e.g. due to guilds or trophic levels;
 @loh_structure_2013]. Similarly, the separate generalized linear models
@@
 
-Compositional data only has enough degrees of freedom to estimate one interaction
-per species pair [@schmidt_modeling_2012], so none of these methods can identify
-the exact nature of the pairwise interactions [e.g. which species in a
-positively-associated pair is facilitating the other]. To estimate asymmetric
-interactions, such as commensalism or predation, ecologists could use time
-series, behavioral observations, manipulative experiments, or natural history.
-These other sources of information could also be used to augment the likelihood
-function with an informative prior distribution, reducing ecologists' eror and
-uncertainty.
+Compositional data only contains enough degrees of freedom to estimate one
+interaction per species pair, so none of these methods can identify the exact
+nature of the pairwise interactions [e.g. which species in a
+positively-associated pair is facilitating the other; cf
+@schmidt_modeling_2012]. To estimate asymmetric interactions, such as
+commensalism or predation, ecologists could use time series, behavioral
+observations, manipulative experiments, or natural history. These other sources
+of information could also be used to augment the likelihood function with an
+informative prior distribution, reducing ecologists' error and uncertainty.
 
@@
 null [@gotelli_empirical_2009], especially when the test statistic is
-effectively just a correlation coefficient [[Figure 3B]]. When a non-null
+effectively just a correlation coefficient Figure 4A. When a non-null
 backdrop is not controlled for, Type I error rates can skyrocket, the apparent
@@
 
-\noindent \textbf{Figure 4: A.} Inferences from the Markov network's 95%
-confidence intervals on the "no-environment" simulated landscapes, as a function
-of the "true" coefficient values. **B.** Inferences from the $Z$-statistics
-estimated by the "Pairs" software for null modeling.
+\noindent \textbf{Figure 4: A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates. **B.** The null
+model's estimates typically matched the correlation coefficient, after
+controlling for landscape size. **C.** For any given interaction strength, the
+null model was much more likely to misclassify its sign with 95% confidence than
+the Markov network was.
 
-**References:**
+\noindent \textbf{References:}
 

### 2015-11-01 Harris_2015/manuscript-materials/draft.tex
and another thing

@@
 \usepackage{ifthen}
-\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+\geometry{verbose,letterpaper,tmargin=2.5cm, bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
 \ifxetex
@@
 
-\subsubsection{Introduction}\label{introduction}
+\noindent\textbf{Introduction}
 
@@
 
+\noindent
 To the extent that nontrophic species interactions (such as competition)
@@
 
-\subsubsection{Methods}\label{methods}
+\noindent\textbf{Methods}
 
+\noindent
 \textbf{Markov networks.} Markov networks provide a framework for
@@
 
-\subsubsection{Results}\label{results}
+\noindent
+\textbf{Results}
 
@@
 correlation-based methods) outperformed those that did not. Of these,
-the Markov network consistently performed the best, especially on the
-smaller landscapes (which are also the most representative of typical
-analyses in this field; Gotelli and Ulrich 2009). As shown in Appendix
-{[}{[}X{]}{]}, the standard errors associated with the estimates in
-Figure 3 are small (less than 0.01), so these differences should not be
-attributed to sampling error.
+the Markov network consistently performed the best. As anticipated by
+Lee and Hastie (2012), the Markov network produced very similar
+estimates to the generalized linear models (Figure 4A), but with higher
+accuracy---especially with smaller data sets. As reviewed in Gotelli and
+Ulrich (2009), most analyses in this field in ecology involve fewer than
+50 sites, so this is the most important regime for model performance. As
+shown in Appendix {[}{[}X{]}{]}, the standard errors associated with the
+estimates in Figure 3 are small (less than 0.01), so these differences
+should not be attributed to sampling error.
 
@@
 distribution model (BayesComm) or the null model (Pairs). The estimates
-from these three approaches were tightly correlated (Figure
-{[}{[}3B{]}{]}), suggesting that the null model only contains a noisy
-version of the same information that could be obtained more easily and
-interpretably with simple correlation coefficients.
+from these three approaches were tightly correlated, after controlling
+for the size of the landscape, suggesting that the null model only
+contains a noisy version of the same information that could be obtained
+more easily and interpretably with simple correlation coefficients
+(Figure 4B).
 
-Finally, we can evaluate the models' statistical inferences. Figure 4
-shows, across a range of true interaction strengths, the probability
-that the null model or the Markov network will predict the wrong sign of
-the interaction with 95\% confidence. The null model makes such errors
-nearly 8 times as often as the Markov network, even though it only
-reject the null hypothesis twice as often overall {[}{[}Appendix{]}{]}.
-The Markov network's errors were also more concentrated around 0, so it
-never misclassified strong interactions like the null model did.
+Finally, we can evaluate the models' statistical inferences (focusing on
+the first two simulation types, for which the true interaction rates are
+easiest to interpret). The Markov network's Type I error rate (defined
+here as its probability of rejecting the null hypothesis for a pair of
+species where \(|\beta_{ij}|<0.1\) depended on the simulation type: 0.03
+for simulations that matched the model's assumptions, versus 0.15 for
+simulations that included environmental heterogeneity
+{[}{[}Appendix{]}{]}. For the null model, these rates were 0.31 and
+0.50, respectively. Figure 4C shows, across a range of true interaction
+strengths, the probability that the null model or the Markov network
+will predict the wrong sign of the interaction with 95\% confidence. The
+null model makes such errors nearly 8 times as often as the Markov
+network, even though it only reject the null hypothesis twice as often
+overall {[}{[}Appendix{]}{]}. The Markov network's errors were also more
+concentrated around 0, so it never misclassified strong interactions
+like the null model did.
 
-\subsubsection{Discussion}\label{discussion}
+\noindent
+\textbf{Discussion}
 
+\noindent
 The results presented above show that Markov networks can reliably
@@
 computationally efficient approximations, but with somewhat lower
-accuracy. The difference in accuracy may be larger for real data sets
-than for the simulated landscapes in Figure 3, however; linear
-approximations to the Markov network make larger errors when the
-interaction matrix is structured (e.g.~due to guilds or trophic levels;
-Loh and Wainwright 2013). Similarly, the separate generalized linear
-models for each species can severely overfit in some cases (Lee and
-Hastie 2012). The full Markov network should thus be preferred to the
-approximations when it is computationally feasible.
+accuracy, especially for typically-sized data sets with small numbers of
+sites (Gotelli and Ulrich 2009). The difference in accuracy may be
+larger for real data sets than for the simulated landscapes in Figure 3,
+however; linear approximations to the Markov network make larger errors
+when the interaction matrix is structured (e.g.~due to guilds or trophic
+levels; Loh and Wainwright 2013). Similarly, the separate generalized
+linear models for each species can severely overfit in some cases (Lee
+and Hastie 2012). The full Markov network should thus be preferred to
+the approximations when it is computationally feasible.
 
-Compositional data only has enough degrees of freedom to estimate one
-interaction per species pair (Schmidt and Murphy 2012), so none of these
-methods can identify the exact nature of the pairwise interactions
-{[}e.g.~which species in a positively-associated pair is facilitating
-the other{]}. To estimate asymmetric interactions, such as commensalism
+Compositional data only contains enough degrees of freedom to estimate
+one interaction per species pair, so none of these methods can identify
+the exact nature of the pairwise interactions (e.g.~which species in a
+positively-associated pair is facilitating the other; cf Schmidt and
+Murphy 2012). To estimate asymmetric interactions, such as commensalism
 or predation, ecologists could use time series, behavioral observations,
@@
 information could also be used to augment the likelihood function with
-an informative prior distribution, reducing ecologists' eror and
+an informative prior distribution, reducing ecologists' error and
 uncertainty.
@@
 a general-purpose null (Gotelli and Ulrich 2009), especially when the
-test statistic is effectively just a correlation coefficient
-{[}{[}Figure 3B{]}{]}. When a non-null backdrop is not controlled for,
-Type I error rates can skyrocket, the apparent sign of the interaction
-can change, and null models can lead to reliably incorrect inferences
-(Figure 1, Figure 4).
+test statistic is effectively just a correlation coefficient Figure 4A.
+When a non-null backdrop is not controlled for, Type I error rates can
+skyrocket, the apparent sign of the interaction can change, and null
+models can lead to reliably incorrect inferences (Figure 1, Figure 4).
 
@@
 
-\noindent \textbf{Figure 4: A.} Inferences from the Markov network's
-95\% confidence intervals on the ``no-environment'' simulated
-landscapes, as a function of the ``true'' coefficient values.
-\textbf{B.} Inferences from the \(Z\)-statistics estimated by the
-``Pairs'' software for null modeling.
+\noindent \textbf{Figure 4: A.} The Markov network's estimated
+interaction coefficients were generally very similar to the GLM
+estimates. \textbf{B.} The null model's estimates typically matched the
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was.
 
-\textbf{References:}
+\noindent \textbf{References:}
 

### 2015-11-01 Harris_2015/show-results.R

@@ -133,10 +133,11 @@ result_summary %>%
 
 
 pdf("manuscript-materials/figures/performance.pdf", width = 7.5, height = 2.5)
-ggplot(result_summary, aes(x = n_sites, y = r2, col = method)) + 
+ggplot(result_summary, aes(x = n_sites, y = r2, col = method, shape = method)) + 
   facet_grid(~simulation_type) + 
   geom_line(size = .5) + 
-  geom_point(size = 2, fill = "white") + 
+  geom_point(size = 2.5, fill = "white") + 
+  scale_shape_manual(values = c(16, 22, 17, 23, 18, 24, 15)) +
   geom_hline(yintercept = 0, size = 1/2) +
   geom_vline(xintercept = 0, size = 1) + 
   #scale_shape_manual(values = c(21, 25, 15, 18)) + 
@@ -193,6 +194,18 @@ markov_summary$sig_pos = markov_summary$lower > 0
 markov_summary$sig_neg = markov_summary$upper < 0
 
 
+# Compare 
+null_cor = x %>%
+  dplyr::select(-lower, -upper, -X) %>%
+  spread(method, estimate) %>%
+  na.omit()
+
+# R-squared
+round(summary(lm(null ~ I(correlation*sqrt(n_sites)), data = null_cor))$r.squared, 2)
+
+
+
+
 my_geom_smooth = function(data, color, ...){
   geom_smooth(
     data = data,
@@ -211,7 +224,7 @@ my_geom_smooth = function(data, color, ...){
 }
 
 
-truth_seq = seq(min(data$truth), max(data$truth), length = 1000)
+truth_seq = seq(min(markov_summary$truth), max(markov_summary$truth), length = 1000)
 
 plotfun = function(..., add){
   if(add){
@@ -238,34 +251,23 @@ y_pairs = error_smoother(pairs_summary)
 y_markov = error_smoother(markov_summary)
 
 
-null_cor = x %>%
-  dplyr::select(-lower, -upper, -X) %>%
-  spread(method, estimate) %>%
-  na.omit()
-
-# R-squared
-round(summary(lm(null ~ I(correlation*sqrt(n_sites)), data = null_cor))$r.squared, 2)
-
-
-pdf("manuscript-materials/figures/error_rates.pdf", height = 8, width = 4)
-par(mfrow = c(2, 1))
-plotfun(
-  truth_seq,
-  y_pairs,
-  type = "l",
[…]

### 2015-11-02 Harris_2015/manuscript-materials/draft.Rmd
etc

@@
 
-**Title:** Estimating species interactions from co-occurrence data with Markov
+**Title:** Inferring species interactions from co-occurrence data with Markov
 networks
@@
 
-**Abstract:** Estimating species interactions from co-occurrence data is one of
+**Abstract:** Inferring species interactions from co-occurrence data is one of
 the most controversial tasks in community ecology. One difficulty is that a
@@
 after controlling for the other species in the network (Figure 2A, Figure 2B).
-For example, if $\beta_{ij} = -2$, then each species' odds of occurrence are
-$exp{2}$ times lower when the other one is present (as compared with otherwise
-comparable sites). Because the relative probability of a presence-absence vector
+For example, if $\beta_{ij} = +2$, then each species' odds of occurrence would be
+$e^2$ times higher when the other one is present (as compared with otherwise
+equivalent sites). Because the relative probability of a presence-absence vector
 increases when positively-associated species co-occur and decreases when
@@
 presence of one competitor is associated with release from another competitor),
-then species' overall co-occurrence patterns can be more complicated, and may
-require summing over the different possible assemblages (Figure 2B) to determine
-the overall outcomes.
+then predicting species' overall co-occurrence rates can be more complicated,
+and may require summing over the different possible assemblages (Figure 2B).
 
@@
 \noindent \textbf{Simulated landscapes.}
-I simulated several sets of landscapes using known parameters as a test bed for
-different statistical methods. The first set of landscapes included the three
-competing species shown in Figure 1. For each of 1000 replicates, I generated a
-landscape with 100 sites by sampling from a probability distribution defined by
-the figure's interaction coefficients (Appendix 1). Each of the methods
-described below was then evaluated on its ability to correctly infer that the
-two shrub species competed with one another, despite their frequent co-occurrence.
+I simulated several sets of landscapes using known parameters to evaluate
+different statistical methods' performance. The first set of landscapes included
+the three competing species shown in Figure 1. For each of 1000 replicates, I
+generated a landscape with 100 sites by sampling from a probability distribution
+defined by the figure's interaction coefficients (Appendix 1). Each of the
+methods described below was then evaluated on its ability to correctly infer
+that the two shrub species competed with one another, despite their frequent
+co-occurrence.
 
@@
 that did not.  Of these, the Markov network consistently performed the best.
-As anticipated by @lee_learning_2012, the Markov network produced very
-similar estimates to the generalized linear models (Figure 4A), but with
-higher accuracy---especially with smaller data sets.  As reviewed in
-@gotelli_empirical_2009, most analyses in this field in ecology involve
-fewer than 50 sites, so this is the most important regime for model
-performance. As shown in Appendix [[X]], the standard errors associated
-with the estimates in Figure 3 are small (less than 0.01), so these
-differences should not be attributed to sampling error.
+As anticipated by @lee_learning_2012, generalized linear models closely
+approximated the Markov network estimates, especially when the data sets were
+very large. As reviewed in @gotelli_empirical_2009, however, most analyses in
+this field involve fewer than 50 sites, so model performance on the smaller
+landscapes where the gap was larger is most important. As shown in Appendix
+[[X]], the standard errors associated with the estimates in Figure 3 are small
+(less than 0.01), so the differences among methods should not be attributed to
+sampling error.
 
@@
 true interaction strengths than either the joint species distribution model
-(BayesComm) or the null model (Pairs). The estimates from these three approaches
-were tightly correlated, after controlling for the size of the landscape,
+(BayesComm) or the null model (Pairs). The estimates from these approaches
+were tightly correlated (after controlling for the size of the landscape)
 suggesting that the null model only contains a noisy version of the
@@
 first two simulation types, for which the true interaction rates are easiest to
-interpret).  The Markov network's Type I error rate (defined here as its
-probability of rejecting the null hypothesis for a pair of species where
-$|\beta_{ij}|<0.1$ depended on the simulation type: 0.03 for simulations that
+interpret).  The Markov network's approximate Type I error rate (defined here as
+its probability of rejecting the null hypothesis for a pair of species where
+$|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for simulations that
 matched the model's assumptions, versus 0.15 for simulations that included
 environmental heterogeneity [[Appendix]].  For the null model, these rates were
-0.31 and 0.50, respectively. Figure 4C shows, across a range of true interaction
-strengths, the probability that the null model or the Markov network will
-predict the wrong sign of the interaction with 95% confidence.  The null model
-makes such errors nearly 8 times as often as the Markov network, even though it
-only reject the null hypothesis twice as often overall [[Appendix]].  The Markov
-network's errors were also more concentrated around 0, so it never misclassified
-strong interactions like the null model did.
+0.31 and 0.50, respectively---far higher than the nominal 0.05 rate. Figure 4C
+shows, across a range of true interaction strengths, the probability that the
+null model or the Markov network will predict the wrong sign of the interaction
+with 95% confidence.  The null model makes such errors nearly 8 times as often
+as the Markov network, even though it only reject the null hypothesis twice as
+often overall [[Appendix]].  The Markov network's errors were also more
+concentrated around 0, so it never misclassified strong interactions like the
+null model did.
 
@@
 full Markov network should thus be preferred to the approximations when
-it is computationally feasible.
+it is computationally tractable.
 
 Compositional data only contains enough degrees of freedom to estimate one
-interaction per species pair, so none of these methods can identify the exact
-nature of the pairwise interactions [e.g. which species in a
-positively-associated pair is facilitating the other; cf
-@schmidt_modeling_2012]. To estimate asymmetric interactions, such as
-commensalism or predation, ecologists could use time series, behavioral
-observations, manipulative experiments, or natural history. These other sources
-of information could also be used to augment the likelihood function with an
-informative prior distribution, reducing ecologists' error and uncertainty.
+interaction per species pair [@schmidt_modeling_2012], so none of these methods
+can identify the exact nature of the pairwise interactions (e.g. which species
+in a positively-associated pair is facilitating the other). To estimate
+asymmetric interactions, such as commensalism or predation, ecologists could use
+time series, behavioral observations, manipulative experiments, or natural
+history. These other sources of information could also be used to augment the
+likelihood function with an informative prior distribution, reducing ecologists'
+error and uncertainty relative to Figure 3's results.
 
 Despite their limitations, Markov networks have enormous potential to improve
-ecological understanding. In particular, they are less vulnerable than some
-of the most commonly-used methods to mistakenly identifying positive species
-interactions between competing species, and can make precise statements about
-the conditions where indirect interactions will overwhelm direct ones. They also
-provide a simple answer to the question of how competition should affect a
-species' overall prevalence, which was a major flashpoint for the null model
-debates in the 1980’s [@roughgarden_competition_1983; @strong_ecological_1984].
-Equation 1 can be used to calculate the expected prevalence of a species in the
-absence of biotic influences [$\frac{e^\alpha}{1 + e^{\alpha}}$; @lee_learning_2012].
-Competition's effect on prevalence in a Markov network can then be calculated by
-subtracting this value from the observed prevalence (cf Figure 2D). This kind of
-insight would have been difficult to obtain without a generative model that
-predicts the consequences of species interactions (as opposed to null
-simulations that omit interactions entirely).
+ecological understanding. In particular, they make many fewer errors than existing
+approaches, and can make precise statements about the conditions where indirect
+interactions will overwhelm direct ones. They also provide a simple answer to
+the question of how competition should affect a species' overall prevalence,
+which has important implications for community-level modeling
+[@strong_ecological_1984]. Specifically, Equation 1 can be used to calculate the
+expected prevalence of a species in the absence of biotic influences as
+$e^\alpha/(e^{0} + e^\alpha)$. Competition's effect on prevalence can then be
+estimated by comparing this value with the observed prevalence (e.g. comparing
+Figure 2D with Figure 2C). This novel quantitative result conflicts with most of
+our null models, which unreasonably assume that prevalence would be the exactly
+same in the absence of competition as it is in the observed data
+[@roughgarden_competition_1983].
 
-Markov networks---particularly the Ising model for binary networks---have been
-studied for nearly a century [@cipra_introduction_1987], and the models are
-well-understood. Using the same framework for species interactions would thus
-allow ecologists to tap into a vast set of existing discoveries and techniques
-for dealing with indirect effects, stability, and alternative stable states.
-Numerous other extensions are possible: for example, the states of the
-interaction network can be modeled as a function of the local abiotic
-environment [@lee_learning_2012], which would help incorporate networks of
-biotic interactions into species distribution models
+Markov networks---particularly the Ising model for binary networks---are very
+well understood, having been studied for nearly a century
+[@cipra_introduction_1987]. Tapping into this framework would thus allow
+ecologists to take advantage of into a vast set of existing discoveries and
+techniques for dealing with indirect effects, stability, and alternative stable
+states. Numerous extensions to the basic network are possible as well. For
+example, the states of the interaction network can be modeled as a function of
+the local abiotic environment [@lee_learning_2012], which would help incorporate
+networks of biotic interactions into species distribution models
 [@pollock_understanding_2014] and lead to a better understanding of the
@@
 null [@gotelli_empirical_2009], especially when the test statistic is
-effectively just a correlation coefficient Figure 4A. When a non-null
+effectively just a correlation coefficient (Figure 4B). When the non-null
 backdrop is not controlled for, Type I error rates can skyrocket, the apparent
-sign of the interaction can change, and null models can lead to reliably
-incorrect inferences (Figure 1, Figure 4).
+sign of the interaction can change, and null models can routinely produce
+misleading inferences (Figure 1, Figure 4C, @gotelli_empirical_2009).
 
@@
 that the models' relative performance on real data sets is similar to the range
-of results shown in Figure 3A, scientists in this field could more than double
+of results shown in Figure 3, scientists in this field could more than double
 their explanatory power by switching from null models to Markov networks (or
@@
 
-\noindent \textbf{Figure 1: A.} A small network of three competing species. The tree (top) tends not to
-co-occur with either of the two shrub species, as indicated by the strongly
-negative coefficient linking them. The two shrub species also compete with one
-another, but more weakly (circled coefficient). **B.** In spite of the
-competitive interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors positively
-correlated (circled). **C.** Controlling for the tree species' presence with a
-conditional (all-else-equal) approach such as a partial covariance or a
-Markov network leads to correct identification of the negative shrub-shrub
-interaction (circled).
+\noindent \textbf{Figure 1: A.} A small network of three competing species.
+The tree (top) tends not to co-occur with either of the two shrub species, as
+indicated by the strongly negative coefficient linking them. The two shrub
+species also compete with one another, but more weakly (circled coefficient).
+**B.** In spite of the competitive interactions between the two shrub species,
+their shared tendency to occur in locations without trees makes their occurrence
+vectors positively correlated (circled). **C.** Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a Markov
+network leads to correct identification of the negative shrub-shrub interaction
+(circled).
 
@@
 ($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
-$\beta_12$ coefficient is consistent with competition between the two species.
+$\beta_{12}$ coefficient is consistent with competition between the two species.
 **B.** The coefficients determine the probabilities of all four possible
@@
 absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
-species $2$ is present ($y_2 = 1$), and includes $\beta_12$ only when both are
+species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
 present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
@@
 frequencies of all possible co-occurrence patterns between the two species of
-interest, as calculated in the previous panel. **D.** If $\beta_{12}$ equaled
-zero (e.g. if the species no longer competed for the same resources), then the
-each species' overall occurrence rates would increase.
+interest, as calculated in the previous panel. **D.** Without competition (i.e.
+with $\beta_{12}=0$, each species would occur more often.
 
 \noindent \textbf{Figure 3:} Proportion of variance in interaction
-coefficients explained by each method versus number of sampled locations.
+coefficients explained by each method versus number of sampled locations across
+the three simulation types.
 
@@
 coefficients were generally very similar to the GLM estimates. **B.** The null
-model's estimates typically matched the correlation coefficient, after
-controlling for landscape size. **C.** For any given interaction strength, the
-null model was much more likely to misclassify its sign with 95% confidence than
-the Markov network was.
+model's estimates typically matched the (negative) correlation coefficient,
+after controlling for landscape size. **C.** For any given interaction strength,
+the null model was much more likely to misclassify its sign with 95% confidence
+than the Markov network was.
 

### 2015-11-02 Harris_2015/manuscript-materials/draft.tex
etc

@@
 
-\textbf{Title:} Estimating species interactions from co-occurrence data
+\textbf{Title:} Inferring species interactions from co-occurrence data
 with Markov networks
@@
 
-\textbf{Abstract:} Estimating species interactions from co-occurrence
+\textbf{Abstract:} Inferring species interactions from co-occurrence
 data is one of the most controversial tasks in community ecology. One
@@
 controlling for the other species in the network (Figure 2A, Figure 2B).
-For example, if \(\beta_{ij} = -2\), then each species' odds of
-occurrence are \(exp{2}\) times lower when the other one is present (as
-compared with otherwise comparable sites). Because the relative
+For example, if \(\beta_{ij} = +2\), then each species' odds of
+occurrence would be \(e^2\) times higher when the other one is present
+(as compared with otherwise equivalent sites). Because the relative
 probability of a presence-absence vector increases when
@@
 presence of one competitor is associated with release from another
-competitor), then species' overall co-occurrence patterns can be more
-complicated, and may require summing over the different possible
-assemblages (Figure 2B) to determine the overall outcomes.
+competitor), then predicting species' overall co-occurrence rates can be
+more complicated, and may require summing over the different possible
+assemblages (Figure 2B).
 
@@
 \noindent \textbf{Simulated landscapes.} I simulated several sets of
-landscapes using known parameters as a test bed for different
-statistical methods. The first set of landscapes included the three
+landscapes using known parameters to evaluate different statistical
+methods' performance. The first set of landscapes included the three
 competing species shown in Figure 1. For each of 1000 replicates, I
@@
 the Markov network consistently performed the best. As anticipated by
-Lee and Hastie (2012), the Markov network produced very similar
-estimates to the generalized linear models (Figure 4A), but with higher
-accuracy---especially with smaller data sets. As reviewed in Gotelli and
-Ulrich (2009), most analyses in this field in ecology involve fewer than
-50 sites, so this is the most important regime for model performance. As
-shown in Appendix {[}{[}X{]}{]}, the standard errors associated with the
-estimates in Figure 3 are small (less than 0.01), so these differences
-should not be attributed to sampling error.
+Lee and Hastie (2012), generalized linear models closely approximated
+the Markov network estimates, especially when the data sets were very
+large. As reviewed in Gotelli and Ulrich (2009), however, most analyses
+in this field involve fewer than 50 sites, so model performance on the
+smaller landscapes where the gap was larger is most important. As shown
+in Appendix {[}{[}X{]}{]}, the standard errors associated with the
+estimates in Figure 3 are small (less than 0.01), so the differences
+among methods should not be attributed to sampling error.
 
@@
 distribution model (BayesComm) or the null model (Pairs). The estimates
-from these three approaches were tightly correlated, after controlling
-for the size of the landscape, suggesting that the null model only
-contains a noisy version of the same information that could be obtained
-more easily and interpretably with simple correlation coefficients
-(Figure 4B).
+from these approaches were tightly correlated (after controlling for the
+size of the landscape) suggesting that the null model only contains a
+noisy version of the same information that could be obtained more easily
+and interpretably with simple correlation coefficients (Figure 4B).
 
@@
 the first two simulation types, for which the true interaction rates are
-easiest to interpret). The Markov network's Type I error rate (defined
-here as its probability of rejecting the null hypothesis for a pair of
-species where \(|\beta_{ij}|<0.1\) depended on the simulation type: 0.03
-for simulations that matched the model's assumptions, versus 0.15 for
-simulations that included environmental heterogeneity
-{[}{[}Appendix{]}{]}. For the null model, these rates were 0.31 and
-0.50, respectively. Figure 4C shows, across a range of true interaction
-strengths, the probability that the null model or the Markov network
-will predict the wrong sign of the interaction with 95\% confidence. The
-null model makes such errors nearly 8 times as often as the Markov
-network, even though it only reject the null hypothesis twice as often
-overall {[}{[}Appendix{]}{]}. The Markov network's errors were also more
+easiest to interpret). The Markov network's approximate Type I error
+rate (defined here as its probability of rejecting the null hypothesis
+for a pair of species where \(|\beta_{ij}|<0.1\)) depended on the
+simulation type: 0.03 for simulations that matched the model's
+assumptions, versus 0.15 for simulations that included environmental
+heterogeneity {[}{[}Appendix{]}{]}. For the null model, these rates were
+0.31 and 0.50, respectively---far higher than the nominal 0.05 rate.
+Figure 4C shows, across a range of true interaction strengths, the
+probability that the null model or the Markov network will predict the
+wrong sign of the interaction with 95\% confidence. The null model makes
+such errors nearly 8 times as often as the Markov network, even though
+it only reject the null hypothesis twice as often overall
+{[}{[}Appendix{]}{]}. The Markov network's errors were also more
 concentrated around 0, so it never misclassified strong interactions
@@
 and Hastie 2012). The full Markov network should thus be preferred to
-the approximations when it is computationally feasible.
+the approximations when it is computationally tractable.
 
 Compositional data only contains enough degrees of freedom to estimate
-one interaction per species pair, so none of these methods can identify
-the exact nature of the pairwise interactions (e.g.~which species in a
-positively-associated pair is facilitating the other; cf Schmidt and
-Murphy 2012). To estimate asymmetric interactions, such as commensalism
-or predation, ecologists could use time series, behavioral observations,
+one interaction per species pair (Schmidt and Murphy 2012), so none of
+these methods can identify the exact nature of the pairwise interactions
+(e.g.~which species in a positively-associated pair is facilitating the
+other). To estimate asymmetric interactions, such as commensalism or
+predation, ecologists could use time series, behavioral observations,
 manipulative experiments, or natural history. These other sources of
@@
 an informative prior distribution, reducing ecologists' error and
-uncertainty.
+uncertainty relative to Figure 3's results.
 
 Despite their limitations, Markov networks have enormous potential to
-improve ecological understanding. In particular, they are less
-vulnerable than some of the most commonly-used methods to mistakenly
-identifying positive species interactions between competing species, and
-can make precise statements about the conditions where indirect
-interactions will overwhelm direct ones. They also provide a simple
-answer to the question of how competition should affect a species'
-overall prevalence, which was a major flashpoint for the null model
-debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
-can be used to calculate the expected prevalence of a species in the
-absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
-and Hastie 2012). Competition's effect on prevalence in a Markov network
-can then be calculated by subtracting this value from the observed
-prevalence (cf Figure 2D). This kind of insight would have been
-difficult to obtain without a generative model that predicts the
-consequences of species interactions (as opposed to null simulations
-that omit interactions entirely).
+improve ecological understanding. In particular, they make many fewer
+errors than existing approaches, and can make precise statements about
+the conditions where indirect interactions will overwhelm direct ones.
+They also provide a simple answer to the question of how competition
+should affect a species' overall prevalence, which has important
+implications for community-level modeling (Strong et al. 1984).
+Specifically, Equation 1 can be used to calculate the expected
+prevalence of a species in the absence of biotic influences as
+\(e^\alpha/(e^{0} + e^\alpha)\). Competition's effect on prevalence can
+then be estimated by comparing this value with the observed prevalence
+(e.g.~comparing Figure 2D with Figure 2C). This novel quantitative
+result conflicts with most of our null models, which unreasonably assume
+that prevalence would be the exactly same in the absence of competition
+as it is in the observed data (Roughgarden 1983).
 
-Markov networks---particularly the Ising model for binary
-networks---have been studied for nearly a century (Cipra 1987), and the
-models are well-understood. Using the same framework for species
-interactions would thus allow ecologists to tap into a vast set of
-existing discoveries and techniques for dealing with indirect effects,
-stability, and alternative stable states. Numerous other extensions are
-possible: for example, the states of the interaction network can be
-modeled as a function of the local abiotic environment (Lee and Hastie
-2012), which would help incorporate networks of biotic interactions into
-species distribution models (Pollock et al. 2014) and lead to a better
+Markov networks---particularly the Ising model for binary networks---are
+very well understood, having been studied for nearly a century (Cipra
+1987). Tapping into this framework would thus allow ecologists to take
+advantage of into a vast set of existing discoveries and techniques for
+dealing with indirect effects, stability, and alternative stable states.
+Numerous extensions to the basic network are possible as well. For
+example, the states of the interaction network can be modeled as a
+function of the local abiotic environment (Lee and Hastie 2012), which
+would help incorporate networks of biotic interactions into species
+distribution models (Pollock et al. 2014) and lead to a better
 understanding of the interplay between biotic and abiotic effects on
@@
 a general-purpose null (Gotelli and Ulrich 2009), especially when the
-test statistic is effectively just a correlation coefficient Figure 4A.
-When a non-null backdrop is not controlled for, Type I error rates can
-skyrocket, the apparent sign of the interaction can change, and null
-models can lead to reliably incorrect inferences (Figure 1, Figure 4).
+test statistic is effectively just a correlation coefficient (Figure
+4B). When the non-null backdrop is not controlled for, Type I error
+rates can skyrocket, the apparent sign of the interaction can change,
+and null models can routinely produce misleading inferences (Figure 1,
+Figure 4C, Gotelli and Ulrich (2009)).
 
@@
 extent that the models' relative performance on real data sets is
-similar to the range of results shown in Figure 3A, scientists in this
+similar to the range of results shown in Figure 3, scientists in this
 field could more than double their explanatory power by switching from
@@
 occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for the tree
-species' presence with a conditional (all-else-equal) approach such as a
-partial covariance or a Markov network leads to correct identification
-of the negative shrub-shrub interaction (circled).
+positively correlated (circled). \textbf{C.} Controlling for trees with
+a conditional (all-else-equal) approach such as a partial covariance or
+a Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled).
 
@@
 occurrence of each species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta_12\) coefficient is
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
 consistent with competition between the two species. \textbf{B.} The
@@
 includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
-and includes \(\beta_12\) only when both are present (\(y_1y_2 = 1\)).
+and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
 The normalizing constant \(Z\), ensures that the four probabilities sum
@@
 frequencies of all possible co-occurrence patterns between the two
-species of interest, as calculated in the previous panel. \textbf{D.} If
-\(\beta_{12}\) equaled zero (e.g.~if the species no longer competed for
-the same resources), then the each species' overall occurrence rates
-would increase.
+species of interest, as calculated in the previous panel. \textbf{D.}
+Without competition (i.e.~with \(\beta_{12}=0\), each species would
+occur more often.
 
 \noindent \textbf{Figure 3:} Proportion of variance in interaction
-coefficients explained by each method versus number of sampled
-locations.
+coefficients explained by each method versus number of sampled locations
+across the three simulation types.
 
@@
 estimates. \textbf{B.} The null model's estimates typically matched the
-correlation coefficient, after controlling for landscape size.
-\textbf{C.} For any given interaction strength, the null model was much
-more likely to misclassify its sign with 95\% confidence than the Markov
-network was.
+(negative) correlation coefficient, after controlling for landscape
+size. \textbf{C.} For any given interaction strength, the null model was
+much more likely to misclassify its sign with 95\% confidence than the
+Markov network was.
 

### 2015-11-02 Harris_2015/show-results.R

@@ -131,10 +131,15 @@ result_summary %>%
   summarise(round(mean(r2), 3))
 
 
+result_summary$simulation_type_long = plyr::revalue(result_summary$simulation_type,
+                                               c(no_env = "constant environment",
+                                                 env = "heterogeneous environment",
+                                                 pop = "population dynamics"))
+
 
 pdf("manuscript-materials/figures/performance.pdf", width = 7.5, height = 2.5)
 ggplot(result_summary, aes(x = n_sites, y = r2, col = method, shape = method)) + 
-  facet_grid(~simulation_type) + 
+  facet_grid(~simulation_type_long) + 
   geom_line(size = .5) + 
   geom_point(size = 2.5, fill = "white") + 
   scale_shape_manual(values = c(16, 22, 17, 23, 18, 24, 15)) +

### 2015-11-02 Harris_2015/manuscript-materials/draft.Rmd
fix appendices, I think

@@
-\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
+\chead{\ifthenelse{\value{page}=1}{}{\textit{Inferring species interactions}}}
 \renewcommand{\headrulewidth}{0pt}
@@
 1600 sites using three increasingly complex models (50 replicates for each
-combination of size and model). As described in Appendix [[X]], I randomly drew
+combination of size and model). As described in Appendix 2, I randomly drew
 the “true” coefficient values for each replicate landscape
@@
 @connor_checkered_2013].  I generated the final set of landscapes by simulating
-the population dynamics of 20 interacting species, as described in Appendix [[X]].
+the population dynamics of 20 interacting species, as described in Appendix 3.
 
@@
 I compared seven techniques for determining the sign and strength of the
-associations between pairs of species from simulated data (Appendix 3).
+associations between pairs of species from simulated data (Appendix 4).
 First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
@@
 regularizer (equivalent to a logistic prior with scale 2) to ensure that the
-model estimates were always finite [[Appendix]].
+model estimates were always finite.
 

### 2015-11-05 Harris_2015/manuscript-materials/response-to-editor.md
add fakedata and updates

Dear Mr. Harris,

Thank you very much for submitting your revised manuscript "Estimating species interactions from observational data with Markov networks" (Statistical Report) for review by Ecology. Many aspects of the presentation are improved. Unfortunately my assessment is that the new community dynamics simulations turned out to be a step in the wrong direction, making it harder rather than easier to assess the new method and compare it to older ones. I can see that this new material resulted from taking Reviewer #1's suggestion to heart, but the bottom line is that it didn't turn out well.

First, I don't think you needed to remove the extensive set of potentially very informative simulations that you had done previously but, if anything, add to them. (I know the 20-page Report limit is tight, but you have no limits in the supplement.) Evaluation of new methods when assumptions are not met builds naturally on first having evaluated the method when its assumptions are met, not skipping over that step. However, you have removed the lion's share of such comparisons, leaving only the simple 3 species case before diving into the community dynamics simulations. I was left unable to know how the method performs on its own terms before being faced with evaluation of how it performs on very complicated terms.

> I appreciate this perspective.  I felt very ambivalent about including these simulations in the first place, and had many of the same concerns that you raise.  I ultimately decided to include them because a substantial number of colleagues that read the preprint had similar concerns to Reviewer 1: they wanted to see how the different methods would perform on more realistic data sets and especially on data sets that were not simulated from a model that resembled a Markov network.  I do agree, however, that it is important to show that the model performs well on its own terms before diving into a more complicated model, and that omitting the simpler simulations made the paper worse. For this reason, I now include three different simulation regimes for the 20-species landscapes: 1. Simulating directly from a Markov network, 2. Simulating from a Markov network with spatial heterogeneity in the α parameters (sometimes called a conditional random field), and 3. Simulating with a dynamic model.

> For this revision, I considered a number of alternative simulation methods for the dynamic model that could address the concerns of Reviewer 1 and my colleagues, but they all had serious problems.  Using a set of GLMs as the generative model (as you suggest below) would either involve directed cycles (if β_ij≠β_ji) or would match the Markov network exactly (if β_ij=β_ji). Allowing directed cycles in the generative model would cause a number of problems. Most importantly, these models would not be identifiable from the presence-absence vectors they generated, as an infinite number of different parameter settings could all generate a given joint distribution. On the other hand, re-labeling the simulation network as a set of GLMs with equality constraints on the coefficients might have alleviated some readers' concerns, but only by hiding the fact that it was actually the same model.

Your proposed community dynamics model would raise a host of concerns. You state it is a large population approximation but then include demographic stochasticity, which is typically of diminishing importance for large populations and outweighed by environmental stochasticity, which you omit.

> I now refer to the differential equation as the model's deterministic backbone, rather than as an approximation.

The choice of parameters to draw from is arbitrary and ultimately discernible only by reading code, and the mutualism term appears to be arbitrary.

> This is true, but it was also true of all the other models I could envision where mutualism was guaranteed not to cause unbounded population explosions regardless of the parameters chosen.

You may disagree, and one could debate all aspects, but that is really the point: one would have wanted a simpler and more standard model so comparison of the statistical methods is not tangled in these issues.

> The simpler models are now included.

I can see you made a serious effort, but the outcome is not something I think readers will gain value from. The ultimate problem is that we are left with no comprehension of how the data generated by these processes suit the assumptions of the various methods. Does the model actually generate data that are well approximated by the Markov network assumptions or not? Trying to understand that would be a theoretical ecology exercise that, for your proposed model, does make sense to include in a paper about the statistical methods.

> Based on the number of readers that have asked for something along these lines (including Reviewer 1), I think that there is some value.  The new Figure 3 shows that the addition of population dynamics and per-capita interactions does weaken the relationship between the simulation coefficients and all of the estimation methods, but that it does not change the rank ordering of the different methods, which is an important result.

I suggest instead staying closer to the statistical models and including some additional cases to probe the method's robustness to violations of its assumptions and/or cases where one of the other parametric models in the competition generates the data. Which kinds of potential violations of assumptions to consider is for you to decide. For example, perhaps the cases of interest for the Markov network model would be if there really are 3-way interactions, or asymmetric interactions, or heterogeneity among sites in the model parameters, that contribute to the data generating process but are ignored in the analysis model. Indeed, describing what would constitute violations would help the reader. Another good way to probe would be to generate data from one of the other parametric models, such as the GLM.

> I now do this with the first two simulation types: the first type matches the model's assumptions and the second one adds heterogeneity among sites.  I now show that this violation decreases the models' R-squared values and increases their Type I error rates.

Note that often violations of assumptions do not create bias in parameter estimates (although sometimes they do) but rather incorrect assessments of uncertainty (inaccurate confidence intervals and Type I error rates in hypothesis tests). You have referred in multiple places to having applied a bootstrap as well as calculating Wald intervals, but it appears to me you have reported almost no results about validity of confidence intervals or hypothesis tests from those (there are a few results for the 3-species case). It would be important to do so.

> I now report approximate Type I error rates

It appears to me that you did not understand my point that you are presenting results that are integrated over parameter distributions, which appears still to be the case. What I meant is the following. Often one would want to see how a statistical method performs as a function of specific parameter values. For example, one might vary the value of an interaction coefficient (beta) from negative (competition) to zero to positive (mutualism) and run a set of simulations for each value of beta. This would allow evaluation of statistical power and accuracy of confidence intervals and Type I error rate as well as RMSE, all as a function of true parameter values. It would also allow assessment of whether the method works better for some parameter scenarios than others, which is not uncommon. Instead you are drawing parameters from distributions and then presenting performance assessments in aggregate from those results, such as the average R-squared over all cases of a given network size. Generally, and clearly in your case, such metrics can be written as expectations (integrations) over the distribution of parameters you sampled from. You are right there is nothing Bayesian about it, but I pointed out a Bayesian may feel justified in doing it because many Bayesian results take the same format. It is a question of presenting aggregated versus disaggregated results. Unfortunately, presenting only aggregated results leaves the reader unable to dig deeper into the disaggregated results. The analogy to sampling from an archipelago is lovely but irrelevant. It appears to me that since you have all the simulations at hand, and since you can place more detailed results in the supplement, you should do so. I leave it to you to decide how to do that.

> Showing how some of the results change with the value of beta is a great idea, and I have added several analyses along these lines in Figure 4 and Appendix 5.  In particular, I show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta, in addition to global estimates of R-squared and Type I error rates.  Unfortunately, I am dealing with a 210-dimensional system (20 alphas and 190 betas), which limits my options.  Holding 209 parameters constant at arbitrary values doesn't seem like the right answer, since the results would depend on which values were chosen.  Apart from the trivial case where all the coefficients except one are fixed at zero (in which case the indirect interactions that make this paper's methods necessary would not exist), there doesn't seem to be any particular line through the 210-dimensional space that deserves this kind of special attention. Likewise, there isn't a good way for completely free-form exploration of this large space.  For this reason, the best compromise in this revision seemed like marginalizing over all but one variable at a time for these analyses.

Some other minor notes

- I don't think you have stated that the betas are symmetric. It appears that beta_ij = beta_ji and hence each could be interpreted as half of the interaction strength. E.g. on line 80, it would be e^4 if I take the preceding model equation as correct and include both beta_ij and beta_ji. Please fix it up if that is not right.

> I've made this issue much clearer in the revision: see lines [[XXXX]].  It should now be unambiguous that the sum in the likelihood should only include each pair once, and that each pair must share a single value for $\beta_{ij}$.

- Both times that I have seen the title anew, the phrase "estimating species interactions from observational data" made me think it would be about time-series method, since that is a common kind of "observational data" from which people try to estimate interactions. I recommend adding a descriptive term like "species composition data" or "observational occurrence data" or "presence/absence data." You decide. This could enter the first Abstract sentence as well.

> This is another good point.  I've changed the title to focus on co-occurrence data.  Now that I've included inferential statistics in the paper and evaluated their frequentist properties, I've also changed the word "estimating" back to "inferring".

- Line 44 is not stated well. You could replace "begin with the assumption" by "test the hypothesis" or insert "that the hypothesis of interest is that [all pairwaise interactions...]" or some such change.

- Line 69: The phrase "how groups of species can co-occur" seems imprecise or incorrect. Maybe "to determine species occurence models"?

- Lines 86-87 are not a good lead-in to the point of the rest of this paragraph. It sounds like you are about to talk about species interactions, again, but really the point of the paragraph is that normalizing the probabilities involves a difficult summation.

- Lines 72 & 101. I was willing to overlook the fact that you didn't define the vector y explicitly in terms of its elements on line 72 because perhaps that is really obvious. However on line 101 you then state maximum likelihood as if it would be done from a single y vector (and a single likelihood term) rather than a matrix of observations. So now I think you need more explicit notation to clarify the arrangements of your variables.

> The revised manuscript now clarifies that each $\vec{y}$ vector represents
the different species that could occur at a single site and that the full
co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape." This should
clarify the relationship betweeen the $\vec{y}$ vectors and the matrix as well
as the relationship between the single-site likelihood and the landscape-level
likelihood (independence among sites).

- Line 130-132. This statement is unclear and unconvincing. Are you saying that the remoteness of the upper range of the prior justifies that the prior has no influence? That would not be correct since it is an issue of prior weightings in the range of heavy posterior weight. The simplest way to show a prior has little influence is to try a couple of different ones.

- Lines 136-137. Bootstrapping is a method to estimate uncertainty of an estimation procedure. It is unclear what you mean by using it to validate if a procedure gives "stable" estimates.

> The bootstrapping results have now been removed and replaced with the confidence intervals.

- Line 154: This statement is not clear.

- Figure 2B: This is odd notation because, if I follow, you have chosen a null symbol (0 with a slash) to indicate the event y_i = 0 but have chosen the label y_i to indicate the event y_i = 1. I think you want P[00], P[10], etc. or P[y1=0, y2=0], P[y1=1, y2=0], etc, or something like that. The caption states "relative probability" where I think you want "probability."

> This is an excellent point. The figure has been fixed (along with the caption).

- Figures in general: would benefit from titles on each subfigure.

> Subfigures now have titles

It's too bad we are in this unusual situation that your effort to respond to a reviewer made the manuscript worse. I will let you make another revision to get back on course. What I need you to do are: (1) remove the community dynamics model, put back in extensive results from the model of interest, and take a new approach to probing the model more thoroughly by including some scenarios with clearly defined violations of assumptions and/or data generated from one of the alternative statistical models. You may decide what to put in the online material since I recognize you are trying to stick to the 20-page report limit. (2) Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation. (3) Provide some disaggregated results. (4) Show how the regression steps work in about a page or two of supplement text, not requiring code reading.

### 2015-11-05 Harris_2015/manuscript-materials/response-to-reviewers.md
add fakedata and updates

\setlength{\parskip}{0pt}

## Responses to Editor's comments

* The reviewers and I appreciate the work you have accomplished. Based on the reviews, we are willing to consider a revised version for publication in the journal, assuming that you are able to modify the manuscript according to the recommendations.

> **Response:** Thank you.

1. I would like to see a more conventional presentation of a new statistical method: state what kind of data needs to be modeled, present the model and estimation method, and then present the simulations and/or examples. This would mean placing your "Simulations" section after the section that should be labeled "Estimation" (rather than "Inferring..."), so that it comes as an evaluation of the method you have just explained. Actually what you have stated is a maximum likelihood estimation method and you have not addressed inference, i.e. how to decide uncertainty or significance.

> **Response:** I've re-ordered the material as you suggested (model, then estimation method, then examples).  I have also made a clearer distinction between parameter estimation and hypothesis testing throughout the manuscript (e.g. lines 185-191 discuss point estimates and inferential statistics separately).

2. On page two please go more directly to the method. As written it sounds like the issue is somehow to adapt partial correlations to non-Gaussian data, and I think this is a rather indirect and confusing explanation. What you have are multivariate 0/1 data and what you are presenting is a joint probability model for such data. This would provide a more direct path toward explaining to your contribution.

> **Response:** I now treat the partial correlations as an approximation to the discrete network (lines 141-154), rather than treating the discrete network as a generalization of partial correlations and introduce the method as a way to model the probability of multivariate presence-absence data under different parameter sets (lines 66-111).

3. Most of lines 84-96 are redundant and could be removed. This would save you space (and the reviewers may or may not have realized you were hewing to the 20-page limit for Statistical Reports).

> **Response:** Removed.

4. Please tread lightly on physics analogies and language and where possible use the corresponding statistical language. For example, the "partition function" is a normalizing constant.

> **Response:** I now refer to the partition function as a normalizing constant (line 386) and no longer discuss the energy function.

5. I think you could use some better paragraph choices, but I am not sure since you have no paragraph indentations. Please do not do that again.

> **Response:** Indentations have been added.  It may be worth updating the Society's [page on "Preparing LaTeX manuscripts for ESA journals"](http://esapubs.org/esapubs/latexTIPsESA.pdf) (e.g. to suggest adding `\setlength{\parindent}{1cm}` at the beginning of the manuscript and `\setlength{\parindent}{0cm}` before the references and figure captions) so that authors know to do this in the future.

6. Provide distinct labels for the two correlation-based methods to carry through the paper (line 137).

> **Response:** The distinctions between the different methods (including the ones added in response to reviewer comments) should now be clear throughout the paper (e.g. lines 192-203).

7. 24 simulations for a scenario is not very much.

> **Response:** I now perform 50 simulations per landscape size (150 total).  The bootstrap resampling performed in Appendix 3 indicates that the estimates reported in the Results section are stable with this many replicates (e.g. the 95% bootstrap CI indicates that Markov networks performed between 14 and 18 percent better than the next best method, which is a narrow enough range to have some confidence in the results).

8. The interpretation of your more complicated simulation and estimation exercise is fairly convoluted. Normally one would like to see a set of scenarios with fixed parameter values, but you are integrating over arbitrary distributions of parameter values in a way that a Bayesian might appreciate but really makes it difficult to know if the relative performance of the methods depends on the parameters. You could provide more elaborate results in the appendix. Please attempt to disentangle some of the results to provide greater understanding.

> **Response:** The simulation and estimation procedure should now be much clearer.  It could be thought of as simulating 150 ecologists that each collect data from a different archipelago, and then asking whether their results are consistent with the underlying biology and with one another.  There is no integration over any distribution, nor is there anything particularly Bayesian about the approach.  The bootstrapping analysis in Appendix 3 in particular produces frequentist confidence intervals showing that the relative performance of the different methods is insensitive to changes in parameter values.

9. Your use of a prior distribution for alpha and beta is a serious concern. It appears to be highly informative towards the range of true parameters that you drew from. On the other hand, I am not sure what you did because you stated you used a logistic distribution, but the paper you cited discusses use of t distributions for parameters of logistic regression. A potential drawback of your method is the number of parameters to estimate, so use of an informed prior is troubling. You must give results without this prior or present a very solid justification for using it or come up with another approach. Given that the reviewers and I find your method potentially valuable, then even if it turns out that estimation is difficult without informative priors, that would be an important result for readers to see clearly.

> **Response:** I've removed the citation, since I agree it wasn't clear. That paper justified the *t* prior in part as an approximation to the logistic, but that wasn't clear from my previous submission.  The new discussion on lines 128-137 is more directly on-point.  I've also substantially weakened the regularizer substantially (well beyond what was done in the Faisal et al. paper pointed out by Reviewer 2) and clarified its likely influence on the results.

10. At lines 163-165, please explain why maximum likelihood estimates should give a model with "exactly the observed occurence frequencies and co-occurrence frequencies".

> **Response:** This is now clarified (lines 105-107).

11. Lines 174-175. Typically statisticians talk about the two steps of estimation followed by inference for a given method. In this sentence you talk about "fitting" (estimation), by which you mean your method, and then "inferring", by which you actually mean other methods for comparison. And you have not actually stated how you would do inference with the new method (presumably likelihood ratios or such). Please re-write.

> **Response:** This distinction should now be clearer throughout the manuscript.  The new manuscript now discusses inferential statistics in two places, using approximate confidence intervals derived from the Hessian matrix (lines 168-169) and the bootstrap distribution Appendix 4.

12. Lines 178-181 and 185-190 need explaining, with rationale and sufficient details for a reader to repeat the work. This can go in the appendix if you prefer.

> **Response:** This material is now explained in the first three Appendices, along with code that a reader could use to replicate the analyses exactly.  The regularizer discussed on lines 178-181 of the old text is now chosen automatically by the `corpcor` package (lines 146-148 of the new text). The description of the "Pairs" model's $Z$-score (lines 157-158) should now be clearer as well.

13. I believe I followed lines 194-202 but they will be confusing for most readers. Again you can use space in the Appendix to clarify.

> **Response:** The section on model evaluation has been rewritten, and should be much clearer (see lines 170-181).  Additionally, the first three Appendices include heavily-commented code for replicating the analysis.

14. In lines 207-211, it is unclear for each method whether you are reporting a percentage of "significant" results or simply the percentage of correct signs (positive or negative). This is confusing because on the one hand the randomization methods aim explicitly for significance, but on the other hand you haven't stated how you would do a hypothesis test for your method (although clearly you could). As a result I am not sure if you are comparing apples to oranges here.

> **Response:** I now discuss the sign of the point estimate separately from its statistical significance.  See lines 183-191.

15. Please consider carefully if "Markov networks" is the label you want to provide for these models. At first glance it seems a trivial or ironic usage of the concept, since by assumption all vertices (species) in the graph are connected, so the Markovian notion of depending only on graph neighbors carries no impact. I am not sure if there is a more precise or otherwise better label, but I ask you to give this some consideration.

> **Response:** I've considered this, and I can't find a better name from the literature.  "Markov random field" still has "Markov" in the name; the term "Ising model" is not helpful to non-physicists and often implies a grid-structured network; "Boltzmann machine" is not helpful outside of one area of machine learning and often implies the inclusion of latent random variables, which are absent here.  My reading of the literature indicates that the terms "Markov network" and "Markov random field" are routinely used to describe fully-connected graphs (e.g. "Learning Factor Graphs in Polynomial Time and Sample Complexity", by Pieter Abbeel, Daphne Koller, and Andrew Y. Ng, bottom of page 1746).

16. Section 1 of the Appendix needs explicit equations

> **Response:** This whole Appendix has been replaced.

17. Your simulations seems to have a curious feature: you are drawing from distributions conditioned on having all species present. It is conceivable that this could introduce bias to your results. The conditional distribution you are actually sampling from may have different correlations (and hence parameter sampling distributions) than the full distribution. Please address this concern.

> **Response:** I no longer condition on the number of species; some landscapes now have fewer than 20 species.

18. Both reviewers note that you have missed substantial relevant literature.

> **Response:** This literature is now addressed more directly (see responses to reviewers below).

## Responses to Reviewer 1's comments

* I very much liked the manuscript: it has a clear point to make. As somebody who has used covariance matrices to make inferences about interactions, I agree that their interpretation can be spurious, for reasons pointed out by the author.

> **Response:** Thank you.

* My first major critical comment is that - if I understood correctly - the data are generated by the same Markov networks that are fitted to the data as one of the competing approaches. Thus it is not at all surprising that the Markov network approach works best in recovering the correct parameter values - the comparison between the approaches is not fair. A more convincing demonstration that the Markov network approach is able to infer interactions (not just associations) would be to use a more mechanistic model to generate the data. For example, a time series model of interactive species where the interactions would happen preferably at the individual level (as they do in nature; predators eat prey, etc.). Or at least a population level model where the current abundances of the species influence their growth rates.

> **Response:** I now simulate my landscapes from a time series model where each species' birth and death rates depend on the abundances of the other species on the landscape (see lines 118-124 and Appendix 2).

* My second major comment is that the ms seems to ignore a large amount of recent work in community ecology that has been focused on estimating co-occurrence [examples of such paper include Pollock, L.J. et al. (2014) Understanding co-occurrence by modelling species simultaneously with a Joint Species Distribution Model (JSDM). Methods in Ecology and Evolution 5, 397-406. Clark, J.S. et al. (2014) More than the sum of the parts: forest climate response from Joint Species Distribution Models. Ecological Applications 24, 990-999. Ovaskainen, O., Hottola, J. and Siitonen, J. 2010. Modeling species co-occurrence by multivariate logistic regression generates new hypotheses on fungal interactions. Ecology 91, 2514-2521]. Thus I think the author should do a better job in describing the state of the art.

> **Response:** I now include a JSDM's correlation matrix as a competing estimator (lines 159-164, Figure 3) and reference Pollock et al. (2014) in a number of places (lines 33, 64, 162-163, and 263).

* My third major comment relates to the fact that all partial covariances can be derived simply by inverting the covariance matrix (which I think is worth mentioning in the paper, now I think it is mentioned only in the Appendix). As many of the earlier methods (e.g. listed above) provide the full posterior distribution of the covariance matrix, it seems highly feasible to link to (some of) those earlier methods by using them to estimate the posterior distribution of the covariance matrix, and then inverting the covariance matrix. That would yield the full posterior distribution of the partial covariances, without the technical problem mentioned in the ms. Such comparison would seem better connected to recent literature than the regularized ridge regression.

> **Response:** I now include partial correlations calculated from the JSDM as well, inverting 1000 samples from the posterior distribution over correlation matrices (lines 163-164, Figure 3).

* I further note that while partial covariances may not have been used much in community ecology, they have been used e.g. in gene and protein interaction networks - I think the author should make links to this literature.

> **Response:** This is now noted on lines 145-146.

## Responses to Reviewer 2's comments

* The problem of species interactions is an important one, and Markov networks do indeed present a very promising avenue of research. The technical competency of the work appears strong.

> **Response:** Thank you.

1) I'd really like to see a real data example in here. The simulations are fine at an introductory level. But I think adding a real example would help illustrate the true worth of the methodology. A real data application (which doesn't have to be large) could show how using Markov networks produces a result which tallies well with our current understanding of interactions between species - in much the same way as the hypothetical example is presented.

> **Response:** I agree that this would be an important step, however I do not have access to any co-occurrence data sets where the signs and strengths of the interactions between species had been independently estimated.  I also don't have enough space to fit such an analysis in the short format for the journal.  Fortunately, the Azaele et al. (2010) paper I cite already includes an application of the method to a real data set, which should make it less important to do so here.

2) I do think there are some key references missing. The following are but two examples of inferring species interaction networks from data: [Aderhold et a. 2012, *Ecological informatics* and Faisal et al. 2010, *Ecological Informatics*]. The reader needs to know how the current paper compares with the above, and other similar works.

> **Response:** The manuscript now cites Faisal et al. (2010) on lines 142 and 151, and makes comparisons two several of the methods proposed in that paper (regularized generalized linear models and gaussian models [in the form of partial correlations]), in addition to the Markov network, a null model, the correlation matrix, and two versions of a latent variable model (see responses to Reviewer 1 above).  I did not find the spaially-explicit methods in Aderhold et al. (2012) to be as relevant for the kinds of spatially-implicit model comparisons being made in the paper, although it is possible that I misunderstood something.

* Of course, it is also perhaps worth noting that Markov networks are common for models of spatial correlation in Ecology.

> **Response:** This is now noted on line 53.

3) The beta symbol in Fig 2A should be beta_12; this then helps connect the specific example in the figure with the formula at the top of page 5.

> **Response:** The symbol has been modified as suggested.

4) Lines 172-173; the value of 0.2 for the ridge parameter should be justified.

> **Response:** The paper no longer uses a pre-specified penalty. Instead, it relies on the `corpcor` package to determine the appropriate shrinkage for each matrix (line 148).

5) Some comments on sensitivity would be welcome - especially with regard to Figure 3. We really should see something more on the uncertainty (or stability) of the estimated networks.

> **Response:** The revised manuscript includes several new analyses to address this issue.  First, the uncertainty in Figure 3's results is estimated with bootstrap resampling of the 50 simulated landscapes in each size class (Appendix 3).  Second, the sensitivity to sampling error is estimated in Appendix 4, using bootstrap resampling of the sites in one landscape.  Third, the manuscript now mentions the approximate confidence intervals that can be produced using the Hessian matrix (lines 168-169).

### 2015-11-05 Harris_2015/manuscript-materials/draft.Rmd
add fakedata and updates

@@
 A Markov network defines the relative probability of observing a given vector of
-species-level presences (1s) and absences (0s), $\vec{y}$, as
+species-level presences (1s) and absences (0s), $\vec{y}$ at a site, as
 
@@
 so that most species pairs interacted negligibly, a few pairs interacted very
-strongly, and competition was three times more common than facilitation.
+strongly, and competition was more common than facilitation.
 
@@
 the environmental or spatial variables that helped shape the landscape [cf
-@connor_checkered_2013].  I generated the final set of landscapes by simulating
-the population dynamics of 20 interacting species, as described in Appendix 3.
+@connor_checkered_2013].  I generated the final set of landscapes with an
+abundance-based model that included per-capita interaction rates instead of
+per-species interaction rates, as described in Appendix 3.
 
@@
 \textbf{Twenty species.}
-Despite some variability across contexts (Figure 3), the four methods that
-controlled for indirect effects (the Markov network, the generalized linear
-models, and the two partial correlation-based methods) outperformed those
-that did not.  Of these, the Markov network consistently performed the best.
-As anticipated by @lee_learning_2012, generalized linear models closely
+In general, each model's performance was highest for large landscapes with
+simple assembly rules and no environmental heterogeneity (Figure 3).
+Despite some variability across contexts, the rank ordering across methods was
+very consistent.  In particular, the four methods that controlled for indirect
+effects (the Markov network, the generalized linear models, and the two partial
+correlation-based methods) always matched or outperformed those that did not.  
+Of these, the Markov network consistently performed the best. As anticipated by
+@lee_learning_2012, generalized linear models closely
 approximated the Markov network estimates, especially when the data sets were
@@
 landscapes where the gap was larger is most important. As shown in Appendix
-[[X]], the standard errors associated with the estimates in Figure 3 are small
+5, the standard errors associated with the estimates in Figure 3 are small
 (less than 0.01), so the differences among methods should not be attributed to
@@
 first two simulation types, for which the true interaction rates are easiest to
-interpret).  The Markov network's approximate Type I error rate (defined here as
-its probability of rejecting the null hypothesis for a pair of species where
-$|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for simulations that
-matched the model's assumptions, versus 0.15 for simulations that included
-environmental heterogeneity [[Appendix]].  For the null model, these rates were
-0.31 and 0.50, respectively---far higher than the nominal 0.05 rate. Figure 4C
-shows, across a range of true interaction strengths, the probability that the
-null model or the Markov network will predict the wrong sign of the interaction
-with 95% confidence.  The null model makes such errors nearly 8 times as often
-as the Markov network, even though it only reject the null hypothesis twice as
-often overall [[Appendix]].  The Markov network's errors were also more
-concentrated around 0, so it never misclassified strong interactions like the
-null model did.
+interpret).  The Markov network's approximate Type I error rate (defined here
+as the probability that 0 fell outside the 95% confidence interval for a pair of
+species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for
+simulations that matched the model's assumptions, versus 0.15 for simulations
+that included environmental heterogeneity (Appendix 5).  Coverage was similar
+for other values of $\beta$. For the null model, these rates were 0.31 and 0.50,
+respectively---far higher than the nominal 0.05 rate. Figure 4C shows, across a
+range of true interaction strengths, the probability that the null model or the
+Markov network will predict the wrong sign of the interaction with 95%
+confidence.  The null model makes such errors nearly 8 times as often as the
+Markov network, even though it only reject the null hypothesis twice as often
+overall (Appendix 5).  The Markov network's errors were also more concentrated
+around 0, so it never misclassified strong interactions like the null model did.
 
@@
 
-\noindent \textbf{Figure 1: A.} A small network of three competing species.
+\noindent \textbf{References:}
+
+\setlength{\parindent}{-1em}
+\setlength{\leftskip}{1em}
+
+\setlength{\parindent}{0cm}
+\setlength{\leftskip}{0cm}
+
+**Figure 1: A.** A small network of three competing species.
 The tree (top) tends not to co-occur with either of the two shrub species, as
@@
 
-\noindent \textbf{Figure 2: A.} A small Markov network, defined by its $\alpha$ and
+**Figure 2: A.** A small Markov network, defined by its $\alpha$ and
 $\beta$ values.  The abiotic environment favors the occurrence of each species
@@
 
-\noindent \textbf{Figure 3:} Proportion of variance in interaction
+**Figure 3:** Proportion of variance in interaction
 coefficients explained by each method versus number of sampled locations across
@@
 
-\noindent \textbf{Figure 4: A.} The Markov network's estimated interaction
+**Figure 4: A.** The Markov network's estimated interaction
 coefficients were generally very similar to the GLM estimates. **B.** The null
@@
 than the Markov network was.
-
-\noindent \textbf{References:}
-
-\setlength{\parindent}{-1em}
-\setlength{\leftskip}{1em}

### 2015-11-05 Harris_2015/manuscript-materials/draft.tex
add fakedata and updates

@@
 \begin{flushleft}
-\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
+\chead{\ifthenelse{\value{page}=1}{}{\textit{Inferring species interactions}}}
 \renewcommand{\headrulewidth}{0pt}
@@
 replicates for each combination of size and model). As described in
-Appendix {[}{[}X{]}{]}, I randomly drew the ``true'' coefficient values
-for each replicate landscape so that most species pairs interacted
-negligibly, a few pairs interacted very strongly, and competition was
-three times more common than facilitation.
+Appendix 2, I randomly drew the ``true'' coefficient values for each
+replicate landscape so that most species pairs interacted negligibly, a
+few pairs interacted very strongly, and competition was three times more
+common than facilitation.
 
@@
 shape the landscape (cf Connor et al. 2013). I generated the final set
-of landscapes by simulating the population dynamics of 20 interacting
-species, as described in Appendix {[}{[}X{]}{]}.
+of landscapes by simulating the population dynamics of 20 species with
+given per-capita interaction rates, as described in Appendix 3.
 
@@
 I compared seven techniques for determining the sign and strength of the
-associations between pairs of species from simulated data (Appendix 3).
+associations between pairs of species from simulated data (Appendix 4).
 First, I used the rosalia package (Harris 2015a) to fit Markov network
@@
 weak regularizer (equivalent to a logistic prior with scale 2) to ensure
-that the model estimates were always finite {[}{[}Appendix{]}{]}.
+that the model estimates were always finite.
 
@@
 \noindent
-\textbf{Twenty species.} Despite some variability across contexts
-(Figure 3), the four methods that controlled for indirect effects (the
+\textbf{Twenty species.} In general, each model's performance was
+highest for large landscapes with simple assembly rules and no
+environmental heterogeneity (Figure 3). Despite some variability across
+contexts, the rank ordering across methods was very consistent. In
+particular, the four methods that controlled for indirect effects (the
 Markov network, the generalized linear models, and the two partial
-correlation-based methods) outperformed those that did not. Of these,
-the Markov network consistently performed the best. As anticipated by
-Lee and Hastie (2012), generalized linear models closely approximated
-the Markov network estimates, especially when the data sets were very
-large. As reviewed in Gotelli and Ulrich (2009), however, most analyses
-in this field involve fewer than 50 sites, so model performance on the
-smaller landscapes where the gap was larger is most important. As shown
-in Appendix {[}{[}X{]}{]}, the standard errors associated with the
-estimates in Figure 3 are small (less than 0.01), so the differences
-among methods should not be attributed to sampling error.
+correlation-based methods) always matched or outperformed those that did
+not.\\Of these, the Markov network consistently performed the best. As
+anticipated by Lee and Hastie (2012), generalized linear models closely
+approximated the Markov network estimates, especially when the data sets
+were very large. As reviewed in Gotelli and Ulrich (2009), however, most
+analyses in this field involve fewer than 50 sites, so model performance
+on the smaller landscapes where the gap was larger is most important. As
+shown in Appendix 5, the standard errors associated with the estimates
+in Figure 3 are small (less than 0.01), so the differences among methods
+should not be attributed to sampling error.
 
@@
 easiest to interpret). The Markov network's approximate Type I error
-rate (defined here as its probability of rejecting the null hypothesis
-for a pair of species where \(|\beta_{ij}|<0.1\)) depended on the
-simulation type: 0.03 for simulations that matched the model's
-assumptions, versus 0.15 for simulations that included environmental
-heterogeneity {[}{[}Appendix{]}{]}. For the null model, these rates were
-0.31 and 0.50, respectively---far higher than the nominal 0.05 rate.
-Figure 4C shows, across a range of true interaction strengths, the
-probability that the null model or the Markov network will predict the
-wrong sign of the interaction with 95\% confidence. The null model makes
-such errors nearly 8 times as often as the Markov network, even though
-it only reject the null hypothesis twice as often overall
-{[}{[}Appendix{]}{]}. The Markov network's errors were also more
-concentrated around 0, so it never misclassified strong interactions
-like the null model did.
+rate (defined here as the probability that 0 fell outside the 95\%
+confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
+depended on the simulation type: 0.03 for simulations that matched the
+model's assumptions, versus 0.15 for simulations that included
+environmental heterogeneity (Appendix 5). Coverage was similar for other
+values of \(\beta\). For the null model, these rates were 0.31 and 0.50,
+respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
+across a range of true interaction strengths, the probability that the
+null model or the Markov network will predict the wrong sign of the
+interaction with 95\% confidence. The null model makes such errors
+nearly 8 times as often as the Markov network, even though it only
+reject the null hypothesis twice as often overall (Appendix 5). The
+Markov network's errors were also more concentrated around 0, so it
+never misclassified strong interactions like the null model did.
 
@@
 
-\noindent \textbf{Figure 1: A.} A small network of three competing
-species. The tree (top) tends not to co-occur with either of the two
-shrub species, as indicated by the strongly negative coefficient linking
-them. The two shrub species also compete with one another, but more
-weakly (circled coefficient). \textbf{B.} In spite of the competitive
-interactions between the two shrub species, their shared tendency to
-occur in locations without trees makes their occurrence vectors
-positively correlated (circled). \textbf{C.} Controlling for trees with
-a conditional (all-else-equal) approach such as a partial covariance or
-a Markov network leads to correct identification of the negative
+\noindent \textbf{References:}
+
+\setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
+
+\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}
+
+\textbf{Figure 1: A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a
+Markov network leads to correct identification of the negative
 shrub-shrub interaction (circled).
 
-\noindent \textbf{Figure 2: A.} A small Markov network, defined by its
-\(\alpha\) and \(\beta\) values. The abiotic environment favors the
-occurrence of each species (\(\alpha >0\)), particularly species 2
+\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
+and \(\beta\) values. The abiotic environment favors the occurrence of
+each species (\(\alpha >0\)), particularly species 2
 (\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
@@
 
-\noindent \textbf{Figure 3:} Proportion of variance in interaction
-coefficients explained by each method versus number of sampled locations
-across the three simulation types.
+\textbf{Figure 3:} Proportion of variance in interaction coefficients
+explained by each method versus number of sampled locations across the
+three simulation types.
 
-\noindent \textbf{Figure 4: A.} The Markov network's estimated
-interaction coefficients were generally very similar to the GLM
-estimates. \textbf{B.} The null model's estimates typically matched the
-(negative) correlation coefficient, after controlling for landscape
-size. \textbf{C.} For any given interaction strength, the null model was
-much more likely to misclassify its sign with 95\% confidence than the
-Markov network was.
-
-\noindent \textbf{References:}
-
-\setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
+\textbf{Figure 4: A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates.
+\textbf{B.} The null model's estimates typically matched the (negative)
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was.
 

### 2015-11-05 Harris_2015/simulation.Rmd
add fakedata and updates

@@
 The following function draws random coefficients for a Markov network of a pre-specified size.
-The probability distribution was chosen so that most species pairs interact weakly, a few
-act strongly, and that most of the interactions were negative rather than positive.
 
 ```{r, cache=TRUE}
-make_coefficients = function(n_spp){
+make_coefficients = function(n_spp, p_neg, mean_alpha){
   # Exponential distribution has lots of mass near 0 but has
@@
 
-  # Species interactions are negative 3/4 of the time in 
-  # these simulations.
+  # Multiply some proportion of the interactions
+  # by -1
   b = true_beta_magnitudes * sample(
@@
     size = length(true_beta_magnitudes), 
-    prob = c(.75, .25),
+    prob = c(p_neg, 1 - p_neg),
     replace = TRUE
@@
   # Species' intercepts are normally distributed
-  a = rnorm(n_spp, -1)
+  a = rnorm(n_spp, mean_alpha)
 
@@
 
-I used the following function to generate "true" parameters with the method above, 
+I used the following R function to generate "true" parameters with the method above, 
 simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.
@@
 
+For the abundance-based simulations, abundance was simulated with a Poisson
+distribution with rate parameter given by 
+$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$.
+
 ```{r, cache = TRUE}
-simulate_data = function(n_spp, n_sites, rep_name, n_gibbs = 1000, n_env = 2, sd = 2){
+simulate_data = function(n_spp, n_sites, rep_name, n_gibbs, n_env, sd, f, rdist, p_neg, mean_alpha){
+  # n_spp:     number of species to include in the landscape
+  # n_sites:   number of sites to include in the landscape
+  # rep_name:  an identifier to use for the landscape replicate
+  # n_gibbs:   number of Gibbs samples to perform
+  # n_env:     number of environmental variables to simulate
+  # sd:        standard deviation of environmental variables (can be zero)
+  # f:         inverse link function (see above for two examples)
+  # rdist:     a function for sampling a random value from a distribution
+  # p_neg:     proportion of negative interactions (e.g. competition)
+  # mean_alpha: the intercept value for the average species
 
   # Determine the "true" parameters for the simulated assemblage
-  par = make_coefficients(n_spp)
+  par = make_coefficients(n_spp, p_neg, mean_alpha)
 
@@
   # Landscape starts as if betas were all zero. Each species' occurrence probability
-  # depends on its alpha value and on the environment (assuming alpha_env is not zero).
+  # or abundance depends on its alpha value and on the environment (assuming alpha_env 
+  # is not zero).
   x = matrix(
-    plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
+    f(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
     nrow = n_sites, 
@@
     for(j in 1:n_spp){
-      x[,j] = rbinom(
+      x[,j] = rdist(
         nrow(x),
-        size = 1,
-        prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
+        f(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
       )
@@
 
+  # Collapse abundance data to presence/absence and store
+  # it as integer values rather than true/false
+  x = x > 0
+  mode(x) = "integer"
+
+  colnames(x) = paste0("V", 1:n_spp)
+
   # Save the results in a "fake data" folder
 
-  file_stem = paste(n_spp, n_sites, rep_name, sep = "-")
+  file_stem = paste(n_sites, rep_name, sep = "-")
+
+  # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
+  x_subset = x[rowSums(x) != 0, colSums(x) != 0]
 
@@
   )
-
-  # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
-  x_subset = x[rowSums(x) != 0, colSums(x) != 0]
 
@@
 ```{r}
+# Define a convenience function for Bernoulli random samples
+rbern = function(n, prob){
+  rbinom(n = n, size = 1, prob = prob)
+}
+```
+
+```{r}
 set.seed(1)
@@
 for(n_sites in c(25, 200, 1600)){
-  for(type in c("env", "no_env")){
+  for(type in c("no_env", "env", "abund")){
     for(i in 1:50){
@@
         n_sites = n_sites, 
+        n_gibbs = 1000,
+        n_env = 2,
         rep_name = paste0(type, i),
-        sd = ifelse(type == "env", 2, 0)
+        sd = ifelse(type == "env", 2, 0),
+        f = if(type == "abund"){function(x){log(1 + exp(x))}}else{plogis},
+        rdist = if(type == "abund"){rpois}else{rbern},
+        p_neg = if(type == "abund"){1}else{0.75},
+        mean_alpha = if(type == "abund"){-1}else{5}
       )

### 2015-11-05 Harris_2015/show-results.R

@@ -256,7 +256,7 @@ y_pairs = error_smoother(pairs_summary)
 y_markov = error_smoother(markov_summary)
 
 
-pdf("manuscript-materials/figures/error_rates.pdf", height = 9, width = 3)
+pdf("manuscript-materials/figures/error_rates.pdf", height = 8.5, width = 8.5/3)
 par(mfrow = c(3, 1))
 with(
   null_cor, 
@@ -322,12 +322,46 @@ with(markov_summary,  mean(sig_neg | sig_pos))
 
 
 # P(reject null | small interaction)
+# (approximate Type I error rates)
 markov_summary %>%
   group_by(simulation_type) %>% 
   filter(abs(truth) < .1) %>% 
-  summarize(Type_1_error_rate = round(mean(sig_neg | sig_pos), 2))
+  summarize(Type_1_error_rate = round(mean(sig_neg | sig_pos), 3))
 
 pairs_summary %>%
   group_by(simulation_type) %>% 
   filter(abs(truth) < .1) %>% 
-  summarize(Type_1_error_rate = round(mean(sig_neg | sig_pos), 2))
+  summarize(Type_1_error_rate = round(mean(sig_neg | sig_pos), 3))
+
+
+
+# Confidence interval coverage across all values
+# What's the probability that any "true" value falls inside the 95% CI?
+markov_summary %>% 
+  group_by(simulation_type) %>% 
+  summarize(round(mean(truth > lower & truth < upper), 3))
+
+
+# RMSE
+markov_summary %>% 
+  group_by(simulation_type) %>%
+  summarize(sqrt(mean((truth - estimate)^2)))
+
+
+
+# Coverage versus true beta value.  The top and bottom 0.5%
+# of the distribution have been omitted to prevent bad behavior
+# by the smoother in the tails.
+markov_summary %>%
+  filter(percent_rank(truth) > .005 & percent_rank(truth) < .995) %>%
+  mutate(covered = truth > lower & truth < upper) %>%
+  ggplot(aes(x = truth, y = as.integer(covered))) + 
+  facet_grid(~simulation_type) + 
+  geom_smooth(method = gam, formula = y ~ s(x), family = binomial) + 
+  theme_bw() + 
+  coord_cartesian(ylim = c(0, 1)) + 
+  geom_vline(xintercept = 0) + 
+  geom_hline(yintercept = 0.95, color = "red") + 
+  ylab("Coverage")
+
+

### 2015-11-12 Harris_2015/simulation.Rmd
fix mean alpha

@@
         p_neg = if(type == "abund"){1}else{0.75},
-        mean_alpha = if(type == "abund"){-1}else{5}
+        mean_alpha = if(type == "abund"){5}else{-1}
       )

### 2015-11-22 Harris_2015/manuscript-materials/draft.Rmd
latest simulations etc

@@
 1600 sites using three increasingly complex models (50 replicates for each
-combination of size and model). As described in Appendix 2, I randomly drew
-the “true” coefficient values for each replicate landscape
-so that most species pairs interacted negligibly, a few pairs interacted very
-strongly, and competition was more common than facilitation.
+combination of size and model; see Appendix 2 for details). I randomly drew the
+“true” coefficient magnitudes for each replicate landscape from exponential
+distributions so that most species pairs interacted negligibly but a few pairs
+interacted strongly enough that their effects could propagate indirectly to
+other species in the network.
 
@@
 abundance-based model that included per-capita interaction rates instead of
-per-species interaction rates, as described in Appendix 3.
+per-species interaction rates.
 
@@
 using the posterior mean correlation [@pollock_understanding_2014], I also used
-the posterior mean *partial* correlation, which might be able to control for
+the posterior mean *partial* correlation, which should control better for
 indirect effects.
@@
 approximated the Markov network estimates, especially when the data sets were
-very large. As reviewed in @gotelli_empirical_2009, however, most analyses in
-this field involve fewer than 50 sites, so model performance on the smaller
-landscapes where the gap was larger is most important. As shown in Appendix
+very large [[Figure]]. As reviewed in @gotelli_empirical_2009, however, most analyses in
+this field of ecology involve fewer than 50 sites, so model performance on the smaller
+landscapes (where the gap was larger) is very important. As shown in Appendix
 5, the standard errors associated with the estimates in Figure 3 are small
@@
 as the probability that 0 fell outside the 95% confidence interval for a pair of
-species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.03 for
-simulations that matched the model's assumptions, versus 0.15 for simulations
-that included environmental heterogeneity (Appendix 5).  Coverage was similar
-for other values of $\beta$. For the null model, these rates were 0.31 and 0.50,
-respectively---far higher than the nominal 0.05 rate. Figure 4C shows, across a
-range of true interaction strengths, the probability that the null model or the
-Markov network will predict the wrong sign of the interaction with 95%
-confidence.  The null model makes such errors nearly 8 times as often as the
-Markov network, even though it only reject the null hypothesis twice as often
-overall (Appendix 5).  The Markov network's errors were also more concentrated
-around 0, so it never misclassified strong interactions like the null model did.
+species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
+simulations that matched the model's assumptions, versus 0.14 for simulations
+that included environmental heterogeneity (see Appendix 5 for confidence
+interval coverage across a range of $\beta_{ij}$ values). The null model's
+Type I error rates were [[XX]] and [[XX]] for the constant and heterogeneous
+landscapse, respectively---far higher than the nominal 0.05 rate. Figure
+4C shows, across a range of true interaction strengths, the probability that
+the null model or the Markov network will predict the wrong sign of the
+interaction with 95% confidence.  The null model makes such errors [[XX]] times
+as often as the Markov network, even though it only reject the null hypothesis
+[[XX]] as often overall (Appendix 5).  The Markov network's errors were also
+more concentrated around 0, so it never misclassified strong interactions like
+the null model did.
 
@@
 The results presented above show that Markov networks can reliably recover
-species' pairwise interactions from observational data, even for cases where
+species' pairwise interactions from species composition data, even for cases where
 environmental heterogeneity and indirect interactions cause ecologists' typical
@@
 
-**Figure 3:** Proportion of variance in interaction
-coefficients explained by each method versus number of sampled locations across
-the three simulation types.
+**Figure 3:** Proportion of variance in interaction coefficients explained by
+each method versus number of sampled locations across the three simulation
+types. For the null model (Pairs), these values include the removal of two
+outliers with $|Z|>1000$ [[Appendix]].
 

### 2015-11-22 Harris_2015/manuscript-materials/draft.tex
latest simulations etc

@@
 A Markov network defines the relative probability of observing a given
-vector of species-level presences (1s) and absences (0s), \(\vec{y}\),
-as
+vector of species-level presences (1s) and absences (0s), \(\vec{y}\) at
+a site, as
 
@@
 200, or 1600 sites using three increasingly complex models (50
-replicates for each combination of size and model). As described in
-Appendix 2, I randomly drew the ``true'' coefficient values for each
-replicate landscape so that most species pairs interacted negligibly, a
-few pairs interacted very strongly, and competition was three times more
-common than facilitation.
+replicates for each combination of size and model; see Appendix 2 for
+details). I randomly drew the ``true'' coefficient magnitudes for each
+replicate landscape from exponential distributions so that most species
+pairs interacted negligibly but a few pairs interacted strongly enough
+that their effects could propagate indirectly to other species in the
+network.
 
@@
 shape the landscape (cf Connor et al. 2013). I generated the final set
-of landscapes by simulating the population dynamics of 20 species with
-given per-capita interaction rates, as described in Appendix 3.
+of landscapes with an abundance-based model that included per-capita
+interaction rates instead of per-species interaction rates.
 
@@
 addition to using the posterior mean correlation (Pollock et al. 2014),
-I also used the posterior mean \emph{partial} correlation, which might
-be able to control for indirect effects.
+I also used the posterior mean \emph{partial} correlation, which should
+control better for indirect effects.
 
@@
 correlation-based methods) always matched or outperformed those that did
-not.\\Of these, the Markov network consistently performed the best. As
+not.\\
+Of these, the Markov network consistently performed the best. As
 anticipated by Lee and Hastie (2012), generalized linear models closely
 approximated the Markov network estimates, especially when the data sets
-were very large. As reviewed in Gotelli and Ulrich (2009), however, most
-analyses in this field involve fewer than 50 sites, so model performance
-on the smaller landscapes where the gap was larger is most important. As
-shown in Appendix 5, the standard errors associated with the estimates
-in Figure 3 are small (less than 0.01), so the differences among methods
-should not be attributed to sampling error.
+were very large {[}{[}Figure{]}{]}. As reviewed in Gotelli and Ulrich
+(2009), however, most analyses in this field of ecology involve fewer
+than 50 sites, so model performance on the smaller landscapes (where the
+gap was larger) is very important. As shown in Appendix 5, the standard
+errors associated with the estimates in Figure 3 are small (less than
+0.01), so the differences among methods should not be attributed to
+sampling error.
 
@@
 confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
-depended on the simulation type: 0.03 for simulations that matched the
-model's assumptions, versus 0.15 for simulations that included
-environmental heterogeneity (Appendix 5). Coverage was similar for other
-values of \(\beta\). For the null model, these rates were 0.31 and 0.50,
-respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
-across a range of true interaction strengths, the probability that the
-null model or the Markov network will predict the wrong sign of the
-interaction with 95\% confidence. The null model makes such errors
-nearly 8 times as often as the Markov network, even though it only
-reject the null hypothesis twice as often overall (Appendix 5). The
-Markov network's errors were also more concentrated around 0, so it
-never misclassified strong interactions like the null model did.
+depended on the simulation type: 0.02 for simulations that matched the
+model's assumptions, versus 0.14 for simulations that included
+environmental heterogeneity (see Appendix 5 for confidence interval
+coverage across a range of \(\beta_{ij}\) values). The null model's Type
+I error rates were {[}{[}XX{]}{]} and {[}{[}XX{]}{]} for the constant
+and heterogeneous landscapse, respectively---far higher than the nominal
+0.05 rate. Figure 4C shows, across a range of true interaction
+strengths, the probability that the null model or the Markov network
+will predict the wrong sign of the interaction with 95\% confidence. The
+null model makes such errors {[}{[}XX{]}{]} times as often as the Markov
+network, even though it only reject the null hypothesis {[}{[}XX{]}{]}
+as often overall (Appendix 5). The Markov network's errors were also
+more concentrated around 0, so it never misclassified strong
+interactions like the null model did.
 
@@
 The results presented above show that Markov networks can reliably
-recover species' pairwise interactions from observational data, even for
-cases where environmental heterogeneity and indirect interactions cause
-ecologists' typical null modeling approaches to reliably fail. Partial
-covariances and generalized linear models can both provide
-computationally efficient approximations, but with somewhat lower
-accuracy, especially for typically-sized data sets with small numbers of
-sites (Gotelli and Ulrich 2009). The difference in accuracy may be
-larger for real data sets than for the simulated landscapes in Figure 3,
-however; linear approximations to the Markov network make larger errors
-when the interaction matrix is structured (e.g.~due to guilds or trophic
-levels; Loh and Wainwright 2013). Similarly, the separate generalized
-linear models for each species can severely overfit in some cases (Lee
-and Hastie 2012). The full Markov network should thus be preferred to
-the approximations when it is computationally tractable.
+recover species' pairwise interactions from species composition data,
+even for cases where environmental heterogeneity and indirect
+interactions cause ecologists' typical null modeling approaches to
+reliably fail. Partial covariances and generalized linear models can
+both provide computationally efficient approximations, but with somewhat
+lower accuracy, especially for typically-sized data sets with small
+numbers of sites (Gotelli and Ulrich 2009). The difference in accuracy
+may be larger for real data sets than for the simulated landscapes in
+Figure 3, however; linear approximations to the Markov network make
+larger errors when the interaction matrix is structured (e.g.~due to
+guilds or trophic levels; Loh and Wainwright 2013). Similarly, the
+separate generalized linear models for each species can severely overfit
+in some cases (Lee and Hastie 2012). The full Markov network should thus
+be preferred to the approximations when it is computationally tractable.
 
@@
 
+\hyperdef{}{refs}{\label{refs}}
+\hyperdef{}{ref-albrechtux5fspatialux5f2001}{\label{ref-albrechtux5fspatialux5f2001}}
 Albrecht, M., and N. J. Gotelli. 2001. Spatial and temporal niche
@@
 
+\hyperdef{}{ref-azaeleux5finferringux5f2010}{\label{ref-azaeleux5finferringux5f2010}}
 Azaele, S., R. Muneepeerakul, A. Rinaldo, and I. Rodriguez-Iturbe. 2010.
@@
 
+\hyperdef{}{ref-brunoux5finclusionux5f2003}{\label{ref-brunoux5finclusionux5f2003}}
 Bruno, J. F., J. J. Stachowicz, and M. D. Bertness. 2003. Inclusion of
@@
 
+\hyperdef{}{ref-cipraux5fintroductionux5f1987}{\label{ref-cipraux5fintroductionux5f1987}}
 Cipra, B. A. 1987. An introduction to the Ising model. American
@@
 
+\hyperdef{}{ref-connorux5fcheckeredux5f2013}{\label{ref-connorux5fcheckeredux5f2013}}
 Connor, E. F., M. D. Collins, and D. Simberloff. 2013. The checkered
@@
 
+\hyperdef{}{ref-diamondux5fislandux5f1975}{\label{ref-diamondux5fislandux5f1975}}
 Diamond, J. M. 1975. The island dilemma: Lessons of modern biogeographic
@@
 
+\hyperdef{}{ref-dodsonux5fcomplementaryux5f1970}{\label{ref-dodsonux5fcomplementaryux5f1970}}
 Dodson, S. I. 1970. COMPLEMENTARY FEEDING NICHES SUSTAINED BY
@@
 
+\hyperdef{}{ref-faisalux5finferringux5f2010}{\label{ref-faisalux5finferringux5f2010}}
 Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
@@
 
+\hyperdef{}{ref-friedmanux5fsparseux5f2008}{\label{ref-friedmanux5fsparseux5f2008}}
 Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
@@
 
+\hyperdef{}{ref-gelfandux5fmodellingux5f2005}{\label{ref-gelfandux5fmodellingux5f2005}}
 Gelfand, A. E., A. M. Schmidt, S. Wu, J. A. Silander, A. Latimer, and A.
@@
 
+\hyperdef{}{ref-gelmanux5fweaklyux5f2008}{\label{ref-gelmanux5fweaklyux5f2008}}
 Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
@@
 
+\hyperdef{}{ref-gilpinux5ffactorsux5f1982}{\label{ref-gilpinux5ffactorsux5f1982}}
 Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
@@
 
+\hyperdef{}{ref-goldingux5fbayescommux5f2015}{\label{ref-goldingux5fbayescommux5f2015}}
 Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian Community
@@
 
+\hyperdef{}{ref-gotelliux5fempiricalux5f2009}{\label{ref-gotelliux5fempiricalux5f2009}}
 Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
@@
 
+\hyperdef{}{ref-harrisux5frosaliaux5f2015}{\label{ref-harrisux5frosaliaux5f2015}}
 Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
@@
 
+\hyperdef{}{ref-harrisux5fgeneratingux5f2015}{\label{ref-harrisux5fgeneratingux5f2015}}
 Harris, D. J. 2015b. Generating realistic assemblages with a Joint
@@
 
+\hyperdef{}{ref-harrisux5foccupancyux5f2011}{\label{ref-harrisux5foccupancyux5f2011}}
 Harris, D. J., K. G. Smith, and P. J. Hanly. 2011. Occupancy is
@@
 
+\hyperdef{}{ref-harrisux5fcontactux5f1974}{\label{ref-harrisux5fcontactux5f1974}}
 Harris, T. E. 1974. Contact Interactions on a Lattice. The Annals of
@@
 
+\hyperdef{}{ref-leeux5flearningux5f2012}{\label{ref-leeux5flearningux5f2012}}
 Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
@@
 
+\hyperdef{}{ref-levineux5fcompetitiveux5f1976}{\label{ref-levineux5fcompetitiveux5f1976}}
 Levine, S. H. 1976. Competitive Interactions in Ecosystems. The American
@@
 
+\hyperdef{}{ref-lewinux5fsantaux5f1983}{\label{ref-lewinux5fsantaux5f1983}}
 Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.
 
+\hyperdef{}{ref-lohux5fstructureux5f2013}{\label{ref-lohux5fstructureux5f2013}}
 Loh, P.-L., and M. J. Wainwright. 2013. Structure estimation for
@@
 
+\hyperdef{}{ref-macarthurux5fpopulationux5f1958}{\label{ref-macarthurux5fpopulationux5f1958}}
 MacArthur, R. H. 1958. Population ecology of some warblers of
@@
 
+\hyperdef{}{ref-murphyux5fmachineux5f2012}{\label{ref-murphyux5fmachineux5f2012}}
 Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
@@
 
+\hyperdef{}{ref-pollockux5funderstandingux5f2014}{\label{ref-pollockux5funderstandingux5f2014}}
 Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K.
@@
 
+\hyperdef{}{ref-rux5fcoreux5fteamux5frux5f2015}{\label{ref-rux5fcoreux5fteamux5frux5f2015}}
 R Core Team. 2015. R: A Language and Environment for Statistical
@@
 
+\hyperdef{}{ref-roughgardenux5fcompetitionux5f1983}{\label{ref-roughgardenux5fcompetitionux5f1983}}
 Roughgarden, J. 1983. Competition and Theory in Community Ecology. The
@@
 
+\hyperdef{}{ref-schaferux5fcorpcorux5f2014}{\label{ref-schaferux5fcorpcorux5f2014}}
 Schäfer, J., R. Opgen-Rhein, V. Zuber, M. Ahdesmäki, A. P. D. Silva, and
@@
 
+\hyperdef{}{ref-schmidtux5fmodelingux5f2012}{\label{ref-schmidtux5fmodelingux5f2012}}
 Schmidt, M., and K. Murphy. 2012. Modeling Discrete Interventional Data
@@
 
+\hyperdef{}{ref-strongux5fecologicalux5f1984}{\label{ref-strongux5fecologicalux5f1984}}
 Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.
@@
 
+\hyperdef{}{ref-tjelmelandux5fmarkovux5f1998}{\label{ref-tjelmelandux5fmarkovux5f1998}}
 Tjelmeland, H., and J. Besag. 1998. Markov Random Fields with
@@
 
+\hyperdef{}{ref-veechux5fprobabilisticux5f2013}{\label{ref-veechux5fprobabilisticux5f2013}}
 Veech, J. A. 2013. A probabilistic model for analysing species
@@
 
+\hyperdef{}{ref-whittamux5fspeciesux5f1981}{\label{ref-whittamux5fspeciesux5f1981}}
 Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
@@
 
+\hyperdef{}{ref-xiaoux5fstrongux5f2015}{\label{ref-xiaoux5fstrongux5f2015}}
 Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the

### 2015-11-22 Harris_2015/show-results.R

@@ -5,7 +5,6 @@ library(tidyr)
 
 
 x = read.csv("estimates.csv", stringsAsFactors = FALSE)
-x$dynamic_simulation = grepl("pop", x$rep_name)
 x$simulation_type = gsub("[0-9]", "", x$rep_name)
 
 
@@ -19,27 +18,27 @@ library(stringr)
 beginnings = grep("Sp1", pairs_txt) + 1
 ends = c(
   grep("^[^ ]", pairs_txt)[-1],
-  length(pairs_txt)
+  length(pairs_txt) + 1
 ) - 1
 
-file_list = readLines("manuscript-materials/fakedata_list.txt")[-1]
 partial_names = sapply(strsplit(grep("^>", pairs_txt, value = TRUE), " +"), function(x) x[[3]])
 filename_lines = grep("^>", pairs_txt)
 
-# File_list agrees with partial_names to the extent possible
-all(sapply(1:450, function(i)grepl(partial_names[i], file_list[i])))
-
-
-split_names = strsplit(file_list, "-")
+# Sort a vector of alphanumeric strings by the numeric component
+# as if they were integers.  For example, V20 is larger than V12,
+# even though V12 comes first alphabetically
+alnum_sort = function(x){
+  raw = as.integer(gsub("[[:alpha:]]", "", x))
+  x[order(raw)]
+}
 
 pairs_results = lapply(
-  1:length(file_list),
+  1:length(filename_lines),
   function(i){
     
-    message(i)
+    n_sites = as.integer(strsplit(partial_names[[i]], "-")[[1]][[1]])
+    rep_name = strsplit(partial_names[[i]], "-")[[1]][[2]]
     
-    n_sites = as.integer(split_names[[i]][[2]])
-    rep_name = split_names[[i]][[3]]
     
     # Find the line where the current data set is mentioned in
     # pairs.txt
@@ -59,11 +58,11 @@ pairs_results = lapply(
       splitted,
       function(x){
         # in the x data frame, species 1 is always a lower number than species 2
-        spp = sort(as.integer(x[3:4]))
+        spp = alnum_sort(x[3:4])
         data.frame(
-          sp1 = paste0("V", spp[1]), 
-          sp2 = paste0("V", spp[2]), 
-          z = as.numeric(x[14]),
+          sp1 = spp[1], 
+          sp2 = spp[2], 
+          z = x[14],
           stringsAsFactors = FALSE
[…]

### 2015-11-22 Harris_2015/manuscript-materials/draft.Rmd
almost done

@@
 ecologists have long known that indirect effects can overwhelm direct ones at
-the landscape level [@dodson_complementary_1970; @levine_competitive_1976], the
+the landscape level [@levine_competitive_1976], the
 vast majority of our methods for drawing inferences from observational data do
@@
 @pollock_understanding_2014]. To the extent that indirect interactions like
-those in Figure 1 are generally important [@dodson_complementary_1970], existing
+those in Figure 1 are generally important, existing
 methods will not provide much evidence regarding species interactions.
@@
 While competition doesn't reliably reduce co-occurrence rates at the
-whole-landscape level (as most methods assume), it nevertheless does leave a
+whole-landscape level (as most methods assume), it does still leave a
 signal in the data (Figure 1C). For example, after partitioning the data set
@@
 estimates of direct (conditional, or all-else-equal) effects.  This kind of
-precision is difficult to obtain from null models, which only simulate the most
-extreme possible scenario: that *all* direct and indirect interactions are
+precision is difficult to obtain from null models, which only test the most
+extreme possible hypothesis: that *all* direct and indirect interactions are
 exactly zero. Nevertheless, null models have dominated this field for more than
@@
 kinds of species assemblages that these relationships produce. Here, I show how
-a set of conditional relationships can be used to determine how groups of species
-can co-occur. Methods for estimating conditional relationships from data are
-discussed in the next section.
+a set of conditional relationships can determine species composition. Methods for
+estimating conditional relationships from data are discussed in the next section.
 
@@
 $e^2$ times higher when the other one is present (as compared with otherwise
-equivalent sites). Because the relative probability of a presence-absence vector
+equivalent sites). The relative probability of a presence-absence vector
 increases when positively-associated species co-occur and decreases when
-negatively-associated species co-occur, the model tends---all else equal---to
-produce assemblages that have many pairs of positively-associated species and
-relatively few pairs of negatively-associated species (exactly as an ecologist
-might expect). Of course, if all else is *not* equal (e.g. Figure 1, where the
-presence of one competitor is associated with release from another competitor),
-then predicting species' overall co-occurrence rates can be more complicated,
-and may require summing over the different possible assemblages (Figure 2B).
+negatively-associated species co-occur. As a result, the model tends---all else
+equal---to produce assemblages where many positively-associated species pairs
+co-occur and few negatively-associated pairs do (just as an ecologist might
+expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
+competitor is associated with release from another competitor), then predicting
+species' overall co-occurrence rates can be more complicated, and may require
+summing over the different possible assemblages (Figure 2B).
 
@@
 an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
-indicating which species are present at each site on the landscape). When the number of species
-is reasonably small, one can compute exact maximum likelihood estimates for all
-of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
+indicating which species are present at each site on the landscape). When the
+number of species is reasonably small (less than about 30), one can find
+exact maximum likelihood estimates for all of the $\alpha$ and $\beta$
+coefficients given a presence-absence matrix by numerically
 optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
 ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
-ensuring that this procedure will always converge on the global maximum. This
-maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
-that would be expected to produce exactly the observed co-occurrence frequencies
-on average [i.e., maximizing the likelihood ensures that the sufficient
-statistics of the model distribution match the sufficient statistics of the data;
-@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
-R programming language [@r_core_team_r_2015] to optimize the Markov network
-parameters.
+ensuring that this procedure will converge on the global maximum. I used the
+rosalia package [@harris_rosalia_2015] for the R programming language
+[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its
+gradients; the package passes these functions to the "BFGS" method in R's
+general-purpose optimizer, which then estimates the Markov network parameters.
 
@@
 landscapes with perfectly-correlated species pairs, I used a regularized
-estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
-`pcor.shrink` function with the default settings [@schafer_corpcor_2014].
+estimate provided by the corpcor package’s `pcor.shrink` function with the
+default settings [@schafer_corpcor_2014].
 
@@
 the data, however [@schmidt_modeling_2012], so I used their average as an
-overall measure of the overall relationship between species $i$ and species $j$.
+overall measure of their relationship.
 
@@
 
-I then evaluated the relationship between each method's estimates and the "true"
-interaction strengths among all of the species pairs from the larger simulated
-landscapes. This determined which of the methods provide a consistent way to
-know how strong species interactions are---regardless of which species were
-present in a particular data set or how many observations were taken. To ensure
-that the different test statistics (e.g. correlations versus $Z$ scores) were on
-a common scale, I rescaled them using linear regression through the origin. I
-then calculated the proportion of variance explained for different combinations
-of model type and landscape size (compared with a null model that assumed all
-interaction strengths to be zero).
+For the larger landscapes, I evaluated the relationship between each method's
+estimates and the "true" interaction strengths. To ensure that the different
+test statistics (e.g. correlations versus $Z$ scores) were on a common scale, I
+rescaled them using linear regression through the origin. I then calculated the
+proportion of variance explained for different combinations of model type and
+landscape size (compared with a null model that assumed all interaction
+strengths to be zero).
 
@@
 correlation-based methods) always matched or outperformed those that did not.
-Of these, the Markov network consistently performed the best. As anticipated by
+The Markov network consistently performed best of all. As anticipated by
 @lee_learning_2012, generalized linear models closely
@@
 Type I error rates were [[XX]] and [[XX]] for the constant and heterogeneous
-landscapse, respectively---far higher than the nominal 0.05 rate. Figure
+landscapes, respectively---far higher than the nominal 0.05 rate. Figure
 4C shows, across a range of true interaction strengths, the probability that
@@
 that the models' relative performance on real data sets is similar to the range
-of results shown in Figure 3, scientists in this field could more than double
+of results shown in Figure 3, scientists in this field could often triple
 their explanatory power by switching from null models to Markov networks (or
@@
 \setlength{\leftskip}{1em}
+
+<!--references go here-->
 

### 2015-11-22 Harris_2015/manuscript-materials/draft.tex
almost done

@@
 \usepackage{ifthen}
-\geometry{verbose,letterpaper,tmargin=2.5cm, bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
+\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
 \ifxetex
@@
 1B). While ecologists have long known that indirect effects can
-overwhelm direct ones at the landscape level (Dodson 1970, Levine 1976),
-the vast majority of our methods for drawing inferences from
-observational data do not control for these effects (e.g. Diamond 1975,
-Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock et al.
-2014). To the extent that indirect interactions like those in Figure 1
-are generally important (Dodson 1970), existing methods will not provide
-much evidence regarding species interactions.
+overwhelm direct ones at the landscape level (Levine 1976), the vast
+majority of our methods for drawing inferences from observational data
+do not control for these effects (e.g. Diamond 1975, Strong et al. 1984,
+Gotelli and Ulrich 2009, Veech 2013, Pollock et al. 2014). To the extent
+that indirect interactions like those in Figure 1 are generally
+important, existing methods will not provide much evidence regarding
+species interactions.
 
 While competition doesn't reliably reduce co-occurrence rates at the
-whole-landscape level (as most methods assume), it nevertheless does
-leave a signal in the data (Figure 1C). For example, after partitioning
-the data set into shaded and unshaded sites, there will be co-occurrence
-deficits in each subset that wouldn't otherwise be apparent. More
-generally, controlling for other species in the network will often be
-important for obtaining reliable estimates of direct (conditional, or
-all-else-equal) effects. This kind of precision is difficult to obtain
-from null models, which only simulate the most extreme possible
-scenario: that \emph{all} direct and indirect interactions are exactly
-zero. Nevertheless, null models have dominated this field for more than
-three decades (Strong et al. 1984, Gotelli and Ulrich 2009).
+whole-landscape level (as most methods assume), it does still leave a
+signal in the data (Figure 1C). For example, after partitioning the data
+set into shaded and unshaded sites, there will be co-occurrence deficits
+in each subset that wouldn't otherwise be apparent. More generally,
+controlling for other species in the network will often be important for
+obtaining reliable estimates of direct (conditional, or all-else-equal)
+effects. This kind of precision is difficult to obtain from null models,
+which only test the most extreme possible hypothesis: that \emph{all}
+direct and indirect interactions are exactly zero. Nevertheless, null
+models have dominated this field for more than three decades (Strong et
+al. 1984, Gotelli and Ulrich 2009).
 
@@
 assemblages that these relationships produce. Here, I show how a set of
-conditional relationships can be used to determine how groups of species
-can co-occur. Methods for estimating conditional relationships from data
-are discussed in the next section.
+conditional relationships can determine species composition. Methods for
+estimating conditional relationships from data are discussed in the next
+section.
 
@@
 occurrence would be \(e^2\) times higher when the other one is present
-(as compared with otherwise equivalent sites). Because the relative
-probability of a presence-absence vector increases when
-positively-associated species co-occur and decreases when
-negatively-associated species co-occur, the model tends---all else
-equal---to produce assemblages that have many pairs of
-positively-associated species and relatively few pairs of
-negatively-associated species (exactly as an ecologist might expect). Of
-course, if all else is \emph{not} equal (e.g.~Figure 1, where the
-presence of one competitor is associated with release from another
-competitor), then predicting species' overall co-occurrence rates can be
-more complicated, and may require summing over the different possible
-assemblages (Figure 2B).
+(as compared with otherwise equivalent sites). The relative probability
+of a presence-absence vector increases when positively-associated
+species co-occur and decreases when negatively-associated species
+co-occur. As a result, the model tends---all else equal---to produce
+assemblages where many positively-associated species pairs co-occur and
+few negatively-associated pairs do (just as an ecologist might expect).
+When all else is \emph{not} equal (e.g.~Figure 1, where the presence of
+one competitor is associated with release from another competitor), then
+predicting species' overall co-occurrence rates can be more complicated,
+and may require summing over the different possible assemblages (Figure
+2B).
 
@@
 present at each site on the landscape). When the number of species is
-reasonably small, one can compute exact maximum likelihood estimates for
-all of the \(\alpha\) and \(\beta\) coefficients given a
-presence-absence matrix by optimizing \(p(\vec{y}; \alpha, \beta)\).
-Fully-observed Markov networks like the ones considered here have
-unimodal likelihood surfaces (Murphy 2012), ensuring that this procedure
-will always converge on the global maximum. This maximum represents the
-unique combination of \(\alpha\) and \(\beta\) coefficients that would
-be expected to produce exactly the observed co-occurrence frequencies on
-average (i.e., maximizing the likelihood ensures that the sufficient
-statistics of the model distribution match the sufficient statistics of
-the data; Murphy 2012). I used the rosalia package (Harris 2015a) for
-the R programming language (R Core Team 2015) to optimize the Markov
-network parameters.
+reasonably small (less than about 30), one can find exact maximum
+likelihood estimates for all of the \(\alpha\) and \(\beta\)
+coefficients given a presence-absence matrix by numerically optimizing
+\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
+ones considered here have unimodal likelihood surfaces (Murphy 2012),
+ensuring that this procedure will converge on the global maximum. I used
+the rosalia package (Harris 2015a) for the R programming language (R
+Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
+gradients; the package passes these functions to the ``BFGS'' method in
+R's general-purpose optimizer, which then estimates the Markov network
+parameters.
 
@@
 for landscapes with perfectly-correlated species pairs, I used a
-regularized estimate based on James-Stein shrinkage, as implemented in
-the corpcor package's \texttt{pcor.shrink} function with the default
-settings (Schäfer et al. 2014).
+regularized estimate provided by the corpcor package's
+\texttt{pcor.shrink} function with the default settings (Schäfer et al.
+2014).
 
@@
 identifiable from the data, however (Schmidt and Murphy 2012), so I used
-their average as an overall measure of the overall relationship between
-species \(i\) and species \(j\).
+their average as an overall measure of their relationship.
 
@@
 
-I then evaluated the relationship between each method's estimates and
-the ``true'' interaction strengths among all of the species pairs from
-the larger simulated landscapes. This determined which of the methods
-provide a consistent way to know how strong species interactions
-are---regardless of which species were present in a particular data set
-or how many observations were taken. To ensure that the different test
-statistics (e.g.~correlations versus \(Z\) scores) were on a common
-scale, I rescaled them using linear regression through the origin. I
-then calculated the proportion of variance explained for different
-combinations of model type and landscape size (compared with a null
-model that assumed all interaction strengths to be zero).
+For the larger landscapes, I evaluated the relationship between each
+method's estimates and the ``true'' interaction strengths. To ensure
+that the different test statistics (e.g.~correlations versus \(Z\)
+scores) were on a common scale, I rescaled them using linear regression
+through the origin. I then calculated the proportion of variance
+explained for different combinations of model type and landscape size
+(compared with a null model that assumed all interaction strengths to be
+zero).
 
@@
 correlation-based methods) always matched or outperformed those that did
-not.\\
-Of these, the Markov network consistently performed the best. As
+not. The Markov network consistently performed best of all. As
 anticipated by Lee and Hastie (2012), generalized linear models closely
@@
 I error rates were {[}{[}XX{]}{]} and {[}{[}XX{]}{]} for the constant
-and heterogeneous landscapse, respectively---far higher than the nominal
+and heterogeneous landscapes, respectively---far higher than the nominal
 0.05 rate. Figure 4C shows, across a range of true interaction
@@
 similar to the range of results shown in Figure 3, scientists in this
-field could more than double their explanatory power by switching from
-null models to Markov networks (or increase it nearly as much with
-linear or generalized linear approximations). Regardless of the methods
-ecologists ultimately choose, controlling for indirect effects could
-clearly improve our understanding of species' direct effects on one
-another and on community structure.
+field could often triple their explanatory power by switching from null
+models to Markov networks (or increase it nearly as much with linear or
+generalized linear approximations). Regardless of the methods ecologists
+ultimately choose, controlling for indirect effects could clearly
+improve our understanding of species' direct effects on one another and
+on community structure.
 
@@
 explained by each method versus number of sampled locations across the
-three simulation types.
+three simulation types. For the null model (Pairs), these values include
+the removal of two outliers with \(|Z|>1000\) {[}{[}Appendix{]}{]}.
 
@@
 7:129--146.
-
-\hyperdef{}{ref-dodsonux5fcomplementaryux5f1970}{\label{ref-dodsonux5fcomplementaryux5f1970}}
-Dodson, S. I. 1970. COMPLEMENTARY FEEDING NICHES SUSTAINED BY
-SIZE-SELECTIVE PREDATION. Limnology and Oceanography 15:131--137.
 

### 2015-11-23 Harris_2015/manuscript-materials/response-to-editor.md
editor

@@
 
-> I appreciate this perspective.
-I felt very ambivalent about including these simulations in the first place, and had many of the same concerns that you raise.
-I ultimately decided to include them because a substantial number of colleagues that read the preprint had similar concerns to Reviewer 1: they wanted to see how the different methods would perform on more realistic data sets and especially on data sets that were not simulated from a model that resembled a Markov network.
-I do agree, however, that it is important to show that the model performs well on its own terms before diving into a more complicated model, and that omitting the simpler simulations made the paper worse.
-For this reason, I now include three different simulation regimes for the 20-species landscapes: 1.
-Simulating directly from a Markov network, 2.
-Simulating from a Markov network with spatial heterogeneity in the α parameters (sometimes called a conditional random field), and 3.
-Simulating with a dynamic model.
-
-> For this revision, I considered a number of alternative simulation methods for the dynamic model that could address the concerns of Reviewer 1 and my colleagues, but they all had serious problems.
-Using a set of GLMs as the generative model (as you suggest below) would either involve directed cycles (if β_ij≠β_ji) or would match the Markov network exactly (if β_ij=β_ji).
-Allowing directed cycles in the generative model would cause a number of problems.
-Most importantly, these models would not be identifiable from the presence-absence vectors they generated, as an infinite number of different parameter settings could all generate a given joint distribution.
-On the other hand, re-labeling the simulation network as a set of GLMs with equality constraints on the coefficients might have alleviated some readers' concerns, but only by hiding the fact that it was actually the same model.
-
 Your proposed community dynamics model would raise a host of concerns.
 You state it is a large population approximation but then include demographic stochasticity, which is typically of diminishing importance for large populations and outweighed by environmental stochasticity, which you omit.
-
-> I now refer to the differential equation as the model's deterministic backbone, rather than as an approximation.
-
 The choice of parameters to draw from is arbitrary and ultimately discernible only by reading code, and the mutualism term appears to be arbitrary.
-
-> This is true, but it was also true of all the other models I could envision where mutualism was guaranteed not to cause unbounded population explosions regardless of the parameters chosen.
-
 You may disagree, and one could debate all aspects, but that is really the point: one would have wanted a simpler and more standard model so comparison of the statistical methods is not tangled in these issues.
-
-> The simpler models are now included.
 
@@
 Trying to understand that would be a theoretical ecology exercise that, for your proposed model, does make sense to include in a paper about the statistical methods.
-
-> Based on the number of readers that have asked for something along these lines (including Reviewer 1), I think that there is some value.
-The new Figure 3 shows that the addition of population dynamics and per-capita interactions does weaken the relationship between the simulation coefficients and all of the estimation methods, but that it does not change the rank ordering of the different methods, which is an important result.
 
@@
 
-> I now do this with the first two simulation types: the first type matches the model's assumptions and the second one adds heterogeneity among sites.
-I now show that this violation decreases the models' R-squared values and increases their Type I error rates.
+> I appreciate this perspective.
+I felt very conflicted about including these simulations in the first place, and had many of the same concerns that you raise.
+I ultimately decided to include them in the previous version because a substantial number of colleagues that read the preprint had similar concerns to Reviewer 1: they wanted to see how the different methods would perform on more realistic data sets (especially on data sets that were not simulated from a model that resembled a Markov network).
+I do agree, however, that it is important to show that the model performs well on its own terms before diving into a more complicated model, and that omitting the simpler simulations made the paper worse.
+I had previously considered the option to use a set of GLMs as the generative model (as you suggest), but decided against it once I realized that this approach would either involve directed cycles and the associated identifiability problems (if β_ij≠β_ji) or would match the Markov network exactly (if β_ij=β_ji).
+
+>For this reason, I now include three different simulation regimes with different properties for the 20-species landscapes: 1.
+Simulating directly from a Markov network, 2.
+Simulating from a Markov network with spatial heterogeneity in the α parameters (sometimes called a conditional random field), and 3.
+Simulating from a process that included abundances and per-capita effects rather than species-level interactions between binary species.
+In order to avoid the problems that you highlighted with the previous submission, I simulated the abundances for the third set of landscapes using Gibbs sampling rather than a population dynamic model.
+To keep the abundances finite without adding arbitrary constraints to the new simulations, I removed mutualistic interactions from the third set of landscapes.
+
+> The overall result is that the reader can now see how the different models perform when A) the species' "true" joint distribution is fully described by univariate and bivariate potentials (e.g.
+a Markov network), when B) the joint distribution also depends on external environmental factors, and when C) the strength of the interaction between two species depends on (unobserved) variation in abundance.
 
@@
 
-> I now report approximate Type I error rates
+> I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.
+The 95% confidence intervals have about 97.5% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity [[lines and Appendix]].
+These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods [[Figure XX]].
 
@@
 
-> Showing how some of the results change with the value of beta is a great idea, and I have added several analyses along these lines in Figure 4 and Appendix 5.
-In particular, I show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta, in addition to global estimates of R-squared and Type I error rates.
-Unfortunately, I am dealing with a 210-dimensional system (20 alphas and 190 betas), which limits my options.
-Holding 209 parameters constant at arbitrary values doesn't seem like the right answer, since the results would depend on which values were chosen.
-Apart from the trivial case where all the coefficients except one are fixed at zero (in which case the indirect interactions that make this paper's methods necessary would not exist), there doesn't seem to be any particular line through the 210-dimensional space that deserves this kind of special attention.
-Likewise, there isn't a good way for completely free-form exploration of this large space.
-For this reason, the best compromise in this revision seemed like marginalizing over all but one variable at a time for these analyses.
+> Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4 and Appendix 5.
+In particular, I now show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).
+If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as an Appendix.
+
+> Beyond these steps, I've considered several other options for further exploration of the parameter space, but none of them would be very useful given the large number of parameters in these models (20 alphas and 190 betas).
+Systematically varying one of the beta coefficients while holding the other 209 parameters constant at arbitrary values would not convincingly yield generalizable insights, as different lines through the 210-dimensional space could produce different results.
 
@@
 
-- I don't think you have stated that the betas are symmetric.
+I don't think you have stated that the betas are symmetric.
 It appears that beta_ij = beta_ji and hence each could be interpreted as half of the interaction strength.
@@
 
-- Both times that I have seen the title anew, the phrase "estimating species interactions from observational data" made me think it would be about time-series method, since that is a common kind of "observational data" from which people try to estimate interactions.
+ Both times that I have seen the title anew, the phrase "estimating species interactions from observational data" made me think it would be about time-series method, since that is a common kind of "observational data" from which people try to estimate interactions.
 I recommend adding a descriptive term like "species composition data" or "observational occurrence data" or "presence/absence data." You decide.
@@
 
-- Line 44 is not stated well.
+Line 44 is not stated well.
 You could replace "begin with the assumption" by "test the hypothesis" or insert "that the hypothesis of interest is that [all pairwaise interactions...]" or some such change.
 
-- Line 69: The phrase "how groups of species can co-occur" seems imprecise or incorrect. Maybe "to determine species occurence models"?
+> I have changed the wording to emphasize hypothesis testing rather than assumptions (lines [[XX]]).
 
-- Lines 86-87 are not a good lead-in to the point of the rest of this paragraph.
+Line 69: The phrase "how groups of species can co-occur" seems imprecise or incorrect. Maybe "to determine species occurence models"?
+
+>This line [[XX]] now describes "how... conditional relationships can determine species composition" rather than referring to "groups".
+
+Lines 86-87 are not a good lead-in to the point of the rest of this paragraph.
 It sounds like you are about to talk about species interactions, again, but really the point of the paragraph is that normalizing the probabilities involves a difficult summation.
+
+>The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines [[xx]]).
 
@@
 
-> The revised manuscript now clarifies that each $\vec{y}$ vector represents
-the different species that could occur at a single site and that the full
-co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors
-indicating which species are present at each site on the landscape." This should
-clarify the relationship betweeen the $\vec{y}$ vectors and the matrix as well
-as the relationship between the single-site likelihood and the landscape-level
-likelihood (independence among sites).
+> The revised manuscript now clarifies that each $\vec{y}$ vector represents the different species that could occur at a single site and that the full co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors indicating which species are present at each site on the landscape." This should clarify the relationship between the $\vec{y}$ vectors and the full matrix as well as the relationship between the single-site likelihood and the landscape-level likelihood (i.e.
+independence among sites implies that the full likelihood equals the sum of the site-level likelihoods).
 
-- Line 130-132.
+Line 130-132.
 This statement is unclear and unconvincing.
@@
 
-- Lines 136-137. Bootstrapping is a method to estimate uncertainty of an estimation procedure. It is unclear what you mean by using it to validate if a procedure gives "stable" estimates.
+> I no longer suggest that the prior has no influence.
+I also make no claims (e.g.
+regarding unbiasedness) that would be invalidated by a non-flat prior.
+Your initial concern about the prior in my first submission was that the logistic distribution with scale 1 was too similar to the exponential distribution with rate 1 used to simulate the "true" distribution.
+Now that the prior is more than twice as wide as the "true" distribution, this is no longer an issue.
 
-> The bootstrapping results have now been removed and replaced with the confidence intervals.
+>With regard to your suggestion that I try different priors, I have now used three different ones and gotten similar results: a flat prior (for the three-species landscapes where the MLE was always finite), and two different logistic priors (for the larger communities).
 
-- Line 154: This statement is not clear.
+Lines 136-137. Bootstrapping is a method to estimate uncertainty of an estimation procedure. It is unclear what you mean by using it to validate if a procedure gives "stable" estimates.
 
-- Figure 2B: This is odd notation because, if I follow, you have chosen a null symbol (0 with a slash) to indicate the event y_i = 0 but have chosen the label y_i to indicate the event y_i = 1.
+> The bootstrapping results have now been removed and replaced with the approximate confidence intervals.
+
+Line 154: This statement is not clear.
+
+Figure 2B: This is odd notation because, if I follow, you have chosen a null symbol (0 with a slash) to indicate the event y_i = 0 but have chosen the label y_i to indicate the event y_i = 1.
 I think you want P[00], P[10], etc.
@@
 
-- Figures in general: would benefit from titles on each subfigure.
+Figures in general: would benefit from titles on each subfigure.
 
@@
 You may decide what to put in the online material since I recognize you are trying to stick to the 20-page report limit.
-(2) Include meaningful results about accuracy of uncertainties.
-These should be at your fingertips since you can calculate confidence intervals for every simulation.
+
+> As described above, I have removed the community dynamics model and now compare all the models' performance on simulated landscapes where all the assumptions are met, and in two cases where different assumptions are violated in clear ways.
+
+(2) Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation.
+
+> As described above, I now estimate the coverage of the confidence intervals, and (in the Appendix) show how coverage varies as a function of the "true" parameter value.
+
 (3) Provide some disaggregated results.
+
+> As described above, I now show how several performance metrics vary as a function of $\beta$ and have offered to include the raw results for all 7 methods across the 450 simulations.
+
 (4) Show how the regression steps work in about a page or two of supplement text, not requiring code reading.
+
+> I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either.
+With regard to the logistic regressions, I've now clarified the symmetry issue: the reader should now be able to tell that I am merely averaging the estimated regression coefficient for y_i versus y_j with the coefficient from the regression of y_j on y_i [[lines]].
+With regard to the

### 2015-11-23 Harris_2015/manuscript-materials/response-to-editor.md
almost done?

@@
 I do agree, however, that it is important to show that the model performs well on its own terms before diving into a more complicated model, and that omitting the simpler simulations made the paper worse.
-I had previously considered the option to use a set of GLMs as the generative model (as you suggest), but decided against it once I realized that this approach would either involve directed cycles and the associated identifiability problems (if β_ij≠β_ji) or would match the Markov network exactly (if β_ij=β_ji).
+I had previously considered the option to use a set of GLMs as the generative model (as you suggest), but decided against it once I realized that this approach would either involve directed cycles and the associated identifiability problems (if $\beta_{ij} \neq \beta_{ji}$) or would match the Markov network exactly (if $\beta_{ij}=\beta_{ji}$).
 
@@
 Simulating directly from a Markov network, 2.
-Simulating from a Markov network with spatial heterogeneity in the α parameters (sometimes called a conditional random field), and 3.
+Simulating from a Markov network with spatial heterogeneity in the $\alpha$ parameters (sometimes called a conditional random field), and 3.
 Simulating from a process that included abundances and per-capita effects rather than species-level interactions between binary species.
@@
 > I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.
-The 95% confidence intervals have about 97.5% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity [[lines and Appendix]].
-These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods [[Figure XX]].
+The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 5).
+These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods (lines 206-213; Figure 4C).
 
@@
 
-> Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4 and Appendix 5.
+> Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 5.
 In particular, I now show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).
-If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as an Appendix.
+If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a sixth Appendix.
 
@@
 
-> I've made this issue much clearer in the revision: see lines [[XXXX]].
+> I've made this issue much clearer in the revision: see lines 68-78.
 It should now be unambiguous that the sum in the likelihood should only include each pair once, and that each pair must share a single value for $\beta_{ij}$.
@@
 
-> I have changed the wording to emphasize hypothesis testing rather than assumptions (lines [[XX]]).
+> I have changed the wording to emphasize hypothesis testing rather than assumptions (lines 40-42).
 
@@
 
->This line [[XX]] now describes "how... conditional relationships can determine species composition" rather than referring to "groups".
+>This line (65-66) now describes "how... conditional relationships can determine species composition" rather than referring to "groups".
 
@@
 
->The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines [[xx]]).
+>The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines 82-85).
 
-- Lines 72 & 101.
+Lines 72 & 101.
 I was willing to overlook the fact that you didn't define the vector y explicitly in terms of its elements on line 72 because perhaps that is really obvious.
@@
 I also make no claims (e.g.
-regarding unbiasedness) that would be invalidated by a non-flat prior.
+regarding unbiasedness) that would be invalidated by a non-flat prior, so the strength of the prior should not affect the validity of any of the results presented.
 Your initial concern about the prior in my first submission was that the logistic distribution with scale 1 was too similar to the exponential distribution with rate 1 used to simulate the "true" distribution.
@@
 
-(2) Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation.
+2 Include meaningful results about accuracy of uncertainties. These should be at your fingertips since you can calculate confidence intervals for every simulation.
 
-> As described above, I now estimate the coverage of the confidence intervals, and (in the Appendix) show how coverage varies as a function of the "true" parameter value.
+> As described above, I now estimate the coverage of the confidence intervals, and (Appendix 5) show how coverage varies as a function of the "true" parameter value.
 
-(3) Provide some disaggregated results.
+3 Provide some disaggregated results.
 
@@
 
-(4) Show how the regression steps work in about a page or two of supplement text, not requiring code reading.
+4 Show how the regression steps work in about a page or two of supplement text, not requiring code reading.
 
-> I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either.
-With regard to the logistic regressions, I've now clarified the symmetry issue: the reader should now be able to tell that I am merely averaging the estimated regression coefficient for y_i versus y_j with the coefficient from the regression of y_j on y_i [[lines]].
-With regard to the
+> I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either:
+
+>With regard to the logistic regressions, I've now clarified the symmetry issue: the reader should now be able to tell that I am merely averaging the estimated coefficient from the regression of y_i on y_j with the coefficient from the regression of y_j on y_i (lines 142-148).
+
+>With regard to fitting the Markov network, the function being optimized is already in Equation 1 and its gradients are clearly derived in Murphy (2012; see pages 676-677 in [the linked PDF from author's website](https://www.cs.ubc.ca/~murphyk/MLbook/pml-print3-ch19.pdf)).
+The manuscript text now clarifies that the role of the package is simply to define these equations in R code and to pass them to R's general purpose optimizer (lines 90-98).

### 2015-11-23 Harris_2015/manuscript-materials/draft.Rmd
almost done?

@@
 [@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its
-gradients; the package passes these functions to the "BFGS" method in R's
-general-purpose optimizer, which then estimates the Markov network parameters.
+gradients [@murphy_machine_2012]; the package passes these functions to the
+"BFGS" method in R's general-purpose optimizer, which then estimates the Markov
+network parameters.
 
@@
 First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
-models, as described above. For the analyses with 20 species, I added a weak
-regularizer (equivalent to a logistic prior with scale 2) to ensure that the
-model estimates were always finite.
+models, as described above. For the analyses with 20 species, a weak regularizer
+(equivalent to a logistic prior with scale 2) ensured that the estimates were
+always finite.
 
@@
 the data, however [@schmidt_modeling_2012], so I used their average as an
-overall measure of their relationship.
+overall measure of the pairwise relationship.
 
@@
 @lee_learning_2012, generalized linear models closely
-approximated the Markov network estimates, especially when the data sets were
-very large [[Figure]]. As reviewed in @gotelli_empirical_2009, however, most analyses in
-this field of ecology involve fewer than 50 sites, so model performance on the smaller
-landscapes (where the gap was larger) is very important. As shown in Appendix
-5, the standard errors associated with the estimates in Figure 3 are small
-(less than 0.01), so the differences among methods should not be attributed to
-sampling error.
+approximated the Markov network estimates (Figure 4A), especially when the data
+sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
+however, most analyses in this field of ecology involve fewer than 50 sites,
+and the gap between the Markov network and GLMs is larger in this context. As
+shown in Appendix 5, the standard errors associated with the estimates in Figure
+3 are small (less than 0.01), so the differences among methods should not be
+attributed to sampling error.
 
@@
 interval coverage across a range of $\beta_{ij}$ values). The null model's
-Type I error rates were [[XX]] and [[XX]] for the constant and heterogeneous
+Type I error rates were 0.30 and 0.50 for the constant and heterogeneous
 landscapes, respectively---far higher than the nominal 0.05 rate. Figure
@@
 the null model or the Markov network will predict the wrong sign of the
-interaction with 95% confidence.  The null model makes such errors [[XX]] times
+interaction with 95% confidence.  The null model makes such errors more than 5 times
 as often as the Markov network, even though it only reject the null hypothesis
-[[XX]] as often overall (Appendix 5).  The Markov network's errors were also
+twice as often overall (Appendix 5).  The Markov network's errors were also
 more concentrated around 0, so it never misclassified strong interactions like
@@
 linear models can both provide computationally efficient approximations, but
-with somewhat lower accuracy, especially for typically-sized data sets with small
-numbers of sites [@gotelli_empirical_2009]. The difference in accuracy may be
+with somewhat lower accuracy [especially for typically-sized data sets with small
+numbers of sites; @gotelli_empirical_2009]. The difference in accuracy may be
 larger for real data sets than for the simulated landscapes in Figure 3,
@@
 history. These other sources of information could also be used to augment the
-likelihood function with an informative prior distribution, reducing ecologists'
+likelihood function with a more informative prior distribution, reducing ecologists'
 error and uncertainty relative to Figure 3's results.
@@
 \setlength{\leftskip}{1em}
+\setlength{\parskip}{0pt}
 
@@
 \setlength{\leftskip}{0cm}
+
+\noindent \textbf{Figures:}
 
@@
 each method versus number of sampled locations across the three simulation
-types. For the null model (Pairs), these values include the removal of two
-outliers with $|Z|>1000$ [[Appendix]].
+types. For the null model (Pairs), two outliers with $|Z|>1000$ were manually
+adjusted to $|Z|=50$ to mitigate their detrimental influence on $R^2$ (Appendix
+5).
 

### 2015-11-23 Harris_2015/manuscript-materials/response-to-editor.md
clean up appendix 5 etc

@@
+> Dear Dr. de Valpine,
+
+> [[stuff]]
+
+> Sincerely,
+
+> David J. Harris
+
 Dear Mr. Harris,
@@
 
-> I appreciate this perspective.
+* I appreciate this perspective.
 I felt very conflicted about including these simulations in the first place, and had many of the same concerns that you raise.
@@
 
-> I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.
+* I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.
 The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 5).
@@
 
-> Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 5.
-In particular, I now show how CI coverage and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).
+* Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 5.
+In particular, I now show how CI coverage, RMSE, and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).
 If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a sixth Appendix.
@@
 
-> I've made this issue much clearer in the revision: see lines 68-78.
+* I've made this issue much clearer in the revision: see lines 68-78.
 It should now be unambiguous that the sum in the likelihood should only include each pair once, and that each pair must share a single value for $\beta_{ij}$.
@@
 
-> This is another good point.
+* This is another good point.
 I've changed the title to focus on co-occurrence data.
@@
 
-> I have changed the wording to emphasize hypothesis testing rather than assumptions (lines 40-42).
+* I have changed the wording to emphasize hypothesis testing rather than assumptions (lines 40-42).
 
@@
 
->This line (65-66) now describes "how... conditional relationships can determine species composition" rather than referring to "groups".
+* Lines 65-66 now describe "how... conditional relationships can determine species composition" rather than referring to "groups".
 
@@
 
->The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines 82-85).
+* The discussion of computational difficulties has been removed to save space, and the lines you noted are now at the end of a paragraph rather than the beginning of one (lines 82-85).
 
@@
 
-> The revised manuscript now clarifies that each $\vec{y}$ vector represents the different species that could occur at a single site and that the full co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors indicating which species are present at each site on the landscape." This should clarify the relationship between the $\vec{y}$ vectors and the full matrix as well as the relationship between the single-site likelihood and the landscape-level likelihood (i.e.
+* The revised manuscript now clarifies that each $\vec{y}$ vector represents the different species that could occur at a single site and that the full co-occurrence matrix is made of "a set of independent $\vec{y}$ vectors indicating which species are present at each site on the landscape" (lines 89-90).
+This should clarify the relationship between the $\vec{y}$ vectors and the full matrix as well as the relationship between the single-site likelihood of Equation 1 and the landscape-level likelihood (i.e.
 independence among sites implies that the full likelihood equals the sum of the site-level likelihoods).
@@
 
-> I no longer suggest that the prior has no influence.
+* I no longer suggest that the prior has no influence.
 I also make no claims (e.g.
@@
 
->With regard to your suggestion that I try different priors, I have now used three different ones and gotten similar results: a flat prior (for the three-species landscapes where the MLE was always finite), and two different logistic priors (for the larger communities).
+> With regard to your suggestion that I try different priors, I have now used three different ones and gotten similar results: a flat prior (for the three-species landscapes where the MLE was always finite), and two different logistic priors (for the larger communities).
 
@@
 
-> The bootstrapping results have now been removed and replaced with the approximate confidence intervals.
+* The bootstrapping results have now been removed and replaced with the approximate confidence intervals.
 
 Line 154: This statement is not clear.
+
+* See lines 144-148 for a clarified version
 
@@
 
-> This is an excellent point. The figure has been fixed (along with the caption).
+* This is an excellent point. The figure has been fixed (along with the caption).
 
@@
 
-> Subfigures now have titles
+* Subfigures now have titles
 
@@
 
-> As described above, I have removed the community dynamics model and now compare all the models' performance on simulated landscapes where all the assumptions are met, and in two cases where different assumptions are violated in clear ways.
+* As described above, I have removed the community dynamics model and now compare all the models' performance on simulated landscapes where all the assumptions are met, and in two cases where different assumptions are violated in clear ways.
 
@@
 
-> As described above, I now estimate the coverage of the confidence intervals, and (Appendix 5) show how coverage varies as a function of the "true" parameter value.
+* As described above, I now estimate the coverage of the confidence intervals, and (Appendix 5) show how coverage varies as a function of the "true" parameter value.
 
@@
 
-> As described above, I now show how several performance metrics vary as a function of $\beta$ and have offered to include the raw results for all 7 methods across the 450 simulations.
+* As described above, I now show how several performance metrics vary as a function of $\beta$ and have offered to include the raw results for all 7 methods across the 450 simulations.
 
@@
 
-> I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either:
+* I wasn't entirely sure whether you were referring to the logistic regression models or to the process of fitting the Markov network, but I believe that I've clarified the discussion of both issues sufficiently that supplemental material is no longer required for either:
 

### 2015-11-23 Harris_2015/show-results.Rmd
clean up appendix 5 etc

# Appendix 5: results

```{r, message = FALSE}
library(dplyr)
library(mgcv)
library(ggplot2)
library(tidyr)
library(knitr)
library(lme4)
```

## Import the results from Appendix 4:

```{r, cache = TRUE}
x = read.csv("estimates.csv", stringsAsFactors = FALSE)
x$simulation_type = gsub("[0-9]", "", x$rep_name)
```

## Import the results from the *Pairs* software:

```{r, cache = TRUE}

pairs_txt = readLines("fakedata/matrices/Pairs.txt")
library(stringr)

# Find areas of the data file that correspond
# to species pairs' results
beginnings = grep("Sp1", pairs_txt) + 1
ends = c(
  grep("^[^ ]", pairs_txt)[-1],
  length(pairs_txt) + 1
) - 1

partial_names = sapply(strsplit(grep("^>", pairs_txt, value = TRUE), " +"), function(x) x[[3]])
filename_lines = grep("^>", pairs_txt)

# Sort a vector of alphanumeric strings by the numeric component
# as if they were integers.  For example, V20 is larger than V12,
# even though V12 comes first alphabetically
alnum_sort = function(x){
  raw = as.integer(gsub("[[:alpha:]]", "", x))
  x[order(raw)]
}

pairs_results = lapply(
  1:length(filename_lines),
  function(i){

    n_sites = as.integer(strsplit(partial_names[[i]], "-")[[1]][[1]])
    rep_name = strsplit(partial_names[[i]], "-")[[1]][[2]]

    # Find the line where the current data set is mentioned in
    # pairs.txt
    filename_line = filename_lines[i]

    # Which chunk of the data file corresponds to this file?
    chunk = min(which(beginnings > filename_line))

    # Split the chunk on whitespace.
    splitted = strsplit(pairs_txt[beginnings[chunk]:ends[chunk]], " +")

    # Pull out the corresponding chunk of the "x" data frame, based on n_sites and rep_name
    x_subset = x[x$n_sites == n_sites & x$rep_name == rep_name & x$method == "correlation", ]

    # Pull out the species numbers and their Z-scores, then join to x_subset
    pairs_results = lapply(
      splitted,
      function(x){
        # in the x data frame, species 1 is always a lower number than species 2
        spp = alnum_sort(x[3:4])
        data.frame(
          sp1 = spp[1], 
          sp2 = spp[2], 
          z = x[14],
          stringsAsFactors = FALSE
        )
      }
    ) %>% 
      bind_rows %>%
      mutate(spp = paste(sp1, sp2, sep = "-"))

    n_spp = 20

    pairs_results$z = as.numeric(pairs_results$z)

    # Re-order the pairs_results to match the other methods
    m = matrix(NA, n_spp, n_spp)
    new_order = match(
      paste0("V", row(m)[upper.tri(m)], "-V", col(m)[upper.tri(m)]),
      pairs_results$spp
    )
    ordered_pairs_results = pairs_results[na.omit(new_order), ]

    ordered_pairs_results = ordered_pairs_results %>%
      filter(sp1 %in% c(x_subset$sp1) & sp2 %in% x_subset$sp2)

    x_subset$estimate = ordered_pairs_results$z
    x_subset$method = "null"

    x_subset
  }
) %>% bind_rows()

# Manually adjust the Z values less than -1000 so that these outliers
# won't completely dominate the analyses below
pairs_results$estimate[pairs_results$estimate < -1000] = -50

x = rbind(x, pairs_results)

```

## Calculate model performance:

```{r, cache = TRUE}
resids = function(data){
  resid(lm(truth ~ estimate + 0, data = data))
}

result_summary = x %>% 
  group_by(method, simulation_type) %>%
  do(data.frame(., resids = resids(.))) %>%
  ungroup %>%
  group_by(method, simulation_type, n_sites) %>%
  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))

result_summary$method = reorder(result_summary$method, -result_summary$r2)

result_summary$simulation_type = reorder(result_summary$simulation_type, -result_summary$r2)

result_summary = result_summary %>% 
  group_by(method) %>% 
  summarise(mean_r2 = round(100 * mean(r2))) %>%
  mutate(method_r2 = paste0(method, " (0.", mean_r2, ")")) %>%
  select(method, method_r2) %>%
  inner_join(result_summary, "method")

result_summary$simulation_type_long = plyr::revalue(
  result_summary$simulation_type,
  c(no_env = "constant environment",
    env = "heterogeneous environment",
    abund = "abundance")
)

result_summary$method_r2 = reorder(result_summary$method_r2, -result_summary$r2)
```

#### Model performance with no environmental variation:

```{r}
result_summary %>% 
  filter(simulation_type == "no_env") %>%
  group_by(method) %>% 
  summarise(mean(r2)) %>% 
  kable(digits = 3)
```

#### Model performance with environmental heterogeneity:

```{r}
result_summary %>% 
  filter(simulation_type == "env") %>%
  group_by(method) %>% 
  summarise(mean(r2)) %>% 
  kable(digits = 3)
```

#### Model performance with per-capita species interactions:

```{r}
result_summary %>% 
  filter(simulation_type == "abund") %>%
  group_by(method) %>% 
  summarise(mean(r2)) %>% 
  kable(digits = 3)
```

## Plot the R-squared results (Figure 3)

```{r}

legend_name = expression(Method~(mean~R^2))

pdf("manuscript-materials/figures/performance.pdf", width = 8, height = 2.5)
ggplot(result_summary, aes(x = n_sites, y = r2, col = method_r2, shape = method_r2)) + 
  facet_grid(~simulation_type_long) + 
  geom_line(size = .5) + 
  geom_point(size = 2.5, fill = "white") + 
  scale_shape_manual(values = c(16, 22, 17, 23, 18, 24, 15), name = legend_name) +
  geom_hline(yintercept = 0, size = 1/2) +
  geom_vline(xintercept = 0, size = 1) + 
  #scale_shape_manual(values = c(21, 25, 15, 18)) + 
  coord_cartesian(ylim = c(-.01, 0.76)) + 
  ylab(expression(R^2)) + 
  xlab("Number of sites (log scale)") + 
  scale_x_log10(breaks = unique(x$n_sites), limits = range(x$n_sites)) +
  theme_bw(base_size = 11) + 
  theme(panel.margin = grid::unit(1.25, "lines")) + 
  theme(panel.border = element_blank(), axis.line = element_blank()) + 
  theme(
    panel.grid.minor = element_blank(), 
    panel.grid.major.y = element_line(color = "lightgray", size = 1/4), 
    panel.grid.major.x = element_blank()
  ) + 
  theme(strip.background = element_blank(), legend.key = element_blank()) + 
  theme(plot.margin = grid::unit(c(.01, .01, .75, .1), "lines")) + 
  theme(
    axis.title.x = element_text(vjust = -0.2, size = 12), 
    axis.title.y = element_text(angle = 0, hjust = -.1, size = 12)
  ) + 
  scale_color_brewer(palette = "Dark2", name = legend_name)
dev.off()
```

## Estimate R-squared uncertainty:

Fit a linear mixed model describing R-squared as a function of method, landscape
size, and simulation type.  Note the small standard errors associated with the
effect of estimation method.

```{r, cache = TRUE}
landscape_estimates = x %>% 
  group_by(method, simulation_type) %>%
  do(data.frame(., resids = resids(.))) %>%
  ungroup %>%
  group_by(method, simulation_type, rep_name, n_sites) %>%
  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))

summary(lmer(r2 ~ method + n_sites + simulation_type + (1|rep_name), data = landscape_estimates))
```

## Summarize the inferential results of the Markov network and the null model:

```{r, cache = TRUE}

pairs_summary = x[x$method == "null" & !grepl("pop", x$rep_name), ]
markov_summary = x[x$method == "Markov network" & !grepl("pop", x$rep_name), ]

# Pairs's Z score is based on C-scores, which are positive when species are 
# disaggregated.  So significantly positive interactions == negative estimates
pairs_summary$sig_pos = pairs_summary$estimate < qnorm(.025)
pairs_summary$sig_neg = pairs_summary$estimate > qnorm(.975)

# Markov network is significant when lower bound is above zero or lower bound is
# below zero
markov_summary$sig_pos = markov_summary$lower > 0
markov_summary$sig_neg = markov_summary$upper < 0
```

## Create Figure 4:

```{r, cache = TRUE}

# Function for smoothing the error rates with a generalized additive model
# and plotting the results
my_geom_smooth = function(data, color, ...){
  geom_smooth(
    data = data,
    aes(
      x = truth, 
      y = as.integer((sig_neg & truth > 0) | (sig_pos & truth < 0)),
      color = color
    ),
    method = gam, 
    family = binomial, 
    formula = y ~ s(x),
    se = FALSE,
    n = 1024,
    ...
  )
}

truth_seq = seq(min(markov_summary$truth), max(markov_summary$truth), length = 1000)

plotfun = function(..., add){
  if(add){
    lines(...)
  }else{
    plot(...)
  }
}

error_smoother = function(data){
  predict(
    gam(
      I((sig_neg & truth > 0) | (sig_pos & truth < 0)) ~ s(truth),
      data = data,
      family = binomial
    ),
    data.frame(truth = truth_seq),
    type = "response"
  )
}

y_pairs = error_smoother(pairs_summary)
y_markov = error_smoother(markov_summary)

pdf("manuscript-materials/figures/error_rates.pdf", height = 8.5, width = 8.5/3)
par(mfrow = c(3, 1))

# Compare estimates
spread_estimates = x %>%
  dplyr::select(-lower, -upper, -X) %>%
  spread(method, estimate) %>%
  na.omit()

# R-squared for null versus correlation
round(summary(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))$r.squared, 2)

# R-squared for glm markov network versus glm
round(summary(lm(`Markov network` ~ GLM, data = spread_estimates))$r.squared, 2)

with(
  spread_estimates, 
  plot(
    `Markov network`, 
    GLM, 
    pch = ".", 
    col = "#00000020",
    ylab = "Markov network estimate",
    xlab = "GLM estimate",
    bty = "l"
  )
)
mtext("A. Markov network estimates\nvs. GLM estimates", adj = 0, side = 3, font = 2, line = 1.2, cex = .9) 
abline(lm(`Markov network` ~ GLM, data = spread_estimates))
text(0, 5, expression(R^2==0.94))

with(
  spread_estimates, 
  plot(
    correlation * sqrt(n_sites), 
    null, 
    pch = ".", 
    col = "#00000020",
    ylab = "Z-score",
    xlab = expression("correlation" %*% sqrt(number~~of~~sites)),
    bty = "l"
  )
)
mtext("B. Null model estimates vs.\nscaled correlation coefficients", adj = 0, side = 3, font = 2, line = 1.2, cex = .9) 
abline(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))
text(10, 12, expression(R^2==0.95))

plotfun(
  truth_seq,
  y_pairs,
  type = "l",
  xlab = "\"True\" interaction strength",
  ylab = "P(confidently predict wrong sign)",
  bty = "l",
  yaxs = "i",
  col = 2,
  add = FALSE,
  ylim = c(0, .4),
  lwd = 2
)
mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, font = 2, line = 1.2, cex = .9) 
plotfun(truth_seq, y_markov, add = TRUE, lwd = 2)
legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), col = c(2, 1), bty = "n")
dev.off()
```

## Summarize inferential statistics:

```{r}
# P(Pairs confidently wrong)
with(pairs_summary,  mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
# P(Markov network confidently wrong)
with(markov_summary, mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))

# P(Pairs rejects null)
with(pairs_summary,  mean(sig_neg | sig_pos))

# P(Markov network rejects null)
with(markov_summary,  mean(sig_neg | sig_pos))

# P(reject null | small interaction) for Makrov network
# (approximate Type I error rates)
markov_summary %>%
  group_by(simulation_type) %>% 
  filter(abs(truth) < .1) %>% 
  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
  kable(digits = 3)

# P(reject null | small interaction) for Pairs
# (approximate Type I error rates)
pairs_summary %>%
  group_by(simulation_type) %>% 
  filter(abs(truth) < .1) %>% 
  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
  kable(digits = 3)

# Confidence interval coverage across all values
# What's the probability that any "true" value falls inside the 95% CI?
markov_summary %>% 
  group_by(simulation_type) %>% 
  summarize(mean(truth > lower & truth < upper)) %>%
  kable(digits = 3)
```

## Plot root mean square error versus "true" beta

Exclude the "abundance" simulations because the interpretation of $\beta$ is different.

```{r, cache = TRUE}
rmses = markov_summary %>% filter(simulation_type != "abund") %>% mutate(rmse = sqrt((truth - estimate)^2))

ggplot(rmses, aes(x = truth, y = rmse)) + geom_point(alpha = .2, size = 1) + geom_smooth() + theme_bw() + coord_cartesian(ylim = c(0, max(rmses$rmse)))
```

## Plot confidence interval coverage versus true $\beta$ value

```{r, cache = TRUE}
# Coverage versus true beta value.  The top and bottom 0.5%
# of the distribution have been omitted to prevent bad behavior
# by the smoother in the tails.
markov_summary %>%
  filter(percent_rank(truth) > .005 & percent_rank(truth) < .995) %>%
  mutate(covered = truth > lower & truth < upper) %>%
  ggplot(aes(x = truth, y = as.integer(covered))) + 
  facet_grid(~simulation_type) + 
  geom_smooth(method = gam, formula = y ~ s(x), family = binomial) + 
  theme_bw() + 
  coord_cartesian(ylim = c(0, 1)) + 
  geom_vline(xintercept = 0) + 
  geom_hline(yintercept = 0.95, color = "red") + 
  ylab("Coverage")

```

### 2015-11-25 Harris_2015/manuscript-materials/response-to-editor.md
as resubmitted

@@
 
-> [[stuff]]
+> Thank you for your insightful and thorough comments.  I have addressed each of them below.
 
@@
 * I now report approximate Type I error rates and the coverage of the model's 95% confidence intervals.
-The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 5).
+The 95% confidence intervals have about 98% coverage when the data was generated from an unmodified Markov network, and about 86% coverage in the presence of environmental heterogeneity (lines 201-206 and Appendix 4).
 These coverage values aren't perfect, but they are reasonable, and they provide far better inferences than ecologists' current methods (lines 206-213; Figure 4C).
@@
 
-* Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 5.
+* Showing the results' dependence on the value of beta is a great idea, and I have added several analyses along these lines in Figure 4C and Appendix 4.
 In particular, I now show how CI coverage, RMSE, and the probability that the model will predict the wrong sign for an interaction vary as a function of beta (in addition to global estimates of R-squared and Type I error rates).
-If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a sixth Appendix.
+If you think that readers would benefit from even more disaggregation, I could include a data file containing the seven models' estimates for all the species pairs across all 450 simulated landscapes as a fifth Appendix.
 
@@
 
-* As described above, I now estimate the coverage of the confidence intervals, and (Appendix 5) show how coverage varies as a function of the "true" parameter value.
+* As described above, I now estimate the coverage of the confidence intervals, and (Appendix 4) show how coverage varies as a function of the "true" parameter value.
 

### 2015-11-25 Harris_2015/manuscript-materials/draft.Rmd
as resubmitted

@@
 I compared seven techniques for determining the sign and strength of the
-associations between pairs of species from simulated data (Appendix 4).
+associations between pairs of species from simulated data (Appendix 3).
 First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
@@
 and the gap between the Markov network and GLMs is larger in this context. As
-shown in Appendix 5, the standard errors associated with the estimates in Figure
+shown in Appendix 4, the standard errors associated with the estimates in Figure
 3 are small (less than 0.01), so the differences among methods should not be
@@
 simulations that matched the model's assumptions, versus 0.14 for simulations
-that included environmental heterogeneity (see Appendix 5 for confidence
+that included environmental heterogeneity (see Appendix 4 for confidence
 interval coverage across a range of $\beta_{ij}$ values). The null model's
@@
 as often as the Markov network, even though it only reject the null hypothesis
-twice as often overall (Appendix 5).  The Markov network's errors were also
+twice as often overall (Appendix 4).  The Markov network's errors were also
 more concentrated around 0, so it never misclassified strong interactions like

### 2015-11-25 Harris_2015/manuscript-materials/draft.tex
as resubmitted

@@
 Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
-gradients; the package passes these functions to the ``BFGS'' method in
-R's general-purpose optimizer, which then estimates the Markov network
-parameters.
+gradients (Murphy 2012); the package passes these functions to the
+``BFGS'' method in R's general-purpose optimizer, which then estimates
+the Markov network parameters.
 
@@
 I compared seven techniques for determining the sign and strength of the
-associations between pairs of species from simulated data (Appendix 4).
+associations between pairs of species from simulated data (Appendix 3).
 First, I used the rosalia package (Harris 2015a) to fit Markov network
-models, as described above. For the analyses with 20 species, I added a
-weak regularizer (equivalent to a logistic prior with scale 2) to ensure
-that the model estimates were always finite.
+models, as described above. For the analyses with 20 species, a weak
+regularizer (equivalent to a logistic prior with scale 2) ensured that
+the estimates were always finite.
 
@@
 identifiable from the data, however (Schmidt and Murphy 2012), so I used
-their average as an overall measure of their relationship.
+their average as an overall measure of the pairwise relationship.
 
@@
 anticipated by Lee and Hastie (2012), generalized linear models closely
-approximated the Markov network estimates, especially when the data sets
-were very large {[}{[}Figure{]}{]}. As reviewed in Gotelli and Ulrich
-(2009), however, most analyses in this field of ecology involve fewer
-than 50 sites, so model performance on the smaller landscapes (where the
-gap was larger) is very important. As shown in Appendix 5, the standard
-errors associated with the estimates in Figure 3 are small (less than
-0.01), so the differences among methods should not be attributed to
-sampling error.
+approximated the Markov network estimates (Figure 4A), especially when
+the data sets were very large (Figure 3). As reviewed in Gotelli and
+Ulrich (2009), however, most analyses in this field of ecology involve
+fewer than 50 sites, and the gap between the Markov network and GLMs is
+larger in this context. As shown in Appendix 4, the standard errors
+associated with the estimates in Figure 3 are small (less than 0.01), so
+the differences among methods should not be attributed to sampling
+error.
 
@@
 model's assumptions, versus 0.14 for simulations that included
-environmental heterogeneity (see Appendix 5 for confidence interval
+environmental heterogeneity (see Appendix 4 for confidence interval
 coverage across a range of \(\beta_{ij}\) values). The null model's Type
-I error rates were {[}{[}XX{]}{]} and {[}{[}XX{]}{]} for the constant
-and heterogeneous landscapes, respectively---far higher than the nominal
-0.05 rate. Figure 4C shows, across a range of true interaction
-strengths, the probability that the null model or the Markov network
-will predict the wrong sign of the interaction with 95\% confidence. The
-null model makes such errors {[}{[}XX{]}{]} times as often as the Markov
-network, even though it only reject the null hypothesis {[}{[}XX{]}{]}
-as often overall (Appendix 5). The Markov network's errors were also
-more concentrated around 0, so it never misclassified strong
-interactions like the null model did.
+I error rates were 0.30 and 0.50 for the constant and heterogeneous
+landscapes, respectively---far higher than the nominal 0.05 rate. Figure
+4C shows, across a range of true interaction strengths, the probability
+that the null model or the Markov network will predict the wrong sign of
+the interaction with 95\% confidence. The null model makes such errors
+more than 5 times as often as the Markov network, even though it only
+reject the null hypothesis twice as often overall (Appendix 4). The
+Markov network's errors were also more concentrated around 0, so it
+never misclassified strong interactions like the null model did.
 
@@
 both provide computationally efficient approximations, but with somewhat
-lower accuracy, especially for typically-sized data sets with small
-numbers of sites (Gotelli and Ulrich 2009). The difference in accuracy
+lower accuracy (especially for typically-sized data sets with small
+numbers of sites; Gotelli and Ulrich 2009). The difference in accuracy
 may be larger for real data sets than for the simulated landscapes in
@@
 manipulative experiments, or natural history. These other sources of
-information could also be used to augment the likelihood function with
-an informative prior distribution, reducing ecologists' error and
+information could also be used to augment the likelihood function with a
+more informative prior distribution, reducing ecologists' error and
 uncertainty relative to Figure 3's results.
@@
 \setlength{\parindent}{0cm}
+\setlength{\parskip}{0pt}
 
@@
 \setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
-
-\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}
-
-\textbf{Figure 1: A.} A small network of three competing species. The
-tree (top) tends not to co-occur with either of the two shrub species,
-as indicated by the strongly negative coefficient linking them. The two
-shrub species also compete with one another, but more weakly (circled
-coefficient). \textbf{B.} In spite of the competitive interactions
-between the two shrub species, their shared tendency to occur in
-locations without trees makes their occurrence vectors positively
-correlated (circled). \textbf{C.} Controlling for trees with a
-conditional (all-else-equal) approach such as a partial covariance or a
-Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled).
-
-\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
-and \(\beta\) values. The abiotic environment favors the occurrence of
-each species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
-consistent with competition between the two species. \textbf{B.} The
-coefficients determine the probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
-is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
-but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
-includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
-and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
-The normalizing constant \(Z\), ensures that the four probabilities sum
-to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest, as calculated in the previous panel. \textbf{D.}
-Without competition (i.e.~with \(\beta_{12}=0\), each species would
-occur more often.
-
-\textbf{Figure 3:} Proportion of variance in interaction coefficients
-explained by each method versus number of sampled locations across the
-three simulation types. For the null model (Pairs), these values include
-the removal of two outliers with \(|Z|>1000\) {[}{[}Appendix{]}{]}.
-
-\textbf{Figure 4: A.} The Markov network's estimated interaction
-coefficients were generally very similar to the GLM estimates.
-\textbf{B.} The null model's estimates typically matched the (negative)
-correlation coefficient, after controlling for landscape size.
-\textbf{C.} For any given interaction strength, the null model was much
-more likely to misclassify its sign with 95\% confidence than the Markov
-network was.
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
+
+\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}
+
+\noindent \textbf{Figures:}
+
+\textbf{Figure 1: A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a
+Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled).
+
+\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
+and \(\beta\) values. The abiotic environment favors the occurrence of
+each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
+consistent with competition between the two species. \textbf{B.} The
+coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
+is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
+but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
+includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
+and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
+The normalizing constant \(Z\), ensures that the four probabilities sum
+to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest, as calculated in the previous panel. \textbf{D.}
+Without competition (i.e.~with \(\beta_{12}=0\), each species would
+occur more often.
+
+\textbf{Figure 3:} Proportion of variance in interaction coefficients
+explained by each method versus number of sampled locations across the
+three simulation types. For the null model (Pairs), two outliers with
+\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
+detrimental influence on \(R^2\) (Appendix 5).
+
+\textbf{Figure 4: A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates.
+\textbf{B.} The null model's estimates typically matched the (negative)
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was.
+
 \end{flushleft}

### 2015-11-25 Harris_2015/recovery.Rmd
fix naming
as resubmitted

@@
 title: "Appendix 3: Estimating species interactions"
-subtitle: \"Estimating species interactions from observational data with Markov networks\"
+subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
 author: David J. Harris
@@
 from Appendix 2 and how each model's performance was evaluated.[^1]
+
+Note that the `pairs` program was run separately (outside of R) with the following options:
+
+  * Batch mode
+  * Sequential swap ("s")
+  * Printing all pairs ("y")
+  * C-score co-occurrence measure ("c")
+  * Default confidence limits (0.05)
+  * Default iterations (100)
+  * Maximum of 20 species
 
@@
 
-mc.cores = 2
+mc.cores = 8
 
@@
 ```
-
-Load in the results from the `pairs` program, run outside of R with
-the following options:
-
-  * Batch mode
-  * Sequential swap ("s")
-  * Printing all pairs ("y")
-  * C-score co-occurrence measure ("c")
-  * Default confidence limits (0.05)
-  * Default iterations (100)
-  * Maximum of 20 species
 
@@
   splitname = strsplit(identifier, "/|-|\\.")[[1]]
-  n_spp = as.integer(splitname[[1]]) # simulated species, NOT present species
-  n_sites = as.integer(splitname[[2]])
-  rep_name = splitname[[3]]
+  n_sites = as.integer(splitname[[1]])
+  rep_name = splitname[[2]]
 
@@
 
-```{r, cache = FALSE, eval = TRUE}
+```{r, cache = FALSE, eval = FALSE}
 

### 2015-11-25 Harris_2015/recovery.tex
as resubmitted

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\RegionMarkerTok}[1]{{#1}}
-\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
-\newcommand{\NormalTok}[1]{{#1}}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\ImportTok}[1]{{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
+\newcommand{\BuiltInTok}[1]{{#1}}
+\newcommand{\ExtensionTok}[1]{{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
+\newcommand{\RegionMarkerTok}[1]{{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
+\newcommand{\NormalTok}[1]{{#1}}
 \setlength{\parindent}{0pt}
@@
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
@@
   \posttitle{\par}
-\subtitle{"Estimating species interactions from observational data with Markov
-networks"}
+\subtitle{Inferring species interactions from co-occurrence data with Markov
+networks}
   \author{David J. Harris}
@@
 
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
 \begin{document}
-
 \maketitle
@@
 
-Initialization:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(dplyr)        }\CommentTok{# For manipulating data structures}
-\KeywordTok{library}\NormalTok{(corpcor)      }\CommentTok{# For regularized partial covariances}
-\KeywordTok{library}\NormalTok{(rosalia)      }\CommentTok{# For Markov networks}
-\KeywordTok{library}\NormalTok{(arm)          }\CommentTok{# For regularized logistic regression}
-\KeywordTok{library}\NormalTok{(BayesComm)    }\CommentTok{# For joint species distribution modeling}
-\KeywordTok{library}\NormalTok{(RColorBrewer) }\CommentTok{# For color palette}
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Load in the results from the \texttt{pairs} program, run outside of R
+Note that the \texttt{pairs} program was run separately (outside of R)
 with the following options:
@@
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
@@
 
+Initialization:
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\NormalTok{pairs_txt =}\StringTok{ }\KeywordTok{readLines}\NormalTok{(}\StringTok{"fakedata/Pairs.txt"}\NormalTok{)}
+\NormalTok{mc.cores =}\StringTok{ }\DecValTok{8}
 
-\CommentTok{# Find areas of the data file that correspond}
-\CommentTok{# to species pairs' results}
-\NormalTok{beginnings =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}\StringTok{"Sp1"}\NormalTok{, pairs_txt) +}\StringTok{ }\DecValTok{1}
-\NormalTok{ends =}\StringTok{ }\KeywordTok{c}\NormalTok{(}
-  \KeywordTok{grep}\NormalTok{(}\StringTok{"^[^ ]"}\NormalTok{, pairs_txt)[-}\DecValTok{1}\NormalTok{],}
-  \KeywordTok{length}\NormalTok{(pairs_txt)}
-\NormalTok{) -}\StringTok{ }\DecValTok{1}
-
-\CommentTok{# The above code fails on the very last line of the file}
-\NormalTok{ends[}\KeywordTok{length}\NormalTok{(ends)] =}\StringTok{ }\NormalTok{ends[}\KeywordTok{length}\NormalTok{(ends)] +}\StringTok{ }\DecValTok{1}
+\KeywordTok{library}\NormalTok{(dplyr)        }\CommentTok{# For manipulating data structures}
+\KeywordTok{library}\NormalTok{(corpcor)      }\CommentTok{# For regularized partial covariances}
+\KeywordTok{library}\NormalTok{(rosalia)      }\CommentTok{# For Markov networks}
+\KeywordTok{library}\NormalTok{(arm)          }\CommentTok{# For regularized logistic regression}
+\KeywordTok{library}\NormalTok{(BayesComm)    }\CommentTok{# For joint species distribution modeling}
+\KeywordTok{library}\NormalTok{(parallel)     }\CommentTok{# for mclapply}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{fit_all =}\StringTok{ }\NormalTok{function(filename)\{}
+\NormalTok{fit_all =}\StringTok{ }\NormalTok{function(identifier)\{}
   \NormalTok{######## Import ########}
 
-  \CommentTok{# Multiplying by one is necessary to prevent silly errors}
-  \CommentTok{# regarding TRUE/FALSE versus 1/0}
-  \NormalTok{raw_obs =}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(filename)[[}\StringTok{"observed"}\NormalTok{]] *}\StringTok{ }\DecValTok{1}
+  \NormalTok{data_filename =}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"fakedata/matrices/"}\NormalTok{, identifier, }\StringTok{".csv"}\NormalTok{)}
+  \NormalTok{truth_filename =}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"fakedata/truths/"}\NormalTok{, identifier, }\StringTok{".txt"}\NormalTok{)}
+
+  \CommentTok{# first column is row numbers; drop it}
+  \NormalTok{raw_obs =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{read.csv}\NormalTok{(data_filename)[ , -}\DecValTok{1}\NormalTok{])}
 
@@
   \NormalTok{x =}\StringTok{ }\NormalTok{raw_obs[ , species_is_variable]}
-  \NormalTok{truth =}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(filename)[[}\StringTok{"truth"}\NormalTok{]][pair_is_variable[}\KeywordTok{upper.tri}\NormalTok{(pair_is_variable)]]}
+  \NormalTok{truth =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(}\KeywordTok{read.table}\NormalTok{(truth_filename))[pair_is_variable[}\KeywordTok{upper.tri}\NormalTok{(pair_is_variable)]]}
 
-  \NormalTok{splitname =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(filename, }\StringTok{"/|-|}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{."}\NormalTok{)[[}\DecValTok{1}\NormalTok{]]}
-  \NormalTok{n_sites =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(splitname[[}\DecValTok{3}\NormalTok{]])}
-  \NormalTok{rep =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(splitname[[}\DecValTok{4}\NormalTok{]])}
+  \NormalTok{splitname =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(identifier, }\StringTok{"/|-|}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{."}\NormalTok{)[[}\DecValTok{1}\NormalTok{]]}
+  \NormalTok{n_sites =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(splitname[[}\DecValTok{1}\NormalTok{]])}
+  \NormalTok{rep_name =}\StringTok{ }\NormalTok{splitname[[}\DecValTok{2}\NormalTok{]]}
+
+  \CommentTok{# Species IDs}
+  \NormalTok{sp1 =}\StringTok{ }\KeywordTok{combn}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(x), }\DecValTok{2}\NormalTok{)[}\DecValTok{1}\NormalTok{, ]}
+  \NormalTok{sp2 =}\StringTok{ }\KeywordTok{combn}\NormalTok{(}\KeywordTok{colnames}\NormalTok{(x), }\DecValTok{2}\NormalTok{)[}\DecValTok{2}\NormalTok{, ]}
 
@@
   \NormalTok{corr =}\StringTok{ }\KeywordTok{cor}\NormalTok{(x)}
-
   \NormalTok{######## GLM ########}
@@
   \NormalTok{######## Markov network ########}
-  \NormalTok{rosie =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{maxit =} \DecValTok{200}\NormalTok{, }\DataTypeTok{trace =} \DecValTok{0}\NormalTok{, }\DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{))}
+  \NormalTok{rosie =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{maxit =} \DecValTok{200}\NormalTok{, }\DataTypeTok{trace =} \DecValTok{0}\NormalTok{, }\DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{), }\DataTypeTok{hessian =} \OtherTok{TRUE}\NormalTok{)}
+
+  \NormalTok{rosie_point =}\StringTok{ }\NormalTok{rosie$beta[}\KeywordTok{upper.tri}\NormalTok{(rosie$beta)]}
+  \NormalTok{rosie_se =}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(}\KeywordTok{diag}\NormalTok{(}\KeywordTok{solve}\NormalTok{(rosie$opt$hessian)))[-(}\DecValTok{1}\NormalTok{:}\KeywordTok{sum}\NormalTok{(species_is_variable))]}
 
@@
 
-  \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\DecValTok{0}
-
-  \NormalTok{for(i in }\DecValTok{1}\NormalTok{:}\KeywordTok{nrow}\NormalTok{(bc$trace$R))\{}
+  \NormalTok{bc_pcors =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
+    \DecValTok{1}\NormalTok{:}\KeywordTok{nrow}\NormalTok{(bc$trace$R), }
+    \NormalTok{function(i)\{}
       \NormalTok{Sigma =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DataTypeTok{nrow =} \KeywordTok{ncol}\NormalTok{(x), }\DataTypeTok{ncol =} \KeywordTok{ncol}\NormalTok{(x))}
@@
       \KeywordTok{diag}\NormalTok{(Sigma) <-}\StringTok{ }\DecValTok{1}  \CommentTok{# Diagonal equals 1 in multivariate probit model}
-
-    \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ `}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{+}\StringTok{ }\KeywordTok{cor2pcor}\NormalTok{(Sigma) /}\StringTok{ }\KeywordTok{nrow}\NormalTok{(bc$trace$R)}
+      \NormalTok{pcor =}\StringTok{ }\KeywordTok{cor2pcor}\NormalTok{(Sigma)}
+      \NormalTok{pcor[}\KeywordTok{upper.tri}\NormalTok{(pcor)]}
     \NormalTok{\}}
-
-  \NormalTok{######## Pairs ########}
-  \CommentTok{# Find the line where the current data set is mentioned in}
-  \CommentTok{# pairs.txt}
-  \NormalTok{filename_line =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}
-    \KeywordTok{paste0}\NormalTok{(}
-      \KeywordTok{gsub}\NormalTok{(}\StringTok{"fakedata/(.*)}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.rds"}\NormalTok{, }\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{1"}\NormalTok{, filename),}
-      \StringTok{"-"}
-    \NormalTok{),}
-    \NormalTok{pairs_txt}
   \NormalTok{)}
 
-  \CommentTok{# Which chunk of the data file corresponds to this file?}
-  \NormalTok{chunk =}\StringTok{ }\KeywordTok{min}\NormalTok{(}\KeywordTok{which}\NormalTok{(beginnings >}\StringTok{ }\NormalTok{filename_line))}
-
-  \CommentTok{# Split the chunk on whitespace.}
-  \NormalTok{splitted =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(pairs_txt[beginnings[chunk]:ends[chunk]], }\StringTok{" +"}\NormalTok{)}
-
-  \CommentTok{# Pull out the species numbers and their Z-scores}
-  \NormalTok{pairs_results =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
-    \NormalTok{splitted,}
-    \NormalTok{function(x)\{}
-      \NormalTok{spp =}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(x[}\DecValTok{3}\NormalTok{:}\DecValTok{4}\NormalTok{]))}
-      \KeywordTok{data.frame}\NormalTok{(}
-        \DataTypeTok{sp1 =} \NormalTok{spp[}\DecValTok{1}\NormalTok{], }
-        \DataTypeTok{sp2 =} \NormalTok{spp[}\DecValTok{2}\NormalTok{], }
-        \DataTypeTok{z =} \KeywordTok{as.numeric}\NormalTok{(x[}\DecValTok{14}\NormalTok{])}
+  \KeywordTok{bind_rows}\NormalTok{(}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"Markov network"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \NormalTok{rosie_point, }
+      \DataTypeTok{lower =} \KeywordTok{qnorm}\NormalTok{(.}\DecValTok{025}\NormalTok{, rosie_point, rosie_se), }
+      \DataTypeTok{upper =} \KeywordTok{qnorm}\NormalTok{(.}\DecValTok{975}\NormalTok{, rosie_point, rosie_se)}
+    \NormalTok{),}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"GLM"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \NormalTok{coef_matrix[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)], }
+      \DataTypeTok{lower =} \OtherTok{NA}\NormalTok{, }
+      \DataTypeTok{upper =} \OtherTok{NA}
+    \NormalTok{),}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"correlation"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \NormalTok{corr[}\KeywordTok{upper.tri}\NormalTok{(corr)], }
+      \DataTypeTok{lower =} \OtherTok{NA}\NormalTok{, }
+      \DataTypeTok{upper =} \OtherTok{NA}
+    \NormalTok{),}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"partial correlation"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \NormalTok{p_corr[}\KeywordTok{upper.tri}\NormalTok{(p_corr)], }
+      \DataTypeTok{lower =} \OtherTok{NA}\NormalTok{, }
+      \DataTypeTok{upper =} \OtherTok{NA}
+    \NormalTok{),}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"partial BayesComm"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \KeywordTok{rowMeans}\NormalTok{(bc_pcors),}
+      \DataTypeTok{lower =} \KeywordTok{apply}\NormalTok{(bc_pcors, }\DecValTok{1}\NormalTok{, quantile, .}\DecValTok{025}\NormalTok{),}
+      \DataTypeTok{upper =} \KeywordTok{apply}\NormalTok{(bc_pcors, }\DecValTok{1}\NormalTok{, quantile, .}\DecValTok{975}\NormalTok{)}
+    \NormalTok{),}
+    \KeywordTok{data_frame}\NormalTok{(}
+      \DataTypeTok{rep_name =} \NormalTok{rep_name,}
+      \DataTypeTok{n_sites =} \NormalTok{n_sites,}
+      \DataTypeTok{sp1 =} \NormalTok{sp1,}
+      \DataTypeTok{sp2 =} \NormalTok{sp2,}
+      \DataTypeTok{method =} \StringTok{"BayesComm"}\NormalTok{,}
+      \DataTypeTok{truth =} \NormalTok{truth,}
+      \DataTypeTok{estimate =} \KeywordTok{colMeans}\NormalTok{(bc$trace$R),}
+      \DataTypeTok{lower =} \KeywordTok{apply}\NormalTok{(bc$trace$R, }\DecValTok{2}\NormalTok{, quantile, .}\DecValTok{025}\NormalTok{),}
+      \DataTypeTok{upper =} \KeywordTok{apply}\NormalTok{(bc$trace$R, }\DecValTok{2}\NormalTok{, quantile, .}\DecValTok{975}\NormalTok{)}
     \NormalTok{)}
-    \NormalTok{\}}
-  \NormalTok{) %>%}\StringTok{ }
-\StringTok{    }\NormalTok{bind_rows %>%}
-\StringTok{    }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{spp =} \KeywordTok{paste}\NormalTok{(sp1, sp2, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{))}
-
-  \CommentTok{# Re-order the pairs_results to match the other methods}
-  \NormalTok{m =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\OtherTok{NA}\NormalTok{, }\KeywordTok{ncol}\NormalTok{(x), }\KeywordTok{ncol}\NormalTok{(x))}
-  \NormalTok{new_order =}\StringTok{ }\KeywordTok{match}\NormalTok{(}
-    \KeywordTok{paste}\NormalTok{(}\KeywordTok{row}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)], }\KeywordTok{col}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)], }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{),}
-    \NormalTok{pairs_results$spp}
-  \NormalTok{)}
-  \NormalTok{ordered_pairs_results =}\StringTok{ }\NormalTok{pairs_results[new_order, ]}
-
-  \NormalTok{######## Output ########}
-  \KeywordTok{data.frame}\NormalTok{(}
-    \DataTypeTok{truth =} \NormalTok{truth, }
-    \DataTypeTok{n_sites =} \NormalTok{n_sites,}
-    \DataTypeTok{rep =} \NormalTok{rep,}
-    \DataTypeTok{sp1 =} \NormalTok{ordered_pairs_results$sp1,}
-    \DataTypeTok{sp2 =} \NormalTok{ordered_pairs_results$sp2,}
-    \StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{p_corr[}\KeywordTok{upper.tri}\NormalTok{(p_corr)],}
-    \DataTypeTok{correlation =} \NormalTok{corr[}\KeywordTok{upper.tri}\NormalTok{(corr)],}
-    \StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{rosie$beta[}\KeywordTok{upper.tri}\NormalTok{(rosie$beta)],}
-    \DataTypeTok{GLM =} \NormalTok{coef_matrix[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)],}
-    \StringTok{`}\DataTypeTok{BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(bc$trace$R),}
-    \StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ `}\DataTypeTok{partial BayesComm}\StringTok{`}\NormalTok{[}\KeywordTok{upper.tri}\NormalTok{(}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`}\NormalTok{)],}
-    \DataTypeTok{Pairs =} \NormalTok{ordered_pairs_results$z}
   \NormalTok{)}
@@
 \begin{Highlighting}[]
-\CommentTok{# Find all the .rds files in the fakedata folder}
-\NormalTok{files =}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"fakedata"}\NormalTok{, }\DataTypeTok{pattern =} \StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.rds$"}\NormalTok{, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{)}
+\CommentTok{# Find all the csv files in the fakedata/matrices folder,}
+\CommentTok{# then drop .csv}
+\NormalTok{identifiers =}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"fakedata/matrices"}\NormalTok{, }\DataTypeTok{pattern =} \StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{) %>%}
+\StringTok{  }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{, }\StringTok{""}\NormalTok{, .)}
 
-\CommentTok{# Run all the analyses on all the files}
-\NormalTok{z =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(files, fit_all) %>%}\StringTok{ }\NormalTok{bind_rows %>%}\StringTok{ }\NormalTok{as.data.frame}
-
-\CommentTok{# Fix a formatting issue with the column names}
-\KeywordTok{colnames}\NormalTok{(z) =}\StringTok{ }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{."}\NormalTok{, }\StringTok{" "}\NormalTok{, }\KeywordTok{colnames}\NormalTok{(z))}
-\end{Highlighting}
-\end{Shaded}
-
-Summarize the results of all 7 methods across the simulated landscapes:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Calculate residuals for each method}
-\NormalTok{resids =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
-  \KeywordTok{colnames}\NormalTok{(z)[-(}\DecValTok{1}\NormalTok{:}\DecValTok{5}\NormalTok{)],}
-  \NormalTok{function(i)\{}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z[,i] +}\StringTok{ }\DecValTok{0}\NormalTok{))\}}
-\NormalTok{)}
-
-\NormalTok{sizes =}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\KeywordTok{unique}\NormalTok{(z$n_sites))}
-
-\CommentTok{# Compute proportion of variance explained, compared with a null that}
-\CommentTok{# assumes all species interactions are 0.}
-\NormalTok{results =}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(}
-  \KeywordTok{t}\NormalTok{(}
-    \KeywordTok{sapply}\NormalTok{(}
-      \NormalTok{sizes,}
-      \NormalTok{function(n)\{}
-        \NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{z))[z$n_sites ==}\StringTok{ }\NormalTok{n]^}\DecValTok{2}\NormalTok{)}
-        \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids[z$n_sites ==}\StringTok{ }\NormalTok{n , ]^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss}
-      \NormalTok{\}}
-    \NormalTok{)}
-  \NormalTok{)}
-\NormalTok{)}
-\CommentTok{# Sort the results in decreasing order}
-\NormalTok{results =}\StringTok{ }\NormalTok{results[}\KeywordTok{order}\NormalTok{(}\KeywordTok{colMeans}\NormalTok{(results), }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
-\end{Highlighting}
-\end{Shaded}
-
-Plot the results:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{colors =}\StringTok{ }\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{8}\NormalTok{, }\StringTok{"Dark2"}\NormalTok{)[}\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{8}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{7}\NormalTok{)]}
-
-\CommentTok{# Set up the plotting canvas}
-\KeywordTok{pdf}\NormalTok{(}\StringTok{"manuscript-materials/figures/performance.pdf"}\NormalTok{, }\DataTypeTok{height =} \DecValTok{8}\NormalTok{, }\DataTypeTok{width =} \DecValTok{5}\NormalTok{)}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{))}
-\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar =} \KeywordTok{c}\NormalTok{(}\DecValTok{5}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{3}\NormalTok{, }\DecValTok{0}\NormalTok{) +}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{)}
-
-\CommentTok{# Plot the model performance}
-\KeywordTok{matplot}\NormalTok{(}
-  \NormalTok{sizes, }
-  \NormalTok{results, }
-  \DataTypeTok{type =} \StringTok{"o"}\NormalTok{, }
-  \DataTypeTok{ylab =} \StringTok{""}\NormalTok{, }
-  \DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\DecValTok{25} \NormalTok{*}\StringTok{ }\NormalTok{.}\DecValTok{9}\NormalTok{, }\DecValTok{1600} \NormalTok{*}\StringTok{ }\DecValTok{8}\NormalTok{),}
-  \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{5} \NormalTok{+}\StringTok{ }\FloatTok{1E-9}\NormalTok{),}
-  \DataTypeTok{pch =} \KeywordTok{c}\NormalTok{(}\DecValTok{15}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{16}\NormalTok{, }\DecValTok{2}\NormalTok{, }\DecValTok{3}\NormalTok{), }
-  \DataTypeTok{lty =} \DecValTok{1}\NormalTok{,}
-  \DataTypeTok{log =} \StringTok{"x"}\NormalTok{,}
-  \DataTypeTok{xlab =} \StringTok{""}\NormalTok{,}
-  \DataTypeTok{axes =} \OtherTok{FALSE}\NormalTok{,}
-  \DataTypeTok{xaxs =} \StringTok{"i"}\NormalTok{,}
-  \DataTypeTok{yaxs =} \StringTok{"i"}\NormalTok{,}
-  \DataTypeTok{col =} \NormalTok{colors,}
-  \DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{,}
-  \DataTypeTok{cex =} \FloatTok{0.9}\NormalTok{,}
-  \DataTypeTok{bty =} \StringTok{"l"}
-\NormalTok{)}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{25}\NormalTok{, }\DecValTok{200}\NormalTok{, }\DecValTok{1600}\NormalTok{))}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, .}\DecValTok{1}\NormalTok{), }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}
-
-\KeywordTok{mtext}\NormalTok{(}\KeywordTok{expression}\NormalTok{(R^}\DecValTok{2}\NormalTok{), }\DataTypeTok{side =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \DecValTok{3}\NormalTok{, }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\StringTok{"Number of sites (log scale)"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{1}\NormalTok{, }\DataTypeTok{line =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{200}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\StringTok{"A."}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{25}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{1.5}\NormalTok{)}
-
-\CommentTok{# Determine how high on the y-axis each method label should be so they}
-\CommentTok{# don't overlap.}
-\NormalTok{heights =}\StringTok{ }\NormalTok{results[}\KeywordTok{nrow}\NormalTok{(results), ]}
-
-\NormalTok{heights$Pairs =}\StringTok{ }\NormalTok{heights$Pairs -}\StringTok{ }\NormalTok{.}\DecValTok{01}
-\NormalTok{heights$correlation =}\StringTok{ }\NormalTok{heights$correlation +}\StringTok{ }\NormalTok{.}\DecValTok{01}
-
-\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial correlation}\StringTok{`} \NormalTok{+}\StringTok{ }\NormalTok{.}\DecValTok{01}
-\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{heights$}\StringTok{`}\DataTypeTok{partial BayesComm}\StringTok{`} \NormalTok{-}\StringTok{ }\NormalTok{.}\DecValTok{01}
-
-\NormalTok{heights$GLM =}\StringTok{ }\NormalTok{heights$GLM -}\StringTok{ }\NormalTok{.}\DecValTok{01}
-
-\KeywordTok{text}\NormalTok{(}\DecValTok{1625}\NormalTok{, heights, }\KeywordTok{colnames}\NormalTok{(results), }\DataTypeTok{pos =} \DecValTok{4}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{0.75}\NormalTok{, }\DataTypeTok{col =} \NormalTok{colors)}
-
-\NormalTok{######}
-
-\CommentTok{# silly code to make the next graph line up prettily with the previous one}
-\NormalTok{full_range =}\StringTok{ }\KeywordTok{log10}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{25} \NormalTok{*}\StringTok{ }\NormalTok{.}\DecValTok{9}\NormalTok{, }\DecValTok{1600} \NormalTok{*}\StringTok{ }\DecValTok{8}\NormalTok{))}
-\NormalTok{base_range =}\StringTok{ }\KeywordTok{log10}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DecValTok{1600}\NormalTok{))}
-
-\NormalTok{base_scaled =}\StringTok{ }\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{base_range /}\StringTok{ }\NormalTok{(base_range[}\DecValTok{2}\NormalTok{] -}\StringTok{ }\NormalTok{base_range[}\DecValTok{1}\NormalTok{])}
-\NormalTok{full_scaled =}\StringTok{ }\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{full_range /}\StringTok{ }\NormalTok{(base_range[}\DecValTok{2}\NormalTok{] -}\StringTok{ }\NormalTok{base_range[}\DecValTok{1}\NormalTok{])}
-
-\CommentTok{# New colors for plot B}
-\NormalTok{colors2 =}\StringTok{ }\KeywordTok{brewer.pal}\NormalTok{(}\DecValTok{10}\NormalTok{, }\StringTok{"Paired"}\NormalTok{)[}\KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{4}\NormalTok{, }\DecValTok{10}\NormalTok{)]}
-\KeywordTok{plot}\NormalTok{(}
-  \NormalTok{z$correlation, }
-  \NormalTok{z$Pairs, }
-  \DataTypeTok{col =} \NormalTok{colors2[}\KeywordTok{factor}\NormalTok{(z$n_sites)],}
-  \DataTypeTok{las =} \DecValTok{1}\NormalTok{,}
-  \DataTypeTok{xlab =} \StringTok{""}\NormalTok{,}
-  \DataTypeTok{ylab =} \KeywordTok{expression}\NormalTok{(}\KeywordTok{italic}\NormalTok{(Z)-score),}
-  \DataTypeTok{bty =} \StringTok{"l"}\NormalTok{,}
-  \DataTypeTok{xaxs =} \StringTok{"i"}\NormalTok{,}
-  \DataTypeTok{axes =} \OtherTok{FALSE}\NormalTok{,}
-  \DataTypeTok{xlim =} \NormalTok{full_scaled -}\StringTok{ }\NormalTok{base_scaled[}\DecValTok{2}\NormalTok{] +}\StringTok{ }\DecValTok{1}
-\NormalTok{)}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{seq}\NormalTok{(-}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{, .}\DecValTok{5}\NormalTok{))}
-\KeywordTok{axis}\NormalTok{(}\DecValTok{2}\NormalTok{, }\KeywordTok{seq}\NormalTok{(-}\DecValTok{50}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DecValTok{10}\NormalTok{), }\DataTypeTok{las =} \DecValTok{1}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\StringTok{"Correlation coefficient"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{1}\NormalTok{, }\DataTypeTok{line =} \FloatTok{2.25}\NormalTok{, }\DataTypeTok{at =} \DecValTok{0}\NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\StringTok{"B."}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.25}\NormalTok{, }\DataTypeTok{at =} \NormalTok{-}\DecValTok{1}\NormalTok{, }\DataTypeTok{cex =} \FloatTok{1.5}\NormalTok{)}
-
-\KeywordTok{legend}\NormalTok{(}
-  \DataTypeTok{x =} \FloatTok{1.05}\NormalTok{, }
-  \DataTypeTok{y =} \KeywordTok{max}\NormalTok{(z$Pairs),}
-  \DataTypeTok{pch =} \DecValTok{1}\NormalTok{,}
-  \DataTypeTok{lty =} \DecValTok{0}\NormalTok{,}
-  \DataTypeTok{lwd =} \DecValTok{2}\NormalTok{,}
-  \DataTypeTok{col =} \NormalTok{colors2[}\DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(}\KeywordTok{unique}\NormalTok{(z$n_sites))], }
-  \DataTypeTok{legend =} \KeywordTok{levels}\NormalTok{(}\KeywordTok{factor}\NormalTok{(z$n_sites)),}
-  \DataTypeTok{title =} \StringTok{"Number of sites"}\NormalTok{,}
-  \DataTypeTok{bty =} \StringTok{"n"}\NormalTok{,}
-  \DataTypeTok{cex =} \FloatTok{0.75}\NormalTok{,}
-  \DataTypeTok{pt.cex =} \DecValTok{1}
-\NormalTok{)}
-\KeywordTok{segments}\NormalTok{(-}\DecValTok{1000}\NormalTok{, }\DecValTok{0}\NormalTok{, }\FloatTok{1.04}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DataTypeTok{col =} \StringTok{"#00000025"}\NormalTok{, }\DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{)}
-\KeywordTok{segments}\NormalTok{(}\DecValTok{0}\NormalTok{, -}\DecValTok{1000}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{1000}\NormalTok{, }\DataTypeTok{col =} \StringTok{"#00000025"}\NormalTok{, }\DataTypeTok{lwd =} \FloatTok{1.25}\NormalTok{)}
-
-\KeywordTok{dev.off}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{manuscript-materials/figures/performance.pdf}
-
-Summarize the results:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# R-squareds computed across *all* landscape types}
-\NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{z))^}\DecValTok{2}\NormalTok{)}
-\KeywordTok{round}\NormalTok{(}\DecValTok{100} \NormalTok{*}\StringTok{ }\KeywordTok{sort}\NormalTok{(}\DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss), }\DecValTok{1}\NormalTok{)}
-
-\CommentTok{# Rank correlations among the methods that estimate marginal relationships}
-\KeywordTok{round}\NormalTok{(}\KeywordTok{cor}\NormalTok{(z[ , }\KeywordTok{c}\NormalTok{(}\DecValTok{7}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{)], }\DataTypeTok{method =} \StringTok{"spearman"}\NormalTok{)[,}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Plot the true interactions versus estimated interactions for two
-methods:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(ggplot2)}
-
-\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Markov network}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
-\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Markov network"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
-
-\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{-z$Pairs}
-\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{-Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
-\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
-\end{Highlighting}
-\end{Shaded}
-
-Bootstrap resample from each of the three landscape size classes,
-generating new sets of 150 landscapes. For each one, calculate the
-proportion of variance explained (compared to a null baseline that
-assumes all species pairs' interaction strengths are zero).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{rep =}\StringTok{ }\DecValTok{12}
-\NormalTok{n =}\StringTok{ }\DecValTok{200}
-\NormalTok{boots =}\StringTok{ }\KeywordTok{replicate}\NormalTok{(}\DecValTok{500}\NormalTok{,}
-          \NormalTok{\{}
-            \NormalTok{boot =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
-              \NormalTok{sizes, }
-              \NormalTok{function(n)\{}\KeywordTok{lapply}\NormalTok{(}
-                \KeywordTok{sample.int}\NormalTok{(}\KeywordTok{max}\NormalTok{(z$rep), }\DataTypeTok{replace =} \OtherTok{TRUE}\NormalTok{),}
-                \NormalTok{function(rep)\{}
-                  \NormalTok{z[z$rep ==}\StringTok{ }\NormalTok{rep &}\StringTok{ }\NormalTok{z$n_sites ==}\StringTok{ }\NormalTok{n, ]}
-                \NormalTok{\}}
-              \NormalTok{) %>%}\StringTok{ }\NormalTok{bind_rows\}}
-            \NormalTok{) %>%}\StringTok{ }\NormalTok{bind_rows}
-
-            \NormalTok{resids =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
-              \KeywordTok{colnames}\NormalTok{(boot)[-(}\DecValTok{1}\NormalTok{:}\DecValTok{5}\NormalTok{)],}
-              \NormalTok{function(i)\{}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(boot$truth ~}\StringTok{ }\NormalTok{boot[[i]] +}\StringTok{ }\DecValTok{0}\NormalTok{))\}}
-            \NormalTok{)}
-            \NormalTok{total_ss =}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{boot))^}\DecValTok{2}\NormalTok{)}
-            \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{colMeans}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\NormalTok{total_ss}
-          \NormalTok{\}}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-Summarize the bootstrap results:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{for(compared in }\KeywordTok{c}\NormalTok{(}\StringTok{"partial correlation"}\NormalTok{, }\StringTok{"correlation"}\NormalTok{, }\StringTok{"GLM"}\NormalTok{, }\StringTok{"BayesComm"}\NormalTok{, }\StringTok{"partial BayesComm"}\NormalTok{, }\StringTok{"Pairs"}\NormalTok{))\{}
-  \NormalTok{CI =}\StringTok{ }\KeywordTok{round}\NormalTok{(}
-    \KeywordTok{quantile}\NormalTok{(}
-      \NormalTok{boots[}\StringTok{"Markov network"}\NormalTok{, ] /}\StringTok{ }\NormalTok{boots[compared, ], }
-      \KeywordTok{c}\NormalTok{(.}\DecValTok{025}\NormalTok{, .}\DecValTok{975}\NormalTok{)), }
-    \DecValTok{2}
-  \NormalTok{)}
-
-  \KeywordTok{message}\NormalTok{(}\StringTok{"R-squared is "} \NormalTok{, CI[}\DecValTok{1}\NormalTok{], }\StringTok{"-"}\NormalTok{, CI[}\DecValTok{2}\NormalTok{], }\StringTok{" times higher than from "}\NormalTok{, compared, }\StringTok{" (95% bootstrap CI)"}\NormalTok{)}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{save}\NormalTok{(results, sizes, }\DataTypeTok{file =} \StringTok{"graph_data.Rdata"}\NormalTok{)}
+\KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }\DataTypeTok{mc.preschedule =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{mc.silent =} \OtherTok{TRUE}\NormalTok{) %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{bind_rows}\NormalTok{() %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{write.csv}\NormalTok{(}\DataTypeTok{file =} \StringTok{"estimates.csv"}\NormalTok{)}
 \end{Highlighting}

### 2015-11-25 Harris_2015/show-results.Rmd
as resubmitted

@@
-# Appendix 5: results
+---
+output: 
+  pdf_document:
+    keep_tex: true
+title: "Appendix 4: Results"
+subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
+author: David J. Harris
+fontsize: 11pt
+---
 
@@
 
-Exclude the "abundance" simulations because the interpretation of $\beta$ is different.
-
 ```{r, cache = TRUE}
+# Exclude the "abundance" simulations because the interpretation of $\beta$ 
+# is different
 rmses = markov_summary %>% filter(simulation_type != "abund") %>% mutate(rmse = sqrt((truth - estimate)^2))

### 2015-11-25 Harris_2015/show-results.tex
as resubmitted

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={David J. Harris},
            pdftitle={Appendix 4: Results},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\ImportTok}[1]{{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{{#1}}}}
\newcommand{\BuiltInTok}[1]{{#1}}
\newcommand{\ExtensionTok}[1]{{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{{#1}}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{{#1}}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{Appendix 4: Results}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
\subtitle{Inferring species interactions from co-occurrence data with Markov
networks}
  \author{David J. Harris}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(dplyr)}
\KeywordTok{library}\NormalTok{(mgcv)}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{library}\NormalTok{(tidyr)}
\KeywordTok{library}\NormalTok{(knitr)}
\KeywordTok{library}\NormalTok{(lme4)}
\end{Highlighting}
\end{Shaded}

\subsection{Import the results from Appendix
4:}\label{import-the-results-from-appendix-4}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{x =}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"estimates.csv"}\NormalTok{, }\DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}\NormalTok{)}
\NormalTok{x$simulation_type =}\StringTok{ }\KeywordTok{gsub}\NormalTok{(}\StringTok{"[0-9]"}\NormalTok{, }\StringTok{""}\NormalTok{, x$rep_name)}
\end{Highlighting}
\end{Shaded}

\subsection{\texorpdfstring{Import the results from the \emph{Pairs}
software:}{Import the results from the Pairs software:}}\label{import-the-results-from-the-pairs-software}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pairs_txt =}\StringTok{ }\KeywordTok{readLines}\NormalTok{(}\StringTok{"fakedata/matrices/Pairs.txt"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(stringr)}

\CommentTok{# Find areas of the data file that correspond}
\CommentTok{# to species pairs' results}
\NormalTok{beginnings =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}\StringTok{"Sp1"}\NormalTok{, pairs_txt) +}\StringTok{ }\DecValTok{1}
\NormalTok{ends =}\StringTok{ }\KeywordTok{c}\NormalTok{(}
  \KeywordTok{grep}\NormalTok{(}\StringTok{"^[^ ]"}\NormalTok{, pairs_txt)[-}\DecValTok{1}\NormalTok{],}
  \KeywordTok{length}\NormalTok{(pairs_txt) +}\StringTok{ }\DecValTok{1}
\NormalTok{) -}\StringTok{ }\DecValTok{1}

\NormalTok{partial_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(}\KeywordTok{grep}\NormalTok{(}\StringTok{"^>"}\NormalTok{, pairs_txt, }\DataTypeTok{value =} \OtherTok{TRUE}\NormalTok{), }\StringTok{" +"}\NormalTok{), function(x) x[[}\DecValTok{3}\NormalTok{]])}
\NormalTok{filename_lines =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}\StringTok{"^>"}\NormalTok{, pairs_txt)}

\CommentTok{# Sort a vector of alphanumeric strings by the numeric component}
\CommentTok{# as if they were integers.  For example, V20 is larger than V12,}
\CommentTok{# even though V12 comes first alphabetically}
\NormalTok{alnum_sort =}\StringTok{ }\NormalTok{function(x)\{}
  \NormalTok{raw =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{gsub}\NormalTok{(}\StringTok{"[[:alpha:]]"}\NormalTok{, }\StringTok{""}\NormalTok{, x))}
  \NormalTok{x[}\KeywordTok{order}\NormalTok{(raw)]}
\NormalTok{\}}

\NormalTok{pairs_results =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
  \DecValTok{1}\NormalTok{:}\KeywordTok{length}\NormalTok{(filename_lines),}
  \NormalTok{function(i)\{}

    \NormalTok{n_sites =}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(partial_names[[i]], }\StringTok{"-"}\NormalTok{)[[}\DecValTok{1}\NormalTok{]][[}\DecValTok{1}\NormalTok{]])}
    \NormalTok{rep_name =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(partial_names[[i]], }\StringTok{"-"}\NormalTok{)[[}\DecValTok{1}\NormalTok{]][[}\DecValTok{2}\NormalTok{]]}

    \CommentTok{# Find the line where the current data set is mentioned in}
    \CommentTok{# pairs.txt}
    \NormalTok{filename_line =}\StringTok{ }\NormalTok{filename_lines[i]}

    \CommentTok{# Which chunk of the data file corresponds to this file?}
    \NormalTok{chunk =}\StringTok{ }\KeywordTok{min}\NormalTok{(}\KeywordTok{which}\NormalTok{(beginnings >}\StringTok{ }\NormalTok{filename_line))}

    \CommentTok{# Split the chunk on whitespace.}
    \NormalTok{splitted =}\StringTok{ }\KeywordTok{strsplit}\NormalTok{(pairs_txt[beginnings[chunk]:ends[chunk]], }\StringTok{" +"}\NormalTok{)}

    \CommentTok{# Pull out the corresponding chunk of the "x" data frame, based on n_sites and rep_name}
    \NormalTok{x_subset =}\StringTok{ }\NormalTok{x[x$n_sites ==}\StringTok{ }\NormalTok{n_sites &}\StringTok{ }\NormalTok{x$rep_name ==}\StringTok{ }\NormalTok{rep_name &}\StringTok{ }\NormalTok{x$method ==}\StringTok{ "correlation"}\NormalTok{, ]}

    \CommentTok{# Pull out the species numbers and their Z-scores, then join to x_subset}
    \NormalTok{pairs_results =}\StringTok{ }\KeywordTok{lapply}\NormalTok{(}
      \NormalTok{splitted,}
      \NormalTok{function(x)\{}
        \CommentTok{# in the x data frame, species 1 is always a lower number than species 2}
        \NormalTok{spp =}\StringTok{ }\KeywordTok{alnum_sort}\NormalTok{(x[}\DecValTok{3}\NormalTok{:}\DecValTok{4}\NormalTok{])}
        \KeywordTok{data.frame}\NormalTok{(}
          \DataTypeTok{sp1 =} \NormalTok{spp[}\DecValTok{1}\NormalTok{], }
          \DataTypeTok{sp2 =} \NormalTok{spp[}\DecValTok{2}\NormalTok{], }
          \DataTypeTok{z =} \NormalTok{x[}\DecValTok{14}\NormalTok{],}
          \DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}
        \NormalTok{)}
      \NormalTok{\}}
    \NormalTok{) %>%}\StringTok{ }
\StringTok{      }\NormalTok{bind_rows %>%}
\StringTok{      }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{spp =} \KeywordTok{paste}\NormalTok{(sp1, sp2, }\DataTypeTok{sep =} \StringTok{"-"}\NormalTok{))}

    \NormalTok{n_spp =}\StringTok{ }\DecValTok{20}

    \NormalTok{pairs_results$z =}\StringTok{ }\KeywordTok{as.numeric}\NormalTok{(pairs_results$z)}

    \CommentTok{# Re-order the pairs_results to match the other methods}
    \NormalTok{m =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\OtherTok{NA}\NormalTok{, n_spp, n_spp)}
    \NormalTok{new_order =}\StringTok{ }\KeywordTok{match}\NormalTok{(}
      \KeywordTok{paste0}\NormalTok{(}\StringTok{"V"}\NormalTok{, }\KeywordTok{row}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)], }\StringTok{"-V"}\NormalTok{, }\KeywordTok{col}\NormalTok{(m)[}\KeywordTok{upper.tri}\NormalTok{(m)]),}
      \NormalTok{pairs_results$spp}
    \NormalTok{)}
    \NormalTok{ordered_pairs_results =}\StringTok{ }\NormalTok{pairs_results[}\KeywordTok{na.omit}\NormalTok{(new_order), ]}

    \NormalTok{ordered_pairs_results =}\StringTok{ }\NormalTok{ordered_pairs_results %>%}
\StringTok{      }\KeywordTok{filter}\NormalTok{(sp1 %in%}\StringTok{ }\KeywordTok{c}\NormalTok{(x_subset$sp1) &}\StringTok{ }\NormalTok{sp2 %in%}\StringTok{ }\NormalTok{x_subset$sp2)}

    \NormalTok{x_subset$estimate =}\StringTok{ }\NormalTok{ordered_pairs_results$z}
    \NormalTok{x_subset$method =}\StringTok{ "null"}

    \NormalTok{x_subset}
  \NormalTok{\}}
\NormalTok{) %>%}\StringTok{ }\KeywordTok{bind_rows}\NormalTok{()}

\CommentTok{# Manually adjust the Z values less than -1000 so that these outliers}
\CommentTok{# won't completely dominate the analyses below}
\NormalTok{pairs_results$estimate[pairs_results$estimate <}\StringTok{ }\NormalTok{-}\DecValTok{1000}\NormalTok{] =}\StringTok{ }\NormalTok{-}\DecValTok{50}

\NormalTok{x =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(x, pairs_results)}
\end{Highlighting}
\end{Shaded}

\subsection{Calculate model
performance:}\label{calculate-model-performance}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{resids =}\StringTok{ }\NormalTok{function(data)\{}
  \KeywordTok{resid}\NormalTok{(}\KeywordTok{lm}\NormalTok{(truth ~}\StringTok{ }\NormalTok{estimate +}\StringTok{ }\DecValTok{0}\NormalTok{, }\DataTypeTok{data =} \NormalTok{data))}
\NormalTok{\}}

\NormalTok{result_summary =}\StringTok{ }\NormalTok{x %>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method, simulation_type) %>%}
\StringTok{  }\KeywordTok{do}\NormalTok{(}\KeywordTok{data.frame}\NormalTok{(., }\DataTypeTok{resids =} \KeywordTok{resids}\NormalTok{(.))) %>%}
\StringTok{  }\NormalTok{ungroup %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method, simulation_type, n_sites) %>%}
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{r2 =} \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{sum}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\KeywordTok{sum}\NormalTok{(truth^}\DecValTok{2}\NormalTok{))}

\NormalTok{result_summary$method =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(result_summary$method, -result_summary$r2)}

\NormalTok{result_summary$simulation_type =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(result_summary$simulation_type, -result_summary$r2)}

\NormalTok{result_summary =}\StringTok{ }\NormalTok{result_summary %>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{mean_r2 =} \KeywordTok{round}\NormalTok{(}\DecValTok{100} \NormalTok{*}\StringTok{ }\KeywordTok{mean}\NormalTok{(r2))) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{method_r2 =} \KeywordTok{paste0}\NormalTok{(method, }\StringTok{" (0."}\NormalTok{, mean_r2, }\StringTok{")"}\NormalTok{)) %>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(method, method_r2) %>%}
\StringTok{  }\KeywordTok{inner_join}\NormalTok{(result_summary, }\StringTok{"method"}\NormalTok{)}

\NormalTok{result_summary$simulation_type_long =}\StringTok{ }\NormalTok{plyr::}\KeywordTok{revalue}\NormalTok{(}
  \NormalTok{result_summary$simulation_type,}
  \KeywordTok{c}\NormalTok{(}\DataTypeTok{no_env =} \StringTok{"constant environment"}\NormalTok{,}
    \DataTypeTok{env =} \StringTok{"heterogeneous environment"}\NormalTok{,}
    \DataTypeTok{abund =} \StringTok{"abundance"}\NormalTok{)}
\NormalTok{)}

\NormalTok{result_summary$method_r2 =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(result_summary$method_r2, -result_summary$r2)}
\end{Highlighting}
\end{Shaded}

\paragraph{Model performance with no environmental
variation:}\label{model-performance-with-no-environmental-variation}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{result_summary %>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "no_env"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
method & mean(r2)\tabularnewline
\midrule
\endhead
Markov network & 0.525\tabularnewline
GLM & 0.472\tabularnewline
partial correlation & 0.403\tabularnewline
partial BayesComm & 0.394\tabularnewline
correlation & 0.291\tabularnewline
null & 0.227\tabularnewline
BayesComm & 0.206\tabularnewline
\bottomrule
\end{longtable}

\paragraph{Model performance with environmental
heterogeneity:}\label{model-performance-with-environmental-heterogeneity}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{result_summary %>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "env"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
method & mean(r2)\tabularnewline
\midrule
\endhead
Markov network & 0.451\tabularnewline
GLM & 0.405\tabularnewline
partial correlation & 0.322\tabularnewline
partial BayesComm & 0.302\tabularnewline
correlation & 0.183\tabularnewline
null & 0.125\tabularnewline
BayesComm & 0.110\tabularnewline
\bottomrule
\end{longtable}

\paragraph{Model performance with per-capita species
interactions:}\label{model-performance-with-per-capita-species-interactions}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{result_summary %>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "abund"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
method & mean(r2)\tabularnewline
\midrule
\endhead
Markov network & 0.384\tabularnewline
GLM & 0.283\tabularnewline
partial correlation & 0.200\tabularnewline
partial BayesComm & 0.166\tabularnewline
correlation & 0.117\tabularnewline
null & 0.075\tabularnewline
BayesComm & 0.060\tabularnewline
\bottomrule
\end{longtable}

\subsection{Plot the R-squared results (Figure
3)}\label{plot-the-r-squared-results-figure-3}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{legend_name =}\StringTok{ }\KeywordTok{expression}\NormalTok{(Method~(mean~R^}\DecValTok{2}\NormalTok{))}

\KeywordTok{pdf}\NormalTok{(}\StringTok{"manuscript-materials/figures/performance.pdf"}\NormalTok{, }\DataTypeTok{width =} \DecValTok{8}\NormalTok{, }\DataTypeTok{height =} \FloatTok{2.5}\NormalTok{)}
\KeywordTok{ggplot}\NormalTok{(result_summary, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{n_sites, }\DataTypeTok{y =} \NormalTok{r2, }\DataTypeTok{col =} \NormalTok{method_r2, }\DataTypeTok{shape =} \NormalTok{method_r2)) +}\StringTok{ }
\StringTok{  }\KeywordTok{facet_grid}\NormalTok{(~simulation_type_long) +}\StringTok{ }
\StringTok{  }\KeywordTok{geom_line}\NormalTok{(}\DataTypeTok{size =} \NormalTok{.}\DecValTok{5}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size =} \FloatTok{2.5}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"white"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_shape_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\DecValTok{16}\NormalTok{, }\DecValTok{22}\NormalTok{, }\DecValTok{17}\NormalTok{, }\DecValTok{23}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{24}\NormalTok{, }\DecValTok{15}\NormalTok{), }\DataTypeTok{name =} \NormalTok{legend_name) +}
\StringTok{  }\KeywordTok{geom_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{2}\NormalTok{) +}
\StringTok{  }\KeywordTok{geom_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }
\StringTok{  }\CommentTok{#scale_shape_manual(values = c(21, 25, 15, 18)) + }
\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-.}\DecValTok{01}\NormalTok{, }\FloatTok{0.76}\NormalTok{)) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\KeywordTok{expression}\NormalTok{(R^}\DecValTok{2}\NormalTok{)) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Number of sites (log scale)"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_x_log10}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{unique}\NormalTok{(x$n_sites), }\DataTypeTok{limits =} \KeywordTok{range}\NormalTok{(x$n_sites)) +}
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{(}\DataTypeTok{base_size =} \DecValTok{11}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{panel.margin =} \NormalTok{grid::}\KeywordTok{unit}\NormalTok{(}\FloatTok{1.25}\NormalTok{, }\StringTok{"lines"}\NormalTok{)) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{panel.border =} \KeywordTok{element_blank}\NormalTok{(), }\DataTypeTok{axis.line =} \KeywordTok{element_blank}\NormalTok{()) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}
    \DataTypeTok{panel.grid.minor =} \KeywordTok{element_blank}\NormalTok{(), }
    \DataTypeTok{panel.grid.major.y =} \KeywordTok{element_line}\NormalTok{(}\DataTypeTok{color =} \StringTok{"lightgray"}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{4}\NormalTok{), }
    \DataTypeTok{panel.grid.major.x =} \KeywordTok{element_blank}\NormalTok{()}
  \NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{strip.background =} \KeywordTok{element_blank}\NormalTok{(), }\DataTypeTok{legend.key =} \KeywordTok{element_blank}\NormalTok{()) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{plot.margin =} \NormalTok{grid::}\KeywordTok{unit}\NormalTok{(}\KeywordTok{c}\NormalTok{(.}\DecValTok{01}\NormalTok{, .}\DecValTok{01}\NormalTok{, .}\DecValTok{75}\NormalTok{, .}\DecValTok{1}\NormalTok{), }\StringTok{"lines"}\NormalTok{)) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme}\NormalTok{(}
    \DataTypeTok{axis.title.x =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{vjust =} \NormalTok{-}\FloatTok{0.2}\NormalTok{, }\DataTypeTok{size =} \DecValTok{12}\NormalTok{), }
    \DataTypeTok{axis.title.y =} \KeywordTok{element_text}\NormalTok{(}\DataTypeTok{angle =} \DecValTok{0}\NormalTok{, }\DataTypeTok{hjust =} \NormalTok{-.}\DecValTok{1}\NormalTok{, }\DataTypeTok{size =} \DecValTok{12}\NormalTok{)}
  \NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_color_brewer}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"Dark2"}\NormalTok{, }\DataTypeTok{name =} \NormalTok{legend_name)}
\KeywordTok{dev.off}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\subsection{Estimate R-squared
uncertainty:}\label{estimate-r-squared-uncertainty}

Fit a linear mixed model describing R-squared as a function of method,
landscape size, and simulation type. Note the small standard errors
associated with the effect of estimation method.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{landscape_estimates =}\StringTok{ }\NormalTok{x %>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method, simulation_type) %>%}
\StringTok{  }\KeywordTok{do}\NormalTok{(}\KeywordTok{data.frame}\NormalTok{(., }\DataTypeTok{resids =} \KeywordTok{resids}\NormalTok{(.))) %>%}
\StringTok{  }\NormalTok{ungroup %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(method, simulation_type, rep_name, n_sites) %>%}
\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\DataTypeTok{r2 =} \DecValTok{1} \NormalTok{-}\StringTok{ }\KeywordTok{sum}\NormalTok{(resids^}\DecValTok{2}\NormalTok{) /}\StringTok{ }\KeywordTok{sum}\NormalTok{(truth^}\DecValTok{2}\NormalTok{))}

\KeywordTok{summary}\NormalTok{(}\KeywordTok{lmer}\NormalTok{(r2 ~}\StringTok{ }\NormalTok{method +}\StringTok{ }\NormalTok{n_sites +}\StringTok{ }\NormalTok{simulation_type +}\StringTok{ }\NormalTok{(}\DecValTok{1}\NormalTok{|rep_name), }\DataTypeTok{data =} \NormalTok{landscape_estimates))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linear mixed model fit by REML ['lmerMod']
## Formula: r2 ~ method + n_sites + simulation_type + (1 | rep_name)
##    Data: landscape_estimates
## 
## REML criterion at convergence: -4945.2
## 
## Scaled residuals: 
##     Min      1Q  Median      3Q     Max 
## -7.3997 -0.6350  0.0758  0.6883  2.7104 
## 
## Random effects:
##  Groups   Name        Variance  Std.Dev.
##  rep_name (Intercept) 0.0008304 0.02882 
##  Residual             0.0113238 0.10641 
## Number of obs: 3150, groups:  rep_name, 150
## 
## Fixed effects:
##                             Estimate Std. Error t value
## (Intercept)               -2.481e-02  7.186e-03   -3.45
## methodcorrelation          6.965e-02  7.094e-03    9.82
## methodGLM                  2.544e-01  7.094e-03   35.85
## methodMarkov network       3.217e-01  7.094e-03   45.35
## methodnull                 1.504e-02  7.094e-03    2.12
## methodpartial BayesComm    1.616e-01  7.094e-03   22.78
## methodpartial correlation  1.796e-01  7.094e-03   25.31
## n_sites                    1.050e-04  2.690e-06   39.04
## simulation_typeenv         8.853e-02  7.402e-03   11.96
## simulation_typeno_env      1.795e-01  7.402e-03   24.25
## 
## Correlation of Fixed Effects:
##             (Intr) mthdcr mthGLM mthdMn mthdnl mtBysC mthdpc n_sits smltn_
## methdcrrltn -0.494                                                        
## methodGLM   -0.494  0.500                                                 
## mthdMrkvntw -0.494  0.500  0.500                                          
## methodnull  -0.494  0.500  0.500  0.500                                   
## mthdprtBysC -0.494  0.500  0.500  0.500  0.500                            
## mthdprtlcrr -0.494  0.500  0.500  0.500  0.500  0.500                     
## n_sites     -0.228  0.000  0.000  0.000  0.000  0.000  0.000              
## smltn_typnv -0.515  0.000  0.000  0.000  0.000  0.000  0.000  0.000       
## smltn_typn_ -0.515  0.000  0.000  0.000  0.000  0.000  0.000  0.000  0.500
\end{verbatim}

\subsection{Summarize the inferential results of the Markov network and
the null
model:}\label{summarize-the-inferential-results-of-the-markov-network-and-the-null-model}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pairs_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "null"} \NormalTok{&}\StringTok{ }\NormalTok{!}\KeywordTok{grepl}\NormalTok{(}\StringTok{"pop"}\NormalTok{, x$rep_name), ]}
\NormalTok{markov_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "Markov network"} \NormalTok{&}\StringTok{ }\NormalTok{!}\KeywordTok{grepl}\NormalTok{(}\StringTok{"pop"}\NormalTok{, x$rep_name), ]}

\CommentTok{# Pairs's Z score is based on C-scores, which are positive when species are }
\CommentTok{# disaggregated.  So significantly positive interactions == negative estimates}
\NormalTok{pairs_summary$sig_pos =}\StringTok{ }\NormalTok{pairs_summary$estimate <}\StringTok{ }\KeywordTok{qnorm}\NormalTok{(.}\DecValTok{025}\NormalTok{)}
\NormalTok{pairs_summary$sig_neg =}\StringTok{ }\NormalTok{pairs_summary$estimate >}\StringTok{ }\KeywordTok{qnorm}\NormalTok{(.}\DecValTok{975}\NormalTok{)}

\CommentTok{# Markov network is significant when lower bound is above zero or lower bound is}
\CommentTok{# below zero}
\NormalTok{markov_summary$sig_pos =}\StringTok{ }\NormalTok{markov_summary$lower >}\StringTok{ }\DecValTok{0}
\NormalTok{markov_summary$sig_neg =}\StringTok{ }\NormalTok{markov_summary$upper <}\StringTok{ }\DecValTok{0}
\end{Highlighting}
\end{Shaded}

\subsection{Create Figure 4:}\label{create-figure-4}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Function for smoothing the error rates with a generalized additive model}
\CommentTok{# and plotting the results}
\NormalTok{my_geom_smooth =}\StringTok{ }\NormalTok{function(data, color, ...)\{}
  \KeywordTok{geom_smooth}\NormalTok{(}
    \DataTypeTok{data =} \NormalTok{data,}
    \KeywordTok{aes}\NormalTok{(}
      \DataTypeTok{x =} \NormalTok{truth, }
      \DataTypeTok{y =} \KeywordTok{as.integer}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)),}
      \DataTypeTok{color =} \NormalTok{color}
    \NormalTok{),}
    \DataTypeTok{method =} \NormalTok{gam, }
    \DataTypeTok{family =} \NormalTok{binomial, }
    \DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x),}
    \DataTypeTok{se =} \OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{n =} \DecValTok{1024}\NormalTok{,}
    \NormalTok{...}
  \NormalTok{)}
\NormalTok{\}}

\NormalTok{truth_seq =}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(markov_summary$truth), }\KeywordTok{max}\NormalTok{(markov_summary$truth), }\DataTypeTok{length =} \DecValTok{1000}\NormalTok{)}

\NormalTok{plotfun =}\StringTok{ }\NormalTok{function(..., add)\{}
  \NormalTok{if(add)\{}
    \KeywordTok{lines}\NormalTok{(...)}
  \NormalTok{\}else\{}
    \KeywordTok{plot}\NormalTok{(...)}
  \NormalTok{\}}
\NormalTok{\}}

\NormalTok{error_smoother =}\StringTok{ }\NormalTok{function(data)\{}
  \KeywordTok{predict}\NormalTok{(}
    \KeywordTok{gam}\NormalTok{(}
      \KeywordTok{I}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)) ~}\StringTok{ }\KeywordTok{s}\NormalTok{(truth),}
      \DataTypeTok{data =} \NormalTok{data,}
      \DataTypeTok{family =} \NormalTok{binomial}
    \NormalTok{),}
    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{truth =} \NormalTok{truth_seq),}
    \DataTypeTok{type =} \StringTok{"response"}
  \NormalTok{)}
\NormalTok{\}}

\NormalTok{y_pairs =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(pairs_summary)}
\NormalTok{y_markov =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(markov_summary)}

\KeywordTok{pdf}\NormalTok{(}\StringTok{"manuscript-materials/figures/error_rates.pdf"}\NormalTok{, }\DataTypeTok{height =} \FloatTok{8.5}\NormalTok{, }\DataTypeTok{width =} \FloatTok{8.5}\NormalTok{/}\DecValTok{3}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{))}

\CommentTok{# Compare estimates}
\NormalTok{spread_estimates =}\StringTok{ }\NormalTok{x %>%}
\StringTok{  }\NormalTok{dplyr::}\KeywordTok{select}\NormalTok{(-lower, -upper, -X) %>%}
\StringTok{  }\KeywordTok{spread}\NormalTok{(method, estimate) %>%}
\StringTok{  }\KeywordTok{na.omit}\NormalTok{()}

\CommentTok{# R-squared for null versus correlation}
\KeywordTok{round}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(null ~}\StringTok{ }\KeywordTok{I}\NormalTok{(correlation*}\KeywordTok{sqrt}\NormalTok{(n_sites)), }\DataTypeTok{data =} \NormalTok{spread_estimates))$r.squared, }\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.95
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# R-squared for glm markov network versus glm}
\KeywordTok{round}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{~}\StringTok{ }\NormalTok{GLM, }\DataTypeTok{data =} \NormalTok{spread_estimates))$r.squared, }\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.94
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{with}\NormalTok{(}
  \NormalTok{spread_estimates, }
  \KeywordTok{plot}\NormalTok{(}
    \StringTok{`}\DataTypeTok{Markov network}\StringTok{`}\NormalTok{, }
    \NormalTok{GLM, }
    \DataTypeTok{pch =} \StringTok{"."}\NormalTok{, }
    \DataTypeTok{col =} \StringTok{"#00000020"}\NormalTok{,}
    \DataTypeTok{ylab =} \StringTok{"Markov network estimate"}\NormalTok{,}
    \DataTypeTok{xlab =} \StringTok{"GLM estimate"}\NormalTok{,}
    \DataTypeTok{bty =} \StringTok{"l"}
  \NormalTok{)}
\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"A. Markov network estimates}\CharTok{\textbackslash{}n}\StringTok{vs. GLM estimates"}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
\KeywordTok{abline}\NormalTok{(}\KeywordTok{lm}\NormalTok{(}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{~}\StringTok{ }\NormalTok{GLM, }\DataTypeTok{data =} \NormalTok{spread_estimates))}
\KeywordTok{text}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\KeywordTok{expression}\NormalTok{(R^}\DecValTok{2}\NormalTok{==}\FloatTok{0.94}\NormalTok{))}

\KeywordTok{with}\NormalTok{(}
  \NormalTok{spread_estimates, }
  \KeywordTok{plot}\NormalTok{(}
    \NormalTok{correlation *}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(n_sites), }
    \NormalTok{null, }
    \DataTypeTok{pch =} \StringTok{"."}\NormalTok{, }
    \DataTypeTok{col =} \StringTok{"#00000020"}\NormalTok{,}
    \DataTypeTok{ylab =} \StringTok{"Z-score"}\NormalTok{,}
    \DataTypeTok{xlab =} \KeywordTok{expression}\NormalTok{(}\StringTok{"correlation"} \NormalTok{%*%}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(number~}\ErrorTok{~}\NormalTok{of~}\ErrorTok{~}\NormalTok{sites)),}
    \DataTypeTok{bty =} \StringTok{"l"}
  \NormalTok{)}
\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"B. Null model estimates vs.}\CharTok{\textbackslash{}n}\StringTok{scaled correlation coefficients"}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
\KeywordTok{abline}\NormalTok{(}\KeywordTok{lm}\NormalTok{(null ~}\StringTok{ }\KeywordTok{I}\NormalTok{(correlation*}\KeywordTok{sqrt}\NormalTok{(n_sites)), }\DataTypeTok{data =} \NormalTok{spread_estimates))}
\KeywordTok{text}\NormalTok{(}\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\KeywordTok{expression}\NormalTok{(R^}\DecValTok{2}\NormalTok{==}\FloatTok{0.95}\NormalTok{))}

\KeywordTok{plotfun}\NormalTok{(}
  \NormalTok{truth_seq,}
  \NormalTok{y_pairs,}
  \DataTypeTok{type =} \StringTok{"l"}\NormalTok{,}
  \DataTypeTok{xlab =} \StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ interaction strength"}\NormalTok{,}
  \DataTypeTok{ylab =} \StringTok{"P(confidently predict wrong sign)"}\NormalTok{,}
  \DataTypeTok{bty =} \StringTok{"l"}\NormalTok{,}
  \DataTypeTok{yaxs =} \StringTok{"i"}\NormalTok{,}
  \DataTypeTok{col =} \DecValTok{2}\NormalTok{,}
  \DataTypeTok{add =} \OtherTok{FALSE}\NormalTok{,}
  \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{4}\NormalTok{),}
  \DataTypeTok{lwd =} \DecValTok{2}
\NormalTok{)}
\KeywordTok{mtext}\NormalTok{(}\StringTok{"C. Error rate vs.}\CharTok{\textbackslash{}n}\StringTok{interaction strength"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
\KeywordTok{plotfun}\NormalTok{(truth_seq, y_markov, }\DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Null model"}\NormalTok{, }\StringTok{"Markov network"}\NormalTok{), }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
\KeywordTok{dev.off}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

\subsection{Summarize inferential
statistics:}\label{summarize-inferential-statistics}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(Pairs confidently wrong)}
\KeywordTok{with}\NormalTok{(pairs_summary,  }\KeywordTok{mean}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.1533758
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(Markov network confidently wrong)}
\KeywordTok{with}\NormalTok{(markov_summary, }\KeywordTok{mean}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.02861541
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(Pairs rejects null)}
\KeywordTok{with}\NormalTok{(pairs_summary,  }\KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.4520535
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(Markov network rejects null)}
\KeywordTok{with}\NormalTok{(markov_summary,  }\KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.2116561
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(reject null | small interaction) for Makrov network}
\CommentTok{# (approximate Type I error rates)}
\NormalTok{markov_summary %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{Type_1_error_rate =} \KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos)) %>%}
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
simulation\_type & Type\_1\_error\_rate\tabularnewline
\midrule
\endhead
abund & 0.220\tabularnewline
env & 0.136\tabularnewline
no\_env & 0.019\tabularnewline
\bottomrule
\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# P(reject null | small interaction) for Pairs}
\CommentTok{# (approximate Type I error rates)}
\NormalTok{pairs_summary %>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{Type_1_error_rate =} \KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos)) %>%}
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
simulation\_type & Type\_1\_error\_rate\tabularnewline
\midrule
\endhead
abund & 0.575\tabularnewline
env & 0.501\tabularnewline
no\_env & 0.297\tabularnewline
\bottomrule
\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Confidence interval coverage across all values}
\CommentTok{# What's the probability that any "true" value falls inside the 95% CI?}
\NormalTok{markov_summary %>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\KeywordTok{mean}\NormalTok{(truth >}\StringTok{ }\NormalTok{lower &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\NormalTok{upper)) %>%}
\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
simulation\_type & mean(truth \textgreater{} lower \& truth \textless{}
upper)\tabularnewline
\midrule
\endhead
abund & 0.868\tabularnewline
env & 0.857\tabularnewline
no\_env & 0.975\tabularnewline
\bottomrule
\end{longtable}

\subsection{\texorpdfstring{Plot root mean square error versus ``true''
beta}{Plot root mean square error versus true beta}}\label{plot-root-mean-square-error-versus-true-beta}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Exclude the "abundance" simulations because the interpretation of $\textbackslash{}beta$ }
\CommentTok{# is different}
\NormalTok{rmses =}\StringTok{ }\NormalTok{markov_summary %>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(simulation_type !=}\StringTok{ "abund"}\NormalTok{) %>%}\StringTok{ }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{rmse =} \KeywordTok{sqrt}\NormalTok{((truth -}\StringTok{ }\NormalTok{estimate)^}\DecValTok{2}\NormalTok{))}

\KeywordTok{ggplot}\NormalTok{(rmses, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{rmse)) +}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \NormalTok{.}\DecValTok{2}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }\KeywordTok{geom_smooth}\NormalTok{() +}\StringTok{ }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{max}\NormalTok{(rmses$rmse)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## geom_smooth: method="auto" and size of largest group is >=1000, so using gam with formula: y ~ s(x, bs = "cs"). Use 'method = x' to change the smoothing method.
\end{verbatim}

\includegraphics{show-results_files/figure-latex/unnamed-chunk-13-1.pdf}

\subsection{\texorpdfstring{Plot confidence interval coverage versus
true \(\beta\)
value}{Plot confidence interval coverage versus true \textbackslash{}beta value}}\label{plot-confidence-interval-coverage-versus-true-beta-value}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Coverage versus true beta value.  The top and bottom 0.5%}
\CommentTok{# of the distribution have been omitted to prevent bad behavior}
\CommentTok{# by the smoother in the tails.}
\NormalTok{markov_summary %>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{percent_rank}\NormalTok{(truth) >}\StringTok{ }\NormalTok{.}\DecValTok{005} \NormalTok{&}\StringTok{ }\KeywordTok{percent_rank}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{995}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{covered =} \NormalTok{truth >}\StringTok{ }\NormalTok{lower &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\NormalTok{upper) %>%}
\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \KeywordTok{as.integer}\NormalTok{(covered))) +}\StringTok{ }
\StringTok{  }\KeywordTok{facet_grid}\NormalTok{(~simulation_type) +}\StringTok{ }
\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \NormalTok{gam, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x), }\DataTypeTok{family =} \NormalTok{binomial) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)) +}\StringTok{ }
\StringTok{  }\KeywordTok{geom_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{geom_hline}\NormalTok{(}\DataTypeTok{yintercept =} \FloatTok{0.95}\NormalTok{, }\DataTypeTok{color =} \StringTok{"red"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"Coverage"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{show-results_files/figure-latex/unnamed-chunk-14-1.pdf}

\end{document}

### 2015-11-25 Harris_2015/simulation.Rmd
as resubmitted

@@
-# Analyses for Figure 3
+---
+output: pdf_document
+title: "Appendix 2: Simulating 20-species landscapes"
+subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
+author: David J. Harris
+fontsize: 11pt
+---
 
@@
 I used the following R function to generate "true" parameters with the method above, 
-simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.
+then simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and finally save the results to a "fakedata" folder.
 
@@
 
+Loop to simulate and save all the landscapes:
+
 ```{r}

### 2015-11-25 Harris_2015/manuscript-materials/draft.tex
use the right version for the pdf and tex

@@
 fewer than 50 sites, and the gap between the Markov network and GLMs is
-larger in this context. As shown in Appendix 4, the standard errors
+larger in this context. As shown in Appendix 5, the standard errors
 associated with the estimates in Figure 3 are small (less than 0.01), so
@@
 model's assumptions, versus 0.14 for simulations that included
-environmental heterogeneity (see Appendix 4 for confidence interval
+environmental heterogeneity (see Appendix 5 for confidence interval
 coverage across a range of \(\beta_{ij}\) values). The null model's Type
@@
 more than 5 times as often as the Markov network, even though it only
-reject the null hypothesis twice as often overall (Appendix 4). The
+reject the null hypothesis twice as often overall (Appendix 5). The
 Markov network's errors were also more concentrated around 0, so it

### 2015-11-25 Harris_2015/recovery.tex
as actually submitted to Ecology

@@
   \NormalTok{######## Markov network ########}
-  \NormalTok{rosie =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{maxit =} \DecValTok{200}\NormalTok{, }\DataTypeTok{trace =} \DecValTok{0}\NormalTok{, }\DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{), }\DataTypeTok{hessian =} \OtherTok{TRUE}\NormalTok{)}
+  \NormalTok{rosie =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{maxit =} \DecValTok{200}\NormalTok{, }\DataTypeTok{trace =} \DecValTok{0}\NormalTok{, }
+                  \DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{), }\DataTypeTok{hessian =} \OtherTok{TRUE}\NormalTok{)}
 
@@
 
-\KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }\DataTypeTok{mc.preschedule =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{mc.silent =} \OtherTok{TRUE}\NormalTok{) %>%}\StringTok{ }
+\KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }
+         \DataTypeTok{mc.preschedule =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{mc.silent =} \OtherTok{TRUE}\NormalTok{) %>%}\StringTok{ }
 \StringTok{  }\KeywordTok{bind_rows}\NormalTok{() %>%}\StringTok{ }

### 2015-11-25 Harris_2015/show-results.Rmd
as actually submitted to Ecology

@@
 
-partial_names = sapply(strsplit(grep("^>", pairs_txt, value = TRUE), " +"), function(x) x[[3]])
+partial_names = sapply(
+  strsplit(grep("^>", pairs_txt, value = TRUE), " +"), 
+  function(x) x[[3]]
+)
 filename_lines = grep("^>", pairs_txt)
@@
 
-summary(lmer(r2 ~ method + n_sites + simulation_type + (1|rep_name), data = landscape_estimates))
+summary(
+  lmer(
+    r2 ~ method + n_sites + simulation_type + (1|rep_name), 
+    data = landscape_estimates
+  )
+)
 ```
@@
 )
-mtext("A. Markov network estimates\nvs. GLM estimates", adj = 0, side = 3, font = 2, line = 1.2, cex = .9) 
+mtext("A. Markov network estimates\nvs. GLM estimates", adj = 0, side = 3, 
+      font = 2, line = 1.2, cex = .9) 
 abline(lm(`Markov network` ~ GLM, data = spread_estimates))
@@
 )
-mtext("B. Null model estimates vs.\nscaled correlation coefficients", adj = 0, side = 3, font = 2, line = 1.2, cex = .9) 
+mtext("B. Null model estimates vs.\nscaled correlation coefficients", 
+      adj = 0, side = 3, font = 2, line = 1.2, cex = .9) 
 abline(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))
@@
 )
-mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, font = 2, line = 1.2, cex = .9) 
+mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, 
+      font = 2, line = 1.2, cex = .9) 
 plotfun(truth_seq, y_markov, add = TRUE, lwd = 2)
-legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), col = c(2, 1), bty = "n")
+legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
+       col = c(2, 1), bty = "n")
 dev.off()
@@
 # is different
-rmses = markov_summary %>% filter(simulation_type != "abund") %>% mutate(rmse = sqrt((truth - estimate)^2))
+rmses = markov_summary %>% 
+  filter(simulation_type != "abund") %>% 
+  mutate(rmse = sqrt((truth - estimate)^2))
 
-ggplot(rmses, aes(x = truth, y = rmse)) + geom_point(alpha = .2, size = 1) + geom_smooth() + theme_bw() + coord_cartesian(ylim = c(0, max(rmses$rmse)))
+ggplot(rmses, aes(x = truth, y = rmse)) + 
+  geom_point(alpha = .2, size = 1) + 
+  geom_smooth() + 
+  theme_bw() + 
+  coord_cartesian(ylim = c(0, max(rmses$rmse)))
 ```

### 2015-11-25 Harris_2015/show-results.tex
as actually submitted to Ecology

@@
 
-\NormalTok{partial_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}\KeywordTok{strsplit}\NormalTok{(}\KeywordTok{grep}\NormalTok{(}\StringTok{"^>"}\NormalTok{, pairs_txt, }\DataTypeTok{value =} \OtherTok{TRUE}\NormalTok{), }\StringTok{" +"}\NormalTok{), function(x) x[[}\DecValTok{3}\NormalTok{]])}
+\NormalTok{partial_names =}\StringTok{ }\KeywordTok{sapply}\NormalTok{(}
+  \KeywordTok{strsplit}\NormalTok{(}\KeywordTok{grep}\NormalTok{(}\StringTok{"^>"}\NormalTok{, pairs_txt, }\DataTypeTok{value =} \OtherTok{TRUE}\NormalTok{), }\StringTok{" +"}\NormalTok{), }
+  \NormalTok{function(x) x[[}\DecValTok{3}\NormalTok{]]}
+\NormalTok{)}
 \NormalTok{filename_lines =}\StringTok{ }\KeywordTok{grep}\NormalTok{(}\StringTok{"^>"}\NormalTok{, pairs_txt)}
@@
 
-\KeywordTok{summary}\NormalTok{(}\KeywordTok{lmer}\NormalTok{(r2 ~}\StringTok{ }\NormalTok{method +}\StringTok{ }\NormalTok{n_sites +}\StringTok{ }\NormalTok{simulation_type +}\StringTok{ }\NormalTok{(}\DecValTok{1}\NormalTok{|rep_name), }\DataTypeTok{data =} \NormalTok{landscape_estimates))}
+\KeywordTok{summary}\NormalTok{(}
+  \KeywordTok{lmer}\NormalTok{(}
+    \NormalTok{r2 ~}\StringTok{ }\NormalTok{method +}\StringTok{ }\NormalTok{n_sites +}\StringTok{ }\NormalTok{simulation_type +}\StringTok{ }\NormalTok{(}\DecValTok{1}\NormalTok{|rep_name), }
+    \DataTypeTok{data =} \NormalTok{landscape_estimates}
+  \NormalTok{)}
+\NormalTok{)}
 \end{Highlighting}
@@
 Markov network estimates}\CharTok{\textbackslash{}n}\StringTok{vs.
-GLM estimates"}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
+GLM estimates"}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }
+      \DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
 \KeywordTok{abline}\NormalTok{(}\KeywordTok{lm}\NormalTok{(}\StringTok{`}\DataTypeTok{Markov network}\StringTok{`} \NormalTok{~}\StringTok{ }\NormalTok{GLM, }\DataTypeTok{data =} \NormalTok{spread_estimates))}
@@
 \NormalTok{)}
-\KeywordTok{mtext}\NormalTok{(}\StringTok{"B.
-Null model estimates vs.}\CharTok{\textbackslash{}n}\StringTok{scaled correlation coefficients"}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
+\KeywordTok{mtext}\NormalTok{(}\StringTok{"B. Null model estimates vs.}\CharTok{\textbackslash{}n}\StringTok{scaled correlation coefficients"}\NormalTok{, }
+      \DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
 \KeywordTok{abline}\NormalTok{(}\KeywordTok{lm}\NormalTok{(null ~}\StringTok{ }\KeywordTok{I}\NormalTok{(correlation*}\KeywordTok{sqrt}\NormalTok{(n_sites)), }\DataTypeTok{data =} \NormalTok{spread_estimates))}
@@
 \KeywordTok{mtext}\NormalTok{(}\StringTok{"C.
-Error rate vs.}\CharTok{\textbackslash{}n}\StringTok{interaction strength"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }\DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
+Error rate vs.}\CharTok{\textbackslash{}n}\StringTok{interaction strength"}\NormalTok{, }\DataTypeTok{side =} \DecValTok{3}\NormalTok{, }\DataTypeTok{adj =} \DecValTok{0}\NormalTok{, }
+      \DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
 \KeywordTok{plotfun}\NormalTok{(truth_seq, y_markov, }\DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
-\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Null model"}\NormalTok{, }\StringTok{"Markov network"}\NormalTok{), }\DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Null model"}\NormalTok{, }\StringTok{"Markov network"}\NormalTok{), }
+       \DataTypeTok{col =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{bty =} \StringTok{"n"}\NormalTok{)}
 \KeywordTok{dev.off}\NormalTok{()}
@@
 \CommentTok{# is different}
-\NormalTok{rmses =}\StringTok{ }\NormalTok{markov_summary %>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(simulation_type !=}\StringTok{ "abund"}\NormalTok{) %>%}\StringTok{ }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{rmse =} \KeywordTok{sqrt}\NormalTok{((truth -}\StringTok{ }\NormalTok{estimate)^}\DecValTok{2}\NormalTok{))}
+\NormalTok{rmses =}\StringTok{ }\NormalTok{markov_summary %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type !=}\StringTok{ "abund"}\NormalTok{) %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{rmse =} \KeywordTok{sqrt}\NormalTok{((truth -}\StringTok{ }\NormalTok{estimate)^}\DecValTok{2}\NormalTok{))}
 
-\KeywordTok{ggplot}\NormalTok{(rmses, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{rmse)) +}\StringTok{ }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \NormalTok{.}\DecValTok{2}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }\KeywordTok{geom_smooth}\NormalTok{() +}\StringTok{ }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{max}\NormalTok{(rmses$rmse)))}
+\KeywordTok{ggplot}\NormalTok{(rmses, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{rmse)) +}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \NormalTok{.}\DecValTok{2}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{() +}\StringTok{ }
+\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
+\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{max}\NormalTok{(rmses$rmse)))}
 \end{Highlighting}

### 2015-11-25 Harris_2015/manuscript-materials/draft.Rmd
for submission to biorxiv

@@
+---
+output: pdf_document
+---
 \chead{\ifthenelse{\value{page}=1}{}{\textit{Inferring species interactions}}}
@@
 methods will not provide much evidence regarding species interactions.
+
+![**A.** A small network of three competing species.
+The tree (top) tends not to co-occur with either of the two shrub species, as
+indicated by the strongly negative coefficient linking them. The two shrub
+species also compete with one another, but more weakly (circled coefficient).
+**B.** In spite of the competitive interactions between the two shrub species,
+their shared tendency to occur in locations without trees makes their occurrence
+vectors positively correlated (circled). **C.** Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a Markov
+network leads to correct identification of the negative shrub-shrub interaction
+(circled).](figures/figure_1.pdf)
 
@@
 summing over the different possible assemblages (Figure 2B).
+
+![**A.** A small Markov network, defined by its $\alpha$ and
+$\beta$ values.  The abiotic environment favors the occurrence of each species
+($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
+$\beta_{12}$ coefficient is consistent with competition between the two species.
+**B.** The coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
+to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
+absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
+species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
+present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
+probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
+frequencies of all possible co-occurrence patterns between the two species of
+interest, as calculated in the previous panel. **D.** Without competition (i.e.
+with $\beta_{12}=0$, each species would occur more often.](figures/figure_2.pdf)
 
@@
 
+![Proportion of variance in interaction coefficients explained by
+each method versus number of sampled locations across the three simulation
+types. For the null model (Pairs), two outliers with $|Z|>1000$ were manually
+adjusted to $|Z|=50$ to mitigate their detrimental influence on $R^2$ (Appendix
+5).](figures/performance.pdf)
+
+![**A.** The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates. **B.** The null
+model's estimates typically matched the (negative) correlation coefficient,
+after controlling for landscape size. **C.** For any given interaction strength,
+the null model was much more likely to misclassify its sign with 95% confidence
+than the Markov network was.](figures/error_rates.pdf)
+
 Of the methods that did not control for indirect effects, Figure 3 shows that
@@
 \setlength{\parskip}{0pt}
-
-<!--references go here-->
-
-\setlength{\parindent}{0cm}
-\setlength{\leftskip}{0cm}
-
-\noindent \textbf{Figures:}
-
-**Figure 1: A.** A small network of three competing species.
-The tree (top) tends not to co-occur with either of the two shrub species, as
-indicated by the strongly negative coefficient linking them. The two shrub
-species also compete with one another, but more weakly (circled coefficient).
-**B.** In spite of the competitive interactions between the two shrub species,
-their shared tendency to occur in locations without trees makes their occurrence
-vectors positively correlated (circled). **C.** Controlling for trees with a
-conditional (all-else-equal) approach such as a partial covariance or a Markov
-network leads to correct identification of the negative shrub-shrub interaction
-(circled).
-
-**Figure 2: A.** A small Markov network, defined by its $\alpha$ and
-$\beta$ values.  The abiotic environment favors the occurrence of each species
-($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
-$\beta_{12}$ coefficient is consistent with competition between the two species.
-**B.** The coefficients determine the probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. $\alpha_1$ is added
-to the exponent whenever Species 1 is present ($y_1 = 1$), but not when it is
-absent ($y_1 = 0$).  Similarly, the exponent includes $\alpha_2$ only when
-species $2$ is present ($y_2 = 1$), and includes $\beta_{12}$ only when both are
-present ($y_1y_2 = 1$). The normalizing constant $Z$, ensures that the four
-probabilities sum to 1.  In this case, $Z$ is about 18.5.  **C.** The expected
-frequencies of all possible co-occurrence patterns between the two species of
-interest, as calculated in the previous panel. **D.** Without competition (i.e.
-with $\beta_{12}=0$, each species would occur more often.
-
-**Figure 3:** Proportion of variance in interaction coefficients explained by
-each method versus number of sampled locations across the three simulation
-types. For the null model (Pairs), two outliers with $|Z|>1000$ were manually
-adjusted to $|Z|=50$ to mitigate their detrimental influence on $R^2$ (Appendix
-5).
-
-**Figure 4: A.** The Markov network's estimated interaction
-coefficients were generally very similar to the GLM estimates. **B.** The null
-model's estimates typically matched the (negative) correlation coefficient,
-after controlling for landscape size. **C.** For any given interaction strength,
-the null model was much more likely to misclassify its sign with 95% confidence
-than the Markov network was.

### 2015-11-25 Harris_2015/manuscript-materials/draft.tex
for submission to biorxiv

--- 2015-11-25 Harris_2015/manuscript-materials/draft.tex
@@
 \geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
 \date{}
+
+\def\tightlist{}
 
@@
 species interactions.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/figure_1.pdf}
+\caption{\textbf{A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a
+Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled).}
+\end{figure}
 
@@
 2B).
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/figure_2.pdf}
+\caption{\textbf{A.} A small Markov network, defined by its \(\alpha\)
+and \(\beta\) values. The abiotic environment favors the occurrence of
+each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
+consistent with competition between the two species. \textbf{B.} The
+coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
+is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
+but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
+includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
+and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
+The normalizing constant \(Z\), ensures that the four probabilities sum
+to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest, as calculated in the previous panel. \textbf{D.}
+Without competition (i.e.~with \(\beta_{12}=0\), each species would
+occur more often.}
+\end{figure}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/performance.pdf}
+\caption{Proportion of variance in interaction coefficients explained by
+each method versus number of sampled locations across the three
+simulation types. For the null model (Pairs), two outliers with
+\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
+detrimental influence on \(R^2\) (Appendix 5).}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{figures/error_rates.pdf}
+\caption{\textbf{A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates.
+\textbf{B.} The null model's estimates typically matched the (negative)
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was.}
+\end{figure}
+
 Of the methods that did not control for indirect effects, Figure 3 shows
@@
 \setlength{\parindent}{0cm}
-\setlength{\parskip}{0pt}
 
@@
 \setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
+\setlength{\parskip}{0pt}
 
@@
 Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
-
-\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}
-
-\noindent \textbf{Figures:}
-
-\textbf{Figure 1: A.} A small network of three competing species. The
-tree (top) tends not to co-occur with either of the two shrub species,
-as indicated by the strongly negative coefficient linking them. The two
-shrub species also compete with one another, but more weakly (circled
-coefficient). \textbf{B.} In spite of the competitive interactions
-between the two shrub species, their shared tendency to occur in
-locations without trees makes their occurrence vectors positively
-correlated (circled). \textbf{C.} Controlling for trees with a
-conditional (all-else-equal) approach such as a partial covariance or a
-Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled).
-
-\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
-and \(\beta\) values. The abiotic environment favors the occurrence of
-each species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
-consistent with competition between the two species. \textbf{B.} The
-coefficients determine the probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
-is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
-but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
-includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
-and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
-The normalizing constant \(Z\), ensures that the four probabilities sum
-to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest, as calculated in the previous panel. \textbf{D.}
-Without competition (i.e.~with \(\beta_{12}=0\), each species would
-occur more often.
-
-\textbf{Figure 3:} Proportion of variance in interaction coefficients
-explained by each method versus number of sampled locations across the
-three simulation types. For the null model (Pairs), two outliers with
-\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
-detrimental influence on \(R^2\) (Appendix 5).
-
-\textbf{Figure 4: A.} The Markov network's estimated interaction
-coefficients were generally very similar to the GLM estimates.
-\textbf{B.} The null model's estimates typically matched the (negative)
-correlation coefficient, after controlling for landscape size.
-\textbf{C.} For any given interaction strength, the null model was much
-more likely to misclassify its sign with 95\% confidence than the Markov
-network was.
-
 \end{flushleft}

### 2016-02-11 Harris_2015/manuscript-materials/draft.Rmd
editor comments

--- 2015-11-25 Harris_2015/manuscript-materials/draft.Rmd
@@
 the direct and indirect consequences of any possible species interaction matrix.
-Interactions in these models can also be estimated from observed co-occurrence
+Interactions in these models can be estimated from observed co-occurrence
 rates via maximum likelihood, controlling for indirect effects. Using simulated
@@
 obscured by their shared tendency to occur in unshaded areas (Figure 1B). While
-ecologists have long known that indirect effects can overwhelm direct ones at
-the landscape level [@levine_competitive_1976], the
-vast majority of our methods for drawing inferences from observational data do
-not control for these effects [e.g. @diamond_island_1975;
+ecologists have long known that indirect effects can overwhelm direct ones 
+[@levine_competitive_1976], most methods for drawing inferences from 
+co-occurrence data do not control for these effects [e.g. @diamond_island_1975;
 @strong_ecological_1984; @gotelli_empirical_2009; @veech_probabilistic_2013;
-@pollock_understanding_2014]. To the extent that indirect interactions like
-those in Figure 1 are generally important, existing
-methods will not provide much evidence regarding species interactions.
+@pollock_understanding_2014]. As a result, ecologists do not have tools that allow
+them to isolate direct interactions from indirect effects.
 
@@
 whole-landscape level (as most methods assume), it does still leave a
-signal in the data (Figure 1C). For example, after partitioning the data set
-into shaded and unshaded sites, there will be co-occurrence deficits in
-each subset that wouldn't otherwise be apparent. More generally, controlling for
-other species in the network will often be important for obtaining reliable
-estimates of direct (conditional, or all-else-equal) effects.  This kind of
-precision is difficult to obtain from null models, which only test the most
-extreme possible hypothesis: that *all* direct and indirect interactions are
-exactly zero. Nevertheless, null models have dominated this field for more than
-three decades [@strong_ecological_1984; @gotelli_empirical_2009].
+signal in the data (Figure 1C). After controlling for the presence
+of the tree species (e.g. by splitting the data set into shaded and unshaded 
+sites or by using a model that estimates conditional relationships amongs 
+species), the two shrubs do have a negative association, with fewer co-occurrences 
+in unshaded areas than would be expected if they occurred independently of one
+another.
 
@@
 and for estimating them from observed presence-absence matrices. Markov networks
-have been used in many scientific fields in similar contexts for decades, from
+have been used in diverse scientific fields in similar contexts for decades, from
 physics [where nearby particles interact magnetically; @cipra_introduction_1987]
@@
 about the direct and indirect consequences of an interaction matrix, and also to
-make inferences about the species interactions that contributed to an observed
-set of co-occurrences. Using simulated data sets where the "true" interactions
-are known, I compare this approach with several existing methods. Finally, I
+make inferences about the interaction matrix based on co-occurrence rates. Using 
+simulated data sets where the "true" interactions are known, I compare this 
+approach with several existing methods. Finally, I
 discuss opportunities for extending the approach presented here to other
@@
 competitor is associated with release from another competitor), then predicting
-species' overall co-occurrence rates can be more complicated, and may require
-summing over the different possible assemblages (Figure 2B).
+species' overall co-occurrence rates would require normalizing the relative 
+probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).
 
@@
 an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
-indicating which species are present at each site on the landscape). When the
-number of species is reasonably small (less than about 30), one can find
-exact maximum likelihood estimates for all of the $\alpha$ and $\beta$
-coefficients given a presence-absence matrix by numerically
-optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
-ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
-ensuring that this procedure will converge on the global maximum. I used the
-rosalia package [@harris_rosalia_2015] for the R programming language
-[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its
-gradients [@murphy_machine_2012]; the package passes these functions to the
-"BFGS" method in R's general-purpose optimizer, which then estimates the Markov
-network parameters.
+indicating which species are present at each site on the landscape). When
+Equation 1 can be normalized, one can find exact maximum likelihood estimates for
+$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like the 
+ones considered here have unimodal likelihood surfaces [@murphy_machine_2012], so 
+these estimates will represent the global optimum. I used the rosalia package 
+[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
+calculate $p(\vec{y}; \alpha, \beta)$ and its gradients [@murphy_machine_2012]; 
+the package passes these functions to the "BFGS" method in R's general-purpose 
+optimizer, which then estimates the Markov network parameters.
 
@@
 
-I then generated landscapes with up to 20 interacting species at 25, 200, or
+I then simulated landscapes with up to 20 interacting species at 25, 200, or
 1600 sites using three increasingly complex models (50 replicates for each
-combination of size and model; see Appendix 2 for details). I randomly drew the
-“true” coefficient magnitudes for each replicate landscape from exponential
-distributions so that most species pairs interacted negligibly but a few pairs
-interacted strongly enough that their effects could propagate indirectly to
-other species in the network.
+combination of size and model). The simplest set of simulated landscapes were
+generated with Gibbs samples from Equation 1, and provided a test of methods'
+ability to isolate direct species-level interactions from a web of indirect 
+ones. For each replicate, I randomly drew the “true” $\beta$ coefficient
+magnitudes an exponential distribution with rate 1 so that
+most species pairs interacted negligibly but a few interactions were strong
+enough to propagate through the newtwork. I randomly assigned 25% of the
+interactions to be positive; the remainder were negative.
 
-The first set of 20-species landscapes, like the landscapes with three species,
-were generated directly from a Markov network to ensure that the model could
-recover the parameters used to generate the "observed" co-occurrence data.
-Then, I added two environmental factors that varied from location to location
-across the simulated landscapes, and simulated a new set of co-occurrence data
-so that species' $\alpha$ coefficients depended on the local environment.
-The latter set of simulated landscapes provide an important test of the methods’
-ability to distinguish co-occurrence patterns that were generated from pairwise
-biotic interactions from those that were generated by external forces like
-abiotic environmental filtering. This task was made especially difficult
-because---as with most analyses of presence-absence data for co-occurrence
-patterns---the inference procedure did not have access to any information about
-the environmental or spatial variables that helped shape the landscape [cf
-@connor_checkered_2013].  I generated the final set of landscapes with an
-abundance-based model that included per-capita interaction rates instead of
-per-species interaction rates.
+The next set of landscapes provided a way to assess each method's ability to 
+identify direct interactions in the presence of environmental heterogeneity.
+Here, the $\beta$ coefficients were calculated as above, but each species'
+$\alpha$ value depended linearly on two normally-distributed environmental 
+factors, which were drawn from independent Gaussians for each site. Once the 
+local $\alpha$ values were calculated, independent Gibbs samplers determined 
+the species composition for each site based on Equation 1.
+
+In the third set of landscapes, I simulated each species' abundance (instead of
+just their presence/absence), and interactions between species occurred on a 
+per-capita basis. To avoid runaway mutualisms leading to infinite abundance,
+all interaction coefficients were negative in these simulations. If a method
+performs well on these landscapes, it would indicate some robustness to the
+biolocical mechanisms underlying species' interactiosn with one another.
 
@@
 I also evaluated six alternative methods: five from the existing literature,
-plus a novel combination of two of these methods. The first alternative interaction
-metric was the sample correlation between species' presence-absence vectors,
-which summarizes their marginal association.  Next, I used partial correlations,
-which summarize species' conditional relationships. This approach is
-common in molecular biology [@friedman_sparse_2008], but is rare in ecology
-(see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions).
-In the context of non-Gaussian data, the partial correlation can be thought of
-as a computationally efficient approximation to the full Markov network model
-[@loh_structure_2013]. Because partial correlations are undefined for
-landscapes with perfectly-correlated species pairs, I used a regularized
-estimate provided by the corpcor package’s `pcor.shrink` function with the
-default settings [@schafer_corpcor_2014].
+plus a novel combination of two of these methods. The first alternative 
+interaction metric was the sample correlation between species' presence-absence
+vectors, which summarizes their marginal association.  Next, I used partial 
+correlations, which summarize species' conditional relationships. This approach, 
+which is  closely related to linear regression, is common in molecular biology
+[@friedman_sparse_2008], but is rare in ecology (see @albrecht_spatial_2001 and
+@faisal_inferring_2010 for two exceptions). In the context of non-Gaussian data,
+the partial correlation can be thought of as a computationally efficient
+approximation to the full Markov network model [@loh_structure_2013]. Partial
+correlations are undefined for landscapes with perfectly-correlated species
+pairs, so I used the regularized estimate provided by the corpcor package’s 
+`pcor.shrink` function with the default settings [@schafer_corpcor_2014].
 
@@
 
-The next method, described in @gotelli_empirical_2009, involved
-simulating new landscapes from a null model that retains the row and column
-sums of the original matrix [@strong_ecological_1984]. I used the $Z$-scores
-computed by the Pairs software described in @gotelli_empirical_2009 as my null
-model-based estimator of species interactions.
+The next method used the Pairs software described in @gotelli_empirical_2009.
+This program simulating new landscapes from a null model that retains the row 
+and column sums of the original matrix [@strong_ecological_1984] and calculates
+$Z$-scores to summarize a species pair's deviation from this null.
 
@@
 proportion of variance explained for different combinations of model type and
-landscape size (compared with a null model that assumed all interaction
+landscape size (compared with a baseline model that assumed all interaction
 strengths to be zero).
+
+For the null model and the Markov network, the probability of rejecting the
+null hypothesis of zero interaction was estimated across a range of "true"
+interaction strengths using a generalized additive model with binomial responses
+(reject versus not; Appendix 4). The Type I error rate was defined as this
+model's estimated probability of rejecting the null hypothesis when the "true"
+interaction was zero.
 
@@
 sets were very large (Figure 3). As reviewed in @gotelli_empirical_2009,
-however, most analyses in this field of ecology involve fewer than 50 sites,
-and the gap between the Markov network and GLMs is larger in this context. As
-shown in Appendix 4, the standard errors associated with the estimates in Figure
-3 are small (less than 0.01), so the differences among methods should not be
-attributed to sampling error.
+however, most analyses in this field of ecology involve fewer than 50 sites;
+in this context, the gap between the methods was larger. As shown in Appendix 4,
+the standard errors associated with the estimates in Figure 3 are small (less
+than 0.01), so the differences among methods should not be attributed to 
+sampling error.
 
@@
 true interaction strengths than either the joint species distribution model
-(BayesComm) or the null model (Pairs). The estimates from these approaches
-were tightly correlated (after controlling for the size of the landscape)
-suggesting that the null model only contains a noisy version of the
-same information that could be obtained more easily and interpretably with
-simple correlation coefficients (Figure 4B).
+(BayesComm) or the null model (Pairs). 95% of the variance the Pairs test 
+statistic was explained by correlation coefficients (controlling for landscape
+size; Figure 4B); much of the remaining variance is due to sampling error.
 
 Finally, we can evaluate the models' statistical inferences (focusing on the
-first two simulation types, for which the true interaction rates are easiest to
-interpret).  The Markov network's approximate Type I error rate (defined here
-as the probability that 0 fell outside the 95% confidence interval for a pair of
-species where $|\beta_{ij}|<0.1$) depended on the simulation type: 0.02 for
-simulations that matched the model's assumptions, versus 0.14 for simulations
-that included environmental heterogeneity (see Appendix 4 for confidence
-interval coverage across a range of $\beta_{ij}$ values). The null model's
-Type I error rates were 0.30 and 0.50 for the constant and heterogeneous
-landscapes, respectively---far higher than the nominal 0.05 rate. Figure
-4C shows, across a range of true interaction strengths, the probability that
-the null model or the Markov network will predict the wrong sign of the
-interaction with 95% confidence.  The null model makes such errors more than 5 times
-as often as the Markov network, even though it only reject the null hypothesis
-twice as often overall (Appendix 4).  The Markov network's errors were also
-more concentrated around 0, so it never misclassified strong interactions like
-the null model did.
+first two simulation types, where the interaction coefficients are easiest to
+interpret).  The Markov network's Type I error rate depended on the 
+simulation type: [[0.02]] for simulations that matched the model's assumptions,
+versus [[0.14]] for simulations that included environmental heterogeneity (see 
+Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
+values). The null model's Type I error rates were [[0.30]] and [[0.50]] for the
+constant and heterogeneous landscapes, respectively---far higher than the 
+nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
+strengths, the probability that the null model or the Markov network will 
+predict the wrong sign of the interaction with 95% confidence.  The null model 
+makes such errors more than [[5]] times as often as the Markov network, even
+though it only rejects the null hypothesis [[twice]] as often overall (Appendix
+4). The Markov network's errors were also more concentrated around 0, so it 
+never misclassified strong interactions like the null model did.
 
@@
 
-<!--references go here-->
-
-\setlength{\parindent}{0cm}
-\setlength{\leftskip}{0cm}
-
-\noindent \textbf{Figures:}
-
 **Figure 1: A.** A small network of three competing species.
@@
 network leads to correct identification of the negative shrub-shrub interaction
-(circled).
+(circled). See Appendix 1 and the results for "three species" for more details.
 

### 2016-02-11 Harris_2015/manuscript-materials/draft.tex
editor comments

--- 2015-11-25 Harris_2015/manuscript-materials/draft.tex
@@
 
+\def\tightlist{}
+
 \begin{document}
@@
 predict the direct and indirect consequences of any possible species
-interaction matrix. Interactions in these models can also be estimated
-from observed co-occurrence rates via maximum likelihood, controlling
-for indirect effects. Using simulated landscapes with known pairwise
+interaction matrix. Interactions in these models can be estimated from
+observed co-occurrence rates via maximum likelihood, controlling for
+indirect effects. Using simulated landscapes with known pairwise
 interaction strengths, I evaluated Markov networks and six existing
@@
 1B). While ecologists have long known that indirect effects can
-overwhelm direct ones at the landscape level (Levine 1976), the vast
-majority of our methods for drawing inferences from observational data
-do not control for these effects (e.g. Diamond 1975, Strong et al. 1984,
-Gotelli and Ulrich 2009, Veech 2013, Pollock et al. 2014). To the extent
-that indirect interactions like those in Figure 1 are generally
-important, existing methods will not provide much evidence regarding
-species interactions.
+overwhelm direct ones (Levine 1976), most methods for drawing inferences
+from co-occurrence data do not control for these effects (e.g. Diamond
+1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock
+et al. 2014). As a result, ecologists do not have tools that allow them
+to isolate direct interactions from indirect effects.
 
@@
 whole-landscape level (as most methods assume), it does still leave a
-signal in the data (Figure 1C). For example, after partitioning the data
-set into shaded and unshaded sites, there will be co-occurrence deficits
-in each subset that wouldn't otherwise be apparent. More generally,
-controlling for other species in the network will often be important for
-obtaining reliable estimates of direct (conditional, or all-else-equal)
-effects. This kind of precision is difficult to obtain from null models,
-which only test the most extreme possible hypothesis: that \emph{all}
-direct and indirect interactions are exactly zero. Nevertheless, null
-models have dominated this field for more than three decades (Strong et
-al. 1984, Gotelli and Ulrich 2009).
+signal in the data (Figure 1C). After controlling for the presence of
+the tree species (e.g.~by splitting the data set into shaded and
+unshaded sites or by using a model that estimates conditional
+relationships amongs species), the two shrubs do have a negative
+association, with fewer co-occurrences in unshaded areas than would be
+expected if they occurred independently of one another.
 
@@
 from observed presence-absence matrices. Markov networks have been used
-in many scientific fields in similar contexts for decades, from physics
-(where nearby particles interact magnetically; Cipra 1987) to spatial
-statistics (where adjacent grid cells have correlated values; Harris
-1974, Gelfand et al. 2005). While community ecologists explored some
-related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
-used severe approximations that led to unintelligible results (e.g.
+in diverse scientific fields in similar contexts for decades, from
+physics (where nearby particles interact magnetically; Cipra 1987) to
+spatial statistics (where adjacent grid cells have correlated values;
+Harris 1974, Gelfand et al. 2005). While community ecologists explored
+some related approaches in the 1980's (Whittam and Siegel-Causey 1981),
+they used severe approximations that led to unintelligible results (e.g.
 ``probabilities'' greater than one; Gilpin and Diamond 1982).
@@
 predictions about the direct and indirect consequences of an interaction
-matrix, and also to make inferences about the species interactions that
-contributed to an observed set of co-occurrences. Using simulated data
-sets where the ``true'' interactions are known, I compare this approach
-with several existing methods. Finally, I discuss opportunities for
-extending the approach presented here to other problems in community
-ecology, e.g.~quantifying the overall effect of species interactions on
-occurrence rates (Roughgarden 1983) and disentangling the effects of
-biotic versus abiotic interactions on species composition (Pollock et
-al. 2014).
+matrix, and also to make inferences about the interaction matrix based
+on co-occurrence rates. Using simulated data sets where the ``true''
+interactions are known, I compare this approach with several existing
+methods. Finally, I discuss opportunities for extending the approach
+presented here to other problems in community ecology, e.g.~quantifying
+the overall effect of species interactions on occurrence rates
+(Roughgarden 1983) and disentangling the effects of biotic versus
+abiotic interactions on species composition (Pollock et al. 2014).
 
@@
 one competitor is associated with release from another competitor), then
-predicting species' overall co-occurrence rates can be more complicated,
-and may require summing over the different possible assemblages (Figure
-2B).
+predicting species' overall co-occurrence rates would require
+normalizing the relative probabilities of possible \(\vec{y}\) vectors
+so that they sum to one (Figure 2B).
 
@@
 a set of independent \(\vec{y}\) vectors indicating which species are
-present at each site on the landscape). When the number of species is
-reasonably small (less than about 30), one can find exact maximum
-likelihood estimates for all of the \(\alpha\) and \(\beta\)
-coefficients given a presence-absence matrix by numerically optimizing
-\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
-ones considered here have unimodal likelihood surfaces (Murphy 2012),
-ensuring that this procedure will converge on the global maximum. I used
-the rosalia package (Harris 2015a) for the R programming language (R
-Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
+present at each site on the landscape). When Equation 1 can be
+normalized, one can find exact maximum likelihood estimates for
+\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
+networks like the ones considered here have unimodal likelihood surfaces
+(Murphy 2012), so these estimates will represent the global optimum. I
+used the rosalia package (Harris 2015a) for the R programming language
+(R Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
 gradients (Murphy 2012); the package passes these functions to the
@@
 
-I then generated landscapes with up to 20 interacting species at 25,
+I then simulated landscapes with up to 20 interacting species at 25,
 200, or 1600 sites using three increasingly complex models (50
-replicates for each combination of size and model; see Appendix 2 for
-details). I randomly drew the ``true'' coefficient magnitudes for each
-replicate landscape from exponential distributions so that most species
-pairs interacted negligibly but a few pairs interacted strongly enough
-that their effects could propagate indirectly to other species in the
-network.
+replicates for each combination of size and model). The simplest set of
+simulated landscapes were generated with Gibbs samples from Equation 1,
+and provided a test of methods' ability to isolate direct species-level
+interactions from a web of indirect ones. For each replicate, I randomly
+drew the ``true'' \(\beta\) coefficient magnitudes an exponential
+distribution with rate 1 so that most species pairs interacted
+negligibly but a few interactions were strong enough to propagate
+through the newtwork. I randomly assigned 25\% of the interactions to be
+positive; the remainder were negative.
 
-The first set of 20-species landscapes, like the landscapes with three
-species, were generated directly from a Markov network to ensure that
-the model could recover the parameters used to generate the ``observed''
-co-occurrence data. Then, I added two environmental factors that varied
-from location to location across the simulated landscapes, and simulated
-a new set of co-occurrence data so that species' \(\alpha\) coefficients
-depended on the local environment. The latter set of simulated
-landscapes provide an important test of the methods' ability to
-distinguish co-occurrence patterns that were generated from pairwise
-biotic interactions from those that were generated by external forces
-like abiotic environmental filtering. This task was made especially
-difficult because---as with most analyses of presence-absence data for
-co-occurrence patterns---the inference procedure did not have access to
-any information about the environmental or spatial variables that helped
-shape the landscape (cf Connor et al. 2013). I generated the final set
-of landscapes with an abundance-based model that included per-capita
-interaction rates instead of per-species interaction rates.
+The next set of landscapes provided a way to assess each method's
+ability to identify direct interactions in the presence of environmental
+heterogeneity. Here, the \(\beta\) coefficients were calculated as
+above, but each species' \(\alpha\) value depended linearly on two
+normally-distributed environmental factors, which were drawn from
+independent Gaussians for each site. Once the local \(\alpha\) values
+were calculated, independent Gibbs samplers determined the species
+composition for each site based on Equation 1.
+
+In the third set of landscapes, I simulated each species' abundance
+(instead of just their presence/absence), and interactions between
+species occurred on a per-capita basis. To avoid runaway mutualisms
+leading to infinite abundance, all interaction coefficients were
+negative in these simulations. If a method performs well on these
+landscapes, it would indicate some robustness to the biolocical
+mechanisms underlying species' interactiosn with one another.
 
@@
 association. Next, I used partial correlations, which summarize species'
-conditional relationships. This approach is common in molecular biology
-(Friedman et al. 2008), but is rare in ecology (see Albrecht and Gotelli
-(2001) and Faisal et al. (2010) for two exceptions). In the context of
-non-Gaussian data, the partial correlation can be thought of as a
-computationally efficient approximation to the full Markov network model
-(Loh and Wainwright 2013). Because partial correlations are undefined
-for landscapes with perfectly-correlated species pairs, I used a
-regularized estimate provided by the corpcor package's
-\texttt{pcor.shrink} function with the default settings (Schäfer et al.
-2014).
+conditional relationships. This approach, which is closely related to
+linear regression, is common in molecular biology (Friedman et al.
+2008), but is rare in ecology (see Albrecht and Gotelli (2001) and
+Faisal et al. (2010) for two exceptions). In the context of non-Gaussian
+data, the partial correlation can be thought of as a computationally
+efficient approximation to the full Markov network model (Loh and
+Wainwright 2013). Partial correlations are undefined for landscapes with
+perfectly-correlated species pairs, so I used the regularized estimate
+provided by the corpcor package's \texttt{pcor.shrink} function with the
+default settings (Schäfer et al. 2014).
 
@@
 
-The next method, described in Gotelli and Ulrich (2009), involved
-simulating new landscapes from a null model that retains the row and
-column sums of the original matrix (Strong et al. 1984). I used the
-\(Z\)-scores computed by the Pairs software described in Gotelli and
-Ulrich (2009) as my null model-based estimator of species interactions.
+The next method used the Pairs software described in Gotelli and Ulrich
+(2009). This program simulating new landscapes from a null model that
+retains the row and column sums of the original matrix (Strong et al.
+1984) and calculates \(Z\)-scores to summarize a species pair's
+deviation from this null.
 
@@
 explained for different combinations of model type and landscape size
-(compared with a null model that assumed all interaction strengths to be
-zero).
+(compared with a baseline model that assumed all interaction strengths
+to be zero).
+
+For the null model and the Markov network, the probability of rejecting
+the null hypothesis of zero interaction was estimated across a range of
+``true'' interaction strengths using a generalized additive model with
+binomial responses (reject versus not; Appendix 4). The Type I error
+rate was defined as this model's estimated probability of rejecting the
+null hypothesis when the ``true'' interaction was zero.
 
@@
 Ulrich (2009), however, most analyses in this field of ecology involve
-fewer than 50 sites, and the gap between the Markov network and GLMs is
-larger in this context. As shown in Appendix 4, the standard errors
-associated with the estimates in Figure 3 are small (less than 0.01), so
-the differences among methods should not be attributed to sampling
-error.
+fewer than 50 sites; in this context, the gap between the methods was
+larger. As shown in Appendix 4, the standard errors associated with the
+estimates in Figure 3 are small (less than 0.01), so the differences
+among methods should not be attributed to sampling error.
 
@@
 of species' true interaction strengths than either the joint species
-distribution model (BayesComm) or the null model (Pairs). The estimates
-from these approaches were tightly correlated (after controlling for the
-size of the landscape) suggesting that the null model only contains a
-noisy version of the same information that could be obtained more easily
-and interpretably with simple correlation coefficients (Figure 4B).
+distribution model (BayesComm) or the null model (Pairs). 95\% of the
+variance the Pairs test statistic was explained by correlation
+coefficients (controlling for landscape size; Figure 4B); much of the
+remaining variance is due to sampling error.
 
 Finally, we can evaluate the models' statistical inferences (focusing on
-the first two simulation types, for which the true interaction rates are
-easiest to interpret). The Markov network's approximate Type I error
-rate (defined here as the probability that 0 fell outside the 95\%
-confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
-depended on the simulation type: 0.02 for simulations that matched the
-model's assumptions, versus 0.14 for simulations that included
-environmental heterogeneity (see Appendix 4 for confidence interval
-coverage across a range of \(\beta_{ij}\) values). The null model's Type
-I error rates were 0.30 and 0.50 for the constant and heterogeneous
-landscapes, respectively---far higher than the nominal 0.05 rate. Figure
-4C shows, across a range of true interaction strengths, the probability
-that the null model or the Markov network will predict the wrong sign of
-the interaction with 95\% confidence. The null model makes such errors
-more than 5 times as often as the Markov network, even though it only
-reject the null hypothesis twice as often overall (Appendix 4). The
+the first two simulation types, where the interaction coefficients are
+easiest to interpret). The Markov network's Type I error rate depended
+on the simulation type: {[}{[}0.02{]}{]} for simulations that matched
+the model's assumptions, versus {[}{[}0.14{]}{]} for simulations that
+included environmental heterogeneity (see Appendix 4 for confidence
+interval coverage across a range of \(\beta_{ij}\) values). The null
+model's Type I error rates were {[}{[}0.30{]}{]} and {[}{[}0.50{]}{]}
+for the constant and heterogeneous landscapes, respectively---far higher
+than the nominal 0.05 rate. Figure 4C shows, across a range of true
+interaction strengths, the probability that the null model or the Markov
+network will predict the wrong sign of the interaction with 95\%
+confidence. The null model makes such errors more than {[}{[}5{]}{]}
+times as often as the Markov network, even though it only rejects the
+null hypothesis {[}{[}twice{]}{]} as often overall (Appendix 4). The
 Markov network's errors were also more concentrated around 0, so it
@@
 \setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
+\setlength{\parskip}{0pt}
+
+\textbf{Figure 1: A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for trees with a
+conditional (all-else-equal) approach such as a partial covariance or a
+Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled). See Appendix 1 and the results for
+``three species'' for more details.
+
+\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
+and \(\beta\) values. The abiotic environment favors the occurrence of
+each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
+consistent with competition between the two species. \textbf{B.} The
+coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
+is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
+but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
+includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
+and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
+The normalizing constant \(Z\), ensures that the four probabilities sum
+to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest, as calculated in the previous panel. \textbf{D.}
+Without competition (i.e.~with \(\beta_{12}=0\), each species would
+occur more often.
+
+\textbf{Figure 3:} Proportion of variance in interaction coefficients
+explained by each method versus number of sampled locations across the
+three simulation types. For the null model (Pairs), two outliers with
+\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
+detrimental influence on \(R^2\) (Appendix 5).
+
+\textbf{Figure 4: A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates.
+\textbf{B.} The null model's estimates typically matched the (negative)
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was.
 

### 2016-02-11 stochastic/chapter-3.Rmd
some chapter 3 work

@@
 ---
-output: pdf_document
-title: "Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation"
+output: 
+  pdf_document:
+    fig_caption: yes
+title: "Estimating species interactions in large, abiotically structured communities using Markov networks and stochastic approximation"
+author: "David J. Harris"
+bibliography: "My Library.bib"
+csl: ecology.csl
+fontsize: 12pt
 ---
+
+```{r, echo = FALSE}
+# bibtex ids with colons in them (eg R package names) don't work. Remove the pattern :_
+cmd = 'sed -i "" "s/:_/_/g" My\\ Library.bib'
+
+system(cmd)
+```
 
@@
 Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and thus attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions.
-In this paper, I introduce stochastic approximation as an alternative method for fitting these models that addresses both of these problems, making it feasible to use Markov networksin cases where each species responds to multiple abiotic factors and hundreds of competitors or mutualists.
+In this paper, I introduce stochastic approximation as an alternative method for fitting these models that addresses both of these problems, making it feasible to use Markov networks in cases where each species responds to multiple abiotic factors and hundreds of competitors or mutualists.
 While stochastic approximation does introduce some sampling noise during the optimization process, it still converges to the maximum likelihood estimate for species' interaction coefficients with probability one.
@@
 @friedman_sparse_2008).
-The discrepancy is due to the fact that ecologists have chosen to focus, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al.
-2014; see Faisal et al.
-2010 and Harris 2015 for two recent exceptions).
+The discrepancy is due to the fact that ecologists have chosen to focus, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (@pollock_understanding_2014; see @faisal_inferring_2010 and @harris_inferring_2015 for two recent exceptions).
 As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (e.g.
 abiotic influences and indirect biotic effects; Figure 1).
-When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015].
+When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_inferring_2015].
 While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still lack a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.
@@
 Rather than describing the overall relationship between two variables across all conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.
-@harris_estimating_2015 found that the partial correlation can do extract accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
+@harris_inferring_2015 found that the partial correlation can do extract accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.
 
-Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @harris_estimating_2015).
+Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @harris_inferring_2015).
 A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network).
@@
 the Markov network has the most information entropy of any possible model satisfying its constraints; @azaele_inferring_2010].
-As shown in @harris_estimating_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.
+As shown in @harris_inferring_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.
 
 Unfortunately, Markov networks have an intractable likelihood function whose computational difficulty more than doubles each time a new species is added to the model.
-This exponential growth in the likelihood function's complexity means that the methods that worked for @azaele_inferring_2010 and @harris_estimating_2015 with networks of 20 species would be completely infeasible for networks with 50 species, requiring over a billion times more computational effort ($2^{50}/2^{20} \approx 10^{9}$).
+This exponential growth in the likelihood function's complexity means that the methods that worked for @azaele_inferring_2010 and @harris_inferring_2015 with networks of 20 species would be completely infeasible for networks with 50 species, requiring over a billion times more computational effort ($2^{50}/2^{20} \approx 10^{9}$).
 Extending the method to account for abiotic variation among sites on the landscape would require repeating these expensive computations independently for every site, increasing the computational burden even further.
@@
 
-As discussed in @azaele_inferring_2010 and @harris_estimating_2015, Markov networks such as the Ising model [@cipra_introduction_1987] can be used to describe community structure as follows.
+As discussed in @azaele_inferring_2010 and @harris_inferring_2015, Markov networks such as the Ising model [@cipra_introduction_1987] can be used to describe community structure as follows.
 Each species is represented by a binary random variable, describing whether it is present (1) or absent (0).
@@
 
-The Markov networks fit in @harris_estimating_2015 treated the $\alpha_i$ terms from Equations 1 and 2 as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.
+The Markov networks fit in @harris_inferring_2015 treated the $\alpha_i$ terms from Equations 1 and 2 as constants, meaning that the models expected each species to have the same conditional occurrence probability in any location with a given set of facilitators and competitors.
 In this paper, I extend the model so that the local abiotic conditions can also affect species occurrence probabilities directly.
@@
 
-The approximate likelihood gradients in Step 2 match the ones from @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
+The approximate likelihood gradients in Step 2 match the ones from @harris_inferring_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations.
 These gradients were augmented with a momentum term [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 2.0 (for environmental responses) or 0.5 (for pairwise interactions).
@@
 These landscapes had few enough species that the `rosalia` package for estimating Markov networks [@harris_rosalia_2015] could find the penalized maximum likelihood estimates for the Markov network in a reasonable amount of time.
-For each of these simulated landscapes, I then fit the same Markov network model to the simulated data using the exact approach from @harris_estimating_2015 and the stochastic approximation method described above.
+For each of these simulated landscapes, I then fit the same Markov network model to the simulated data using the exact approach from @harris_inferring_2015 and the stochastic approximation method described above.
 
@@
 
-For decades, ecologists have relied on poor test statistics for inferring species interactions from observational data [@harris_estimating_2015].
-As shown here and in @harris_estimating_2015, however, these inferences can only be made reliably by methods that control for other species in the network, such as partial correlations or Markov networks.
+For decades, ecologists have relied on poor test statistics for inferring species interactions from observational data [@harris_inferring_2015].
+As shown here and in @harris_inferring_2015, however, these inferences can only be made reliably by methods that control for other species in the network, such as partial correlations or Markov networks.
 The biggest computational problem with large Markov networks is the impossibility of averaging over all their possible states, but these results demonstrate that this step can be avoided in ecological analyses.
@@
 
-The largest downside of using stochastic approximation instead of the exact methods introduced in @harris_estimating_2015 is the difficulty in generating confidence intervals.
+The largest downside of using stochastic approximation instead of the exact methods introduced in @harris_inferring_2015 is the difficulty in generating confidence intervals.
 The `rosalia` package can produce confidence intervals based on the Hessian matrix, but it is not feasible to calculate this matrix for large networks.
@@
 
+# Acknowledgements
+
+This work benefited greatly from discussions with A. Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
+
 # References

### 2016-02-11 stochastic/chapter-3.tex
some chapter 3 work

@@
+\documentclass[12pt,]{article}
+\usepackage[compact]{titlesec}
+\usepackage{fancyhdr}
+\usepackage{hanging}
+\pagestyle{fancy}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage{setspace}
+\usepackage{lineno}
+\linenumbers
+\usepackage{setspace}
+\usepackage{geometry}
+\usepackage{ifthen}
+\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={David J. Harris},
+            pdftitle={Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{4pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
+
+\title{Estimating species interactions in large, abiotically structured
+communities with Markov networks and stochastic approximation}
+\author{David J. Harris}
+\date{}
+
+\begin{document}
+\maketitle
+
+\begin{spacing}{1.9}
+\begin{flushleft}
 \section{Abstract}\label{abstract}
@@
 ecological questions, computational infeasibility, or all three (Connor
-and Simberloff 1979; Gilpin and Diamond 1982; Strong et al. 1984;
+and Simberloff 1979, Gilpin and Diamond 1982, Strong et al. 1984,
 Hastings 1987). As documented below, ecologists have, for the most part,
@@
 solved the analogous problem of estimating interactions among thousands
-of genes from their co-expression levels (e.g. Friedman, Hastie, and
-Tibshirani (2008)). The discrepancy is due to the fact that ecologists
-have chosen to focus, almost exclusively, on overall co-occurrence rates
-(Connor and Simberloff 1979; Gotelli and Ulrich 2009; Gilpin and Diamond
-1982; Veech 2013) or on closely-related values such as correlation
-coefficients (Pollock et al. 2014; see Faisal et al. 2010 and Harris
-2015 for two recent exceptions). As ecologists know, however, the
-correlation between two species will often reflect other factors beyond
-their direct pairwise interactions (e.g.~abiotic influences and indirect
-biotic effects; Figure 1). When these factors are not accounted for,
-this reliance on overall co-occurrence rates can reliably lead to
-incorrect inferences (D. J. Harris 2015). While some methods have been
-proposed for incorporating a small number of specific factors (such as
-geographic or environmental dissimilarity) into ecologists' null models
-at a time (Lessard et al. 2011), we still lack a good way to scale these
+of genes from their co-expression levels (e.g. Friedman et al. (2008)).
+The discrepancy is due to the fact that ecologists have chosen to focus,
+almost exclusively, on overall co-occurrence rates (Connor and
+Simberloff 1979, Gilpin and Diamond 1982, Gotelli and Ulrich 2009, Veech
+2013) or on closely-related values such as correlation coefficients
+(Pollock et al. 2014; see Faisal et al. 2010 and Harris 2015 for two
+recent exceptions). As ecologists know, however, the correlation between
+two species will often reflect other factors beyond their direct
+pairwise interactions (e.g.~abiotic influences and indirect biotic
+effects; Figure 1). When these factors are not accounted for, this
+reliance on overall co-occurrence rates can reliably lead to incorrect
+inferences (Harris 2015a). While some methods have been proposed for
+incorporating a small number of specific factors (such as geographic or
+environmental dissimilarity) into ecologists' null models at a time
+(Lessard et al. 2011), we still lack a good way to scale these
 approaches up for cases with many biotic and abiotic factors acting
@@
 coefficients) describe the portion of the relationship that remains
-after the other variables in the data have been accounted for. D. J.
-Harris (2015) found that the partial correlation can do extract accurate
+after the other variables in the data have been accounted for. Harris
+(2015a) found that the partial correlation can do extract accurate
 information about pairwise species interactions from binary
@@
 fields) to estimate conditional relationships from binary data by
-maximum entropy or maximum likelihood (Azaele et al. (2010); D. J.
-Harris (2015)). A Markov network defines a probability distribution over
+maximum entropy or maximum likelihood (Azaele et al. (2010); Harris
+(2015a)). A Markov network defines a probability distribution over
 possible binary species assemblages, and its coefficients (including one
@@
 has the most information entropy of any possible model satisfying its
-constraints; Azaele et al. 2010). As shown in D. J. Harris (2015),
-Markov networks do a better job of estimating species' influences on one
+constraints; Azaele et al. 2010). As shown in Harris (2015a), Markov
+networks do a better job of estimating species' influences on one
 another better than a number of existing methods, particularly when
@@
 function's complexity means that the methods that worked for Azaele et
-al. (2010) and D. J. Harris (2015) with networks of 20 species would be
+al. (2010) and Harris (2015a) with networks of 20 species would be
 completely infeasible for networks with 50 species, requiring over a
@@
 In this paper, I present a different way of optimizing the likelihood,
-called ``stochastic approximation'' (Robbins and Monro 1951;
+called ``stochastic approximation'' (Robbins and Monro 1951,
 Salakhutdinov and Hinton 2012), which replaces the intractable
@@
 stochastic approximation provides strong guarantees for eventual
-convergence to the maximum likelihood estimate (Younes 1999;
+convergence to the maximum likelihood estimate (Younes 1999,
 Salakhutdinov and Hinton 2012). This change in approach makes it
@@
 
-As discussed in Azaele et al. (2010) and D. J. Harris (2015), Markov
-networks such as the Ising model (Cipra 1987) can be used to describe
-community structure as follows. Each species is represented by a binary
-random variable, describing whether it is present (1) or absent (0). A
-species' conditional probability of presence under a given set of biotic
-and abiotic conditions depends on its coefficients. The coefficients
-linking species to one another describe their conditional associations:
-all else equal, a species will occur less often in the presence of its
+As discussed in Azaele et al. (2010) and Harris (2015a), Markov networks
+such as the Ising model (Cipra 1987) can be used to describe community
+structure as follows. Each species is represented by a binary random
+variable, describing whether it is present (1) or absent (0). A species'
+conditional probability of presence under a given set of biotic and
+abiotic conditions depends on its coefficients. The coefficients linking
+species to one another describe their conditional associations: all else
+equal, a species will occur less often in the presence of its
 competitors and other species with which it is negatively associated,
@@
 any given combination of presences and absences, up to a multiplicative
-constant (Lee and Hastie 2012; Murphy 2012):
+constant (Lee and Hastie 2012, Murphy 2012):
 
@@
 
-The Markov networks fit in D. J. Harris (2015) treated the \(\alpha_i\)
-terms from Equations 1 and 2 as constants, meaning that the models
-expected each species to have the same conditional occurrence
-probability in any location with a given set of facilitators and
-competitors. In this paper, I extend the model so that the local abiotic
-conditions can also affect species occurrence probabilities directly.
-Specifically, the \(\alpha_i\) terms in these analyses are linear
-combinations of the local environmental conditions, represented by
-\(x_1\) through \(x_5\). By estimating the coefficients associated with
-each of these \(x\) variables for each species, we can account for their
-responses to the abiotic environment as well as to one another. In other
-fields, a model like this one, where a Markov network's parameters
-depend on external factors is called a conditional random field (Lee and
-Hastie 2012); these models have been used in a variety of contexts, from
-language models to image processing (Murphy 2012).
+The Markov networks fit in Harris (2015a) treated the \(\alpha_i\) terms
+from Equations 1 and 2 as constants, meaning that the models expected
+each species to have the same conditional occurrence probability in any
+location with a given set of facilitators and competitors. In this
+paper, I extend the model so that the local abiotic conditions can also
+affect species occurrence probabilities directly. Specifically, the
+\(\alpha_i\) terms in these analyses are linear combinations of the
+local environmental conditions, represented by \(x_1\) through \(x_5\).
+By estimating the coefficients associated with each of these \(x\)
+variables for each species, we can account for their responses to the
+abiotic environment as well as to one another. In other fields, a model
+like this one, where a Markov network's parameters depend on external
+factors is called a conditional random field (Lee and Hastie 2012);
+these models have been used in a variety of contexts, from language
+models to image processing (Murphy 2012).
 
@@
 co-occurrences between each species, and the cross-product between the
-species and environment matrices (Azaele et al. 2010; Lee and Hastie
-2012; Murphy 2012). Finding the maximum likelihood estimate for the
+species and environment matrices (Azaele et al. 2010, Lee and Hastie
+2012, Murphy 2012). Finding the maximum likelihood estimate for the
 model parameters is equivalent to minimizing the discrepancy between the
@@
 exactly, averaging over all possible presence-absence combinations.
-Stochastic approximation (Robbins and Monro 1951; Salakhutdinov and
+Stochastic approximation (Robbins and Monro 1951, Salakhutdinov and
 Hinton 2012) instead estimates the expected values of the sufficient
@@
 
-1: simulate a set of assemblages from the current model parameters and
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  Simulate a set of assemblages from the current model parameters and
   calculate sufficient statistics for the sample.
-
-2: subtract the simulated sufficient statistics from the observed ones
-to calculate the approximate likelihood gradient
-
-3: Adjust the model parameters to climb the approximate gradient, using
-a schedule of step sizes that satisfies the criteria in Chapter 6 of
+\item
+  Subtract the simulated sufficient statistics from the observed ones to
+  calculate the approximate likelihood gradient
+\item
+  Adjust the model parameters to climb the approximate gradient, using a
+  schedule of step sizes that satisfies the criteria in Chapter 6 of
   Powell (2007).
+\end{enumerate}
 
@@
 convergence to the maximum likelihood estimate rather than preventing it
-altogether (Younes 1999; Salakhutdinov and Hinton 2012).
+altogether (Younes 1999, Salakhutdinov and Hinton 2012).
 
-The approximate likelihood gradients in Step 2 match the ones from D. J.
-Harris (2015), except that they are averaged over a set of Monte Carlo
+The approximate likelihood gradients in Step 2 match the ones from
+Harris (2015a), except that they are averaged over a set of Monte Carlo
 samples rather than over all possible presence-absence combinations.
@@
 regularizers based on a logistic prior with location 0 and scale 2.0
-(for environmental responses) or 0.5 (for pairwise interactions).
-
-The step size parameter in Step 3, \(\alpha_t\), decreased after each
+(for environmental responses) or 0.5 (for pairwise interactions). The
+step size parameter in Step 3, \(\alpha_t\), decreased after each
 iteration according to a generalized harmonic sequence,
-\(\alpha_t = \alpha_0 1000/(999 + t)\), which satisfies the criteria in
-Powell (2007). The scaling factor, \(\alpha_0\), was 1.0 for species'
-intercepts and environmental responses. Because of the enormous number
-of pairwise interaction coefficients, I set \(\alpha_0\) to 0.1 for the
-\(\beta\) parameters.
+\(\alpha_t = 1000/(999 + t)\), which satisfies the criteria in Powell
+(2007).
 
@@
 enough species that the \texttt{rosalia} package for estimating Markov
-networks (David J. Harris 2015) could find the penalized maximum
-likelihood estimates for the Markov network in a reasonable amount of
-time. For each of these simulated landscapes, I then fit the same Markov
-network model to the simulated data using the exact approach from D. J.
-Harris (2015) and the stochastic approximation method described above.
+networks (Harris 2015b) could find the penalized maximum likelihood
+estimates for the Markov network in a reasonable amount of time. For
+each of these simulated landscapes, I then fit the same Markov network
+model to the simulated data using the exact approach from Harris (2015a)
+and the stochastic approximation method described above.
 
@@
 For decades, ecologists have relied on poor test statistics for
-inferring species interactions from observational data (D. J. Harris
-2015). As shown here and in D. J. Harris (2015), however, these
-inferences can only be made reliably by methods that control for other
-species in the network, such as partial correlations or Markov networks.
-The biggest computational problem with large Markov networks is the
-impossibility of averaging over all their possible states, but these
-results demonstrate that this step can be avoided in ecological
-analyses. Stochastic approximation is able to recover the same estimates
-as exact methods for smaller networks, while scaling gracefully to
-networks with hundreds of species.
+inferring species interactions from observational data (Harris 2015a).
+As shown here and in Harris (2015a), however, these inferences can only
+be made reliably by methods that control for other species in the
+network, such as partial correlations or Markov networks. The biggest
+computational problem with large Markov networks is the impossibility of
+averaging over all their possible states, but these results demonstrate
+that this step can be avoided in ecological analyses. Stochastic
+approximation is able to recover the same estimates as exact methods for
+smaller networks, while scaling gracefully to networks with hundreds of
+species.
 
 The largest downside of using stochastic approximation instead of the
-exact methods introduced in D. J. Harris (2015) is the difficulty in
+exact methods introduced in Harris (2015a) is the difficulty in
 generating confidence intervals. The \texttt{rosalia} package can
@@
 Alternatively, ecologists may be able to take advantage of the advanced
-Monte Carlo techniques introduced in Murray, Ghahramani, and MacKay
-(2012) to sample from the posterior distribution over possible parameter
-values, which could be used to generate credible intervals for the
-parameter estimates. Other advanced Monte Carlo methods (e.g.
-Salakhutdinov 2009) may also speed up convergence to the maximum
-likelihood estimate in cases where serial autocorrelation in the Gibbs
-sampler is too strong for effectively sampling the space of possible
-landscapes.
+Monte Carlo techniques introduced in Murray et al. (2012) to sample from
+the posterior distribution over possible parameter values, which could
+be used to generate credible intervals for the parameter estimates.
+Other advanced Monte Carlo methods (e.g. Salakhutdinov 2009) may also
+speed up convergence to the maximum likelihood estimate in cases where
+serial autocorrelation in the Gibbs sampler is too strong for
+effectively sampling the space of possible landscapes.
 
@@
 presence can affect the relationship between two other species (Whittam
-and Siegel-Causey 1981; Tjelmeland and Besag 1998).
+and Siegel-Causey 1981, Tjelmeland and Besag 1998).
 
@@
 
-Azaele, S, R Muneepeerakul, A Rinaldo, and I Rodriguez-Iturbe. 2010.
-``Inferring Plant Ecosystem Organization from Species Occurrences.''
-\emph{Journal of Theoretical Biology} 262 (2): 323--29.
+\hyperdef{}{refs}{\label{refs}}
+\hyperdef{}{ref-azaeleux5finferringux5f2010}{\label{ref-azaeleux5finferringux5f2010}}
+Azaele, S., R. Muneepeerakul, A. Rinaldo, and I. Rodriguez-Iturbe. 2010.
+Inferring plant ecosystem organization from species occurrences. Journal
+of theoretical biology 262:323--329.
 
-Bickel, PJ, and K Doksum. 1977. \emph{Mathematical Statistics: Basic
-Ideas and Selected Topics}. San Francisco: Holden---Day.
+\hyperdef{}{ref-bickelux5fmathematicalux5f1977}{\label{ref-bickelux5fmathematicalux5f1977}}
+Bickel, P., and K. Doksum. 1977. Mathematical Statistics: Basic Ideas
+and Selected Topics. San Francisco: Holden---Day.
 
-Cipra, Barry A. 1987. ``An Introduction to the Ising Model.''
-\emph{American Mathematical Monthly} 94 (10): 937--59.
+\hyperdef{}{ref-cipraux5fintroductionux5f1987}{\label{ref-cipraux5fintroductionux5f1987}}
+Cipra, B. A. 1987. An introduction to the Ising model. American
+Mathematical Monthly 94:937--959.
 
-Connor, Edward F, and Daniel Simberloff. 1979. ``The Assembly of Species
-Communities: Chance or Competition?'' \emph{Ecology}, 1132--40.
+\hyperdef{}{ref-connorux5fassemblyux5f1979}{\label{ref-connorux5fassemblyux5f1979}}
+Connor, E. F., and D. Simberloff. 1979. The assembly of species
+communities: Chance or competition? Ecology:1132--1140.
 
-Faisal, Ali, Frank Dondelinger, Dirk Husmeier, and Colin M. Beale. 2010.
-``Inferring Species Interaction Networks from Species Abundance Data: A
-Comparative Evaluation of Various Statistical and Machine Learning
-Methods.'' \emph{Ecological Informatics} 5 (6): 451--64.
-doi:\href{http://dx.doi.org/http://dx.doi.org/10.1016/j.ecoinf.2010.06.005}{http://dx.doi.org/10.1016/j.ecoinf.2010.06.005}.
+\hyperdef{}{ref-faisalux5finferringux5f2010}{\label{ref-faisalux5finferringux5f2010}}
+Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
+Inferring species interaction networks from species abundance data: A
+comparative evaluation of various statistical and machine learning
+methods. Ecological Informatics 5:451--464.
 
-Friedman, Jerome, Trevor Hastie, and Robert Tibshirani. 2008. ``Sparse
-Inverse Covariance Estimation with the Graphical Lasso.''
-\emph{Biostatistics} 9 (3): 432--41.
-doi:\href{http://dx.doi.org/10.1093/biostatistics/kxm045}{10.1093/biostatistics/kxm045}.
+\hyperdef{}{ref-friedmanux5fsparseux5f2008}{\label{ref-friedmanux5fsparseux5f2008}}
+Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
+covariance estimation with the graphical lasso. Biostatistics
+9:432--441.
 
-Gilpin, Michael E., and Jared M. Diamond. 1982. ``Factors Contributing
-to Non-Randomness in Species Co-Occurrences on Islands.''
-\emph{Oecologia} 52 (1): 75--84.
-doi:\href{http://dx.doi.org/10.1007/BF00349014}{10.1007/BF00349014}.
+\hyperdef{}{ref-gilpinux5ffactorsux5f1982}{\label{ref-gilpinux5ffactorsux5f1982}}
+Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
+non-randomness in species Co-occurrences on Islands. Oecologia
+52:75--84.
 
-Gotelli, Nicholas J., and Werner Ulrich. 2009. ``The Empirical Bayes
-Approach as a Tool to Identify Non-Random Species Associations.''
-\emph{Oecologia} 162 (2): 463--77.
-doi:\href{http://dx.doi.org/10.1007/s00442-009-1474-y}{10.1007/s00442-009-1474-y}.
+\hyperdef{}{ref-gotelliux5fempiricalux5f2009}{\label{ref-gotelliux5fempiricalux5f2009}}
+Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
+tool to identify non-random species associations. Oecologia
+162:463--477.
 
-Harris, D. J. 2015. ``Estimating Species Interactions from Observational
-Data with Markov Networks.'' \emph{BioRxiv}.
-doi:\href{http://dx.doi.org/10.1101/018861}{10.1101/018861}.
+\hyperdef{}{ref-harrisux5festimatingux5f2015}{\label{ref-harrisux5festimatingux5f2015}}
+Harris, D. J. 2015a. Estimating species interactions from observational
+data with Markov networks. bioRxiv.
 
-Harris, David J. 2015. \emph{Rosalia: Exact Inference for Small Binary
-Markov Networks. R Package Version 0.1.0}. Zenodo.
+\hyperdef{}{ref-harrisux5frosaliaux5f2015}{\label{ref-harrisux5frosaliaux5f2015}}
+Harris, D. J. 2015b. Rosalia: Exact inference for small binary Markov
+networks. R package version 0.1.0. Zenodo.
 http://dx.doi.org/10.5281/zenodo.17808.
-\url{http://dx.doi.org/10.5281/zenodo.17808}.
 
-Hastings, Alan. 1987. ``Can Competition Be Detected Using Species
-Co-Occurrence Data?'' \emph{Ecology}, 117--23.
+\hyperdef{}{ref-hastingsux5fcanux5f1987}{\label{ref-hastingsux5fcanux5f1987}}
+Hastings, A. 1987. Can competition be detected using species
+co-occurrence data? Ecology:117--123.
 
-Hinton, Geoffrey E. 2012. ``A Practical Guide to Training Restricted
-Boltzmann Machines.'' In \emph{Neural Networks: Tricks of the Trade},
-599--619. Springer.
+\hyperdef{}{ref-hintonux5fpracticalux5f2012}{\label{ref-hintonux5fpracticalux5f2012}}
+Hinton, G. E. 2012. A practical guide to training restricted boltzmann
+machines. Pages 599--619 \emph{in} Neural Networks: Tricks of the Trade.
+Springer.
 
-Kissling, W. D., Carsten F. Dormann, Jürgen Groeneveld, Thomas Hickler,
-Ingolf Kühn, Greg J. McInerny, José M. Montoya, et al. 2012. ``Towards
-Novel Approaches to Modelling Biotic Interactions in Multispecies
-Assemblages at Large Spatial Extents.'' \emph{Journal of Biogeography}
-39 (12): 2163--78.
-doi:\href{http://dx.doi.org/10.1111/j.1365-2699.2011.02663.x}{10.1111/j.1365-2699.2011.02663.x}.
+\hyperdef{}{ref-kisslingux5ftowardsux5f2012}{\label{ref-kisslingux5ftowardsux5f2012}}
+Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
+J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
+Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
+Towards novel approaches to modelling biotic interactions in
+multispecies assemblages at large spatial extents. Journal of
+Biogeography 39:2163--2178.
 
-Lee, Jason D., and Trevor J. Hastie. 2012. ``Learning Mixed Graphical
-Models.'' \emph{CERN Document Server}.
-\url{http://cds.cern.ch/record/1451206}.
+\hyperdef{}{ref-leeux5flearningux5f2012}{\label{ref-leeux5flearningux5f2012}}
+Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
+Models.
 
-Lessard, Jean-Philippe, Michael K. Borregaard, James A. Fordyce, Carsten
-Rahbek, Michael D. Weiser, Robert R. Dunn, and Nathan J. Sanders. 2011.
-``Strong Influence of Regional Species Pools on Continent-Wide
-Structuring of Local Communities.'' \emph{Proceedings of the Royal
-Society of London B: Biological Sciences}.
-doi:\href{http://dx.doi.org/10.1098/rspb.2011.0552}{10.1098/rspb.2011.0552}.
+\hyperdef{}{ref-lessardux5fstrongux5f2011}{\label{ref-lessardux5fstrongux5f2011}}
+Lessard, J.-P., M. K. Borregaard, J. A. Fordyce, C. Rahbek, M. D.
+Weiser, R. R. Dunn, and N. J. Sanders. 2011. Strong influence of
+regional species pools on continent-wide structuring of local
+communities. Proceedings of the Royal Society of London B: Biological
+Sciences.
 
-Lewin, Roger. 1983. ``Santa Rosalia Was a Goat.'' \emph{Science} 221
-(4611): 636--39.
+\hyperdef{}{ref-lewinux5fsantaux5f1983}{\label{ref-lewinux5fsantaux5f1983}}
+Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.
 
-Murphy, Kevin P. 2012. \emph{Machine Learning: A Probabilistic
-Perspective}. The MIT Press.
+\hyperdef{}{ref-murphyux5fmachineux5f2012}{\label{ref-murphyux5fmachineux5f2012}}
+Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
+MIT Press.
 
-Murray, Iain, Zoubin Ghahramani, and David MacKay. 2012. ``MCMC for
-Doubly-Intractable Distributions.'' \emph{ArXiv Preprint
-ArXiv:1206.6848}.
+\hyperdef{}{ref-murrayux5fmcmcux5f2012}{\label{ref-murrayux5fmcmcux5f2012}}
+Murray, I., Z. Ghahramani, and D. MacKay. 2012. MCMC for
+doubly-intractable distributions. arXiv preprint arXiv:1206.6848.
 
-Pollock, Laura J., Reid Tingley, William K. Morris, Nick Golding, Robert
-B. O'Hara, Kirsten M. Parris, Peter A. Vesk, and Michael A. McCarthy.
-2014. ``Understanding Co-Occurrence by Modelling Species Simultaneously
-with a Joint Species Distribution Model (JSDM).'' \emph{Methods in
-Ecology and Evolution}, n/a--/a.
-doi:\href{http://dx.doi.org/10.1111/2041-210X.12180}{10.1111/2041-210X.12180}.
+\hyperdef{}{ref-pollockux5funderstandingux5f2014}{\label{ref-pollockux5funderstandingux5f2014}}
+Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K.
+M. Parris, P. A. Vesk, and M. A. McCarthy. 2014. Understanding
+co-occurrence by modelling species simultaneously with a Joint Species
+Distribution Model (JSDM). Methods in Ecology and Evolution:n/a--n/a.
 
-Powell, Warren B. 2007. \emph{Approximate Dynamic Programming: Solving
-the Curses of Dimensionality}. John Wiley \& Sons.
+\hyperdef{}{ref-powellux5fapproximateux5f2007}{\label{ref-powellux5fapproximateux5f2007}}
+Powell, W. B. 2007. Approximate Dynamic Programming: Solving the curses
+of dimensionality. John Wiley \& Sons.
 
-Robbins, Herbert, and Sutton Monro. 1951. ``A Stochastic Approximation
-Method.'' \emph{The Annals of Mathematical Statistics} 22 (3): 400--407.
-doi:\href{http://dx.doi.org/10.1214/aoms/1177729586}{10.1214/aoms/1177729586}.
+\hyperdef{}{ref-robbinsux5fstochasticux5f1951}{\label{ref-robbinsux5fstochasticux5f1951}}
+Robbins, H., and S. Monro. 1951. A Stochastic Approximation Method. The
+Annals of Mathematical Statistics 22:400--407.
 
-Salakhutdinov, Ruslan. 2009. ``Learning in Markov Random Fields Using
-Tempered Transitions.'' \emph{Advances in Neural Information Processing
-Systems} 22: 1598--1606.
-\url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.163.9244\&rep=rep1\&type=pdf}.
+\hyperdef{}{ref-salakhutdinovux5flearningux5f2009}{\label{ref-salakhutdinovux5flearningux5f2009}}
+Salakhutdinov, R. 2009. Learning in Markov random fields using tempered
+transitions. Advances in neural information processing systems
+22:1598--1606.
 
-Salakhutdinov, Ruslan, and Geoffrey Hinton. 2012. ``An Efficient
-Learning Procedure for Deep Boltzmann Machines.'' \emph{Neural
-Computation} 24 (8): 1967--2006.
-\url{http://www.mitpressjournals.org/doi/pdf/10.1162/NECO_a_00311}.
+\hyperdef{}{ref-salakhutdinovux5fefficientux5f2012}{\label{ref-salakhutdinovux5fefficientux5f2012}}
+Salakhutdinov, R., and G. Hinton. 2012. An efficient learning procedure
+for deep Boltzmann machines. Neural Computation 24:1967--2006.
 
-Schmidt, Mark, and Kevin Murphy. 2012. ``Modeling Discrete
-Interventional Data Using Directed Cyclic Graphical Models.''
-\emph{ArXiv Preprint ArXiv:1205.2617}.
+\hyperdef{}{ref-schmidtux5fmodelingux5f2012}{\label{ref-schmidtux5fmodelingux5f2012}}
+Schmidt, M., and K. Murphy. 2012. Modeling Discrete Interventional Data
+using Directed Cyclic Graphical Models. arXiv preprint arXiv:1205.2617.
 
-Strong, Donald R, Daniel Simberloff, Lawrence G Abele, and Anne B
-Thistle. 1984. \emph{Ecological Communities: Conceptual Issues and the
-Evidence}. Princeton University Press.
+\hyperdef{}{ref-strongux5fecologicalux5f1984}{\label{ref-strongux5fecologicalux5f1984}}
+Strong, D. R., D. Simberloff, L. G. Abele, and A. B. Thistle. 1984.
+Ecological communities: Conceptual issues and the evidence. Princeton
+University Press.
 
-Tjelmeland, Håkon, and Julian Besag. 1998. ``Markov Random Fields with
-Higher-Order Interactions.'' \emph{Scandinavian Journal of Statistics}
-25 (3): 415--33.
-doi:\href{http://dx.doi.org/10.1111/1467-9469.00113}{10.1111/1467-9469.00113}.
+\hyperdef{}{ref-tjelmelandux5fmarkovux5f1998}{\label{ref-tjelmelandux5fmarkovux5f1998}}
+Tjelmeland, H., and J. Besag. 1998. Markov Random Fields with
+Higher-order Interactions. Scandinavian Journal of Statistics
+25:415--433.
 
-Veech, Joseph A. 2013. ``A Probabilistic Model for Analysing Species
-Co-Occurrence.'' \emph{Global Ecology and Biogeography} 22 (2): 252--60.
-doi:\href{http://dx.doi.org/10.1111/j.1466-8238.2012.00789.x}{10.1111/j.1466-8238.2012.00789.x}.
+\hyperdef{}{ref-veechux5fprobabilisticux5f2013}{\label{ref-veechux5fprobabilisticux5f2013}}
+Veech, J. A. 2013. A probabilistic model for analysing species
+co-occurrence. Global Ecology and Biogeography 22:252--260.
 
-Whittam, Thomas S., and Douglas Siegel-Causey. 1981. ``Species
-Interactions and Community Structure in Alaskan Seabird Colonies.''
-\emph{Ecology} 62 (6): 1515--24.
-doi:\href{http://dx.doi.org/10.2307/1941508}{10.2307/1941508}.
+\hyperdef{}{ref-whittamux5fspeciesux5f1981}{\label{ref-whittamux5fspeciesux5f1981}}
+Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
+Community Structure in Alaskan Seabird Colonies. Ecology 62:1515--1524.
 
-Younes, Laurent. 1999. ``On the Convergence of Markovian Stochastic
-Algorithms with Rapidly Decreasing Ergodicity Rates.''
-\emph{Stochastics: An International Journal of Probability and
-Stochastic Processes} 65 (3-4): 177--228.
+\hyperdef{}{ref-younesux5fconvergenceux5f1999}{\label{ref-younesux5fconvergenceux5f1999}}
+Younes, L. 1999. On the convergence of Markovian stochastic algorithms
+with rapidly decreasing ergodicity rates. Stochastics: An International
+Journal of Probability and Stochastic Processes 65:177--228.
+\end{flushleft}
+\end{spacing}
+
+\end{document}

### 2016-02-11 stochastic/stochastic-recovery.Rmd
some chapter 3 work

@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 library(rosalia)
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 set.seed(1)
@@
 
-```{r,eval=FALSE, cache = TRUE}
+```{r,, cache = TRUE}
 # Initialize the `y` matrix.
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 # Calculate sufficient statistics of the data
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 # Initialize the simulated landscape for stochastic approximation
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 # Very weak priors on alpha terms, somewhat stronger on beta terms
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 initial_learning_rate = 1 # step size at start of optimization
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 library(cowplot)
@@
     coord_equal(),
+
   ggplot(NULL, aes(x = c(alpha_env), y = c(true_alpha_env))) + 
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 small_n_spp = 20
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 exact_time = system.time({
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 maxit_small = 50000
@@
 
-```{r,eval=FALSE}
+```{r,cache = TRUE}
 pdf("stochastic/convergence.pdf", width = 10, height = 7)
@@
 plot(small_times, mses, type = "l", 
-     ylab = "mean square deviation from MLE\n(log scale)", 
+     ylab = "mean square deviation from MLE", 
      xlab = "time (seconds)", 

### 2016-02-11 Harris_2015/manuscript-materials/draft.Rmd
chapter 2 wording

@@
 expected prevalence of a species in the absence of biotic influences as
-$e^\alpha/(e^{0} + e^\alpha)$. Competition's effect on prevalence can then be
+$e^{\alpha_i}/(e^{0} + e^{\alpha_i})$. Competition's effect on prevalence can then be
 estimated by comparing this value with the observed prevalence (e.g. comparing
-Figure 2D with Figure 2C). This novel quantitative result conflicts with most of
+Figure 2D with Figure 2C). This novel quantitative result undermines most of
 our null models, which unreasonably assume that prevalence would be the exactly
@@
 example, the states of the interaction network can be modeled as a function of
-the local abiotic environment [@lee_learning_2012], which would help incorporate
-networks of biotic interactions into species distribution models
-[@pollock_understanding_2014] and lead to a better understanding of the
-interplay between biotic and abiotic effects on community structure.
-Alternatively, models could allow one species to alter the relationship between
-two other species [@tjelmeland_markov_1998; cf @bruno_inclusion_2003].
+the local abiotic environment [@lee_learning_2012; @harris_estimating_2016], 
+which would lead to a better understanding of the interplay between biotic and 
+abiotic effects on community structure. Alternatively, models could allow one 
+species to alter the relationship between two other species 
+[@tjelmeland_markov_1998; @whittam_species_1981; cf @bruno_inclusion_2003].
 

### 2016-02-11 Harris_2015/manuscript-materials/draft.tex
chapter 2 wording

@@
 prevalence of a species in the absence of biotic influences as
-\(e^\alpha/(e^{0} + e^\alpha)\). Competition's effect on prevalence can
-then be estimated by comparing this value with the observed prevalence
-(e.g.~comparing Figure 2D with Figure 2C). This novel quantitative
-result conflicts with most of our null models, which unreasonably assume
-that prevalence would be the exactly same in the absence of competition
-as it is in the observed data (Roughgarden 1983).
+\(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
+prevalence can then be estimated by comparing this value with the
+observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
+novel quantitative result conflicts with most of our null models, which
+unreasonably assume that prevalence would be the exactly same in the
+absence of competition as it is in the observed data (Roughgarden 1983).
 
@@
 example, the states of the interaction network can be modeled as a
-function of the local abiotic environment (Lee and Hastie 2012), which
-would help incorporate networks of biotic interactions into species
-distribution models (Pollock et al. 2014) and lead to a better
-understanding of the interplay between biotic and abiotic effects on
-community structure. Alternatively, models could allow one species to
-alter the relationship between two other species (Tjelmeland and Besag
-1998, cf Bruno et al. 2003).
+function of the local abiotic environment (Lee and Hastie 2012,
+{\textbf{???}}), which would lead to a better understanding of the
+interplay between biotic and abiotic effects on community structure.
+Alternatively, models could allow one species to alter the relationship
+between two other species (Whittam and Siegel-Causey 1981, Tjelmeland
+and Besag 1998, cf Bruno et al. 2003).
 
@@
 \hyperdef{}{ref-cipraux5fintroductionux5f1987}{\label{ref-cipraux5fintroductionux5f1987}}
-Cipra, B. A. 1987. An introduction to the Ising model. American
+Cipra, B. A. 1987. An introduction to the ising model. American
 Mathematical Monthly 94:937--959.
@@
 \hyperdef{}{ref-gelmanux5fweaklyux5f2008}{\label{ref-gelmanux5fweaklyux5f2008}}
-Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
-Informative Default Prior Distribution for Logistic and Other Regression
-Models. The Annals of Applied Statistics 2:1360--1383.
+Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A weakly
+informative default prior distribution for logistic and other regression
+models. The Annals of Applied Statistics 2:1360--1383.
 
@@
 Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
-non-randomness in species Co-occurrences on Islands. Oecologia
+non-randomness in species co-occurrences on islands. Oecologia
 52:75--84.
@@
 \hyperdef{}{ref-goldingux5fbayescommux5f2015}{\label{ref-goldingux5fbayescommux5f2015}}
-Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian Community
-Ecology Analysis.
+Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian community
+ecology analysis.
 
 \hyperdef{}{ref-gotelliux5fempiricalux5f2009}{\label{ref-gotelliux5fempiricalux5f2009}}
-Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
+Gotelli, N. J., and W. Ulrich. 2009. The empirical bayes approach as a
 tool to identify non-random species associations. Oecologia
@@
 \hyperdef{}{ref-harrisux5frosaliaux5f2015}{\label{ref-harrisux5frosaliaux5f2015}}
-Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
-networks. R package version 0.1.0. Zenodo.
+Harris, D. J. 2015a. Rosalia: Exact inference for small binary markov
+networks. r package version 0.1.0. Zenodo.
 http://dx.doi.org/10.5281/zenodo.17808.
@@
 \hyperdef{}{ref-harrisux5fgeneratingux5f2015}{\label{ref-harrisux5fgeneratingux5f2015}}
-Harris, D. J. 2015b. Generating realistic assemblages with a Joint
-Species Distribution Model. Methods in Ecology and Evolution.
+Harris, D. J. 2015b. Generating realistic assemblages with a joint
+species distribution model. Methods in Ecology and Evolution.
 
@@
 \hyperdef{}{ref-harrisux5fcontactux5f1974}{\label{ref-harrisux5fcontactux5f1974}}
-Harris, T. E. 1974. Contact Interactions on a Lattice. The Annals of
+Harris, T. E. 1974. Contact interactions on a lattice. The Annals of
 Probability 2:969--988.
@@
 \hyperdef{}{ref-leeux5flearningux5f2012}{\label{ref-leeux5flearningux5f2012}}
-Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
-Models.
+Lee, J. D., and T. J. Hastie. 2012. Learning mixed graphical models.
 
 \hyperdef{}{ref-levineux5fcompetitiveux5f1976}{\label{ref-levineux5fcompetitiveux5f1976}}
-Levine, S. H. 1976. Competitive Interactions in Ecosystems. The American
+Levine, S. H. 1976. Competitive interactions in ecosystems. The American
 Naturalist 110:903--910.
@@
 \hyperdef{}{ref-lewinux5fsantaux5f1983}{\label{ref-lewinux5fsantaux5f1983}}
-Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.
+Lewin, R. 1983. Santa rosalia was a goat. Science 221:636--639.
 
@@
 \hyperdef{}{ref-murphyux5fmachineux5f2012}{\label{ref-murphyux5fmachineux5f2012}}
-Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
+Murphy, K. P. 2012. Machine learning: A probabilistic perspective. The
 MIT Press.
@@
 M. Parris, P. A. Vesk, and M. A. McCarthy. 2014. Understanding
-co-occurrence by modelling species simultaneously with a Joint Species
-Distribution Model (JSDM). Methods in Ecology and Evolution:n/a--n/a.
+co-occurrence by modelling species simultaneously with a joint species
+distribution model (JSDM). Methods in Ecology and Evolution:n/a--n/a.
 
 \hyperdef{}{ref-rux5fcoreux5fteamux5frux5f2015}{\label{ref-rux5fcoreux5fteamux5frux5f2015}}
-R Core Team. 2015. R: A Language and Environment for Statistical
-Computing. R Foundation for Statistical Computing, Vienna, Austria.
+R Core Team. 2015. R: A language and environment for statistical
+computing. R Foundation for Statistical Computing, Vienna, Austria.
 
 \hyperdef{}{ref-roughgardenux5fcompetitionux5f1983}{\label{ref-roughgardenux5fcompetitionux5f1983}}
-Roughgarden, J. 1983. Competition and Theory in Community Ecology. The
+Roughgarden, J. 1983. Competition and theory in community ecology. The
 American Naturalist 122:583--601.
@@
 Schäfer, J., R. Opgen-Rhein, V. Zuber, M. Ahdesmäki, A. P. D. Silva, and
-K. Strimmer. 2014. Corpcor: Efficient Estimation of Covariance and
-(Partial) Correlation.
+K. Strimmer. 2014. Corpcor: Efficient estimation of covariance and
+(partial) correlation.
 
 \hyperdef{}{ref-schmidtux5fmodelingux5f2012}{\label{ref-schmidtux5fmodelingux5f2012}}
-Schmidt, M., and K. Murphy. 2012. Modeling Discrete Interventional Data
-using Directed Cyclic Graphical Models. arXiv preprint arXiv:1205.2617.
+Schmidt, M., and K. Murphy. 2012. Modeling discrete interventional data
+using directed cyclic graphical models. arXiv preprint arXiv:1205.2617.
 
@@
 \hyperdef{}{ref-tjelmelandux5fmarkovux5f1998}{\label{ref-tjelmelandux5fmarkovux5f1998}}
-Tjelmeland, H., and J. Besag. 1998. Markov Random Fields with
-Higher-order Interactions. Scandinavian Journal of Statistics
+Tjelmeland, H., and J. Besag. 1998. Markov random fields with
+higher-order interactions. Scandinavian Journal of Statistics
 25:415--433.
@@
 \hyperdef{}{ref-whittamux5fspeciesux5f1981}{\label{ref-whittamux5fspeciesux5f1981}}
-Whittam, T. S., and D. Siegel-Causey. 1981. Species Interactions and
-Community Structure in Alaskan Seabird Colonies. Ecology 62:1515--1524.
+Whittam, T. S., and D. Siegel-Causey. 1981. Species interactions and
+community structure in alaskan seabird colonies. Ecology 62:1515--1524.
 
@@
 Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
-Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
+maximum entropy theory of ecology. The American Naturalist 185:E70--E80.
 \end{flushleft}

### 2016-02-12 Harris_2015/manuscript-materials/draft.Rmd
clarify optimization

@@
 $\beta_{ij}$ is the amount that the co-occurrence of species $i$ and species $j$
-contributes to the log-probability; it controls the conditional relationship
+contributes to the log-probability; it determines the conditional relationship
 between two species, i.e. the probability that they will be found together,
@@
 Equation 1 can be normalized, one can find exact maximum likelihood estimates for
-$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like the 
-ones considered here have unimodal likelihood surfaces [@murphy_machine_2012], so 
-these estimates will represent the global optimum. I used the rosalia package 
-[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to 
-calculate $p(\vec{y}; \alpha, \beta)$ and its gradients [@murphy_machine_2012]; 
-the package passes these functions to the "BFGS" method in R's general-purpose 
-optimizer, which then estimates the Markov network parameters.
+$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like 
+the ones considered here have unimodal likelihood surfaces 
+[@murphy_machine_2012], so these estimates will represent the global optimum.
+
+When the number of species is larger than about 30, noralizing Equation 1 
+can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
+they want to fit a Markov network model. For the analyses presented here, I used 
+the rosalia package [@harris_rosalia_2015] for the R programming language
+[@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its gradients
+exactly [@murphy_machine_2012]; the package passes these functions to the "BFGS" 
+method in R's general-purpose optimizer, which finds values for $\alpha$ and 
+$\beta$.
 
@@
 that the two shrub species competed with one another, despite their frequent
-co-occurrence.
+co-occurrence in unshaded areas.
 

### 2016-02-12 Harris_2015/manuscript-materials/draft.tex
clarify optimization

@@
 networks like the ones considered here have unimodal likelihood surfaces
-(Murphy 2012), so these estimates will represent the global optimum. I
-used the rosalia package (Harris 2015a) for the R programming language
-(R Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
-gradients (Murphy 2012); the package passes these functions to the
-``BFGS'' method in R's general-purpose optimizer, which then estimates
-the Markov network parameters.
+(Murphy 2012), so these estimates will represent the global optimum.
+
+When the number of species is larger than about 30, noralizing Equation
+1 can become intractable, and researchers will either need to
+approximate it (Lee and Hastie 2012) or approximate its gradient
+({\textbf{???}}). Such approximations are beyond the scope of this
+paper, however. For the analyses presented here, I used the rosalia
+package (Harris 2015a) for the R programming language (R Core Team 2015)
+to calculate \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly
+(Murphy 2012); the package passes these functions to the ``BFGS'' method
+in R's general-purpose optimizer, which finds values for \(\alpha\) and
+\(\beta\).
 
@@
 ability to correctly infer that the two shrub species competed with one
-another, despite their frequent co-occurrence.
+another, despite their frequent co-occurrence in unshaded areas.
 
@@
 observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
-novel quantitative result conflicts with most of our null models, which
+novel quantitative result undermines most of our null models, which
 unreasonably assume that prevalence would be the exactly same in the

### 2016-02-18 Harris_2015/manuscript-materials/draft.Rmd
many changes

@@
 surprising indirect consequences. For example, the negative correlation between
-two competing species can be reversed in the presence of a third species that is
-capable of outcompeting both of them. Here, I apply models from statistical
-physics, called Markov networks or Markov random fields, that can predict
+two competing species can be reversed in the presence of a third species that 
+outcompetes both of them. Here, I apply models from statistical physics, called 
+Markov networks or Markov random fields, that can predict
 the direct and indirect consequences of any possible species interaction matrix.
@@
 rates via maximum likelihood, controlling for indirect effects. Using simulated
-landscapes with known pairwise interaction strengths, I evaluated Markov
-networks and six existing approaches. The Markov networks consistently
-outperformed other methods, correctly isolating direct interactions between
-species pairs even when indirect interactions or abiotic factors largely
-overpowered them. Two computationally efficient approximations, based on
-controlling for indirect effects with linear or generalized linear models, also
-performed well. Indirect effects reliably caused a common null modeling approach
-to produce incorrect inferences, however.
+landscapes with known interactions, I evaluated Markov networks and six existing
+approaches. The Markov networks consistently outperformed other methods, 
+correctly isolating direct interactions between species pairs even when indirect
+interactions or abiotic factors largely overpowered them. Two computationally 
+efficient approximations, based on controlling for indirect effects with linear 
+or generalized linear models, also performed well. Null models showed no 
+evidence of being able to control for indirect effects, and reliably yielded 
+incorrect inferences when such effects were present.
 
@@
 \noindent \textbf{Simulated landscapes.}
-I simulated several sets of landscapes using known parameters to evaluate
-different statistical methods' performance. The first set of landscapes included
-the three competing species shown in Figure 1. For each of 1000 replicates, I
-generated a landscape with 100 sites by sampling from a probability distribution
-defined by the figure's interaction coefficients (Appendix 1). Each of the
-methods described below was then evaluated on its ability to correctly infer
-that the two shrub species competed with one another, despite their frequent
-co-occurrence in unshaded areas.
+I simulated several sets of landscapes using known parameters so that model
+estimates could be compared with "true" values. The first set of landscapes 
+included the three competing species shown in Figure 1. For each of 1000 
+replicates, I generated a landscape with 100 sites by sampling from a 
+probability distribution defined by the figure's interaction coefficients 
+(Appendix 1). Each of the methods described below was then evaluated on its 
+ability to correctly infer that the two shrub species competed with one another,
+despite their frequent co-occurrence in unshaded areas.
 
@@
 combination of size and model). The simplest set of simulated landscapes were
-generated with Gibbs samples from Equation 1, and provided a test of methods'
-ability to isolate direct species-level interactions from a web of indirect 
-ones. For each replicate, I randomly drew the “true” $\beta$ coefficient
-magnitudes an exponential distribution with rate 1 so that
-most species pairs interacted negligibly but a few interactions were strong
-enough to propagate through the newtwork. I randomly assigned 25% of the
-interactions to be positive; the remainder were negative.
+generated with Gibbs samples from Equation 1. For each replicate, I randomly 
+drew the “true” $\beta$ coefficient magnitudes from an exponential distribution 
+with rate 1 so that most species pairs interacted negligibly but a few 
+interactions were strong enough to propagate through the newtwork. I randomly 
+assigned 25% of the interactions to be positive; the remainder were negative.
 
@@
 per-capita basis. To avoid runaway mutualisms leading to infinite abundance,
-all interaction coefficients were negative in these simulations. If a method
-performs well on these landscapes, it would indicate some robustness to the
-biolocical mechanisms underlying species' interactiosn with one another.
+all interaction coefficients were negative in these simulations. Good 
+performance on these landscapes would indicate some robustness to the
+mechanistic details of species interactions.
 
@@
 First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
-models, as described above. For the analyses with 20 species, a weak regularizer
-(equivalent to a logistic prior with scale 2) ensured that the estimates were
-always finite.
+models, as described above. For the analyses with 20 species, a 
+weakly-informative regularizer (equivalent to a logistic prior with location 0
+and scale 2) ensured that the estimates were always finite.
 
@@
 @faisal_inferring_2010 for two exceptions). In the context of non-Gaussian data,
-the partial correlation can be thought of as a computationally efficient
-approximation to the full Markov network model [@loh_structure_2013]. Partial
-correlations are undefined for landscapes with perfectly-correlated species
-pairs, so I used the regularized estimate provided by the corpcor package’s 
-`pcor.shrink` function with the default settings [@schafer_corpcor_2014].
+the partial correlation (or partial covariance) can be thought of as a 
+computationally efficient approximation to the full Markov network model
+[@loh_structure_2013]. Partial correlations are undefined for landscapes with
+perfectly-correlated species pairs, so I used the regularized estimate provided 
+by the corpcor package’s `pcor.shrink` function with the default settings
+[@schafer_corpcor_2014].
 
@@
 environmental heterogeneity and indirect interactions cause ecologists' typical
-null modeling approaches to reliably fail. Partial covariances and generalized
+null modeling approaches to reliably fail. Partial correlations and generalized
 linear models can both provide computationally efficient approximations, but
@@
 vectors positively correlated (circled). **C.** Controlling for trees with a
-conditional (all-else-equal) approach such as a partial covariance or a Markov
+conditional (all-else-equal) approach such as a partial correlation or a Markov
 network leads to correct identification of the negative shrub-shrub interaction

### 2016-02-18 Harris_2015/manuscript-materials/draft.tex
many changes

@@
 example, the negative correlation between two competing species can be
-reversed in the presence of a third species that is capable of
-outcompeting both of them. Here, I apply models from statistical
-physics, called Markov networks or Markov random fields, that can
-predict the direct and indirect consequences of any possible species
-interaction matrix. Interactions in these models can be estimated from
-observed co-occurrence rates via maximum likelihood, controlling for
-indirect effects. Using simulated landscapes with known pairwise
-interaction strengths, I evaluated Markov networks and six existing
-approaches. The Markov networks consistently outperformed other methods,
-correctly isolating direct interactions between species pairs even when
-indirect interactions or abiotic factors largely overpowered them. Two
-computationally efficient approximations, based on controlling for
-indirect effects with linear or generalized linear models, also
-performed well. Indirect effects reliably caused a common null modeling
-approach to produce incorrect inferences, however.
+reversed in the presence of a third species that outcompetes both of
+them. Here, I apply models from statistical physics, called Markov
+networks or Markov random fields, that can predict the direct and
+indirect consequences of any possible species interaction matrix.
+Interactions in these models can be estimated from observed
+co-occurrence rates via maximum likelihood, controlling for indirect
+effects. Using simulated landscapes with known interactions, I evaluated
+Markov networks and six existing approaches. The Markov networks
+consistently outperformed other methods, correctly isolating direct
+interactions between species pairs even when indirect interactions or
+abiotic factors largely overpowered them. Two computationally efficient
+approximations, based on controlling for indirect effects with linear or
+generalized linear models, also performed well. Null models showed no
+evidence of being able to control for indirect effects, and reliably
+yielded incorrect inferences when such effects were present.
 
@@
 co-occurrence of species \(i\) and species \(j\) contributes to the
-log-probability; it controls the conditional relationship between two
+log-probability; it determines the conditional relationship between two
 species, i.e.~the probability that they will be found together, after
@@
 approximate it (Lee and Hastie 2012) or approximate its gradient
-({\textbf{???}}). Such approximations are beyond the scope of this
-paper, however. For the analyses presented here, I used the rosalia
-package (Harris 2015a) for the R programming language (R Core Team 2015)
-to calculate \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly
-(Murphy 2012); the package passes these functions to the ``BFGS'' method
-in R's general-purpose optimizer, which finds values for \(\alpha\) and
+({\textbf{???}}) if they want to fit a Markov network model. For the
+analyses presented here, I used the rosalia package (Harris 2015a) for
+the R programming language (R Core Team 2015) to calculate
+\(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
+the package passes these functions to the ``BFGS'' method in R's
+general-purpose optimizer, which finds values for \(\alpha\) and
 \(\beta\).
@@
 \noindent \textbf{Simulated landscapes.} I simulated several sets of
-landscapes using known parameters to evaluate different statistical
-methods' performance. The first set of landscapes included the three
-competing species shown in Figure 1. For each of 1000 replicates, I
-generated a landscape with 100 sites by sampling from a probability
+landscapes using known parameters so that model estimates could be
+compared with ``true'' values. The first set of landscapes included the
+three competing species shown in Figure 1. For each of 1000 replicates,
+I generated a landscape with 100 sites by sampling from a probability
 distribution defined by the figure's interaction coefficients (Appendix
@@
 replicates for each combination of size and model). The simplest set of
-simulated landscapes were generated with Gibbs samples from Equation 1,
-and provided a test of methods' ability to isolate direct species-level
-interactions from a web of indirect ones. For each replicate, I randomly
-drew the ``true'' \(\beta\) coefficient magnitudes an exponential
-distribution with rate 1 so that most species pairs interacted
-negligibly but a few interactions were strong enough to propagate
-through the newtwork. I randomly assigned 25\% of the interactions to be
-positive; the remainder were negative.
+simulated landscapes were generated with Gibbs samples from Equation 1.
+For each replicate, I randomly drew the ``true'' \(\beta\) coefficient
+magnitudes from an exponential distribution with rate 1 so that most
+species pairs interacted negligibly but a few interactions were strong
+enough to propagate through the newtwork. I randomly assigned 25\% of
+the interactions to be positive; the remainder were negative.
 
@@
 leading to infinite abundance, all interaction coefficients were
-negative in these simulations. If a method performs well on these
-landscapes, it would indicate some robustness to the biolocical
-mechanisms underlying species' interactiosn with one another.
+negative in these simulations. Good performance on these landscapes
+would indicate some robustness to the mechanistic details of species
+interactions.
 
@@
 First, I used the rosalia package (Harris 2015a) to fit Markov network
-models, as described above. For the analyses with 20 species, a weak
-regularizer (equivalent to a logistic prior with scale 2) ensured that
-the estimates were always finite.
+models, as described above. For the analyses with 20 species, a
+weakly-informative regularizer (equivalent to a logistic prior with
+location 0 and scale 2) ensured that the estimates were always finite.
 
@@
 Faisal et al. (2010) for two exceptions). In the context of non-Gaussian
-data, the partial correlation can be thought of as a computationally
-efficient approximation to the full Markov network model (Loh and
-Wainwright 2013). Partial correlations are undefined for landscapes with
-perfectly-correlated species pairs, so I used the regularized estimate
-provided by the corpcor package's \texttt{pcor.shrink} function with the
-default settings (Schäfer et al. 2014).
+data, the partial correlation (or partial covariance) can be thought of
+as a computationally efficient approximation to the full Markov network
+model (Loh and Wainwright 2013). Partial correlations are undefined for
+landscapes with perfectly-correlated species pairs, so I used the
+regularized estimate provided by the corpcor package's
+\texttt{pcor.shrink} function with the default settings (Schäfer et al.
+2014).
 
@@
 interactions cause ecologists' typical null modeling approaches to
-reliably fail. Partial covariances and generalized linear models can
+reliably fail. Partial correlations and generalized linear models can
 both provide computationally efficient approximations, but with somewhat
@@
 correlated (circled). \textbf{C.} Controlling for trees with a
-conditional (all-else-equal) approach such as a partial covariance or a
+conditional (all-else-equal) approach such as a partial correlation or a
 Markov network leads to correct identification of the negative

### 2016-02-18 Harris_2015/recovery.Rmd
as actually submitted to Ecology
many changes

@@
 
-[^1]: The PDF version of this document has been manually altered to omit 150
-lines of output from the `corpcor` package of the form "`## Estimating optimal shrinkage intensity lambda (correlation matrix): 0.0326`"
-
 Initialization:
@@
 
-```{r, cache = FALSE, eval = TRUE}
+```{r, cache = FALSE, eval = FALSE}
 fit_all = function(identifier){
@@
   ######## Markov network ########
-  rosie = rosalia(x, maxit = 200, trace = 0, prior = make_logistic_prior(scale = 2), hessian = TRUE)
+  rosie = rosalia(x, maxit = 200, trace = 0, 
+                  prior = make_logistic_prior(scale = 2), hessian = TRUE)
 
@@
 
-mclapply(identifiers, fit_all, mc.cores = mc.cores, mc.preschedule = FALSE, mc.silent = TRUE) %>% 
+mclapply(identifiers, fit_all, mc.cores = mc.cores, 
+         mc.preschedule = FALSE, mc.silent = TRUE) %>% 
   bind_rows() %>% 
@@
 ```
+
+### Logistic prior used for the Markov network
+
+From the R help file for the logistic distribution:
+
+>The Logistic distribution with location = m and scale = s has... density
+
+>> f(x) = 1/s exp((x-m)/s) (1 + exp((x-m)/s))^-2.
+
+> It is a long-tailed distribution with mean m and variance $\pi$^2 /3 s^2.
+
+As noted in the main text, I used location zero and scale 2 (plotted in
+red below). The "true" parameter distribution for $\beta$ is drawn in black,
+for reference.  These distributions have different means and the prior is
+substantially wider than the distribution of true parameters (mean absolute
+deviation from zero about `r round(mean(abs(rlogis(1E7, scale = 2))), 1)`
+times larger).
+
+```{r}
+curve(ifelse(x < 0, .75 * dexp(abs(x)), .25 * dexp(x)), from = -15, to = 15,
+      n = 1000, ylab = "Density")
+curve(dlogis(x, location = 0, scale = 2), add = TRUE, col = "red")
+```
+
+The log of this density was added to the log-likelihood to calculate an
+un-normalized log-posterior, which is optimized by the `optim` function in the
+`stats` package.
+
+This prior distribution is "weakly informative" in the sense of Gelman et al.
+(2008. Annals of Applied Statistics, "A weakly informative default prior 
+distribution for logistic and other regression models"), as shown below:
+the logistic regularizer pulls the estimated values toward zero less strongly
+than Gelman et al.'s weakly-informative Cauchy prior does.
+
+```{r}
+x = as.matrix(
+  read.csv("fakedata/matrices/25-no_env1.csv")[,-1]
+)
+
+# Don't estimate parameters for species that were never observed
+x = x[ , colMeans(x) != 0]
+
+# Estimate parameters using a weakly informative logistic prior
+logistic_fit = rosalia(x, prior = make_logistic_prior(scale = 2),
+                             trace = FALSE)
+
+# Estimate parameters using Gelman et al.'s weakly informative Cauchy prior
+cauchy_fit = rosalia(x, prior = make_cauchy_prior(scale = 2.5),
+                           trace = FALSE)
+
+plot(logistic_fit$beta, cauchy_fit$beta)
+abline(0,1, h = 0, v = 0, col = "darkgray")
+```

### 2016-02-18 Harris_2015/recovery.tex
many changes

@@
 \newcommand{\NormalTok}[1]{{#1}}
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \setlength{\parindent}{0pt}
@@
 simulated data from Appendix 2 and how each model's performance was
-evaluated.\footnote{The PDF version of this document has been manually
-  altered to omit 150 lines of output from the \texttt{corpcor} package
-  of the form
-  ``\texttt{\#\#\ Estimating\ optimal\ shrinkage\ intensity\ lambda\ (correlation\ matrix):\ 0.0326}''}
+evaluated.{[}\^{}1{]}
 
@@
 
+\subsubsection{Logistic prior used for the Markov
+network}\label{logistic-prior-used-for-the-markov-network}
+
+From the R help file for the logistic distribution:
+
+\begin{quote}
+The Logistic distribution with location = m and scale = s has\ldots{}
+density
+\end{quote}
+
+\begin{quote}
+\begin{quote}
+f(x) = 1/s exp((x-m)/s) (1 + exp((x-m)/s))\^{}-2.
+\end{quote}
+\end{quote}
+
+\begin{quote}
+It is a long-tailed distribution with mean m and variance \(\pi\)\^{}2
+/3 s\^{}2.
+\end{quote}
+
+As noted in the main text, I used location zero and scale 2 (plotted in
+red below). The ``true'' parameter distribution for \(\beta\) is drawn
+in black, for reference. These distributions have different means and
+the prior is substantially wider than the distribution of true
+parameters (mean absolute deviation from zero about 2.8 times larger).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{curve}\NormalTok{(}\KeywordTok{ifelse}\NormalTok{(x <}\StringTok{ }\DecValTok{0}\NormalTok{, .}\DecValTok{75} \NormalTok{*}\StringTok{ }\KeywordTok{dexp}\NormalTok{(}\KeywordTok{abs}\NormalTok{(x)), .}\DecValTok{25} \NormalTok{*}\StringTok{ }\KeywordTok{dexp}\NormalTok{(x)), }\DataTypeTok{from =} \NormalTok{-}\DecValTok{15}\NormalTok{, }\DataTypeTok{to =} \DecValTok{15}\NormalTok{,}
+      \DataTypeTok{n =} \DecValTok{1000}\NormalTok{, }\DataTypeTok{ylab =} \StringTok{"Density"}\NormalTok{)}
+\KeywordTok{curve}\NormalTok{(}\KeywordTok{dlogis}\NormalTok{(x, }\DataTypeTok{location =} \DecValTok{0}\NormalTok{, }\DataTypeTok{scale =} \DecValTok{2}\NormalTok{), }\DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{col =} \StringTok{"red"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{recovery_files/figure-latex/unnamed-chunk-4-1.pdf}
+
+The log of this density was added to the log-likelihood to calculate an
+un-normalized log-posterior, which is optimized by the \texttt{optim}
+function in the \texttt{stats} package.
+
+This prior distribution is ``weakly informative'' in the sense of Gelman
+et al. (2008. Annals of Applied Statistics, ``A weakly informative
+default prior distribution for logistic and other regression models''),
+as shown below: the logistic regularizer pulls the estimated values
+toward zero less strongly than Gelman et al.'s weakly-informative Cauchy
+prior does.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{x =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(}
+  \KeywordTok{read.csv}\NormalTok{(}\StringTok{"fakedata/matrices/25-no_env1.csv"}\NormalTok{)[,-}\DecValTok{1}\NormalTok{]}
+\NormalTok{)}
+
+\CommentTok{# Don't estimate parameters for species that were never observed}
+\NormalTok{x =}\StringTok{ }\NormalTok{x[ , }\KeywordTok{colMeans}\NormalTok{(x) !=}\StringTok{ }\DecValTok{0}\NormalTok{]}
+
+\CommentTok{# Estimate parameters using a weakly informative logistic prior}
+\NormalTok{logistic_fit =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{prior =} \KeywordTok{make_logistic_prior}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{2}\NormalTok{),}
+                             \DataTypeTok{trace =} \OtherTok{FALSE}\NormalTok{)}
+
+\CommentTok{# Estimate parameters using Gelman et al.'s weakly informative Cauchy prior}
+\NormalTok{cauchy_fit =}\StringTok{ }\KeywordTok{rosalia}\NormalTok{(x, }\DataTypeTok{prior =} \KeywordTok{make_cauchy_prior}\NormalTok{(}\DataTypeTok{scale =} \FloatTok{2.5}\NormalTok{),}
+                           \DataTypeTok{trace =} \OtherTok{FALSE}\NormalTok{)}
+
+\KeywordTok{plot}\NormalTok{(logistic_fit$beta, cauchy_fit$beta)}
+\KeywordTok{abline}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{, }\DataTypeTok{h =} \DecValTok{0}\NormalTok{, }\DataTypeTok{v =} \DecValTok{0}\NormalTok{, }\DataTypeTok{col =} \StringTok{"darkgray"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{recovery_files/figure-latex/unnamed-chunk-5-1.pdf}
+
 \end{document}

### 2016-02-18 Harris_2015/show-results.Rmd
many changes

@@
 # Sort a vector of alphanumeric strings by the numeric component
-# as if they were integers.  For example, V20 is larger than V12,
-# even though V12 comes first alphabetically
+# as if they were integers.  For example, V20 should be larger 
+# than V12, even though V12 comes first alphabetically
 alnum_sort = function(x){
@@
 
-    # Pull out the corresponding chunk of the "x" data frame, based on n_sites and rep_name
-    x_subset = x[x$n_sites == n_sites & x$rep_name == rep_name & x$method == "correlation", ]
+    # Pull out the corresponding chunk of the "x" data frame, based on n_sites 
+    # and rep_name
+    is_correct_sim = x$n_sites == n_sites & x$rep_name == rep_name
+    x_subset = x[is_correct_sim & x$method == "correlation", ]
 
@@
 
-## Calculate model performance:
+## Calculate model performance (R-squared):
+
+For each combination of `method` and `simulation_type`, fit a linear model.
+Then, for each combination of `method`, `simulation_type`, and `n_sites`,
+report the proportion of variance explained by the corresponding linear model.
 
@@
 
+# Set the ordering of the data frame based on R-squared
 result_summary$method = reorder(result_summary$method, -result_summary$r2)
+result_summary$simulation_type = reorder(
+  result_summary$simulation_type, 
+  -result_summary$r2
+)
 
-result_summary$simulation_type = reorder(result_summary$simulation_type, -result_summary$r2)
-
+# Add a column that includes method name and its mean R-squared,
+# for plotting purposes below
 result_summary = result_summary %>% 
@@
 
+# Rename the simulation types for clearer graph labels
 result_summary$simulation_type_long = plyr::revalue(
@@
 
-#### Model performance with no environmental variation:
+#### Calculate average R-squared across methods and simulation types
 
@@
 result_summary %>% 
-  filter(simulation_type == "no_env") %>%
-  group_by(method) %>% 
+  group_by(method, simulation_type) %>% 
   summarise(mean(r2)) %>% 
+  spread(simulation_type, `mean(r2)`) %>%
   kable(digits = 3)
@@
 
-#### Model performance with environmental heterogeneity:
-
-```{r}
-result_summary %>% 
-  filter(simulation_type == "env") %>%
-  group_by(method) %>% 
-  summarise(mean(r2)) %>% 
-  kable(digits = 3)
-```
-
-#### Model performance with per-capita species interactions:
-
-```{r}
-result_summary %>% 
-  filter(simulation_type == "abund") %>%
-  group_by(method) %>% 
-  summarise(mean(r2)) %>% 
-  kable(digits = 3)
-```
-
-## Plot the R-squared results (Figure 3)
+## Save the finer-grained R-squared results as Figure 3
 
@@
   geom_vline(xintercept = 0, size = 1) + 
-  #scale_shape_manual(values = c(21, 25, 15, 18)) + 
   coord_cartesian(ylim = c(-.01, 0.76)) + 
@@
     data = landscape_estimates
-  )
+  ),
+  correlation = FALSE
 )
@@
 
-## Summarize the inferential results of the Markov network and the null model:
+## Identify statistical significance for the Markov network and Pairs
+
+For Pairs, compare the Z-scores with Gaussian quantiles for 95% coverage;
+for the Markov network, use the approximate confidence intervals estimated in
+Appendix 3.
 
@@
 # Pairs's Z score is based on C-scores, which are positive when species are 
-# disaggregated.  So significantly positive interactions == negative estimates
+# disaggregated.  So significantly negative scores imply positive interactions
+# and vice versa
 pairs_summary$sig_pos = pairs_summary$estimate < qnorm(.025)
@@
 markov_summary$sig_neg = markov_summary$upper < 0
+
+confidently_right = function(data){
+  (data$sig_pos & data$truth > 0) | (data$sig_neg & data$truth < 0)
+}
+
+confidently_wrong = function(data){
+  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
+}
 ```
@@
 ```{r, cache = TRUE}
-
-# Function for smoothing the error rates with a generalized additive model
-# and plotting the results
-my_geom_smooth = function(data, color, ...){
-  geom_smooth(
-    data = data,
-    aes(
-      x = truth, 
-      y = as.integer((sig_neg & truth > 0) | (sig_pos & truth < 0)),
-      color = color
-    ),
-    method = gam, 
-    family = binomial, 
-    formula = y ~ s(x),
-    se = FALSE,
-    n = 1024,
-    ...
-  )
-}
 
@@
 
-error_smoother = function(data){
+error_smoother = function(data, f){
   predict(
     gam(
-      I((sig_neg & truth > 0) | (sig_pos & truth < 0)) ~ s(truth),
+      f(data) ~ s(truth),
       data = data,
@@
 
-y_pairs = error_smoother(pairs_summary)
-y_markov = error_smoother(markov_summary)
+p_right_pairs = error_smoother(pairs_summary, confidently_right)
+p_wrong_pairs = error_smoother(pairs_summary, confidently_wrong)
+
+p_right_markov = error_smoother(markov_summary, confidently_right)
+p_wrong_markov = error_smoother(markov_summary, confidently_wrong)
+
+par(mfrow = c(1, 2))
+plot(truth_seq, p_right_pairs + p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
+lines(truth_seq, p_right_markov + p_wrong_markov, col = 2)
+abline(h = 0.05)
+
+plot(truth_seq, p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
+lines(truth_seq, p_wrong_markov, col = 2)
+par(mfrow = c(1, 1))
+
+y_markov = error_smoother(markov_summary, confidently_wrong)
 
@@
   facet_grid(~simulation_type) + 
-  geom_smooth(method = gam, formula = y ~ s(x), family = binomial) + 
+  geom_smooth(method = gam, formula = y ~ s(x), method.args = list(family = "binomial")) + 
   theme_bw() + 

### 2016-02-18 Harris_2015/show-results.tex
many changes

@@
 
-    \CommentTok{# Pull out the corresponding chunk of the "x" data frame, based on n_sites and rep_name}
-    \NormalTok{x_subset =}\StringTok{ }\NormalTok{x[x$n_sites ==}\StringTok{ }\NormalTok{n_sites &}\StringTok{ }\NormalTok{x$rep_name ==}\StringTok{ }\NormalTok{rep_name &}\StringTok{ }\NormalTok{x$method ==}\StringTok{ "correlation"}\NormalTok{, ]}
+    \CommentTok{# Pull out the corresponding chunk of the "x" data frame, based on n_sites }
+    \CommentTok{# and rep_name}
+    \NormalTok{is_correct_sim =}\StringTok{ }\NormalTok{x$n_sites ==}\StringTok{ }\NormalTok{n_sites &}\StringTok{ }\NormalTok{x$rep_name ==}\StringTok{ }\NormalTok{rep_name}
+    \NormalTok{x_subset =}\StringTok{ }\NormalTok{x[is_correct_sim &}\StringTok{ }\NormalTok{x$method ==}\StringTok{ "correlation"}\NormalTok{, ]}
 
@@
 
-\NormalTok{result_summary$simulation_type =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(result_summary$simulation_type, -result_summary$r2)}
+\NormalTok{result_summary$simulation_type =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(}
+  \NormalTok{result_summary$simulation_type, }
+  \NormalTok{-result_summary$r2}
+\NormalTok{)}
 
@@
 
-\paragraph{Model performance with no environmental
-variation:}\label{model-performance-with-no-environmental-variation}
+\paragraph{Calculate average R-squared across methods and simulation
+types}\label{calculate-average-r-squared-across-methods-and-simulation-types}
 
@@
 \NormalTok{result_summary %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "no_env"}\NormalTok{) %>%}
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{group_by}\NormalTok{(method, simulation_type) %>%}\StringTok{ }
 \StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{spread}\NormalTok{(simulation_type, }\StringTok{`}\DataTypeTok{mean(r2)}\StringTok{`}\NormalTok{) %>%}
 \StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
@@
 
-\begin{longtable}[c]{@{}lr@{}}
+\begin{longtable}[c]{@{}lrrr@{}}
 \toprule
-method & mean(r2)\tabularnewline
+method & no\_env & env & abund\tabularnewline
 \midrule
 \endhead
-Markov network & 0.525\tabularnewline
-GLM & 0.472\tabularnewline
-partial correlation & 0.403\tabularnewline
-partial BayesComm & 0.394\tabularnewline
-correlation & 0.291\tabularnewline
-null & 0.227\tabularnewline
-BayesComm & 0.206\tabularnewline
+Markov network & 0.525 & 0.451 & 0.384\tabularnewline
+GLM & 0.472 & 0.405 & 0.283\tabularnewline
+partial correlation & 0.403 & 0.322 & 0.200\tabularnewline
+partial BayesComm & 0.394 & 0.302 & 0.166\tabularnewline
+correlation & 0.291 & 0.183 & 0.117\tabularnewline
+null & 0.227 & 0.125 & 0.075\tabularnewline
+BayesComm & 0.206 & 0.110 & 0.060\tabularnewline
 \bottomrule
@@
 
-\paragraph{Model performance with environmental
-heterogeneity:}\label{model-performance-with-environmental-heterogeneity}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{result_summary %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "env"}\NormalTok{) %>%}
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[c]{@{}lr@{}}
-\toprule
-method & mean(r2)\tabularnewline
-\midrule
-\endhead
-Markov network & 0.451\tabularnewline
-GLM & 0.405\tabularnewline
-partial correlation & 0.322\tabularnewline
-partial BayesComm & 0.302\tabularnewline
-correlation & 0.183\tabularnewline
-null & 0.125\tabularnewline
-BayesComm & 0.110\tabularnewline
-\bottomrule
-\end{longtable}
-
-\paragraph{Model performance with per-capita species
-interactions:}\label{model-performance-with-per-capita-species-interactions}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{result_summary %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type ==}\StringTok{ "abund"}\NormalTok{) %>%}
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(method) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarise}\NormalTok{(}\KeywordTok{mean}\NormalTok{(r2)) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[c]{@{}lr@{}}
-\toprule
-method & mean(r2)\tabularnewline
-\midrule
-\endhead
-Markov network & 0.384\tabularnewline
-GLM & 0.283\tabularnewline
-partial correlation & 0.200\tabularnewline
-partial BayesComm & 0.166\tabularnewline
-correlation & 0.117\tabularnewline
-null & 0.075\tabularnewline
-BayesComm & 0.060\tabularnewline
-\bottomrule
-\end{longtable}
-
-\subsection{Plot the R-squared results (Figure
-3)}\label{plot-the-r-squared-results-figure-3}
+\subsection{Save the R-squared results as Figure
+3}\label{save-the-r-squared-results-as-figure-3}
 
@@
 
-\subsection{Summarize the inferential results of the Markov network and
-the null
-model:}\label{summarize-the-inferential-results-of-the-markov-network-and-the-null-model}
+\subsection{Identify statistical significance for the Markov network and
+Pairs}\label{identify-statistical-significance-for-the-markov-network-and-pairs}
 
@@
 
-\begin{verbatim}
-## geom_smooth: method="auto" and size of largest group is >=1000, so using gam with formula: y ~ s(x, bs = "cs"). Use 'method = x' to change the smoothing method.
-\end{verbatim}
-
-\includegraphics{show-results_files/figure-latex/unnamed-chunk-13-1.pdf}
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-11-1.pdf}
 
@@
 \StringTok{  }\KeywordTok{facet_grid}\NormalTok{(~simulation_type) +}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \NormalTok{gam, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x), }\DataTypeTok{family =} \NormalTok{binomial) +}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \NormalTok{gam, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x), }\DataTypeTok{method.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{family =} \StringTok{"binomial"}\NormalTok{)) +}\StringTok{ }
 \StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
@@
 
-\includegraphics{show-results_files/figure-latex/unnamed-chunk-14-1.pdf}
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-12-1.pdf}
 

### 2016-02-18 Harris_2015/simulation.Rmd
many changes

@@
 The following function draws random coefficients for a Markov network of a pre-specified size.
+
+The $alpha$ coefficients controlling species' intercepts were drawn from 
+Gaussian distributions with unit variance and a mean given by `mean_alpha`.
+In the R code at the end of this document, `mean_alpha` was defined as $-1$ for presence-absence simulations and $5$ for the abundance simulations.
+
+The $\beta$ coefficients' magnitudes were drawn from an exponential 
+distribution with rate parameter equal to 1.  The sign of each coefficient was
+randomly sampled, based on the value of `p_neg`, defined below. as 0.75 (75%
+negative) for the presence-absence simulations and 1.0 (100% negative) for the
+simulations that included abundance.
 
@@
 
-I used the following R function to generate "true" parameters with the method above, 
-then simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and finally save the results to a "fakedata" folder.
-
+The following function calls the function above to generate "true" parameters, 
+then uses them to simulate a landscape with Markov chain Monte Carlo, and 
+finally saves the results to a "fakedata" folder.
 I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
 In each round of Gibbs sampling, I cycled through all the species, randomly
-updating each one's presence/absence vector in response to its conditional
-occurrence probability:
+updating each one's binary presence/absence vector with random samples from
+a Bernoulli distribution whose conditional probability was defined by
 
@@
 
-where the logistic function is $\frac{1}{1+e^{-x}}$.
+where the logistic function is $\frac{1}{1+e^{-x}}$. For efficiency, 
+$\sum_j\beta_{ij}y_j$ is calculated with matrix multiplication in the code
+below.
 
-For the abundance-based simulations, abundance was simulated with a Poisson
-distribution with rate parameter given by 
-$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$.
+For simulations that included environmental heterogeneity, $\alpha_i$ varied 
+from site to site as follows:
+
+* `n_env` environmental variables were simulated for each site; these were
+drawn from zero-mean Gaussians with standard deviation given by `sd`. In the 
+R code at the end of this document, both `n_env` and `sd` were set to 2, 
+corresponding to an environment with two important axes of variation and 
+had a standard deviation of two.
+
+* The local value of $\alpha_i$ for species $i$ is given by the species-level
+$\alpha$ defined by the `make_coefficients` function, plus a linear combination
+of the environmental variables. For efficiency, this linear combination is 
+calculated with matrix multiplication. The weights associated with this linear 
+combination (`alpha_env`) are drawn from a Gaussian distribution with zero mean 
+and unit variance.
+
+For the abundance-based simulations, each species' conditional abundance was
+simulated with a Poisson distribution instead of a Bernoulli distribution. The
+rate parameter of this distribution was given by 
+$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$. (The more conventional 
+exponential inverse link was not used because preliminary experiments showed that
+it led to very large abundances that prevented the Gibbs sampler from mixing 
+effectively.)
+
+Each simulated landscape was saved after the Gibbs sampler had updated the 
+value of each species `n_gibbs` times; see the end of this document for the
+value used.
 
@@
   # or abundance depends on its alpha value and on the environment (assuming alpha_env 
-  # is not zero).
+  # is not set to zero).
   x = matrix(
@@
 
-Loop to simulate and save all the landscapes:
+Loop to simulate and save 50 landscapes for each combination of simulation type
+and landscape size.
 

### 2016-02-19 stochastic/stochastic-recovery.Rmd
don't cache

@@
 
-```{r,cache = TRUE}
+```{r}
 library(rosalia)
@@
 
-```{r,cache = TRUE}
+```{r}
 set.seed(1)
@@
 
-```{r,, cache = TRUE}
+```{r}
 # Initialize the `y` matrix.
@@
 
-```{r,cache = TRUE}
+```{r}
 # Calculate sufficient statistics of the data
@@
 
-```{r,cache = TRUE}
+```{r}
 # Initialize the simulated landscape for stochastic approximation
@@
 
-```{r,cache = TRUE}
+```{r}
 # Very weak priors on alpha terms, somewhat stronger on beta terms
@@
 
-```{r,cache = TRUE}
+```{r}
 initial_learning_rate = 1 # step size at start of optimization
@@
 
-```{r,cache = TRUE}
+```{r}
 library(cowplot)
@@
 
-```{r,cache = TRUE}
+```{r}
 small_n_spp = 20
@@
 
-```{r,cache = TRUE}
+```{r}
 exact_time = system.time({
@@
 
-```{r,cache = TRUE}
+```{r}
 maxit_small = 50000
@@
 
-```{r,cache = TRUE}
+```{r}
 pdf("stochastic/convergence.pdf", width = 10, height = 7)

### 2016-02-22 Harris_2015/manuscript-materials/draft.Rmd
more manuscript and appendices

@@
 interactions or abiotic factors largely overpowered them. Two computationally 
-efficient approximations, based on controlling for indirect effects with linear 
-or generalized linear models, also performed well. Null models showed no 
-evidence of being able to control for indirect effects, and reliably yielded 
-incorrect inferences when such effects were present.
+efficient approximations, based on controlling for indirect effects with partial
+correlations or generalized linear models, also performed well. Null models 
+showed no evidence of being able to control for indirect effects, and reliably 
+yielded incorrect inferences when such effects were present.
 
@@
 
-In the third set of landscapes, I simulated each species' abundance (instead of
-just their presence/absence), and interactions between species occurred on a 
-per-capita basis. To avoid runaway mutualisms leading to infinite abundance,
+In the final set of landscapes, I simulated each species' abundance (instead of
+just their presence/absence); here, interactions between species occurred on a 
+per-capita basis. To prevent runaway mutualisms leading to infinite abundance,
 all interaction coefficients were negative in these simulations. Good 
@@
 interaction strengths using a generalized additive model with binomial responses
-(reject versus not; Appendix 4). The Type I error rate was defined as this
-model's estimated probability of rejecting the null hypothesis when the "true"
-interaction was zero.
+(reject versus not; Appendix 4). The probability of rejection when the "true"
+value of $\beta$ was zero was defined as the Type I error rate.
 
@@
 interpret).  The Markov network's Type I error rate depended on the 
-simulation type: [[0.02]] for simulations that matched the model's assumptions,
-versus [[0.14]] for simulations that included environmental heterogeneity (see 
+simulation type: 0.02 for simulations that matched the model's assumptions,
+versus 0.14 for simulations that included environmental heterogeneity (see 
 Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
-values). The null model's Type I error rates were [[0.30]] and [[0.50]] for the
+values). The null model's Type I error rates were 0.31 and 0.53 for the
 constant and heterogeneous landscapes, respectively---far higher than the 
@@
 predict the wrong sign of the interaction with 95% confidence.  The null model 
-makes such errors more than [[5]] times as often as the Markov network, even
-though it only rejects the null hypothesis [[twice]] as often overall (Appendix
+makes such errors more than 5 times as often as the Markov network, even
+though it only rejects the null hypothesis twice as often overall (Appendix
 4). The Markov network's errors were also more concentrated around 0, so it 
-never misclassified strong interactions like the null model did.
+never misclassified strong interactions like the null model did (Figure 4C).
 

### 2016-02-22 Harris_2015/manuscript-materials/draft.tex
more manuscript and appendices

@@
 abiotic factors largely overpowered them. Two computationally efficient
-approximations, based on controlling for indirect effects with linear or
-generalized linear models, also performed well. Null models showed no
-evidence of being able to control for indirect effects, and reliably
-yielded incorrect inferences when such effects were present.
+approximations, based on controlling for indirect effects with partial
+correlations or generalized linear models, also performed well. Null
+models showed no evidence of being able to control for indirect effects,
+and reliably yielded incorrect inferences when such effects were
+present.
 
@@
 1 can become intractable, and researchers will either need to
-approximate it (Lee and Hastie 2012) or approximate its gradient
-({\textbf{???}}) if they want to fit a Markov network model. For the
-analyses presented here, I used the rosalia package (Harris 2015a) for
-the R programming language (R Core Team 2015) to calculate
+approximate it (Lee and Hastie 2012) or approximate its gradient (Harris
+2016) if they want to fit a Markov network model. For the analyses
+presented here, I used the rosalia package (Harris 2015a) for the R
+programming language (R Core Team 2015) to calculate
 \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
@@
 
-In the third set of landscapes, I simulated each species' abundance
-(instead of just their presence/absence), and interactions between
-species occurred on a per-capita basis. To avoid runaway mutualisms
+In the final set of landscapes, I simulated each species' abundance
+(instead of just their presence/absence); here, interactions between
+species occurred on a per-capita basis. To prevent runaway mutualisms
 leading to infinite abundance, all interaction coefficients were
@@
 ``true'' interaction strengths using a generalized additive model with
-binomial responses (reject versus not; Appendix 4). The Type I error
-rate was defined as this model's estimated probability of rejecting the
-null hypothesis when the ``true'' interaction was zero.
+binomial responses (reject versus not; Appendix 4). The probability of
+rejection when the ``true'' value of \(\beta\) was zero was defined as
+the Type I error rate.
 
@@
 easiest to interpret). The Markov network's Type I error rate depended
-on the simulation type: {[}{[}0.02{]}{]} for simulations that matched
-the model's assumptions, versus {[}{[}0.14{]}{]} for simulations that
-included environmental heterogeneity (see Appendix 4 for confidence
-interval coverage across a range of \(\beta_{ij}\) values). The null
-model's Type I error rates were {[}{[}0.30{]}{]} and {[}{[}0.50{]}{]}
-for the constant and heterogeneous landscapes, respectively---far higher
-than the nominal 0.05 rate. Figure 4C shows, across a range of true
-interaction strengths, the probability that the null model or the Markov
-network will predict the wrong sign of the interaction with 95\%
-confidence. The null model makes such errors more than {[}{[}5{]}{]}
-times as often as the Markov network, even though it only rejects the
-null hypothesis {[}{[}twice{]}{]} as often overall (Appendix 4). The
-Markov network's errors were also more concentrated around 0, so it
-never misclassified strong interactions like the null model did.
+on the simulation type: 0.02 for simulations that matched the model's
+assumptions, versus 0.14 for simulations that included environmental
+heterogeneity (see Appendix 4 for confidence interval coverage across a
+range of \(\beta_{ij}\) values). The null model's Type I error rates
+were 0.31 and 0.53 for the constant and heterogeneous landscapes,
+respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
+across a range of true interaction strengths, the probability that the
+null model or the Markov network will predict the wrong sign of the
+interaction with 95\% confidence. The null model makes such errors more
+than 5 times as often as the Markov network, even though it only rejects
+the null hypothesis twice as often overall (Appendix 4). The Markov
+network's errors were also more concentrated around 0, so it never
+misclassified strong interactions like the null model did (Figure 4C).
 
@@
 example, the states of the interaction network can be modeled as a
-function of the local abiotic environment (Lee and Hastie 2012,
-{\textbf{???}}), which would lead to a better understanding of the
-interplay between biotic and abiotic effects on community structure.
+function of the local abiotic environment (Lee and Hastie 2012, Harris
+2016), which would lead to a better understanding of the interplay
+between biotic and abiotic effects on community structure.
 Alternatively, models could allow one species to alter the relationship
@@
 
+\hyperdef{}{ref-harrisux5festimatingux5f2016}{\label{ref-harrisux5festimatingux5f2016}}
+Harris, D. J. 2016. Estimating species interactions in large,
+abiotically structured communities using markov networks and stochastic
+approximation. bioRxiv.
+
 \hyperdef{}{ref-harrisux5foccupancyux5f2011}{\label{ref-harrisux5foccupancyux5f2011}}

### 2016-02-22 Harris_2015/show-results.Rmd
more manuscript and appendices

@@
 
+# A: Import packages and data
+
 ```{r, message = FALSE}
 library(dplyr)
+library(magrittr)
 library(mgcv)
@@
 
-## Import the results from Appendix 4:
+## Import the results from Appendix 3:
 
@@
 
+# B: Compare model estimates
+
 ## Calculate model performance (R-squared):
@@
 
-#### Calculate average R-squared across methods and simulation types
+#### Average the R-squared values across methods and simulation types
 
@@
 
-## Estimate R-squared uncertainty:
+## Estimate uncertainty in mean R-squared:
 
@@
 
+## Plot estimates versus "true" values across methods and lanscape sizes
+
+The gray line indicates the 1:1 line, where models recover an unbiased estimate
+of the "true" $\beta$ coefficients.  The GLM and Markov network methods estimate
+the coefficients on the original scale, and approach unbiasedness as the number
+of sites increases. The four methods based on (partial) correlations are
+constrained to return values between -1 and 1, and do not return values on the
+original scale.  The covariance and partial covariance (which are not scaled in
+this way) may avoid this problem in future analyses (Loh and Wainwright 2013; 
+see main text for full citation).
+
+Note that the null model (Pairs) is not included in these graphs, as its
+test statistic depends strongly on the size of the landscape and fills a range 
+that would not fit on these graphs.
+
+```{r, dev="png", dpi=500, cache = TRUE}
+ggplot(filter(x, simulation_type == "no_env", method != "null"),
+       aes(x = truth, y = estimate, color = factor(n_sites))) + 
+  facet_wrap(~method) +
+  geom_point(alpha = 0.01) +
+  geom_abline(intercept = 0, slope = 1, size = 2, col = "darkgray") + 
+  geom_smooth(method = "lm", fullrange = FALSE) + 
+  coord_equal() + 
+  scale_color_discrete(name = "n_sites")
+```
+
+# C: Inferential statistics
+
 ## Identify statistical significance for the Markov network and Pairs
@@
 markov_summary$sig_neg = markov_summary$upper < 0
-
-confidently_right = function(data){
-  (data$sig_pos & data$truth > 0) | (data$sig_neg & data$truth < 0)
-}
-
-confidently_wrong = function(data){
-  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
-}
 ```
 
-## Create Figure 4:
+The `error_smoother` is a function that fits a generalized additive model to 
+estimate the probability that a model estimate will match some criterion (e.g. 
+statistical significance) for some parameter as a function of that parameter's
+"true" value.  This function is used in Figure 4C and in calculating Type I
+error rates below.
 
-```{r, cache = TRUE}
+```{r}
 
@@
 
-plotfun = function(..., add){
-  if(add){
-    lines(...)
-  }else{
-    plot(...)
-  }
-}
-
-error_smoother = function(data, f){
+error_smoother = function(data, f, values = truth_seq){
   predict(
@@
     ),
-    data.frame(truth = truth_seq),
+    data.frame(truth = values),
     type = "response"
@@
 }
+```
 
-p_right_pairs = error_smoother(pairs_summary, confidently_right)
-p_wrong_pairs = error_smoother(pairs_summary, confidently_wrong)
+## Create Figure 4
 
-p_right_markov = error_smoother(markov_summary, confidently_right)
-p_wrong_markov = error_smoother(markov_summary, confidently_wrong)
+Panels A and B plot the point estimates or test statistics for different models 
+against one another.
 
-par(mfrow = c(1, 2))
-plot(truth_seq, p_right_pairs + p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
-lines(truth_seq, p_right_markov + p_wrong_markov, col = 2)
-abline(h = 0.05)
+Panel C shows the probability of rejecting the null hypothesis (of no 
+interaction between two species) in the opposite direction of the true value
+(see main text).
 
-plot(truth_seq, p_wrong_pairs, type = "l", ylim = c(0, 1), yaxs = "i")
-lines(truth_seq, p_wrong_markov, col = 2)
-par(mfrow = c(1, 1))
-
-y_markov = error_smoother(markov_summary, confidently_wrong)
+```{r, cache = TRUE}
 
@@
 par(mfrow = c(3, 1))
+
+# Calculate p(confidently wrong) for each model as a function of
+# the true values
+confidently_wrong = function(data){
+  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
+}
+y_markov = error_smoother(markov_summary, confidently_wrong)
+y_pairs = error_smoother(pairs_summary, confidently_wrong)
 
@@
 # R-squared for null versus correlation
-round(summary(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))$r.squared, 2)
+round(
+  summary(lm(null ~ I(correlation*sqrt(n_sites)), data = spread_estimates))$r.squared, 
+  2
+)
 
@@
 
-plotfun(
+plot(
   truth_seq,
@@
   col = 2,
-  add = FALSE,
   ylim = c(0, .4),
@@
       font = 2, line = 1.2, cex = .9) 
-plotfun(truth_seq, y_markov, add = TRUE, lwd = 2)
+lines(truth_seq, y_markov, lwd = 2)
 legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
@@
 
+#### P(Pairs confidently wrong):
 ```{r}
-# P(Pairs confidently wrong)
 with(pairs_summary,  mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
-# P(Markov network confidently wrong)
+```
+
+#### P(Markov network confidently wrong)
+```{r}
 with(markov_summary, mean((sig_neg & truth > 0) | (sig_pos & truth < 0)))
+```
 
-# P(Pairs rejects null)
+#### P(Pairs rejects null)
+```{r}
 with(pairs_summary,  mean(sig_neg | sig_pos))
+```
 
-# P(Markov network rejects null)
+#### P(Markov network rejects null)
+```{r}
 with(markov_summary,  mean(sig_neg | sig_pos))
+```
 
-# P(reject null | small interaction) for Makrov network
-# (approximate Type I error rates)
+## Type I error rates:
+
+Type I error rates (probability of rejecting the null hypothesis given a 
+"true" interaction strength of zero, according to the spline model defined by
+`error_smoother` above). The smoother focuses on the 
+`r markov_summary %>% summarize(mean(abs(truth) < 1)) %>% round(., 2) %>% multiply_by(100)`%
+of the data points whose "true" values fall between -1 and 1, because its 
+estimates near 0 are more accurate when they aren't affected by distant data 
+points.
+
+#### Markov network Type I error rates
+```{r}
 markov_summary %>% 
+  filter(abs(truth) < 1) %>% 
   group_by(simulation_type) %>% 
-  filter(abs(truth) < .1) %>% 
-  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
-  kable(digits = 3)
-
-# P(reject null | small interaction) for Pairs
-# (approximate Type I error rates)
-pairs_summary %>%
-  group_by(simulation_type) %>% 
-  filter(abs(truth) < .1) %>% 
-  summarize(Type_1_error_rate = mean(sig_neg | sig_pos)) %>%
-  kable(digits = 3)
-
-# Confidence interval coverage across all values
-# What's the probability that any "true" value falls inside the 95% CI?
-markov_summary %>% 
-  group_by(simulation_type) %>% 
-  summarize(mean(truth > lower & truth < upper)) %>%
+  do(data.frame(
+    `Type I error rate` = 
+      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
+  )) %>%
   kable(digits = 3)
@@
 
-## Plot root mean square error versus "true" beta
-
-```{r, cache = TRUE}
-# Exclude the "abundance" simulations because the interpretation of $\beta$ 
-# is different
-rmses = markov_summary %>% 
-  filter(simulation_type != "abund") %>% 
-  mutate(rmse = sqrt((truth - estimate)^2))
-
-ggplot(rmses, aes(x = truth, y = rmse)) + 
-  geom_point(alpha = .2, size = 1) + 
-  geom_smooth() + 
-  theme_bw() + 
-  coord_cartesian(ylim = c(0, max(rmses$rmse)))
+#### Null model Type I error rates
+```{r}
+pairs_summary %>% 
+  filter(abs(truth) < 1) %>% 
+  group_by(simulation_type) %>% 
+  do(data.frame(
+    `Type I error rate` = 
+      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
+  )) %>%
+  kable(digits = 3)
 ```
 
-## Plot confidence interval coverage versus true $\beta$ value
+## Plot Markov network confidence interval coverage versus true $\beta$ value
 
-```{r, cache = TRUE}
-# Coverage versus true beta value.  The top and bottom 0.5%
-# of the distribution have been omitted to prevent bad behavior
-# by the smoother in the tails.
-markov_summary %>%
+The top and bottom 0.5% of the distribution have been omitted to prevent bad 
+behavior by the smoother in the tails. The horizontal red line shows the nominal
+95% coverage rate and the black vertical line marks where the "true" value
+of $\beta$ was zero. The Type I error rates calculated above for the Markov 
+network closely match (one minus) the values of these curves at zero.
+
+```{r}
+coverage_data = markov_summary %>%
   filter(percent_rank(truth) > .005 & percent_rank(truth) < .995) %>%
   mutate(covered = truth > lower & truth < upper) %>%
-  ggplot(aes(x = truth, y = as.integer(covered))) + 
+  mutate(simulation_type = factor(simulation_type, c("no_env", "env", "abund")))
+
+ggplot(coverage_data, aes(x = truth, y = as.integer(covered))) + 
   facet_grid(~simulation_type) + 
-  geom_smooth(method = gam, formula = y ~ s(x), method.args = list(family = "binomial")) + 
+  geom_smooth(method = gam, formula = y ~ s(x), method.args = list(family = binomial)) + 
   theme_bw() + 
-  coord_cartesian(ylim = c(0, 1)) + 
+  coord_cartesian(ylim = c(0, 1), expand = FALSE) + 
   geom_vline(xintercept = 0) + 

### 2016-02-22 Harris_2015/show-results.tex
more manuscript and appendices

@@
 
+\section{A: Import packages and data}\label{a-import-packages-and-data}
+
 \begin{Shaded}
@@
 \KeywordTok{library}\NormalTok{(dplyr)}
+\KeywordTok{library}\NormalTok{(magrittr)}
 \KeywordTok{library}\NormalTok{(mgcv)}
@@
 \subsection{Import the results from Appendix
-4:}\label{import-the-results-from-appendix-4}
+3:}\label{import-the-results-from-appendix-3}
 
@@
 \CommentTok{# Sort a vector of alphanumeric strings by the numeric component}
-\CommentTok{# as if they were integers.  For example, V20 is larger than V12,}
-\CommentTok{# even though V12 comes first alphabetically}
+\CommentTok{# as if they were integers.  For example, V20 should be larger }
+\CommentTok{# than V12, even though V12 comes first alphabetically}
 \NormalTok{alnum_sort =}\StringTok{ }\NormalTok{function(x)\{}
@@
 
-\subsection{Calculate model
-performance:}\label{calculate-model-performance}
+\section{B: Compare model estimates}\label{b-compare-model-estimates}
+
+\subsection{Calculate model performance
+(R-squared):}\label{calculate-model-performance-r-squared}
+
+For each combination of \texttt{method} and \texttt{simulation\_type},
+fit a linear model. Then, for each combination of \texttt{method},
+\texttt{simulation\_type}, and \texttt{n\_sites}, report the proportion
+of variance explained by the corresponding linear model.
 
@@
 
+\CommentTok{# Set the ordering of the data frame based on R-squared}
 \NormalTok{result_summary$method =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(result_summary$method, -result_summary$r2)}
-
 \NormalTok{result_summary$simulation_type =}\StringTok{ }\KeywordTok{reorder}\NormalTok{(}
@@
 
+\CommentTok{# Add a column that includes method name and its mean R-squared,}
+\CommentTok{# for plotting purposes below}
 \NormalTok{result_summary =}\StringTok{ }\NormalTok{result_summary %>%}\StringTok{ }
@@
 
+\CommentTok{# Rename the simulation types for clearer graph labels}
 \NormalTok{result_summary$simulation_type_long =}\StringTok{ }\NormalTok{plyr::}\KeywordTok{revalue}\NormalTok{(}
@@
 
-\paragraph{Calculate average R-squared across methods and simulation
-types}\label{calculate-average-r-squared-across-methods-and-simulation-types}
+\paragraph{Average the R-squared values across methods and simulation
+types}\label{average-the-r-squared-values-across-methods-and-simulation-types}
 
@@
 
-\subsection{Save the R-squared results as Figure
-3}\label{save-the-r-squared-results-as-figure-3}
+\subsection{Save the finer-grained R-squared results as Figure
+3}\label{save-the-finer-grained-r-squared-results-as-figure-3}
 
@@
 \StringTok{  }\KeywordTok{geom_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\CommentTok{#scale_shape_manual(values = c(21, 25, 15, 18)) + }
 \StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(-.}\DecValTok{01}\NormalTok{, }\FloatTok{0.76}\NormalTok{)) +}\StringTok{ }
@@
 
-\subsection{Estimate R-squared
-uncertainty:}\label{estimate-r-squared-uncertainty}
+\subsection{Estimate uncertainty in mean
+R-squared:}\label{estimate-uncertainty-in-mean-r-squared}
 
@@
     \DataTypeTok{data =} \NormalTok{landscape_estimates}
-  \NormalTok{)}
+  \NormalTok{),}
+  \DataTypeTok{correlation =} \OtherTok{FALSE}
 \NormalTok{)}
@@
 ## simulation_typeno_env      1.795e-01  7.402e-03   24.25
-## 
-## Correlation of Fixed Effects:
-##             (Intr) mthdcr mthGLM mthdMn mthdnl mtBysC mthdpc n_sits smltn_
-## methdcrrltn -0.494                                                        
-## methodGLM   -0.494  0.500                                                 
-## mthdMrkvntw -0.494  0.500  0.500                                          
-## methodnull  -0.494  0.500  0.500  0.500                                   
-## mthdprtBysC -0.494  0.500  0.500  0.500  0.500                            
-## mthdprtlcrr -0.494  0.500  0.500  0.500  0.500  0.500                     
-## n_sites     -0.228  0.000  0.000  0.000  0.000  0.000  0.000              
-## smltn_typnv -0.515  0.000  0.000  0.000  0.000  0.000  0.000  0.000       
-## smltn_typn_ -0.515  0.000  0.000  0.000  0.000  0.000  0.000  0.000  0.500
 \end{verbatim}
+
+\subsection{\texorpdfstring{Plot estimates versus ``true'' values across
+methods and lanscape
+sizes}{Plot estimates versus true values across methods and lanscape sizes}}\label{plot-estimates-versus-true-values-across-methods-and-lanscape-sizes}
+
+The gray line indicates the 1:1 line, where models recover an unbiased
+estimate of the ``true'' \(\beta\) coefficients. The GLM and Markov
+network methods estimate the coefficients on the original scale, and
+approach unbiasedness as the number of sites increases. The four methods
+based on (partial) correlations are constrained to return values between
+-1 and 1, and do not return values on the original scale. The covariance
+and partial covariance (which are not scaled in this way) may avoid this
+problem in future analyses (Loh and Wainwright 2013; see main text for
+full citation).
+
+Note that the null model (Pairs) is not included in these graphs, as its
+test statistic depends strongly on the size of the landscape and fills a
+range that would not fit on these graphs.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{filter}\NormalTok{(x, simulation_type ==}\StringTok{ "no_env"}\NormalTok{, method !=}\StringTok{ "null"}\NormalTok{),}
+       \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{estimate, }\DataTypeTok{color =} \KeywordTok{factor}\NormalTok{(n_sites))) +}\StringTok{ }
+\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(~method) +}
+\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.01}\NormalTok{) +}
+\StringTok{  }\KeywordTok{geom_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{size =} \DecValTok{2}\NormalTok{, }\DataTypeTok{col =} \StringTok{"darkgray"}\NormalTok{) +}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \StringTok{"lm"}\NormalTok{, }\DataTypeTok{fullrange =} \OtherTok{FALSE}\NormalTok{) +}\StringTok{ }
+\StringTok{  }\KeywordTok{coord_equal}\NormalTok{() +}\StringTok{ }
+\StringTok{  }\KeywordTok{scale_color_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"n_sites"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-8-1.png}
+
+\section{C: Inferential statistics}\label{c-inferential-statistics}
 
@@
 Pairs}\label{identify-statistical-significance-for-the-markov-network-and-pairs}
+
+For Pairs, compare the Z-scores with Gaussian quantiles for 95\%
+coverage; for the Markov network, use the approximate confidence
+intervals estimated in Appendix 3.
 
@@
 \CommentTok{# Pairs's Z score is based on C-scores, which are positive when species are }
-\CommentTok{# disaggregated.  So significantly positive interactions == negative estimates}
+\CommentTok{# disaggregated.  So significantly negative scores imply positive interactions}
+\CommentTok{# and vice versa}
 \NormalTok{pairs_summary$sig_pos =}\StringTok{ }\NormalTok{pairs_summary$estimate <}\StringTok{ }\KeywordTok{qnorm}\NormalTok{(.}\DecValTok{025}\NormalTok{)}
@@
 
-\subsection{Create Figure 4:}\label{create-figure-4}
+The \texttt{error\_smoother} is a function that fits a generalized
+additive model to estimate the probability that a model estimate will
+match some criterion (e.g. statistical significance) for some parameter
+as a function of that parameter's ``true'' value. This function is used
+in Figure 4C and in calculating Type I error rates below.
 
@@
 \begin{Highlighting}[]
-\CommentTok{# Function for smoothing the error rates with a generalized additive model}
-\CommentTok{# and plotting the results}
-\NormalTok{my_geom_smooth =}\StringTok{ }\NormalTok{function(data, color, ...)\{}
-  \KeywordTok{geom_smooth}\NormalTok{(}
-    \DataTypeTok{data =} \NormalTok{data,}
-    \KeywordTok{aes}\NormalTok{(}
-      \DataTypeTok{x =} \NormalTok{truth, }
-      \DataTypeTok{y =} \KeywordTok{as.integer}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)),}
-      \DataTypeTok{color =} \NormalTok{color}
-    \NormalTok{),}
-    \DataTypeTok{method =} \NormalTok{gam, }
-    \DataTypeTok{family =} \NormalTok{binomial, }
-    \DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x),}
-    \DataTypeTok{se =} \OtherTok{FALSE}\NormalTok{,}
-    \DataTypeTok{n =} \DecValTok{1024}\NormalTok{,}
-    \NormalTok{...}
-  \NormalTok{)}
-\NormalTok{\}}
-
 \NormalTok{truth_seq =}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(markov_summary$truth), }\KeywordTok{max}\NormalTok{(markov_summary$truth), }\DataTypeTok{length =} \DecValTok{1000}\NormalTok{)}
 
-\NormalTok{plotfun =}\StringTok{ }\NormalTok{function(..., add)\{}
-  \NormalTok{if(add)\{}
-    \KeywordTok{lines}\NormalTok{(...)}
-  \NormalTok{\}else\{}
-    \KeywordTok{plot}\NormalTok{(...)}
-  \NormalTok{\}}
-\NormalTok{\}}
-
-\NormalTok{error_smoother =}\StringTok{ }\NormalTok{function(data)\{}
+\NormalTok{error_smoother =}\StringTok{ }\NormalTok{function(data, f, }\DataTypeTok{values =} \NormalTok{truth_seq)\{}
   \KeywordTok{predict}\NormalTok{(}
     \KeywordTok{gam}\NormalTok{(}
-      \KeywordTok{I}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)) ~}\StringTok{ }\KeywordTok{s}\NormalTok{(truth),}
+      \KeywordTok{f}\NormalTok{(data) ~}\StringTok{ }\KeywordTok{s}\NormalTok{(truth),}
       \DataTypeTok{data =} \NormalTok{data,}
@@
     \NormalTok{),}
-    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{truth =} \NormalTok{truth_seq),}
+    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{truth =} \NormalTok{values),}
     \DataTypeTok{type =} \StringTok{"response"}
@@
 \NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
 
-\NormalTok{y_pairs =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(pairs_summary)}
-\NormalTok{y_markov =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(markov_summary)}
+\subsection{Create Figure 4}\label{create-figure-4}
 
+Panels A and B plot the point estimates or test statistics for different
+models against one another.
+
+Panel C shows the probability of rejecting the null hypothesis (of no
+interaction between two species) in the opposite direction of the true
+value (see main text).
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \KeywordTok{pdf}\NormalTok{(}\StringTok{"manuscript-materials/figures/error_rates.pdf"}\NormalTok{, }\DataTypeTok{height =} \FloatTok{8.5}\NormalTok{, }\DataTypeTok{width =} \FloatTok{8.5}\NormalTok{/}\DecValTok{3}\NormalTok{)}
 \KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{3}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+
+\CommentTok{# Calculate p(confidently wrong) for each model as a function of}
+\CommentTok{# the true values}
+\NormalTok{confidently_wrong =}\StringTok{ }\NormalTok{function(data)\{}
+  \NormalTok{(data$sig_pos &}\StringTok{ }\NormalTok{data$truth <}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(data$sig_neg &}\StringTok{ }\NormalTok{data$truth >}\StringTok{ }\DecValTok{0}\NormalTok{)}
+\NormalTok{\}}
+\NormalTok{y_markov =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(markov_summary, confidently_wrong)}
+\NormalTok{y_pairs =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(pairs_summary, confidently_wrong)}
 
@@
 \CommentTok{# R-squared for null versus correlation}
-\KeywordTok{round}\NormalTok{(}\KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(null ~}\StringTok{ }\KeywordTok{I}\NormalTok{(correlation*}\KeywordTok{sqrt}\NormalTok{(n_sites)), }\DataTypeTok{data =} \NormalTok{spread_estimates))$r.squared, }\DecValTok{2}\NormalTok{)}
+\KeywordTok{round}\NormalTok{(}
+  \KeywordTok{summary}\NormalTok{(}\KeywordTok{lm}\NormalTok{(null ~}\StringTok{ }\KeywordTok{I}\NormalTok{(correlation*}\KeywordTok{sqrt}\NormalTok{(n_sites)), }\DataTypeTok{data =} \NormalTok{spread_estimates))$r.squared, }
+  \DecValTok{2}
+\NormalTok{)}
 \end{Highlighting}
@@
 
-\KeywordTok{plotfun}\NormalTok{(}
+\KeywordTok{plot}\NormalTok{(}
   \NormalTok{truth_seq,}
@@
   \DataTypeTok{col =} \DecValTok{2}\NormalTok{,}
-  \DataTypeTok{add =} \OtherTok{FALSE}\NormalTok{,}
   \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{4}\NormalTok{),}
@@
       \DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
-\KeywordTok{plotfun}\NormalTok{(truth_seq, y_markov, }\DataTypeTok{add =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{lines}\NormalTok{(truth_seq, y_markov, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
 \KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Null model"}\NormalTok{, }\StringTok{"Markov network"}\NormalTok{), }
@@
 
+\paragraph{P(Pairs confidently wrong):}\label{ppairs-confidently-wrong}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(Pairs confidently wrong)}
 \KeywordTok{with}\NormalTok{(pairs_summary,  }\KeywordTok{mean}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)))}
@@
 
+\paragraph{P(Markov network confidently
+wrong)}\label{pmarkov-network-confidently-wrong}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(Markov network confidently wrong)}
 \KeywordTok{with}\NormalTok{(markov_summary, }\KeywordTok{mean}\NormalTok{((sig_neg &}\StringTok{ }\NormalTok{truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(sig_pos &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\DecValTok{0}\NormalTok{)))}
@@
 
+\paragraph{P(Pairs rejects null)}\label{ppairs-rejects-null}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(Pairs rejects null)}
 \KeywordTok{with}\NormalTok{(pairs_summary,  }\KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos))}
@@
 
+\paragraph{P(Markov network rejects
+null)}\label{pmarkov-network-rejects-null}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(Markov network rejects null)}
 \KeywordTok{with}\NormalTok{(markov_summary,  }\KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos))}
@@
 
+\subsection{Type I error rates:}\label{type-i-error-rates}
+
+Type I error rates (probability of rejecting the null hypothesis given a
+``true'' interaction strength of zero, according to the spline model
+defined by \texttt{error\_smoother} above). The smoother focuses on the
+64\% of the data points whose ``true'' values fall between -1 and 1,
+because its estimates near 0 are more accurate when they aren't affected
+by distant data points.
+
+\paragraph{Markov network Type I error
+rates}\label{markov-network-type-i-error-rates}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(reject null | small interaction) for Makrov network}
-\CommentTok{# (approximate Type I error rates)}
-\NormalTok{markov_summary %>%}
+\NormalTok{markov_summary %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
 \StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{Type_1_error_rate =} \KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos)) %>%}
+\StringTok{  }\KeywordTok{do}\NormalTok{(}\KeywordTok{data.frame}\NormalTok{(}
+    \StringTok{`}\DataTypeTok{Type I error rate}\StringTok{`} \NormalTok{=}\StringTok{ }
+\StringTok{      }\KeywordTok{error_smoother}\NormalTok{(., function(x)\{x$sig_pos |}\StringTok{ }\NormalTok{x$sig_neg\}, }\DataTypeTok{values =} \DecValTok{0}\NormalTok{)}
+  \NormalTok{)) %>%}
 \StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
@@
 \toprule
-simulation\_type & Type\_1\_error\_rate\tabularnewline
+simulation\_type & Type.I.error.rate\tabularnewline
 \midrule
 \endhead
-abund & 0.220\tabularnewline
-env & 0.136\tabularnewline
-no\_env & 0.019\tabularnewline
+abund & 0.258\tabularnewline
+env & 0.142\tabularnewline
+no\_env & 0.024\tabularnewline
 \bottomrule
@@
 
+\paragraph{Null model Type I error
+rates}\label{null-model-type-i-error-rates}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{# P(reject null | small interaction) for Pairs}
-\CommentTok{# (approximate Type I error rates)}
-\NormalTok{pairs_summary %>%}
+\NormalTok{pairs_summary %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
 \StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{1}\NormalTok{) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\DataTypeTok{Type_1_error_rate =} \KeywordTok{mean}\NormalTok{(sig_neg |}\StringTok{ }\NormalTok{sig_pos)) %>%}
+\StringTok{  }\KeywordTok{do}\NormalTok{(}\KeywordTok{data.frame}\NormalTok{(}
+    \StringTok{`}\DataTypeTok{Type I error rate}\StringTok{`} \NormalTok{=}\StringTok{ }
+\StringTok{      }\KeywordTok{error_smoother}\NormalTok{(., function(x)\{x$sig_pos |}\StringTok{ }\NormalTok{x$sig_neg\}, }\DataTypeTok{values =} \DecValTok{0}\NormalTok{)}
+  \NormalTok{)) %>%}
 \StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
@@
 \toprule
-simulation\_type & Type\_1\_error\_rate\tabularnewline
+simulation\_type & Type.I.error.rate\tabularnewline
 \midrule
 \endhead
-abund & 0.575\tabularnewline
-env & 0.501\tabularnewline
-no\_env & 0.297\tabularnewline
+abund & 0.597\tabularnewline
+env & 0.530\tabularnewline
+no\_env & 0.308\tabularnewline
 \bottomrule
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Confidence interval coverage across all values}
-\CommentTok{# What's the probability that any "true" value falls inside the 95% CI?}
-\NormalTok{markov_summary %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(simulation_type) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{summarize}\NormalTok{(}\KeywordTok{mean}\NormalTok{(truth >}\StringTok{ }\NormalTok{lower &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\NormalTok{upper)) %>%}
-\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{digits =} \DecValTok{3}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\subsection{\texorpdfstring{Plot Markov network confidence interval
+coverage versus true \(\beta\)
+value}{Plot Markov network confidence interval coverage versus true \textbackslash{}beta value}}\label{plot-markov-network-confidence-interval-coverage-versus-true-beta-value}
 
-\begin{longtable}[c]{@{}lr@{}}
-\toprule
-simulation\_type & mean(truth \textgreater{} lower \& truth \textless{}
-upper)\tabularnewline
-\midrule
-\endhead
-abund & 0.868\tabularnewline
-env & 0.857\tabularnewline
-no\_env & 0.975\tabularnewline
-\bottomrule
-\end{longtable}
-
-\subsection{\texorpdfstring{Plot root mean square error versus ``true''
-beta}{Plot root mean square error versus true beta}}\label{plot-root-mean-square-error-versus-true-beta}
+The top and bottom 0.5\% of the distribution have been omitted to
+prevent bad behavior by the smoother in the tails. The horizontal red
+line shows the nominal 95\% coverage rate and the black vertical line
+marks where the ``true'' value of \(\beta\) was zero. The Type I error
+rates calculated above for the Markov network closely match (one minus)
+the values of these curves at zero.
 
@@
 \begin{Highlighting}[]
-\CommentTok{# Exclude the "abundance" simulations because the interpretation of $\textbackslash{}beta$ }
-\CommentTok{# is different}
-\NormalTok{rmses =}\StringTok{ }\NormalTok{markov_summary %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(simulation_type !=}\StringTok{ "abund"}\NormalTok{) %>%}\StringTok{ }
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{rmse =} \KeywordTok{sqrt}\NormalTok{((truth -}\StringTok{ }\NormalTok{estimate)^}\DecValTok{2}\NormalTok{))}
-
-\KeywordTok{ggplot}\NormalTok{(rmses, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{rmse)) +}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \NormalTok{.}\DecValTok{2}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{max}\NormalTok{(rmses$rmse)))}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{show-results_files/figure-latex/unnamed-chunk-11-1.pdf}
-
-\subsection{\texorpdfstring{Plot confidence interval coverage versus
-true \(\beta\)
-value}{Plot confidence interval coverage versus true \textbackslash{}beta value}}\label{plot-confidence-interval-coverage-versus-true-beta-value}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Coverage versus true beta value.  The top and bottom 0.5%}
-\CommentTok{# of the distribution have been omitted to prevent bad behavior}
-\CommentTok{# by the smoother in the tails.}
-\NormalTok{markov_summary %>%}
+\NormalTok{coverage_data =}\StringTok{ }\NormalTok{markov_summary %>%}
 \StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{percent_rank}\NormalTok{(truth) >}\StringTok{ }\NormalTok{.}\DecValTok{005} \NormalTok{&}\StringTok{ }\KeywordTok{percent_rank}\NormalTok{(truth) <}\StringTok{ }\NormalTok{.}\DecValTok{995}\NormalTok{) %>%}
 \StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{covered =} \NormalTok{truth >}\StringTok{ }\NormalTok{lower &}\StringTok{ }\NormalTok{truth <}\StringTok{ }\NormalTok{upper) %>%}
-\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \KeywordTok{as.integer}\NormalTok{(covered))) +}\StringTok{ }
+\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{simulation_type =} \KeywordTok{factor}\NormalTok{(simulation_type, }\KeywordTok{c}\NormalTok{(}\StringTok{"no_env"}\NormalTok{, }\StringTok{"env"}\NormalTok{, }\StringTok{"abund"}\NormalTok{)))}
+
+\KeywordTok{ggplot}\NormalTok{(coverage_data, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \KeywordTok{as.integer}\NormalTok{(covered))) +}\StringTok{ }
 \StringTok{  }\KeywordTok{facet_grid}\NormalTok{(~simulation_type) +}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \NormalTok{gam, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x), }\DataTypeTok{method.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{family =} \StringTok{"binomial"}\NormalTok{)) +}\StringTok{ }
+\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \NormalTok{gam, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\KeywordTok{s}\NormalTok{(x), }\DataTypeTok{method.args =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{family =} \NormalTok{binomial)) +}\StringTok{ }
 \StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{)) +}\StringTok{ }
+\StringTok{  }\KeywordTok{coord_cartesian}\NormalTok{(}\DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{expand =} \OtherTok{FALSE}\NormalTok{) +}\StringTok{ }
 \StringTok{  }\KeywordTok{geom_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{) +}\StringTok{ }
@@
 
-\includegraphics{show-results_files/figure-latex/unnamed-chunk-12-1.pdf}
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-18-1.pdf}
 

### 2016-02-22 Harris_2015/manuscript-materials/draft.Rmd
cleanup

@@
 interactions or abiotic factors largely overpowered them. Two computationally 
-efficient approximations, based on controlling for indirect effects with partial
+efficient approximations, which controlled for indirect effects with partial
 correlations or generalized linear models, also performed well. Null models 
@@
 and for estimating them from observed presence-absence matrices. Markov networks
-have been used in diverse scientific fields in similar contexts for decades, from
+have been used in diverse scientific fields for decades, from
 physics [where nearby particles interact magnetically; @cipra_introduction_1987]
@@
 co-occur and few negatively-associated pairs do (just as an ecologist might
-expect). When all else is *not* equal (e.g. Figure 1, where the presence of one
-competitor is associated with release from another competitor), then predicting
-species' overall co-occurrence rates would require normalizing the relative 
-probabilities of possible $\vec{y}$ vectors so that they sum to one (Figure 2B).
+expect).
 
@@
 indicating which species are present at each site on the landscape). When
-Equation 1 can be normalized, one can find exact maximum likelihood estimates for
-$\alpha$ and $\beta$ using numerical optimization. Fully-observed networks like 
-the ones considered here have unimodal likelihood surfaces 
-[@murphy_machine_2012], so these estimates will represent the global optimum.
+Equation 1 can be normalized (Figure 2B), one can find exact maximum likelihood
+estimates for $\alpha$ and $\beta$ by numerically optimizing
+$p(\vec{y}|\alpha, \beta)$. Fully-observed networks like the ones considered 
+here have unimodal likelihood surfaces [@murphy_machine_2012], so these
+estimates will represent the global optimum.
 
@@
 can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
-they want to fit a Markov network model. For the analyses presented here, I used 
-the rosalia package [@harris_rosalia_2015] for the R programming language
+they want to fit a Markov network model. For the analyses presented here, where
+the number of species did not exceed 20, approximations were not necessary. Instead,
+I used the rosalia package [@harris_rosalia_2015] for the R programming language
 [@r_core_team_r_2015] to calculate $p(\vec{y}; \alpha, \beta)$ and its gradients
@@
 Here, the $\beta$ coefficients were calculated as above, but each species'
-$\alpha$ value depended linearly on two normally-distributed environmental 
-factors, which were drawn from independent Gaussians for each site. Once the 
-local $\alpha$ values were calculated, independent Gibbs samplers determined 
-the species composition for each site based on Equation 1.
+$\alpha$ value depended linearly on two environmental factors, which were drawn 
+from independent Gaussians for each site. Once the local $\alpha$ values were
+calculated, independent Gibbs samplers determined the species composition for 
+each site based on Equation 1.
 
 In the final set of landscapes, I simulated each species' abundance (instead of
-just their presence/absence); here, interactions between species occurred on a 
-per-capita basis. To prevent runaway mutualisms leading to infinite abundance,
-all interaction coefficients were negative in these simulations. Good 
-performance on these landscapes would indicate some robustness to the
-mechanistic details of species interactions.
+just presence/absence); furthermore, interactions between species occurred on a 
+per-capita basis in these simulations. To prevent runaway mutualisms leading to
+infinite abundance, all interaction coefficients were negative in these 
+simulations. Good performance on these landscapes would indicate some robustness 
+to the mechanistic details of species interactions.
 
@@
 weakly-informative regularizer (equivalent to a logistic prior with location 0
-and scale 2) ensured that the estimates were always finite.
+and scale 2) ensured that the estimates were always finite (Appendix 2).
 
@@
 The next method used the Pairs software described in @gotelli_empirical_2009.
-This program simulating new landscapes from a null model that retains the row 
+This program simulates new landscapes from a null model that retains the row 
 and column sums of the original matrix [@strong_ecological_1984] and calculates
@@
 though it only rejects the null hypothesis twice as often overall (Appendix
-4). The Markov network's errors were also more concentrated around 0, so it 
+4). The Markov network's errors were also more concentrated around 0, and it 
 never misclassified strong interactions like the null model did (Figure 4C).
@@
 numbers of sites; @gotelli_empirical_2009]. The difference in accuracy may be
-larger for real data sets than for the simulated landscapes in Figure 3,
-however; linear approximations to the Markov network make larger errors when
+even larger for real data sets than for the simulated landscapes in Figure 3; 
+linear approximations to the Markov network make larger errors when
 the interaction matrix is structured [e.g. due to guilds or trophic levels;

### 2016-02-22 Harris_2015/manuscript-materials/draft.tex
cleanup

@@
 abiotic factors largely overpowered them. Two computationally efficient
-approximations, based on controlling for indirect effects with partial
+approximations, which controlled for indirect effects with partial
 correlations or generalized linear models, also performed well. Null
@@
 from observed presence-absence matrices. Markov networks have been used
-in diverse scientific fields in similar contexts for decades, from
-physics (where nearby particles interact magnetically; Cipra 1987) to
-spatial statistics (where adjacent grid cells have correlated values;
-Harris 1974, Gelfand et al. 2005). While community ecologists explored
-some related approaches in the 1980's (Whittam and Siegel-Causey 1981),
-they used severe approximations that led to unintelligible results (e.g.
+in diverse scientific fields for decades, from physics (where nearby
+particles interact magnetically; Cipra 1987) to spatial statistics
+(where adjacent grid cells have correlated values; Harris 1974, Gelfand
+et al. 2005). While community ecologists explored some related
+approaches in the 1980's (Whittam and Siegel-Causey 1981), they used
+severe approximations that led to unintelligible results (e.g.
 ``probabilities'' greater than one; Gilpin and Diamond 1982).
@@
 few negatively-associated pairs do (just as an ecologist might expect).
-When all else is \emph{not} equal (e.g.~Figure 1, where the presence of
-one competitor is associated with release from another competitor), then
-predicting species' overall co-occurrence rates would require
-normalizing the relative probabilities of possible \(\vec{y}\) vectors
-so that they sum to one (Figure 2B).
 
@@
 present at each site on the landscape). When Equation 1 can be
-normalized, one can find exact maximum likelihood estimates for
-\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
-networks like the ones considered here have unimodal likelihood surfaces
-(Murphy 2012), so these estimates will represent the global optimum.
+normalized (Figure 2B), one can find exact maximum likelihood estimates
+for \(\alpha\) and \(\beta\) by numerically optimizing
+\(p(\vec{y}|\alpha, \beta)\). Fully-observed networks like the ones
+considered here have unimodal likelihood surfaces (Murphy 2012), so
+these estimates will represent the global optimum.
 
@@
 2016) if they want to fit a Markov network model. For the analyses
-presented here, I used the rosalia package (Harris 2015a) for the R
-programming language (R Core Team 2015) to calculate
-\(p(\vec{y}; \alpha, \beta)\) and its gradients exactly (Murphy 2012);
-the package passes these functions to the ``BFGS'' method in R's
-general-purpose optimizer, which finds values for \(\alpha\) and
+presented here, where the number of species did not exceed 20,
+approximations were not necessary. Instead, I used the rosalia package
+(Harris 2015a) for the R programming language (R Core Team 2015) to
+calculate \(p(\vec{y}; \alpha, \beta)\) and its gradients exactly
+(Murphy 2012); the package passes these functions to the ``BFGS'' method
+in R's general-purpose optimizer, which finds values for \(\alpha\) and
 \(\beta\).
@@
 above, but each species' \(\alpha\) value depended linearly on two
-normally-distributed environmental factors, which were drawn from
-independent Gaussians for each site. Once the local \(\alpha\) values
-were calculated, independent Gibbs samplers determined the species
-composition for each site based on Equation 1.
+environmental factors, which were drawn from independent Gaussians for
+each site. Once the local \(\alpha\) values were calculated, independent
+Gibbs samplers determined the species composition for each site based on
+Equation 1.
 
 In the final set of landscapes, I simulated each species' abundance
-(instead of just their presence/absence); here, interactions between
-species occurred on a per-capita basis. To prevent runaway mutualisms
-leading to infinite abundance, all interaction coefficients were
-negative in these simulations. Good performance on these landscapes
-would indicate some robustness to the mechanistic details of species
-interactions.
+(instead of just presence/absence); furthermore, interactions between
+species occurred on a per-capita basis in these simulations. To prevent
+runaway mutualisms leading to infinite abundance, all interaction
+coefficients were negative in these simulations. Good performance on
+these landscapes would indicate some robustness to the mechanistic
+details of species interactions.
 
@@
 weakly-informative regularizer (equivalent to a logistic prior with
-location 0 and scale 2) ensured that the estimates were always finite.
+location 0 and scale 2) ensured that the estimates were always finite
+(Appendix 2).
 
@@
 The next method used the Pairs software described in Gotelli and Ulrich
-(2009). This program simulating new landscapes from a null model that
+(2009). This program simulates new landscapes from a null model that
 retains the row and column sums of the original matrix (Strong et al.
@@
 the null hypothesis twice as often overall (Appendix 4). The Markov
-network's errors were also more concentrated around 0, so it never
+network's errors were also more concentrated around 0, and it never
 misclassified strong interactions like the null model did (Figure 4C).
@@
 numbers of sites; Gotelli and Ulrich 2009). The difference in accuracy
-may be larger for real data sets than for the simulated landscapes in
-Figure 3, however; linear approximations to the Markov network make
-larger errors when the interaction matrix is structured (e.g.~due to
-guilds or trophic levels; Loh and Wainwright 2013). Similarly, the
-separate generalized linear models for each species can severely overfit
-in some cases (Lee and Hastie 2012). The full Markov network should thus
-be preferred to the approximations when it is computationally tractable.
+may be even larger for real data sets than for the simulated landscapes
+in Figure 3; linear approximations to the Markov network make larger
+errors when the interaction matrix is structured (e.g.~due to guilds or
+trophic levels; Loh and Wainwright 2013). Similarly, the separate
+generalized linear models for each species can severely overfit in some
+cases (Lee and Hastie 2012). The full Markov network should thus be
+preferred to the approximations when it is computationally tractable.
 

### 2016-02-22 Harris_2015/simulation.Rmd
more manuscript and appendices
cleanup

@@
 
-The following function draws random coefficients for a Markov network of a pre-specified size.
+The following function draws random coefficients for a Markov network of a pre-specified size (20 species in these examples).
 
@@
 Gaussian distributions with unit variance and a mean given by `mean_alpha`.
-In the R code at the end of this document, `mean_alpha` was defined as $-1$ for presence-absence simulations and $5$ for the abundance simulations.
+In the R code at the end of this document, `mean_alpha` is defined as $-1$ 
+for presence-absence simulations and $5$ for the abundance simulations.
 
@@
   # a long tail.
-  true_beta_magnitudes = rexp(choose(n_spp, 2))
+  true_beta_magnitudes = rexp(choose(n_spp, 2), rate = 1)
 

### 2016-03-14 Harris_2015/manuscript-materials/draft.Rmd
final before resubmission to Ecology?

@@
 landscapes with known interactions, I evaluated Markov networks and six existing
-approaches. The Markov networks consistently outperformed other methods, 
+approaches. Markov networks consistently outperformed the other methods, 
 correctly isolating direct interactions between species pairs even when indirect
@@
 disentangling the effects of biotic versus abiotic interactions on species
-composition [@pollock_understanding_2014].
+composition [@harris_estimating_2016].
 
@@
 $p(\vec{y}|\alpha, \beta)$. Fully-observed networks like the ones considered 
-here have unimodal likelihood surfaces [@murphy_machine_2012], so these
-estimates will represent the global optimum.
+here have unimodal likelihood surfaces [@murphy_machine_2012], so optimizers 
+will always find the global optimum.
 
 When the number of species is larger than about 30, noralizing Equation 1 
-can become intractable, and researchers will either need to approximate it [@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
+can become intractable, and researchers will either need to approximate it 
+[@lee_learning_2012] or approximate its gradient [@harris_estimating_2016] if
 they want to fit a Markov network model. For the analyses presented here, where
@@
 1600 sites using three increasingly complex models (50 replicates for each
-combination of size and model). The simplest set of simulated landscapes were
-generated with Gibbs samples from Equation 1. For each replicate, I randomly 
-drew the “true” $\beta$ coefficient magnitudes from an exponential distribution 
-with rate 1 so that most species pairs interacted negligibly but a few 
-interactions were strong enough to propagate through the newtwork. I randomly 
-assigned 25% of the interactions to be positive; the remainder were negative.
+combination of size and model; Appendix 2). The simplest set of simulated 
+landscapes were generated with Gibbs samples from Equation 1. For each 
+replicate, I randomly drew the “true” $\beta$ coefficient magnitudes from an 
+exponential distribution with rate 1 so that most species pairs interacted 
+negligibly but a few interactions were strong enough to propagate through the 
+newtwork. I randomly assigned 25% of the interactions to be positive; the 
+remainder were negative.
 
@@
 just presence/absence); furthermore, interactions between species occurred on a 
-per-capita basis in these simulations. To prevent runaway mutualisms leading to
+per-capita basis in these simulations (i.e. each species' effect on the others 
+is proportional to its abundance). To prevent runaway mutualisms leading to
 infinite abundance, all interaction coefficients were negative in these 
@@
 weakly-informative regularizer (equivalent to a logistic prior with location 0
-and scale 2) ensured that the estimates were always finite (Appendix 2).
+and scale 2) ensured that the estimates were always finite (Appendix 3).
 
@@
 [@lee_learning_2012].  Following @faisal_inferring_2010, I fit regularized 
-logistic regression models [@gelman_weakly_2008] for each species, using the
-other species on the landscape as predictors. This produced two interaction
-estimates for each species pair; one for the effect of species $i$ on species
-$j$ and one for the reverse. These two coefficients are not identifiable from
-the data, however [@schmidt_modeling_2012], so I used their average as an
-overall measure of the pairwise relationship.
+logistic regression models [@gelman_weakly_2008] for each 
+species, using the other species as predictors. This produced 
+two interaction estimates for each species pair (one for the effect of species 
+$i$ on species $j$ and one for the reverse). These two estimates were very 
+tightly correlated (mean Pearson correlation of 0.95, Appendix 3); their 
+arithmetic mean provided a consensus estimate of the overall interaction.
 
@@
 null hypothesis of zero interaction was estimated across a range of "true"
-interaction strengths using a generalized additive model with binomial responses
-(reject versus not; Appendix 4). The probability of rejection when the "true"
-value of $\beta$ was zero was defined as the Type I error rate.
+interaction strengths using a kernel smoother (Appendix 4). The probability of 
+rejection when the "true" value of $\beta$ was zero was defined as the Type I 
+error rate. Because the coefficients' interpretation is different for the 
+abundance simulations than for the other two types, its error rates were not
+analyzed in this way.
 
@@
 
-Finally, we can evaluate the models' statistical inferences (focusing on the
+Finally, we can evaluate the models' inferential statistics (focusing on the
 first two simulation types, where the interaction coefficients are easiest to
-interpret).  The Markov network's Type I error rate depended on the 
-simulation type: 0.02 for simulations that matched the model's assumptions,
-versus 0.14 for simulations that included environmental heterogeneity (see 
-Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
-values). The null model's Type I error rates were 0.31 and 0.53 for the
-constant and heterogeneous landscapes, respectively---far higher than the 
-nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
+interpret).  The Markov network's Type I error rate was 0.02 for simulations
+that matched the model's assumptions, and 0.14 for simulations that included
+environmental heterogeneity (see  Appendix 4 for confidence interval coverage
+across a range of $\beta_{ij}$ values). In contrast, the null model's Type I 
+error rates were 0.30 and 0.51, respectively---far higher than the nominal 0.05 
+rate. Figure 4C shows, across a range of true interaction 
 strengths, the probability that the null model or the Markov network will 
 predict the wrong sign of the interaction with 95% confidence.  The null model 
-makes such errors more than 5 times as often as the Markov network, even
+makes such errors more than 8 times as often as the Markov network, even
 though it only rejects the null hypothesis twice as often overall (Appendix
-4). The Markov network's errors were also more concentrated around 0, and it 
+4). The Markov network's errors were also more concentrated around 0, as it 
 never misclassified strong interactions like the null model did (Figure 4C).
@@
 
-Despite their limitations, Markov networks have enormous potential to improve
-ecological understanding. In particular, they make many fewer errors than existing
+Markov networks have enormous potential to improve our understanding of species 
+interactions. In particular, they make many fewer errors than existing
 approaches, and can make precise statements about the conditions where indirect
@@
 Finally, the results presented here have important implications for
-ecologists' continued use of null models for studying species interactions. Null
-and neutral models can be useful for clarifying our thinking
-[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
-null model must be interpreted with care [@roughgarden_competition_1983]. Even
-in small networks with three species, it may simply not be possible to implicate
-specific ecological processes like competition by rejecting a general-purpose
-null [@gotelli_empirical_2009], especially when the test statistic is
-effectively just a correlation coefficient (Figure 4B). When the non-null
-backdrop is not controlled for, Type I error rates can skyrocket, the apparent
-sign of the interaction can change, and null models can routinely produce
-misleading inferences (Figure 1, Figure 4C, @gotelli_empirical_2009).
+ecologists' continued use of null models for studying species interactions. 
+When the non-null backdrop is not controlled for, Type I error rates can 
+skyrocket, the apparent sign of the interaction can change, and null models can 
+routinely produce misleading inferences (Figure 1, Figure 4C, 
+@gotelli_empirical_2009). Null and neutral models can be useful for clarifying 
+our thinking [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a 
+given null model must be interpreted with care [@roughgarden_competition_1983]. 
+Even in small networks with three species, it may simply not be possible to 
+implicate specific ecological processes like competition by rejecting a 
+general-purpose null [@gotelli_empirical_2009], especially when the test 
+statistic is effectively just a correlation coefficient (Figure 4B).
 
@@
 the null model was much more likely to misclassify its sign with 95% confidence
-than the Markov network was.
+than the Markov network was. As with the other analyses based on inferential
+statistics, this panel only shows data from the first two simulation types.

### 2016-03-14 Harris_2015/manuscript-materials/draft.tex
final before resubmission to Ecology?

@@
 effects. Using simulated landscapes with known interactions, I evaluated
-Markov networks and six existing approaches. The Markov networks
-consistently outperformed other methods, correctly isolating direct
+Markov networks and six existing approaches. Markov networks
+consistently outperformed the other methods, correctly isolating direct
 interactions between species pairs even when indirect interactions or
@@
 (Roughgarden 1983) and disentangling the effects of biotic versus
-abiotic interactions on species composition (Pollock et al. 2014).
+abiotic interactions on species composition (Harris 2016).
 
@@
 considered here have unimodal likelihood surfaces (Murphy 2012), so
-these estimates will represent the global optimum.
+optimizers will always find the global optimum.
 
@@
 200, or 1600 sites using three increasingly complex models (50
-replicates for each combination of size and model). The simplest set of
-simulated landscapes were generated with Gibbs samples from Equation 1.
-For each replicate, I randomly drew the ``true'' \(\beta\) coefficient
-magnitudes from an exponential distribution with rate 1 so that most
-species pairs interacted negligibly but a few interactions were strong
-enough to propagate through the newtwork. I randomly assigned 25\% of
-the interactions to be positive; the remainder were negative.
+replicates for each combination of size and model; Appendix 2). The
+simplest set of simulated landscapes were generated with Gibbs samples
+from Equation 1. For each replicate, I randomly drew the ``true''
+\(\beta\) coefficient magnitudes from an exponential distribution with
+rate 1 so that most species pairs interacted negligibly but a few
+interactions were strong enough to propagate through the newtwork. I
+randomly assigned 25\% of the interactions to be positive; the remainder
+were negative.
 
@@
 (instead of just presence/absence); furthermore, interactions between
-species occurred on a per-capita basis in these simulations. To prevent
-runaway mutualisms leading to infinite abundance, all interaction
-coefficients were negative in these simulations. Good performance on
-these landscapes would indicate some robustness to the mechanistic
-details of species interactions.
+species occurred on a per-capita basis in these simulations (i.e.~each
+species' effect on the others is proportional to its abundance). To
+prevent runaway mutualisms leading to infinite abundance, all
+interaction coefficients were negative in these simulations. Good
+performance on these landscapes would indicate some robustness to the
+mechanistic details of species interactions.
 
@@
 location 0 and scale 2) ensured that the estimates were always finite
-(Appendix 2).
+(Appendix 3).
 
@@
 regression models (Gelman et al. 2008) for each species, using the other
-species on the landscape as predictors. This produced two interaction
-estimates for each species pair; one for the effect of species \(i\) on
-species \(j\) and one for the reverse. These two coefficients are not
-identifiable from the data, however (Schmidt and Murphy 2012), so I used
-their average as an overall measure of the pairwise relationship.
+species as predictors. This produced two interaction estimates for each
+species pair (one for the effect of species \(i\) on species \(j\) and
+one for the reverse). These two estimates were very tightly correlated
+(mean Pearson correlation of 0.95, Appendix 3); their arithmetic mean
+provided a consensus estimate of the overall interaction.
 
@@
 the null hypothesis of zero interaction was estimated across a range of
-``true'' interaction strengths using a generalized additive model with
-binomial responses (reject versus not; Appendix 4). The probability of
-rejection when the ``true'' value of \(\beta\) was zero was defined as
-the Type I error rate.
+``true'' interaction strengths using a kernel smoother (Appendix 4). The
+probability of rejection when the ``true'' value of \(\beta\) was zero
+was defined as the Type I error rate. Because the coefficients'
+interpretation is different for the abundance simulations than for the
+other two types, its error rates were not analyzed in this way.
 
@@
 
-Finally, we can evaluate the models' statistical inferences (focusing on
+Finally, we can evaluate the models' inferential statistics (focusing on
 the first two simulation types, where the interaction coefficients are
-easiest to interpret). The Markov network's Type I error rate depended
-on the simulation type: 0.02 for simulations that matched the model's
-assumptions, versus 0.14 for simulations that included environmental
-heterogeneity (see Appendix 4 for confidence interval coverage across a
-range of \(\beta_{ij}\) values). The null model's Type I error rates
-were 0.31 and 0.53 for the constant and heterogeneous landscapes,
-respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
-across a range of true interaction strengths, the probability that the
-null model or the Markov network will predict the wrong sign of the
-interaction with 95\% confidence. The null model makes such errors more
-than 5 times as often as the Markov network, even though it only rejects
-the null hypothesis twice as often overall (Appendix 4). The Markov
-network's errors were also more concentrated around 0, and it never
-misclassified strong interactions like the null model did (Figure 4C).
+easiest to interpret). The Markov network's Type I error rate was 0.02
+for simulations that matched the model's assumptions, and 0.14 for
+simulations that included environmental heterogeneity (see Appendix 4
+for confidence interval coverage across a range of \(\beta_{ij}\)
+values). In contrast, the null model's Type I error rates were 0.30 and
+0.51, respectively---far higher than the nominal 0.05 rate. Figure 4C
+shows, across a range of true interaction strengths, the probability
+that the null model or the Markov network will predict the wrong sign of
+the interaction with 95\% confidence. The null model makes such errors
+more than 8 times as often as the Markov network, even though it only
+rejects the null hypothesis twice as often overall (Appendix 4). The
+Markov network's errors were also more concentrated around 0, as it
+never misclassified strong interactions like the null model did (Figure
+4C).
 
@@
 
-Despite their limitations, Markov networks have enormous potential to
-improve ecological understanding. In particular, they make many fewer
-errors than existing approaches, and can make precise statements about
-the conditions where indirect interactions will overwhelm direct ones.
-They also provide a simple answer to the question of how competition
-should affect a species' overall prevalence, which has important
-implications for community-level modeling (Strong et al. 1984).
-Specifically, Equation 1 can be used to calculate the expected
-prevalence of a species in the absence of biotic influences as
+Markov networks have enormous potential to improve our understanding of
+species interactions. In particular, they make many fewer errors than
+existing approaches, and can make precise statements about the
+conditions where indirect interactions will overwhelm direct ones. They
+also provide a simple answer to the question of how competition should
+affect a species' overall prevalence, which has important implications
+for community-level modeling (Strong et al. 1984). Specifically,
+Equation 1 can be used to calculate the expected prevalence of a species
+in the absence of biotic influences as
 \(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
@@
 ecologists' continued use of null models for studying species
-interactions. Null and neutral models can be useful for clarifying our
-thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
-particular null model must be interpreted with care (Roughgarden 1983).
-Even in small networks with three species, it may simply not be possible
-to implicate specific ecological processes like competition by rejecting
-a general-purpose null (Gotelli and Ulrich 2009), especially when the
-test statistic is effectively just a correlation coefficient (Figure
-4B). When the non-null backdrop is not controlled for, Type I error
-rates can skyrocket, the apparent sign of the interaction can change,
-and null models can routinely produce misleading inferences (Figure 1,
-Figure 4C, Gotelli and Ulrich (2009)).
+interactions. When the non-null backdrop is not controlled for, Type I
+error rates can skyrocket, the apparent sign of the interaction can
+change, and null models can routinely produce misleading inferences
+(Figure 1, Figure 4C, Gotelli and Ulrich (2009)). Null and neutral
+models can be useful for clarifying our thinking (Harris et al. 2011,
+Xiao et al. 2015), but deviations from a given null model must be
+interpreted with care (Roughgarden 1983). Even in small networks with
+three species, it may simply not be possible to implicate specific
+ecological processes like competition by rejecting a general-purpose
+null (Gotelli and Ulrich 2009), especially when the test statistic is
+effectively just a correlation coefficient (Figure 4B).
 
@@
 more likely to misclassify its sign with 95\% confidence than the Markov
-network was.
+network was. As with the other analyses based on inferential statistics,
+this panel only shows data from the first two simulation types.
 

### 2016-03-14 Harris_2015/recovery.Rmd
final before resubmission to Ecology?

@@
 This document describes how the different models were fit to the simulated data 
-from Appendix 2 and how each model's performance was evaluated.[^1]
+from Appendix 2 and how each model's performance was evaluated.
 
@@
 
-```{r, cache = FALSE, eval = FALSE}
+```{r, cache = FALSE, eval = TRUE}
 
@@
   gsub("\\.csv$", "", .)
+```
 
+```{r, eval = FALSE}
 mclapply(identifiers, fit_all, mc.cores = mc.cores, 
@@
 
-### Logistic prior used for the Markov network
+## Logistic prior used for the Markov network
 
@@
 ```
+
+## GLM coefficient correlations
+
+The GLM method produces two estimates for each species pair.  These estimates
+tend to be very similar, so most reasonable methods for generating a consensus
+estimate should produce very similar results.
+
+```{r, cache=TRUE}
+get_glm_cor = function(identifier){
+  data_filename = paste0("fakedata/matrices/", identifier, ".csv")
+
+  # first column is row numbers; drop it
+  raw_obs = as.matrix(read.csv(data_filename)[ , -1])
+
+  # Identify species that are never present (or never absent) so they 
+  # can be dropped
+  species_is_variable = diag(var(raw_obs)) > 0
+  pair_is_variable = tcrossprod(species_is_variable) > 0
+
+  x = raw_obs[ , species_is_variable]
+
+  coef_matrix = matrix(0, ncol(x), ncol(x))
+
+  # Fill in the coefficient matrix, column-by-column
+  for(i in 1:ncol(x)){
+    if(var(x[,i]) > 0){
+      coefs = coef(bayesglm(x[,i] ~ x[ , -i], family = binomial))[-1]
+      coef_matrix[i, -i] = coefs
+    }
+  }
+
+  # Compare the estimates from the upper triangle with the estimates from 
+  # the estimates in the lower triangle (which becomes the upper triangle after 
+  # transposing)
+  cor(coef_matrix[upper.tri(coef_matrix)], t(coef_matrix)[upper.tri(coef_matrix)])
+}
+
+glm_correlations = unlist(mclapply(identifiers, get_glm_cor))
+
+mean(glm_correlations)
+
+hist(glm_correlations)
+```

### 2016-03-14 Harris_2015/recovery.tex
final before resubmission to Ecology?

@@
 \StringTok{  }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{, }\StringTok{""}\NormalTok{, .)}
+\end{Highlighting}
+\end{Shaded}
 
+\begin{Shaded}
+\begin{Highlighting}[]
 \KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }
@@
 
-\subsubsection{Logistic prior used for the Markov
+\subsection{Logistic prior used for the Markov
 network}\label{logistic-prior-used-for-the-markov-network}
@@
 
-\includegraphics{recovery_files/figure-latex/unnamed-chunk-4-1.pdf}
+\includegraphics{recovery_files/figure-latex/unnamed-chunk-5-1.pdf}
 
@@
 
-\includegraphics{recovery_files/figure-latex/unnamed-chunk-5-1.pdf}
+\includegraphics{recovery_files/figure-latex/unnamed-chunk-6-1.pdf}
+
+\subsection{GLM coefficient
+correlations}\label{glm-coefficient-correlations}
+
+The GLM method produces two estimates for each species pair. These
+estimates tend to be very similar, so most reasonable methods for
+generating a consensus estimate should produce very similar results.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{get_glm_cor =}\StringTok{ }\NormalTok{function(identifier)\{}
+  \NormalTok{data_filename =}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"fakedata/matrices/"}\NormalTok{, identifier, }\StringTok{".csv"}\NormalTok{)}
+
+  \CommentTok{# first column is row numbers; drop it}
+  \NormalTok{raw_obs =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(}\KeywordTok{read.csv}\NormalTok{(data_filename)[ , -}\DecValTok{1}\NormalTok{])}
+
+  \CommentTok{# Identify species that are never present (or never absent) so they }
+  \CommentTok{# can be dropped}
+  \NormalTok{species_is_variable =}\StringTok{ }\KeywordTok{diag}\NormalTok{(}\KeywordTok{var}\NormalTok{(raw_obs)) >}\StringTok{ }\DecValTok{0}
+  \NormalTok{pair_is_variable =}\StringTok{ }\KeywordTok{tcrossprod}\NormalTok{(species_is_variable) >}\StringTok{ }\DecValTok{0}
+
+  \NormalTok{x =}\StringTok{ }\NormalTok{raw_obs[ , species_is_variable]}
+
+  \NormalTok{coef_matrix =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{ncol}\NormalTok{(x), }\KeywordTok{ncol}\NormalTok{(x))}
+
+  \CommentTok{# Fill in the coefficient matrix, column-by-column}
+  \NormalTok{for(i in }\DecValTok{1}\NormalTok{:}\KeywordTok{ncol}\NormalTok{(x))\{}
+    \NormalTok{if(}\KeywordTok{var}\NormalTok{(x[,i]) >}\StringTok{ }\DecValTok{0}\NormalTok{)\{}
+      \NormalTok{coefs =}\StringTok{ }\KeywordTok{coef}\NormalTok{(}\KeywordTok{bayesglm}\NormalTok{(x[,i] ~}\StringTok{ }\NormalTok{x[ , -i], }\DataTypeTok{family =} \NormalTok{binomial))[-}\DecValTok{1}\NormalTok{]}
+      \NormalTok{coef_matrix[i, -i] =}\StringTok{ }\NormalTok{coefs}
+    \NormalTok{\}}
+  \NormalTok{\}}
+
+  \CommentTok{# Compare the estimates from the upper triangle with the estimates from }
+  \CommentTok{# the estimates in the lower triangle (which becomes the upper triangle after }
+  \CommentTok{# transposing)}
+  \KeywordTok{cor}\NormalTok{(coef_matrix[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)], }\KeywordTok{t}\NormalTok{(coef_matrix)[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)])}
+\NormalTok{\}}
+
+\NormalTok{glm_correlations =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(}\KeywordTok{mclapply}\NormalTok{(identifiers, get_glm_cor))}
+
+\KeywordTok{mean}\NormalTok{(glm_correlations)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 0.9546387
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{hist}\NormalTok{(glm_correlations)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{recovery_files/figure-latex/unnamed-chunk-7-1.pdf}
 

### 2016-03-14 Harris_2015/show-results.Rmd
final before resubmission to Ecology?

@@
     keep_tex: true
+    toc: true
 title: "Appendix 4: Results"
@@
 
-```{r, cache = TRUE}
+```{r, cache = FALSE}
 resids = function(data){
@@
 result_summary$method_r2 = reorder(result_summary$method_r2, -result_summary$r2)
+```
+
+#### Plot the regressions used above to calculate R^2
+
+The plots below each correspond to one of the regressions performed above to
+calculate R^2 for a combination of simulation method and estimation method.
+The null model ("Pairs") was not included in these plots because its Z-scores
+were frequently too large to fit on the same scale as the other methods' estimates.
+Below, the thinner dashed lines represent the 1:1 line, where the expected estimate
+matches the "true" value, and the thicker solid lines represent a linear regression
+through the origin, as calculated above. In general, the four correlation- or
+partial correlation-based methods had slopes that were steeper than the 1:1 line,
+because their estimates were constrained between -1 and 1. The GLM and Markov network
+usually had slopes that were closer to 1.
+
+(Warnings of the form "Removed N rows containing missing values
+(geom_smooth)" can be ignored: they are simply noting that the regression lines 
+were cut off in some panels of the figure).
+
+```{r, dev="png", dpi=500, cache = TRUE}
+plot_regressions = function(type, title){
+  x[sample.int(nrow(x)) , ] %>%
+  subset(simulation_type == type & method != "null") %>% 
+  ggplot(aes(x = estimate, y = truth)) +
+    facet_wrap(~method) +
+    ylim(min(x$truth), max(x$truth)) + 
+    geom_point(aes(color = factor(n_sites)), size = 0.2) + 
+    geom_smooth(method = "lm", color = "black", formula = y ~ 0 + x, 
+                fullrange = TRUE) + 
+    geom_abline(intercept = 0, slope = 1, linetype = 2) +
+    theme_bw() + 
+    ggtitle(title)
+}
+
+plot_regressions("no_env", "'Baseline' regressions")
+plot_regressions("env", "'Environment' regressions")
+plot_regressions("abund", "'Abundance' regressions")
 ```
@@
 
-```{r, cache = TRUE}
+```{r, cache = FALSE}
 landscape_estimates = x %>% 
@@
 
-## Plot estimates versus "true" values across methods and lanscape sizes
+# C: Inferential statistics
 
-The gray line indicates the 1:1 line, where models recover an unbiased estimate
-of the "true" $\beta$ coefficients.  The GLM and Markov network methods estimate
-the coefficients on the original scale, and approach unbiasedness as the number
-of sites increases. The four methods based on (partial) correlations are
-constrained to return values between -1 and 1, and do not return values on the
-original scale.  The covariance and partial covariance (which are not scaled in
-this way) may avoid this problem in future analyses (Loh and Wainwright 2013; 
-see main text for full citation).
-
-Note that the null model (Pairs) is not included in these graphs, as its
-test statistic depends strongly on the size of the landscape and fills a range 
-that would not fit on these graphs.
-
-```{r, dev="png", dpi=500, cache = TRUE}
-ggplot(filter(x, simulation_type == "no_env", method != "null"),
-       aes(x = truth, y = estimate, color = factor(n_sites))) + 
-  facet_wrap(~method) +
-  geom_point(alpha = 0.01) +
-  geom_abline(intercept = 0, slope = 1, size = 2, col = "darkgray") + 
-  geom_smooth(method = "lm", fullrange = FALSE) + 
-  coord_equal() + 
-  scale_color_discrete(name = "n_sites")
-```
-
-# C: Inferential statistics
+Note that, as mentioned in the main text, inferential statistics are only 
+calculated for the `env` and `no_env` simulations (and not for the `abund`
+simulations).
 
@@
 
-```{r, cache = TRUE}
+```{r, cache = FALSE}
 
-pairs_summary = x[x$method == "null" & !grepl("pop", x$rep_name), ]
-markov_summary = x[x$method == "Markov network" & !grepl("pop", x$rep_name), ]
+pairs_summary = x[x$method == "null" & x$simulation_type != "abund", ]
+markov_summary = x[x$method == "Markov network" & x$simulation_type != "abund", ]
 
@@
 
-The `error_smoother` is a function that fits a generalized additive model to 
+The `error_smoother` is a function that fits a kernel smoother model to 
 estimate the probability that a model estimate will match some criterion (e.g. 
@@
 error_smoother = function(data, f, values = truth_seq){
-  predict(
-    gam(
-      f(data) ~ s(truth),
-      data = data,
-      family = binomial
-    ),
-    data.frame(truth = values),
-    type = "response"
+  out = ksmooth(
+    x = data$truth,
+    y = as.numeric(f(data)),
+    x.points = values,
+    kernel = "normal",
+    bandwidth = 0.25
   )
+
+  out$y[is.na(out$y)] = 0
+
+  out
 }
@@
 }
-y_markov = error_smoother(markov_summary, confidently_wrong)
-y_pairs = error_smoother(pairs_summary, confidently_wrong)
+smoothed_markov = error_smoother(markov_summary, confidently_wrong)
+smoothed_pairs = error_smoother(pairs_summary, confidently_wrong)
 
@@
 plot(
-  truth_seq,
-  y_pairs,
+  smoothed_pairs$x,
+  smoothed_pairs$y,
   type = "l",
@@
   col = 2,
-  ylim = c(0, .4),
+  ylim = c(0, .3),
   lwd = 2
@@
       font = 2, line = 1.2, cex = .9) 
-lines(truth_seq, y_markov, lwd = 2)
+lines(smoothed_markov$x, smoothed_markov$y, lwd = 2)
 legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
@@
 Type I error rates (probability of rejecting the null hypothesis given a 
-"true" interaction strength of zero, according to the spline model defined by
-`error_smoother` above). The smoother focuses on the 
-`r markov_summary %>% summarize(mean(abs(truth) < 1)) %>% round(., 2) %>% multiply_by(100)`%
-of the data points whose "true" values fall between -1 and 1, because its 
-estimates near 0 are more accurate when they aren't affected by distant data 
-points.
+"true" interaction strength of zero, according to the kernel smoother defined by
+`error_smoother` above).
 
@@
     `Type I error rate` = 
-      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
+      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)$y
   )) %>%
@@
 The top and bottom 0.5% of the distribution have been omitted to prevent bad 
-behavior by the smoother in the tails. The horizontal red line shows the nominal
-95% coverage rate and the black vertical line marks where the "true" value
-of $\beta$ was zero. The Type I error rates calculated above for the Markov 
-network closely match (one minus) the values of these curves at zero.
+behavior by the smoother (a generalized additive model) in the tails. The 
+horizontal red line shows the nominal 95% coverage rate and the black vertical 
+line marks where the "true" value of $\beta$ was zero. The Type I error rates 
+calculated above for the Markov network closely match (one minus) the values of 
+these curves at zero.
 
@@
 ```
+
+## Plot power versus true $\beta$ value
+
+Calculate p(confidently detect sign of $\beta$) as a function of
+the "true"" values.  Pairs has a somewhat higher rate of rejecting the null 
+hypothesis in the correct direction than the Markov network, but the benefit of 
+this increase in power is more than offset by the large increases in Type I 
+errors and sign errors demonstrated above and in Figure 4 of the main text.
+
+```{r}
+confidently_right = function(data){
+  (data$sig_pos & data$truth > 0) | (data$sig_neg & data$truth < 0)
+}
+
+plot(error_smoother(markov_summary, confidently_right), 
+     type = "l", 
+     xlim = c(-5, 5),
+     xlab = "'True' coefficient value",
+     ylab = "Power",
+     ylim = c(0, 1),
+     yaxs = "i"
+)
+lines(error_smoother(pairs_summary, confidently_right), col = 2)
+legend("topleft", legend = c("Markov network", "Pairs"), lty = 1, col = 1:2)
+```

### 2016-03-14 Harris_2015/show-results.tex
final before resubmission to Ecology?

@@
 
+{
+\hypersetup{linkcolor=black}
+\setcounter{tocdepth}{2}
+\tableofcontents
+}
 \section{A: Import packages and data}\label{a-import-packages-and-data}
@@
 
+\paragraph{Plot the regressions used above to calculate
+R\^{}2}\label{plot-the-regressions-used-above-to-calculate-r2}
+
+The plots below each correspond to one of the regressions performed
+above to calculate R\^{}2 for a combination of simulation method and
+estimation method. The null model (``Pairs'') was not included in these
+plots because its Z-scores were frequently too large to fit on the same
+scale as the other methods' estimates. Below, the thinner dashed lines
+represent the 1:1 line, where the expected estimate matches the ``true''
+value, and the thicker solid lines represent a linear regression through
+the origin, as calculated above. In general, the four correlation- or
+partial correlation-based methods had slopes that were steeper than the
+1:1 line, because their estimates were constrained between -1 and 1. The
+GLM and Markov network usually had slopes that were closer to 1.
+
+(Warnings of the form ``Removed N rows containing missing values
+(geom\_smooth)'' can be ignored: they are simply noting that the
+regression lines were cut off in some panels of the figure).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{plot_regressions =}\StringTok{ }\NormalTok{function(type, title)\{}
+  \NormalTok{x[}\KeywordTok{sample.int}\NormalTok{(}\KeywordTok{nrow}\NormalTok{(x)) , ] %>%}
+\StringTok{  }\KeywordTok{subset}\NormalTok{(simulation_type ==}\StringTok{ }\NormalTok{type &}\StringTok{ }\NormalTok{method !=}\StringTok{ "null"}\NormalTok{) %>%}\StringTok{ }
+\StringTok{  }\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{estimate, }\DataTypeTok{y =} \NormalTok{truth)) +}
+\StringTok{    }\KeywordTok{facet_wrap}\NormalTok{(~method) +}
+\StringTok{    }\KeywordTok{ylim}\NormalTok{(}\KeywordTok{min}\NormalTok{(x$truth), }\KeywordTok{max}\NormalTok{(x$truth)) +}\StringTok{ }
+\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =} \KeywordTok{factor}\NormalTok{(n_sites)), }\DataTypeTok{size =} \FloatTok{0.2}\NormalTok{) +}\StringTok{ }
+\StringTok{    }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \StringTok{"lm"}\NormalTok{, }\DataTypeTok{color =} \StringTok{"black"}\NormalTok{, }\DataTypeTok{formula =} \NormalTok{y ~}\StringTok{ }\DecValTok{0} \NormalTok{+}\StringTok{ }\NormalTok{x, }
+                \DataTypeTok{fullrange =} \OtherTok{TRUE}\NormalTok{) +}\StringTok{ }
+\StringTok{    }\KeywordTok{geom_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{linetype =} \DecValTok{2}\NormalTok{) +}
+\StringTok{    }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
+\StringTok{    }\KeywordTok{ggtitle}\NormalTok{(title)}
+\NormalTok{\}}
+
+\KeywordTok{plot_regressions}\NormalTok{(}\StringTok{"no_env"}\NormalTok{, }\StringTok{"'Baseline' regressions"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 227 rows containing missing values (geom_smooth).
+\end{verbatim}
+
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-5-1.png}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot_regressions}\NormalTok{(}\StringTok{"env"}\NormalTok{, }\StringTok{"'Environment' regressions"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 188 rows containing missing values (geom_smooth).
+\end{verbatim}
+
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-5-2.png}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{plot_regressions}\NormalTok{(}\StringTok{"abund"}\NormalTok{, }\StringTok{"'Abundance' regressions"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: Removed 155 rows containing missing values (geom_smooth).
+\end{verbatim}
+
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-5-3.png}
+
 \paragraph{Average the R-squared values across methods and simulation
@@
 
-\subsection{\texorpdfstring{Plot estimates versus ``true'' values across
-methods and lanscape
-sizes}{Plot estimates versus true values across methods and lanscape sizes}}\label{plot-estimates-versus-true-values-across-methods-and-lanscape-sizes}
+\section{C: Inferential statistics}\label{c-inferential-statistics}
 
-The gray line indicates the 1:1 line, where models recover an unbiased
-estimate of the ``true'' \(\beta\) coefficients. The GLM and Markov
-network methods estimate the coefficients on the original scale, and
-approach unbiasedness as the number of sites increases. The four methods
-based on (partial) correlations are constrained to return values between
--1 and 1, and do not return values on the original scale. The covariance
-and partial covariance (which are not scaled in this way) may avoid this
-problem in future analyses (Loh and Wainwright 2013; see main text for
-full citation).
-
-Note that the null model (Pairs) is not included in these graphs, as its
-test statistic depends strongly on the size of the landscape and fills a
-range that would not fit on these graphs.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{ggplot}\NormalTok{(}\KeywordTok{filter}\NormalTok{(x, simulation_type ==}\StringTok{ "no_env"}\NormalTok{, method !=}\StringTok{ "null"}\NormalTok{),}
-       \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{truth, }\DataTypeTok{y =} \NormalTok{estimate, }\DataTypeTok{color =} \KeywordTok{factor}\NormalTok{(n_sites))) +}\StringTok{ }
-\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(~method) +}
-\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.01}\NormalTok{) +}
-\StringTok{  }\KeywordTok{geom_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \DecValTok{1}\NormalTok{, }\DataTypeTok{size =} \DecValTok{2}\NormalTok{, }\DataTypeTok{col =} \StringTok{"darkgray"}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_smooth}\NormalTok{(}\DataTypeTok{method =} \StringTok{"lm"}\NormalTok{, }\DataTypeTok{fullrange =} \OtherTok{FALSE}\NormalTok{) +}\StringTok{ }
-\StringTok{  }\KeywordTok{coord_equal}\NormalTok{() +}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_color_discrete}\NormalTok{(}\DataTypeTok{name =} \StringTok{"n_sites"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{show-results_files/figure-latex/unnamed-chunk-8-1.png}
-
-\section{C: Inferential statistics}\label{c-inferential-statistics}
+Note that, as mentioned in the main text, inferential statistics are
+only calculated for the \texttt{env} and \texttt{no\_env} simulations
+(and not for the \texttt{abund} simulations).
 
@@
 \begin{Highlighting}[]
-\NormalTok{pairs_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "null"} \NormalTok{&}\StringTok{ }\NormalTok{!}\KeywordTok{grepl}\NormalTok{(}\StringTok{"pop"}\NormalTok{, x$rep_name), ]}
-\NormalTok{markov_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "Markov network"} \NormalTok{&}\StringTok{ }\NormalTok{!}\KeywordTok{grepl}\NormalTok{(}\StringTok{"pop"}\NormalTok{, x$rep_name), ]}
+\NormalTok{pairs_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "null"} \NormalTok{&}\StringTok{ }\NormalTok{x$simulation_type !=}\StringTok{ "abund"}\NormalTok{, ]}
+\NormalTok{markov_summary =}\StringTok{ }\NormalTok{x[x$method ==}\StringTok{ "Markov network"} \NormalTok{&}\StringTok{ }\NormalTok{x$simulation_type !=}\StringTok{ "abund"}\NormalTok{, ]}
 
@@
 
-The \texttt{error\_smoother} is a function that fits a generalized
-additive model to estimate the probability that a model estimate will
-match some criterion (e.g. statistical significance) for some parameter
-as a function of that parameter's ``true'' value. This function is used
-in Figure 4C and in calculating Type I error rates below.
+The \texttt{error\_smoother} is a function that fits a kernel smoother
+model to estimate the probability that a model estimate will match some
+criterion (e.g. statistical significance) for some parameter as a
+function of that parameter's ``true'' value. This function is used in
+Figure 4C and in calculating Type I error rates below.
 
@@
 \NormalTok{error_smoother =}\StringTok{ }\NormalTok{function(data, f, }\DataTypeTok{values =} \NormalTok{truth_seq)\{}
-  \KeywordTok{predict}\NormalTok{(}
-    \KeywordTok{gam}\NormalTok{(}
-      \KeywordTok{f}\NormalTok{(data) ~}\StringTok{ }\KeywordTok{s}\NormalTok{(truth),}
-      \DataTypeTok{data =} \NormalTok{data,}
-      \DataTypeTok{family =} \NormalTok{binomial}
-    \NormalTok{),}
-    \KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{truth =} \NormalTok{values),}
-    \DataTypeTok{type =} \StringTok{"response"}
+  \NormalTok{out =}\StringTok{ }\KeywordTok{ksmooth}\NormalTok{(}
+    \DataTypeTok{x =} \NormalTok{data$truth,}
+    \DataTypeTok{y =} \KeywordTok{as.numeric}\NormalTok{(}\KeywordTok{f}\NormalTok{(data)),}
+    \DataTypeTok{x.points =} \NormalTok{values,}
+    \DataTypeTok{kernel =} \StringTok{"normal"}\NormalTok{,}
+    \DataTypeTok{bandwidth =} \FloatTok{0.25}
   \NormalTok{)}
+
+  \NormalTok{out$y[}\KeywordTok{is.na}\NormalTok{(out$y)] =}\StringTok{ }\DecValTok{0}
+
+  \NormalTok{out}
 \NormalTok{\}}
@@
 \NormalTok{\}}
-\NormalTok{y_markov =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(markov_summary, confidently_wrong)}
-\NormalTok{y_pairs =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(pairs_summary, confidently_wrong)}
+\NormalTok{smoothed_markov =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(markov_summary, confidently_wrong)}
+\NormalTok{smoothed_pairs =}\StringTok{ }\KeywordTok{error_smoother}\NormalTok{(pairs_summary, confidently_wrong)}
 
@@
 \KeywordTok{plot}\NormalTok{(}
-  \NormalTok{truth_seq,}
-  \NormalTok{y_pairs,}
+  \NormalTok{smoothed_pairs$x,}
+  \NormalTok{smoothed_pairs$y,}
   \DataTypeTok{type =} \StringTok{"l"}\NormalTok{,}
@@
   \DataTypeTok{col =} \DecValTok{2}\NormalTok{,}
-  \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{4}\NormalTok{),}
+  \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{3}\NormalTok{),}
   \DataTypeTok{lwd =} \DecValTok{2}
@@
       \DataTypeTok{font =} \DecValTok{2}\NormalTok{, }\DataTypeTok{line =} \FloatTok{1.2}\NormalTok{, }\DataTypeTok{cex =} \NormalTok{.}\DecValTok{9}\NormalTok{) }
-\KeywordTok{lines}\NormalTok{(truth_seq, y_markov, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{lines}\NormalTok{(smoothed_markov$x, smoothed_markov$y, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{)}
 \KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{lwd =} \DecValTok{2}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Null model"}\NormalTok{, }\StringTok{"Markov network"}\NormalTok{), }
@@
 \begin{verbatim}
-## [1] 0.1533758
+## [1] 0.1234429
 \end{verbatim}
@@
 \begin{verbatim}
-## [1] 0.02861541
+## [1] 0.01529933
 \end{verbatim}
@@
 \begin{verbatim}
-## [1] 0.4520535
+## [1] 0.4640395
 \end{verbatim}
@@
 \begin{verbatim}
-## [1] 0.2116561
+## [1] 0.22179
 \end{verbatim}
@@
 Type I error rates (probability of rejecting the null hypothesis given a
-``true'' interaction strength of zero, according to the spline model
-defined by \texttt{error\_smoother} above). The smoother focuses on the
-64\% of the data points whose ``true'' values fall between -1 and 1,
-because its estimates near 0 are more accurate when they aren't affected
-by distant data points.
+``true'' interaction strength of zero, according to the kernel smoother
+defined by \texttt{error\_smoother} above).
 
@@
     \StringTok{`}\DataTypeTok{Type I error rate}\StringTok{`} \NormalTok{=}\StringTok{ }
-\StringTok{      }\KeywordTok{error_smoother}\NormalTok{(., function(x)\{x$sig_pos |}\StringTok{ }\NormalTok{x$sig_neg\}, }\DataTypeTok{values =} \DecValTok{0}\NormalTok{)}
+\StringTok{      }\KeywordTok{error_smoother}\NormalTok{(., function(x)\{x$sig_pos |}\StringTok{ }\NormalTok{x$sig_neg\}, }\DataTypeTok{values =} \DecValTok{0}\NormalTok{)$y}
   \NormalTok{)) %>%}
@@
 \endhead
-abund & 0.258\tabularnewline
-env & 0.142\tabularnewline
+env & 0.138\tabularnewline
 no\_env & 0.024\tabularnewline
@@
 \endhead
-abund & 0.597\tabularnewline
-env & 0.530\tabularnewline
-no\_env & 0.308\tabularnewline
+env & 0.510\tabularnewline
+no\_env & 0.298\tabularnewline
 \bottomrule
@@
 The top and bottom 0.5\% of the distribution have been omitted to
-prevent bad behavior by the smoother in the tails. The horizontal red
-line shows the nominal 95\% coverage rate and the black vertical line
-marks where the ``true'' value of \(\beta\) was zero. The Type I error
-rates calculated above for the Markov network closely match (one minus)
-the values of these curves at zero.
+prevent bad behavior by the smoother (a generalized additive model) in
+the tails. The horizontal red line shows the nominal 95\% coverage rate
+and the black vertical line marks where the ``true'' value of \(\beta\)
+was zero. The Type I error rates calculated above for the Markov network
+closely match (one minus) the values of these curves at zero.
 
@@
 
+\subsection{\texorpdfstring{Plot power versus true \(\beta\)
+value}{Plot power versus true \textbackslash{}beta value}}\label{plot-power-versus-true-beta-value}
+
+Calculate p(confidently detect sign of \(\beta\)) as a function of the
+``true''" values. Pairs has a somewhat higher rate of rejecting the null
+hypothesis in the correct direction than the Markov network, but the
+benefit of this increase in power is more than offset by the large
+increases in Type I errors and sign errors demonstrated above and in
+Figure 4 of the main text.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{confidently_right =}\StringTok{ }\NormalTok{function(data)\{}
+  \NormalTok{(data$sig_pos &}\StringTok{ }\NormalTok{data$truth >}\StringTok{ }\DecValTok{0}\NormalTok{) |}\StringTok{ }\NormalTok{(data$sig_neg &}\StringTok{ }\NormalTok{data$truth <}\StringTok{ }\DecValTok{0}\NormalTok{)}
+\NormalTok{\}}
+
+\KeywordTok{plot}\NormalTok{(}\KeywordTok{error_smoother}\NormalTok{(markov_summary, confidently_right), }
+     \DataTypeTok{type =} \StringTok{"l"}\NormalTok{, }
+     \DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(-}\DecValTok{5}\NormalTok{, }\DecValTok{5}\NormalTok{),}
+     \DataTypeTok{xlab =} \StringTok{"'True' coefficient value"}\NormalTok{,}
+     \DataTypeTok{ylab =} \StringTok{"Power"}\NormalTok{,}
+     \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{),}
+     \DataTypeTok{yaxs =} \StringTok{"i"}
+\NormalTok{)}
+\KeywordTok{lines}\NormalTok{(}\KeywordTok{error_smoother}\NormalTok{(pairs_summary, confidently_right), }\DataTypeTok{col =} \DecValTok{2}\NormalTok{)}
+\KeywordTok{legend}\NormalTok{(}\StringTok{"topleft"}\NormalTok{, }\DataTypeTok{legend =} \KeywordTok{c}\NormalTok{(}\StringTok{"Markov network"}\NormalTok{, }\StringTok{"Pairs"}\NormalTok{), }\DataTypeTok{lty =} \DecValTok{1}\NormalTok{, }\DataTypeTok{col =} \DecValTok{1}\NormalTok{:}\DecValTok{2}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{show-results_files/figure-latex/unnamed-chunk-19-1.pdf}
+
 \end{document}

### 2016-03-14 Harris_2015/simulation.Rmd
final before resubmission to Ecology?

@@
 
-The $alpha$ coefficients controlling species' intercepts were drawn from 
+The $\alpha$ coefficients controlling species' intercepts were drawn from 
 Gaussian distributions with unit variance and a mean given by `mean_alpha`.
@@
 distribution with rate parameter equal to 1.  The sign of each coefficient was
-randomly sampled, based on the value of `p_neg`, defined below. as 0.75 (75%
+randomly sampled, based on the value of `p_neg`, defined below as 0.75 (75%
 negative) for the presence-absence simulations and 1.0 (100% negative) for the
@@
 
-The following function calls the function above to generate "true" parameters, 
+The next function calls the function above to generate "true" parameters, 
 then uses them to simulate a landscape with Markov chain Monte Carlo, and 
@@
 I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
-In each round of Gibbs sampling, I cycled through all the species, randomly
+In each round of Gibbs sampling, I cycled through all the species,
 updating each one's binary presence/absence vector with random samples from
@@
 simulated with a Poisson distribution instead of a Bernoulli distribution. The
-rate parameter of this distribution was given by 
-$\mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$. (The more conventional 
-exponential inverse link was not used because preliminary experiments showed that
-it led to very large abundances that prevented the Gibbs sampler from mixing 
-effectively.)
+rate parameter of this distribution (which corresponds to the expected abundance)
+was given by
+
+$$y_i \sim \mathrm{log}(1 + exp(\alpha_i + \sum_j\beta_{ij}y_j))$$.
+
+(The more conventional exponential inverse link was not used because preliminary 
+experiments showed that it led to very large $y$ values, which prevented the 
+Gibbs sampler from mixing effectively.)
 
@@
 ```{r, cache = TRUE}
-simulate_data = function(n_spp, n_sites, rep_name, n_gibbs, n_env, sd, f, rdist, p_neg, mean_alpha){
+simulate_data = function(n_spp, n_sites, rep_name, n_gibbs, n_env, sd, f, 
+                         rdist, p_neg, mean_alpha){
   # n_spp:     number of species to include in the landscape
@@
 
-  # Environmental states are normally distributed
+  # Environmental states are normally distributed with mean=0 and sd=1
   env = matrix(rnorm(n_sites * n_env), ncol = n_env)
@@
 
-  # Landscape starts as if betas were all zero. Each species' occurrence probability
-  # or abundance depends on its alpha value and on the environment (assuming alpha_env 
-  # is not set to zero).
+  # Landscape starts as if betas were all zero. Each species' occurrence 
+  # probability or abundance depends on its alpha value and on the 
+  # environment (assuming alpha_env is not set to zero).
   x = matrix(
@@
 
-Loop to simulate and save 50 landscapes for each combination of simulation type
-and landscape size.
+A loop to simulate and save 50 landscapes for each combination of simulation type
+and landscape size, using the functions defined above:
 
-```{r}
+```{r, eval = FALSE}
 set.seed(1)

### 2016-03-15 Harris_2015/manuscript-materials/draft.tex
as resubmitted to Ecology

@@
 
-\textbf{Figure 1: A.} A small network of three competing species. The
-tree (top) tends not to co-occur with either of the two shrub species,
-as indicated by the strongly negative coefficient linking them. The two
-shrub species also compete with one another, but more weakly (circled
-coefficient). \textbf{B.} In spite of the competitive interactions
-between the two shrub species, their shared tendency to occur in
-locations without trees makes their occurrence vectors positively
-correlated (circled). \textbf{C.} Controlling for trees with a
-conditional (all-else-equal) approach such as a partial correlation or a
-Markov network leads to correct identification of the negative
-shrub-shrub interaction (circled). See Appendix 1 and the results for
-``three species'' for more details.
-
-\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
-and \(\beta\) values. The abiotic environment favors the occurrence of
-each species (\(\alpha >0\)), particularly species 2
-(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
-consistent with competition between the two species. \textbf{B.} The
-coefficients determine the probabilities of all four possible
-presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
-is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
-but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
-includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
-and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
-The normalizing constant \(Z\), ensures that the four probabilities sum
-to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
-frequencies of all possible co-occurrence patterns between the two
-species of interest, as calculated in the previous panel. \textbf{D.}
-Without competition (i.e.~with \(\beta_{12}=0\), each species would
-occur more often.
-
-\textbf{Figure 3:} Proportion of variance in interaction coefficients
-explained by each method versus number of sampled locations across the
-three simulation types. For the null model (Pairs), two outliers with
-\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
-detrimental influence on \(R^2\) (Appendix 5).
-
-\textbf{Figure 4: A.} The Markov network's estimated interaction
-coefficients were generally very similar to the GLM estimates.
-\textbf{B.} The null model's estimates typically matched the (negative)
-correlation coefficient, after controlling for landscape size.
-\textbf{C.} For any given interaction strength, the null model was much
-more likely to misclassify its sign with 95\% confidence than the Markov
-network was. As with the other analyses based on inferential statistics,
-this panel only shows data from the first two simulation types.
-
 \hyperdef{}{refs}{\label{refs}}
@@
 Harris, D. J. 2016. Estimating species interactions in large,
-abiotically structured communities using markov networks and stochastic
-approximation. bioRxiv.
+abiotically structured communities with markov networks and stochastic
+approximation. \emph{in} Multi-process statistical modeling of species'
+joint distributions. Figshare.
+http://doi.org/10.6084/m9.figshare.3114226.v1.
 
@@
 maximum entropy theory of ecology. The American Naturalist 185:E70--E80.
+
+\setlength{\parindent}{0em} \setlength{\leftskip}{0em}
+
+\textbf{Figure Captions}
+
+\textbf{Figure 1: A.} A small network of three competing species. The
+tree (top) tends not to co-occur with either of the two shrub species,
+as indicated by the strongly negative coefficient linking them. The two
+shrub species also compete with one another, but more weakly (circled
+coefficient). \textbf{B.} In spite of the competitive interactions
+between the two shrub species, their shared tendency to occur in
+locations without trees makes their occurrence vectors positively
+correlated (circled). \textbf{C.} Controlling for trees with a
+conditional (all-else-equal) approach such as a partial correlation or a
+Markov network leads to correct identification of the negative
+shrub-shrub interaction (circled). See Appendix 1 and the results for
+``three species'' for more details.
+
+\textbf{Figure 2: A.} A small Markov network, defined by its \(\alpha\)
+and \(\beta\) values. The abiotic environment favors the occurrence of
+each species (\(\alpha >0\)), particularly species 2
+(\(\alpha_2 > \alpha_1\)). The negative \(\beta_{12}\) coefficient is
+consistent with competition between the two species. \textbf{B.} The
+coefficients determine the probabilities of all four possible
+presence-absence combinations for Species 1 and Species 2. \(\alpha_1\)
+is added to the exponent whenever Species 1 is present (\(y_1 = 1\)),
+but not when it is absent (\(y_1 = 0\)). Similarly, the exponent
+includes \(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)),
+and includes \(\beta_{12}\) only when both are present (\(y_1y_2 = 1\)).
+The normalizing constant \(Z\), ensures that the four probabilities sum
+to 1. In this case, \(Z\) is about 18.5. \textbf{C.} The expected
+frequencies of all possible co-occurrence patterns between the two
+species of interest, as calculated in the previous panel. \textbf{D.}
+Without competition (i.e.~with \(\beta_{12}=0\), each species would
+occur more often.
+
+\textbf{Figure 3:} Proportion of variance in interaction coefficients
+explained by each method versus number of sampled locations across the
+three simulation types. For the null model (Pairs), two outliers with
+\(|Z|>1000\) were manually adjusted to \(|Z|=50\) to mitigate their
+detrimental influence on \(R^2\) (Appendix 5).
+
+\textbf{Figure 4: A.} The Markov network's estimated interaction
+coefficients were generally very similar to the GLM estimates.
+\textbf{B.} The null model's estimates typically matched the (negative)
+correlation coefficient, after controlling for landscape size.
+\textbf{C.} For any given interaction strength, the null model was much
+more likely to misclassify its sign with 95\% confidence than the Markov
+network was. As with the other analyses based on inferential statistics,
+this panel only shows data from the first two simulation types.
+
 \end{flushleft}
